### 2016-08-24 agda.tex
copied material from POPL submission

% !TEX root = hazelnut-popl17.tex
So far, we have given an overview of the most important judgements and rules in
the semantics of Hazelnut, and stated the critical metatheorem, Sensibility, and several auxiliary ``checks''. In a few
cases, we have informally sketched out why these metatheorems will hold. In order to formally establish that our design meets our stated objectives, we
have mechanized the semantics of Hazelnut using the Agda proof assistant \cite{norell:thesis} (also see the Agda Wiki, hosted
at \url{http://wiki.portal.chalmers.se/agda/}.)

This development is available in the supplemental material. At the time of submission, the full set of rules is available and the proofs of Theorem \ref{thrm:actsafe}, Theorem \ref{lemma:movement-erasure} and a number of technical lemmas that we elided or mentioned informally in the text are complete. We also have nearly all of the cases of Reachability and Constructability complete, and we see no major barrier to the remaining cases -- we prioritized developing a clear exposition of our contributions after proving representative cases and finding that they were straightforward. The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.

The documentation 
includes a more detailed discussion of the technical representation
decisions that we made. The main idea is standard: we encode each judgement as a
dependent type. The rules defining the judgements become the constructors of this
type, and derivations are terms of these type. This is a rich
setting that allows proofs to take advantage of pattern matching on the
shape of derivations, closely matching standard on-paper proofs. No proof automation was used, because the proof structure itself is likely to be interesting to researchers who plan to build upon our work. 
% The formalization differs from the calculus defined here in a few small 
% ways. Most interestingly, instead of giving separate movement rules for each 
% form of Z-expression, we abstract
% over different corresponding forms that happen to have the same arity, e.g. 
% additions and applications. Formalizing this intuition
% reduces the number of cases we need to consider 
% somewhat, but more importantly allows us to write a slightly more
% general calculus -- it will be easier to extend Hazelnut with more 
% interesting language features with this generic infrastructure in place.
We adopt Barendregt's convention for
bound variables \cite{urban}. Hazelnut's semantics does not need substitution, so we do not need to adopt more sophisticated encodings (e.g. \cite{lh09unibind,Pouillard11}.)

\subsection{Determinism}
A useful property would be \emph{action determinism}, i.e. that performing an action produces a unique result. Formally, this would be established as follows:
% \begin{theorem}[Action Determinism] All of the following hold:
% \label{thrm:actdet}
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
  $\zexp' = \zexp''$ and $\htau' = \htau''$.
% \item If all of

  % \begin{quote}
  %   \begin{enumerate}
  %   \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
  %   \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
  %   \item $\tcompat{\htau}{\htau'}$, and
  %   \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
  %     $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
  %   \end{enumerate}
  % \end{quote}
  % hold, then $\zexp' = \zexp''$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
  \zexp''$.
\end{enumerate}
% \end{theorem}

This is not a theorem of the system as described in this paper. The reason is quite subtle: the semantics of two construction actions, $\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the analytic case than they do in the synthetic case -- the difference has to do with what the generated ascription contains. Both cases can technically ``fire'' due to action subsumption. The mechanization includes a ``proof'' of determinism for all cases but these two. Resolving this ambiguity formally requires some rather inelegant machinery to exclude these particular base cases from action subsumption, so we are content to admonish implementors who apply action subsumption when another analytic rule would do.

### 2016-08-24 example.tex
copied material from POPL submission

% !TEX root = hazelnut-popl17.tex
The reader is encouraged to follow along with the examples that we will discuss in this section using the implementation.

%
\subsection{Example 1: Constructing the Increment Function}

Figure~\ref{fig:first-example} shows an edit sequence that constructs the increment function, of type $\tarr{\tnum}{\tnum}$, starting from the empty hole via the indicated sequence of {actions}. We will introduce Hazelnut's formal syntax and define the referenced rules in Sec. \ref{sec:hazel}. First, let us build some high-level intuitions.

%In the first task (Lines 1-9), the user constructs the identity function over numbers. In the second task (Lines 10-19), the user applies this function (assumed to be bound to a variable, $id$), to the number expression $\hnum{3}$.
% Each of these tasks is carried out interactively, through the sequence of \emph{actions} shown in the  column labeled \textbf{Next Action}. For reference, we cite the relevant rules from Sec. \ref{sec:hazel} in the final column.

% The second and third columns of the
% table show the program as it is being constructed in two forms. The second column shows it as an \textbf{H-expression}, which is an expression that can contain \emph{holes}, delimited by $\llparenthesis$ and $\rrparenthesis$. The third column shows a corresponding \textbf{Z-expression}. Z-expressions are H-expressions with a single focus on some sub-term, delimited by $\triangleright$ and $\triangleleft$. The focus need not be on a hole.
% on working on filling just one of the holes.
% Each action produces a new Z-expression, but this may or may not correspond to a new H-expression (in particular, some actions only move the focus, without changing the structure of the term.)
% to the hole in
%focus in the Z-Expression to produce the next line, which may or may not
%produce a substantively different H-Expression.

The edit state in Hazelnut is a {Z-expression}, $\zexp$. Every Z-expression has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the {cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively. For example, on Line 1, the empty expression hole, $\hhole{}$, is under the cursor.

Actions act relative to the cursor. The first action that we perform is $\aConstruct{\flam{x}}$, which instantiates the hole with a lambda abstraction binding the variable $x$. This results in the Z-expression on Line 2, consisting of a lambda abstraction with an arrow type ascription. The cursor is placed on the argument type hole. Type holes are typeset like empty expression holes for visual consistency.

The actions on Lines 2-4 complete the type ascription. In particular, the $\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and the $\aMove{\dNext}$ action moves the cursor to the next hole.

\begin{figure}[t!]
\begin{center}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} & 
%\textbf{H-Expression} & 
% \textbf{Type} & 
\textbf{Next Action} & \textbf{Rule}
\\
\hline
1 &
\zwsel{\hhole{}} & 
% \hhole{} &
% \tehole 
% &
\aConstruct{\flam{x}} & 
\text{(\ref{r:conelamhole})}
\\ 2 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} & 
% \hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
% \tarr{\tehole}{\tehole} &
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 3 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
% \tarr{\tnum}{\tehole} & 
\aMove{\dNext{}} & 
\text{({\ref{rule:move-nextSib-arr}})}
\\ 4 &
\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
&
% \text{\textquotedbl}&
% \tarr{\tnum}{\tehole} & 
\aConstruct{\fnum{}} & 
\text{(\ref{r:contnum})}
\\ 5 &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} & 
% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} &
\aMove{\dParent{}} & 
\text{(\ref{rule:move-parent-arr-right})}
\\ 6 &
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
&
% \text{\textquotedbl}&
% \tarr{\tnum}{\tnum} & 
\aMove{\dParent{}} & 
\text{(\ref{rule:move-parent-asc-right})}
\\ 7 & 
\zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} & 
\aMove{\dChild} & 
\text{(\ref{r:movefirstchild})}
\\ 8 &
% &
\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} & 
% \tarr{\tnum}{\tnum} &
% &
\aMove{\dChild{}} & 
\text{(\ref{r:movefirstchild-lam})}
\\ 9 &
% &
\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} & 
\aConstruct{\fvar{x}} & 
\text{(\ref{r:conevar})}
\\ 10 &
% \hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
% &
\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} & 
{\aConstruct{\fplus}}
&
\text{(\ref{rule:construct-plus-compat})}
\\ 11 & 
\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} & 
\aConstruct{\fnumlit{1}} & 
\text{(\ref{r:conenumnum})}
\\ 12 & 
\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} & 
\textrm{---} & 
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Constructing an increment function in Hazelnut.}
\label{fig:first-example}
\end{figure}

The actions on Lines 5-8 move the cursor to the function body. For simplicity, Hazelnut defines only a small collection of primitive movement actions. In practice, an editor would also define compound movement actions, e.g. an action that moves the cursor directly to the next remaining hole, in terms of these primitive movement actions.

Lines 9-11 complete the function body by first constructing the variable $x$, then  constructing the plus form, and finally constructing the number $\hnum{1}$. Notice that we did not need to construct the function body in an ``outside-in'' manner, i.e. we constructed $x$ before constructing the outer plus form. Luckily, the transient function bodies, $x$ and $x + \hhole{}$, could be checked against type $\tnum$ (as we will detail in Sec. \ref{sec:holes}.)

\subsection{Example 2: Applying the Increment Function}

\begin{figure}[t!]
\begin{center}
\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{54px}}
% \vspace{-6px}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
% \\
\hline
\# & \textbf{Z-Expression} & 
%\textbf{H-Expression} & 
% \textbf{Type} & 
\textbf{Next Action} & \textbf{Rule}
\\
\hline
13 &
\zwsel{\hhole{}} & 
% \hhole{} &
% \tehole 
% &
\aConstruct{\fvar{incr}} \hphantom{~\,\,~~}& 
\text{(\ref{r:conevar})}
\\ 14 & 
\zwsel{incr} & 
\aConstruct{\fap} & 
\text{(\ref{r:coneapfn})}
\\ 15 & 
incr(\zwsel{\hhole{}}) & 
\aConstruct{\fvar{incr}} & 
\text{(\ref{r:conevar2})}
\\ 16 & 
incr(\hhole{\zwsel{incr}}) & 
\aConstruct{\fap} & 
\text{(\ref{r:coneapfn})}
\\ 17 & 
incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} & 
\aConstruct{\fnumlit{3}} & 
\text{(\ref{r:conenumnum})}
\\ 18 & 
incr(\hhole{incr(\zwsel{\hnum{3}})}) & 
\aMove{\dParent}& 
\text{(\ref{r:moveparent-ap2})}
\\ 19 & 
incr(\hhole{\zwsel{incr(\hnum{3})}}) & 
\aMove{\dParent} & 
\text{(\ref{r:moveparent-hole})}
\\ 20 &
incr(\zwsel{\hhole{incr(\hnum{3})}})& 
\aFinish & 
\text{(\ref{r:finishana})}
\\ 21 & 
incr(\zwsel{incr(\hnum{3})}) & 
\textrm{---} & 
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Appyling the increment function.}
\label{fig:second-example}
\end{figure}

Figure \ref{fig:second-example} shows an edit sequence that constructs the expression $incr(incr(\hnum{3}))$, where $incr$ is assumed bound to the increment function from Figure \ref{fig:first-example}.

We begin on Line 13 by constructing the variable $incr$. Line 14 then constructs the application form with $incr$ in function position, leaving the cursor on a hole in the argument position. Notice that we did not need to construct the outer application form before identifying the function being applied. Again, the transient edit state happens to be well-typed, so we needed to make no special allowances for this order of actions.

We now need to apply $incr$ again, so we perform the same action on Line 15 as we did on Line 13, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic structure editor, performing such an action would result in the following edit state: 
\[
incr(\zwsel{incr})
\]
This edit state is ill-typed (after \emph{cursor erasure}): the argument of $incr$ must be of type $\tnum$ but here it is of type $\tarr{\tnum}{\tnum}$. Hazelnut cannot allow such an edit state to arise.

The programmer could alternatively have performed the $\aConstruct{\fap}$ action on Line 15. This would result in the following edit state, which is well-typed according to the static semantics that we will define in the next section:
\[
incr(\hhole{}(\zwsel{\hhole{}}))
\]
The problem is that the programmer is not able to identify the intended function before constructing the function application form. This stands in contrast to Lines 13-14. % This is qualitatively unnatural.

Hazelnut's action semantics addresses this problem: rather than disallowing the $\aConstruct{\fvar{incr}}$ action on Line 15, it leaves $incr$ inside a hole:
\[
incr(\hhole{\zwsel{incr}})
\]
This defers the type consistency check, exactly as an empty hole in the same position does. One way to think about non-empty holes is as an internalization of the ``squiggly underline'' that text or syntactic structure editors display to indicate a type inconsistency. By internalizing this concept, the presence of a type inconsistency does not leave the entire program formally meaningless.

The expression inside a non-empty hole must itself be well-typed, so the programmer can continue to edit it. Lines 16-17 proceed to apply the inner mention of $incr$ to a number literal, $\hnum{3}$. Finally, Lines 18-19 move the cursor to the non-empty hole and Line 20 performs the $\aFinish$ action. The $\aFinish$ action removes the hole if the type of the expression inside the hole is consistent with the expected type, as it now is. This results in the final edit state on Line 20, as desired. In practice, the editor might automatically perform the $\aFinish$ action as soon as it becomes possible, but for simplicity, Hazelnut formally requires that it be performed explicitly.

% So far, editing has proceeded in an essentially type-directed, outside-in fashion -- the user first specified the type of the function, then produced a body of that type by the action on Line 8. Lines 10-12 similarly begin in a type-directed manner with the user giving an explicit type ascription, indicating that the expression that they are constructing will have type $\tnum$.

% However, on Line 13, the user performs the $\aConstruct{\fvar{id}}$ action. Notice that $id$ has type $\tarr{\tnum}{\tnum}$, which is not consistent with the type $\tnum$ given in the ascription. Na\"ively, this would produce a type error, leaving the program in a well-formed but semantically undefined state. One way to avoid this state is to simply not make this action available in the program configuration on Line 12. This is inflexible, forcing an outside-in approach to program construction (i.e. the user would need to construct the function application form before constructing the variable $id$.) Instead, Hazelnut permits this action, but places the variable $id$ inside a hole. This defers the consistency check that would normally occur: a hole can be checked against any type, as long as its contents have some type. The cursor is placed inside the hole. The user then proceeds to apply $id$ to the number expression $\hnum{3}$. At this point, the expression inside the hole has a type consistent with the ascription, so the user can \emph{finish} the hole. In our simple formalism, this requires moving the cursor to the hole (in practice, the system might find the nearest parent of hole form.) The result is the complete, well-typed program shown on Line 19 (notice that \emph{complete} is distinct from \emph{closed} -- the variable $id$ is free on Line 19, so this is not a closed program.)

%% The third column~(\textbf{Next Action}) lists the first user action:
%% Constructing a lambda abstraction using variable~$x$.
%% %
%% The final column~(\textbf{Semantics}) indicates the semantic rule for this
%% action, Rule (\ref{r:conelamhole}), which gives general semantics for
%% introducing lambda terms into holes.
%% %
%% In Section~\ref{sec:hazel}, we list this rule, and the other rules used in
%% this final column. In total, these rules give a formal semantics to the
%% user actions, which relate each line's Z-Expression to the Z-Expression on
%% the subsequent line.

%% In addition to introducing the lambda term, and its variable, the
%% first user action~$\aConstruct{\flam{x}}$ also introduces a type
%% ascription for this function, as an arrow type, with holes for the
%% type of its domain and codomain.
%% %
%% The actions for Lines~2--5 consist of the user filling these holes
%% with the basetype $\tnum{}$.
%% %
%% To do so, the user constructs the type constructor twice (Lines 2 and
%% 4), and navigates between the holes with a move action (Line~3).
%% %
%% Generally, the move action~$\dNext$ moves the focus from one
%% sub-structure to the next sibling sub-structure of the (common) parent
%% structure; in this case, it moves from the domain type of the arrow
%% type to the codomain of the arrow type.
%% %

### 2016-08-24 formally.tex
copied material from POPL submission

% !TEX root = hazelnut-popl17.tex

The previous section introduced Hazelnut by example. In this section, we systematically introduce the following  structures:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are types and expressions with {holes}. H-types classify H-expressions according to Hazelnut's \textbf{bidirectional static semantics}.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose\- a \emph{cursor} onto H-types and H-expressions, respectively (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Every Z-type (resp. Z-expression) corresponds to an H-type (resp. H-expression) by \emph{cursor erasure}.
\item \textbf{Actions} (Sec. \ref{sec:actions}), which act relative to the cursor according to Hazelnut's \textbf{bidirectional action semantics}. The action semantics enjoys a rich metatheory. Of particular note, the \emph{sensibility theorem} establishes that every edit state is well-typed after cursor erasure.
\end{itemize}

Our overview below omits certain ``uninteresting'' details. The supplement includes the complete collection of rules, in definitional order. These rules, along with the proofs of two key metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement. We will summarize this effort in Sec. \ref{sec:mech}.% The supplement contains the Agda sources and additional technical details.

\subsection{H-types and H-expressions}\label{sec:holes}
Figure \ref{fig:hexp-syntax} defines the syntax of H-types, $\htau$, and H-expressions, $\hexp$. Most forms correspond directly to those of the simply typed lambda calculus (STLC) extended with a single base type, $\tnum$, of numbers (cf. \cite{pfpl}.) The number expression corresponding to the mathematical number $n$ is drawn $\hnum{n}$, and for simplicity, we define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$. The form $\hexp : \htau$ is an explicit \emph{type ascription}.

In addition to these standard forms, \emph{type holes} and \emph{empty expression holes} are both drawn $\hehole$ and \emph{non-empty expression holes} are drawn $\hhole{\hexp}$. We do not need non-empty type holes because every H-type is a valid classifier of H-expressions.

Types and expressions that contain no holes are \emph{complete types} and \emph{complete expressions}, respectively. Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete (see supplement.) We are not concerned here with defining a dynamic semantics for Hazelnut, but the dynamics for complete H-expressions would be entirely standard (in Sec. \ref{sec:rw} we discuss the multi-dimensional design space around the question of evaluating incomplete H-expressions.)% The complete types and expressions correspond immediately to the types and expressions of the STLC with numbers. %Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.

\begin{figure}[t]
$\arraycolsep=4pt\begin{array}{lllllll}
\mathsf{HTyp} & \htau & ::= &
  \tarr{\htau}{\htau} ~\vert~
  \tnum ~\vert~
  \tehole\\
\mathsf{HExp} & \hexp & ::= &
  \hexp : \htau ~\vert~
  x ~\vert~
  \hlam{x}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  \hnum{n} ~\vert~
  \hadd{\hexp}{\hexp} ~\vert~
  \hehole ~\vert~
  \hhole{\hexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
\label{fig:hexp-syntax}
\end{figure}
\begin{figure}
\noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\tau$ and $\tau'$ are consistent}
% \begin{subequations}%\label{rules:tcompat}
% \begin{equation}%\label{rule:tcompat-comm}
% \inferrule
% %[TCSym]
% {
%   \tcompat{\htau}{\htau'}
% }{
%   \tcompat{\htau'}{\htau}
% }
% \end{equation}
\begin{subequations}\label{eqns:consistency}
\begin{mathpar}
% \begin{equation}
\inferrule{ }{
  \tcompat{\tehole}{\htau}
}%~\text{(1a)}
% \end{equation}
% \end{minipage}
% \begin{minipage}{0.15\linewidth}
% \begin{equation}%\label{rule:tcompat-hole}
~~~~~~~~
\inferrule{ }{
  \tcompat{\htau}{\tehole}
}%~\text{(1b)}
% \end{equation}
% \end{minipage}
% \begin{minipage}{0.15\linewidth}
% \begin{equation}
~~~~~~~~
\inferrule{ }{
  \tcompat{\htau}{\htau}
}%~\text{(1c)}
% \end{equation}
% \end{minipage}
% \begin{minipage}{0.15\linewidth}
% \begin{equation}
~~~~~~~~
\inferrule{
  \tcompat{\htau_1}{\htau_1'}\\
  \tcompat{\htau_2}{\htau_2'}
}{
  \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
}~~~~~~~~\text{(\ref*{eqns:consistency}a-d)}
% \end{equation}
% \end{minipage}
\end{mathpar}
\end{subequations}
% \begin{equation}%\label{rule:tcompat-num}
% \end{equation}
% \begin{equation}%\label{rule:tcompat-arr}
% \end{equation}
% \end{subequations}
\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\tau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
\begin{subequations}
\begin{minipage}{0.5\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation}
\end{minipage}
\begin{minipage}{0.5\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
}
\end{equation}
\end{minipage}
% \end{subequations}
% \end{mathpar}
% \noindent\fbox{$\tincompat{\htau}{\htau'}$}
% \begin{subequations}
%   % \begin{equation}
%   %   \inferrule{
%   %     \tincompat{\htau}{\htau'}
%   %   }{
%   %     \tincompat{\htau'}{\htau}
%   %   }
%   % \end{equation}
%   \begin{equation}
%     \inferrule{ }{
%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tnum}
%     }
%   \end{equation}
%   \begin{equation}
%     \inferrule{ }{
%       \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
%     }
%   \end{equation}
%   \begin{equation}
%     \inferrule{
%       \tincompat{\htau_1}{\htau_1'}
%     }{
%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
%     }
%   \end{equation}
%   \begin{equation}
%     \inferrule{
%       \tincompat{\htau_2}{\htau_2'}
%     }{
%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
%     }
%   \end{equation}
\end{subequations}
\caption{H-type consistency and matched arrow types.}
\label{fig:type-consistency}
\end{figure}

\begin{figure}
\fbox{$\hana{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ analyzes against $\htau$}
\begin{subequations}
\begin{equation}\label{rule:ana-subsume}
\inferrule{
  \hsyn{\hGamma}{\hexp}{\htau'}\\
  \tcompat{\htau}{\htau'}
}{
  \hana{\hGamma}{\hexp}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-lam}
\inferrule{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
}
\end{equation}
\end{subequations}
\fbox{$\hsyn{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ synthesizes $\htau$}
\begin{subequations}
\begin{equation}\label{rule:syn-asc}
\inferrule{
  \hana{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hexp : \htau}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-var}
\inferrule{ }{
  \hsyn{\hGamma, x : \htau}{x}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-ap}
\inferrule{
  \hsyn{\hGamma}{\hexp_1}{\htau}\\
  \arrmatch{\htau}{\tarr{\htau_2}{\htau'}}\\
  \hana{\hGamma}{\hexp_2}{\htau_2}
}{
  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau'}
}
\end{equation}
\begin{equation}\label{rule:syn-num}
\inferrule{ }{
  \hsyn{\hGamma}{\hnum{n}}{\tnum}
}
\end{equation}
\begin{equation}\label{rule:syn-plus}
\inferrule{
  \hana{\hGamma}{\hexp_1}{\tnum}\\
  \hana{\hGamma}{\hexp_2}{\tnum}
}{
  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
}
\end{equation}
\begin{equation}\label{rule:syn-ehole}
\inferrule{ }{
  \hsyn{\hGamma}{\hehole}{\tehole}
}
\end{equation}
\begin{equation}\label{rule:syn-hole}
\inferrule{
  \hsyn{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
}
\end{equation}
\end{subequations}
\caption{Analysis and synthesis.}
\label{fig:ana-syn}
\end{figure}

Hazelnut's static semantics is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial} around the two mutually defined judgements in Figure \ref{fig:ana-syn}. Typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner. We identify the context up to exchange and adopt the standard identification convention for formal structures that differ only up to alpha-equivalence.

Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, establish that $\hexp$ can appear where an expression of type $\htau$ is expected. Derivations of the type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, infer a type from $\hexp$, which is necessary in positions where an expected type is not available (e.g. at the top level.) Algorithmically, the type is an ``input'' of the type analysis judgement, but an ``output'' of the type synthesis judgement. %The rules describe a \emph{local type inference} scheme, i.e. type ascriptions are unnecessary when an expression is being analyzed against a known type.
Making a judgemental distinction between these two notions will be essential in our action semantics (Sec. \ref{sec:actions}.)

 %We use the metavariable $\Gamma$ for \emph{complete typing contexts}, i.e. typing contexts where each hypothesis mentions only complete types.

\begin{subequations}\label{rules:syn-ana}
Type synthesis is stronger than type analysis in that if an expression is able to synthesize a type, it can also be analyzed against that type, or any other \emph{consistent} type, according to the \emph{subsumption rule}, Rule (\ref{rule:ana-subsume}).

The \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$, that appears as a premise in the subsumption rule is a reflexive and symmetric (but not transitive) relation between H-types defined judgementally in Figure \ref{fig:type-consistency}. This relation coincides with equality for complete H-types. Two incomplete H-types are consistent if they differ only at positions where a hole appears in either type. The type hole is therefore consistent with every type. This notion of H-type consistency coincides with the notion of type consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ (i.e. unknown) type \cite{Siek06a}.

Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions, $\hlam{x}{\hexp}$. There is no type synthesis rule that applies to this form, so lambda abstractions can appear only in analytic position, i.e. where an expected type is known.\footnote{It is possible to also define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.} Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation*}
The problem is that this standard rule alone leaves us unable to analyze lambda abstractions against the type hole, because the type hole is not immediately of the form $\tarr{\htau_1}{\htau_2}$. There are two plausible solutions to this problem. One solution would be to define a second rule specifically for this case:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \tehole}{\hexp}{\tehole}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tehole}
}
\end{equation*}
Intead, we combine these two possible rules into a single rule through the simple auxiliary \emph{matched arrow type} judgement, $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure \ref{fig:type-consistency}. This judgement leaves arrow types alone and assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$. It is easy to see that the two rules above are admissible by appeal to Rule (\ref{rule:syn-lam}) and the matched arrow type judgement. Encouragingly, this judgement also arises in the study of gradual type systems \cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.

Rule (\ref{rule:syn-asc}) defines type synthesis of expressions of ascription form, $\hexp : \htau$. This allows the user to explicitly state  a type for the ascribed expression to be analyzed against.

Rule (\ref{rule:syn-var}) is the standard rule for variables.

Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function application. It also makes use of the matched function type judgement to combine what would otherwise need to be two rules for function application -- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$ synthesizes the type hole. Indeed, Siek and Taha needed two application rules for the same fundamental reason \cite{Siek06a}. Later work on gradual typing introduced this notion of type matching to resolve this redundancy.

Rule (\ref{rule:syn-num}) states that numbers synthesize the $\tnum$ type. Rule (\ref{rule:syn-plus}) states that $\hexp_1 + \hexp_2$ behaves like a function over numbers.

The rules described so far are sufficient to type complete H-expressions. The two remaining rules give H-expressions with holes a well-defined static semantics.

Rule (\ref{rule:syn-ehole}) states that the empty expression hole synthesizes the type hole.

A non-empty hole contains an H-expression that is ``under construction'', as described in Sec. \ref{sec:example}. According to Rule (\ref{rule:syn-hole}), this inner  expression must synthesize some type, but, like the empty hole,  non-empty holes synthesize the type hole.

% \begin{equation}\label{rule:tcompat-comm}
% \inferrule{
%   \tcompat{\htau}{\htau'}
% }{
%   \tcompat{\htau'}{\htau}
% }
% \end{equation}
% \begin{equation}\label{rule:tcompat-num}
% \inferrule{ }{
%   \tcompat{\tnum}{\tnum}
% }
% \end{equation}
% \begin{equation}\label{rule:tcompat-arr}
% \inferrule{
%   \tcompat{\htau_1}{\htau_1'}\\
%   \tcompat{\htau_2}{\htau_2'}
% }{
%   \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
% }
% \end{equation}

Given these rules, it is instructive to derive the following (by subsumption):
\[\hana{incr : \tarr{\tnum}{\tnum}}{\hhole{incr}}{\tnum}\]
% This is the key premise necessary to synthesize the $\tnum$ type for the H-expression that results from cursor erasure -- defined next -- of the Z-expression on Line 16 of Fig. \ref{fig:second-example}. %In other words, this mechanism is essential if  users are to able to construct a program in anything but an ``outside in'' fashion.

% The final rule handles function applications where the expression in function position synthesizes a hole type, rather than an arrow type. We treat it as if it had instead synthesized $\tarr{\tehole}{\tehole}$:
% \begin{equation}\label{rule:syn-ap-2}
% \inferrule{
%   \hsyn{\hGamma}{\hexp_1}{\tehole}\\
%   \hana{\hGamma}{\hexp_2}{\tehole}
% }{
%   \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}
% }
% \end{equation}

% \todo{redundant text w/ citation}
% The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}. Their system (which was not bidirectionally typed nor an editor model) also needed to define two rules for function application. In general, when a premise requires that a synthesized type be of a particular form, we need a special case where the synthesized hole type is treated instead as if it were the ``holey-est'' type of that form.\footnote{Alternatively, we might add a rule that allows expressions that synthesize hole type to then non-deterministically synthesize any other type, but maintaining determinism is useful in practice, so we avoid this approach.}

\end{subequations}
\subsection{Z-types and Z-expressions}\label{sec:cursors}
\newcommand{\cvert}{{\,{\vert}\,}}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=2pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} \cvert
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} \cvert
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} \cvert
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau \cvert
  \hexp : \ztau \cvert
  \hlam{x}{\zexp} \cvert
  \hap{\zexp}{\hexp} \cvert
  \hap{\hexp}{\zexp} \cvert
  \hadd{\zexp}{\hexp} \cvert
  \hadd{\hexp}{\zexp} \cvert
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions.}
\label{fig:zexp-syntax}
\end{figure}

Hazelnut's action semantics operates over Z-types, $\ztau$, and Z-expressions, $\zexp$. Figure \ref{fig:zexp-syntax} defines the syntax of Z-types and Z-expressions. A Z-type (resp. Z-expression) represents an H-type (resp. H-expression) with a single superimposed \emph{cursor}.

The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on. All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the cursor will be found. Any other sub-term is ``dotted'', i.e. it is either an H-type or H-expression. Taken together, every Z-type and Z-expression contains exactly one selected H-type or H-expression by construction. This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)

We write $\removeSel{\ztau}$ for the H-type constructed by erasing the cursor from $\ztau$, which we refer to as the \emph{cursor erasure} of $\ztau$. This straightforward metafunction is defined as follows:
\begin{align*}
%\removeSel{(\zlsel{\htau})} & = \htau\\
\removeSel{(\zwsel{\htau})} & = \htau\\
%\removeSel{(\zrsel{\htau})} & = \htau\\
\removeSel{(\tarr{\ztau}{\htau})} & = \tarr{\removeSel{\ztau}}{\htau}\\
\removeSel{(\tarr{\htau}{\ztau})} & = \tarr{\htau}{\removeSel{\ztau}}
\end{align*}

Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by erasing the cursor from the Z-expression $\zexp$, i.e. the cursor erasure of $\zexp$. The definition of this metafunction is analogous, so we omit it for concision (see supplement.)
% \begin{align*}
% %\removeSel{(\zlsel{\hexp})} & = \hexp\\
% \removeSel{(\zwsel{\hexp})} & = \hexp\\
% %\removeSel{(\zrsel{\hexp})} & = \hexp\\
% \removeSel{(\zexp : \htau)} & = \removeSel{\zexp} : \htau\\
% \removeSel{(\hexp : \ztau)} & = \hexp : \removeSel{\ztau}\\
% \removeSel{(\hlam{x}{\zexp})} & = \hlam{x}{\removeSel{\zexp}}\\ogoo
% \removeSel{(\hap{\zexp}{\hexp})} & = \hap{\removeSel{\zexp}}{\hexp}\\
% \removeSel{(\hap{\hexp}{\zexp})} & = \hap{\hexp}{\removeSel{\zexp}}\\
% \removeSel{(\hadd{\zexp}{\hexp})} & = \hadd{\removeSel{\zexp}}{\hexp}\\
% \removeSel{(\hadd{\hexp}{\zexp})} & = \hadd{\hexp}{\removeSel{\zexp}}\\
% \removeSel{\hhole{\zexp}} &= \hhole{\removeSel{\zexp}}
% \end{align*}

\subsection{Actions}\label{sec:actions}

We now arrive at the heart of Hazelnut: its \emph{bidirectional action semantics}.  
Figure \ref{fig:action-syntax} defines the syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, and \emph{shapes}, $\psi$.

Expression actions are governed by two mutually defined judgements, 1) the \emph{synthetic action judgement}:
\[
\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}
\]
and 2) \emph{the analytic action judgement}:
\[
\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
\]

In some Z-expressions, the cursor is in a type ascription, so we also need a \emph{type action judgement}, pronounced ``performing $\alpha$ on $\ztau$ results in $\ztau'$'':
\[
\performTyp{\ztau}{\alpha}{\ztau'}
\]

% The type action judgement appears as a premise in the rules that handle Z-expressions where the cursor is inside a . 
% \[\begin{array}{ll}
% %\textbf{Judgement Form} & \textbf{Description}\\
% & 
% \end{array}\]
% \[
% \begin{array}{l}
% \fbox{$\performTyp{\ztau}{\alpha}{\ztau'}$} ~~ \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
% \fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$} \\
% \quad \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
% \quad \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
% \fbox{$\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$} \\
% \quad \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
% \quad \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
% \quad \text{against $\htau$}
% \end{array}\]
\subsubsection{Sensibility}

Before giving the rules for these judgements, let us state the key metatheorem: \emph{sensibility}. This metatheorem deeply informs the design of the rules, given starting in Sec. \ref{sec:action-subsumption}. Its proof, which is mechanized in Agda, is by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules. 
\begin{theorem}[Action Sensibility]\label{thrm:actsafe} Both of the following hold:\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ 
   then
  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{enumerate}
\end{theorem}
\noindent In other words, if an edit state (i.e. a Z-expression) is statically meaningful, i.e. its cursor erasure is well-typed, then performing an action on it leaves the resulting edit state statically meaningful. In particular, the first clause of Theorem \ref{thrm:actsafe} establishes that when an action is performed on an edit state whose cursor erasure synthesizes an H-type, the result is an edit state whose cursor erasure also synthesizes some (possibly different) H-type. The second clause establishes that when an action is performed using the analytic action judgement on an edit state whose cursor erasure analyzes against some H-type, the result is a Z-expression whose cursor erasure also analyzes against the same H-type. % Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.

\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{llcllll}
\mathsf{Action} & \alpha & ::= &
  \aMove{\delta} ~\vert~
  %\aSelect{\delta} ~\vert~
  \aDel ~\vert~
  %\aReplace{\htau} ~\vert~
  %\aReplace{\hexp} ~\vert~
  \aConstruct{\psi} ~\vert~
  \aFinish\\
\mathsf{Dir} & \delta & ::= &
  \dChild ~\vert~
  \dParent ~\vert~
  \dNext \\ % ~\vert~
  % \dPrev\\
\mathsf{Shape} & \psi & ::= &
  \farr ~\vert~
  \fnum \\
& & \vert &
  \fasc ~\vert~
  \fvar{x} ~\vert~
  \flam{x} ~\vert~
  \fap ~\vert~
  \farg ~\vert~
  \fnumlit{n} ~\vert~
  \fplus\\
& & \vert & 
  {\color{gray}\fnehole}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of actions.}
\label{fig:action-syntax}
\end{figure}

% \subsubsection{Determinism}
% Another useful property that we maintain is \emph{action determinism}, i.e. that performing an action should produce a unique result. Formally, this is established as follows:
% \begin{theorem}[Action Determinism] All of the following hold:
% \label{thrm:actdet}
% \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
% \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
% \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
%   $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
%   $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
%   $\zexp' = \zexp''$ and $\htau' = \htau''$.
% % \item If all of

%   \begin{quote}
%     \begin{enumerate}
%     \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
%     \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
%     \item $\tcompat{\htau}{\htau'}$, and
%     \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
%       $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
%     \end{enumerate}
%   \end{quote}
%   hold, then $\zexp' = \zexp''$.
% % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
% %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
% %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
% %   \zexp''$.
% % \end{enumerate}
% \end{theorem}

\subsubsection{Type Inconsistency}
In some of the rules below, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can define this notion either directly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement. The supplement does the latter. The key rule establishes that arrow types are inconsistent with $\tnum$:
% \begin{subequations}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau}{\htau'}
  %   }{
  %     \tincompat{\htau'}{\htau}
  %   }
  % \end{equation}
  \begin{equation*}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation*}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_1}{\htau_1'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_2}{\htau_2'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
% \end{subequations}
% The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.

\subsubsection{Action Subsumption}\label{sec:action-subsumption}

The action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
\begin{equation}\label{rule:action-subsume}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}%\\\\
    % \zexp, \alpha~\mathsf{subsumes}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.  The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e. Rule (\ref{rule:ana-subsume}). Algorithmically, subsumption should be the rule of last resort.

% We need to make some exceptions in this rule to maintain determinism, because certain construction actions, described in Sec. \ref{sec:construction}, behave slightly differently depending on whether a type is known. We do so by defining a simple judgement $\alpha~\mathsf{subsumes}$ that excludes these action forms:\todo{put into supp}
% \begin{subequations}
% \begin{equation}\label{rule:subsumes}
% \inferrule{
%   \alpha \neq \aConstruct{\fasc}\\
%   (\forall x) \alpha \neq \aConstruct{\flam{x}} 
% }{
%   \zexp, \alpha~\mathsf{subsumes}
% }
% \end{equation}
% \begin{equation}\label{rule:subsumes}
% \inferrule{
%   \zexp \neq \zwsel{\hexp}
% }{
%   \zexp, \alpha~\mathsf{subsumes}
% }
% \end{equation}
% \end{subequations}
% Algorithmically, this captures the notion that subsumption should be the case of last resort.

\subsubsection{Relative Movement}\label{sec:movement} 
The rules below define relative movement within Z-types. They should be self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }{
      \aMove{\dChild}
    }{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-parent-arr-right}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-nextSib-arr}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{{\htau_2}}
    }{
      \aMove{\dNext}
    }{
      {\tarr{\htau_1}{\zwsel{\htau_2}}}
    }
  }
\end{equation}
% \begin{equation}
%   \inferrule{ }{
%     \performTyp{
%       \tarr{{\htau_1}}{\zwsel{\htau_2}}
%     }{
%       \aMove{\dPrev}
%     }{
%       {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
%     }
%   }
% \end{equation}
% \begin{equation}
% \inferrule{
%   \performTyp{
%     \ztau
%   }{
%     \aMove{\delta}
%   }{
%     \ztau'
%   }
% }{
%   \performTyp{
%     \tarr{\ztau}{\htau}
%   }{
%     \aMove{\delta}
%   }{
%     \tarr{\ztau'}{\htau}
%   }
% }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{
%       \ztau
%     }{
%       \aMove{\delta}
%     }{
%       \ztau'
%     }
%   }{
%     \performTyp{
%       \tarr{\htau}{\ztau}
%     }{
%       \aMove{\delta}
%     }{
%       \tarr{\htau}{\ztau}
%     }
%   }
% \end{equation}
\end{subequations}
Two more rules are needed to recurse into the zipper structure. We define these zipper rules in an action-independent manner in Sec. \ref{sec:zipper-cases}. 
% The final two rules above recurse into the zipper structure.

The rules for relative movement within Z-expressions are similarly straightforward. Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
% \begin{figure}
% \judgbox{\performMove{\zexp}{\aMove{\delta}}{\zexp'}}{(Additional movement-action rules:)}
% \begin{displaymath}
% \begin{array}{@{}rcl}
%   %\multicolumn{3}{l}{\textbf{Movement Actions for Sum Type Forms}:}
%   %\\
%   \TABperformMove
%       {\zwsel{\hexp : \htau}}
%       {\aMove{\dChild}}
%       {      \zwsel{\hexp} : \htau}
%   \\
%   \TABperformMove
%       {
%       \zwsel{\hexp} : \htau
%     }{
%       \aMove{\dParent}
%     }{
%       \zwsel{\hexp : \htau}
%     }
%   \\
%   \TABperformMove
%     {
%       \hexp : \zwsel{\htau}
%     }{
%       \aMove{\dParent}
%     }{
%       \zwsel{\hexp : \htau}
%     }
%   \\
%   \TABperformMove
%       {
%       \zwsel{\hexp} : \htau
%     }{
%       \aMove{\dNext}
%     }{
%       \hexp : \zwsel{\htau}
%     }
%   \\[2mm]
%   \TABperformMove
%       {      {\tsum{{\htau_1}}{{\zwsel{\htau_2}}}}}
%       {\aMove{\dParent}}
%       {\zwsel{\tsum{{\htau_1}}{\htau_2}}}
%   \\[2mm]
%   \TABperformMove
%       {\zwsel{\hinj{i}{\hexp}}}
%       {\aMove{\dChild}}
%       {\hinj{i}{\zwsel{\hexp}}}      
%   \\
%   \TABperformMove
%       {\hinj{i}{\zwsel{\hexp}}}      
%       {\aMove{\dParent}}
%       {\zwsel{\hinj{i}{\hexp}}}
%   \\[2mm]
%   \TABperformMove
%       {\zwsel{\hcase{\hexp}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dChild}}
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dNext}}
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dNext}}
%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
%       {\aMove{\dPrev}}
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dPrev}}
%       {      {\hcase{{\zwsel{\hexp}}}{x}{{\hexp_1}}{y}{{\hexp_2}}}}
%   \\
%   \TABperformMove
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\hexp_1}{y}{\zwsel{\hexp_2}}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
% \end{array}
% \end{displaymath}
% \caption{Movement action semantics for sum type forms.}
% \end{figure}
The expression movement judgement is defined as follows.

\paragraph{Ascription}
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-parent-asc-right}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
% \begin{equation}
%   \label{r:moveprevsib}
%   \inferrule{ }{
%     \performTyp{
%       \hexp : \zwsel{\htau}
%     }{
%       \aMove{\dPrev}
%     }{
%       \zwsel{\hexp} : \htau
%     }
%   }
% \end{equation}
% \begin{equation}
% \inferrule{
%   \performTyp{
%     \zexp
%   }{
%     \aMove{\delta}
%   }{
%     \zexp'
%   }
% }{
%   \performTyp{
%     \zexp : \htau
%   }{
%     \aMove{\delta}
%   }{
%     \zexp' : \htau
%   }
% }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{
%       \ztau
%     }{
%       \aMove{\delta}
%     }{
%       \ztau'
%     }
%   }{
%     \performTyp{
%       \hexp : \ztau
%     }{
%       \aMove{\delta}
%     }{
%       \hexp : \ztau'
%     }
%   }
% \end{equation}

\paragraph{Lambda}\vspace{-3px}
\begin{equation}\label{r:movefirstchild-lam}
\inferrule{ }{
  \performMove{
    \zwsel{\hlam{x}{\hexp}}
  }{
    \aMove{\dChild}
  }{
    \hlam{x}{\zwsel{\hexp}}
  }
}
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performMove{
      \hlam{x}{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hlam{x}{\hexp}}
    }
  }
\end{equation}
\paragraph{Application}\vspace{-5px}
\begin{equation}
  \inferrule{ }{
    \performMove{
      \zwsel{\hap{\hexp_1}{\hexp_2}}
    }{
      \aMove{\dChild}
    }{
      \hap{\zwsel{\hexp_1}}{\hexp_2}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performMove{
      \hap{\zwsel{\hexp_1}}{\hexp_2}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hap{\hexp_1}{\hexp_2}}
    }
  }
\end{equation}
\begin{equation}\label{r:moveparent-ap2}
  \inferrule{ }{
    \performMove{
      \hap{{\hexp_1}}{\zwsel{\hexp_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hap{\hexp_1}{\hexp_2}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performMove{
      \hap{\zwsel{\hexp_1}}{\hexp_2}
    }{
      \aMove{\dNext}
    }{
      \hap{\hexp_1}{\zwsel{\hexp_2}}
    }
  }
\end{equation}
% \begin{equation}
%   \inferrule{ }{
%     \performMove{
%       \hap{\hexp_1}{\zwsel{\hexp_2}}
%     }{
%       \aMove{\dPrev}
%     }{
%       \hap{\zwsel{\hexp_1}}{\hexp_2}
%     }
%   }
% \end{equation}

\paragraph{Plus}
\begin{equation}
  \inferrule{ }{
    \performMove{
      \zwsel{\hadd{\hexp_1}{\hexp_2}}
    }{
      \aMove{\dChild}
    }{
      \hadd{\zwsel{\hexp_1}}{\hexp_2}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performMove{
      \hadd{\zwsel{\hexp_1}}{\hexp_2}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hadd{\hexp_1}{\hexp_2}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performMove{
      \hadd{{\hexp_1}}{\zwsel{\hexp_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hadd{\hexp_1}{\hexp_2}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performMove{
      \hadd{\zwsel{\hexp_1}}{\hexp_2}
    }{
      \aMove{\dNext}
    }{
      \hadd{\hexp_1}{\zwsel{\hexp_2}}
    }
  }
\end{equation}
% \begin{equation}
%   \inferrule{ }{
%     \performMove{
%       \hadd{\hexp_1}{\zwsel{\hexp_2}}
%     }{
%       \aMove{\dPrev}
%     }{
%       \hadd{\zwsel{\hexp_1}}{\hexp_2}
%     }
%   }
% \end{equation}

\paragraph{Non-Empty Hole}
\begin{equation}
\inferrule{ }{
  \performMove{
    \zwsel{\hhole{\hexp}}
  }{
    \aMove{\dChild}
  }{
    \hhole{\zwsel{\hexp}}
  }
}
\end{equation}
\begin{equation}\label{r:moveparent-hole}
  \inferrule{ }{
    \performMove{
      \hhole{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hhole{\hexp}}
    }
  }
\end{equation}
Again, additional rules are needed to recurse into the zipper structure, but we will define these zipper rules in an action-independent manner in Sec. \ref{sec:zipper-cases}. 
\end{subequations}

The rules above are numerous and fairly uninteresting. That makes them quite hazardous -- we might make a mistake absent-mindedly. One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.

\begin{theorem}[Movement Erasure Invariance]\label{lemma:movement-erasure} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px] 
\item If $\performMove{\ztau}{\aMove{\delta}}{\ztau'}$ then $\removeSel{\ztau}=\removeSel{\ztau'}$.
\item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then $\removeSel{\zexp}=\removeSel{\zexp'}$.
\end{enumerate}
\end{theorem}
\noindent Theorem \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.

Another useful check is to establish \emph{reachability}, i.e. that it is possible, through a sequence of movement actions, to move the cursor from any position to any other position within a well-typed H-expression.

This requires developing machinery for reasoning about sequences of actions. There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define a syntax for lists of actions, $\bar{\alpha}$, together with iterated action judgements. To keep the core of the action semantics small, we take the latter approach in Figure \ref{fig:multistep}.

A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in the supplement, establishes that $\bar\alpha$ consists only of actions of the form $\aMove{\delta}$.

With these definitions, we can state reachability as follows:

\begin{theorem}[Reachability]\label{thrm:reachability} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\removeSel{\ztau}=\removeSel{\ztau'}$ then there exists some $\bar\alpha$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performTypI{\ztau}{\bar\alpha}{\ztau'}$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$. 
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performSynI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zexp'}{\htau}$.
\end{enumerate}
\end{theorem}

\begin{figure}
$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\vspace{4px}\\
\fbox{$\performTypI{\ztau}{\bar{\alpha}}{\ztau'}$}

\vspace{-10px}\begin{subequations}
\begin{minipage}{0.35\linewidth}
\begin{equation}
\inferrule{ }{
    \performTypI{\ztau}{\cdot}{\ztau}
}
\end{equation}
\end{minipage}
\begin{minipage}{0.65\linewidth}
\begin{equation}
\inferrule{
  \performTyp{\ztau}{\alpha}{\ztau'}\\
  \performTypI{\ztau'}{\bar{\alpha}}{\ztau''}
}{
  \performTypI{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
}
\end{equation}
\end{minipage}
\end{subequations}

\fbox{$\performSynI{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
\vspace{-10px}

\begin{subequations}\label{rules:iterated-syn}
\hspace{-12px}
% \begin{minipage}{0.88\linewidth}
\begin{mathpar}
\inferrule{ }{
  \performSynI{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
}
~~~~~~
\inferrule{
  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
  \performSynI{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
}{
  \performSynI{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
}
~~~~~
\text{(\ref*{rules:iterated-syn}a-b)}
\end{mathpar}
% \end{minipage}
% \begin{minipage}{0.12\linewidth}
% (11a-b)
% \end{minipage}
\end{subequations}

\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
\vspace{-12px}
\begin{subequations}\label{rules:iterated-ana}
\begin{mathpar}
~~~~~~~\inferrule{ }{
  \performAnaI{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
}
~~~~~~~~~~~~~~
\inferrule{
  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
  \performAnaI{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
}{
  \performAnaI{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
}
~~~~~~~~~~
\text{(\ref*{rules:iterated-ana}a-b)}
\end{mathpar}
\end{subequations}
\caption{Iterated Action Judgements}
\label{fig:multistep}
\end{figure}

The simplest way to prove Theorem \ref{thrm:reachability} is to  break it into two lemmas. Lemma \ref{lemma:reach-up} establishes that you can always move the cursor to the outermost position in an expression. This serves as a check on our $\aMove{\dParent}$ rules.
\begin{lemma}[Reach Up]\label{lemma:reach-up} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTypI{\ztau}{\bar\alpha}{\zwsel{\htau}}$.
\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}$. 
\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performSynI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
\end{enumerate}
\end{lemma}
Lemma \ref{lemma:reach-down} establishes that you can always move the cursor from the outermost position to any other position. This serves as a check on our $\aMove{\dChild}$ and $\aMove{\dNext}$ rules. 
\begin{lemma}[Reach Down]\label{lemma:reach-down} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTypI{\zwsel{\htau}}{\bar\alpha}{\ztau}$.
\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}$.
\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ nand $\performSynI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.
\end{enumerate}
\end{lemma}
Theorem \ref{thrm:reachability} follows by straightforward composition of these two lemmas. %(The witness to the existential is not, of course, the shortest path between any two cursor locations.)

% \paragraph{Reachability}

% We give the following theorem to demonstrate that the rules for movement do
% indeed capture the process of moving the focus to any editable position
% within a term. Intuitively, given two zippered expressions that are the
% same up to the erasure of focus, its proof gives a list of composable
% actions that, when applied to the first term produces the second.

% As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
% every action in such a list, it is possible consider moving to an ill-typed
% term. Therefore the statement here matches the mutually recursive structure
% of the bidirectional typing judgements.

% \begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
% \label{thrm:reach}
% Both of the following hold:
% \begin{enumerate}
%   \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
%    $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
%    = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
%    ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
%    = \zexp'$.

%   \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
%    $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
%    = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
%    ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
%    = \zexp'$.
% \end{enumerate}
% \end{theorem}

\subsubsection{Deletion} The $\aDel$ action inserts an empty hole at the cursor, deleting what was there before.

The type action rule for $\aDel$ is self-explanatory:
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\htau}
    }{
      \aDel
    }{
      \zwsel{\tehole}
    }
  }
\end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performTyp{\tarr{\ztau}{\htau}}{\aDel}{\tarr{\ztau'}{\htau}}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performTyp{\tarr{\htau}{\ztau}}{\aDel}{\tarr{\htau}{\ztau'}}
%   }
% \end{equation}

Deletion within a Z-expression is similarly straightforward:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
  }
\end{equation}
%\end{subequations}
% The base case turns into a hole:
%\begin{subequations}
% \begin{equation}
% \inferrule{ }{
%   \performDel{\zwsel{\hexp}}{\hehole}
% }
% \end{equation}
% The rules for the recursive ascription case is shown below. The other recursive cases are analagous:
% \begin{equation}
%   \inferrule{
%     \performDel{\zexp}{\zexp'}
%   }{
%     \performDel{\zexp : \htau}{\zexp' : \htau}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performDel{\hexp : \ztau}{\hexp : \ztau'}
%   }
% \end{equation}
\end{subequations}

\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ at the cursor.

\paragraph{Types} The $\aConstruct{\farr}$ action constructs an arrow type. The H-type under the cursor becomes the argument type, and the cursor is placed on an empty return type hole:
\begin{subequations}
  \begin{equation}
    \label{r:contarr}
  \inferrule{ }{
    \performTyp{
      \zwsel{\htau}
    }{
      \aConstruct{\farr}
    }{
      \tarr{\htau}{\zwsel{\tehole}}
    }
  }
\end{equation}

The $\aConstruct{\fnum}$ action replaces an empty type hole under the cursor with the $\tnum$ type:
  \begin{equation}
    \label{r:contnum}
  \inferrule{ }{
    \performTyp{
      \zwsel{\tehole}
    }{
      \aConstruct{\fnum}
    }{
      \zwsel{\tnum}
    }
  }
\end{equation}

% Construction proceeds recursively down the zipper:
%   \begin{equation}
%     \label{r:contarrL}
%   \inferrule{
%     \performTyp{\ztau}{\aConstruct{\psi}}{\ztau'}
%   }{
%     \performTyp{
%       \tarr{\ztau}{\htau}
%     }{
%       \aConstruct{\psi}
%     }{
%       \tarr{\ztau'}{\htau}
%     }
%   }
% \end{equation}
%   \begin{equation}
%     \label{r:contarrR}
%   \inferrule{
%     \performTyp{\ztau}{\aConstruct{\psi}}{\ztau'}
%   }{
%     \performTyp{
%       \tarr{\htau}{\ztau}
%     }{
%       \aConstruct{\psi}
%     }{
%       \tarr{\htau}{\ztau'}
%     }
%   }
% \end{equation}
\end{subequations}

\begin{subequations}

\paragraph{Ascription} The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type. In the first case, the synthesized type appears in the ascription:
\begin{equation}
  \label{r:constructasc}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}{\htau}
  }
\end{equation}
In the second case, the type provided for analysis appears in the ascription:
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}
  }
\end{equation}
% This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).

\paragraph{Variables} The $\aConstruct{\fvar{x}}$ action places the variable $x$ into an empty hole. If that hole is being asked to synthesize a type, then the result synthesizes the hypothesized type:
\begin{equation}
  \label{r:conevar}
  \inferrule{ }{
    \performSyn{\hGamma, x : \htau}{\zwsel{\hehole}}{\tehole}{\aConstruct{\fvar{x}}}{\zwsel{x}}{\htau}
  }
\end{equation}
If the hole is being analyzed against a type that is consistent with the hypothesized type, then the action semantics goes through the {action subsumption rule} described in Sec. \ref{sec:action-subsumption}. If the hole is being analyzed against a type that is inconsistent with the hypothesized type, $x$ is placed inside a hole:
\begin{equation}
 \label{r:conevar2}
  \inferrule{
    \tincompat{\htau}{\htau'}
  }{
    \performAna{\hGamma, x : \htau'}{\zwsel{\hehole}}{\htau}{\aConstruct{\fvar{x}}}{\hhole{\zwsel{x}}}
  }
\end{equation}
The rule above featured on Line 15 of Figure \ref{fig:second-example}.

\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole. If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
\begin{equation}
  \label{r:conelamhole}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
      {\tarr{\tehole}{\tehole}}
  }
\end{equation}
The type ascription is necessary because lambda expressions do not synthesize a type. If the empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:
\begin{equation}\label{rule:performAna-lam-1}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\zwsel{\hehole}}}
  }
\end{equation}
% Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).

Finally, if the empty hole is being analyzed against a type that has no matched arrow type, expressed in the premise as inconsistency with $\tarr{\tehole}{\tehole}$, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
is inserted inside a hole, which defers the type inconsistency as previously discussed:
\begin{equation}\label{rule:performAna-construct-lam-2}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hhole{
        \hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}
      }}
  }
\end{equation}

\paragraph{Application} The $\aConstruct{\fap}$ action applies the expression under the cursor. The following rule handles the case where the synthesized type has matched function type:
\begin{equation}
  \label{r:coneapfn}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fap}}
      {\hap{\hexp}{\zwsel{\hehole}}}
      {\htau_2}
  }
\end{equation}
If the expression under the cursor synthesizes a type that is inconsistent with an arrow type, then we must place that expression inside a hole to maintain Theorem \ref{sec:holes}:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fap}}
      {\hap{\hhole{\hexp}}{\zwsel{\hehole}}}
      {\tehole}
  }
\end{equation}

The $\aConstruct{\farg}$ action instead places the expression under the cursor in the argument position of an application form. Because the function position is always an empty hole in this situation, we need only a single rule:
\begin{equation}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\farg}}
      {\hap{\zwsel{\hehole}}{\hexp}}
      {\tehole}
  }
\end{equation}

\paragraph{Numbers} The $\aConstruct{\fnumlit{n}}$ action replaces an empty hole with the number expression $\hnum{n}$. If the empty hole is being asked to synthesize a type, then the rule is straightforward:
\begin{equation}
  \label{r:conenumnum}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\fnumlit{n}}}
      {\zwsel{\hnum{n}}}
      {\tnum}
  }
\end{equation}
If the empty hole is being analyzed against a type that is inconsistent with $\tnum$, then we must place the number expression inside the hole:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tnum}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\fnumlit{n}}}
      {\hhole{\zwsel{\hnum{n}}}}
  }
\end{equation}

The $\aConstruct{\fplus}$ action constructs a plus expression with the expression under the cursor as its first argument. If that expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
\begin{equation}\label{rule:construct-plus-compat}
  \inferrule{
    \tcompat{\htau}{\tnum}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fplus}}
      {\hadd{\hexp}{\zwsel{\hehole}}}
      {\tnum}
  }
\end{equation}
Otherwise, we must place that expression inside a hole:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tnum}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fplus}}
      {\hadd{\hhole{\hexp}}{\zwsel{\hehole}}}
      {\tnum}
  }
\end{equation}

\paragraph{Non-Empty Holes} The final shape is $\fnehole$. This explicitly places the expression under the cursor inside a hole:
\begin{equation}
\inferrule{ }{
  \performSyn
    {\hGamma}
    {\zwsel{\hexp}}
    {\htau}
    {\aConstruct{\fnehole}}
    {\hhole{\zwsel{\hexp}}}
    {\tehole}
}
\end{equation}\end{subequations}

The $\fnehole$ shape is grayed out in Figure \ref{fig:action-syntax} because we do not expect the programmer to perform it explicitly -- other actions automatically insert holes when a type inconsistency would arise. Its inclusion is mainly to make it easier to state another ``checksum'' theorem: \emph{constructability}

\paragraph{Constructability}
To check that we have defined ``enough'' construct actions, we need to establish that we can start from an empty hole and arrive at any well-typed expression with, for simplicity, the cursor on the outside (Lemma \ref{lemma:reach-down} allows us to then move the cursor anywhere else.) As with reachability, we rely on the iterated action judgements defined in Figure \ref{fig:multistep}.
\begin{theorem}[Constructability]\label{thrm:constructability} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item For every $\htau$ there exists $\bar\alpha$ such that $\performTypI{\zwsel{\tehole}}{\bar\alpha}{\zwsel{\htau}}$.
\item If $\hana{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that: $$\performAnaI{\hGamma}{\zwsel{\hhole{}}}{\htau}{\bar\alpha}{\zwsel{\hexp}}$$
\item If $\hsyn{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that: $$\performSynI{\hGamma}{\zwsel{\hhole{}}}{\tehole}{\bar\alpha}{\zwsel{\hexp}}{\htau}$$
\end{enumerate}
\end{theorem}

\subsubsection{Finishing}
The final action we need to consider is $\aFinish$, which finishes the non-empty hole under the cursor.

If the non-empty hole appears in synthetic position, then it can always be finished:
\begin{subequations}
  \begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\hexp}{\htau'}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\tehole}
      {\aFinish}
      {\zwsel{\hexp}}
      {\htau'}
  }
\end{equation}

If the non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the enveloped expression is consistent with the type that the hole is being analyzed against. This amounts to analyzing the enveloped expression against the provided type (by subsumption):
\begin{equation}\label{r:finishana}
  \inferrule{
    \hana{\hGamma}{\hexp}{\htau}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\htau}
      {\aFinish}
      {\zwsel{\hexp}}
  }
\end{equation}
\end{subequations}

\subsubsection{Zipper Cases}\label{sec:zipper-cases} The rules defined so far handle the base cases, i.e. the cases where the action has ``reached'' the expression under the cursor. We also need to define the recursive cases, which propagate the action into the subtree where the cursor appears, as encoded by the zipper structure. For types, the zipper rules are straightforward:
\begin{subequations}
\begin{equation}
    %\label{r:contarrL}
  \inferrule{
    \performTyp{\ztau}{\alpha}{\ztau'}
  }{
    \performTyp{
      \tarr{\ztau}{\htau}
    }{
      \alpha
    }{
      \tarr{\ztau'}{\htau}
    }
  }
\end{equation}
  \begin{equation}
    %\label{r:contarrR}
  \inferrule{
    \performTyp{\ztau}{\alpha}{\ztau'}
  }{
    \performTyp{
      \tarr{\htau}{\ztau}
    }{
      \alpha
    }{
      \tarr{\htau}{\ztau'}
    }
  }
\end{equation}
\end{subequations}
For expressions, the zipper rules essentially follow the structure of the corresponding rules in the statics.

\begin{subequations}
In particular, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):
\begin{equation}
\inferrule{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \performAna
    {\hGamma, x : \htau_1}
    {\zexp}
    {\htau_2}
    {\alpha}
    {\zexp'}
}{
  \performAna
    {\hGamma}
    {\hlam{x}{\zexp}}
    {\htau}
    {\alpha}
    {\hlam{x}{\zexp'}}
}
\end{equation}

When the cursor is in the expression position of an ascription, we use the analytic  judgement, mirroring Rule (\ref{rule:syn-asc}):
\begin{equation}
\inferrule{
  \performAna
    {\hGamma}
    {\zexp}
    {\htau}
    {\alpha}
    {\zexp'}
}{
  \performSyn
    {\hGamma}
    {\zexp : \htau}
    {\htau}
    {\alpha}
    {\zexp' : \htau}
    {\htau}
}
\end{equation}

When the cursor is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
\begin{equation}\label{rule:zipper-asc}
\inferrule{
  \performTyp{\ztau}{\alpha}{\ztau'}\\
  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
}{
  \performSyn
    {\hGamma}
    {\hexp : \ztau}
    {\removeSel{\ztau}}
    {\alpha}
    {\hexp : \ztau'}
    {\removeSel{\ztau'}}
}
\end{equation}

When the cursor is in the function position of an application, the rule mirrors Rule (\ref{rule:syn-ap}):
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
    \performSyn
      {\hGamma}
      {\zexp}
      {\htau_2}
      {\alpha}
      {\zexp'}
      {\htau_3}\\\\
    \arrmatch{\htau_3}{\tarr{\htau_4}{\htau_5}}\\
    \hana{\hGamma}{\hexp}{\htau_4}
  }{
    \performSyn
      {\hGamma}
      {\hap{\zexp}{\hexp}}
      {\htau_1}
      {\alpha}
      {\hap{\zexp'}{\hexp}}
      {\htau_5}
  }
\end{equation}
% \begin{equation}
%   \inferrule{
%     \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
%     \performSyn
%       {\hGamma}
%       {\zexp}
%       {\htau_2}
%       {\alpha}
%       {\zexp'}
%       {\tehole}\\
%     \hana{\hGamma}{\hexp}{\tehole}
%   }{
%     \performSyn
%       {\hGamma}
%       {\hap{\zexp}{\hexp}}
%       {\htau_1}
%       {\alpha}
%       {\hap{\zexp'}{\hexp}}
%       {\tehole}
%   }
% \end{equation}

The situation is similar when the cursor is in argument position:
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\hexp}{\htau_2}\\
    \arrmatch{\htau_2}{\tarr{\htau_3}{\htau_4}}\\
    \performAna
      {\hGamma}
      {\zexp}
      {\htau_3}
      {\alpha}
      {\zexp'}
  }{
    \performSyn
      {\hGamma}
      {\hap{\hexp}{\zexp}}
      {\htau_1}
      {\alpha}
      {\hap{\hexp}{\zexp'}}
      {\htau_4}
  }
\end{equation}
% \begin{equation}
%   \inferrule{
%     \hsyn{\hGamma}{\hexp}{\tehole}\\
%     \performAna
%       {\hGamma}
%       {\zexp}
%       {\tehole}
%       {\alpha}
%       {\zexp'}
%   }{
%     \performSyn
%       {\hGamma}
%       {\hap{\hexp}{\zexp}}
%       {\tehole}
%       {\alpha}
%       {\hap{\hexp}{\zexp'}}
%       {\tehole}
%   }
% \end{equation}

The rules for the addition operator mirror Rule (\ref{rule:syn-plus}):
\begin{equation}
  \inferrule{
    \performAna
      {\hGamma}
      {\zexp}
      {\tnum}
      {\alpha}
      {\zexp'}
  }{
    \performSyn
      {\hGamma}
      {\hadd{\zexp}{\hexp}}
      {\tnum}
      {\alpha}
      {\hadd{\zexp'}{\hexp}}
      {\tnum}
  }
\end{equation}
\begin{equation}
  \inferrule{
    \performAna
      {\hGamma}
      {\zexp}
      {\tnum}
      {\alpha}
      {\zexp'}
  }{
    \performSyn
      {\hGamma}
      {\hadd{\hexp}{\zexp}}
      {\tnum}
      {\alpha}
      {\hadd{\hexp}{\zexp'}}
      {\tnum}
  }
\end{equation}

Finally, if the cursor is inside a non-empty hole, the relevant zipper rule mirrors Rule (\ref{rule:syn-ehole}):
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
    \performSyn
      {\hGamma}
      {\zexp}
      {\htau}
      {\alpha}
      {\zexp'}
      {\htau'}\\
  }{
    \performSyn
      {\hGamma}
      {\hhole{\zexp}}
      {\tehole}
      {\alpha}
      {\hhole{\zexp'}}
      {\tehole}
  }
\end{equation}
% \begin{equation}
%   \inferrule{
%     \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
%     \performSyn
%       {\hGamma}
%       {\zexp}
%       {\htau}
%       {\alpha}
%       {\zwsel{\hehole}}
%       {\tehole}\\
%   }{
%     \performSyn
%       {\hGamma}
%       {\hhole{\zexp}}
%       {\tehole}
%       {\alpha}
%       {\zwsel{\hehole}}
%       {\tehole}
%   }
% \end{equation}

Theorem \ref{thrm:actsafe} directly checks the correctness of these rules. Moreover, the zipper rules arise ubiquitously in derivations of edit steps, so the proofs of the other ``check'' theorems, e.g. Reachability and Constructability, serve as a check that none of these rules have been missed.
\end{subequations}

### 2016-08-24 hazelnut-popl17.tex
copied material from POPL submission

\def\OPTIONConf{1}%
%\newif\iftr\trfalse
\newif\iftr\trtrue
%\documentclass[12pt]{article}
\RequirePackage{etex}
\documentclass[preprint,9pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{multicol}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.9}
\usepackage{soul}
\setulcolor{red}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\usepackage{listings}
\lstset{tabsize=2,
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

\usepackage{joshuadunfield}
\usepackage{llproof}
\usepackage{rulelinks}

%\usepackage{balance} %% Balance last page's two columns

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}
\makeatletter

\renewcommand\topfraction{0.85}
\renewcommand\bottomfraction{0.85}
\renewcommand\textfraction{0.1}
\renewcommand\floatpagefraction{0.85}

\AtBeginDocument{%
 \abovedisplayskip=2pt
 \abovedisplayshortskip=0pt
 \belowdisplayskip=2pt
 \belowdisplayshortskip=0pt
}

\setlength{\floatsep}{10pt}
\setlength{\textfloatsep}{12pt}

\usepackage[compact]{titlesec}
\titlespacing*{\section}{0pt}{4pt}{2pt}
\titlespacing*{\subsection}{0pt}{4pt}{2pt}
\titlespacing*{\subsubsection}{0pt}{4pt}{2pt}
\titlespacing*{\paragraph}{0pt}{4pt}{2pt}
\setlength{\skip\footins}{3pt plus 1px minus 5px}

\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink

\usepackage{enumitem}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=4px plus 2px minus 2px
  \thm@postskip=\thm@preskip % or whatever, if you don't want them to be equal
}
\makeatother

\usepackage{todonotes}
\usepackage{xcolor}
%\usepackage{adjustbox}

\input{macros}
\begin{document}

\conferenceinfo{-}{-}
\copyrightyear{-}
\copyrightdata{[to be supplied]}

%\titlebanner{}        % These are ignored unless
\preprintfooter{Draft}   % 'preprint' option specified.

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}
% \title{Hazelnut:\\A Bidirectionally Typed Structure Editor Calculus}

%\authorinfo{\vspace{-2px}}{}{}
%\authorinfo{~}{~}{\vspace{-10px}}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
%         {Carnegie Mellon University}
%         {\{comar, aldrich\}@cs.cmu.edu}
\iftr
\authorinfo{
        Cyrus Omar$^1$
        \and Ian Voysey$^1$
        \and Michael Hilton$^2$
        \and Jonathan Aldrich$^1$
        \and Matthew A. Hammer$^3$
}
{
 $^1$~Carnegie Mellon University
 \and
 $^2$~Oregon State University
 \and
 $^3$~University of Colorado Boulder\vspace{-10px}
}
{
%\email{comar@cs.cmu.edu}\\
%\texttt{http://www.cs.cmu.edu/\homedir comar/}
%\email{hiltonm@eecs.oregonstate.edu}\\
}
\else
\authorinfo{~}{~}{\vspace{-10px}}
\fi

%
%\authorrunning{Omar et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
%\email{Matthew.Hammer@colorado.edu}

\maketitle
\begin{abstract}
% Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker. Not all text survives this journey. In particular, text that arises transiently, or when the programmer makes a mistake, is often malformed or ill-typed.
% Not all text that arises during the programming process survives the journey through a parser and typechecker. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
% Contending with malformed program text or well-formed but meaningless syntax trees is difficult for programmers and their tools alike.

%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
% Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.

\emph{Structure editors} allow programmers to edit the tree structure of a program directly. This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they do not need to contend with malformed program text.

This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: it's {edit actions} operate over statically meaningful (i.e. well-typed) terms.  
Na\"ively, this prohibition on ill-typed edit states would force the programmer to construct terms in a rigid ``outside-in'' manner. To avoid this problem, the {action semantics} automatically places terms assigned a type that is inconsistent with the expected type {inside} a {hole}. This safely defers the type consistency check until the term inside the hole is \emph{finished}.

Hazelnut is a foundational type-theoretic account of typed structure editing, rather than an end-user tool itself. To that end, we describe how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to the calculus. We also discuss various plausible evaluation strategies for terms with holes, and in so doing reveal connections with gradual typing and contextual modal type theory (the Curry-Howard interpretation of contextual modal logic.) Finally, we  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml}.

%Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation

\section{Introduction}\label{sec:introduction}
\input{intro}

\section{Programming in Hazelnut}\label{sec:example}
\input{example}

\section{Hazelnut, Formally}
\label{sec:hazel}
\input{formally}

\subsection{Mechanization}
\label{sec:mech}\label{sec:mt}
\input{agda}

% \clearpage

% \section{Dynamics of Hazelnut}\label{sec:dynamics}

% \subsection{Complete Expressions}
% ... canonical forms and type safety with e complete in it ...

% \subsection{Holes as TODOs}
% ... add an e err and define new type safety ...

% \subsection{Indeterminate Expressions}
% \input{dynamics}

% \clearpage

\section{Extending Hazelnut}\label{sec:extending}
%% \todo{write this section} Add sum types.
%% H-expression
%% Z-expression + Erasure rules
%% Statics
%% Movement actions
%% Construct actions
\input{sumtypes}

\section{Implementation}
\label{sec:impl}
\input{implementation}

\section{Related Work and Discussion}\label{sec:rw}
%\subsection{Structure Editors}
\subsection{Structure Editors}
Syntactic structure editors have a long history -- the Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in 1981. Novice programmers have been a common target for structure editors. For example, 
GNOME~\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates. 
Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students. Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure editor targeted at children ages 8 to 16.  TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students. An implementation of Hazelnut might be useful in teaching students about the typed lambda calculus, though that has not been our explicit aim with this work.

Not all structure editors are for educational purposes. For example,
mbeddr \cite{voelter_mbeddr:_2012} is a structure editor for a C-based programming language (nominally, for programming embedded systems.)  
Lamdu~\cite{lamdu}, like Hazelnut, is a structure editor for a statically typed functional language. It is designed for use by professional programmers.

The examples given so far either define a language with a trivial static semantics, or do not attempt to maintain well-typedness as an edit invariant. This can pose problems, for reasons discussed in the Introduction. One apparent exception is Unison~\cite{unison}, a structure editor for a typed functional language similar to Haskell. Like Hazelnut, it seems to define some notion of well-typedness for expressions with holes (though there is no technical documentation on virtually any aspect of its design.) Unlike Hazelnut, it does not have a notion analagous to Hazelnut's notion of a non-empty hole. As such, programmers must construct programs in a rigid outside-in manner, as discussed in Sec. \ref{sec:example}. Another system with the same problem is Polymorphic Blocks, a block-based user interface where the structure of block connectors encodes a type \cite{DBLP:conf/chi/LernerFG15}.

We fundamentally differ from these projects in our design philosophy: we consider it essential to start by building type theoretic foundations, which are independent of nearly all decisions about the user interface. In contrast, these editors have developed innovative user interfaces (e.g. see the discussion in \cite{DBLP:conf/sle/VolterSBK14}) but lack a principled foundational calculus. In this respect, we follow the philosophical approach taken by languages that are rooted in the type theoretic tradition and have gone to great effort to develop a clear metatheory, like Standard ML \cite{mthm97-for-dart,Harper00atype-theoretic}.  In the future, we hope that these lines of research will merge to produce a human-usable typed structure editor with sound formal foundations. Our contribution, then, is in defining and analyzing the theoretical properties of a small foundational calculus that could be extended to achieve this vision.

Some structure editor generators do operate on formal or semi-formal definitions of an underlying language. For example, the Synthesizer Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an attribute grammar-based language implementation that then can be used to generate a structured editor. CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language. Barista is a programmatic toolkit for building structure editors \cite{ko_barista:_2006}. mbeddr is built on top of the commercial JetBrains MPS framework for constructing structure editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}. These systems do not give a semantics to the edit states of the structure editor itself, or maintain non-trivial edit invariants, as Hazelnut does. 
%These early systems were developed  of the systems are rooted in the type-theoretic tradition.

Related to structure editors are value editors, which operate directly on simple values (but not generally expressions or functions) of a programming language. For example, Eros is a typed value editor based in Haskell \cite{DBLP:conf/icfp/Elliott07}.

\subsection{Gradual Type Systems}
A significant contribution of this paper is the discovery of a clear technical relationship between typed structure editing and gradual typing. In particular, the machinery necessary to give a reasonable semantics to type holes -- i.e. type consistency and type matching -- coincides with that developed in the study of gradual type systems for functional languages. The pioneering work of Siek and Taha \cite{Siek06a} introduced type consistency. Subsequent work developed the concept of type matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}. In retrospect, this relationship is perhaps unsurprising: gradual typing is, notionally, also motivated by the idea of iterated development of a program where every intermediate state is well-defined in some sense. %Typed structure editing endows this intuitive notion with technical force.

Recent work has discovered a systematic procedure for generating a ``gradual version'' of a standard type system \cite{DBLP:conf/popl/CiminiS16}. This system, called the Gradualizer, operates on a logic program that specifies a simple type assignment system with some additional annotations to generate a corresponding specification of a gradual type system. The authors leave support for working with bidirectional type systems as future work. This suggests the possibility of an analagous ``Editorializer'' that generates a specification of a typed structure editor from a simple language definition. Our exposition in Sec. \ref{sec:extending} certainly suggests that many of the necessary definitions follow seemingly mechanically from the definition of the static semantics, and the relationship with gradual typing suggests that many of the technical details of this transformation may already exist in the Gradualizer. One possibility we have explored informally is to use Agda's reflection features to implement such a system.

An aspect of gradual typing that we did not touch on directly here is its concern with  assigning a dynamics to programs where type information is not known, by inserting dynamic type casts. This would correspond to assigning a dynamics to Hazelnut expressions with type holes such that a run-time error occurs when a type hole is found to be unfillable through evaluation.

\subsection{Type Holes as Unification Variables}
Another possible interpretation of type holes is as explicit unification variables. In other words, we might define the dynamics of Hazelnut such that a program cannot be run if some automated type inference procedure cannot statically fill in any type holes that arise. For a simple calculus, e.g. the STLC upon which Hazelnut is based, type inference is known to be decidable \cite{damas1982principal}. In more complex settings, e.g. in a dependently typed language, a partial decision procedure may still be useful in this regard, both at edit-time and (just prior to) run-time. Indeed, text editor modes for proof assistants, e.g. for Agda, attempt to do exactly this for indicated ``type holes'' (and do not always succeed.)

\subsection{Exceptions}
Expression holes, too, could dynamically be interpreted in several ways. One straightforward approach would be to dynamically raise an exception if evaluation encounters an expression hole. Indeed, placing \textt{raise Unimplemented} or similar in portions of an expression that are under construction is a common practice across programming languages.

\subsection{Type-Directed Program Synthesis}
Some text editor modes, e.g. those for proof assistants like Agda \cite{norell:thesis} and Idris \cite{brady2013idris}, support a more explicit hole-based programming model where indicated expression holes are treated as sites where the system can be asked to automatically generate an expression of an appropriate type. These systems are not statically well-defined themselves (though see below.)

The Graphite system used an informal model of typed holes in Java to allow library providers to associate interactive code generation interfaces with types \cite{Omar:2012:ACC:2337223.2337324}. %RThis system was limited in that these interfaces could not themselves manipulate terms.

More generally, the topic of type-directed program synthesis an active area of research, e.g. \cite{DBLP:conf/pldi/OseraZ15}. By maintaining static well-definedness throughout the editing process, Hazelnut provides researchers interested in editor-integrated type-directed program synthesis  with a formal foundation upon which to build.% such systems could be integrated directly into editors.

\subsection{Contextual Modal Type Theory}
Expression holes can also be understood by invoking the notion of a \emph{metavariable} as found in contextual modal type theory (CMTT) \cite{DBLP:journals/tocl/NanevskiPP08}. In particular, expression holes have types and are surrounded by contexts, just as metavariables in CMTT are associated with types and contexts. This helps to clarify the logical meaning of a typing derivation in Hazelnut -- it conveys well-typedness relative to an (implicit) modal context that extracts each expression hole's type and context. The modal context must be emptied -- i.e. the expression holes must be instantiated with expressions of the proper type in the proper context -- before the expression can be considered complete. This corresponds to the notion of modal necessity in contextual modal logic.

Making the modal context explicit in our semantics is not technically useful given our stated purpose -- interactive program editing is not merely hole filling in Hazelnut (i.e. the cursor need not be on a hole). Moreover, the hole's type and context become apparent as our action semantics traverses the zipper structure on each step. Some interactive proof assistants, however, support a tactic model based directly on hole filling, so the connection to CMTT and similar systems is more useful. For example,  Beluga \cite{DBLP:conf/flops/Pientka10} is based on dependent CMTT and aspects of Idris' editor support \cite{brady2013idris} are based on McBride's OLEG \cite{mcbride2000dependently} and Lee and Friedman have explored a lambda calculus with contexts for a similar purpose \cite{DBLP:conf/icfp/LeeF96}.

One interesting avenue of future work is to elaborate expression holes to CMTT's closures, i.e. CMTT terms of the form $\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable associated with each hole and $\text{id}(\Gamma)$ is the explicit identity substitution. This would allow us to evaluate expressions with holes such that the closure ``accumulates'' substitutions explicitly. When evaluation gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped with a notion of progress under a non-empty modal context), it would then be possible for the programmer to choose holes from amongst the visible holes (which may have been duplicated) to edit in their original context. Once finished, the CMTT hole instantiation operation, together with a metatheorem that establishes that reduction commutes with instantiation, would enable an ``edit and resume'' feature with a clear formal basis. This notion of reduction commuting with instantiation has also been studied in other calculi \cite{DBLP:journals/entcs/Sands97}. Being able to edit a running program also has connections to less formal work on ``live programming'' interfaces \cite{burckhardt2013s,lamdu}. %, e.g.

\subsection{Evaluation Strategies: A High-Dimensional Space}
To summarize, we have discussed three different evaluation strategies in the presence of type holes:
\begin{enumerate}[noitemsep]
\item ...as preventing evaluation (the standard approach.)
\item ...as unknown types, in the gradual sense.
\item ...as unification variables.
\end{enumerate}
In addition, we have discussed four different evaluation strategies in the presence of expression holes:
\begin{enumerate}[noitemsep]
\item ...as preventing evaluation (the standard approach.)
\item ...as causing exceptions.
\item ...as sites for automatic program synthesis.
\item ...as the closures of CMTT.
\end{enumerate}

Every combination of these choices could well be considered in the design of a full-scale programming system in the spirit of Hazelnut. Indeed, the user might be given several options from among these combinations, depending on their usage scenario. Many of these warrant further inquiry.

% \todo{Eclipse parsing partial programs // other heuristics}

% \paragraph{Gradual typing: Has Matched Relation}
% %
% A. Rastogi, A. Chaudhuri, and B. Hosmer. The ins and outs of gradual type inference.
% \\
% Definition 3.1 is similar to our ``has matched'' relations:
% \\
% \url{https://www.cs.umd.edu/~avik/papers/iogti.pdf}

% This paper defines ``has matched'' relation for arrow, just like us:
% \\
% \url{http://cimini.info/publications/Gradualizer_Draft.pdf}
% \\
% They credit this technique to the following paper:
% \\
% \url{http://www.cs.ubc.ca/~rxg/ptsgp.pdf}
% \\
% But, the Rastogi paper came earlier.

% \todo{TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}}

% Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g. Emacs.) These systems also have notions of holes and use types to guide the user toward filling these holes. These  systems are also, to our knowledge, not formally well-defined but rather exist only as part of system implementations.

% Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified. For example, while Lamdu has many interesting features, there is no theoretical basis presented for their work -- it is a rather large body of Haskell code with an unclear (and indeed, often somewhat perplexing, in our experience) action model. Unison is also a rather large body of Haskell code, though its action model appears superficially more similar to ours. We maintain what we believe to be a stronger action sensibility invariant than Unison (i.e. in Unison, one must construct expressions from the outside-in.) These systems are rich sources of interesting ideas, however -- there is room enough for many different approaches in this (re-)emerging space.

% \todo{TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}}

% \todo{TODO: cite} \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}

%Drag-and-drop / for novices: lots of examples, e.g. Alice and others
%
%Contemporary: Lamdu, MPS/Mbeddr, TouchDevelop
%
%Hybrid: Cyrus' active code completion paper

%\subsection{Refactoring Models}
%(Michael, can you fill this section out?)

%\subsection{Formal Editor Models}
%Need to do a search to see what else has been done...

\section{Conclusion}
\label{sec:future}
This paper presented Hazelnut, a type theoretic structure editor calculus. Our aim was to take a principled approach to its design by formally defining its static semantics as well as its action semantics and developing a rich metatheory. Moreover, we have mechanized substantial portions of the metatheory, including the crucial Sensibility theorem that establishes that every edit state is statically meaningful.

In addition to simplifying the job of an editor designer, typed structure editors also promise to increase the speed of development by eliminating redundant syntax and supporting higher-level primitive actions. However, we did not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g. whether a keyboard or a mouse is in use. Indeed, we consider it a virtue of this work that such implementation details do not enter into our design.

% By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
% By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?  In progress, unfinished, incomple?} and yet still in a well defined type state. 
% This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.\todo{integrate this}

\subsection{Future Work}
\subsubsection{Richer Languages}
Hazelnut is, obviously, a very limited language at its core. So the most obvious avenue for future work is to increase the expressive power of this language by extension. Our plan is to simultaneously maintain a mechanization and implementation (following, for example, Standard ML) as we proceed, ultimately producing the first large-scale, formally verified bidirectionally typed structure editor.

It is interesting to note that the demarcation between the language and the editor is fuzzy (indeed, non-existent) in Hazelnut. There may well be interesting opportunities in language design when the language is being codesigned with a typed structure editor. It may be that certain language features are unnecessary given a sufficiently advanced type-aware structure editor (e.g. SML's \texttt{open}?), while other features may only be practical with editor support. We intend to use Hazelnut and derivative systems thereof as a platform for rigorously exploring such questions.

\subsubsection{Editor Services}
There are various aspects of the editor model that we have not yet formalized. For example, our action model does not consider how actions are actually entered using, for example, key combinations or chords. In practice, we would want also to rank available actions in some reasonable manner (perhaps based on usage data gathered from other users or code repositories.) Designing a rigorous typed probability model over actions and H-expressions is one avenue of research that we have started to explore, with intriguing initial results.

\subsubsection{Programmable Actions}
Our language of actions is intentionally primitive. However, even now it acts much like a simple imperative command language. This suggests future expansion to, for example, a true tactic language. Alternatively, it may be more useful to develop the notion of an \emph{action macro}, whereby functional programs could themselves be lifted to the level of actions to compute non-trivial compound actions.

\subsubsection{Views}
Another research direction is in exploring how types can be used to control the presentation of expressions in the editor. For example, following an approach developed in a textual setting of developing \emph{type-specific languages} (TSLs), it should be possible to have the type that an expression is being analyzed against define alternative display forms and interaction modes \cite{TSLs}.

It should also be possible to develop a semantics of semantic comments, i.e. comments that mention semantic structures. These would be subject to the same operations, e.g. renaming, as other structures, helping to address the problem of comments becoming inconsistent with code.

\subsubsection{Collaborative Programming}
Finally, we did not consider any aspects of \emph{collaborative programming}, such as a packaging system, a differencing algorithm for use in a source control system, support for multiple simultaneous focii for different users, and so on. These are all interesting avenues for future work.

\subsubsection{Empirical Evaluation}
Although we make few empirical claims in this paper, it is ultimately an empirical question as to whether structure editors, and typed structure editors, are practical. We hope to conduct user studies once a richer semantics has been developed.

\subsubsection{More Theory}
Connections with gradual type systems and CMTT, discussed in the previous section, seem likely to continue to be revealing.

The notion of having one of many possible locations within a term under a cursor has a very strong intuitive connection to the proof theoretic notion of focusing \cite{Simmons11tr}. Building closer connections with proof theory (and category theory) is likely to be a fruitful avenue of further inquiry.

% We already discussed a connection to gradual typing \cite{Siek06a}. We hope to explore this connection more thoroughly. In particular, it may be possible to better support exploratory and live programming by allowing even programs with holes in them to execute as long as those holes are only in the type portions, by deferring to the semantics given in work on gradual typing.

% It may also be possible to give a dynamics to incomplete expressions. Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and
% $\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
% explored an equational theory for terms with holes, but intend to once our
% formalization effort is more mature.\todo{fix bib}

% \todo{new action form that makes actions extensible given proofs of
% admissiblity of a derived form, like prevSib}

\begin{quote}
\emph{In any case, these are but steps toward more graphical program-description
systems, for we will not forever stay confined to mere strings of symbols.}

--- Marvin Minsky, Turing Award lecture \cite{DBLP:journals/jacm/Minsky70}
\end{quote}

\clearpage
% We recommend abbrvnat bibliography style.
\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.
\bibliography{bibliography}
%\softraggedright
%P. Q. Smith, and X. Y. Jones. ...reference text...

\iftr
\clearpage
%\onecolumn
\appendix
\input{appendix}
\else
% No Appendix Here
\fi

\end{document}

### 2016-08-24 implementation.tex
copied material from POPL submission

% !TEX root = hazelnut-popl17.tex
\subsection{Implementation Concepts}
Central to any implementation of Hazelnut is a stream of edit states whose cursor erasures synthesize types under an empty context according to the synthetic action judgement, $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$. The middle row of Figure \ref{fig:impl-overview} diagrams this stream of edit states. By Theorem \ref{thrm:actsafe}, the editor does not need to typecheck each edit state (though in some of the rules, e.g. Rule (\ref{rule:zipper-asc}) which handles the situation where the type in a type ascription changes, portions of the program do need to be typechecked.) For example, the reader is encouraged to re-examine the examples in Figure \ref{fig:first-example} and \ref{fig:second-example} -- the cursor erasure of each edit state synthesizes a type.

The programmer examines a view generated from each edit state and produces actions in some implementation-defined manner (e.g. using a keyboard, mouse, touchscreen, voice interface, or neural implant.) Each new action causes a new abstract edit state to arise according to an implementation of the action semantics. This then causes a new view to arise. This is a simple event-based functional reactive programming model \cite{Wan:2000:FRP:349299.349331}.

If an action is not well-defined according to Hazelnut's action semantics, the implementation must reject it. In fact, the implementation is encouraged to present an ``action palette'' that either hides or visibly disables actions that are not well-defined (see below.)

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{impl-overview2}
\caption{Implementation Concepts}
\label{fig:impl-overview}
\end{figure}

\subsection{HZ}
We have developed a simple reference implementation, HZ, of Hazelnut.
In order to reach a wide audience, we decided to implement HZ in the web browser.
In order to take advantage of a mature implementation of the FRP model, we chose to implement HZ using OCaml\footnote{\url{https://ocaml.org/}}, the \texttt{js\_of\_ocaml} compiler and associated libraries \cite{DBLP:conf/ml/Balat06}\footnote{\url{http://ocsigen.org/js\_of\_ocaml/}} and the OCaml React library\footnote{\url{http://erratique.ch/software/react}}.

At the time of the writing, 
HZ's view computation renders the model as a string embedded into HTML that approximates our typeset notation in this paper. The action palette is a collection of buttons and text boxes, which are disabled when an action is not well-defined or an invalid action argument is entered. We determine this by simply attempting to perform the action internally and catching the exception that is raised when an action is undefined. This design, of course, is not meant to score marks for usability or performance. Instead, as a simple reference implementation, we expect its source code to be of use to others who are interested in layering a more fluid user interface atop the core semantics (which have been implemented in a style that follows the presentation in this paper closely), or extensions thereof.

### 2016-08-24 intro.tex
copied material from POPL submission

% !TEX root = hazelnut-popl17.tex

%% There are some benefits to this approach, to be sure, but the structural
%% mismatch between programs and their textual representations also imposes
%% various burdens.  For example, the primitive edit actions available in a
%% text editor (e.g. inserting or deleting a character or word) do not
%% always correspond to sensible structural transformations.

% spj: describe the problem; state our contributions; STOP. one page max.

% When constructing a program or proof in a language with rich type
% structure, skilled programmers generally follow a \emph{type discipline}
% where they first determine the type of the expression that they are
% constructing in order to constrain the mental search space that they are
% operating within.

% For example, if the programmer knows that an expression of type
% $\tarr{\tnum}{\tnum}$ is needed, then it is often the case (though, of course, not
% necessarily the case) that the expression will take the form $$\hlam{\mathit{x}}{e}$$
% for some variable $x$ and function body $e$. If the programmer chooses this
% form, then after picking a suitable variable name, her focus will be on
% constructing a suitable body, $e$. Following the type discipline, $e$ must
% be of type $\tnum$, and so this process can begin anew.

% The problem is that when using a text
% editor to construct a program, it is easy, and indeed necessary, to deviate from this disciplined process.  Rather, text editors operate on sequences of
% characters (i.e. \emph{text}.)

%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
% Programming languages, and therefore programs themselves, are rich mathematical structures.
Programmers typically construct and manipulate well-typed expressions only indirectly, by editing text that is first parsed according to a textual syntax and then typechecked according to a static semantics. This indirection has practical benefits, to be sure -- text editors and other text-based tools benefit from decades of development effort -- but it also introduces some fundamental complexity into the programming process.

% Every day, programmers use text-based tools to construct and manipulate programs.  
% These programs are written in languages which defined by a textual syntax. 
% While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.

% It also complicates matters for tool designers because tools are often confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}. This may be because the programmer is in the midst of a sequence of 
% edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake. The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g. by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)
%Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
%Most programming languages define a textual syntax. This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs. While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.

First, it requires that programmers learn the subtleties of the textual syntax (e.g. the precedence of the various forms.) This can be particularly challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258}. %For example, one study of novice programmers found syntax errors (e.g. unbalanced parentheses) to be the most common class of errors.
% Too much detail?

The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers. In particular, program editors must contend with meaningless text on a regular basis. In a dataset analyzed by Yoon and Myers consisting of 1460 hours of edit logs, 44.2\% of edit states were syntactically malformed \cite{6883030}. Some additional percentage of edit states were well-formed but ill-typed (the dataset did not contain enough information to determine the exact percentage.)  
Some of these meaningless edit states arose because the programmer was in the midst of a sequence of edits that left the edit state transiently meaningless, while others arose because the programmer made a logical mistake. Because the language definitions is silent about these edit states, it is difficult to design useful editor services, e.g. syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \cite{mens2004survey}. Editors must either disable these editor services when they encounter meaningless edit states or develop \emph{ad hoc} heuristics, which can be misleading. %, or because the programmer has made a mistake. 
% The syntax definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action

% Many editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or continue past a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer. % This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.

% stuck editing a representation of the program instead of the structures
% themselves. The editor does not restrict what the programmer may do: you
% can delete characters that belong, insert ones that don't, forget things
% that were needed, and so. There's nothing stopping us from accidentally
% writing $$\lambda \mathit{x:num}.\mathit{(x,x)}$$ even though it's obvious
% that building a pair can't hope to form a natural number.

% The type structure of the language makes this sort of error obvious: it's
% not that you're adding characters that make your program incorrect, or even
% malformed; you're adding characters that can't possibly create a structure
% you want because of the type. Simply put, the primitive operations
% available in text editors do not always correspond to sensible
% transformations on the structure of the program.
These complications have motivated a long line of research into \emph{structure editors}, i.e. program editors where every edit state is a program structure. % Eliminating text eliminates the possibility of syntax errors.

Most structure editors are \emph{syntactic structure editors}, i.e. the edit state is a syntax tree with \emph{holes} that stand for branches of the tree that have yet to be constructed, and the edit actions are context-free tree transformations. For example, Scratch is a syntactic structure editor that has achieved success as a tool for teaching children how to program \cite{Resnick:2009:SP:1592761.1592779}.

The Scratch language has a trivial static semantics, but researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics. For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}. Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design. The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed. As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or deploy \emph{ad hoc} heuristics.
%Indeed, it is not entirely clear what it should mean for terms with holes to be statically meaningful. 
% Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.

% \begin{itemize}
% \item

% First, there is an all too familiar problem: these languages do not have a clear formal semantics and metatheory, i.e. ``the implementation is the specification''. This makes it difficult to reason methodically about types and binding, so, in turn, it is difficult to design reliable editor services.

% Even if a formal semantics and metatheory for the underlying language were forthcoming (e.g. following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}), it would go only part of the way towards addressing the challenges faced by structure editor designers. First, a standard semantics assigns no formal meaning to \emph{incomplete terms}, i.e. terms with holes. Moreover, a syntactic structure editor  -- so incorporating holes into the static semantics would in any case be only a minor salve. Heuristics analagous to those that pervade textual program editors would still be needed to contend with the ill-typed edit states.

This paper develops a principled solution to this problem. We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}. Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e. well-typed) H-expression with a single superimposed \emph{cursor}. We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because our encoding follows Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g. a function argument), only actions consistent with that type are permitted.

Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner. For example, the programmer would often need to construct the outer function application form before identifying the intended function. To address this problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole if the expression's type is inconsistent with the expected type. This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}. In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remain under construction. %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}. %Actions act at a programmer-indicate subtree, called the \emph{focus} (which is defined following Huet's zipper pattern.)

% By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state. 
%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics. Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state.

%However, these syntactically correct states can be semantically meaningless (i.e. undefined), because the language definitions generally only give meaning to complete, well-typed terms. (e.g., when a branching element is introduced, the program does not become semantically correct until both branches are complete)
% This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
% Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
%Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g. Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .

The remainder of this paper is organized as follows:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item We begin in Sec. 
    \ref{sec:example} with two examples of edit sequences to develop the reader's intuitions.  
\item We then give a detailed overview of Hazelnut's semantics and metatheory, and our mechanization in the Agda proof assistant, in Sec.  \ref{sec:hazel}.
\item Hazelnut is a {foundational} calculus, i.e. a calculus that language and editor designers are expected to extend with higher level constructs. We extend Hazelnut with simple sum types in Sec.  \ref{sec:extending} to demonstrate how Hazelnut's rich metatheory guides such a development. 
\item In Sec.  \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.

% \item However, we consider the space of possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}. 
% so doing, we discover interesting connections with gradual typing and contextual modal type theory. The former provides an interpretation of type holes, and the latter provides a logical interpretation of expression holes. \todo{revise this one}

\item In Sec.  \ref{sec:rw}, we summarize related work. In particular, much of the technical machinery needed to handle type holes coincides with machinery developed in the study of gradual type systems. Similarly, expression holes can be interpreted as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work. %This suggests a principled design for an ``edit and resume'' feature.

\item We conclude in Sec.  \ref{sec:future} by summarizing our vision of a principled science of structure editor design rooted in type theory, and suggest a number of future directions.
\end{itemize} 
The supplemental material includes 1) a typeset listing of Hazelnut's rules in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our reference implementation of Hazelnut, both in source form and pre-compiled to JavaScript for use in a web browser.

### 2017-05-03 hazelnut-dynamics.tex
minor stuff
hollowed out repository

--- 2016-08-24 hazelnut-popl17.tex
@@
-\def\OPTIONConf{1}%
-%\newif\iftr\trfalse
-\newif\iftr\trtrue
-%\documentclass[12pt]{article}
-\RequirePackage{etex}
-\documentclass[preprint,9pt]{sigplanconf}
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-\usepackage[T1]{fontenc}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{ stmaryrd }
+\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
+%% For double-blind review submission
+\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
+%% For single-blind review submission
+%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
+%% For final camera-ready submission
+%\documentclass[acmlarge]{acmart}\settopmatter{}
+
+%% Note: Authors migrating a paper from PACMPL format to traditional
+%% SIGPLAN proceedings format should change 'acmlarge' to
+%% 'sigplan,10pt'.
+
+%% Some recommended packages.
+\usepackage{booktabs}   %% For formal tables:
+                        %% http://ctan.org/pkg/booktabs
+\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
+                        %% http://ctan.org/pkg/subcaption
+
+%% Cyrus packages
+\usepackage{microtype}
+\usepackage{todonotes}
+\usepackage{mdframed}
+\usepackage{colortab}
 \usepackage{mathpartir}
-\usepackage{stmaryrd}
-\usepackage{wasysym}
-\usepackage{multicol}
-\usepackage{extarrows}
-\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
-\definecolor{light-gray}{gray}{0.9}
-\usepackage{soul}
-\setulcolor{red}
-\usepackage{mathpazo}
-\usepackage{colortab}
-\usepackage{url}
-\usepackage{todonotes}
-\usepackage{listings}
-\lstset{tabsize=2,
-basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
-\usepackage{microtype}
-\sloppy
-\def \TirNameStyle #1{\small\rulename{#1}}
-\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}
+\usepackage{enumitem}
+\usepackage{bbm}
 
-\usepackage{joshuadunfield}
-\usepackage{llproof}
-\usepackage{rulelinks}
+% \newtheorem{theorem}{Theorem}[chapter]
+% \newtheorem{lemma}[theorem]{Lemma}
+% \newtheorem{corollary}[theorem]{Corollary}
+% \newtheorem{definition}[theorem]{Definition}
+\newtheorem{assumption}[theorem]{Assumption}
+\newtheorem{condition}[theorem]{Condition}
 
-%\usepackage{balance} %% Balance last page's two columns
+%% Listings
+\makeatletter\if@ACM@journal\makeatother
+%% Journal information (used by PACMPL format)
+%% Supplied to authors by publisher for camera-ready submission
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{1}
+\acmArticle{1}
+\acmYear{2017}
+\acmMonth{1}
+\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+\else\makeatother
+%% Conference information (used by SIGPLAN proceedings format)
+%% Supplied to authors by publisher for camera-ready submission
+\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
+\acmYear{2017}
+\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+\fi
 
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}{Corollary}
-\newtheorem{definition}{Definition}
-\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}
-\makeatletter
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission
+\setcopyright{none}             %% For review submission
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\copyrightyear{2017}           %% If different from \acmYear
 
-\renewcommand\topfraction{0.85}
-\renewcommand\bottomfraction{0.85}
-\renewcommand\textfraction{0.1}
-\renewcommand\floatpagefraction{0.85}
-
-\AtBeginDocument{%
- \abovedisplayskip=2pt
- \abovedisplayshortskip=0pt
- \belowdisplayskip=2pt
- \belowdisplayshortskip=0pt
-}
-
-\setlength{\floatsep}{10pt}
-\setlength{\textfloatsep}{12pt}
-
-\usepackage[compact]{titlesec}
-\titlespacing*{\section}{0pt}{4pt}{2pt}
-\titlespacing*{\subsection}{0pt}{4pt}{2pt}
-\titlespacing*{\subsubsection}{0pt}{4pt}{2pt}
-\titlespacing*{\paragraph}{0pt}{4pt}{2pt}
-\setlength{\skip\footins}{3pt plus 1px minus 5px}
-
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-
-\usepackage{enumitem}
-\makeatletter
-\def\thm@space@setup{%
-  \thm@preskip=4px plus 2px minus 2px
-  \thm@postskip=\thm@preskip % or whatever, if you don't want them to be equal
-}
-\makeatother
-
-\usepackage{todonotes}
-\usepackage{xcolor}
-%\usepackage{adjustbox}
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
 
 \input{macros}
+
 \begin{document}
 
-\conferenceinfo{-}{-}
-\copyrightyear{-}
-\copyrightdata{[to be supplied]}
+%% Title information
+\title{Hazelnut Live: Type-Theoretic Foundations for Running Incomplete Programs}         %% [Short Title] is optional;
+                                        %% when present, will be used in
+                                        %% header instead of Full Title.
+% \titlenote{with title note}             %% \titlenote is optional;
+                                        %% can be repeated if necessary;
+                                        %% contents suppressed with 'anonymous'
+% \subtitle{Subtitle}                     %% \subtitle is optional
+% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
+                                        %% can be repeated if necessary;
+                                        %% contents suppressed with 'anonymous'
 
-%\titlebanner{}        % These are ignored unless
-\preprintfooter{Draft}   % 'preprint' option specified.
+%% Author information
+%% Contents and number of authors suppressed with 'anonymous'.
+%% Each author should be introduced by \author, followed by
+%% \authornote (optional), \orcid (optional), \affiliation, and
+%% \email.
+%% An author may have multiple affiliations and/or emails; repeat the
+%% appropriate command.
+%% Many elements are not rendered, but should be provided for metadata
+%% extraction tools.
 
-\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
- Calculus}
-% \title{Hazelnut:\\A Bidirectionally Typed Structure Editor Calculus}
+%% Author with single affiliation.
+\author{First1 Last1}
+\authornote{with author1 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  \position{Position1}
+  \department{Department1}              %% \department is recommended
+  \institution{Institution1}            %% \institution is required
+  \streetaddress{Street1 Address1}
+  \city{City1}
+  \state{State1}
+  \postcode{Post-Code1}
+  \country{Country1}
+}
+\email{first1.last1@inst1.edu}          %% \email is recommended
 
-%\authorinfo{\vspace{-2px}}{}{}
-%\authorinfo{~}{~}{\vspace{-10px}}
-%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
-%         {Carnegie Mellon University}
-%         {\{comar, aldrich\}@cs.cmu.edu}
-\iftr
-\authorinfo{
-        Cyrus Omar$^1$
-        \and Ian Voysey$^1$
-        \and Michael Hilton$^2$
-        \and Jonathan Aldrich$^1$
-        \and Matthew A. Hammer$^3$
+%% Author with two affiliations and emails.
+\author{First2 Last2}
+\authornote{with author2 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  \position{Position2a}
+  \department{Department2a}             %% \department is recommended
+  \institution{Institution2a}           %% \institution is required
+  \streetaddress{Street2a Address2a}
+  \city{City2a}
+  \state{State2a}
+  \postcode{Post-Code2a}
+  \country{Country2a}
 }
-{
- $^1$~Carnegie Mellon University
- \and
- $^2$~Oregon State University
- \and
- $^3$~University of Colorado Boulder\vspace{-10px}
+\email{first2.last2@inst2a.com}         %% \email is recommended
+\affiliation{
+  \position{Position2b}
+  \department{Department2b}             %% \department is recommended
+  \institution{Institution2b}           %% \institution is required
+  \streetaddress{Street3b Address2b}
+  \city{City2b}
+  \state{State2b}
+  \postcode{Post-Code2b}
+  \country{Country2b}
 }
-{
-%\email{comar@cs.cmu.edu}\\
-%\texttt{http://www.cs.cmu.edu/\homedir comar/}
-%\email{hiltonm@eecs.oregonstate.edu}\\
-}
-\else
-\authorinfo{~}{~}{\vspace{-10px}}
-\fi
+\email{first2.last2@inst2b.org}         %% \email is recommended
 
-%
-%\authorrunning{Omar et al.} % abbreviated author list (for running head)
-%
-%%%% list of authors for the TOC (use if author list has to be modified)
-%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
-%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
-%
-%\email{Matthew.Hammer@colorado.edu}
+%% Paper note
+%% The \thanks command may be used to create a "paper note" ---
+%% similar to a title note or an author note, but not explicitly
+%% associated with a particular element.  It will appear immediately
+%% above the permission/copyright statement.
+% \thanks{with paper note}                %% \thanks is optional
+                                        %% can be repeated if necesary
+                                        %% contents suppressed with 'anonymous'
 
-\maketitle
+%% Abstract
+%% Note: \begin{abstract}...\end{abstract} environment must come
+%% before \maketitle command
 \begin{abstract}
-% Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker.
-Not all text survives this journey.
-In particular, text that arises transiently, or when the programmer makes a mistake, is often malformed or ill-typed.
-% Not all text that arises during the programming process survives the journey through a parser and typechecker.
-In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
-% Contending with malformed program text or well-formed but meaningless syntax trees is difficult for programmers and their tools alike.
-
-%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
-% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}.
-They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
-% Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.
-
-\emph{Structure editors} allow programmers to edit the tree structure of a program directly.
-This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they do not need to contend with malformed program text.
-
-This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: it's {edit actions} operate over statically meaningful (i.e.
-well-typed) terms.
-Na\"ively, this prohibition on ill-typed edit states would force the programmer to construct terms in a rigid ``outside-in'' manner.
-To avoid this problem, the {action semantics} automatically places terms assigned a type that is inconsistent with the expected type {inside} a {hole}.
-This safely defers the type consistency check until the term inside the hole is \emph{finished}.
-
-Hazelnut is a foundational type-theoretic account of typed structure editing, rather than an end-user tool itself.
-To that end, we describe how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to the calculus.
-We also discuss various plausible evaluation strategies for terms with holes, and in so doing reveal connections with gradual typing and contextual modal type theory (the Curry-Howard interpretation of contextual modal logic.) Finally, we  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program.
-Our reference implementation is written using \lstinline{js_of_ocaml}.
-
-%Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
+\todo{abstract}
 \end{abstract}
 
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
-%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
-%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
-%\keywords
-%extensible languages; module systems; type abstraction; typed compilation; type-level computation
+%% 2012 ACM Computing Classification System (CSS) concepts
+%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+% \begin{CCSXML}
+% <ccs2012>
+% <concept>
+% <concept_id>10011007.10011006.10011008</concept_id>
+% <concept_desc>Software and its engineering~General programming languages</concept_desc>
+% <concept_significance>500</concept_significance>
+% </concept>
+% <concept>
+% <concept_id>10003456.10003457.10003521.10003525</concept_id>
+% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
+% <concept_significance>300</concept_significance>
+% </concept>
+% </ccs2012>
+% \end{CCSXML}
 
-\section{Introduction}\label{sec:introduction}
-\input{intro}
+% \ccsdesc[500]{Software and its engineering~General programming languages}
+% \ccsdesc[300]{Social and professional topics~History of programming languages}
+%% End of generated code
 
-\section{Programming in Hazelnut}\label{sec:example}
-\input{example}
+%% Keywords
+%% comma separated list
+% \keywords{keyword1, keyword2, keyword3}  %% \keywords is optional
 
-\section{Hazelnut, Formally}
-\label{sec:hazel}
-\input{formally}
+%% \maketitle
+%% Note: \maketitle command must come after title commands, author
+%% commands, abstract environment, Computing Classification System
+%% environment and commands, and keywords command.
+\maketitle
 
-\subsection{Mechanization}
-\label{sec:mech}\label{sec:mt}
-\input{agda}
+\newcommand{\introSec}{Introduction}
+\section{\protect\introSec}
+\label{sec:intro}
+\todo{intro}
 
-% \clearpage
-
-% \section{Dynamics of Hazelnut}\label{sec:dynamics}
-
-% \subsection{Complete Expressions}
-% ... canonical forms and type safety with e complete in it ...
-
-% \subsection{Holes as TODOs}
-% ... add an e err and define new type safety ...
-
-% \subsection{Indeterminate Expressions}
-% \input{dynamics}
-
-% \clearpage
-
-\section{Extending Hazelnut}\label{sec:extending}
-%% \todo{write this section} Add sum types.
-%% H-expression
-%% Z-expression + Erasure rules
-%% Statics
-%% Movement actions
-%% Construct actions
-\input{sumtypes}
-
-\section{Implementation}
-\label{sec:impl}
-\input{implementation}
-
-\section{Related Work and Discussion}\label{sec:rw}
-%\subsection{Structure Editors}
-\subsection{Structure Editors}
-Syntactic structure editors have a long history -- the Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in 1981.
-Novice programmers have been a common target for structure editors.
-For example,
-GNOME~\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates. 
-Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students.
-Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure editor targeted at children ages 8 to 16.
-TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students.
-An implementation of Hazelnut might be useful in teaching students about the typed lambda calculus, though that has not been our explicit aim with this work.
-
-Not all structure editors are for educational purposes. For example,
-mbeddr \cite{voelter_mbeddr:_2012} is a structure editor for a C-based programming language (nominally, for programming embedded systems.)  
-Lamdu~\cite{lamdu}, like Hazelnut, is a structure editor for a statically typed functional language. It is designed for use by professional programmers.
-
-The examples given so far either define a language with a trivial static semantics, or do not attempt to maintain well-typedness as an edit invariant.
-This can pose problems, for reasons discussed in the Introduction.
-One apparent exception is Unison~\cite{unison}, a structure editor for a typed functional language similar to Haskell.
-Like Hazelnut, it seems to define some notion of well-typedness for expressions with holes (though there is no technical documentation on virtually any aspect of its design.) Unlike Hazelnut, it does not have a notion analagous to Hazelnut's notion of a non-empty hole.
-As such, programmers must construct programs in a rigid outside-in manner, as discussed in Sec.
-\ref{sec:example}.
-Another system with the same problem is Polymorphic Blocks, a block-based user interface where the structure of block connectors encodes a type \cite{DBLP:conf/chi/LernerFG15}.
-
-We fundamentally differ from these projects in our design philosophy: we consider it essential to start by building type theoretic foundations, which are independent of nearly all decisions about the user interface.
-In contrast, these editors have developed innovative user interfaces (e.g.
-see the discussion in \cite{DBLP:conf/sle/VolterSBK14}) but lack a principled foundational calculus.
-In this respect, we follow the philosophical approach taken by languages that are rooted in the type theoretic tradition and have gone to great effort to develop a clear metatheory, like Standard ML \cite{mthm97-for-dart,Harper00atype-theoretic}.
-In the future, we hope that these lines of research will merge to produce a human-usable typed structure editor with sound formal foundations.
-Our contribution, then, is in defining and analyzing the theoretical properties of a small foundational calculus that could be extended to achieve this vision.
-
-Some structure editor generators do operate on formal or semi-formal definitions of an underlying language.
-For example, the Synthesizer Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an attribute grammar-based language implementation that then can be used to generate a structured editor.
-CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language.
-Barista is a programmatic toolkit for building structure editors \cite{ko_barista:_2006}.
-mbeddr is built on top of the commercial JetBrains MPS framework for constructing structure editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}.
-These systems do not give a semantics to the edit states of the structure editor itself, or maintain non-trivial edit invariants, as Hazelnut does.
-%These early systems were developed  of the systems are rooted in the type-theoretic tradition.
-
-Related to structure editors are value editors, which operate directly on simple values (but not generally expressions or functions) of a programming language.
-For example, Eros is a typed value editor based in Haskell \cite{DBLP:conf/icfp/Elliott07}.
-
-\subsection{Gradual Type Systems}
-A significant contribution of this paper is the discovery of a clear technical relationship between typed structure editing and gradual typing.
-In particular, the machinery necessary to give a reasonable semantics to type holes -- i.e.
-type consistency and type matching -- coincides with that developed in the study of gradual type systems for functional languages.
-The pioneering work of Siek and Taha \cite{Siek06a} introduced type consistency.
-Subsequent work developed the concept of type matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}.
-In retrospect, this relationship is perhaps unsurprising: gradual typing is, notionally, also motivated by the idea of iterated development of a program where every intermediate state is well-defined in some sense.
-%Typed structure editing endows this intuitive notion with technical force.
-
-Recent work has discovered a systematic procedure for generating a ``gradual version'' of a standard type system \cite{DBLP:conf/popl/CiminiS16}.
-This system, called the Gradualizer, operates on a logic program that specifies a simple type assignment system with some additional annotations to generate a corresponding specification of a gradual type system.
-The authors leave support for working with bidirectional type systems as future work.
-This suggests the possibility of an analagous ``Editorializer'' that generates a specification of a typed structure editor from a simple language definition.
-Our exposition in Sec.
-\ref{sec:extending} certainly suggests that many of the necessary definitions follow seemingly mechanically from the definition of the static semantics, and the relationship with gradual typing suggests that many of the technical details of this transformation may already exist in the Gradualizer.
-One possibility we have explored informally is to use Agda's reflection features to implement such a system.
-
-An aspect of gradual typing that we did not touch on directly here is its concern with  assigning a dynamics to programs where type information is not known, by inserting dynamic type casts.
-This would correspond to assigning a dynamics to Hazelnut expressions with type holes such that a run-time error occurs when a type hole is found to be unfillable through evaluation.
-
-\subsection{Type Holes as Unification Variables}
-Another possible interpretation of type holes is as explicit unification variables.
-In other words, we might define the dynamics of Hazelnut such that a program cannot be run if some automated type inference procedure cannot statically fill in any type holes that arise.
-For a simple calculus, e.g.
-the STLC upon which Hazelnut is based, type inference is known to be decidable \cite{damas1982principal}.
-In more complex settings, e.g.
-in a dependently typed language, a partial decision procedure may still be useful in this regard, both at edit-time and (just prior to) run-time.
-Indeed, text editor modes for proof assistants, e.g.
-for Agda, attempt to do exactly this for indicated ``type holes'' (and do not always succeed.)
-
-\subsection{Exceptions}
-Expression holes, too, could dynamically be interpreted in several ways.
-One straightforward approach would be to dynamically raise an exception if evaluation encounters an expression hole.
-Indeed, placing \textt{raise Unimplemented} or similar in portions of an expression that are under construction is a common practice across programming languages.
-
-\subsection{Type-Directed Program Synthesis}
-Some text editor modes, e.g.
-those for proof assistants like Agda \cite{norell:thesis} and Idris \cite{brady2013idris}, support a more explicit hole-based programming model where indicated expression holes are treated as sites where the system can be asked to automatically generate an expression of an appropriate type.
-These systems are not statically well-defined themselves (though see below.)
-
-The Graphite system used an informal model of typed holes in Java to allow library providers to associate interactive code generation interfaces with types \cite{Omar:2012:ACC:2337223.2337324}.
-%RThis system was limited in that these interfaces could not themselves manipulate terms.
-
-More generally, the topic of type-directed program synthesis an active area of research, e.g.
-\cite{DBLP:conf/pldi/OseraZ15}.
-By maintaining static well-definedness throughout the editing process, Hazelnut provides researchers interested in editor-integrated type-directed program synthesis  with a formal foundation upon which to build.% such systems could be integrated directly into editors.
-
-\subsection{Contextual Modal Type Theory}
-Expression holes can also be understood by invoking the notion of a \emph{metavariable} as found in contextual modal type theory (CMTT) \cite{DBLP:journals/tocl/NanevskiPP08}.
-In particular, expression holes have types and are surrounded by contexts, just as metavariables in CMTT are associated with types and contexts.
-This helps to clarify the logical meaning of a typing derivation in Hazelnut -- it conveys well-typedness relative to an (implicit) modal context that extracts each expression hole's type and context.
-The modal context must be emptied -- i.e.
-the expression holes must be instantiated with expressions of the proper type in the proper context -- before the expression can be considered complete.
-This corresponds to the notion of modal necessity in contextual modal logic.
-
-Making the modal context explicit in our semantics is not technically useful given our stated purpose -- interactive program editing is not merely hole filling in Hazelnut (i.e.
-the cursor need not be on a hole).
-Moreover, the hole's type and context become apparent as our action semantics traverses the zipper structure on each step.
-Some interactive proof assistants, however, support a tactic model based directly on hole filling, so the connection to CMTT and similar systems is more useful.
-For example,  Beluga \cite{DBLP:conf/flops/Pientka10} is based on dependent CMTT and aspects of Idris' editor support \cite{brady2013idris} are based on McBride's OLEG \cite{mcbride2000dependently} and Lee and Friedman have explored a lambda calculus with contexts for a similar purpose \cite{DBLP:conf/icfp/LeeF96}.
-
-One interesting avenue of future work is to elaborate expression holes to CMTT's closures, i.e.
-CMTT terms of the form $\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable associated with each hole and $\text{id}(\Gamma)$ is the explicit identity substitution.
-This would allow us to evaluate expressions with holes such that the closure ``accumulates'' substitutions explicitly.
-When evaluation gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped with a notion of progress under a non-empty modal context), it would then be possible for the programmer to choose holes from amongst the visible holes (which may have been duplicated) to edit in their original context.
-Once finished, the CMTT hole instantiation operation, together with a metatheorem that establishes that reduction commutes with instantiation, would enable an ``edit and resume'' feature with a clear formal basis.
-This notion of reduction commuting with instantiation has also been studied in other calculi \cite{DBLP:journals/entcs/Sands97}.
-Being able to edit a running program also has connections to less formal work on ``live programming'' interfaces \cite{burckhardt2013s,lamdu}.
-%, e.g.
-
-\subsection{Evaluation Strategies: A High-Dimensional Space}
-To summarize, we have discussed three different evaluation strategies in the presence of type holes:
-\begin{enumerate}[noitemsep]
-\item ...as preventing evaluation (the standard approach.)
-\item ...as unknown types, in the gradual sense.
-\item ...as unification variables.
-\end{enumerate}
-In addition, we have discussed four different evaluation strategies in the presence of expression holes:
-\begin{enumerate}[noitemsep]
-\item ...as preventing evaluation (the standard approach.)
-\item ...as causing exceptions.
-\item ...as sites for automatic program synthesis.
-\item ...as the closures of CMTT.
-\end{enumerate}
-
-Every combination of these choices could well be considered in the design of a full-scale programming system in the spirit of Hazelnut.
-Indeed, the user might be given several options from among these combinations, depending on their usage scenario.
-Many of these warrant further inquiry.
-
-% \todo{Eclipse parsing partial programs // other heuristics}
-
-% \paragraph{Gradual typing: Has Matched Relation}
-% %
-% A. Rastogi, A. Chaudhuri, and B. Hosmer. The ins and outs of gradual type inference.
-% \\
-% Definition 3.1 is similar to our ``has matched'' relations:
-% \\
-% \url{https://www.cs.umd.edu/~avik/papers/iogti.pdf}
-
-% This paper defines ``has matched'' relation for arrow, just like us:
-% \\
-% \url{http://cimini.info/publications/Gradualizer_Draft.pdf}
-% \\
-% They credit this technique to the following paper:
-% \\
-% \url{http://www.cs.ubc.ca/~rxg/ptsgp.pdf}
-% \\
-% But, the Rastogi paper came earlier.
-
-% \todo{TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}}
-
-% Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g.
-Emacs.) These systems also have notions of holes and use types to guide the user toward filling these holes.
-These  systems are also, to our knowledge, not formally well-defined but rather exist only as part of system implementations.
-
-% Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified.
-For example, while Lamdu has many interesting features, there is no theoretical basis presented for their work -- it is a rather large body of Haskell code with an unclear (and indeed, often somewhat perplexing, in our experience) action model.
-Unison is also a rather large body of Haskell code, though its action model appears superficially more similar to ours.
-We maintain what we believe to be a stronger action sensibility invariant than Unison (i.e.
-in Unison, one must construct expressions from the outside-in.) These systems are rich sources of interesting ideas, however -- there is room enough for many different approaches in this (re-)emerging space.
-
-% \todo{TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}}
-
-% \todo{TODO: cite} \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}
-
-%Drag-and-drop / for novices: lots of examples, e.g. Alice and others
-%
-%Contemporary: Lamdu, MPS/Mbeddr, TouchDevelop
-%
-%Hybrid: Cyrus' active code completion paper
-
-%\subsection{Refactoring Models}
-%(Michael, can you fill this section out?)
-
-%\subsection{Formal Editor Models}
-%Need to do a search to see what else has been done...
-
-\section{Conclusion}
-\label{sec:future}
-This paper presented Hazelnut, a type theoretic structure editor calculus.
-Our aim was to take a principled approach to its design by formally defining its static semantics as well as its action semantics and developing a rich metatheory.
-Moreover, we have mechanized substantial portions of the metatheory, including the crucial Sensibility theorem that establishes that every edit state is statically meaningful.
-
-In addition to simplifying the job of an editor designer, typed structure editors also promise to increase the speed of development by eliminating redundant syntax and supporting higher-level primitive actions.
-However, we did not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
-whether a keyboard or a mouse is in use.
-Indeed, we consider it a virtue of this work that such implementation details do not enter into our design.
-
-% By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
-% By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?
-In progress, unfinished, incomple?} and yet still in a well defined type state.
-% This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.\todo{integrate this}
-
-\subsection{Future Work}
-\subsubsection{Richer Languages}
-Hazelnut is, obviously, a very limited language at its core.
-So the most obvious avenue for future work is to increase the expressive power of this language by extension.
-Our plan is to simultaneously maintain a mechanization and implementation (following, for example, Standard ML) as we proceed, ultimately producing the first large-scale, formally verified bidirectionally typed structure editor.
-
-It is interesting to note that the demarcation between the language and the editor is fuzzy (indeed, non-existent) in Hazelnut.
-There may well be interesting opportunities in language design when the language is being codesigned with a typed structure editor.
-It may be that certain language features are unnecessary given a sufficiently advanced type-aware structure editor (e.g.
-SML's \texttt{open}?), while other features may only be practical with editor support.
-We intend to use Hazelnut and derivative systems thereof as a platform for rigorously exploring such questions.
-
-\subsubsection{Editor Services}
-There are various aspects of the editor model that we have not yet formalized.
-For example, our action model does not consider how actions are actually entered using, for example, key combinations or chords.
-In practice, we would want also to rank available actions in some reasonable manner (perhaps based on usage data gathered from other users or code repositories.) Designing a rigorous typed probability model over actions and H-expressions is one avenue of research that we have started to explore, with intriguing initial results.
-
-\subsubsection{Programmable Actions}
-Our language of actions is intentionally primitive.
-However, even now it acts much like a simple imperative command language.
-This suggests future expansion to, for example, a true tactic language.
-Alternatively, it may be more useful to develop the notion of an \emph{action macro}, whereby functional programs could themselves be lifted to the level of actions to compute non-trivial compound actions.
-
-\subsubsection{Views}
-Another research direction is in exploring how types can be used to control the presentation of expressions in the editor.
-For example, following an approach developed in a textual setting of developing \emph{type-specific languages} (TSLs), it should be possible to have the type that an expression is being analyzed against define alternative display forms and interaction modes \cite{TSLs}.
-
-It should also be possible to develop a semantics of semantic comments, i.e.
-comments that mention semantic structures.
-These would be subject to the same operations, e.g.
-renaming, as other structures, helping to address the problem of comments becoming inconsistent with code.
-
-\subsubsection{Collaborative Programming}
-Finally, we did not consider any aspects of \emph{collaborative programming}, such as a packaging system, a differencing algorithm for use in a source control system, support for multiple simultaneous focii for different users, and so on.
-These are all interesting avenues for future work.
-
-\subsubsection{Empirical Evaluation}
-Although we make few empirical claims in this paper, it is ultimately an empirical question as to whether structure editors, and typed structure editors, are practical.
-We hope to conduct user studies once a richer semantics has been developed.
-
-\subsubsection{More Theory}
-Connections with gradual type systems and CMTT, discussed in the previous section, seem likely to continue to be revealing.
-
-The notion of having one of many possible locations within a term under a cursor has a very strong intuitive connection to the proof theoretic notion of focusing \cite{Simmons11tr}.
-Building closer connections with proof theory (and category theory) is likely to be a fruitful avenue of further inquiry.
-
-% We already discussed a connection to gradual typing \cite{Siek06a}.
-We hope to explore this connection more thoroughly.
-In particular, it may be possible to better support exploratory and live programming by allowing even programs with holes in them to execute as long as those holes are only in the type portions, by deferring to the semantics given in work on gradual typing.
-
-% It may also be possible to give a dynamics to incomplete expressions.
-Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}.
-In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and
-% $\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
-% explored an equational theory for terms with holes, but intend to once our
-% formalization effort is more mature.\todo{fix bib}
-
-% \todo{new action form that makes actions extensible given proofs of
-% admissiblity of a derived form, like prevSib}
-
-\begin{quote}
-\emph{In any case, these are but steps toward more graphical program-description
-systems, for we will not forever stay confined to mere strings of symbols.}
-
---- Marvin Minsky, Turing Award lecture \cite{DBLP:journals/jacm/Minsky70}
-\end{quote}
-
-\clearpage
-% We recommend abbrvnat bibliography style.
-\bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-\bibliography{bibliography}
-%\softraggedright
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\iftr
-\clearpage
-%\onecolumn
-\appendix
-\input{appendix}
-\else
-% No Appendix Here
-\fi
+\begin{figure}
 

### 2017-05-03 calculus.tex
added figure outlines and basic judgment forms

% !TEX root = hazelnut-dynamics.tex

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}
\todo{exposition}
\todo{add sums and products}

\begin{figure}[h]
...
\caption{Syntax of H-expressions and H-types.}
\end{figure}

\begin{figure}[h]
\judgbox{\tconsistent{\htau_1}{\htau_2}}{$\htau_1$ is consistent with $\htau_2$}
...

\judgbox{\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}}{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
...
\caption{Type consistency and matched types.}
\end{figure}

\begin{figure}[h]
\judgbox{\expandSyn{\hGamma}{\hexp}{M}{\htau}{\Delta}}{ }
...

\judgbox{\expandAna{\hGamma}{\hexp}{M}{\htau}{\Delta}}{ }
...
\caption{Bidirectionally Typed Expansion}
\end{figure}

\begin{figure}[h!]
\judgbox{\hasType{\hGamma}{M}{\htau}{\Delta}}{$M$ has type $\htau$ with holes $\Delta$}
...
\caption{Declarative Statics}
\end{figure}

\begin{figure}[h!]
\judgbox{\isValue{M}}{$M$ is a value}
...

\judgbox{\isIndet{M}}{$M$ is indeterminate}
...

\judgbox{\stepsTo{M}{N}}{$M$ steps to $N$}
...
\caption{Structural Dynamics}
\end{figure}

### 2017-05-03 intro.tex
added figure outlines and basic judgment forms

@@
-% !TEX root = hazelnut-popl17.tex
-
-%% There are some benefits to this approach, to be sure, but the structural
-%% mismatch between programs and their textual representations also imposes
-%% various burdens.  For example, the primitive edit actions available in a
-%% text editor (e.g. inserting or deleting a character or word) do not
-%% always correspond to sensible structural transformations.
-
-% spj: describe the problem; state our contributions; STOP. one page max.
-
-% When constructing a program or proof in a language with rich type
-% structure, skilled programmers generally follow a \emph{type discipline}
-% where they first determine the type of the expression that they are
-% constructing in order to constrain the mental search space that they are
-% operating within.
-
-% For example, if the programmer knows that an expression of type
-% $\tarr{\tnum}{\tnum}$ is needed, then it is often the case (though, of course, not
-% necessarily the case) that the expression will take the form $$\hlam{\mathit{x}}{e}$$
-% for some variable $x$ and function body $e$. If the programmer chooses this
-% form, then after picking a suitable variable name, her focus will be on
-% constructing a suitable body, $e$. Following the type discipline, $e$ must
-% be of type $\tnum$, and so this process can begin anew.
-
-% The problem is that when using a text
-% editor to construct a program, it is easy, and indeed necessary, to deviate from this disciplined process.  Rather, text editors operate on sequences of
-% characters (i.e. \emph{text}.)
-
-%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
-% Programming languages, and therefore programs themselves, are rich mathematical structures.
-Programmers typically construct and manipulate well-typed expressions only indirectly, by editing text that is first parsed according to a textual syntax and then typechecked according to a static semantics.
-This indirection has practical benefits, to be sure -- text editors and other text-based tools benefit from decades of development effort -- but it also introduces some fundamental complexity into the programming process.
-
-% Every day, programmers use text-based tools to construct and manipulate programs.  
-% These programs are written in languages which defined by a textual syntax. 
-% While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.
-
-% It also complicates matters for tool designers because tools are often confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}.
-This may be because the programmer is in the midst of a sequence of
-% edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake.
-The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g.
-by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)
-%Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
-%Most programming languages define a textual syntax.
-This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs.
-While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.
-
-First, it requires that programmers learn the subtleties of the textual syntax (e.g.
-the precedence of the various forms.) This can be particularly challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258}.
-%For example, one study of novice programmers found syntax errors (e.g.
-unbalanced parentheses) to be the most common class of errors.
-% Too much detail?
-
-The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers.
-In particular, program editors must contend with meaningless text on a regular basis.
-In a dataset analyzed by Yoon and Myers consisting of 1460 hours of edit logs, 44.2\% of edit states were syntactically malformed \cite{6883030}.
-Some additional percentage of edit states were well-formed but ill-typed (the dataset did not contain enough information to determine the exact percentage.)
-Some of these meaningless edit states arose because the programmer was in the midst of a sequence of edits that left the edit state transiently meaningless, while others arose because the programmer made a logical mistake.
-Because the language definitions is silent about these edit states, it is difficult to design useful editor services, e.g.
-syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \cite{mens2004survey}.
-Editors must either disable these editor services when they encounter meaningless edit states or develop \emph{ad hoc} heuristics, which can be misleading.
-%, or because the programmer has made a mistake.
-% The syntax definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action
-
-% Many editors have developed \emph{ad hoc} workarounds for this problem, e.g.
-they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or continue past a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer.
-% This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.
-
-% stuck editing a representation of the program instead of the structures
-% themselves. The editor does not restrict what the programmer may do: you
-% can delete characters that belong, insert ones that don't, forget things
-% that were needed, and so. There's nothing stopping us from accidentally
-% writing $$\lambda \mathit{x:num}.\mathit{(x,x)}$$ even though it's obvious
-% that building a pair can't hope to form a natural number.
-
-% The type structure of the language makes this sort of error obvious: it's
-% not that you're adding characters that make your program incorrect, or even
-% malformed; you're adding characters that can't possibly create a structure
-% you want because of the type. Simply put, the primitive operations
-% available in text editors do not always correspond to sensible
-% transformations on the structure of the program.
-These complications have motivated a long line of research into \emph{structure editors}, i.e.
-program editors where every edit state is a program structure.
-% Eliminating text eliminates the possibility of syntax errors.
-
-Most structure editors are \emph{syntactic structure editors}, i.e.
-the edit state is a syntax tree with \emph{holes} that stand for branches of the tree that have yet to be constructed, and the edit actions are context-free tree transformations.
-For example, Scratch is a syntactic structure editor that has achieved success as a tool for teaching children how to program \cite{Resnick:2009:SP:1592761.1592779}.
-
-The Scratch language has a trivial static semantics, but researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics.
-For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}.
-Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design.
-The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed.
-As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or deploy \emph{ad hoc} heuristics.
-%Indeed, it is not entirely clear what it should mean for terms with holes to be statically meaningful. 
-% Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.
-
-% \begin{itemize}
-% \item
-
-% First, there is an all too familiar problem: these languages do not have a clear formal semantics and metatheory, i.e.
-``the implementation is the specification''.
-This makes it difficult to reason methodically about types and binding, so, in turn, it is difficult to design reliable editor services.
-
-% Even if a formal semantics and metatheory for the underlying language were forthcoming (e.g.
-following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}), it would go only part of the way towards addressing the challenges faced by structure editor designers.
-First, a standard semantics assigns no formal meaning to \emph{incomplete terms}, i.e.
-terms with holes.
-Moreover, a syntactic structure editor  -- so incorporating holes into the static semantics would in any case be only a minor salve.
-Heuristics analagous to those that pervade textual program editors would still be needed to contend with the ill-typed edit states.
-
-This paper develops a principled solution to this problem.
-We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}.
-Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e.
-well-typed) H-expression with a single superimposed \emph{cursor}.
-We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because our encoding follows Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
-a function argument), only actions consistent with that type are permitted.
-
-Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner.
-For example, the programmer would often need to construct the outer function application form before identifying the intended function.
-To address this problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole if the expression's type is inconsistent with the expected type.
-This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}.
-In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remain under construction.
-%In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}.
-%Actions act at a programmer-indicate subtree, called the \emph{focus} (which is defined following Huet's zipper pattern.)
-
-% By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state.
-%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics.
-Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state.
-
-%However, these syntactically correct states can be semantically meaningless (i.e.
-undefined), because the language definitions generally only give meaning to complete, well-typed terms.
-(e.g., when a branching element is introduced, the program does not become semantically correct until both branches are complete)
-% This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
-% Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
-%Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g.
-Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .
-
-The remainder of this paper is organized as follows:
-\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
-\item We begin in Sec. 
-    \ref{sec:example} with two examples of edit sequences to develop the reader's intuitions.  
-\item We then give a detailed overview of Hazelnut's semantics and metatheory, and our mechanization in the Agda proof assistant, in Sec.  \ref{sec:hazel}.
-\item Hazelnut is a {foundational} calculus, i.e.
-a calculus that language and editor designers are expected to extend with higher level constructs.
-We extend Hazelnut with simple sum types in Sec.
-\ref{sec:extending} to demonstrate how Hazelnut's rich metatheory guides such a development.
-\item In Sec.
-\ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program.
-Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.
-
-% \item However, we consider the space of possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}. 
-% so doing, we discover interesting connections with gradual typing and contextual modal type theory.
-The former provides an interpretation of type holes, and the latter provides a logical interpretation of expression holes.
-\todo{revise this one}
-
-\item In Sec.
-\ref{sec:rw}, we summarize related work.
-In particular, much of the technical machinery needed to handle type holes coincides with machinery developed in the study of gradual type systems.
-Similarly, expression holes can be interpreted as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work.
-%This suggests a principled design for an ``edit and resume'' feature.
-
-\item We conclude in Sec.  \ref{sec:future} by summarizing our vision of a principled science of structure editor design rooted in type theory, and suggest a number of future directions.
-\end{itemize} 
-The supplemental material includes 1) a typeset listing of Hazelnut's rules in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our reference implementation of Hazelnut, both in source form and pre-compiled to JavaScript for use in a web browser.
+% !TEX root = hazelnut-dynamics.tex
+\newcommand{\introSec}{Introduction}
+\section{\protect\introSec}
+\label{sec:intro}
+\todo{intro}

### 2017-05-28 calculus.tex
added figure labels and copied over syntax and type consistency defns from popl paper
macro for declarative H-expressions
draft of dynamics for nums, arrow, and sums
fix typos; add is-final to premises of i-add_1 and i-add_2
decorations on holes in dynamics
substitution
dunno
grammar
syntax figure caption
updated type consistency and matched types judgements with products and sums
commented out various complexities
algorithmic expansion rules
type assignment rules

@@
 \todo{exposition}
-\todo{add sums and products}
 
 \begin{figure}[h]
-...
-\caption{Syntax of H-expressions and H-types.}
+$\arraycolsep=4pt\begin{array}{rllllll}
+\mathsf{HTyp} & \htau & ::= &
+  b ~\vert~
+  \tarr{\htau}{\htau} ~\vert~
+  % \tprod{\htau}{\htau} ~\vert~
+  % \tsum{\htau}{\htau} ~\vert~
+  \tehole\\
+\mathsf{HExp} & \hexp & ::= &
+  c ~\vert~ 
+  x ~\vert~
+  \halam{x}{\htau}{\hexp} ~\vert~
+  \hap{\hexp}{\hexp} ~\vert~
+  % \hpair{\hexp}{\hexp} ~\vert~
+  % \hprj{i}{\hexp} ~\vert~
+  % \hinj{i}{\hexp} ~\vert~
+  % \hcase{\hexp}{x}{\hexp}{x}{\hexp} ~\vert~
+  % \hadd{\hexp}{\hexp} ~\vert~
+  \hehole{u} ~\vert~
+  \hhole{\hexp}{u} ~\vert~
+  {\hlam{x}{\hexp}} ~\vert~
+  \hexp : \htau\\
+ \mathsf{DHExp} & \dexp  & ::= &
+  c ~\vert~
+  x ~\vert~
+  {\halam{x}{\htau}{\dexp}} ~\vert~
+  \hap{\dexp}{\dexp} ~\vert~
+  % \hpair{\dexp}{\dexp} ~\vert~
+  % \hprj{i}{\dexp} ~\vert~
+  % \hinj{i}{\dexp} ~\vert~
+  % \hcase{\dexp}{x}{\dexp}{x}{\dexp} ~\vert~
+  % \hadd{\dexp}{\dexp} ~\vert~
+  \dehole{\mvar}{\subst} ~\vert~
+  \dhole{\dexp}{\mvar}{\subst} ~\vert~
+  \dcast{\dexp}{\htau}\\
+\end{array}$
+\caption{Syntax of H-types, H-expressions and dynamic H-expressions. We write $x$ to range over variables, $u$ over hole names and $\sigma$ over finite substitutions.}
+\label{fig:HTyp}
+\label{fig:HExp}
 \end{figure}
@@
 \judgbox{\tconsistent{\htau_1}{\htau_2}}{$\htau_1$ is consistent with $\htau_2$}
-...
+\begin{mathpar}
+\inferrule{ }{
+  \tconsistent{\tehole}{\htau}
+}
+
+\inferrule{ }{
+  \tconsistent{\htau}{\tehole}
+}
+
+\inferrule{ }{
+  \tconsistent{\htau}{\htau}
+}
+
+\inferrule{
+  \tconsistent{\htau_1}{\htau_1'}\\
+  \tconsistent{\htau_2}{\htau_2'}
+}{
+  \tconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
+}
+%
+% \inferrule{
+%   \tconsistent{\htau_1}{\htau_1'}\\
+%   \tconsistent{\htau_2}{\htau_2'}
+% }{
+%   \tconsistent{\tprod{\htau_1}{\htau_2}}{\tprod{\htau_1'}{\htau_2'}}
+% }
+%
+% \inferrule{
+%   \tconsistent{\htau_1}{\htau_1'}\\
+%   \tconsistent{\htau_2}{\htau_2'}
+% }{
+%   \tconsistent{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1'}{\htau_2'}}
+% }
+\end{mathpar}
 
 \judgbox{\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}}{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
-...
-\caption{Type consistency and matched types.}
+\begin{mathpar}
+\inferrule{ }{
+  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
+}
+
+\inferrule{ }{
+  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
+}
+\end{mathpar}
+
+% \judgbox{\prodmatch{\htau}{\tprod{\htau_1}{\htau_2}}}{$\htau$ has matched product type $\tprod{\htau_1}{\htau_2}$}
+% \begin{mathpar}
+% \inferrule{ }{
+%   \prodmatch{\tehole}{\tprod{\tehole}{\tehole}}
+% }
+
+% \inferrule{ }{
+%   \prodmatch{\tprod{\htau_1}{\htau_2}}{\tprod{\htau_1}{\htau_2}}
+% }
+% \end{mathpar}
+
+% \judgbox{\summatch{\htau}{\tsum{\htau_1}{\htau_2}}}{$\htau$ has matched sum type $\tsum{\htau_1}{\htau_2}$}
+% \begin{mathpar}
+% \inferrule{ }{
+%   \summatch{\tehole}{\tsum{\tehole}{\tehole}}
+% }
+
+% \inferrule{ }{
+%   \summatch{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1}{\htau_2}}
+% }
+% \end{mathpar}
+\caption{Type Consistency and Matched Types}
+\label{fig:tconsistent}
+\label{fig:arrmatch}
 \end{figure}
@@
 \begin{figure}[h]
-\judgbox{\expandSyn{\hGamma}{\hexp}{M}{\htau}{\Delta}}{ }
-...
+\judgbox{\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}}{ }
+\begin{mathpar}
+\inferrule[expand-constant]{ }{
+  \expandSyn{\hGamma}{c}{b}{c}{\emptyset}
+}
 
-\judgbox{\expandAna{\hGamma}{\hexp}{M}{\htau}{\Delta}}{ }
-...
-\caption{Bidirectionally Typed Expansion}
+\inferrule[expand-var]{ }{
+  \expandSyn{\hGamma, x : \htau}{x}{\htau}{x}{\emptyset}
+}
+
+\inferrule[expand-alam]{
+  \expandSyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\Delta}
+}{
+  \expandSyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\Delta}
+}
+
+\inferrule[expand-ap-1]{
+  \hsyn{\hGamma}{\hexp_1}{\tehole}\\
+  \expandAna{\hGamma}{\hexp_2}{\tehole}{\dexp_2}{\htau_2}{\Delta_2}\\
+  \expandAna{\hGamma}{\hexp_1}{\tarr{\htau_2}{\tehole}}{\dexp_1}{\htau_1}{\Delta_1}
+}{
+  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}{\hap{(\dcast{\dexp_1}{\tarr{\htau_2}{\tehole}})}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
+}
+
+\inferrule[expand-ap-2]{
+  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
+  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau'_2}{\Delta_2}\\
+  \htau_2 \neq \htau'_2
+}{
+  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dcast{\dexp_2}{\htau_2}}}{\Dunion{\Delta_1}{\Delta_2}}
+}
+
+\inferrule[expand-ap-3]{
+  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
+  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau_2}{\Delta_2}
+}{
+  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
+}\\
+%
+% \inferrule[expand-pair]{
+%   \expandSyn{\hGamma}{\hexp_1}{\htau_1}{\dexp_1}{\Delta_1}\\
+%   \expandSyn{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\Delta_2}
+% }{
+%   \expandSyn{\hGamma}{\hpair{\hexp_1}{\hexp_2}}{\tprod{\htau_1}{\htau_2}}{\hpair{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
+% }
+%
+% \inferrule[expand-prj]{
+%   a
+% }{
+%   b
+% }
+%
+% (inj)
+
+%
+% \inferrule[expand-plus]{ }{
+%   \expandSyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}{\hadd{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
+% }
+%
+\inferrule[expand-syn-empty-hole]{ }{
+  \expandSyn{\hGamma}{\hehole{u}}{\tehole}{\dehole{u}{\idof{\hGamma}}}{\Dbinding{u}{\hGamma}{\tehole}}
+}
+
+\inferrule[expand-syn-non-empty-hole]{
+  \expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}
+}{
+  \expandSyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\Delta, \Dbinding{u}{\hGamma}{\tehole}}
+}\\
+%
+\inferrule[expand-asc-1]{
+  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}\\
+  \htau \neq \htau'
+}{
+  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dcast{\dexp}{\htau}}{\Delta}
+}
+
+\inferrule[expand-asc-2]{
+  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}
+}{
+  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dexp}{\Delta}
+}
+\end{mathpar}
+
+\judgbox{\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}}{ }
+\begin{mathpar}
+\inferrule[expand-lam]{
+  \expandAna{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\htau'_2}{\Delta}
+}{
+  \expandAna{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2'}}{\Delta}
+}
+
+\inferrule[expand-subsume]{
+  \hexp \neq \hehole{u}\\
+  \hexp \neq \hhole{\hexp'}{u}\\\\
+  \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
+  \tconsistent{\htau}{\htau'}
+}{
+  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}
+}
+
+\inferrule[expand-ana-empty-hole]{ }{
+  \expandAna{\hGamma}{\hehole{u}}{\htau}{\dehole{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
+}
+
+\inferrule[expand-ana-non-empty-hole]{ }{
+  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
+}
+\end{mathpar}
+\caption{Algorithmic Expansion}
+\label{fig:expandSyn}
+\label{fig:expandAna}
 \end{figure}
@@
 \begin{figure}[h!]
-\judgbox{\hasType{\hGamma}{M}{\htau}{\Delta}}{$M$ has type $\htau$ with holes $\Delta$}
-...
-\caption{Declarative Statics}
+\judgbox{\hasType{\Delta}{\hGamma}{\dexp}{\htau}}{$\dexp$ is assigned type $\htau$}
+\begin{mathpar}
+\inferrule[ty-assign-const]{ }{
+  \hasType{\Delta}{\hGamma}{c}{b}
+}
+
+\inferrule[ty-assign-var]{ }{
+	\hasType{\Delta}{\hGamma, x : \htau}{x}{\htau}
+}
+
+\inferrule[ty-assign-lam]{
+  \hasType{\Delta}{\hGamma, x : \htau_1}{\dexp}{\htau_2}
+}{
+  \hasType{\Delta}{\hGamma}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2}}
+}
+
+\inferrule[ty-assign-ap]{
+  \hasType{\Delta}{\hGamma}{\dexp_1}{\htau_1}\\
+  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
+  \hasType{\Delta}{\hGamma}{\dexp_2}{\htau_2}
+}{
+  \hasType{\Delta}{\hGamma}{\hap{\dexp_1}{\dexp_2}}{\htau}
+}
+
+\inferrule[ty-assign-empty-hole]{
+  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
+}{
+  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dehole{u}{\sigma}}{\htau}
+}
+
+\inferrule[ty-assign-non-empty-hole]{
+  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
+  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
+}{
+  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dhole{\dexp}{u}{\sigma}}{\htau}
+}
+
+\inferrule[ty-assign-cast]{
+  \hasType{\Delta}{\Gamma}{\dexp}{\htau'}\\
+  \tconsistent{\htau}{\htau'}
+}{
+  \hasType{\Delta}{\hGamma}{\dcast{\dexp}{\htau}}{\htau}
+}
+\end{mathpar}
+\caption{Type Assignment for Dynamic H-Expressions}
+\label{fig:hasType}
 \end{figure}
 
+\begin{definition}[Substitution Type Assignment]
+$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+\end{definition}
+
 \begin{figure}[h!]
-\judgbox{\isValue{M}}{$M$ is a value}
-...
+\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
+\[
+\begin{array}{lcl}
+[\dexp_1 / x] \dehole{\mvar}{\subst}
+&=&
+\dehole{\mvar}{[\dexp_1 / x] \cdot \subst}
+\\
+{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}}
+&=& 
+\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}
+\\
+{[\dexp_1 / x] \dlam{y}{\dexp_2}}
+&=&
+{\dlam{y}{[\dexp_1 / x] \dexp_2}}
+\\
+{[\dexp_1 / x] \dapP{e_2}{e_3}}
+&=&
+{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
+\\
+{[\dexp_1 / x] \dinj{i}{e_2}}
+&=&
+{\dinj{i}{[\dexp_1 / x] e_2}}
+\\
+{[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
+&=&
+{\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
+\\
+{[\dexp_1 / x] \dnum{n}}
+&=&
+{\dnum{n}}
+\\
+{[\dexp_1 / x] \dadd{e_2}{e_3}}
+&=&
+{\dadd{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
+\end{array}
+\]
+\end{figure}
 
-\judgbox{\isIndet{M}}{$M$ is indeterminate}
-...
+\begin{figure}
+\judgbox{\isValue{\dexp}}{$\dexp$ is a value}
+\begin{mathpar}
+\Infer{v-lam}
+{ }{\isValue{\hlam{x}{\dexp}}}
+\and
+\Infer{v-num}
+{ }{\isValue{\hnum{n}}}
+\and
+\Infer{v-inj}
+{\isValue{\dexp}}{\isValue{\hinj{i}{\dexp}}}
+\end{mathpar}
 
-\judgbox{\stepsTo{M}{N}}{$M$ steps to $N$}
-...
+\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
+\begin{mathpar}
+\Infer{i-ehole}
+{ }
+{\isIndet{\dehole{\mvar}{\subst}}}
+\and
+\Infer{i-hole}
+{\isFinal{\dexp}}
+{\isIndet{\dhole{\dexp}{\mvar}{\subst}}}
+\\
+\Infer{i-add$_1$}
+{\isIndet{\dexp_1} \\ \isFinal{\dexp_2}}
+{\isIndet{\dadd{\dexp_1}{\dexp_2}}}
+\and
+\Infer{i-add$_2$}
+{\isFinal{\dexp_1} \\ \isIndet{\dexp_2}}
+{\isIndet{\dadd{\dexp_1}{\dexp_2}}}
+\and
+\Infer{i-app}
+{\isIndet{\dexp_1}}
+{\isIndet{\dap{\dexp_1}{\dexp_2}}}
+\and
+\Infer{i-case}
+{\isIndet{\dexp}}
+{\isIndet{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}}
+\end{mathpar}
+
+\judgbox{\isErr{\dexp}}{$\dexp$ is error}
+\begin{mathpar}
+\Infer{e-cast}
+% XXX use macro type assignment judgement here
+{ \vdash e : \htau_2 \\ \tinconsistent{\htau_1}{\htau_2}}
+{\isErr{\dcast{\dexp}{\htau_1}}}
+\and
+\Infer{e-add$_1$}
+{\isErr{\dexp_1}}
+{\isErr{\dadd{\dexp_1}{\dexp_2}}}
+\and
+\Infer{e-add$_2$}
+{\isErr{\dexp_2}}
+{\isErr{\dadd{\dexp_1}{\dexp_2}}}
+\and
+\Infer{e-app}
+{\isErr{\dexp_1}}
+{\isErr{\dap{\dexp_1}{\dexp_2}}}
+\and
+\Infer{e-case}
+{\isErr{\dexp}}
+{\isErr{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}}
+\end{mathpar}
+
+\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
+\begin{mathpar}
+\Infer{f-val}
+{\isValue{\dexp}}{\isFinal{\dexp}}
+\and
+\Infer{f-indet}
+{\isIndet{\dexp}}{\isFinal{\dexp}}
+\and
+\Infer{f-err}
+{\isErr{\dexp}}{\isFinal{\dexp}}
+\end{mathpar}
+
+\judgbox{\stepsTo{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
+\begin{mathpar}
+\Infer{s-hole}
+{\stepsTo{\dexp_1}{\dexp_2} }
+{\stepsTo{\dhole{\dexp_1}{\mvar}{\subst}}{\dhole{\dexp_2}{\mvar}{\subst}}}
+\and
+\Infer{s-cast}
+% XXX use macro type assignment judgement here
+{ \vdash e : \htau_2 \\ \tconsistent{\htau_1}{\htau_2}}
+{\stepsTo{\dcast{\dexp}{\htau_1}}{\dexp}}
+\\
+\Infer{s-add$_1$}
+{\stepsTo{\dexp_1}{\dexp_1'}}
+{\stepsTo{\dadd{\dexp_1}{\dexp_2}}{\dadd{\dexp_1'}{\dexp_2}}}
+\and
+\Infer{s-add$_2$}
+{\isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
+{\stepsTo{\dadd{\dexp_1}{\dexp_2}}{\dadd{\dexp_1}{\dexp_2'}}}
+\and
+\Infer{s-add$_3$}
+{ n_1 + n_2 = n_3 }
+{\stepsTo{\dadd{\dnum{n_1}}{\dnum{n_2}}}{\dnum{n_3}}}
+%\and
+\\
+\Infer{s-app$_1$}
+{\stepsTo{\dexp_1}{\dexp_1'}}
+{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
+\and
+\Infer{s-app$_2$}
+{ \isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
+{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
+\and
+\Infer{s-app$_\beta$}
+{ \isFinal{\dexp_2} }
+{\stepsTo{\dapP{\dlam{x}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
+%\and
+\\
+\Infer{s-case$_1$}
+{\stepsTo{\dexp}{\dexp'}}
+{\stepsTo{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}{\dcase{\dexp'}{x}{\dexp_x}{y}{\dexp_y}}}
+\and
+\Infer{s-case$_2$}
+{ }
+{\stepsTo{\dcase{\dinj{1}{\dexp}}{x}{\dexp_x}{y}{\dexp_y}}{ [\dexp/x]\dexp_x }}
+\and
+\Infer{s-case$_3$}
+{ }
+{\stepsTo{\dcase{\dinj{2}{\dexp}}{x}{\dexp_x}{y}{\dexp_y}}{ [\dexp/y]\dexp_y }}
+\end{mathpar}
 \caption{Structural Dynamics}
+\label{fig:isValue}
+\label{fig:isIndet}
+\label{fig:stepsTo}
 \end{figure}

### 2017-06-02 calculus.tex
bidirectional typing rules
fixes to substitution definition
some tweaks to dynamics, with questions
updating def of err; new question: Do we need both Γ and Δ to judge it?
wrote down major core metatheorems

@@
 % \end{mathpar}
-\caption{Type Consistency and Matched Types}
+\caption{Type Consistency and Matched Types \todo{inconsistent}}
+\label{fig:tconsistent}
+\label{fig:arrmatch}
+\end{figure}
+
+\begin{figure}[h]
+\judgbox{\hsyn{\hGamma}{\hexp}{\htau}}{$\htau$ synthesizes type $\htau$}
+\begin{mathpar}
+\inferrule[syn-const]{ }{
+  \hsyn{\hGamma}{c}{b}
+}
+
+\inferrule[syn-var]{ }{
+  \hsyn{\hGamma, x : \htau}{x}{\htau}
+}
+
+\inferrule[syn-alam]{
+  \hsyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}
+}{
+  \hsyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}
+}
+
+\inferrule[syn-ap]{
+  \hsyn{\hGamma}{\hexp_1}{\htau_1}\\
+  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
+  \hana{\hGamma}{\hexp_2}{\htau_2}
+}{
+  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
+}
+
+\inferrule[syn-empty-hole]{ }{
+  \hsyn{\hGamma}{\hehole{u}}{\tehole}
+}
+
+\inferrule[syn-non-empty-hole]{
+  \hsyn{\hGamma}{\hexp}{\htau}
+}{
+  \hsyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}
+}
+
+\inferrule[syn-ascription]{
+  \hana{\hGamma}{\hexp}{\htau}
+}{
+  \hsyn{\hGamma}{\hexp : \htau}{\htau}
+}
+\end{mathpar}
+
+\judgbox{\hana{\hGamma}{\hexp}{\htau}}{$\htau$ analyzes against type $\htau$}
+\begin{mathpar}
+\inferrule[ana-lam]{
+  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
+  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
+}{
+  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
+}
+
+\inferrule[subsumption]{
+  \hsyn{\hGamma}{\hexp}{\htau}\\
+  \tconsistent{\htau}{\htau'}
+}{
+  \hana{\hGamma}{\hexp}{\htau'}
+}
+\end{mathpar}
+\caption{Bidirectional Typing of H-Expressions}
 \label{fig:tconsistent}
@@
 \begin{array}{lcl}
-[\dexp_1 / x] \dehole{\mvar}{\subst}
+[\dexp_1 / x] c
 &=&
-\dehole{\mvar}{[\dexp_1 / x] \cdot \subst}
+c\\
+{[\dexp_1 / x] x}
+&=&
+\dexp_1\\
+{[\dexp_1 / x] y}
+&=&
+y\\
+{[\dexp_1 / x] \dlam{y}{\dexp_2}}
+&=&
+{\dlam{y}{[\dexp_1 / x] \dexp_2}}
+\\
+{[\dexp_1 / x] \dap{e_2}{e_3}}
+&=&
+{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
+\\
+% {[\dexp_1 / x] \dinj{i}{e_2}}
+% &=&
+% {\dinj{i}{[\dexp_1 / x] e_2}}
+% \\
+% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
+% &=&
+% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
+% \\
+{[\dexp_1 / x]} \dehole{\mvar}{\subst}
+&=&
+\dehole{\mvar}{[\dexp_1 / x] \subst}
 \\
@@
 \dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}
-\\
-{[\dexp_1 / x] \dlam{y}{\dexp_2}}
-&=&
-{\dlam{y}{[\dexp_1 / x] \dexp_2}}
-\\
-{[\dexp_1 / x] \dapP{e_2}{e_3}}
-&=&
-{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
-\\
-{[\dexp_1 / x] \dinj{i}{e_2}}
-&=&
-{\dinj{i}{[\dexp_1 / x] e_2}}
-\\
-{[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
-&=&
-{\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
-\\
-{[\dexp_1 / x] \dnum{n}}
-&=&
-{\dnum{n}}
-\\
-{[\dexp_1 / x] \dadd{e_2}{e_3}}
-&=&
-{\dadd{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
 \end{array}
@@
 \begin{mathpar}
-\Infer{v-lam}
-{ }{\isValue{\hlam{x}{\dexp}}}
-\and
-\Infer{v-num}
-{ }{\isValue{\hnum{n}}}
-\and
-\Infer{v-inj}
-{\isValue{\dexp}}{\isValue{\hinj{i}{\dexp}}}
+\inferrule[v-constant]{ }{
+  \isValue{c}
+}
+
+\inferrule[v-lam]{ }{
+  \isValue{\halam{x}{\htau}{\dexp}}
+}
 \end{mathpar}
@@
 \begin{mathpar}
-\Infer{i-ehole}
+\inferrule[indet-empty-hole]
 { }
 {\isIndet{\dehole{\mvar}{\subst}}}
-\and
-\Infer{i-hole}
+
+\inferrule[indet-non-empty-hole]
 {\isFinal{\dexp}}
 {\isIndet{\dhole{\dexp}{\mvar}{\subst}}}
-\\
-\Infer{i-add$_1$}
-{\isIndet{\dexp_1} \\ \isFinal{\dexp_2}}
-{\isIndet{\dadd{\dexp_1}{\dexp_2}}}
-\and
-\Infer{i-add$_2$}
-{\isFinal{\dexp_1} \\ \isIndet{\dexp_2}}
-{\isIndet{\dadd{\dexp_1}{\dexp_2}}}
-\and
-\Infer{i-app}
-{\isIndet{\dexp_1}}
+
+\inferrule[indet-app]
+{\isIndet{\dexp_1}\\
+% \isFinal{\dexp_2}~\text{\todo{??}}}
+ \isFinal{\dexp_2}}
 {\isIndet{\dap{\dexp_1}{\dexp_2}}}
-\and
-\Infer{i-case}
+
+\inferrule[indet-cast]
 {\isIndet{\dexp}}
-{\isIndet{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}}
+{\isIndet{\dcast{\dexp}{\htau}}}
+
 \end{mathpar}
 
-\judgbox{\isErr{\dexp}}{$\dexp$ is error}
+\judgbox{\isErr{\Delta}{\dexp}}{Under $\Delta$, $\dexp$ contains a cast error}
 \begin{mathpar}
-\Infer{e-cast}
+\inferrule[err-nemp-hole~\todo{??}]
+{\subst \sim_? \Gamma_2
+\\ 
+\isErr{\Delta; \Gamma_1, \Gamma_2}{\dexp}}
+{\isErr{\Delta; \Gamma_1}{\dhole{\dexp}{\mvar}{\subst}}}
+
+\inferrule[err-cast-error~\todo{??}]
 % XXX use macro type assignment judgement here
-{ \vdash e : \htau_2 \\ \tinconsistent{\htau_1}{\htau_2}}
-{\isErr{\dcast{\dexp}{\htau_1}}}
-\and
-\Infer{e-add$_1$}
-{\isErr{\dexp_1}}
-{\isErr{\dadd{\dexp_1}{\dexp_2}}}
-\and
-\Infer{e-add$_2$}
-{\isErr{\dexp_2}}
-{\isErr{\dadd{\dexp_1}{\dexp_2}}}
-\and
-\Infer{e-app}
-{\isErr{\dexp_1}}
-{\isErr{\dap{\dexp_1}{\dexp_2}}}
-\and
-\Infer{e-case}
-{\isErr{\dexp}}
-{\isErr{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}}
+{\hasType{\Delta}{\todo{\emptyset}}{e}{\htau_2}\\
+ \tinconsistent{\htau_1}{\htau_2}}
+{\isErr{\Delta}{\dcast{\dexp}{\htau_1}}}
+
+\inferrule[err-lam~\todo{??}]
+% XXX use macro type assignment judgement here
+{\isErr{\Delta; \Gamma, x:\htau_1}{\dexp}}
+{\isErr{\Delta; \Gamma}{\halam{x}{\htau}{\dexp}}}
+
+\\
+
+\inferrule[err-app-1]
+{
+ \isErr{\Delta}{\dexp_1}
+ %\\\
+ %isFinal{\dexp_2}~\text{\todo{??}}
+}
+{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}
+
+\inferrule[err-app-2]
+{
+ %\isIndet{\dexp_1}
+ %\\
+ \isErr{\Delta}{\dexp_2}
+}
+{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}
+
+\inferrule[err-cast]
+% XXX use macro type assignment judgement here
+{\isErr{\Delta}{\dexp}}
+{\isErr{\Delta}{\dcast{\dexp}{\htau_1}}}
+
 \end{mathpar}
@@
 {\isIndet{\dexp}}{\isFinal{\dexp}}
-\and
-\Infer{f-err}
-{\isErr{\dexp}}{\isFinal{\dexp}}
+%% No longer a rule: Errors are not indet anymore:
+%% \and
+%% \Infer{f-err}
+%% {\isErr{\dexp}}{\isFinal{\dexp}}
 \end{mathpar}
@@
 \\
-\Infer{s-add$_1$}
-{\stepsTo{\dexp_1}{\dexp_1'}}
-{\stepsTo{\dadd{\dexp_1}{\dexp_2}}{\dadd{\dexp_1'}{\dexp_2}}}
-\and
-\Infer{s-add$_2$}
-{\isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
-{\stepsTo{\dadd{\dexp_1}{\dexp_2}}{\dadd{\dexp_1}{\dexp_2'}}}
-\and
-\Infer{s-add$_3$}
-{ n_1 + n_2 = n_3 }
-{\stepsTo{\dadd{\dnum{n_1}}{\dnum{n_2}}}{\dnum{n_3}}}
-%\and
-\\
 \Infer{s-app$_1$}
@@
 {\stepsTo{\dapP{\dlam{x}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
-%\and
-\\
-\Infer{s-case$_1$}
-{\stepsTo{\dexp}{\dexp'}}
-{\stepsTo{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}{\dcase{\dexp'}{x}{\dexp_x}{y}{\dexp_y}}}
-\and
-\Infer{s-case$_2$}
-{ }
-{\stepsTo{\dcase{\dinj{1}{\dexp}}{x}{\dexp_x}{y}{\dexp_y}}{ [\dexp/x]\dexp_x }}
-\and
-\Infer{s-case$_3$}
-{ }
-{\stepsTo{\dcase{\dinj{2}{\dexp}}{x}{\dexp_x}{y}{\dexp_y}}{ [\dexp/y]\dexp_y }}
 \end{mathpar}
@@
 \end{figure}
+
+\clearpage
+\subsection{Metatheory}
+
+...
+
+\begin{theorem}[Expandability \todo{??}] ~
+  \begin{enumerate}
+    \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ for some $\dexp$ and $\Delta$.
+    \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ for some $\dexp$ and $\htau'$ and $\Delta$.
+  \end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Correspondence \todo{??}] ~
+  \begin{enumerate}
+    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ then $\hsyn{\hGamma}{\hexp}{\htau}$.
+    \item If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then $\hana{\hGamma}{\hexp}{\htau}$.
+  \end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Typed Expansion] ~
+  \begin{enumerate}
+    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+    \item If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
+  \end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Expansion Unicity] ~
+  \begin{enumerate}
+    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$ then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
+    \item If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$ and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$ then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
+  \end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Type Assignment Unicity]
+If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$ then $\htau=\htau'$.
+\end{theorem}
+
+\begin{theorem}[Progress]
+If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\stepsTo{\dexp}{\dexp'}$.
+\end{theorem}
+
+\begin{theorem}[Preservation]
+  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and $\stepsTo{\dexp}{\dexp'}$ then $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \todo{or maybe up to consistency?}
+\end{theorem}
+
+\todo{complete progress + preservation}
+
+\todo{canonical forms}
+
+\todo{indeterminate forms}

### 2017-06-09 calculus.tex
fixed expand-ana-non-empty-hole
changing the rule names to match with the corresponding agda constructor names. this entails a change in naming convention, unfortunately: in agda, it's common to name functions with lower case and hyphens as in `subst-commute` or similar, and name constructors in the ML tradition with all caps as in `LCons`. so i've followed that convention.
roughly speaking, the first `n-1` letters in the all-caps prefix of a
constructor name denote the judgement it comes from, and the rest denotes
the form it applies to. if more than one rule applies to the same form,
then th

@@
 \begin{mathpar}
-\inferrule{ }{
+\inferrule[TCHole1]{ }{
   \tconsistent{\tehole}{\htau}
@@
 
-\inferrule{ }{
+\inferrule[TCHole2]{ }{
   \tconsistent{\htau}{\tehole}
@@
 
-\inferrule{ }{
+\inferrule[TCRefl]{ }{
   \tconsistent{\htau}{\htau}
@@
 
-\inferrule{
+\inferrule[TCArr]{
   \tconsistent{\htau_1}{\htau_1'}\\
@@
 
+\judgbox{\tinconsistent{\htau}{\htau'}}{$\htau$ is inconsistent with $\htau'$}
+\begin{mathpar}
+    \inferrule[ICBaseArr1]{ }{
+      \tinconsistent{\tb}{\tarr{\htau_1}{\htau_2}}
+    }
+
+    \inferrule[ICBaseArr2]{ }{
+      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tb}
+    }
+
+    \inferrule[ICArr1]{
+      \tinconsistent{\htau_1}{\htau_3}
+    }{
+      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_3}{\htau_4}}
+    }
+
+    \inferrule[ICArr2]{
+      \tinconsistent{\htau_2}{\htau_4}
+    }{
+      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_3}{\htau_4}}
+    }
+\end{mathpar}
+
 \judgbox{\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}}{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
 \begin{mathpar}
-\inferrule{ }{
+\inferrule[MAHole]{ }{
   \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
@@
 
-\inferrule{ }{
+\inferrule[MAArr]{ }{
   \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
@@
 % \end{mathpar}
-\caption{Type Consistency and Matched Types \todo{inconsistent}}
+\caption{Type Consistency, Inconsistency, and Matching}
 \label{fig:tconsistent}
@@
 \begin{mathpar}
-\inferrule[syn-const]{ }{
+\inferrule[SConst]{ }{
   \hsyn{\hGamma}{c}{b}
@@
 
-\inferrule[syn-var]{ }{
+\inferrule[SVar]{ }{
   \hsyn{\hGamma, x : \htau}{x}{\htau}
@@
 
-\inferrule[syn-alam]{
+\inferrule[SLam]{
   \hsyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}
@@
 
-\inferrule[syn-ap]{
+\inferrule[SAp]{
   \hsyn{\hGamma}{\hexp_1}{\htau_1}\\
@@
 
-\inferrule[syn-empty-hole]{ }{
+\inferrule[SEHole]{ }{
   \hsyn{\hGamma}{\hehole{u}}{\tehole}
@@
 
-\inferrule[syn-non-empty-hole]{
+\inferrule[SNEHole]{
   \hsyn{\hGamma}{\hexp}{\htau}
@@
 
-\inferrule[syn-ascription]{
+\inferrule[SAsc]{
   \hana{\hGamma}{\hexp}{\htau}
@@
 \begin{mathpar}
-\inferrule[ana-lam]{
+\inferrule[ALam]{
   \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
@@
 
-\inferrule[subsumption]{
+\inferrule[ASubsume]{
   \hsyn{\hGamma}{\hexp}{\htau}\\
@@
 
+\todo{definition of complete types and terms? do we care about completeness
+  of $\hexp$s or $\dexp$s?}
+
 \begin{figure}[h]
@@
 \begin{mathpar}
-\inferrule[expand-constant]{ }{
+\inferrule[ESConst]{ }{
   \expandSyn{\hGamma}{c}{b}{c}{\emptyset}
@@
 
-\inferrule[expand-var]{ }{
+\inferrule[ESVar]{ }{
   \expandSyn{\hGamma, x : \htau}{x}{\htau}{x}{\emptyset}
@@
 
-\inferrule[expand-alam]{
+\inferrule[ESLam]{
   \expandSyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\Delta}
@@
 
-\inferrule[expand-ap-1]{
+\inferrule[ESAp1]{
   \hsyn{\hGamma}{\hexp_1}{\tehole}\\
@@
 
-\inferrule[expand-ap-2]{
+\inferrule[ESAp2]{
   \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
@@
 
-\inferrule[expand-ap-3]{
+\inferrule[ESAp3]{
   \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
@@
 %
-\inferrule[expand-syn-empty-hole]{ }{
+\inferrule[ESEHole]{ }{
   \expandSyn{\hGamma}{\hehole{u}}{\tehole}{\dehole{u}{\idof{\hGamma}}}{\Dbinding{u}{\hGamma}{\tehole}}
@@
 
-\inferrule[expand-syn-non-empty-hole]{
+\inferrule[ESNEHole]{
   \expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}
@@
 %
-\inferrule[expand-asc-1]{
+\inferrule[ESAsc1]{
   \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}\\
@@
 
-\inferrule[expand-asc-2]{
+\inferrule[ESAsc2]{
   \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}
@@
 \begin{mathpar}
-\inferrule[expand-lam]{
+\inferrule[EALam]{
   \expandAna{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\htau'_2}{\Delta}
@@
 
-\inferrule[expand-subsume]{
+\inferrule[EASubsume]{
   \hexp \neq \hehole{u}\\
@@
 
-\inferrule[expand-ana-empty-hole]{ }{
+\inferrule[EAEHole]{ }{
   \expandAna{\hGamma}{\hehole{u}}{\htau}{\dehole{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
@@
 
-\inferrule[expand-ana-non-empty-hole]{ }{
-  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
+\inferrule[EANEHole]{
+  \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
+  \tconsistent{\htau}{\htau'}
+}{
+  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\htau}{\Delta, \Dbinding{u}{\hGamma}{\htau}}
 }
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
 \judgbox{\hasType{\Delta}{\hGamma}{\dexp}{\htau}}{$\dexp$ is assigned type $\htau$}
 \begin{mathpar}
-\inferrule[ty-assign-const]{ }{
+\inferrule[TAConst]{ }{
   \hasType{\Delta}{\hGamma}{c}{b}
@@
 
-\inferrule[ty-assign-var]{ }{
+\inferrule[TAVar]{ }{
 	\hasType{\Delta}{\hGamma, x : \htau}{x}{\htau}
@@
 
-\inferrule[ty-assign-lam]{
+\inferrule[TALam]{
   \hasType{\Delta}{\hGamma, x : \htau_1}{\dexp}{\htau_2}
@@
 
-\inferrule[ty-assign-ap]{
+\inferrule[TAAp]{
   \hasType{\Delta}{\hGamma}{\dexp_1}{\htau_1}\\
@@
 
-\inferrule[ty-assign-empty-hole]{
+\inferrule[TAEHole]{
   \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
@@
 
-\inferrule[ty-assign-non-empty-hole]{
+\inferrule[TANEHole]{
   \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
@@
 
-\inferrule[ty-assign-cast]{
+\inferrule[TACast]{
   \hasType{\Delta}{\Gamma}{\dexp}{\htau'}\\
@@
 
-\begin{definition}[Substitution Type Assignment]
+\begin{figure}[!ht]
+  \begin{definition}
     $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
   \end{definition}
+  \caption{substitution type assignment}
+\end{figure}
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
 \judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
@@
 y\\
-{[\dexp_1 / x] \dlam{y}{\dexp_2}}
+{[\dexp_1 / x] \dlam{y}{\htau}{\dexp_2}}
 &=&
-{\dlam{y}{[\dexp_1 / x] \dexp_2}}
+{\dlam{y}{\htau}{[\dexp_1 / x] \dexp_2}}
 \\
@@
 
-\begin{figure}
+\begin{figure}[!ht]
 \judgbox{\isValue{\dexp}}{$\dexp$ is a value}
 \begin{mathpar}
-\inferrule[v-constant]{ }{
+\inferrule[VConst]{ }{
   \isValue{c}
@@
 
-\inferrule[v-lam]{ }{
+\inferrule[VLam]{ }{
   \isValue{\halam{x}{\htau}{\dexp}}
@@
 \begin{mathpar}
-\inferrule[indet-empty-hole]
+\inferrule[IEHole]
 { }
@@
 
-\inferrule[indet-non-empty-hole]
+\inferrule[INEHole]
 {\isFinal{\dexp}}
@@
 
-\inferrule[indet-app]
+\inferrule[IAp]
 {\isIndet{\dexp_1}\\
@@
 
-\inferrule[indet-cast]
+\inferrule[ICast]
 {\isIndet{\dexp}}
@@
 
+\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
+\begin{mathpar}
+\Infer{FVal}
+{\isValue{\dexp}}{\isFinal{\dexp}}
+\and
+\Infer{FIndet}
+{\isIndet{\dexp}}{\isFinal{\dexp}}
+%% No longer a rule: Errors are not indet anymore:
+%% \and
+%% \Infer{f-err}
+%% {\isErr{\dexp}}{\isFinal{\dexp}}
+\end{mathpar}
+
 \judgbox{\isErr{\Delta}{\dexp}}{Under $\Delta$, $\dexp$ contains a cast error}
 \begin{mathpar}
-\inferrule[err-nemp-hole~\todo{??}]
+\inferrule[ERNEHole~\todo{??}]
 {\subst \sim_? \Gamma_2
@@
 
-\inferrule[err-cast-error~\todo{??}]
+\inferrule[ERCastError~\todo{??}]
 % XXX use macro type assignment judgement here
@@
 
-\inferrule[err-lam~\todo{??}]
+\inferrule[ERLam~\todo{??}]
 % XXX use macro type assignment judgement here
@@
 
-\inferrule[err-app-1]
+\inferrule[ERAp1]
 {
@@
 
-\inferrule[err-app-2]
+\inferrule[ERAp2]
 {
@@
 
-\inferrule[err-cast]
+\inferrule[ERCast]
 % XXX use macro type assignment judgement here
@@
 
-\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
-\begin{mathpar}
-\Infer{f-val}
-{\isValue{\dexp}}{\isFinal{\dexp}}
-\and
-\Infer{f-indet}
-{\isIndet{\dexp}}{\isFinal{\dexp}}
-%% No longer a rule: Errors are not indet anymore:
-%% \and
-%% \Infer{f-err}
-%% {\isErr{\dexp}}{\isFinal{\dexp}}
-\end{mathpar}
-
 \judgbox{\stepsTo{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
 \begin{mathpar}
-\Infer{s-hole}
+\Infer{STHole}
 {\stepsTo{\dexp_1}{\dexp_2} }
@@
 \and
-\Infer{s-cast}
+\Infer{STCast}
 % XXX use macro type assignment judgement here
@@
 \\
-\Infer{s-app$_1$}
+\Infer{STAp1}
 {\stepsTo{\dexp_1}{\dexp_1'}}
@@
 \and
-\Infer{s-app$_2$}
+\Infer{STAp2}
 { \isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
@@
 \and
-\Infer{s-app$_\beta$}
+\Infer{STAp$\beta$}
 { \isFinal{\dexp_2} }
-{\stepsTo{\dapP{\dlam{x}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
+{\stepsTo{\dapP{\dlam{x}{\htau}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
 \end{mathpar}
@@
 
-...
+(these theorems should be taken as fairly provisional)
 

### 2017-06-09 calculus.tex
fixed cast syntax, cast stepping rule
added missing requirement for substitution typing

@@
   \dhole{\dexp}{\mvar}{\subst} ~\vert~
-  \dcast{\dexp}{\htau}\\
+  \dcast{\htau}{\dexp}\\
 \end{array}$
@@
 }{
-  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}{\hap{(\dcast{\dexp_1}{\tarr{\htau_2}{\tehole}})}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
+  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}{\hap{(\dcast{\tarr{\htau_2}{\tehole}}{\dexp_1})}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
 }
@@
 }{
-  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dcast{\dexp_2}{\htau_2}}}{\Dunion{\Delta_1}{\Delta_2}}
+  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dcast{\htau_2}{\dexp_2}}}{\Dunion{\Delta_1}{\Delta_2}}
 }
@@
 }{
-  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dcast{\dexp}{\htau}}{\Delta}
+  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dcast{\htau}{\dexp}}{\Delta}
 }
@@
 }{
-  \hasType{\Delta}{\hGamma}{\dcast{\dexp}{\htau}}{\htau}
+  \hasType{\Delta}{\hGamma}{\dcast{\htau}{\dexp}}{\htau}
 }
@@
   \begin{definition}
-    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Delta}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
   \end{definition}
@@
 {\isIndet{\dexp}}
-{\isIndet{\dcast{\dexp}{\htau}}}
+{\isIndet{\dcast{\htau}{\dexp}}}
 
@@
  \tinconsistent{\htau_1}{\htau_2}}
-{\isErr{\Delta}{\dcast{\dexp}{\htau_1}}}
+{\isErr{\Delta}{\dcast{\htau_1}{\dexp}}}
 
@@
 {\isErr{\Delta}{\dexp}}
-{\isErr{\Delta}{\dcast{\dexp}{\htau_1}}}
+{\isErr{\Delta}{\dcast{\htau_1}{\dexp}}}
 
@@
 \Infer{STCast}
-% XXX use macro type assignment judgement here
-{ \vdash e : \htau_2 \\ \tconsistent{\htau_1}{\htau_2}}
-{\stepsTo{\dcast{\dexp}{\htau_1}}{\dexp}}
+{\hasType{\Delta}{\emptyset}{\dexp}{\htau_2} \\
+ \tconsistent{\tau_1}{\tau_2}}
+{\stepsTo{\dcast{\htau_1}{\dexp}}{\dexp}}
 \\

### 2017-06-09 calculus.tex
simplifying the names in the bnfs to not have holes, using d for terms that support a dynamic semantics. this will soon match the agda. also quashing a few LaTeX errors and warnings

@@
 
-\begin{figure}[h]
+\begin{figure}[!ht]
 $\arraycolsep=4pt\begin{array}{rllllll}
@@
 
-\begin{figure}[h]
+\begin{figure}[!ht]
 \judgbox{\tconsistent{\htau_1}{\htau_2}}{$\htau_1$ is consistent with $\htau_2$}
@@
 
-\begin{figure}[h]
+\begin{figure}[!ht]
 \judgbox{\hsyn{\hGamma}{\hexp}{\htau}}{$\htau$ synthesizes type $\htau$}
@@
 \caption{Bidirectional Typing of H-Expressions}
-\label{fig:tconsistent}
-\label{fig:arrmatch}
 \end{figure}
@@
 
-\begin{figure}[h]
+\begin{figure}[!ht]
 \judgbox{\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}}{ }
@@
   \begin{definition}
-    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Delta}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
   \end{definition}

### 2017-06-09 calculus.tex
syntactic consistency
moving around to match use order
adding Delta |- in err and step

@@
 
-\judgbox{\tinconsistent{\htau}{\htau'}}{$\htau$ is inconsistent with $\htau'$}
+\judgbox{\tinconsistent{\htau_1}{\htau_2}}{$\htau_1$ is inconsistent with $\htau_2$}
 \begin{mathpar}
@@
 \begin{figure}[!ht]
+  \begin{definition}
+    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+  \end{definition}
+  \caption{substitution type assignment}
+  \label{fig:subassign}
+\end{figure}
+
+\begin{figure}[!ht]
 \judgbox{\hasType{\Delta}{\hGamma}{\dexp}{\htau}}{$\dexp$ is assigned type $\htau$}
@@
 \label{fig:hasType}
-\end{figure}
-
-\begin{figure}[!ht]
-  \begin{definition}
-    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
-  \end{definition}
-  \caption{substitution type assignment}
 \end{figure}
@@
 
-\judgbox{\stepsTo{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
+\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{under $\Delta$, $\dexp_1$ steps to $\dexp_2$}
 \begin{mathpar}
 \Infer{STHole}
-{\stepsTo{\dexp_1}{\dexp_2} }
-{\stepsTo{\dhole{\dexp_1}{\mvar}{\subst}}{\dhole{\dexp_2}{\mvar}{\subst}}}
+{\stepsToD{\Delta}{\dexp_1}{\dexp_2} }
+{\stepsToD{\Delta}{\dhole{\dexp_1}{\mvar}{\subst}}{\dhole{\dexp_2}{\mvar}{\subst}}}
 \and
@@
  \tconsistent{\tau_1}{\tau_2}}
-{\stepsTo{\dcast{\htau_1}{\dexp}}{\dexp}}
+{\stepsToD{\Delta}{\dcast{\htau_1}{\dexp}}{\dexp}}
 \\
 \Infer{STAp1}
-{\stepsTo{\dexp_1}{\dexp_1'}}
-{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
+{\stepsToD{\Delta}{\dexp_1}{\dexp_1'}}
+{\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
 \and
 \Infer{STAp2}
-{ \isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
-{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
+{ \isFinal{\dexp_1} \\ \stepsToD{\Delta}{\dexp_2}{\dexp_2'}}
+{\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
 \and
@@
 { \isFinal{\dexp_2} }
-{\stepsTo{\dapP{\dlam{x}{\htau}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
+{\stepsToD{\Delta}{\dapP{\dlam{x}{\htau}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
 \end{mathpar}
@@
 \begin{theorem}[Progress]
-If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\stepsTo{\dexp}{\dexp'}$.
+If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\stepsToD{\Delta}{\dexp}{\dexp'}$.
 \end{theorem}
@@
 \begin{theorem}[Preservation]
-  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and $\stepsTo{\dexp}{\dexp'}$ then $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \todo{or maybe up to consistency?}
+  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and $\stepsToD{\Delta}{\dexp}{\dexp'}$ then $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \todo{or maybe up to consistency?}
 \end{theorem}

### 2017-09-21 calculus.tex
todo
moving premises around to match other two ap rules more closely
small changes in working on expandablity with @cyrus-
stubbing in hole for how to expand lambdas that are being checked against hole type
filling in todos for ealamhole rule
adding marks to the holes in dhexps. this is present in the agda, but i cant promise absolute coherence between the two this instant. im sure its close, but there may be a little skew as were still working these out.
fixed definition of substitution type assignment

--- 2017-06-09 calculus.tex
@@
   \begin{definition}
-    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Delta}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Gamma'}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
   \end{definition}

### 2017-09-21 calculus.tex

--- 2017-06-09 calculus.tex
@@
   \hexp : \htau\\
+\mathsf{Mark} & \markname{} & ::= &
+  \evaled{} ~\vert~  \unevaled{}\\
  \mathsf{DHExp} & \dexp  & ::= &
@@
   % \hadd{\dexp}{\dexp} ~\vert~
-  \dehole{\mvar}{\subst} ~\vert~
-  \dhole{\dexp}{\mvar}{\subst} ~\vert~
+  \dehole{\mvar}{\subst}{\markname} ~\vert~
+  \dhole{\dexp}{\mvar}{\subst}{\markname} ~\vert~
   \dcast{\htau}{\dexp}\\
@@
   \hsyn{\hGamma}{\hexp_1}{\tehole}\\
-  \expandAna{\hGamma}{\hexp_2}{\tehole}{\dexp_2}{\htau_2}{\Delta_2}\\
-  \expandAna{\hGamma}{\hexp_1}{\tarr{\htau_2}{\tehole}}{\dexp_1}{\htau_1}{\Delta_1}
+  \expandAna{\hGamma}{\hexp_1}{\tarr{\htau_2}{\tehole}}{\dexp_1}{\htau_1}{\Delta_1}\\
+  \expandAna{\hGamma}{\hexp_2}{\tehole}{\dexp_2}{\htau_2}{\Delta_2}
 }{
@@
 \inferrule[ESEHole]{ }{
-  \expandSyn{\hGamma}{\hehole{u}}{\tehole}{\dehole{u}{\idof{\hGamma}}}{\Dbinding{u}{\hGamma}{\tehole}}
+  \expandSyn{\hGamma}{\hehole{u}}{\tehole}{\dehole{u}{\idof{\hGamma}}{\unevaled}}{\Dbinding{u}{\hGamma}{\tehole}}
 }
@@
 }{
-  \expandSyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\Delta, \Dbinding{u}{\hGamma}{\tehole}}
+  \expandSyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}{\dhole{\dexp}{u}{\idof{\hGamma}}{\unevaled}}{\Delta, \Dbinding{u}{\hGamma}{\tehole}}
 }\\
@@
 
+\inferrule[EALamHole]{
+  \expandAna{\hGamma, x : \tehole}{\hexp}{\tehole}{\dexp}{\htau}{\Delta}
+}{
+  \expandAna{\hGamma}{\hlam{x}{\hexp}}{\tehole}{\halam{x}{\tehole}{\dexp}}{\tarr{\tehole}{\htau}}{\Delta}
+}
+
 \inferrule[EASubsume]{
@@
 \inferrule[EAEHole]{ }{
-  \expandAna{\hGamma}{\hehole{u}}{\htau}{\dehole{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
+  \expandAna{\hGamma}{\hehole{u}}{\htau}{\dehole{u}{\idof{\hGamma}}{\unevaled}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
 }
@@
   \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
-  \tconsistent{\htau}{\htau'}
 }{
-  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\htau}{\Delta, \Dbinding{u}{\hGamma}{\htau}}
+  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}{\unevaled}}{\htau}{\Delta, \Dbinding{u}{\hGamma}{\htau}}
 }
@@
 }{
-  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dehole{u}{\sigma}}{\htau}
+  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dehole{u}{\sigma}{\markname}}{\htau}
 }
@@
 }{
-  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dhole{\dexp}{u}{\sigma}}{\htau}
+  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dhole{\dexp}{u}{\sigma}{\markname}}{\htau}
 }
@@
 \label{fig:hasType}
+\end{figure}
+
+\begin{figure}[!ht]
+  \begin{definition}
+    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Gamma'}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+  \end{definition}
+  \caption{substitution type assignment}
 \end{figure}
@@
 % \\
-{[\dexp_1 / x]} \dehole{\mvar}{\subst}
+{[\dexp_1 / x]} \dehole{\mvar}{\subst}{\todo{??}}
 &=&
-\dehole{\mvar}{[\dexp_1 / x] \subst}
+\dehole{\mvar}{[\dexp_1 / x] \subst}{\todo{??}}
 \\
-{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}}
+{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{\todo{??}}}
 &=&
-\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}
+\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{\todo{??}}
 \end{array}
@@
 { }
-{\isIndet{\dehole{\mvar}{\subst}}}
+{\isIndet{\dehole{\mvar}{\subst}{\todo{??}}}}
 
@@
 {\isFinal{\dexp}}
-{\isIndet{\dhole{\dexp}{\mvar}{\subst}}}
+{\isIndet{\dhole{\dexp}{\mvar}{\subst}{\todo{??}}}}
 
@@
 \end{mathpar}
+
+\todo{the judgmental forms are not consistent in these
+rules; the contexts change. need to figure this out.}
 
@@
 
-\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{under $\Delta$, $\dexp_1$ steps to $\dexp_2$}
+\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{under $\Delta$, $\dexp_1$
+steps to $\dexp_2$}
 \begin{mathpar}
@@
 {\stepsToD{\Delta}{\dexp_1}{\dexp_2} }
-{\stepsToD{\Delta}{\dhole{\dexp_1}{\mvar}{\subst}}{\dhole{\dexp_2}{\mvar}{\subst}}}
+{\stepsToD{\Delta}{\dhole{\dexp_1}{\mvar}{\subst}{\todo{??}}}{\dhole{\dexp_2}{\mvar}{\subst}{\todo{??}}}}
 \and
@@
 \todo{indeterminate forms}
+
+\todo{theorem that special casing the holes in EANEHole gives you ``the
+most descriptive hole types'' for some sense of what that means -- they'd
+all just be hole other wise}

### 2017-09-26 hazelnut-dynamics.tex
added figure outlines and basic judgment forms
added figure labels and copied over syntax and type consistency defns from popl paper
dunno
fixed cast syntax, cast stepping rule
popl wants acmsmall not acmlarge so changed it
simpler title

@@
 %% For double-blind review submission
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
+\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
 %% For single-blind review submission
@@
 \usepackage{microtype}
-\usepackage{todonotes}
 \usepackage{mdframed}
@@
 \usepackage{bbm}
+\usepackage{stmaryrd}
+\usepackage{mathtools}
+\usepackage{leftidx}
+\newcommand{\todo}[1]{{\color{red} #1}}
+
+%% Joshua Dunfield macros
+\def\OPTIONConf{1}%
+\usepackage{joshuadunfield}
 
@@
 %% Title information
-\title{Hazelnut Live: Type-Theoretic Foundations for Running Incomplete Programs}         %% [Short Title] is optional;
+\title{Live Programming with Typed Holes}         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 
-\newcommand{\introSec}{Introduction}
-\section{\protect\introSec}
-\label{sec:intro}
-\todo{intro}
-
-\begin{figure}
+\input{intro}
+\input{calculus}
 

### 2017-09-26 calculus.tex
fixed definition of substitution
d err judgement and fixes to other auxiliary judgments in the dynamics
fixes to auxiliary judgements and added proper rules for marks
step rule names
context membership tweaks per discussion with ian last week
Adding missing rule for substituion into casts
texed up rules for contextual dynamics (minus rule names)
added missing mark transitions and related rules
updating with most of the rule names, pending a chat with @cyrus- for a couple more
adding some possible theorem statements, ideas from Slack

@@
 
-\inferrule[ESVar]{ }{
-  \expandSyn{\hGamma, x : \htau}{x}{\htau}{x}{\emptyset}
+\inferrule[ESVar]{
+  x : \htau \in \hGamma
+}{
+  \expandSyn{\hGamma}{x}{\htau}{x}{\emptyset}
 }
@@
 
-\inferrule[TAVar]{ }{
-	\hasType{\Delta}{\hGamma, x : \htau}{x}{\htau}
+\inferrule[TAVar]{
+  x : \htau \in \hGamma
+}{
+	\hasType{\Delta}{\hGamma}{x}{\htau}
 }
@@
 \inferrule[TAEHole]{
+  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\
   \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
 }{
-  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dehole{u}{\sigma}{\markname}}{\htau}
+  \hasType{\Delta}{\hGamma}{\dehole{u}{\sigma}{\markname}}{\htau}
 }
@@
 \inferrule[TANEHole]{
+  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\\\
   \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
@@
 }{
-  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dhole{\dexp}{u}{\sigma}{\markname}}{\htau}
+  \hasType{\Delta}{\hGamma}{\dhole{\dexp}{u}{\sigma}{\markname}}{\htau}
 }
@@
 \[
-\begin{array}{lcl}
+\begin{array}{lcll}
 [\dexp_1 / x] c
@@
 &=&
-y\\
+y & (y \neq x)\\
 {[\dexp_1 / x] \dlam{y}{\htau}{\dexp_2}}
@@
 {\dlam{y}{\htau}{[\dexp_1 / x] \dexp_2}}
+& (y \neq x)
 \\
-{[\dexp_1 / x] \dap{e_2}{e_3}}
+{[\dexp_1 / x] \dap{d_2}{d_3}}
 &=&
-{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
+{\dapP{[\dexp_1 / x] d_2}{[\dexp_1 / x] d_3}}
 \\
@@
 % \\
-{[\dexp_1 / x]} \dehole{\mvar}{\subst}{\todo{??}}
+{[\dexp_1 / x]} \dehole{\mvar}{\subst}{m}
 &=&
-\dehole{\mvar}{[\dexp_1 / x] \subst}{\todo{??}}
+\dehole{\mvar}{[\dexp_1 / x] \subst}{m}
 \\
-{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{\todo{??}}}
+{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{m}}
 &=&
-\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{\todo{??}}
+\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{m}
+\\
+{[\dexp_1 / x] \dcast{\htau}{\dexp}}
+&=&
+\dcast{\htau}{[\dexp_1 / x] \dexp}
 \end{array}
@@
 { }
-{\isIndet{\dehole{\mvar}{\subst}{\todo{??}}}}
+{\isIndet{\dehole{\mvar}{\subst}{\evaled}}}
 
@@
 {\isFinal{\dexp}}
-{\isIndet{\dhole{\dexp}{\mvar}{\subst}{\todo{??}}}}
+{\isIndet{\dhole{\dexp}{\mvar}{\subst}{\evaled}}}
 
@@
 \begin{mathpar}
-\Infer{FVal}
+\inferrule[FVal]
 {\isValue{\dexp}}{\isFinal{\dexp}}
 \and
-\Infer{FIndet}
+\inferrule[FIndet]
 {\isIndet{\dexp}}{\isFinal{\dexp}}
-%% No longer a rule: Errors are not indet anymore:
-%% \and
-%% \Infer{f-err}
-%% {\isErr{\dexp}}{\isFinal{\dexp}}
 \end{mathpar}
 
-\todo{the judgmental forms are not consistent in these
-rules; the contexts change. need to figure this out.}
+\judgbox{\isErr{\Delta}{\dexp}}{$\dexp$ contains a cast error}
+\begin{mathpar}
+\inferrule[ECastError]{
+  \hasType{\Delta}{\emptyset}{\dexp}{\htau_2}\\
+  \tinconsistent{\htau_1}{\htau_2}
+}{
+  \isErr{\Delta}{\dcast{\htau_1}{\dexp}}
+}
 
-\judgbox{\isErr{\Delta}{\dexp}}{Under $\Delta$, $\dexp$ contains a cast error}
-\begin{mathpar}
-\inferrule[ERNEHole~\todo{??}]
-{\subst \sim_? \Gamma_2
-\\
-\isErr{\Delta; \Gamma_1, \Gamma_2}{\dexp}}
-{\isErr{\Delta; \Gamma_1}{\dhole{\dexp}{\mvar}{\subst}}}
+\inferrule[EConst]{
+}{
+  \isErr{\Delta}{\dcast{\tau}{\dconst}}
+}
 
-\inferrule[ERCastError~\todo{??}]
-% XXX use macro type assignment judgement here
-{\hasType{\Delta}{\todo{\emptyset}}{e}{\htau_2}\\
- \tinconsistent{\htau_1}{\htau_2}}
-{\isErr{\Delta}{\dcast{\htau_1}{\dexp}}}
+\inferrule[EAp1]{
+  \isErr{\Delta}{\dexp_1}
+}{
+  \isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}
+}
 
-\inferrule[ERLam~\todo{??}]
-% XXX use macro type assignment judgement here
-{\isErr{\Delta; \Gamma, x:\htau_1}{\dexp}}
-{\isErr{\Delta; \Gamma}{\halam{x}{\htau}{\dexp}}}
+\inferrule[EAp2]{
+  \isErr{\Delta}{\dexp_2}
+}{
+  \isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}
+}
 
-\\
+\inferrule[ENEHole]{
+  \isErr{\Delta}{\dexp}
+}{
+  \isErr{\Delta}{\dhole{\dexp}{\mvar}{\subst}{m}}
+}
 
-\inferrule[ERAp1]
-{
- \isErr{\Delta}{\dexp_1}
- %\\\
- %isFinal{\dexp_2}~\text{\todo{??}}
+\inferrule[ECastProp]{
+  \isErr{\Delta}{\dexp}
+}{
+  \isErr{\Delta}{\dcast{\htau_1}{\dexp}}
 }
-{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}
+\end{mathpar}
+\caption{Auxiliary Judgments for the Dynamics}
+\label{fig:isValue}
+\label{fig:isIndet}
+\label{fig:isFinal}
+\label{fig:isErr}
+\end{figure}
 
-\inferrule[ERAp2]
-{
- %\isIndet{\dexp_1}
- %\\
- \isErr{\Delta}{\dexp_2}
-}
-{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}
-
-\inferrule[ERCast]
-% XXX use macro type assignment judgement here
-{\isErr{\Delta}{\dexp}}
-{\isErr{\Delta}{\dcast{\htau_1}{\dexp}}}
-
-\end{mathpar}
-
-\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{under $\Delta$, $\dexp_1$
+\begin{figure}
+\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$
 steps to $\dexp_2$}
 \begin{mathpar}
-\Infer{STHole}
+\inferrule[STEHoleEvaled]
+{ }
+{\stepsToD{\Delta}{\dehole{\mvar}{\subst}{\unevaled}}{\dehole{\mvar}{\subst}{\evaled}}}
+
+\inferrule[STNEHoleStep]
 {\stepsToD{\Delta}{\dexp_1}{\dexp_2} }
-{\stepsToD{\Delta}{\dhole{\dexp_1}{\mvar}{\subst}{\todo{??}}}{\dhole{\dexp_2}{\mvar}{\subst}{\todo{??}}}}
-\and
-\Infer{STCast}
-{\hasType{\Delta}{\emptyset}{\dexp}{\htau_2} \\
+{\stepsToD{\Delta}{\dhole{\dexp_1}{\mvar}{\subst}{\unevaled}}{\dhole{\dexp_2}{\mvar}{\subst}{\unevaled}}}
+
+\inferrule[STNEHoleEvaled]
+{\isFinal{\dexp}}
+{\stepsToD{\Delta}{\dhole{\dexp}{\mvar}{\subst}{\unevaled}}{\dhole{\dexp}{\mvar}{\subst}{\evaled}}}
+
+\inferrule[STCast]
+{
+\isValue{\dexp}\\
+\hasType{\Delta}{\emptyset}{\dexp}{\htau_2} \\
 \tconsistent{\tau_1}{\tau_2}}
 {\stepsToD{\Delta}{\dcast{\htau_1}{\dexp}}{\dexp}}
-\\
-\Infer{STAp1}
+
+\inferrule[STApStep1]
 {\stepsToD{\Delta}{\dexp_1}{\dexp_1'}}
 {\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
-\and
-\Infer{STAp2}
+
+\inferrule[STApStep2]
 { \isFinal{\dexp_1} \\ \stepsToD{\Delta}{\dexp_2}{\dexp_2'}}
 {\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
-\and
-\Infer{STAp$\beta$}
+
+\inferrule[STApBeta]
 { \isFinal{\dexp_2} }
@@
 \caption{Structural Dynamics}
-\label{fig:isValue}
-\label{fig:isIndet}
 \label{fig:stepsTo}
+\end{figure}
+
+\begin{figure}
+$\arraycolsep=4pt\begin{array}{rllllll}
+\mathsf{EvalCtx} & \evalctx & ::= &
+  \evalhole ~\vert~
+  \halam{x}{\htau}{\evalctx} ~\vert~
+  \hap{\evalctx}{\dexp} ~\vert~
+  \hap{\dexp}{\evalctx} ~\vert~
+  \dhole{\evalctx}{\mvar}{\subst}{\markname} ~\vert~
+  \dcast{\htau}{\evalctx}
+\end{array}$
+
+\judgbox{\isevalctx{\evalctx}}{$\evalctx$ is an evaluation context}
+\begin{mathpar}
+\inferrule[ECDot]{ }{
+  \isevalctx{\evalhole}
+}
+
+\inferrule[ECLam]{
+  \isevalctx{\evalctx}
+}{
+  \isevalctx{\halam{x}{\htau}{\evalctx}}
+}
+
+\inferrule[ECAp1]{
+  \isFinal{\dexp}\\
+  \isevalctx{\evalctx}
+}{
+  \isevalctx{\hap{\dexp}{\evalctx}}
+}
+
+\inferrule[ECAp2]{
+  \isevalctx{\evalctx}
+}{
+  \isevalctx{\hap{\evalctx}{\dexp}}
+}
+
+\inferrule[ECNEHole]{
+  \isevalctx{\evalctx}
+}{
+  \isevalctx{\dhole{\evalctx}{\mvar}{\subst}{\markname}}
+}
+
+\inferrule[ECCast]{
+  \isevalctx{\evalctx}
+}{
+  \isevalctx{\dcast{\htau}{\evalctx}}
+}
+\end{mathpar}
+
+\judgbox{\selectEvalCtx{d}{\evalctx}{d'}}{$d$ is the result of filling the
+hole in $\evalctx$ with $d'$}
+\begin{mathpar}
+\inferrule[FDot]{ }{
+  \selectEvalCtx{d}{\evalhole}{d}
+}
+
+%% \todo{d used to be c -- TODO IEV}
+
+\inferrule[FLam]{
+  \selectEvalCtx{d}{\evalctx}{d'}
+}{
+  \selectEvalCtx{\halam{x}{\htau}{d}}{\halam{x}{\htau}{\evalctx}}{d'}
+}
+
+\inferrule[FAp1]{
+  \isFinal{d_1}\\
+  \selectEvalCtx{d_2}{\evalctx}{d_2'}
+}{
+  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{d_1}{\evalctx}}{d_2'}
+}
+
+\inferrule[FAp2]{
+  \selectEvalCtx{d_1}{\evalctx}{d_1'}
+}{
+  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{\evalctx}{d_2}}{d_1'}
+}
+
+%% I think these rules are not needed, if you make FDot work for any dhexp
+%% d. The only reason they might be needed is because they're specified to
+%% have check marks, not x marks, but i don't think that's important. i
+%% think the more general FDot makes sense. -- TODO IEV
+%%
+%%
+%% \inferrule[\todo{??}]{ }{
+%%   \selectEvalCtx{\dehole{\mvar}{\subst}{\evaled}}{\evalhole}{\dehole{\mvar}{\subst}{\evaled}}
+%% }
+
+%% \inferrule[\todo{??}]{ }{
+%%   \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\evaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\evaled}}
+%% }
+
+\inferrule[\todo{??}]{
+  \selectEvalCtx{d}{\evalctx}{d'}
+}{
+  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\dhole{\evalctx}{\mvar}{\subst}{\unevaled}}{d'}
+}
+
+\inferrule[\todo{??}]{
+  \isFinal{d}
+}{
+  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\unevaled}}
+}
+
+\inferrule[FCast]{
+  \selectEvalCtx{d}{\evalctx}{d'}
+}{
+  \selectEvalCtx{\dcast{\htau}{d}}{\dcast{\htau}{\evalctx}}{d'}
+}
+\end{mathpar}
+
+\judgbox{\reducesE{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$
+transitions to $\dexp_2$}
+\begin{mathpar}
+\inferrule[ITLam]{
+  \isFinal{d_2}
+}{
+  \reducesE{\Delta}{\hap{(\halam{x}{\htau}{d})}{d_2}}{[d_2/x]d}
+}
+
+\inferrule[ITCast]{
+  \isFinal{d}\\
+  \hasType{\Delta}{\emptyset}{d}{\tau_2}\\
+  \tconsistent{\tau_1}{\tau_2}
+}{
+  \reducesE{\Delta}{\dcast{\htau_1}{d}}{d}
+}
+
+\inferrule[ITEHole]{ }{
+  \reducesE{\Delta}{\dehole{\mvar}{\subst}{\unevaled}}{\dehole{\mvar}{\subst}{\evaled}}
+}
+
+\inferrule[ITNEHole]{
+  \isFinal{d}
+}{
+  \reducesE{\Delta}{\dhole{d}{\mvar}{\subst}{\unevaled}}{\dhole{d}{\mvar}{\subst}{\evaled}}
+}
+\end{mathpar}
+
+\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
+\begin{mathpar}
+\inferrule[Step]{
+  \selectEvalCtx{d}{\evalctx}{d_0}\\
+  \reducesE{\Delta}{d_0}{d_0'}\\
+  \selectEvalCtx{d'}{\evalctx}{d_0'}
+}{
+  \stepsToD{\Delta}{\dexp}{\dexp'}
+}
+\end{mathpar}
+\caption{Contextual Dynamics}
+\label{fig:contextual-dynamics}
 \end{figure}
@@
 \begin{theorem}[Progress]
-If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\stepsToD{\Delta}{\dexp}{\dexp'}$.
+If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
+$\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or
+$\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.
 \end{theorem}
@@
 \begin{theorem}[Preservation]
-  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and $\stepsToD{\Delta}{\dexp}{\dexp'}$ then $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \todo{or maybe up to consistency?}
+  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
+  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
+  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \todo{or maybe up to
+  consistency?}
 \end{theorem}
 
-\todo{complete progress + preservation}
+\begin{theorem}[Complete Progress \todo{??}]
+If $\hcomplete{\hexp}$ and
+$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ then either
+$\isValue{\dexp}$ or $\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.
+
+\end{theorem}
+
+\begin{theorem}[Complete Preservation \todo{??}]
+If $\hcomplete{\hexp}$ and
+$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ and
+$\stepsToD{\Delta}{\dexp}{\dexp'}$ then
+$\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$
+\end{theorem}
+
+\begin{theorem}[Commutativity \todo{state this}]
+\end{theorem}
+
+\begin{theorem}[Maximum Informativity]
+If the expansion produces $t1$, and there exists another possible type choice
+$t2$, then $t1 \sim t2$ and $t1 JOIN t2 = t1$
+\end{theorem}\footnote{idea is that special casing the holes in EANEHole gives you ``the
+most descriptive hole types'' for some sense of what that means -- they'd
+all just be hole other wise. from Matt:
+\begin{quote}
+It sounds like we need a something akin to an abstract domain (a lattice),
+where hole has the least information, and a fully-defined type (without
+holes) has the most information.  You can imagine that this lattice really
+expands the existing definition we have of type consistency, which is
+merely the predicate that says whether two types are comparable
+(“join-able”) in this lattice.  lattice join is the operation that goes
+through the structure of two (consistent) types, and chooses the structure
+that is more defined (i.e., non-hole, if given the choice between hole and
+non-hole).
+
+The rule choosenonhole below is the expansion of this consistency rule that
+we already have (hole consistent with everything)
+\end{quote}}
+
+\begin{verbatim}
+t not hole
+-------------------- :: choose-non-hole
+hole JOIN t  = t
+\end{verbatim}
+\begin{verbatim}
+------------ :: hole-consistent-with-everything
+hole ~ t
+\end{verbatim}
 
@@
 \todo{indeterminate forms}
-
-\todo{theorem that special casing the holes in EANEHole gives you ``the
-most descriptive hole types'' for some sense of what that means -- they'd
-all just be hole other wise}

### 2018-01-11 calculus.tex
adding rule names, cleaning up after talking to cyrus
missing? FHCastFinal rule
removing icast rule
changes?

@@
   \hexp : \htau\\
-\mathsf{Mark} & \markname{} & ::= &
-  \evaled{} ~\vert~  \unevaled{}\\
+% \mathsf{Mark} & \markname{} & ::= &
+%   \evaled{} ~\vert~  \unevaled{}\\
  \mathsf{DHExp} & \dexp  & ::= &
@@
   % \hadd{\dexp}{\dexp} ~\vert~
-  \dehole{\mvar}{\subst}{\markname} ~\vert~
-  \dhole{\dexp}{\mvar}{\subst}{\markname} ~\vert~
+  \dehole{\mvar}{\subst}{} ~\vert~
+  \dhole{\dexp}{\mvar}{\subst}{} ~\vert~
   \dcast{\htau}{\dexp}\\
@@
 }{
-  \hasType{\Delta}{\hGamma}{\dehole{u}{\sigma}{\markname}}{\htau}
+  \hasType{\Delta}{\hGamma}{\dehole{u}{\sigma}{}}{\htau}
 }
@@
 }{
-  \hasType{\Delta}{\hGamma}{\dhole{\dexp}{u}{\sigma}{\markname}}{\htau}
+  \hasType{\Delta}{\hGamma}{\dhole{\dexp}{u}{\sigma}{}}{\htau}
 }
@@
 % \\
-{[\dexp_1 / x]} \dehole{\mvar}{\subst}{m}
+{[\dexp_1 / x]} \dehole{\mvar}{\subst}{}
 &=&
-\dehole{\mvar}{[\dexp_1 / x] \subst}{m}
+\dehole{\mvar}{[\dexp_1 / x] \subst}{}
 \\
-{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{m}}
+{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{}}
 &=&
-\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{m}
+\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{}
 \\
@@
 }{
-  \isErr{\Delta}{\dhole{\dexp}{\mvar}{\subst}{m}}
+  \isErr{\Delta}{\dhole{\dexp}{\mvar}{\subst}{}}
 }
@@
 
+%%   \halam{x}{\htau}{\evalctx} ~\vert~
+
 \begin{figure}
@@
   \evalhole ~\vert~
-  \halam{x}{\htau}{\evalctx} ~\vert~
   \hap{\evalctx}{\dexp} ~\vert~
   \hap{\dexp}{\evalctx} ~\vert~
-  \dhole{\evalctx}{\mvar}{\subst}{\markname} ~\vert~
+  \dhole{\evalctx}{\mvar}{\subst}{} ~\vert~
   \dcast{\htau}{\evalctx}
@@
 
-\inferrule[ECLam]{
-  \isevalctx{\evalctx}
-}{
-  \isevalctx{\halam{x}{\htau}{\evalctx}}
-}
+%% \inferrule[ECLam]{
+%%   \isevalctx{\evalctx}
+%% }{
+%%   \isevalctx{\halam{x}{\htau}{\evalctx}}
+%% }
 
@@
 }{
-  \isevalctx{\dhole{\evalctx}{\mvar}{\subst}{\markname}}
+  \isevalctx{\dhole{\evalctx}{\mvar}{\subst}{}}
 }
@@
 \begin{mathpar}
-\inferrule[FDot]{ }{
-  \selectEvalCtx{d}{\evalhole}{d}
+\inferrule[FHConst]{ }{
+  \selectEvalCtx{c}{\evalhole}{c}
 }
 
-%% \todo{d used to be c -- TODO IEV}
+%% \inferrule[FLam]{
+%%   \selectEvalCtx{d}{\evalctx}{d'}
+%% }{
+%%   \selectEvalCtx{\halam{x}{\htau}{d}}{\halam{x}{\htau}{\evalctx}}{d'}
+%% }
 
-\inferrule[FLam]{
-  \selectEvalCtx{d}{\evalctx}{d'}
-}{
-  \selectEvalCtx{\halam{x}{\htau}{d}}{\halam{x}{\htau}{\evalctx}}{d'}
-}
-
-\inferrule[FAp1]{
+\inferrule[FHAp1]{
   \isFinal{d_1}\\
@@
 
-\inferrule[FAp2]{
+\inferrule[FHAp2]{
   \selectEvalCtx{d_1}{\evalctx}{d_1'}
@@
 
-%% I think these rules are not needed, if you make FDot work for any dhexp
-%% d. The only reason they might be needed is because they're specified to
-%% have check marks, not x marks, but i don't think that's important. i
-%% think the more general FDot makes sense. -- TODO IEV
-%%
-%%
-%% \inferrule[\todo{??}]{ }{
-%%   \selectEvalCtx{\dehole{\mvar}{\subst}{\evaled}}{\evalhole}{\dehole{\mvar}{\subst}{\evaled}}
-%% }
+\inferrule[FHEHole]{ }{
+  \selectEvalCtx{\dehole{\mvar}{\subst}{}}{\evalhole}{\dehole{\mvar}{\subst}{}}
+}
 
-%% \inferrule[\todo{??}]{ }{
-%%   \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\evaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\evaled}}
-%% }
+\inferrule[FHNEHoleEvaled]{ }{
+  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\evaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\evaled}}
+}
 
-\inferrule[\todo{??}]{
+\inferrule[FHNEHoleInside]{
   \selectEvalCtx{d}{\evalctx}{d'}
@@
 
-\inferrule[\todo{??}]{
+\inferrule[FHNEHoleFinal]{
   \isFinal{d}
@@
 
-\inferrule[FCast]{
+\inferrule[FHCast]{
   \selectEvalCtx{d}{\evalctx}{d'}
@@
   \selectEvalCtx{\dcast{\htau}{d}}{\dcast{\htau}{\evalctx}}{d'}
+}
+
+\inferrule[FHCastFinal]{
+  \isFinal{d}
+}{
+  \selectEvalCtx{\dcast{\htau}{d}}{\evalhole}{\dcast{\htau}{d}}
 }
@@
 \begin{theorem}[Progress]
-If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
-$\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or
-$\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.
+If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\stepsToD{\Delta}{\dexp}{\dexp'}$ or 
+$\isValue{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\isIndet{\dexp}$. 
+% for some $\dexp'$.
 \end{theorem}

### 2018-02-06 hazelnut-dynamics.tex
updated to latest acmart cls file and updated title

@@
 % \newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\newtheorem{condition}[theorem]{Condition}
+% \newtheorem{assumption}[theorem]{Assumption}
+% \newtheorem{condition}[theorem]{Condition}
 
@@
 \acmArticle{1}
-\acmYear{2017}
+\acmYear{2018}
 \acmMonth{1}
@@
 %% Supplied to authors by publisher for camera-ready submission
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
+\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
+\acmYear{2018}
 \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
@@
 %% Title information
-\title{Live Programming with Typed Holes}         %% [Short Title] is optional;
+\title{Live Functional Programming with Typed Holes}         %% [Short Title] is optional;
                                         %% when present, will be used in

### 2018-02-06 calculus.tex
pasted in relevant blurbs from LIVE2017 paper and grant to use as starting text

@@
 \label{sec:calculus}
-\todo{exposition}
+\cy{exposition}
 
@@
 
-\todo{definition of complete types and terms? do we care about completeness
+\cy{definition of complete types and terms? do we care about completeness
   of $\hexp$s or $\dexp$s?}
@@
 {\isIndet{\dexp_1}\\
-% \isFinal{\dexp_2}~\text{\todo{??}}}
+% \isFinal{\dexp_2}~\text{\cy{??}}}
  \isFinal{\dexp_2}}
@@
 
-\inferrule[ICast]
-{\isIndet{\dexp}}
-{\isIndet{\dcast{\htau}{\dexp}}}
+%% \inferrule[ICast]
+%% {\isIndet{\dexp}}
+%% {\isIndet{\dcast{\htau}{\dexp}}}
 
@@
 
-\begin{theorem}[Expandability \todo{??}] ~
+\begin{theorem}[Expandability \cy{??}] ~
   \begin{enumerate}
@@
 
-\begin{theorem}[Correspondence \todo{??}] ~
+\begin{theorem}[Correspondence \cy{??}] ~
   \begin{enumerate}
@@
   $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
-  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \todo{or maybe up to
+  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \cy{or maybe up to
   consistency?}
@@
 
-\begin{theorem}[Complete Progress \todo{??}]
+\begin{theorem}[Complete Progress \cy{??}]
 If $\hcomplete{\hexp}$ and
@@
 
-\begin{theorem}[Complete Preservation \todo{??}]
+\begin{theorem}[Complete Preservation \cy{??}]
 If $\hcomplete{\hexp}$ and
@@
 
-\begin{theorem}[Commutativity \todo{state this}]
+\begin{theorem}[Commutativity \cy{state this}]
 \end{theorem}
@@
 
-\todo{canonical forms}
+\cy{canonical forms}
 
-\todo{indeterminate forms}
+\cy{indeterminate forms}

### 2018-02-06 hazelnut-dynamics.tex
pasted in relevant blurbs from LIVE2017 paper and grant to use as starting text

@@
 %% 'sigplan,10pt'.
+
+\bibliographystyle{ACM-Reference-Format}
 
@@
 \usepackage{leftidx}
-\newcommand{\todo}[1]{{\color{red} #1}}
+\usepackage{todonotes}
+
+\usepackage{listings}%
+\lstloadlanguages{ML}
+\lstset{tabsize=2, 
+basicstyle=\footnotesize\ttfamily, 
+% keywordstyle=\sffamily,
+commentstyle=\itshape\ttfamily\color{gray}, 
+stringstyle=\ttfamily\color{purple},
+mathescape=false,escapechar=\#,
+numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
+morekeywords={string, float, int},
+classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount
+}
+\let\li\lstinline
 
@@
 
+\input{commands}
 \input{macros}
@@
 \begin{abstract}
-\todo{abstract}
+\cy{blurb from intro of grant below}
+
+\noindent
+Program editors are increasingly incorporating \emph{live programming
+services}~\cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90,McDirmid:2007,burckhardt2013s}, which interleave editing with
+evaluation of the program.
+%
+However, live programming services typically cannot run incomplete
+programs, or they raise an exception as soon as a hole is encountered
+during evaluation.
+%
+%% When they % encounter an incomplete program, a common strategy is
+%to simply % disable the live programming service, or revert to the
+%last complete % edit state.
+%
+This deprives the programmer, and also other editor services, of
+potentially valuable feedback about the dynamic behavior of the
+visible program.
+Our second track of research will address this gap by defining a dynamic semantics for incomplete programs.
+We plan to focus the first year of research efforts largely on this track of research, as indicated in \autoref{fig:tracks}.
+% Evaluation of incomplete programs will proceed as far as possible past the hole, and in the process, will track the environment around each hole instance.
+This semantics will support live programming services like that outlined in our example above, which provides the programmer with a result on every, or nearly every, edit and allows the programmer to choose a hole instance in the computed result of evaluation and inspect its environment (in the sidebar of \autoref{fig:intro-example}) while they determine how to best fill that hole.
+Tracking each hole instance's environment will also enhance the editor services specified by the remaining two tracks of research, \ie{}~the action suggestion panel (\trackThree{}) and the direct manipulation programming service (\trackFour{}).
+% We will prove that in certain situations, hole filling commutes with evaluation.
+% This formal observation will lead to a decrease in the cost of providing continuous feedback to the programmer on every edit by avoiding redundant recomputation.
+Our preliminary work, described in a presentation at the \emph{LIVE 2017}
+workshop~\cite{HazelLIVE}, has discovered encouraging connections between our proposed dynamics and previous research on contextual modal type theory (the Curry-Howard interpretation of contextual modal logic) \cite{Nanevski2008} and gradual type theory \cite{Siek06a}.
+
 \end{abstract}
@@
 
+\bibliography{references,all.short,hazel_NSF}
+
 \end{document}

### 2018-02-06 intro.tex
wrote down major core metatheorems
pasted in relevant blurbs from LIVE2017 paper and grant to use as starting text

@@
 \label{sec:intro}
-\todo{intro}
+\cy{intro from LIVE 2017 paper is below}
+
+Broadly speaking, live programming environments are those that granularly interleave editing and evaluation \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90,McDirmid:2007:LUL:1297027.1297073,burckhardt2013s}.
+In the words of \citet{burckhardt2013s}, live programming environments 
+``promise to narrow the temporal and perceptive gap 
+between program development and code execution''. Examples of live programming environments include {lab notebook environments},
+e.g. the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
+programmer to interactively edit and evaluate program fragments organized into a
+sequence of cells (an extension of the read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007:LUL:1297027.1297073}, Sketch-n-Sketch \cite{DBLP:conf/pldi/ChughHSA16} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples.
+
+The problem that has motivated much of our recent work is that most  
+programming environments, live programming environments included,  provide feedback via various editor services only once the program being edited is syntactically well-formed and, when relevant, well-typed.
+This leaves a ``temporal and perceptive gap'', because programmers sometimes leave a program malformed or ill-typed for extended periods of time, e.g.
+as they think about what to enter at the cursor, or as they work on a different part of the program.
+
+In view of this general problem, we recently developed a \emph{structure editor calculus} called Hazelnut where every edit state consists of a well-formed and statically meaningful, i.e.
+well-typed, incomplete expression, which we take to mean an expression with holes \cite{popl-paper}.
+This calculus addressed fundamental questions relevant to editor services that operate statically, but there was no solution in that paper to the problems faced by editor services that also require knowledge of the dynamic meaning of an incomplete program, as would be relevant to a live programming environment.
+For example, consider a \emph{stepper} (a.k.a.
+a \emph{stepwise debugger}), like that available to Haskell programmers in the GHCi system \cite{GHC-stepper} and other systems \cite{DBLP:conf/haskell/MarlowIPG07,kar13566}, to OCaml programmers with recent work on \texttt{ocamli} by \citet{ocaml-stepper} and to Standard ML programmers \cite{DBLP:journals/jfp/TolmachA95}.
+A stepper requires that the expression being stepped be assigned dynamic meaning according to a small-step operational semantics \cite{DBLP:journals/jlp/Plotkin04a,pfpl}, but no such semantics was defined for incomplete expressions that arise when using Hazelnut, or any of these other systems.
+Defining such a semantics was left as future work in the Hazelnut paper, and in a subsequent ``vision paper'' \cite{snapl17-paper}.
+The purpose of this paper is to sketch out our progress toward a theoretically well-grounded solution.
+
+\cy{intro from grant section is below}
+
+Live programming environments granularly interleave editing and evaluation,
+``promis[ing] to narrow the temporal and perceptive gap 
+between program development and code execution''~\cite{burckhardt2013s}.
+Examples of live programming environments include {lab notebook environments},
+like the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
+programmer to interactively edit and evaluate program fragments organized into a
+sequence of cells (an extension of the ubiquitous read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples.
+\cy{find somewhere to cite these tanimoto papers \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90}}
+
+Live programming environments typically cannot run incomplete programs, or they simply halt or raise an exception as soon as they encounter a hole during evaluation.
+This leaves a ``temporal and perceptive gap'' because the results of even those computations that do not depend critically on the hole are not available to the programmer.
+Our second track of research will address this gap by defining a dynamic semantics for incomplete programs that proceeds as far as possible around holes.
+Every program assigned static meaning by the semantics of Track 1 will also be assigned dynamic meaning by the semantics of Track 2.
+
+For example, in \autoref{fig:intro-example} we saw that mapping the incomplete function \li{weighted_average} over the \li{grades} list produced an incomplete result.
+If holes simply raised exceptions, then the system would have aborted evaluation as soon as \li{weighted_average} was applied for the first time by \li{map}, and the programmer would not have been able to see that the result was a list of two elements, nor see the value of the sub-expression \li{10.0 *.
+g.hw1} for each element \li{g} of the \li{grades} list.
+Even programs with type inconsistencies can be evaluated under our proposed dynamics, because, as discussed in Track 1, non-empty holes operate as membranes around type inconsistencies.
+Taken together, the results from Tracks 1 and 2 may help to address a common complaint: that a static type discipline makes it difficult to do exploratory programming.
+
+Our proposed dynamics goes beyond staged evaluation (\eg{}~\cite{Taha:1999}), partial evaluation (\eg{}~\cite{Jones:1993uq}) and symbolic evaluation (\eg{}~\cite{King:1976}) in that we track the dynamic environment (\ie{}~the substitutions performed) around each instance of an expression hole in the result.
+This allows the programmer to see, in the sidebar of \autoref{fig:intro-example}, the actual values that the variables in scope take on everywhere that the hole they are filling ends up during evaluation.
+This provides a hole-oriented specification of a standard feature of debuggers: inspection of the environment at designated points in the program.
+Finally, this environmental information is useful (though not required) for the edit action suggestion service of Track 3 and the direct manipulation programming service of Track 4.

### 2018-02-07 hazelnut-dynamics.tex
draft of an abstract (probably a bit too long and meandering/unclear, but its a start)

--- 2018-02-06 hazelnut-dynamics.tex
@@
 \begin{abstract}
-\todo{abstract}
+\emph{Live programming environments} are designed to provide programmers (and, in some cases, audiences) with rapid and continuous feedback about a program's dynamic behavior throughout the editing process.
+The problem is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas programming environments often encounter edit states that do not correspond to complete programs.
+Recent work by \citet{popl-paper} has made progress on this problem of working with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies.
+This paper contributes a complementary dynamic semantics, i.e.
+one that assigns dynamic meaning to every incomplete program accepted by the type system in the prior work.
+Of course, it would not be entirely satisfying to assign trivial dynamic meaning to incomplete programs, e.g.
+by simply raising an exception as soon as evaluation first encounters a hole (as in some existing systems).
+Instead, evaluation proceeds ``around'' expression holes, performing as much computation as is possible while also tracking the dynamic environment around each expression hole as it flows through the program.
+These hole environments can be reported directly to the programmer as they work to fill the hole, and they also support an ``edit-and-resume'' feature for edits that amount to hole filling, rather than requiring that evaluation restart after each such edit.
+Formally, our semantics incorporates features from both gradual type theory (to support evaluation of terms with type holes) and contextual modal type theory (to support environment tracking), combined with technical novelties necessary to maintain type safety in the presence of holes and prove the commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound.
+The primary focus of this paper is on these theoretical properties, which we have mechanized using the Agda proof assistant.
+We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions, based on those defined in the prior work by \citet{popl-paper}, that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type.
+This property, together with the type safety property we prove, results in a simple live functional programming environment where \emph{every edit state has both static and dynamic meaning}.
+
+%In our system, evaluation treats failed casts much like it treats expression holes (rather than immediately failing with a cast error, as in gradual type theory).
+Prior work on contextual modal type theory did not develop an operational semantics for programs with free metavariables, which correspond to programs with holes in our formulation.
+
+% These incomplete edit states are sometimes transient, but at other times, they persist for extended periods of time, e.g.
+when the programmer is filling in the branches of a large case analysis one-by-one, or when an edit causes type errors to appear throughout a program.
+%This gap can cause live programming services to lag behind the programmer's edits, in many cases for an extended period of time.
+
+%
+
+%
+% However, live programming services typically cannot run incomplete
+% programs, or they raise an exception as soon as a hole is encountered
+% during evaluation.
+% %
+% %% When they % encounter an incomplete program, a common strategy is
+% %to simply % disable the live programming service, or revert to the
+% %last complete % edit state.
+% %
+% This deprives the programmer, and also other editor services, of
+% potentially valuable feedback about the dynamic behavior of the
+% visible program.
+Our second track of research will address this gap by defining a dynamic semantics for incomplete programs.
+We plan to focus the first year of research efforts largely on this track of research, as indicated in \autoref{fig:tracks}.
+% % Evaluation of incomplete programs will proceed as far as possible past the hole, and in the process, will track the environment around each hole instance.
+% This semantics will support live programming services like that outlined in our example above, which provides the programmer with a result on every, or nearly every, edit and allows the programmer to choose a hole instance in the computed result of evaluation and inspect its environment (in the sidebar of \autoref{fig:intro-example}) while they determine how to best fill that hole.
+% Tracking each hole instance's environment will also enhance the editor services specified by the remaining two tracks of research, \ie{}~the action suggestion panel (\trackThree{}) and the direct manipulation programming service (\trackFour{}).
+% % We will prove that in certain situations, hole filling commutes with evaluation.
+% % This formal observation will lead to a decrease in the cost of providing continuous feedback to the programmer on every edit by avoiding redundant recomputation.
+% Our preliminary work, described in a presentation at the \emph{LIVE 2017}
+% workshop~\cite{HazelLIVE}, has discovered encouraging connections between our proposed dynamics and previous research on contextual modal type theory (the Curry-Howard interpretation of contextual modal logic) \cite{Nanevski2008} and gradual type theory \cite{Siek06a}.
+
 \end{abstract}
@@
 \input{intro}
+\input{examples}
 \input{calculus}
+\input{related-work}
+\input{discussion}
+
+\bibliography{references,all.short,hazel_NSF}
+
+\clearpage
+\appendix
+\cy{for the submission, appendix should be separated into a separate supplemental PDF}
+
+\input{extensions}
 

### 2018-02-07 discussion.tex
added missing .tex files

% !TEX root = hazelnut-dynamics.tex
\newcommand{\discussionSection}{Discussion}
\section{\protect\discussionSection} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:discussion}

\todo{discussion}

### 2018-02-07 examples.tex
added missing .tex files

% !TEX root = hazelnut-dynamics.tex
\newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:examples}

\todo{examples}

### 2018-02-07 extensions.tex
added missing .tex files

% !TEX root = hazelnut-dynamics.tex
\newcommand{\extensionsSec}{Extensions to Hazelnut Live}
\section{\protect\extensionsSec} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:extensions}
\subsection{Product Types}

\subsection{Sum Types}

\subsection{Recursive Types}

\subsection{Polymorphism}

\subsection{Mutable References}
\cy{if we add references, would have to talk about how commutativity isn't conserved}

### 2018-02-07 related-work.tex
added missing .tex files

% !TEX root = hazelnut-dynamics.tex
\newcommand{\relatedWorkSection}{Related Work}
\section{\protect\relatedWorkSection} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:relatedWork}

\todo{related work}

### 2018-02-07 hazelnut-dynamics.tex
some tweaks to abstract

@@
 These hole environments can be reported directly to the programmer as they work to fill the hole, and they also support an ``edit-and-resume'' feature for edits that amount to hole filling, rather than requiring that evaluation restart after each such edit.
-Formally, our semantics incorporates features from both gradual type theory (to support evaluation of terms with type holes) and contextual modal type theory (to support environment tracking), combined with technical novelties necessary to maintain type safety in the presence of holes and prove the commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound.
+Formally, our semantics incorporates features from both gradual type theory (to support evaluation of terms with type holes) and contextual modal type theory (CMTT, to support environment tracking), combined with technical novelties necessary to maintain type safety in the presence of expression holes (i.e.
+free metavariables, from the perspective of CMTT) and prove the commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound.
 The primary focus of this paper is on these theoretical properties, which we have mechanized using the Agda proof assistant.
-We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions, based on those defined in the prior work by \citet{popl-paper}, that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type.
-This property, together with the type safety property we prove, results in a simple live functional programming environment where \emph{every edit state has both static and dynamic meaning}.
+We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions based on those defined in the prior work by \citet{popl-paper} that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type.
+This property, together with the type safety property we prove, results in a live functional programming environment where feedback never lags the edit state, i.e.
+\emph{every edit state is guaranteed to have non-trivial static and dynamic meaning}.
 

### 2018-02-07 hazelnut-dynamics.tex
tightening up the abstract

@@
 \begin{abstract}
-\emph{Live programming environments} are designed to provide programmers (and, in some cases, audiences) with rapid and continuous feedback about a program's dynamic behavior throughout the editing process.
-The problem is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas programming environments often encounter edit states that do not correspond to complete programs.
-Recent work by \citet{popl-paper} has made progress on this problem of working with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies.
-This paper contributes a complementary dynamic semantics, i.e.
-one that assigns dynamic meaning to every incomplete program accepted by the type system in the prior work.
-Of course, it would not be entirely satisfying to assign trivial dynamic meaning to incomplete programs, e.g.
-by simply raising an exception as soon as evaluation first encounters a hole (as in some existing systems).
-Instead, evaluation proceeds ``around'' expression holes, performing as much computation as is possible while also tracking the dynamic environment around each expression hole as it flows through the program.
-These hole environments can be reported directly to the programmer as they work to fill the hole, and they also support an ``edit-and-resume'' feature for edits that amount to hole filling, rather than requiring that evaluation restart after each such edit.
-Formally, our semantics incorporates features from both gradual type theory (to support evaluation of terms with type holes) and contextual modal type theory (CMTT, to support environment tracking), combined with technical novelties necessary to maintain type safety in the presence of expression holes (i.e.
-free metavariables, from the perspective of CMTT) and prove the commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound.
-The primary focus of this paper is on these theoretical properties, which we have mechanized using the Agda proof assistant.
-We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions based on those defined in the prior work by \citet{popl-paper} that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type.
-This property, together with the type safety property we prove, results in a live functional programming environment where feedback never lags the edit state, i.e.
+\emph{Live programming environments} aim to provide programmers (and, in many cases, audiences) with continuous feedback about a program's dynamic behavior throughout the editing process.
+The problem that motivates this paper is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas program editors frequently encounter edit states that are incomplete.
+Recent work by \citet{popl-paper} has made progress on the problem of working directly with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies.
+This paper develops a corresponding dynamic semantics.
+Rather than aborting with an exception when evaluation encounters a hole (as in  several existing systems), evaluation proceeds ``around'' expression holes, performing as much computation as is possible and tracking the dynamic environment around each hole as it flows through the program.
+These accumulated hole environments can be reported directly to the programmer by the editor as they work to fill holes.
+Hole environments also enable an ``edit-and-resume'' feature that avoids the need to restart evaluation after edits that amount to hole filling.
+Formally, the semantics combines features from gradual type theory (to handle incomplete types) and contextual modal type theory (CMTT, which provides a logical foundation for hole environment tracking) with novel structure necessary to prove key metatheoretic properties, including 1) a type safety property that accounts for expression holes (i.e.
+free metavariables, from the perspective of CMTT), and 2) commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound.
+The primary focus of this paper is on these formal properties, which we have mechanized using the Agda proof assistant.
+We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type, based on the edit action calculus in the prior work by \citet{popl-paper}.
+Taken together, the result is a live functional programming environment where feedback is truly continuous, i.e.
+it is available for every possible edit state.
+% i.e.
 \emph{every edit state is guaranteed to have non-trivial static and dynamic meaning}.

### 2018-02-08 hazelnut-dynamics.tex
incorporated feedback from @matthewhammer about abstract, and put linebreaks between sentences in the tex file

@@
 \begin{abstract}
-\emph{Live programming environments} aim to provide programmers (and, in many cases, audiences) with continuous feedback about a program's dynamic behavior throughout the editing process.
-The problem that motivates this paper is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas program editors frequently encounter edit states that are incomplete.
+\emph{Live programming environments} aim to provide programmers (and, in some cases, audiences) with continuous feedback about a program's dynamic behavior throughout the editing process. 
+The problem that motivates this paper is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas program editors frequently encounter incomplete edit states.
 Recent work by \citet{popl-paper} has made progress on the problem of working directly with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies.
-This paper develops a corresponding dynamic semantics.
+This paper develops a corresponding dynamic semantics for incomplete functional programs. 
 Rather than aborting with an exception when evaluation encounters a hole (as in  several existing systems), evaluation proceeds ``around'' expression holes, performing as much computation as is possible and tracking the dynamic environment around each hole as it flows through the program.
 These accumulated hole environments can be reported directly to the programmer by the editor as they work to fill holes. 
-Hole environments also enable an ``edit-and-resume'' feature that avoids the need to restart evaluation after edits that amount to hole filling.
-Formally, the semantics combines features from gradual type theory (to handle incomplete types) and contextual modal type theory (CMTT, which provides a logical foundation for hole environment tracking) with novel structure necessary to prove key metatheoretic properties, including 1) a type safety property that accounts for expression holes (i.e.
+They also enable an ``edit-and-resume'' feature that avoids the need to restart evaluation after edits that amount to hole filling. 
+Formally, our semantics combines features from gradual type theory (to handle incomplete types) and contextual modal type theory (CMTT, which provides a logical foundation for hole environment tracking) with novel structure necessary to prove key metatheoretic properties, including 1) a type safety property that accounts for expression holes (i.e.
 free metavariables, from the perspective of CMTT), and 2) commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound.
-The primary focus of this paper is on these formal properties, which we have mechanized using the Agda proof assistant.
+The primary focus of this paper is on this formal development, which we have mechanized using the Agda proof assistant. 
 We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type, based on the edit action calculus in the prior work by \citet{popl-paper}.
-Taken together, the result is a live functional programming environment where feedback is truly continuous, i.e.
+Taken together, the result is a proof-of-concept live functional programming environment where dynamic feedback is truly continuous, i.e.
 it is available for every possible edit state.
@@
 %This gap can cause live programming services to lag behind the programmer's edits, in many cases for an extended period of time.
-
-%
-
-%
-% However, live programming services typically cannot run incomplete
-% programs, or they raise an exception as soon as a hole is encountered
-% during evaluation.
-% %
-% %% When they % encounter an incomplete program, a common strategy is
-% %to simply % disable the live programming service, or revert to the
-% %last complete % edit state.
-% %
-% This deprives the programmer, and also other editor services, of
-% potentially valuable feedback about the dynamic behavior of the
-% visible program.
-Our second track of research will address this gap by defining a dynamic semantics for incomplete programs.
-We plan to focus the first year of research efforts largely on this track of research, as indicated in \autoref{fig:tracks}.
-% % Evaluation of incomplete programs will proceed as far as possible past the hole, and in the process, will track the environment around each hole instance.
-% This semantics will support live programming services like that outlined in our example above, which provides the programmer with a result on every, or nearly every, edit and allows the programmer to choose a hole instance in the computed result of evaluation and inspect its environment (in the sidebar of \autoref{fig:intro-example}) while they determine how to best fill that hole.
-% Tracking each hole instance's environment will also enhance the editor services specified by the remaining two tracks of research, \ie{}~the action suggestion panel (\trackThree{}) and the direct manipulation programming service (\trackFour{}).
-% % We will prove that in certain situations, hole filling commutes with evaluation.
-% % This formal observation will lead to a decrease in the cost of providing continuous feedback to the programmer on every edit by avoiding redundant recomputation.
-% Our preliminary work, described in a presentation at the \emph{LIVE 2017}
-% workshop~\cite{HazelLIVE}, has discovered encouraging connections between our proposed dynamics and previous research on contextual modal type theory (the Curry-Howard interpretation of contextual modal logic) \cite{Nanevski2008} and gradual type theory \cite{Siek06a}.
-
 \end{abstract}

### 2018-02-08 related-work.tex
incorporated feedback from @matthewhammer about abstract, and put linebreaks between sentences in the tex file
braindump outline of related work we should consider mentioning

@@
 
-\todo{related work}
+\begin{itemize}
+	\item Hazelnut paper
+	\item simply typed underdeterminism paper
+	\item work on partial evaluation and staging (incl. connections to modal logic -- ``modal analysis of staged computation'')
+	\item symbolic evaluation
+	\item full beta reduction
+	\item Agda
+	\item Idris
+	\item GHC holes
+	\item CMTT
+	\item gradual typing (and dynamic typing)
+		\begin{itemize}
+			\item siek and taha paper
+			\item snapl15 paper
+			\item gradualizer paper
+			\item maybe other things, e.g. several papers by ron garcia
+		\end{itemize}
+	\item DuctileJ stuff -- \url{https://homes.cs.washington.edu/~mernst/pubs/ductile-icse2011.pdf}
+	\item OLEG from McBride's thesis
+	\item Visual Studio (and others) support for edit-and-resume
+	\item Scratch lets you just skip over statement holes
+	\item prior work on confluence for the lambda calculus
+	\item work on debuggers that allow you to inspect environments
+		\begin{itemize}
+		\item might be something relevant in the paper ``A Debugger for Standard ML'' 
+		\item "Visualizing the evaluation of functional programs for debugging" by Whitington and Ridge
+		\item "A lightweight interactive debugger for Haskell'' and ``Multiple-View Tracing for Haskell: a New Hat'' might be relevant
+		\item ocamli -- \url{https://github.com/johnwhitington/ocamli}
+		\item Better supporting debugging aids learning a novel programming language. -- Scaffidi at VLHCC 2017
+		\item quote from Wadler in ``Why no one uses functional languages'':
+			\begin{quote}
+			“...there are few debuggers or
+profilers for strict [functional] languages, perhaps because constructing them is not considered
+research. This is a shame, since such tools are sorely needed, and there remains much of
+interest to learn about their construction and use.
+\end{quote}
+		\end{itemize}
+	\item papers that show up in a search for ``typed holes'' -- \url{https://scholar.google.com/scholar?hl=en&as_sdt=0%2C39&q=%22typed+holes%22&btnG=}
+	\item maybe also search for ``partial programs'' 
+	\item roly pererra work on program slicing
+	\item should say something about how holes show up in program synthesis under various names (what?) 
+	\item ``Achieving flexibility in direct-manipulation programming environments by relaxing the edit-time grammar'' -- \url{http://ieeexplore.ieee.org/document/1509511/}
+	\item ``Call-by-value is dual to call-by-name'' might be relevant? \url{http://homepages.inf.ed.ac.uk/wadler/papers/dual/dual.pdf}
+	\item mention how unspecified evaluation order is something that people do when talking about parallelism?
+\end{itemize}

### 2018-02-08 intro.tex
draft of an abstract (probably a bit too long and meandering/unclear, but its a start)
some thoughts to add to intro, somewhere

@@
 \newcommand{\introSec}{Introduction}
-\section{\protect\introSec}
+\section{\protect\introSec} % don't like the all-caps thing that the template does, so protecting it from that
 \label{sec:intro}
@@
 The purpose of this paper is to sketch out our progress toward a theoretically well-grounded solution.
+
+\matt{below are misc points that may be worth mentioning in section 1
+  (somehow, somewhere), and weaving into the continuation of the story
+  above:}
+
+Type holes vs expression holes:
+
+The dynamics of type holes is gradual typing, including run-time cast
+insertion and execution.
+%
+The dynamics of expression holes consists of defining
+\emph{indeterminate} expressions, as placeholders for subterms that
+traditionally would be deemed ``stuck'' .
+
+In live programming, we expect that programs will generally have holes
+in both terms and types (CITE?); however, to understand the mechanisms
+of \HazelnutLive, it is instructive to imagine situations where the
+only holes or in terms, or in types.
+%
+  When we only use type holes (not common in live programming, AFAIK),
+  we recover gradual typing and do not need indeterminate forms
+  (except for cast errors, which we classify this way).  
+  % 
+  When we only use expression holes, we do not need type casts, or
+  the other dynamic mechanisms of gradual typing.
 

### 2018-02-08 intro.tex
minor

@@
 of \HazelnutLive, it is instructive to imagine situations where the
-only holes or in terms, or in types.
+only holes are in terms, or are in types.
 %
-  When we only use type holes (not common in live programming, AFAIK),
-  we recover gradual typing and do not need indeterminate forms
-  (except for cast errors, which we classify this way).  
+When we only use expression holes and no type holes, we do not need
+type casts, or the other dynamic mechanisms of gradual typing.
 % 
-  When we only use expression holes, we do not need type casts, or
-  the other dynamic mechanisms of gradual typing.
+When we only use type holes, we recover gradual typing and do not need
+indeterminate forms (except for cast errors, which we classify this
+way).
 

### 2018-02-08 intro.tex
cast errors

@@
 % 
-When we only use type holes, we recover gradual typing and do not need
-indeterminate forms (except for cast errors, which we classify this
-way).
+When we only use type holes, we recover gradual typing and do not use
+indeterminate forms until and unless cast errors arise.
+
+Cast errors as indeterminate forms:
+
+Instead of being ``stuck'', cast errors in \HazelnutLive are
+fully-reduced indeterminate terms.
 

### 2018-02-08 intro.tex
braindump in intro

@@
 
-Type holes vs expression holes:
+\textbf{Type holes vs expression holes:}
 
@@
 
-Cast errors as indeterminate forms:
+\textbf{Cast errors as indeterminate forms:}
 
@@
 fully-reduced indeterminate terms.
+
+\textbf{Indeterminate forms as exceptions:}
+
+TODO --- transfer thoughts from 2016 grant writing
+
+\textbf{Holes as breakpoints:}
+
+\begin{enumerate}
+
+\item 
+  \texttt{map somefun list}
+
+Suppose the programmer wishes to map with \texttt{somefun} that they
+either did not author themselves, or they forgot how the function
+works:
+
+\item 
+  \textt{map ($\lambda \texttt{x}. \hhole{\texttt{x}}{a}$) list}
+
+First, the programmer can intercept the function's inputs.  These
+hole~$a$ above functions like a ``breakpoint'', and shows the
+programmer me all of the inputs to \texttt{somefun}.
+
+TODO -- show the result of running above.
+
+\item
+  \textt{map ($\lambda \texttt{x}. \hhole{\texttt{somefun x}}{a}$) list}
+
+  Next, suppose the programmer wishes to see all of the the outputs of
+  \texttt{somefun}.  To do so, they put that (single static)
+  expression into a hole, as above.
+
+\item
+  \textt{map ($\lambda \texttt{x}. \hhole{(x, \texttt{somefun x})}{a}$) list}
+
+  Finally, the programmer can see input-output pairs without breaking
+  the typing derivation of their program globally, since holes are
+  permitted to go \emph{anywhere}.  (They may have also exploited this
+  fact to show the inputs, in the first step).
+
+\end{enumerate}
 

### 2018-02-08 abstract.tex
separate abstract.tex file

--- 2018-02-08 hazelnut-dynamics.tex
@@
-\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
-%% For double-blind review submission
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission
-%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission
-%\documentclass[acmlarge]{acmart}\settopmatter{}
-
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format should change 'acmlarge' to
-%% 'sigplan,10pt'.
-
-\bibliographystyle{ACM-Reference-Format}
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-
-%% Cyrus packages
-\usepackage{microtype}
-\usepackage{mdframed}
-\usepackage{colortab}
-\usepackage{mathpartir}
-\usepackage{enumitem}
-\usepackage{bbm}
-\usepackage{stmaryrd}
-\usepackage{mathtools}
-\usepackage{leftidx}
-\usepackage{todonotes}
-
-\usepackage{listings}%
-\lstloadlanguages{ML}
-\lstset{tabsize=2, 
-basicstyle=\footnotesize\ttfamily, 
-% keywordstyle=\sffamily,
-commentstyle=\itshape\ttfamily\color{gray}, 
-stringstyle=\ttfamily\color{purple},
-mathescape=false,escapechar=\#,
-numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
-morekeywords={string, float, int},
-classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount
-}
-\let\li\lstinline
-
-%% Joshua Dunfield macros
-\def\OPTIONConf{1}%
-\usepackage{joshuadunfield}
-
-% \newtheorem{theorem}{Theorem}[chapter]
-% \newtheorem{lemma}[theorem]{Lemma}
-% \newtheorem{corollary}[theorem]{Corollary}
-% \newtheorem{definition}[theorem]{Definition}
-% \newtheorem{assumption}[theorem]{Assumption}
-% \newtheorem{condition}[theorem]{Condition}
-
-%% Listings
-\makeatletter\if@ACM@journal\makeatother
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2018}
-\acmMonth{1}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2018}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission
-\setcopyright{none}             %% For review submission
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2017}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\input{commands}
-\input{macros}
-
-\begin{document}
-
-%% Title information
-\title{Live Functional Programming with Typed Holes}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-% \titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-% \subtitle{Subtitle}                     %% \subtitle is optional
-% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-
-%% Author with single affiliation.
-\author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}
-}
-\email{first1.last1@inst1.edu}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}
-}
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
-
-%% Paper note
-%% The \thanks command may be used to create a "paper note" ---
-%% similar to a title note or an author note, but not explicitly
-%% associated with a particular element.  It will appear immediately
-%% above the permission/copyright statement.
-% \thanks{with paper note}                %% \thanks is optional
-                                        %% can be repeated if necesary
-                                        %% contents suppressed with 'anonymous'
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
 \begin{abstract}
@@
 Recent work by \citet{popl-paper} has made progress on the problem of working directly with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies.
-This paper develops a corresponding dynamic semantics for incomplete functional programs. 
+This paper develops a corresponding dynamic semantics for this class of incomplete functional programs. 
 Rather than aborting with an exception when evaluation encounters a hole (as in  several existing systems), evaluation proceeds ``around'' expression holes, performing as much computation as is possible and tracking the dynamic environment around each hole as it flows through the program.
@@
 \end{abstract}
-
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-% \begin{CCSXML}
-% <ccs2012>
-% <concept>
-% <concept_id>10011007.10011006.10011008</concept_id>
-% <concept_desc>Software and its engineering~General programming languages</concept_desc>
-% <concept_significance>500</concept_significance>
-% </concept>
-% <concept>
-% <concept_id>10003456.10003457.10003521.10003525</concept_id>
-% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
-% <concept_significance>300</concept_significance>
-% </concept>
-% </ccs2012>
-% \end{CCSXML}
-
-% \ccsdesc[500]{Software and its engineering~General programming languages}
-% \ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-% \keywords{keyword1, keyword2, keyword3}  %% \keywords is optional
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
-\maketitle
-
-\input{intro}
-\input{examples}
-\input{calculus}
-\input{related-work}
-\input{discussion}
-
-\bibliography{references,all.short,hazel_NSF}
-
-\clearpage
-\appendix
-\cy{for the submission, appendix should be separated into a separate supplemental PDF}
-
-\input{extensions}
-
-\end{document}

### 2018-02-08 hazelnut-dynamics.tex
removed possibly redundant bibliographystyle command per @ivoysey's comment
braindump outline of related work we should consider mentioning
some thoughts to add to intro, somewhere
separate abstract.tex file

@@
 
-\bibliographystyle{ACM-Reference-Format}
+% \bibliographystyle{ACM-Reference-Format}
 
@@
 \usepackage{todonotes}
+\usepackage{xspace}
 
@@
 \usepackage{joshuadunfield}
+
+% A macro for the name of the system being described by ``this paper''
+\newcommand{\HazelnutLive}{\textsf{Hazelnut Live}\xspace}
 
@@
 %% before \maketitle command
-\begin{abstract}
-\emph{Live programming environments} aim to provide programmers (and, in some cases, audiences) with continuous feedback about a program's dynamic behavior throughout the editing process. 
-The problem that motivates this paper is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas program editors frequently encounter incomplete edit states.
-Recent work by \citet{popl-paper} has made progress on the problem of working directly with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies.
-This paper develops a corresponding dynamic semantics for incomplete functional programs. 
-Rather than aborting with an exception when evaluation encounters a hole (as in  several existing systems), evaluation proceeds ``around'' expression holes, performing as much computation as is possible and tracking the dynamic environment around each hole as it flows through the program.
-These accumulated hole environments can be reported directly to the programmer by the editor as they work to fill holes. 
-They also enable an ``edit-and-resume'' feature that avoids the need to restart evaluation after edits that amount to hole filling. 
-Formally, our semantics combines features from gradual type theory (to handle incomplete types) and contextual modal type theory (CMTT, which provides a logical foundation for hole environment tracking) with novel structure necessary to prove key metatheoretic properties, including 1) a type safety property that accounts for expression holes (i.e.
-free metavariables, from the perspective of CMTT), and 2) commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound.
-The primary focus of this paper is on this formal development, which we have mechanized using the Agda proof assistant. 
-We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type, based on the edit action calculus in the prior work by \citet{popl-paper}.
-Taken together, the result is a proof-of-concept live functional programming environment where dynamic feedback is truly continuous, i.e.
-it is available for every possible edit state.
-% i.e.
-\emph{every edit state is guaranteed to have non-trivial static and dynamic meaning}.
-
-%In our system, evaluation treats failed casts much like it treats expression holes (rather than immediately failing with a cast error, as in gradual type theory).
-Prior work on contextual modal type theory did not develop an operational semantics for programs with free metavariables, which correspond to programs with holes in our formulation.
-
-% These incomplete edit states are sometimes transient, but at other times, they persist for extended periods of time, e.g.
-when the programmer is filling in the branches of a large case analysis one-by-one, or when an edit causes type errors to appear throughout a program.
-%This gap can cause live programming services to lag behind the programmer's edits, in many cases for an extended period of time.
-\end{abstract}
+\input{abstract}
 

### 2018-02-09 abstract.tex
wrapping and separating lines; no changes to text
edits to abstract

@@
 \begin{abstract}
-\emph{Live programming environments} aim to provide programmers (and, in some cases, audiences) with continuous feedback about a program's dynamic behavior throughout the editing process. 
-The problem that motivates this paper is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas program editors frequently encounter incomplete edit states.
-Recent work by \citet{popl-paper} has made progress on the problem of working directly with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies.
-This paper develops a corresponding dynamic semantics for this class of incomplete functional programs. 
-Rather than aborting with an exception when evaluation encounters a hole (as in  several existing systems), evaluation proceeds ``around'' expression holes, performing as much computation as is possible and tracking the dynamic environment around each hole as it flows through the program.
-These accumulated hole environments can be reported directly to the programmer by the editor as they work to fill holes. 
-They also enable an ``edit-and-resume'' feature that avoids the need to restart evaluation after edits that amount to hole filling. 
-Formally, our semantics combines features from gradual type theory (to handle incomplete types) and contextual modal type theory (CMTT, which provides a logical foundation for hole environment tracking) with novel structure necessary to prove key metatheoretic properties, including 1) a type safety property that accounts for expression holes (i.e.
-free metavariables, from the perspective of CMTT), and 2) commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound.
-The primary focus of this paper is on this formal development, which we have mechanized using the Agda proof assistant. 
-We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type, based on the edit action calculus in the prior work by \citet{popl-paper}.
-Taken together, the result is a proof-of-concept live functional programming environment where dynamic feedback is truly continuous, i.e.
-it is available for every possible edit state.
-% i.e.
-\emph{every edit state is guaranteed to have non-trivial static and dynamic meaning}.
+
+\emph{Live programming environments} aim to provide programmers (and, in some
+cases, audiences) with continuous feedback about a program's dynamic behavior
+throughout the editing process.
+%
+Unfortunately, because programming languages typically assign dynamic meaning
+only to programs that are \emph{complete}---syntactically well-formed and free
+of static type errors---live programming features are often severely restricted,
+if available at all, when programs are incomplete.
+%
+Incomplete edit states are frequently encountered when writing programs, so they
+are, arguably, when live programming services would be most valuable.
+%
+Defining a semantics for incomplete functional programs would help bridge the
+gap between well-formed states and allowing live programming environments to
+provide continuous, uninterrupted, and meaningful feedback.
+
+Recent work by \citet{popl-paper} makes progress toward this goal by developing
+a type system for lambda terms with \emph{holes} in both expression and type
+position, with empty holes standing for missing expressions or types, and
+non-empty holes operating as ``membranes'' around type inconsistencies.
+%
+The contribution of this paper is a corresponding \emph{dynamic semantics for
+functional programs with holes}.
+%
+Rather than aborting with an exception when evaluation encounters a hole (as in
+several existing systems), evaluation proceeds ``around'' expression holes,
+performing as much computation as is possible and tracking the dynamic
+environment around each hole as it flows through the program.  These accumulated
+hole environments can be reported directly to the programmer by the editor as
+they work to fill holes.  They also enable an ``edit-and-resume'' feature that
+avoids the need to restart evaluation after edits that amount to hole filling.
+
+Formally, our semantics combines features from gradual type theory (to handle
+incomplete types) and contextual modal type theory (CMTT, which provides a
+logical foundation for hole environment tracking) with novel structure necessary
+to prove key metatheoretic properties, including (1) a type safety property that
+accounts for expression holes (i.e. free metavariables, from the perspective of
+CMTT), and (2) commutativity and confluence properties that guarantee that the
+``edit-and-resume'' feature is sound.
+%
+We have mechanized our formal development using the Agda proof assistant.
+%
+We additionally describe a simple reference implementation, called Hazelnut
+Live, that exposes a language of structured edit actions that insert holes as
+necessary to guarantee that every edit state has some (possibly incomplete)
+type, based on the edit action calculus in the prior work by \citet{popl-paper}.
+%
+Taken together, the result is a proof-of-concept live functional programming
+environment where dynamic feedback is truly continuous, i.e. it is available for
+every possible edit state. % i.e. \emph{every edit state is guaranteed to have
+non-trivial static and dynamic meaning}.
 

### 2018-02-09 abstract.tex
linebreak issue
rough notes for examples section

@@
 Defining a semantics for incomplete functional programs would help bridge the
-gap between well-formed states and allowing live programming environments to
+gap between well-formed states and allow live programming environments to
 provide continuous, uninterrupted, and meaningful feedback.
@@
 environment where dynamic feedback is truly continuous, i.e. it is available for
-every possible edit state. % i.e. \emph{every edit state is guaranteed to have
-non-trivial static and dynamic meaning}.
+every possible edit state.
+% i.e. \emph{every edit state is guaranteed to have
+% non-trivial static and dynamic meaning}.
 

### 2018-02-09 examples.tex
rough notes for examples section

@@
 
-\todo{examples}
+We introduce \HazelnutLive{} in this section using several examples,
+demonstrating how running incomplete programs---with missing expressions and/or
+type-inconsistent expressions---can be useful.
+%
+We will also demonstrate how running even complete programs marked with holes is
+useful as a form of debugging.
+
+\parahead{Live Programming with Incomplete Expressions}
+
+\emph{Example 1: Computing Weighted Averages.}
+%
+type for grades.
+%
+all student data entered in a list of records.
+
+\begin{lstlisting}
+type student_rec =
+  { name: string,
+    hw1: float, hw2: float, hw3: float,
+    midterm: float, final: float };
+
+let grades : list(student_rec) =
+  [ { name: "Alice", hw1: 76.0, ... }
+  , { name: "Bob", hw1: 88.0, ... }
+  , ...
+  ];
+
+let weighted_average(g: student_rec) =
+  ??;
+
+let weighted_averages = List.map weighted_average grades;
+\end{lstlisting}
+
+end of the term, now time to assign grades.
+%
+weighted average of scores.
+%
+start by weighting hw1 to be 10\%. the rest of the expression is a hole, but can
+run the expression, and see that the weighted averages are too large
+%
+
+\begin{lstlisting}
+let weighted_average(g: student_rec) =
+  (10.0 *. g.hw1) + . ??;
+\end{lstlisting}
+
+oops, should be using numbers in the range 0 to 1 rather than 1 to 100 for
+percentage points.
+%
+by seeing the output as the expression is being written, this error is caught
+early and avoids the same mistake being made several times.
+
+finish the weights, and get the weighted averages.
+
+\emph{Example 2: Assigning Letter Grades.}
+%
+next, the teacher must map weighted averages to letter grades.
+%
+using the following cutoffs type.
+
+\begin{lstlisting}
+type grade_cutoffs =
+  { a: float, b: float, c: float, d: float, f: float };
+
+let cutoffs =
+  { a = ??, b = ??, c = ??, d = ??, f = ?? };
+\end{lstlisting}
+
+initially all holes, because it will be different year-to-year based on the
+data, differences in course difficulty, and to satisfy fairness criteria.
+%
+write some code that does the mapping.
+
+\begin{lstlisting}
+let letter_grade(n: weighted_average) =
+  if n >= cutoffs.a then "A" else
+  if n >= cutoffs.b then "B" else
+  if n >= cutoffs.c then "C" else
+  if n >= cutoffs.d then "D" else
+  if n >= cutoffs.f then "F" else
+  "Incomplete";
+
+let sorted_weighted_averages = List.sort weighted_averages;
+
+let letter_grades = List.map letter_grade sorted_weighted_averages;
+\end{lstlisting}
+
+since each constant in the multi-way if is a hole, none of the weighted averages
+resolve to a letter.
+%
+to start figuring out the constants, first sort the weighted averages.
+%
+by looking at the data, there is a natural gap between 92 and 89.
+%
+so, choose to to use 92 as the cutoff.
+%
+see XXX A's pop out in the lettergrades list.
+%
+before assigning other cutoffs, want to get a sense if whether this is even a
+good choice for A.
+%
+jump to writing a distribution function.
+
+\begin{lstlisting}
+let distribution(list: list(float)) =
+  let n = List.length letter_grades in
+  List.map
+    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list) /. n)))
+    ["A","B","C","D","F","Incomplete"];
+\end{lstlisting}
+
+this shows that the percentage of As is XXX\%, which seems too small.
+%
+look back at the sorted data, and see a cluster around 89, and then another gap
+between 88 and 85.
+%
+so, tweak A to be 88.
+%
+the percentage is updated to XXX\%.
+
+now, continue down the list of sorted averages, picking the next cutoff.
+%
+this series of edits fills one hole, then another, then another.
+%
+so after each, the suspended computation from before can simply be resumed.
+%
+in cases where a previous cutoff is modified (which is no longer marked by any
+hole), our semantics provides no caching.
+%
+however, typical incremental computation can be incorporated to determine which
+situations such changes lead to partial reuse of previous computation.
+%
+we do not model this in our core calculus, since our focus is incremental
+computation for the specific new mechanism we study, namely, evaluating around
+expression holes.
+
+throughout this example, we described a workflow where the programmers goes back
+and forth between different parts of the program.
+%
+this is only possible to do smoothly because of our treatment of expression
+holes.
+%
+treating any one of them as an exception, as is the common ad-hoc approach to
+expression holes, prevents the rest of the useful computation from proceeding
+and being reused when that hole is filled.
+
+\parahead{Live Programming with Type Errors}
+
+Understanding and debugging static type errors is notoriously difficult,
+particularly for novices.
+%
+A variety of approaches have been
+proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
+and explain type errors.
+%
+One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
+that demonstrates a run-time failure, and then displays a compressed execution
+trace to the user as a graph.
+%
+By running type-incorrect programs as far as possible (in particular, until they
+would go wrong), our semantics can provide a similar benefit operationally.
+%
+\rkc{they define a partial evaluation with a form of holes. compare.}
+
+\emph{Example 3: Sum List.}
+%
+Consider the following buggy program of numbers reported by \citet{XXX} that
+attempts to sum a list numbers.
+%
+The problem is that the base case produces a list rather than a number (in
+particular, 0).
+
+\begin{lstlisting}
+sumList : list(int) -> int
+sumList [] = ?[]?
+sumList (n:ns) = n + sumList ns
+\end{lstlisting}
+
+\noindent
+%
+Because the list expression on line 2 does not have time \li{int} as required,
+it is wrapped in a (non-empty) hole.
+%
+Rather than trying to debug the error based on the static error, the programmer
+may wish to run a call to the function, say \li{sumList(2)}, in which case
+\HazelnutLive{} runs and produces the indeterminate value \li{1 + 2 + ?[]?}
+(\cf{} the trace displayed by \citet{Seidel2016}).
+%
+By observing that the hole expression is being added to the number \li{2}, he
+realizes that it needs to be a number, specifically, \li{0}.
+
+%% TODO fold error from Erwig paper.
+%% %
+%% see that final call on stack does have the right answer, but
+%% it's wrapped in a singleton list when the expected type is not
+%% a list.
+%% %
+%% fix is to remove the list, the rest of the computation remains
+%% the same, but b/c they were all wrapped in holes, need to re-run.
+%% %
+%% (add some mechanism for type-consistent non-empty-holes...)
+
+\parasection{Example 4: Stutter}
+%
+As another example, consider the following function which attempts to produce a
+list where every element is repeated twice (borrowed from \citep{XXX}).
+%
+The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
+it produces a \li{list(list(int)} instead.
+
+\begin{lstlisting}
+stutter : list(int) -> list(int)
+stutter xs = List.foldr (\x acc -> ?([x,x] : acc)?) [] xs
+\end{lstlisting}
+
+\noindent
+%
+The editor has a choice about which expression to ``blame'' for the error; the
+entire application that forms the body of the lambda is analyzed against the
+return type \li{list(int)}, so that is a reasonable choice for the editor to
+make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
+both as intended and that only the function \li{(:)} is type-consistent.
+%
+Although one could imagine a setting in which a user would perform this
+reasoning, let's assume the simplest approach for marking consistencies that
+wraps the entire application.
+%
+Running this on \li{stutter [1,2,3]} produces the indeterminate value
+\li{?  [1,1] : (? [2,2] : (? [3:3] : [] ?) ?) ?}, which shows the unfolding of
+\li{List.foldr}.
+%
+Inside the innermost hole, the value is shown to be \li{[[3,3]]}; the user
+realizes that there are too many levels of nesting.
+%
+He tries append in place of cons, which addresses the type inconsistency and
+produces the desired result.
+
+\parahead{Live Programming for Debugging}
+
+\rkc{not sure about the quicksort example yet}
+
+%% \begin{lstlisting}
+%% quicksort : list('a) -> list('a)
+%% quicksort [] = []
+%% quicksort (x:xs) =
+%%   let (left, right) = partition ((<) x) xs in
+%%   let (left', right') = (quicksort left, quicksort right) in
+%%   left' @ [x] @ right
+%% \end{lstlisting}

### 2018-02-20 abstract.tex
Section 2, wip

@@
 %
-We additionally describe a simple reference implementation, called Hazelnut
-Live, that exposes a language of structured edit actions that insert holes as
+We additionally describe a simple reference implementation, called \HazelnutLive{},
+that exposes a language of structured edit actions that insert holes as
 necessary to guarantee that every edit state has some (possibly incomplete)

### 2018-02-20 examples.tex
Section 2, wip

--- 2018-02-07 examples.tex
@@
 % !TEX root = hazelnut-dynamics.tex
-\newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
+%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
+%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
+\newcommand{\examplesSec}{\HazelnutLive, By Example}
 \section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
@@
 
-\todo{examples}
+\newcommand{\overviewExample}[2]{\paragraph{Example {#1}: {#2}}}
+
+We introduce \HazelnutLive{} in this section using several examples,
+demonstrating how running incomplete programs---with missing expressions and/or
+type-inconsistent expressions---can be useful.
+%
+We will also demonstrate how running even complete programs with subexpressions
+marked as holes may be useful in debugging.
+
+\HazelnutLive{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
+alternative Reason syntax \cite{reason-what} extended to support holes.
+%
+As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
+literate program \cite{knuth1984literate}: rich text comments appear at the top
+level and code appears within cells embedded into these comments.
+%
+The reader is encouraged to work through these examples in our prototype
+implementation or watch video demonstrations, both available in
+\suppMaterials{}.\footnote{
+%
+The programs in this section are written using minor syntactic conveniences that
+are not currently supported in our implementation.
+%
+For example, the implementation of recursive types in our current implementation
+requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
+inferred in standard ways (by pairing them with constructors and deconstructors
+for algebraic datatypes) in future work.
+%
+More details about other minor syntactic differences are described in
+\suppMaterials{}.
+%
+}
+
+\subsection{Live Programming with Incomplete Expressions}
+
+The process of writing new program fragments includes many states in which the
+program under construction is, by definition, incomplete.
+%
+Furthermore, it is natural to ``jump back and forth'' between different places
+in the program, iteratively and alternately developing the missing pieces.
+%
+The primary benefit of our approach is that such programs can be evaluated in
+order to provide useful feedback during this editing workflow.
+
+\input{fig-grades-example}
+
+Consider a teacher who is in the midst of developing a \HazelnutLive{} notebook
+(depicted in \autoref{fig:grades-example}) to compute final student grades at
+the end of a term.
+%
+In the first cell in \autoref{fig:grades-example}, the teacher defines a record
+type, \li{student_data}, for each student's course data.
+%
+In our simplified example, this includes the student's name, of type
+\li{string}, and five grades, each of type \li{float}.
+
+\overviewExample{1}{Computing Weighted Averages}
+%
+The teacher begins writing a function called \li{weighted_average} that will
+compute a weighted average, of type \li{float}, for each \li{student_data}
+record.
+%
+The body of \li{weighted_average} is not yet written---marked with the
+\emph{empty hole} on line \rkc{XXX}--- but the teacher jumps ahead to map the
+function over the \li{grades} list.
+
+\begin{lstlisting}
+let weighted_average(g: student_data) =
+  ??;
+
+let weighted_averages = List.map weighted_average grades;
+\end{lstlisting}
+
+\noindent
+%
+\HazelnutLive{} evaluates this incomplete program, showing its
+\emph{indeterminate} result in the sidebar on the right half of
+\autoref{fig:grades-example}; this value is a list whose length is the same as
+\li{grades}, where each element is the (indeterminate) result of evaluating the
+hole expression inside \li{weighted_average}.
+%
+Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
+(\ie{}~\li{a}) and values it produces are assigned corresponding unique dynamic
+identifiers (\ie{}~\li{a.1}, \li{a.2}, etc{}).
+%
+In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
+using placeholder expressions like \li{raise "Not yet implemented."}, would halt
+the program as soon it is evaluated (in this case, inside the definition of
+\li{List.map} when the function is applied to the first element of \li{grades}.
+%
+By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
+least see that the resulting expression does indeed evaluate to a list with the
+correct number of (indeterminate) values.
+
+Now the teacher returns to \li{weighted_average} to work on the missing
+expression, with the goal of computing a weighted sum of each of a student's
+grades.
+%
+The teacher builds the summand weighting the \li{hw1} score to account for 10\%
+of the weighted average; the rest of the computation is a new hole expression
+(\li{??_b}), yet to be filled in.
+
+\begin{lstlisting}
+let weighted_average(g: student_data) =
+  (10.0 *. g.hw1) +. ??_b;
+\end{lstlisting}
+
+\noindent
+%
+\HazelnutLive{} immediately runs this incomplete program, displaying the updated
+list of indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
+%
+Right away, the teacher recognizes that the values are too large; they should be
+at most \li{100.0}.
+%
+The teacher realizes that representing percentage points as \li{float}s requires
+that the constant on line \rkc{XXX} ought to be \li{0.10} instead.
+%
+Because of this live feedback, the teacher corrects this error right away and
+avoids making similar programming errors in the rest of the calculation.
+%
+The teacher continues to build the rest of the arithmetic expression until it is
+complete (there are no longer any expression holes), and the result of
+immediately running the finished program shows that each of values in the final
+result list is in the range \li{0.0} to \li{100.0}.
+
+\overviewExample{2}{Assigning Letter Grades}
+%
+The teacher's next task is to map the weighted averages to the letter grades A
+through F (we will consider only ``whole'' letter grades, for simplicity).
+%
+The \li{grade_cutoffs} record type describes the minimum cutoff for each of
+these six possible.
+%
+Initially, each of values in the \li{cutoffs} is a hole.
+
+\begin{lstlisting}
+type grade_cutoffs =
+  { a: float, b: float, c: float, d: float, f: float };
+
+let cutoffs =
+  { a = ??_a, b = ??_b, c = ??_c, d = ??_d, f = ??_f };
+\end{lstlisting}
+
+\noindent
+%
+%% initially all holes, because it will be different year-to-year based on the %
+%data, differences in course difficulty, and to satisfy fairness criteria.
+%
+Before starting to fill in the cutoffs, the teacher jumps ahead to write a
+function \li{letter_grade} that will make the connection between \li{cutoffs}
+and \li{weighted_averages}.
+%
+The teacher sorts the \li{weighted_averages} because she intends to look at the
+data to help select the cutoff values, and then maps them to \li{letter_grades}.
+
+\begin{lstlisting}
+let letter_grade(n: weighted_average) =
+  if n >= cutoffs.a then "A" else
+  if n >= cutoffs.b then "B" else
+  if n >= cutoffs.c then "C" else
+  if n >= cutoffs.d then "D" else
+  if n >= cutoffs.f then "F" else "Incomplete";
+
+let sorted_weighted_averages = List.sort weighted_averages;
+
+let letter_grades = List.map letter_grade sorted_weighted_averages;
+\end{lstlisting}
+
+\noindent
+%
+When \HazelnutLive{} runs this program, the guard of the outermost conditional,
+\li{n >= cutoffs.a}, is indeterminate because \li{cutoffs.a} is.
+%
+Therefore, each of the indeterminate expressions in \li{letter_grades} is the
+entire expression body, albeit with different bindings for \li{n}.
+%
+Displaying such ``large'' indeterminate expressions can quickly consume all
+available screen space, overwhelming the user with too much information.
+%
+\autoref{fig:XXX} shows how \HazelnutLive{} renders large indeterminate
+expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
+or hole) or list of small values) simply as \li{..} to save space.
+%
+Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
+indeterminate expression as a tooltip.
+%
+\autoref{sec:discussion} discusses this and other user interface concerns when
+trying to display useful live feedback without overwhelming the user.
+%
+These usability factors are beyond the scope of our work, which is to define
+semantic foundations on which such user interfaces can be built.
+
+To start deciding \li{cutoffs}, the teacher looks at the \li{weighted_averages}
+sorted in descending order.
+%
+The data shows a natural gap between \li{92.2} and \li{89.5}.
+%
+So, she chooses to use \li{92.0} as the cutoff for A.
+%
+Resuming the computation from before, \HazelnutLive{} resolves the conditional
+expressions for the first \rkc{XXX} indeterminate expressions, because each of
+those \li{n} values was greater than \li{cutoffs.a}.
+%
+The remaining \rkc{XXX} expressions also proceeded to evaluate the first guard,
+and are now indeterminate at the guard for the second conditional.
+
+Before assigning other cutoff values, the teacher would like to get a sense of
+whether this first choice is a good one.
+%
+She jumps ahead to to write a function that computes the distribution of letter
+grades.
+
+\begin{lstlisting}
+let compute_distribution(list: list(float)) =
+  let n = List.length letter_grades in
+  List.map
+    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list)) /. n))
+    ["A","B","C","D","F","Incomplete"];
+
+let distribution = compute_distribution(letter_grades);
+\end{lstlisting}
+
+\noindent
+%
+Running \li{compute_distribution} shows that the percentage of As is
+\rkc{XXX\%}, which is smaller than what the teacher would like.
+%
+Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
+and then another gap between \li{88.2} and {85.5}.
+%
+So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
+%
+Because this edit is outside of a hole expression, \HazelnutLive{} discards the
+previous execution state and reevaluates the entire program.
+%
+Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
+be applied to seek opportunities for reuse even when non-hole expressions are
+modified.
+%
+Because the focus of work is on the novel implications of running programs with
+holes, our calculus and implementation supports caching ``edit-and-resume'' only
+for the novel situation in which evaluation proceeds around holes.
+%
+After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
+matches the teacher's intention.
+
+In this fashion, the teacher continues down the list of sorted averages,
+determining appropriate values for each cutoff.
+%
+Whenever the teacher is only filling in the ``next'' cutoff, the computation
+from before can simply be resumed.
+%
+Overall, throughout the workflow described in these two examples, the programmer
+can continue to evaluate the program, and receive meaningful feedback, while
+going back and forth between different pieces under development.
+
+\rkc{STOPPING HERE FOR NOW}
+
+%% \subsection{Live Programming with Type Errors}
+\subsection{Live Programming with Static Type Errors}
+
+Understanding and debugging static type errors is notoriously difficult,
+particularly for novices.
+%
+A variety of approaches have been
+proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
+and explain type errors.
+%
+One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
+that demonstrates a run-time failure, and then displays a compressed execution
+trace to the user as a graph.
+%
+By running type-incorrect programs as far as possible (in particular, until they
+would go wrong), our semantics can provide a similar benefit operationally.
+%
+\rkc{they define a partial evaluation with a form of holes. compare.}
+
+\emph{Example 3: Sum List.}
+%
+Consider the following buggy program of numbers reported by \citet{XXX} that
+attempts to sum a list numbers.
+%
+The problem is that the base case produces a list rather than a number (in
+particular, 0).
+
+\begin{lstlisting}
+sumList : list(int) -> int
+sumList [] = ?[]?
+sumList (n:ns) = n + sumList ns
+\end{lstlisting}
+
+\noindent
+%
+Because the list expression on line 2 does not have type \li{int} as required,
+it is wrapped in a (non-empty) hole.
+%
+Rather than trying to debug the error based on the static error, the programmer
+may wish to run a call to the function, say \li{sumList(2)}, in which case
+\HazelnutLive{} runs and produces the indeterminate value \li{1 + 2 + ?[]?}
+(\cf{} the trace displayed by \citet{Seidel2016}).
+%
+By observing that the hole expression is being added to the number \li{2}, he
+realizes that it needs to be a number, specifically, \li{0}.
+
+%% TODO fold error from Erwig paper.
+%% %
+%% see that final call on stack does have the right answer, but
+%% it's wrapped in a singleton list when the expected type is not
+%% a list.
+%% %
+%% fix is to remove the list, the rest of the computation remains
+%% the same, but b/c they were all wrapped in holes, need to re-run.
+%% %
+%% (add some mechanism for type-consistent non-empty-holes...)
+
+\parasection{Example 4: Stutter}
+%
+As another example, consider the following function which attempts to produce a
+list where every element is repeated twice (borrowed from \citep{XXX}).
+%
+The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
+it produces a \li{list(list(int)} instead.
+
+\begin{lstlisting}
+stutter : list(int) -> list(int)
+stutter xs = List.foldr (\x acc -> ?[x,x]? : acc) [] xs
+\end{lstlisting}
+
+\noindent
+%
+The editor has a choice about which expression to ``blame'' for the error; the
+entire application that forms the body of the lambda is analyzed against the
+return type \li{list(int)}, so that is a reasonable choice for the editor to
+make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
+both as intended and that only the function \li{(:)} is type-consistent.
+%
+Although one could imagine a setting in which a user would perform this
+reasoning, let's assume the simplest approach for marking consistencies that
+wraps the entire application.
+%
+Running this on \li{stutter [1,2,3]} produces the indeterminate value
+\li{?  [1,1] : (? [2,2] : (? [3:3] : [] ?) ?) ?}, which shows the unfolding of
+\li{List.foldr}.
+%
+Inside the innermost hole, the value is shown to be \li{[[3,3]]}; the user
+realizes that there are too many levels of nesting.
+%
+He tries append in place of cons, which addresses the type inconsistency and
+produces the desired result.
+
+\subsection{Live Programming with Gradual Type Errors}
+
+\parasection{Example 5: \rkc{...}}
+
+\rkc{gradual typing error, cast error, run around}
+
+\begin{lstlisting}
+foo : bool -> ? -> int
+foo b =
+  if b then
+    x + 10
+  else
+    x 10
+
+foo false 1 + 2 + 2
+\end{lstlisting}
+
+this demonstrates how, when no expression holes, our system is like gradual
+typing, but with the ability to run around cast errors.
+%
+just as it may be useful to report multiple type errors, this allows ability to
+report multiple dynamic errors, as well additional dynamic behavior around them.
+%
+could be incorporated into other gradual type systems
+
+\subsection{Live Programming for Debugging}
+
+\parasection{Example 6: \rkc{...}}
+
+\rkc{breakpoints, understanding input/output behavior}
+
+\parasection{Example 6: Quicksort}
+
+\begin{lstlisting}
+quicksort : list('a) -> list('a)
+quicksort [] = []
+quicksort (x:xs) =
+  let (left, right) = span ((>) x) xs in
+  ??
+\end{lstlisting}
+
+programmer has some intended relationship between names and values they bind.
+%
+live results show that left has larger elements and right has smaller.
+%
+right away, can see that need to change filter predicate. (alternatively, the
+names could have been swapped, but that order of names was the natural way
+to think about it.)
+%
+but next, still see that right has some smaller elements.
+%
+should pick partition instead of span.
+
+\begin{lstlisting}
+quicksort : list('a) -> list('a)
+quicksort [] = []
+quicksort (x:xs) =
+  let (left, right) = partition ((<) x) xs in
+  let (left', right') = (quicksort left, quicksort right) in
+  ? (left' @ [x] @ right) ?
+\end{lstlisting}
+
+now, put the lists back together.
+%
+doesn't do what is expected.
+%
+live feedback shows right' is computed correctly, but ah, it's not being used
+in the return list.
+%
+fix, and re-run, keeping the final expression in a hole.
+%
+can visualize that the nested appends are being built up correctly, and that the
+result would be the correct list.
+%
+finalize the hole and resume. everything but the nested appends can be reused.
+%
+\emph{hole environment traces} or \emph{hole environment traces}

### 2018-02-20 discussion.tex
Section 2, wip

@@
 \todo{discussion}
+
+\rkc{something about UI choices for displaying large indeterminate expressions}
+
+\rkc{something about UI choices for displaying hole environment traces}
+
+\rkc{something about patterns of use for debugging, such as inserting holes
+around all expressions in tail position}
+
+\rkc{\citet{Seidel2016} define a partial evaluation with a form of holes. compare.
+compare ``flattened'' sumList indeterminate result to their trace. could be
+useful to study whether evaluating the type-correct parts of the evaluation
+trace help students focus on the part that has the error.}

### 2018-02-20 examples.tex
Section 2, wip

@@
 %
-Furthermore, it is natural to ``jump back and forth'' between different places
-in the program, iteratively and alternately developing the missing pieces.
+It is natural for a programmer to ``jump back and forth'' between different
+places in the program, iteratively and alternately developing the missing
+pieces.
 %
@@
 %
-The teacher begins writing a function called \li{weighted_average} that will
+The teacher begins writing a \li{weighted_average} function that will
 compute a weighted average, of type \li{float}, for each \li{student_data}
@@
 Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
-(\ie{}~\li{a}) and values it produces are assigned corresponding unique dynamic
-identifiers (\ie{}~\li{a.1}, \li{a.2}, etc{}).
+(\ie{}~\li{a}) and that the indeterminate results it produces are assigned
+corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2}, etc.).
 %
@@
 %
-The teacher builds the summand weighting the \li{hw1} score to account for 10\%
+The teacher uses a \HazelnutLive{} edit action to create a sum expression,
+filling in the first summand by weighting the \li{hw1} score to account for 10\%
 of the weighted average; the rest of the computation is a new hole expression
@@
 The \li{grade_cutoffs} record type describes the minimum cutoff for each of
-these six possible.
+these six possible grades.
 %
-Initially, each of values in the \li{cutoffs} is a hole.
+Initially, each value in the \li{cutoffs} record is a hole.
 
@@
 %
-The teacher sorts the \li{weighted_averages} because she intends to look at the
-data to help select the cutoff values, and then maps them to \li{letter_grades}.
+Because she intends to look at the data to help select the cutoff values, the
+teacher sorts the \li{weighted_averages} and then maps them to
+\li{letter_grades}.
 
@@
 %
-When \HazelnutLive{} runs this program, the guard of the outermost conditional,
-\li{n >= cutoffs.a}, is indeterminate because \li{cutoffs.a} is.
+When \HazelnutLive{} runs this program, the guard of the outermost conditional
+(\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
+is.
 %
@@
 Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
-indeterminate expression as a tooltip.
+indeterminate expression---as well as the evaluation environment that surrounds
+it---as a tooltip.
 %
@@
 Running \li{compute_distribution} shows that the percentage of As is
-\rkc{XXX\%}, which is smaller than what the teacher would like.
+\rkc{XXX\%}, which is smaller than what the teacher would like;
+the remaining percentages are all indeterminate.
 %
@@
 
-\rkc{STOPPING HERE FOR NOW}
-
 %% \subsection{Live Programming with Type Errors}
 \subsection{Live Programming with Static Type Errors}
+
+In the previous section, we described example programs that were incomplete
+because of \emph{missing} expressions.
+%
+Next, we describe example programs that are incomplete because of
+\emph{type-inconsistent} expressions.
+%
+For example, consider the following two definitions.
+
+\begin{lstlisting}
+let bad_bool : bool = ?? 0 ??_bad_bool;
+
+let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
+\end{lstlisting}
+
+\noindent
+%
+These two definitions are ill-typed under standard typing disciplines.
+%
+In contrast, \citet{popl-paper} present a bidirectional type system that assigns
+types to both, by wrapping type-inconsistent expressions (\li{0} on line
+\rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
+%
+Non-empty holes prevent local type inconsistencies from polluting the rest of
+the program surrounding it, which may or may not itself contain additional
+inconsistencies.
 
@@
 %
-By running type-incorrect programs as far as possible (in particular, until they
-would go wrong), our semantics can provide a similar benefit operationally.
+In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
+non-empty holes) as far as possible, in particular, until they
+would go wrong.
 %
-\rkc{they define a partial evaluation with a form of holes. compare.}
+Thus, our operational semantics provides a similar benefit to the approach of
+\citet{Seidel2016}.
 
-\emph{Example 3: Sum List.}
+\overviewExample{3}{Sum List}
 %
-Consider the following buggy program of numbers reported by \citet{XXX} that
-attempts to sum a list numbers.
+Consider the following buggy program (observed during an undergraduate
+functional programming course~\cite{Seidel2016}) that attempts to sum a list
+integers.
 %
-The problem is that the base case produces a list rather than a number (in
-particular, 0).
+The error is that the base case produces a list rather than an integer.
 
@@
 Because the list expression on line 2 does not have type \li{int} as required,
-it is wrapped in a (non-empty) hole.
+it is wrapped in a (non-empty) hole by the bidirectional type
+checker~\cite{popl-paper}.
 %
 Rather than trying to debug the error based on the static error, the programmer
-may wish to run a call to the function, say \li{sumList(2)}, in which case
-\HazelnutLive{} runs and produces the indeterminate value \li{1 + 2 + ?[]?}
-(\cf{} the trace displayed by \citet{Seidel2016}).
+may wish to trying running the function anyway by calling, say, \li{sumList(2)}.
 %
-By observing that the hole expression is being added to the number \li{2}, he
-realizes that it needs to be a number, specifically, \li{0}.
+\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
+%
+By observing that the hole expression is being added to the integer \li{3}, he
+realizes that it needs to be an integer, specifically, \li{0}.
+%
+Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
+produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
+successfully proceeded to evaluate despite the error elsewhere.
 
@@
 
-\parasection{Example 4: Stutter}
+\overviewExample{4}{Stutter}
 %
-As another example, consider the following function which attempts to produce a
-list where every element is repeated twice (borrowed from \citep{XXX}).
+Consider the following function which attempts to produce a
+list where every element is repeated twice (borrowed from \citet{Osera2015}).
 %
@@
 %
-The editor has a choice about which expression to ``blame'' for the error; the
-entire application that forms the body of the lambda is analyzed against the
-return type \li{list(int)}, so that is a reasonable choice for the editor to
-make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
-both as intended and that only the function \li{(:)} is type-consistent.
+The bidirectional type checker of \citet{popl-paper} wraps the expression
+\li{[x,x]} inside a non-empty hole.
 %
-Although one could imagine a setting in which a user would perform this
-reasoning, let's assume the simplest approach for marking consistencies that
-wraps the entire application.
+%% The editor has a choice about which expression to ``blame'' for the error; the
+%% entire application that forms the body of the lambda is analyzed against the
+%% return type \li{list(int)}, so that is a reasonable choice for the editor to
+%% make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
+%% both as intended and that only the function \li{(:)} is type-consistent.
+%% %
+%% Although one could imagine a setting in which a user would perform this
+%% reasoning, let's assume the simplest approach for marking consistencies that
+%% wraps the entire application.
 %
-Running this on \li{stutter [1,2,3]} produces the indeterminate value
-\li{?  [1,1] : (? [2,2] : (? [3:3] : [] ?) ?) ?}, which shows the unfolding of
-\li{List.foldr}.
+Running this on \li{stutter [1,2,3]} produces the indeterminate result
+
+\begin{lstlisting}
+?  [1,1] : (? [2,2] : (? [[3:3]] ?) ?) ?,
+\end{lstlisting}
+
+\noindent
 %
-Inside the innermost hole, the value is shown to be \li{[[3,3]]}; the user
-realizes that there are too many levels of nesting.
+which shows the unfolding of \li{List.foldr}.
 %
-He tries append in place of cons, which addresses the type inconsistency and
-produces the desired result.
+We refer to nested indeterminate computations like this as \rkc{\emph{hole
+environment traces} or \emph{hole environment trees}}.
+%
+The result of the innermost indeterminate expression is \li{[[3,3]]}.
+%
+The user realizes that there are too many levels of nesting, so
+he replaces the \li{(:)} with \li{(@)}, which addresses the type inconsistency
+and, when reevaluated, produces the desired result.
+
+\rkc{STOPPING HERE FOR NOW}
 
@@
 
-\parasection{Example 5: \rkc{...}}
+\overviewExample{5}{\rkc{...}}
 
@@
 
-\parasection{Example 6: \rkc{...}}
+\overviewExample{6}{\rkc{...}}
 
@@
 \emph{hole environment traces} or \emph{hole environment traces}
+
+\paragraph{Recap}
+%
+In the above four subsections, respectively, we considered programs with:
+%
+(1) empty expression holes,
+%
+(2) non-empty expression holes for ill-typed expressions,
+%
+(3) type holes, and
+%
+(4) non-empty expression holes for well-typed expressions.
+%
+Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
+that allows programs with combinations of these kinds of incompleteness to be
+evaluated.

### 2018-02-20 examples.tex
Section 2, wip

@@
 
-\rkc{STOPPING HERE FOR NOW}
-
 \subsection{Live Programming with Gradual Type Errors}
 
-\overviewExample{5}{\rkc{...}}
+An alternative to marking type-inconsistent expressions as sources of error
+(with non-empty holes) is to use \emph{gradual types}~(\eg{}~\cite{XXX,XXX,XXX})
+to sidestep the choice of a assigning single static type, instead deferring to
+dynamic casts.
+%
+\HazelnutLive{} inherits the notion of \emph{type holes} from
+\citet{popl-paper}, which serves a similar static purpose as the unknown type in
+gradual type systems.
 
-\rkc{gradual typing error, cast error, run around}
+\overviewExample{5}{Dynamically Typed Negation}
+
+Consider the \li{negate} function (adapted from \cite{ChughPOPL2012}), which
+does not have a conventional static type (without using expression holes),
+because the argument \li{x} is used at type \li{int} on line \rkc{XXX} and
+function type on line \rkc{XXX}.
+%
+Therefore, the declared type of \li{x} is the hole type \li{??}, which allows
+\li{x} to be used at conflicting types by expanding each use with \emph{casts}
+that ensure the safety of each dynamic use.
+%
+The return type also uses the hole type \li{??}.
 
 \begin{lstlisting}
-foo : bool -> ? -> int
-foo b =
-  if b then
-    x + 10
-  else
-    x 10
+negate : bool -> ?? -> ??
+negate b x =
+  if b
+    then 0 - x     // expanded to: (0 - (x<?? => int>))<int => ??>
+    else not x     // expanded to: (not (x<?? => bool>))<bool => ??>
 
-foo false 1 + 2 + 2
+(negate false 1) + 2 + 3 + (negate true 4)
 \end{lstlisting}
 
-this demonstrates how, when no expression holes, our system is like gradual
-typing, but with the ability to run around cast errors.
+\noindent
 %
-just as it may be useful to report multiple type errors, this allows ability to
-report multiple dynamic errors, as well additional dynamic behavior around them.
+As in prior gradually typed languages~\cite{XXX},
+evaluating the expression \li{negate true 1} on line \rkc{XXX} leads to
+\li{(not (1<int => ??><?? => bool>))<bool => ??>}, where the inner casts
+lead to the \emph{cast error} \li{1<?? =/=> bool>} because \li{1} cannot be
+safely treated as having type \li{bool}.
 %
-could be incorporated into other gradual type systems
+Unlike in prior gradually typed languages, however, \HazelnutLive{} evaluates
+around the cast error, making progress on the \li{2 + 3 + negate false 4}
+expression that surrounds the error; the final indeterminate result is
+\li{(not (1<?? =/=> bool>))<bool => ??> + 1}.
+%
+Just like it is useful for static type checkers to report multiple errors, our
+approach allows us to report multiple dynamic cast errors, and otherwise make
+progress on expressions that do not depend on failed casts.
+%
+This approach can be incorporated into existing gradually typed languages.
+
+\rkc{STOPPING HERE FOR NOW}
 

### 2018-02-20 fig-grades-example.tex

\begin{figure}[t]

\begin{lstlisting}
type student_data =
  { name: string, 
    hw1: float, hw2: float, hw3: float, 
    midterm: float, final: float };

let grades : list(student_data) =
  [ { name: "Alice", hw1: 76.0, hw2: 93.0, hw3: 91.0, midterm: 77.0, final: 95.0 }
  , { name: "Bob"  , hw1: 88.0, hw2: 75.0, hw3: 80.0, midterm: 89.0, final: 87.0 }
  ];
\end{lstlisting}

\rkc{Place this code fragment (and some pieces of the subsequent fragments) in a
single mockup or screenshot.}

\caption{\rkc{...}}
\label{fig:grades-example}
\end{figure}

### 2018-02-20 implementation.tex

@@
-% !TEX root = hazelnut-popl17.tex
-\subsection{Implementation Concepts}
-Central to any implementation of Hazelnut is a stream of edit states whose cursor erasures synthesize types under an empty context according to the synthetic action judgement, $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$.
-The middle row of Figure \ref{fig:impl-overview} diagrams this stream of edit states.
-By Theorem \ref{thrm:actsafe}, the editor does not need to typecheck each edit state (though in some of the rules, e.g.
-Rule (\ref{rule:zipper-asc}) which handles the situation where the type in a type ascription changes, portions of the program do need to be typechecked.) For example, the reader is encouraged to re-examine the examples in Figure \ref{fig:first-example} and \ref{fig:second-example} -- the cursor erasure of each edit state synthesizes a type.
+% !TEX root = hazelnut-dynamics.tex
 
-The programmer examines a view generated from each edit state and produces actions in some implementation-defined manner (e.g.
-using a keyboard, mouse, touchscreen, voice interface, or neural implant.) Each new action causes a new abstract edit state to arise according to an implementation of the action semantics.
-This then causes a new view to arise.
-This is a simple event-based functional reactive programming model \cite{Wan:2000:FRP:349299.349331}.
+\newcommand{\implementationSec}{Implementation}
+\section{\protect\implementationSec}
+\label{sec:implementation}
 
-If an action is not well-defined according to Hazelnut's action semantics, the implementation must reject it.
-In fact, the implementation is encouraged to present an ``action palette'' that either hides or visibly disables actions that are not well-defined (see below.)
+\rkc{there should be at least a brief section about the implementation, right?}
 
-\begin{figure}
-\centering
-\includegraphics[width=\columnwidth]{impl-overview2}
-\caption{Implementation Concepts}
-\label{fig:impl-overview}
-\end{figure}
-
-\subsection{HZ}
-We have developed a simple reference implementation, HZ, of Hazelnut.
-In order to reach a wide audience, we decided to implement HZ in the web browser.
-In order to take advantage of a mature implementation of the FRP model, we chose to implement HZ using OCaml\footnote{\url{https://ocaml.org/}}, the \texttt{js\_of\_ocaml} compiler and associated libraries \cite{DBLP:conf/ml/Balat06}\footnote{\url{http://ocsigen.org/js\_of\_ocaml/}} and the OCaml React library\footnote{\url{http://erratique.ch/software/react}}.
-
-At the time of the writing, 
-HZ's view computation renders the model as a string embedded into HTML that approximates our typeset notation in this paper.
-The action palette is a collection of buttons and text boxes, which are disabled when an action is not well-defined or an invalid action argument is entered.
-We determine this by simply attempting to perform the action internally and catching the exception that is raised when an action is undefined.
-This design, of course, is not meant to score marks for usability or performance.
-Instead, as a simple reference implementation, we expect its source code to be of use to others who are interested in layering a more fluid user interface atop the core semantics (which have been implemented in a style that follows the presentation in this paper closely), or extensions thereof.
+\rkc{describe the UI that allows a mix of text/structured edits that ensure
+well-formed edit states}

### 2018-02-21 examples.tex
Section 2, first full rough draft

@@
 \begin{lstlisting}
-sumList : list(int) -> int
-sumList [] = ?[]?
-sumList (n:ns) = n + sumList ns
+sum_list : list(int) -> int
+sum_list [] = ?[]?
+sum_list (n:ns) = n + sum_list ns
 \end{lstlisting}
@@
 Rather than trying to debug the error based on the static error, the programmer
-may wish to trying running the function anyway by calling, say, \li{sumList(2)}.
+may wish to trying running the function anyway by calling, say, \li{sum_list(2)}.
 %
@@
 
-An alternative to marking type-inconsistent expressions as sources of error
-(with non-empty holes) is to use \emph{gradual types}~(\eg{}~\cite{XXX,XXX,XXX})
-to sidestep the choice of a assigning single static type, instead deferring to
-dynamic casts.
+\emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept
+programs that would otherwise be rejected by static type systems---either
+because type inference cannot reconstruct a valid type assignment, or because
+there may not be a unique valid type assignment at all.
 %
+In either case, gradual type systems allow types to contain \emph{unknown
+types}, and partially unknown types can be used in contexts that require
+different types, so long as they are \emph{consistent} (intuitively, equal
+modulo the unknown parts).
+%
+Dynamic casts are then inserted to ensure that these remaining static
+discrepancies are not violated at run-time.
+
 \HazelnutLive{} inherits the notion of \emph{type holes} from
@@
 gradual type systems.
+%
+In contrast to prior gradually typed languages, however, \HazelnutLive{} can
+evaluate ``around'' cast errors in the same way as it does for expression holes.
 
@@
 Consider the \li{negate} function (adapted from \cite{ChughPOPL2012}), which
-does not have a conventional static type (without using expression holes),
-because the argument \li{x} is used at type \li{int} on line \rkc{XXX} and
-function type on line \rkc{XXX}.
+cannot be assigned a static type in a conventional type system---either a
+bidirectional one, like in \HazelnutLive{}, or one with ML-style,
+unification-based inference---because the argument \li{x} is used at type
+\li{int} on line \rkc{XXX} and \li{bool} on line \rkc{XXX}.
 %
 Therefore, the declared type of \li{x} is the hole type \li{??}, which allows
-\li{x} to be used at conflicting types by expanding each use with \emph{casts}
-that ensure the safety of each dynamic use.
+\li{x} to be used at the conflicting types, with each use expanded into an
+expression wrapped in a \emph{cast} that will dynamically check for safety.
 %
-The return type also uses the hole type \li{??}.
+The return type also uses the hole type \li{??}, leading to additional casts in
+the expansion.
 
@@
 As in prior gradually typed languages~\cite{XXX},
-evaluating the expression \li{negate true 1} on line \rkc{XXX} leads to
+evaluating the expression \li{negate false 1} on line \rkc{XXX} leads to
 \li{(not (1<int => ??><?? => bool>))<bool => ??>}, where the inner casts
@@
 
-\rkc{STOPPING HERE FOR NOW}
-
 \subsection{Live Programming for Debugging}
 
-\overviewExample{6}{\rkc{...}}
+Previous examples highlighted how running different kinds of incomplete
+programs---ones with missing expressions, type-inconsistent expressions, or
+missing types---can be useful.
+%
+Our final example demonstrates how expression holes may be useful even for
+complete programs, in order to facilitate debugging and program understanding.
 
-\rkc{breakpoints, understanding input/output behavior}
+%% \overviewExample{6}{\rkc{...}}
+%%
+%% \rkc{breakpoints, understanding input/output behavior}
 
-\parasection{Example 6: Quicksort}
+\overviewExample{6}{Quicksort}
+
+Consider the following buggy implementation of quicksort, which contains several
+errors.
+%
+The student wants to debug the implementation and so inserts an expression
+around the return expression on line \rkc{XXX}, and runs
+\li{quicksort [X,X,X,X,X,X]}.
 
@@
   let (left, right) = span ((>) x) xs in
-  ??
+  let (left', right') = (quicksort left, quicksort right) in
+  ? (left' @ [x] @ right) ?
 \end{lstlisting}
 
-programmer has some intended relationship between names and values they bind.
+\noindent
 %
-live results show that left has larger elements and right has smaller.
+Based on the live environment for the initial call, the student notices that the
+values in \li{left} are larger than the pivot \li{x}, but \li{left} was intended
+to bind smaller values.
 %
-right away, can see that need to change filter predicate. (alternatively, the
-names could have been swapped, but that order of names was the natural way
-to think about it.)
+This suggests that the student got the filtering predicate backwards, so he
+replaces it with \li{(<)} on line \rkc{XXX}.
 %
-but next, still see that right has some smaller elements.
+After re-running the program and viewing the live output again, \li{left} now
+contains smaller values than \li{x}, but not all of them; \li{right} has some
+values that are larger and some that are smaller.
 %
-should pick partition instead of span.
+The student uses type-based search to find other functions, besides \li{span},
+of type \li{('a -> bool) -> 'a list -> ('a list, 'a list)} and finds
+\li{partition}.
+%
+It may be easier for the student to simply try the function rather than studying
+the documentation, so he edits line \rkc{XXX} to call \li{partition} instead,
+re-runs, and observes that both \li{left} and \li{right} now seem to achieve the
+intended partitioning.
+
+Despite these two fixes, the result expression is not yet sorted.
+%
+In particular, the values greater than the pivot \li{x} are not completely
+sorted.
+%
+Viewing the values of bindings of \li{left'} and \li{right'}, the student sees
+that both seem to have correct prefixes of sorted lists, but that they are
+incorrect after the pivots.
+%
+This suggests that something may be wrong with the handling of \li{right'}.
+%
+Indeed, the student takes a look for uses of \li{right'} and sees that line
+\rkc{XXX} mistakenly uses \li{right} instead.
+%
+Making this change and re-running now produces the desired result, built up as a
+hole environment tree that shows how all of the recursive results will be
+combined by a series of nested appends.
+%
+Happy with the results of the program, the student removes the hole around the
+expression on line \rkc{XXX}, and evaluation can proceed to perform the nested
+appends, producing the final sorted result.
+%
+\autoref{sec:discussion} discusses how debugging tools may be designed to
+systematically insert hole expressions in particular places to achieve certain
+debugging and program understanding strategies.
+%
+%% When run on a larger input, this projection of the full evaluation trace shows a
+%% nesting structure---which could be rendered as visualization to help
+%% understanding, as described in \autoref{sec:discussion}---whose depth is clearly
+%% much smaller than the length of the input list; this suggests that the
+%% algorithm, indeed, achieves some sort of sublinear complexity, as intended.
+%
 
@@
   let (left', right') = (quicksort left, quicksort right) in
-  ? (left' @ [x] @ right) ?
+  ? (left' @ [x] @ right') ?
 \end{lstlisting}
-
-now, put the lists back together.
-%
-doesn't do what is expected.
-%
-live feedback shows right' is computed correctly, but ah, it's not being used
-in the return list.
-%
-fix, and re-run, keeping the final expression in a hole.
-%
-can visualize that the nested appends are being built up correctly, and that the
-result would be the correct list.
-%
-finalize the hole and resume. everything but the nested appends can be reused.
-%
-\emph{hole environment traces} or \emph{hole environment traces}
 

### 2018-02-24 calculus.tex
breaking judgement figures into separate files

@@
 
-\begin{figure}[!ht]
-$\arraycolsep=4pt\begin{array}{rllllll}
-\mathsf{HTyp} & \htau & ::= &
-  b ~\vert~
-  \tarr{\htau}{\htau} ~\vert~
-  % \tprod{\htau}{\htau} ~\vert~
-  % \tsum{\htau}{\htau} ~\vert~
-  \tehole\\
-\mathsf{HExp} & \hexp & ::= &
-  c ~\vert~
-  x ~\vert~
-  \halam{x}{\htau}{\hexp} ~\vert~
-  \hap{\hexp}{\hexp} ~\vert~
-  % \hpair{\hexp}{\hexp} ~\vert~
-  % \hprj{i}{\hexp} ~\vert~
-  % \hinj{i}{\hexp} ~\vert~
-  % \hcase{\hexp}{x}{\hexp}{x}{\hexp} ~\vert~
-  % \hadd{\hexp}{\hexp} ~\vert~
-  \hehole{u} ~\vert~
-  \hhole{\hexp}{u} ~\vert~
-  {\hlam{x}{\hexp}} ~\vert~
-  \hexp : \htau\\
-% \mathsf{Mark} & \markname{} & ::= &
-%   \evaled{} ~\vert~  \unevaled{}\\
- \mathsf{DHExp} & \dexp  & ::= &
-  c ~\vert~
-  x ~\vert~
-  {\halam{x}{\htau}{\dexp}} ~\vert~
-  \hap{\dexp}{\dexp} ~\vert~
-  % \hpair{\dexp}{\dexp} ~\vert~
-  % \hprj{i}{\dexp} ~\vert~
-  % \hinj{i}{\dexp} ~\vert~
-  % \hcase{\dexp}{x}{\dexp}{x}{\dexp} ~\vert~
-  % \hadd{\dexp}{\dexp} ~\vert~
-  \dehole{\mvar}{\subst}{} ~\vert~
-  \dhole{\dexp}{\mvar}{\subst}{} ~\vert~
-  \dcast{\htau}{\dexp}\\
-\end{array}$
-\caption{Syntax of H-types, H-expressions and dynamic H-expressions. We write $x$ to range over variables, $u$ over hole names and $\sigma$ over finite substitutions.}
-\label{fig:HTyp}
-\label{fig:HExp}
-\end{figure}
+\input{fig-syntax}
 
-\begin{figure}[!ht]
-\judgbox{\tconsistent{\htau_1}{\htau_2}}{$\htau_1$ is consistent with $\htau_2$}
-\begin{mathpar}
-\inferrule[TCHole1]{ }{
-  \tconsistent{\tehole}{\htau}
-}
+\input{fig-type-consistency}
 
-\inferrule[TCHole2]{ }{
-  \tconsistent{\htau}{\tehole}
-}
+\input{fig-bidirectional-typing}
 
-\inferrule[TCRefl]{ }{
-  \tconsistent{\htau}{\htau}
-}
+\input{fig-expansion}
 
-\inferrule[TCArr]{
-  \tconsistent{\htau_1}{\htau_1'}\\
-  \tconsistent{\htau_2}{\htau_2'}
-}{
-  \tconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-}
-%
-% \inferrule{
-%   \tconsistent{\htau_1}{\htau_1'}\\
-%   \tconsistent{\htau_2}{\htau_2'}
-% }{
-%   \tconsistent{\tprod{\htau_1}{\htau_2}}{\tprod{\htau_1'}{\htau_2'}}
-% }
-%
-% \inferrule{
-%   \tconsistent{\htau_1}{\htau_1'}\\
-%   \tconsistent{\htau_2}{\htau_2'}
-% }{
-%   \tconsistent{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1'}{\htau_2'}}
-% }
-\end{mathpar}
+\input{fig-typing-dexp}
 
-\judgbox{\tinconsistent{\htau_1}{\htau_2}}{$\htau_1$ is inconsistent with $\htau_2$}
-\begin{mathpar}
-    \inferrule[ICBaseArr1]{ }{
-      \tinconsistent{\tb}{\tarr{\htau_1}{\htau_2}}
-    }
+\input{fig-substitution}
 
-    \inferrule[ICBaseArr2]{ }{
-      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tb}
-    }
+\input{fig-dynamics-aux}
 
-    \inferrule[ICArr1]{
-      \tinconsistent{\htau_1}{\htau_3}
-    }{
-      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_3}{\htau_4}}
-    }
+\input{fig-dynamics-steps}
 
-    \inferrule[ICArr2]{
-      \tinconsistent{\htau_2}{\htau_4}
-    }{
-      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_3}{\htau_4}}
-    }
-\end{mathpar}
-
-\judgbox{\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}}{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
-\begin{mathpar}
-\inferrule[MAHole]{ }{
-  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
-}
-
-\inferrule[MAArr]{ }{
-  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
-}
-\end{mathpar}
-
-% \judgbox{\prodmatch{\htau}{\tprod{\htau_1}{\htau_2}}}{$\htau$ has matched product type $\tprod{\htau_1}{\htau_2}$}
-% \begin{mathpar}
-% \inferrule{ }{
-%   \prodmatch{\tehole}{\tprod{\tehole}{\tehole}}
-% }
-
-% \inferrule{ }{
-%   \prodmatch{\tprod{\htau_1}{\htau_2}}{\tprod{\htau_1}{\htau_2}}
-% }
-% \end{mathpar}
-
-% \judgbox{\summatch{\htau}{\tsum{\htau_1}{\htau_2}}}{$\htau$ has matched sum type $\tsum{\htau_1}{\htau_2}$}
-% \begin{mathpar}
-% \inferrule{ }{
-%   \summatch{\tehole}{\tsum{\tehole}{\tehole}}
-% }
-
-% \inferrule{ }{
-%   \summatch{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1}{\htau_2}}
-% }
-% \end{mathpar}
-\caption{Type Consistency, Inconsistency, and Matching}
-\label{fig:tconsistent}
-\label{fig:arrmatch}
-\end{figure}
-
-\begin{figure}[!ht]
-\judgbox{\hsyn{\hGamma}{\hexp}{\htau}}{$\htau$ synthesizes type $\htau$}
-\begin{mathpar}
-\inferrule[SConst]{ }{
-  \hsyn{\hGamma}{c}{b}
-}
-
-\inferrule[SVar]{ }{
-  \hsyn{\hGamma, x : \htau}{x}{\htau}
-}
-
-\inferrule[SLam]{
-  \hsyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}
-}{
-  \hsyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}
-}
-
-\inferrule[SAp]{
-  \hsyn{\hGamma}{\hexp_1}{\htau_1}\\
-  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
-  \hana{\hGamma}{\hexp_2}{\htau_2}
-}{
-  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
-}
-
-\inferrule[SEHole]{ }{
-  \hsyn{\hGamma}{\hehole{u}}{\tehole}
-}
-
-\inferrule[SNEHole]{
-  \hsyn{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}
-}
-
-\inferrule[SAsc]{
-  \hana{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hexp : \htau}{\htau}
-}
-\end{mathpar}
-
-\judgbox{\hana{\hGamma}{\hexp}{\htau}}{$\htau$ analyzes against type $\htau$}
-\begin{mathpar}
-\inferrule[ALam]{
-  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
-  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
-}{
-  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
-}
-
-\inferrule[ASubsume]{
-  \hsyn{\hGamma}{\hexp}{\htau}\\
-  \tconsistent{\htau}{\htau'}
-}{
-  \hana{\hGamma}{\hexp}{\htau'}
-}
-\end{mathpar}
-\caption{Bidirectional Typing of H-Expressions}
-\end{figure}
+\input{fig-dynamics-contexts}
 
@@
 
-\begin{figure}[!ht]
-\judgbox{\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}}{ }
-\begin{mathpar}
-\inferrule[ESConst]{ }{
-  \expandSyn{\hGamma}{c}{b}{c}{\emptyset}
-}
-
-\inferrule[ESVar]{
-  x : \htau \in \hGamma
-}{
-  \expandSyn{\hGamma}{x}{\htau}{x}{\emptyset}
-}
-
-\inferrule[ESLam]{
-  \expandSyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\Delta}
-}{
-  \expandSyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\Delta}
-}
-
-\inferrule[ESAp1]{
-  \hsyn{\hGamma}{\hexp_1}{\tehole}\\
-  \expandAna{\hGamma}{\hexp_1}{\tarr{\htau_2}{\tehole}}{\dexp_1}{\htau_1}{\Delta_1}\\
-  \expandAna{\hGamma}{\hexp_2}{\tehole}{\dexp_2}{\htau_2}{\Delta_2}
-}{
-  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}{\hap{(\dcast{\tarr{\htau_2}{\tehole}}{\dexp_1})}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
-}
-
-\inferrule[ESAp2]{
-  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
-  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau'_2}{\Delta_2}\\
-  \htau_2 \neq \htau'_2
-}{
-  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dcast{\htau_2}{\dexp_2}}}{\Dunion{\Delta_1}{\Delta_2}}
-}
-
-\inferrule[ESAp3]{
-  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
-  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau_2}{\Delta_2}
-}{
-  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
-}\\
-%
-% \inferrule[expand-pair]{
-%   \expandSyn{\hGamma}{\hexp_1}{\htau_1}{\dexp_1}{\Delta_1}\\
-%   \expandSyn{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\Delta_2}
-% }{
-%   \expandSyn{\hGamma}{\hpair{\hexp_1}{\hexp_2}}{\tprod{\htau_1}{\htau_2}}{\hpair{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
-% }
-%
-% \inferrule[expand-prj]{
-%   a
-% }{
-%   b
-% }
-%
-% (inj)
-
-%
-% \inferrule[expand-plus]{ }{
-%   \expandSyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}{\hadd{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
-% }
-%
-\inferrule[ESEHole]{ }{
-  \expandSyn{\hGamma}{\hehole{u}}{\tehole}{\dehole{u}{\idof{\hGamma}}{\unevaled}}{\Dbinding{u}{\hGamma}{\tehole}}
-}
-
-\inferrule[ESNEHole]{
-  \expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}
-}{
-  \expandSyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}{\dhole{\dexp}{u}{\idof{\hGamma}}{\unevaled}}{\Delta, \Dbinding{u}{\hGamma}{\tehole}}
-}\\
-%
-\inferrule[ESAsc1]{
-  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}\\
-  \htau \neq \htau'
-}{
-  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dcast{\htau}{\dexp}}{\Delta}
-}
-
-\inferrule[ESAsc2]{
-  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}
-}{
-  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dexp}{\Delta}
-}
-\end{mathpar}
-
-\judgbox{\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}}{ }
-\begin{mathpar}
-\inferrule[EALam]{
-  \expandAna{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\htau'_2}{\Delta}
-}{
-  \expandAna{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2'}}{\Delta}
-}
-
-\inferrule[EALamHole]{
-  \expandAna{\hGamma, x : \tehole}{\hexp}{\tehole}{\dexp}{\htau}{\Delta}
-}{
-  \expandAna{\hGamma}{\hlam{x}{\hexp}}{\tehole}{\halam{x}{\tehole}{\dexp}}{\tarr{\tehole}{\htau}}{\Delta}
-}
-
-\inferrule[EASubsume]{
-  \hexp \neq \hehole{u}\\
-  \hexp \neq \hhole{\hexp'}{u}\\\\
-  \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
-  \tconsistent{\htau}{\htau'}
-}{
-  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}
-}
-
-\inferrule[EAEHole]{ }{
-  \expandAna{\hGamma}{\hehole{u}}{\htau}{\dehole{u}{\idof{\hGamma}}{\unevaled}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
-}
-
-\inferrule[EANEHole]{
-  \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
-}{
-  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}{\unevaled}}{\htau}{\Delta, \Dbinding{u}{\hGamma}{\htau}}
-}
-\end{mathpar}
-\caption{Algorithmic Expansion}
-\label{fig:expandSyn}
-\label{fig:expandAna}
-\end{figure}
-
-\begin{figure}[!ht]
+%% \begin{figure}[!ht]
   \begin{definition}
@@
   \end{definition}
-  \caption{substitution type assignment}
-  \label{fig:subassign}
-\end{figure}
+%%   \caption{substitution type assignment}
+%%   \label{fig:subassign}
+%% \end{figure}
 
-\begin{figure}[!ht]
-\judgbox{\hasType{\Delta}{\hGamma}{\dexp}{\htau}}{$\dexp$ is assigned type $\htau$}
-\begin{mathpar}
-\inferrule[TAConst]{ }{
-  \hasType{\Delta}{\hGamma}{c}{b}
-}
-
-\inferrule[TAVar]{
-  x : \htau \in \hGamma
-}{
-	\hasType{\Delta}{\hGamma}{x}{\htau}
-}
-
-\inferrule[TALam]{
-  \hasType{\Delta}{\hGamma, x : \htau_1}{\dexp}{\htau_2}
-}{
-  \hasType{\Delta}{\hGamma}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2}}
-}
-
-\inferrule[TAAp]{
-  \hasType{\Delta}{\hGamma}{\dexp_1}{\htau_1}\\
-  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
-  \hasType{\Delta}{\hGamma}{\dexp_2}{\htau_2}
-}{
-  \hasType{\Delta}{\hGamma}{\hap{\dexp_1}{\dexp_2}}{\htau}
-}
-
-\inferrule[TAEHole]{
-  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\
-  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
-}{
-  \hasType{\Delta}{\hGamma}{\dehole{u}{\sigma}{}}{\htau}
-}
-
-\inferrule[TANEHole]{
-  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\\\
-  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
-  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
-}{
-  \hasType{\Delta}{\hGamma}{\dhole{\dexp}{u}{\sigma}{}}{\htau}
-}
-
-\inferrule[TACast]{
-  \hasType{\Delta}{\Gamma}{\dexp}{\htau'}\\
-  \tconsistent{\htau}{\htau'}
-}{
-  \hasType{\Delta}{\hGamma}{\dcast{\htau}{\dexp}}{\htau}
-}
-\end{mathpar}
-\caption{Type Assignment for Dynamic H-Expressions}
-\label{fig:hasType}
-\end{figure}
-
-\begin{figure}[!ht]
+%% \begin{figure}[!ht]
   \begin{definition}
@@
   \end{definition}
-  \caption{substitution type assignment}
-\end{figure}
-
-\begin{figure}[!ht]
-\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
-\[
-\begin{array}{lcll}
-[\dexp_1 / x] c
-&=&
-c\\
-{[\dexp_1 / x] x}
-&=&
-\dexp_1\\
-{[\dexp_1 / x] y}
-&=&
-y & (y \neq x)\\
-{[\dexp_1 / x] \dlam{y}{\htau}{\dexp_2}}
-&=&
-{\dlam{y}{\htau}{[\dexp_1 / x] \dexp_2}}
-& (y \neq x)
-\\
-{[\dexp_1 / x] \dap{d_2}{d_3}}
-&=&
-{\dapP{[\dexp_1 / x] d_2}{[\dexp_1 / x] d_3}}
-\\
-% {[\dexp_1 / x] \dinj{i}{e_2}}
-% &=&
-% {\dinj{i}{[\dexp_1 / x] e_2}}
-% \\
-% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
-% &=&
-% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
-% \\
-{[\dexp_1 / x]} \dehole{\mvar}{\subst}{}
-&=&
-\dehole{\mvar}{[\dexp_1 / x] \subst}{}
-\\
-{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{}}
-&=&
-\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{}
-\\
-{[\dexp_1 / x] \dcast{\htau}{\dexp}}
-&=&
-\dcast{\htau}{[\dexp_1 / x] \dexp}
-\end{array}
-\]
-\end{figure}
-
-\begin{figure}[!ht]
-\judgbox{\isValue{\dexp}}{$\dexp$ is a value}
-\begin{mathpar}
-\inferrule[VConst]{ }{
-  \isValue{c}
-}
-
-\inferrule[VLam]{ }{
-  \isValue{\halam{x}{\htau}{\dexp}}
-}
-\end{mathpar}
-
-\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
-\begin{mathpar}
-\inferrule[IEHole]
-{ }
-{\isIndet{\dehole{\mvar}{\subst}{\evaled}}}
-
-\inferrule[INEHole]
-{\isFinal{\dexp}}
-{\isIndet{\dhole{\dexp}{\mvar}{\subst}{\evaled}}}
-
-\inferrule[IAp]
-{\isIndet{\dexp_1}\\
-% \isFinal{\dexp_2}~\text{\cy{??}}}
- \isFinal{\dexp_2}}
-{\isIndet{\dap{\dexp_1}{\dexp_2}}}
-
-%% \inferrule[ICast]
-%% {\isIndet{\dexp}}
-%% {\isIndet{\dcast{\htau}{\dexp}}}
-
-\end{mathpar}
-
-\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
-\begin{mathpar}
-\inferrule[FVal]
-{\isValue{\dexp}}{\isFinal{\dexp}}
-\and
-\inferrule[FIndet]
-{\isIndet{\dexp}}{\isFinal{\dexp}}
-\end{mathpar}
-
-\judgbox{\isErr{\Delta}{\dexp}}{$\dexp$ contains a cast error}
-\begin{mathpar}
-\inferrule[ECastError]{
-  \hasType{\Delta}{\emptyset}{\dexp}{\htau_2}\\
-  \tinconsistent{\htau_1}{\htau_2}
-}{
-  \isErr{\Delta}{\dcast{\htau_1}{\dexp}}
-}
-
-\inferrule[EConst]{
-}{
-  \isErr{\Delta}{\dcast{\tau}{\dconst}}
-}
-
-\inferrule[EAp1]{
-  \isErr{\Delta}{\dexp_1}
-}{
-  \isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}
-}
-
-\inferrule[EAp2]{
-  \isErr{\Delta}{\dexp_2}
-}{
-  \isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}
-}
-
-\inferrule[ENEHole]{
-  \isErr{\Delta}{\dexp}
-}{
-  \isErr{\Delta}{\dhole{\dexp}{\mvar}{\subst}{}}
-}
-
-\inferrule[ECastProp]{
-  \isErr{\Delta}{\dexp}
-}{
-  \isErr{\Delta}{\dcast{\htau_1}{\dexp}}
-}
-\end{mathpar}
-\caption{Auxiliary Judgments for the Dynamics}
-\label{fig:isValue}
-\label{fig:isIndet}
-\label{fig:isFinal}
-\label{fig:isErr}
-\end{figure}
-
-\begin{figure}
-\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$
-steps to $\dexp_2$}
-\begin{mathpar}
-\inferrule[STEHoleEvaled]
-{ }
-{\stepsToD{\Delta}{\dehole{\mvar}{\subst}{\unevaled}}{\dehole{\mvar}{\subst}{\evaled}}}
-
-\inferrule[STNEHoleStep]
-{\stepsToD{\Delta}{\dexp_1}{\dexp_2} }
-{\stepsToD{\Delta}{\dhole{\dexp_1}{\mvar}{\subst}{\unevaled}}{\dhole{\dexp_2}{\mvar}{\subst}{\unevaled}}}
-
-\inferrule[STNEHoleEvaled]
-{\isFinal{\dexp}}
-{\stepsToD{\Delta}{\dhole{\dexp}{\mvar}{\subst}{\unevaled}}{\dhole{\dexp}{\mvar}{\subst}{\evaled}}}
-
-\inferrule[STCast]
-{
-\isValue{\dexp}\\
-\hasType{\Delta}{\emptyset}{\dexp}{\htau_2} \\
-\tconsistent{\tau_1}{\tau_2}}
-{\stepsToD{\Delta}{\dcast{\htau_1}{\dexp}}{\dexp}}
-
-\inferrule[STApStep1]
-{\stepsToD{\Delta}{\dexp_1}{\dexp_1'}}
-{\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
-
-\inferrule[STApStep2]
-{ \isFinal{\dexp_1} \\ \stepsToD{\Delta}{\dexp_2}{\dexp_2'}}
-{\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
-
-\inferrule[STApBeta]
-{ \isFinal{\dexp_2} }
-{\stepsToD{\Delta}{\dapP{\dlam{x}{\htau}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
-\end{mathpar}
-\caption{Structural Dynamics}
-\label{fig:stepsTo}
-\end{figure}
+%%   \caption{substitution type assignment}
+%% \end{figure}
 
@@
 
-\begin{figure}
-$\arraycolsep=4pt\begin{array}{rllllll}
-\mathsf{EvalCtx} & \evalctx & ::= &
-  \evalhole ~\vert~
-  \hap{\evalctx}{\dexp} ~\vert~
-  \hap{\dexp}{\evalctx} ~\vert~
-  \dhole{\evalctx}{\mvar}{\subst}{} ~\vert~
-  \dcast{\htau}{\evalctx}
-\end{array}$
-
-\judgbox{\isevalctx{\evalctx}}{$\evalctx$ is an evaluation context}
-\begin{mathpar}
-\inferrule[ECDot]{ }{
-  \isevalctx{\evalhole}
-}
-
-%% \inferrule[ECLam]{
-%%   \isevalctx{\evalctx}
-%% }{
-%%   \isevalctx{\halam{x}{\htau}{\evalctx}}
-%% }
-
-\inferrule[ECAp1]{
-  \isFinal{\dexp}\\
-  \isevalctx{\evalctx}
-}{
-  \isevalctx{\hap{\dexp}{\evalctx}}
-}
-
-\inferrule[ECAp2]{
-  \isevalctx{\evalctx}
-}{
-  \isevalctx{\hap{\evalctx}{\dexp}}
-}
-
-\inferrule[ECNEHole]{
-  \isevalctx{\evalctx}
-}{
-  \isevalctx{\dhole{\evalctx}{\mvar}{\subst}{}}
-}
-
-\inferrule[ECCast]{
-  \isevalctx{\evalctx}
-}{
-  \isevalctx{\dcast{\htau}{\evalctx}}
-}
-\end{mathpar}
-
-\judgbox{\selectEvalCtx{d}{\evalctx}{d'}}{$d$ is the result of filling the
-hole in $\evalctx$ with $d'$}
-\begin{mathpar}
-\inferrule[FHConst]{ }{
-  \selectEvalCtx{c}{\evalhole}{c}
-}
-
-%% \inferrule[FLam]{
-%%   \selectEvalCtx{d}{\evalctx}{d'}
-%% }{
-%%   \selectEvalCtx{\halam{x}{\htau}{d}}{\halam{x}{\htau}{\evalctx}}{d'}
-%% }
-
-\inferrule[FHAp1]{
-  \isFinal{d_1}\\
-  \selectEvalCtx{d_2}{\evalctx}{d_2'}
-}{
-  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{d_1}{\evalctx}}{d_2'}
-}
-
-\inferrule[FHAp2]{
-  \selectEvalCtx{d_1}{\evalctx}{d_1'}
-}{
-  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{\evalctx}{d_2}}{d_1'}
-}
-
-\inferrule[FHEHole]{ }{
-  \selectEvalCtx{\dehole{\mvar}{\subst}{}}{\evalhole}{\dehole{\mvar}{\subst}{}}
-}
-
-\inferrule[FHNEHoleEvaled]{ }{
-  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\evaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\evaled}}
-}
-
-\inferrule[FHNEHoleInside]{
-  \selectEvalCtx{d}{\evalctx}{d'}
-}{
-  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\dhole{\evalctx}{\mvar}{\subst}{\unevaled}}{d'}
-}
-
-\inferrule[FHNEHoleFinal]{
-  \isFinal{d}
-}{
-  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\unevaled}}
-}
-
-\inferrule[FHCast]{
-  \selectEvalCtx{d}{\evalctx}{d'}
-}{
-  \selectEvalCtx{\dcast{\htau}{d}}{\dcast{\htau}{\evalctx}}{d'}
-}
-
-\inferrule[FHCastFinal]{
-  \isFinal{d}
-}{
-  \selectEvalCtx{\dcast{\htau}{d}}{\evalhole}{\dcast{\htau}{d}}
-}
-\end{mathpar}
-
-\judgbox{\reducesE{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$
-transitions to $\dexp_2$}
-\begin{mathpar}
-\inferrule[ITLam]{
-  \isFinal{d_2}
-}{
-  \reducesE{\Delta}{\hap{(\halam{x}{\htau}{d})}{d_2}}{[d_2/x]d}
-}
-
-\inferrule[ITCast]{
-  \isFinal{d}\\
-  \hasType{\Delta}{\emptyset}{d}{\tau_2}\\
-  \tconsistent{\tau_1}{\tau_2}
-}{
-  \reducesE{\Delta}{\dcast{\htau_1}{d}}{d}
-}
-
-\inferrule[ITEHole]{ }{
-  \reducesE{\Delta}{\dehole{\mvar}{\subst}{\unevaled}}{\dehole{\mvar}{\subst}{\evaled}}
-}
-
-\inferrule[ITNEHole]{
-  \isFinal{d}
-}{
-  \reducesE{\Delta}{\dhole{d}{\mvar}{\subst}{\unevaled}}{\dhole{d}{\mvar}{\subst}{\evaled}}
-}
-\end{mathpar}
-
-\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
-\begin{mathpar}
-\inferrule[Step]{
-  \selectEvalCtx{d}{\evalctx}{d_0}\\
-  \reducesE{\Delta}{d_0}{d_0'}\\
-  \selectEvalCtx{d'}{\evalctx}{d_0'}
-}{
-  \stepsToD{\Delta}{\dexp}{\dexp'}
-}
-\end{mathpar}
-\caption{Contextual Dynamics}
-\label{fig:contextual-dynamics}
-\end{figure}
-
 \clearpage
-\subsection{Metatheory}
+\subsection{Metatheory \rkc{TODO}}
 

### 2018-02-25 calculus.tex
some layout tweaks and placeholders
fig-syntax
fig-expansion
fig-typing-dexp

@@
 
+\subsection{Calculus}
+
+\subsubsection{Syntax}
+
 \input{fig-syntax}
 
+{\color{gray}\blindtext}
+
+Shorthand:
+
+$$
+\dcastthree{\dexp}{\htau_1}{\htau_2}{\htau_3} \defeq
+  \dcasttwo{\dcasttwo{\dexp}{\htau_1}{\htau_2}}{\htau_2}{\htau_3}
+$$
+
+Explain that $c$ is a placeholder constant (i.e., unit), $b$ is a placeholder
+type (i.e., unit).
+
+\subsubsection{Statics}
+
 \input{fig-type-consistency}
-
 \input{fig-bidirectional-typing}
 
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+
+A typing context, $\hGamma$, is a list of bindings $x : \htau$.
+
+\subsubsection{Expansion}
+
 \input{fig-expansion}
-
 \input{fig-typing-dexp}
 
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+
+A hole context, $\hDelta$, is a list of bindings $\Dbinding{u}{\hGamma}{\htau}$.
+
+\subsubsection{Dynamics}
+
 \input{fig-substitution}
+\input{fig-dynamics-aux}
+\input{fig-dynamics-steps}
+\input{fig-dynamics-contexts}
 
-\input{fig-dynamics-aux}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
 
-\input{fig-dynamics-steps}
-
-\input{fig-dynamics-contexts}
+\subsubsection{Examples}
 
@@
 %% \begin{figure}[!ht]
-  \begin{definition}
-    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
-  \end{definition}
+%%   \begin{definition}
+%%     $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+%%   \end{definition}
 %%   \caption{substitution type assignment}

### 2018-02-26 calculus.tex
fig-dynamics-aux
fig-dynamics-contexts
metatheory definitions, wip

@@
 
-\input{fig-substitution}
 \input{fig-dynamics-aux}
-\input{fig-dynamics-steps}
+%% \input{fig-dynamics-steps}
 \input{fig-dynamics-contexts}
@@
 {\color{gray}\blindtext}
+
+\begin{lemma}[Focus Formation] ~
+  If $\selectEvalCtx{\dexp}{\evalctx}{\dexp'}$
+  then $\isevalctx{\evalctx}$.
+\end{lemma}
+
+\begin{lemma}[Matched Ground Type Invariant] ~
+  If $\groundmatch{\htau}{\htau'}$
+  then $\isGround{\htau'}$
+  and $\tconsistent{\htau}{\htau'}$
+  and $\htau\neq\htau'$.
+\end{lemma}
 
@@
 \clearpage
-\subsection{Metatheory \rkc{TODO}}
+%% \subsection{Metatheory}
+\subsection{Metatheory: Type Soundness}
 
-(these theorems should be taken as fairly provisional)
+{\color{gray}\blindtext}
 
-\begin{theorem}[Expandability \cy{??}] ~
+\begin{theorem}[Expandability] ~
   \begin{enumerate}
-    \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ for some $\dexp$ and $\Delta$.
-    \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ for some $\dexp$ and $\htau'$ and $\Delta$.
+    \item
+      If $\hsyn{\hGamma}{\hexp}{\htau}$
+      then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      for some $\dexp$ and $\Delta$.
+    \item
+      If $\hana{\hGamma}{\hexp}{\htau}$
+      then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+      for some $\dexp$ and $\htau'$ and $\Delta$.
   \end{enumerate}
@@
 
-\begin{theorem}[Correspondence \cy{??}] ~
+\begin{theorem}[Correspondence] ~
   \begin{enumerate}
-    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ then $\hsyn{\hGamma}{\hexp}{\htau}$.
-    \item If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then $\hana{\hGamma}{\hexp}{\htau}$.
+    \item
+      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      then $\hsyn{\hGamma}{\hexp}{\htau}$.
+    \item
+      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+      then $\hana{\hGamma}{\hexp}{\htau}$.
   \end{enumerate}
@@
   \begin{enumerate}
-    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
-    \item If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
+    \item
+      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+    \item
+      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
   \end{enumerate}
@@
   \begin{enumerate}
-    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$ then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
-    \item If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$ and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$ then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
+    \item
+      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
+      then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
+    \item
+      If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
+      and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
+      then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
   \end{enumerate}
@@
 \begin{theorem}[Type Assignment Unicity]
-If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$ then $\htau=\htau'$.
-\end{theorem}
-
-\begin{theorem}[Progress]
-If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\stepsToD{\Delta}{\dexp}{\dexp'}$ or 
-$\isValue{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\isIndet{\dexp}$. 
-% for some $\dexp'$.
+  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
+  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
+  then $\htau=\htau'$.
 \end{theorem}
@@
   $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
-  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \cy{or maybe up to
-  consistency?}
+  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$.
 \end{theorem}
+
+\begin{theorem}[Progress]
+  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
+  (i) $\stepsToD{}{\dexp}{\dexp'}$ or
+  (ii) $\isIndet{\dexp}$ or
+  (iii) $\isBoxedValue{\dexp}$.
+\end{theorem}
+
+\begin{theorem}[Canonical Value Forms]
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isValue{\dexp}$
+  then $\htau\neq\tehole$ and
+  \begin{enumerate}[label=(\alph*)]
+    \item If $\htau=b$ then $\dexp=c$.
+    \item If $\htau=\tarr{\htau_1}{\htau_2}$
+          then $\dexp=\halam{x}{\htau_1}{\dexp'}$
+          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$.
+  \end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Canonical Boxed Forms]
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isBoxedValue{\dexp}$
+  then
+  \begin{enumerate}[label=(\alph*)]
+    \item If $\htau=b$ then $\dexp=c$.
+    \item If $\htau=\tarr{\htau_1}{\htau_2}$ then either
+      \begin{enumerate}
+        \item[i.]
+          $\dexp=\halam{x}{\htau_1}{\dexp'}$
+          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$, or
+        \item[ii.]
+          $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1'}{\htau_2'}}{\tarr{\htau_1}{\htau_2}}$
+          where $\tarr{\htau_1'}{\htau_2'}\neq\tarr{\htau_1}{\htau_2}$
+          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\tarr{\htau_1'}{\htau_2'}}$.
+      \end{enumerate}
+    \item If $\htau=\tehole$
+          then $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
+          where $\isGround{\htau'}$
+          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$.
+  \end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Canonical Indeterminate Forms]
+  \rkc{TODO}
+
+  \rkc{TODO}
+
+  \rkc{TODO}
+
+  \rkc{TODO}
+
+  \rkc{TODO}
+
+  \rkc{TODO}
+
+  \rkc{TODO}
+
+  \rkc{TODO}
+\end{theorem}
+
+\input{fig-complete}
+
+\autoref{fig:complete} defines complete types and expressions...
 
@@
 \end{theorem}
+
+\subsection{Metatheory: Correctness of Edit-and-Resume}
+
+\input{fig-substitution}
+
+{\color{gray}\blindtext}
 

### 2018-02-26 fig-complete.tex
metatheory definitions, wip

\begin{figure}[t]
\rkc{TODO}
\caption{Complete H-types, H-expressions, and dynamic H-expressions}
\label{fig:complete}
\end{figure}

### 2018-02-26 calculus.tex
fig-complete

@@
 
-\autoref{fig:complete} defines complete types and expressions...
+\autoref{fig:complete} defines \emph{complete} H-types, H-expressions, and
+dynamic H-expressions to be those terms without type or expression holes.
 
-\begin{theorem}[Complete Progress \cy{??}]
-If $\hcomplete{\hexp}$ and
-$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ then either
-$\isValue{\dexp}$ or $\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.
-
+\begin{theorem}[Complete Progress]
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
+  and $\isComplete{\dexp}$
+  then either $\stepsToD{}{\dexp}{\dexp'}$
+  or $\isValue{\dexp}$.
 \end{theorem}
 
-\begin{theorem}[Complete Preservation \cy{??}]
-If $\hcomplete{\hexp}$ and
-$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ and
-$\stepsToD{\Delta}{\dexp}{\dexp'}$ then
-$\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$
+\begin{theorem}[Complete Preservation]
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
+  and $\isComplete{\dexp}$
+  and $\stepsToD{}{\dexp}{\dexp'}$
+  then $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau}$.
+\end{theorem}
+
+\begin{theorem}[Complete Expansion] ~
+  \begin{enumerate}
+    \item
+      If $\isComplete{\hexp}$
+      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      then $\isComplete{\dexp}$.
+    \item
+      If $\isComplete{\hexp}$
+      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+      then $\isComplete{\dexp}$.
+  \end{enumerate}
 \end{theorem}

### 2018-02-26 calculus.tex
multi step, hole instantiation, resumption

@@
 \input{fig-substitution}
+\input{fig-multi-step}
 
@@
 
-\begin{theorem}[Commutativity \cy{state this}]
+\rkc{some of these statements may need additional typing premises}
+
+\begin{theorem}[Instantiation]
+  If $\hasType{\hDelta}{\hGamma}{\dexp}{\tau}$
+  and $\Dbinding{u}{\hGamma'}{\htau'}\in\hDelta$
+  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
+  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
 \end{theorem}
+
+\autoref{fig:multi-step} defines the multi-step relation as the reflexive,
+transitive closure of the single-step relation.
+
+\begin{lemma}[Finality]
+  If $\isFinal{\dexp}$ and $\multiStepsTo{\dexp}{\dexp'}$
+  then $\dexp=\dexp'$.
+\end{lemma}
+
+\begin{theorem}[Commutativity]
+  If $\multiStepsTo{\dexp_0}{\dexp_1}$
+  then $\multiStepsTo{\instantiate{\dexp}{u}{\dexp_0}}
+                     {\instantiate{\dexp}{u}{\dexp_1}}$.
+\end{theorem}
+
+\begin{theorem}[Confluence]
+  If $\multiStepsTo{\dexp}{\dexp_1}$
+  and $\multiStepsTo{\dexp}{\dexp_2}$
+  then there exists $\dexp'$ such that
+  $\multiStepsTo{\dexp_1}{\dexp'}$ and
+  $\multiStepsTo{\dexp_2}{\dexp'}$.
+\end{theorem}
+
+\begin{corollary}[Final Confluence]
+  If $\multiStepsTo{\dexp}{\dexp_1}$
+  and $\isFinal{\dexp_1}$
+  and $\multiStepsTo{\dexp}{\dexp_2}$
+  then $\multiStepsTo{\dexp_2}{\dexp_1}$.
+  \begin{proof}
+    By Confluence and Finality.
+  \end{proof}
+\end{corollary}
+
+\begin{theorem}[Resumption]
+  If $\multiStepsTo{\dexp_1}{\dexp_2}$
+  and $\isFinal{\dexp_2}$
+  and $\multiStepsTo{\instantiate{\dexp_3}{u}{\dexp_1}}{\dexp_4}$
+  and $\isFinal{\dexp_4}$
+  then $\multiStepsTo{\instantiate{\dexp_3}{u}{\dexp_2}}{\dexp_4}$.
+  \begin{proof}
+    By Commutativity,
+    $\multiStepsTo{\instantiate{\dexp_3}{u}{\dexp_1}}
+                  {\instantiate{\dexp_3}{u}{\dexp_2}}$.
+    By Final Confluence, we can conclude.
+  \end{proof}
+\end{theorem}
+
+\begin{comment}
 
@@
 
-\cy{canonical forms}
-
-\cy{indeterminate forms}
+\end{comment}

### 2018-02-28 calculus.tex
note about \dcastthree sugar
Canonical Indeterminate Forms

@@
 $$
+
+\rkc{this syntactic sugar is used in four places: ITCastSucceed, ITCastFail,
+ITGround, and ITExpand. that's not many, and those rules don't look much more
+cluttered without the sugar, so consider eliminating it. if so, just toggle the
+definition of the dcastthree macro to the unsugared option.}
 
@@
 \begin{theorem}[Canonical Indeterminate Forms]
-  \rkc{TODO}
-
-  \rkc{TODO}
-
-  \rkc{TODO}
-
-  \rkc{TODO}
-
-  \rkc{TODO}
-
-  \rkc{TODO}
-
-  \rkc{TODO}
-
-  \rkc{TODO}
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
+  and $\isIndet{\dexp}$
+  then either
+  \begin{enumerate}[label=(\alph*)]
+    \item
+      $\dexp=\dehole{u}{\subst}{}$
+      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
+    \item
+      $\dexp=\dhole{\dexp'}{u}{\subst}{}$
+      and $\isFinal{\dexp'}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$
+      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
+    \item
+      $\dexp=\dap{\dexp_1}{\dexp_2}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2}{\htau}}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2}$
+      and $\isIndet{\dexp_1}$
+      and $\isFinal{\dexp_2}$
+      and $\dexp_1\neq\dcasttwo{\dexp_1}{\tarr{\htau_3}{\htau_4}}
+                                        {\tarr{\htau_3'}{\htau_4'}}$, or
+    %% \item
+    %%   \begin{enumerate}
+    %%     \item blah
+    %%     \item blah
+    %%     \item blah
+    %%   \end{enumerate}
+    \item
+      $\htau=b$
+      and $\dexp=\dcasttwo{\dexp'}{\tehole}{b}$
+      and $\isIndet{\dexp'}$
+      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
+    \item
+      $\htau=b$
+      and $\dexp=\dcastfail{\dexp'}{\htau'}{b}$
+      and $\isGround{\htau'}$
+      and $\htau'\neq{b}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
+    \item
+      $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+      and $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1}{\htau_2}}
+                                  {\tarr{\htau_{11}}{\htau_{12}}}$
+      and $\isIndet{\dexp'}$
+      and $\tarr{\htau_1}{\htau_2}\neq\tarr{\htau_{11}}{\htau_{12}}$, or
+    \item
+      $\htau=\tarr{\tehole}{\tehole}$
+      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+      %% and $\htau_{11}=\tehole$
+      %% and $\htau_{12}=\tehole$
+      and $\dexp=\dcastthree{\dexp'}{\tehole}{\tehole}{\tehole}$
+      and $\isIndet{\dexp'}$
+      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
+    \item
+      $\htau=\tarr{\tehole}{\tehole}$
+      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+      %% and $\htau_{11}=\tehole$
+      %% and $\htau_{12}=\tehole$
+      and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\tehole}{\tehole}}$
+      %% and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\htau_{11}}{\htau_{12}}}$
+      and $\htau'\neq\htau$
+      and $\isGround{\htau'}$
+      and $\isIndet{\dexp'}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
+    \item
+      $\htau=\tehole$
+      and $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
+      and $\isGround{\htau'}$
+      and $\isIndet{\dexp'}$.
+  \end{enumerate}
 \end{theorem}

### 2018-03-07 intro.tex
draft of intro

--- 2018-02-08 intro.tex
@@
 \label{sec:intro}
-\cy{intro from LIVE 2017 paper is below}
 
-Broadly speaking, live programming environments are those that granularly interleave editing and evaluation \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90,McDirmid:2007:LUL:1297027.1297073,burckhardt2013s}.
-In the words of \citet{burckhardt2013s}, live programming environments 
-``promise to narrow the temporal and perceptive gap 
-between program development and code execution''. Examples of live programming environments include {lab notebook environments},
-e.g. the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
-programmer to interactively edit and evaluate program fragments organized into a
-sequence of cells (an extension of the read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007:LUL:1297027.1297073}, Sketch-n-Sketch \cite{DBLP:conf/pldi/ChughHSA16} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples.
+In a traditional programming environment, the user iteratively edits the program
+text, compiles and runs the program, and views the output.
+%
+To make this workflow more usable and efficient, \emph{live programming
+environments} ``promise to narrow the temporal and perceptive gap between
+program development and code execution''~\cite{burckhardt2013s} with
+fine-grained interleaving of editing and evaluation to varying
+degrees~\cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13}.
 
-The problem that has motivated much of our recent work is that most  
-programming environments, live programming environments included,  provide feedback via various editor services only once the program being edited is syntactically well-formed and, when relevant, well-typed.
-This leaves a ``temporal and perceptive gap'', because programmers sometimes leave a program malformed or ill-typed for extended periods of time, e.g.
-as they think about what to enter at the cursor, or as they work on a different part of the program.
+There have been several promising advances in building live programming
+environments in many settings, such as:
+%
+{lab notebook environments}, like the popular
+IPython/Jupyter~\cite{PER-GRA:2007}, which allow the programmer to interactively
+edit and evaluate program fragments organized into a sequence of cells (an
+extension of the ubiquitous read-eval-print loop (REPL));
+%
+spreadsheets, which provide a reactive dataflow programming model;
+%
+live direct manipulation programming environments like SuperGlue
+\cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist}, and the tools
+demonstrated by Bret Victor in his lectures \cite{victor2012inventing};
+%
+the TouchDevelop live UI framework \cite{burckhardt2013s};
+%
+live mobile application development systems like Flutter \cite{flutter};
+%
+live visual image processing languages~\cite{DBLP:journals/vlc/Tanimoto90};
+%
+and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98},
+to name several prominent examples.
 
-In view of this general problem, we recently developed a \emph{structure editor calculus} called Hazelnut where every edit state consists of a well-formed and statically meaningful, i.e.
-well-typed, incomplete expression, which we take to mean an expression with holes \cite{popl-paper}.
-This calculus addressed fundamental questions relevant to editor services that operate statically, but there was no solution in that paper to the problems faced by editor services that also require knowledge of the dynamic meaning of an incomplete program, as would be relevant to a live programming environment.
-For example, consider a \emph{stepper} (a.k.a.
-a \emph{stepwise debugger}), like that available to Haskell programmers in the GHCi system \cite{GHC-stepper} and other systems \cite{DBLP:conf/haskell/MarlowIPG07,kar13566}, to OCaml programmers with recent work on \texttt{ocamli} by \citet{ocaml-stepper} and to Standard ML programmers \cite{DBLP:journals/jfp/TolmachA95}.
-A stepper requires that the expression being stepped be assigned dynamic meaning according to a small-step operational semantics \cite{DBLP:journals/jlp/Plotkin04a,pfpl}, but no such semantics was defined for incomplete expressions that arise when using Hazelnut, or any of these other systems.
-Defining such a semantics was left as future work in the Hazelnut paper, and in a subsequent ``vision paper'' \cite{snapl17-paper}.
-The purpose of this paper is to sketch out our progress toward a theoretically well-grounded solution.
+Despite these advances, however, there is a fundamental limitation which hinders
+the full realization of live programming: traditional programming languages
+cannot evaluate \emph{incomplete programs}, \ie{}, programs that are either
+syntactically ill-formed or ill-typed.
+%
+Without a well-defined semantics to dictate the treatment of incomplete
+programs, various editor services may be provided in ad-hoc, restricted
+forms---if available, at all---during the periods of time between complete
+program states.
+%
+%% ``flickering in and out''
+%
+%% In other words, the ``Edit-Compile-Run'' cycle routinely fails (either in the
+%% first or second steps) during the program editing process.
+%
+The resulting editor experience includes services ``flickering in and out,''
+thus introducing ``temporal and perceptive gaps'' because programmers often
+leave a program malformed or ill-typed for extended periods of time, \eg{}~as
+they think about what to enter at the cursor, or as they work on a different
+part of the program.
+%
+These times are, arguably, when rich, live editor services would be most helpful
+to the user.
 
-\matt{below are misc points that may be worth mentioning in section 1
-  (somehow, somewhere), and weaving into the continuation of the story
-  above:}
+In view of this general problem---the lack of semantics for \emph{incomplete
+programs}---\citet{popl-paper} recently developed a \emph{structure editor
+calculus} called Hazelnut where every edit state consists of a well-formed and
+statically meaningful (\ie{}~well-typed) incomplete expression, in particular,
+one with \emph{expression holes} or \emph{type holes}.
+%
+That work provides a static semantics for incomplete programs but not a
+\emph{dynamic} semantics, without which certain editor services---such as
+\emph{stepwise debuggers} (\eg{}~\cite{XXX}) and advanced \emph{code completion
+suggestions} (\eg{}~\cite{XXX})---cannot be built.
+%
+The topic of this paper is to develop such a dynamic semantics for functional
+programs with holes.
 
-Type holes vs expression holes:
+%% \parahead{Prior Approaches: Exceptions as Expression Holes}
+\parahead{Prior Approaches}
 
-The dynamics of type holes is gradual typing, including run-time cast
-insertion and execution.
+Given the existing lack of dynamic semantic foundations for incomplete programs,
+what alternative mechanisms do programmers have at their disposal to run and
+debug incomplete programs?
+
+\paragraph{Exceptions as Expression Holes}
 %
-The dynamics of expression holes consists of defining
-\emph{indeterminate} expressions, as placeholders for subterms that
-traditionally would be deemed ``stuck'' .
+One common approach for emulating expression holes is to use exceptions (such as
+\verb+Debug.crash "not yet implemented"+ or \verb+raise NotYetImplementedException+).
+%
+Some languages even provide syntax for holes
+%
+(written \verb+_+ in GHC Haskell~\cite{XXX},
+\verb+?+ in Agda~\cite{XXX}, and
+\verb+{x}+ in Idris~\cite{XXX})
+%
+which, semantically, behave similarly to exceptions.
+%
+A placeholder exception serves to appease the static type checker---because an
+error expression is allowed to have any type required by the surrounding
+context---allowing the program to run, while preserving dynamic type safety by
+simply halting execution as soon as the expression appears in evaluation
+position.
 
-In live programming, we expect that programs will generally have holes
-in both terms and types (CITE?); however, to understand the mechanisms
-of \HazelnutLive, it is instructive to imagine situations where the
-only holes are in terms, or are in types.
+Although useful, this ad-hoc approach has limitations.
 %
-When we only use expression holes and no type holes, we do not need
-type casts, or the other dynamic mechanisms of gradual typing.
+Because the program must halt as soon as a placeholder expression is reached (in
+order to ensure type safety), the programmer cannot benefit from seeing how
+other parts of the program, which do not depend on the missing expression,
+continue to evaluate.
 %
-When we only use type holes, we recover gradual typing and do not need
-indeterminate forms (except for cast errors, which we classify this
-way).
+Another limitation is that exceptions typically carry nothing more than the
+stack trace that led to the error.
+%
+But there are other forms of information that could be useful as the programmer
+works to fill the hole---\eg{}, the dynamic environment around the failed
+expression, and the subsequent computations that depend on the missing value
+(which could only be deduced if evaluation continued).
+%
+Thus, the dynamic semantics of exceptions do not adequately address the
+particular needs for expression holes.
 
-\cy{intro from grant section is below}
+\rkc{(mention something about manually commenting out partial expressions,
+rather than something nicer like marking them inside non-empty holes?)}
 
-Live programming environments granularly interleave editing and evaluation,
-``promis[ing] to narrow the temporal and perceptive gap 
-between program development and code execution''~\cite{burckhardt2013s}.
-Examples of live programming environments include {lab notebook environments},
-like the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
-programmer to interactively edit and evaluate program fragments organized into a
-sequence of cells (an extension of the ubiquitous read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples.
-\cy{find somewhere to cite these tanimoto papers \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90}}
+\paragraph{Gradual Typing}
+%
+A well-developed approach for programming with ``type holes'' is \emph{gradual
+typing}~\cite{XXX,XXX,XXX,XXX}, where the \verb+Any+ type can be assigned to
+expressions either when
+%
+(a)~a valid type assignment is unknown and cannot be inferred, or
+%
+(b)~there is no single valid type assignment.
+%
+To ensure type safety of expressions that have been statically assigned the
+\verb+Any+ type, dynamic casts are attached to monitor how the expression is
+used at operations that require specific types.
+%
+A failed dynamic cast halts execution immediately, much like other kinds of
+run-time exceptions.
+%
+As with ad-hoc expression holes, the dynamics of existing gradually typed
+languages~\cite{XXX,XXX,XXX,XXX} do not allow other portions of the program,
+which do not depend on failed casts, to proceed.
 
-Live programming environments typically cannot run incomplete programs, or they simply halt or raise an exception as soon as they encounter a hole during evaluation.
-This leaves a ``temporal and perceptive gap'' because the results of even those computations that do not depend critically on the hole are not available to the programmer.
-Our second track of research will address this gap by defining a dynamic semantics for incomplete programs that proceeds as far as possible around holes.
-Every program assigned static meaning by the semantics of Track 1 will also be assigned dynamic meaning by the semantics of Track 2.
+\parahead{Our Approach: Evaluating ``Around'' Indeterminate Expressions}
 
-For example, in \autoref{fig:intro-example} we saw that mapping the incomplete function \li{weighted_average} over the \li{grades} list produced an incomplete result.
-If holes simply raised exceptions, then the system would have aborted evaluation as soon as \li{weighted_average} was applied for the first time by \li{map}, and the programmer would not have been able to see that the result was a list of two elements, nor see the value of the sub-expression \li{10.0 *.
-g.hw1} for each element \li{g} of the \li{grades} list.
-Even programs with type inconsistencies can be evaluated under our proposed dynamics, because, as discussed in Track 1, non-empty holes operate as membranes around type inconsistencies.
-Taken together, the results from Tracks 1 and 2 may help to address a common complaint: that a static type discipline makes it difficult to do exploratory programming.
+In contrast to using run-time exceptions, as above, our approach is to treat
+expression and type holes with distinct semantic characteristics.
+%
+In particular, we distinguish a new set of \emph{indeterminate expressions},
+which are neither fully-reduced to values, nor stuck, nor erroneous.
+%
+Instead, indeterminate expressions are ``paused'' during evaluation, at which
+point the programmer can
+%
+(a) inspect the dynamic environment at the point at which evaluation left off,
+%
+(b) inspect how the indeterminate expression is used (abstractly) in subsequent
+calculations, and
+%
+(c) observe the evaluation of other parts of the program that do not depend on
+the value of the indeterminate expression.
+%
+Then, if the programmer makes an edit---to fill in an empty expression hole, to
+replace a non-empty hole with a type-correct expression, or to replace a failed
+cast with a type-correct expressions---the paused evaluation can resume.
+%
+Together, these semantic properties can enable a class of live programming
+services---such as a debugging inspector and a suggestion system for filling
+holes based on dynamic environments---that are not possible with run-time
+exceptions.
 
-Our proposed dynamics goes beyond staged evaluation (\eg{}~\cite{Taha:1999}), partial evaluation (\eg{}~\cite{Jones:1993uq}) and symbolic evaluation (\eg{}~\cite{King:1976}) in that we track the dynamic environment (\ie{}~the substitutions performed) around each instance of an expression hole in the result.
-This allows the programmer to see, in the sidebar of \autoref{fig:intro-example}, the actual values that the variables in scope take on everywhere that the hole they are filling ends up during evaluation.
-This provides a hole-oriented specification of a standard feature of debuggers: inspection of the environment at designated points in the program.
-Finally, this environmental information is useful (though not required) for the edit action suggestion service of Track 3 and the direct manipulation programming service of Track 4.
+\parahead{Contributions and Paper Outline}
+
+\newcommand{\contribution}[2]{\paragraph{#1. #2}}
+
+This paper makes the following contributions:
+
+%% \begin{itemize}
+
+%% \item
+%
+\contribution{1}{Dynamic semantics with indeterminate expressions}
+%
+We define a dynamic semantics for functional programs with holes, by
+distinguishing \emph{indeterminate expressions} from those which are fully
+reduced to values and those which are stuck.
+%
+The semantics proceeds by evaluating around indeterminate expressions, allowing
+programs to continue to produce meaningful results despite dynamic cast errors
+(from type holes) or missing expressions (expression holes).
+%
+Formally, our semantics combines features from gradual type theory~\cite{XXX}
+(to handle incomplete types) and contextual modal type theory~\cite{XXX} (CMTT,
+which provides a logical foundation for hole environment tracking).
+%
+With a formal development in Agda, we prove that our calculus enjoys standard
+type safety properties, extended to account for indeterminate expressions.
+%
+(\autoref{sec:calculus})
+
+%% \item
+%
+\contribution{2}{Resumption of suspended evaluation}
+%
+We describe how our semantics allows the evaluation of a program with
+indeterminate expressions to \emph{resume} evaluation, as long as the only edits
+to the program involve filling expression holes.
+%
+This ``edit-and-resume'' feature may be a useful optimization for large,
+long-running programs in a development workflow that makes ``forward progress''
+through filling holes.
+%
+We provide a proof sketch of commutativity and confluence properties that ensure
+the correctness of this feature.
+%
+Our Adga formalization codifies this proof sketch, which assumes several
+unproven lemmas that we expect to be similar to those used to prove confluence
+in standard lambda-calculi~\cite{XXX}.
+%
+(\autoref{sec:resumption})
+
+%% \item
+%
+\contribution{3}{Prototype implementation}
+%
+We extend the Hazelnut system of \citet{popl-paper}, which defined an edit
+action calculus and static semantics for incomplete programs, with our new
+dynamic semantics.
+%
+The resulting tool, \HazelnutLive{}, demonstrates a proof-of-concept where every
+edit results in a program state that can be statically and dynamically analyzed
+in order to provide meaningful feedback to the programmer.
+%
+\HazelnutLive{} can be used to and run the examples in this paper, modulo
+several standard syntactic conveniences that our current implementation does not
+provide.
+%
+(\autoref{sec:implementation})
+
+%% \end{itemize}
+
+\vspace{5pt}
+
+Thus, our work provides an additional step towards the ultimate goal for live
+programming environments to reduce the temporal and perceptive gap between
+program development and code execution.
+%
+Our implementation, examples, and some videos are available in \suppMaterials{}.
+%
+Next, in \autoref{sec:examples}, we describe an overview example to introduce
+the main features in \HazelnutLive{}, before describing each of the above
+contributions in detail.
+%
+We wrap up in \autoref{sec:relatedWork} and \autoref{sec:discussion} with
+discussions of related and future work.
+
+\rkc{Cyrus/Matt: I moved all the ``raw'' material that was here into the
+Appendix. Please take a look there if there's anything else that you'd like to
+include in the Intro.}

### 2018-03-07 misc.tex
draft of intro

\clearpage
\section{Misc}

\subsection{Misc Raw Material for Intro}

\cy{intro from LIVE 2017 paper is below}

Broadly speaking, live programming environments are those that granularly interleave editing and evaluation \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90,McDirmid:2007:LUL:1297027.1297073,burckhardt2013s}. 
In the words of \citet{burckhardt2013s}, live programming environments 
``promise to narrow the temporal and perceptive gap 
between program development and code execution''. Examples of live programming environments include {lab notebook environments},
e.g. the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
programmer to interactively edit and evaluate program fragments organized into a
sequence of cells (an extension of the read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007:LUL:1297027.1297073}, Sketch-n-Sketch \cite{DBLP:conf/pldi/ChughHSA16} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples.

The problem that has motivated much of our recent work is that most  
programming environments, live programming environments included,  provide feedback via various editor services only once the program being edited is syntactically well-formed and, when relevant, well-typed. This leaves a ``temporal and perceptive gap'', because programmers sometimes leave a program malformed or ill-typed for extended periods of time, e.g. as they think about what to enter at the cursor, or as they work on a different part of the program.

In view of this general problem, we recently developed a \emph{structure editor calculus} called Hazelnut where every edit state consists of a well-formed and statically meaningful, i.e. well-typed, incomplete expression, which we take to mean an expression with holes \cite{popl-paper}. This calculus addressed fundamental questions relevant to editor services that operate statically, but there was no solution in that paper to the problems faced by editor services that also require knowledge of the dynamic meaning of an incomplete program, as would be relevant to a live programming environment. For example, consider a \emph{stepper} (a.k.a. a \emph{stepwise debugger}), like that available to Haskell programmers in the GHCi system \cite{GHC-stepper} and other systems \cite{DBLP:conf/haskell/MarlowIPG07,kar13566}, to OCaml programmers with recent work on \texttt{ocamli} by \citet{ocaml-stepper} and to Standard ML programmers \cite{DBLP:journals/jfp/TolmachA95}. A stepper requires that the expression being stepped be assigned dynamic meaning according to a small-step operational semantics \cite{DBLP:journals/jlp/Plotkin04a,pfpl}, but no such semantics was defined for incomplete expressions that arise when using Hazelnut, or any of these other systems. Defining such a semantics was left as future work in the Hazelnut paper, and in a subsequent ``vision paper'' \cite{snapl17-paper}. The purpose of this paper is to sketch out our progress toward a theoretically well-grounded solution.

\matt{below are misc points that may be worth mentioning in section 1
  (somehow, somewhere), and weaving into the continuation of the story
  above:}

\textbf{Type holes vs expression holes:}

The dynamics of type holes is gradual typing, including run-time cast
insertion and execution.
%
The dynamics of expression holes consists of defining
\emph{indeterminate} expressions, as placeholders for subterms that
traditionally would be deemed ``stuck'' .

In live programming, we expect that programs will generally have holes
in both terms and types (CITE?); however, to understand the mechanisms
of \HazelnutLive, it is instructive to imagine situations where the
only holes are in terms, or are in types.
%
When we only use expression holes and no type holes, we do not need
type casts, or the other dynamic mechanisms of gradual typing.
% 
When we only use type holes, we recover gradual typing and do not use
indeterminate forms until and unless cast errors arise.

\textbf{Cast errors as indeterminate forms:}

Instead of being ``stuck'', cast errors in \HazelnutLive are
fully-reduced indeterminate terms.

\textbf{Indeterminate forms as exceptions:}

TODO --- transfer thoughts from 2016 grant writing

\textbf{Holes as breakpoints:}

\begin{enumerate}

\item 
  \texttt{map somefun list}

Suppose the programmer wishes to map with \texttt{somefun} that they
either did not author themselves, or they forgot how the function
works:

\item 
  \textt{map ($\lambda \texttt{x}. \hhole{\texttt{x}}{a}$) list}

First, the programmer can intercept the function's inputs.  These
hole~$a$ above functions like a ``breakpoint'', and shows the
programmer me all of the inputs to \texttt{somefun}.

TODO -- show the result of running above.

\item
  \textt{map ($\lambda \texttt{x}. \hhole{\texttt{somefun x}}{a}$) list}

  Next, suppose the programmer wishes to see all of the the outputs of
  \texttt{somefun}.  To do so, they put that (single static)
  expression into a hole, as above.

\item
  \textt{map ($\lambda \texttt{x}. \hhole{(x, \texttt{somefun x})}{a}$) list}

  Finally, the programmer can see input-output pairs without breaking
  the typing derivation of their program globally, since holes are
  permitted to go \emph{anywhere}.  (They may have also exploited this
  fact to show the inputs, in the first step).

\end{enumerate}

\cy{intro from grant section is below}

Live programming environments granularly interleave editing and evaluation,
``promis[ing] to narrow the temporal and perceptive gap 
between program development and code execution''~\cite{burckhardt2013s}.
Examples of live programming environments include {lab notebook environments},
like the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
programmer to interactively edit and evaluate program fragments organized into a
sequence of cells (an extension of the ubiquitous read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples. \cy{find somewhere to cite these tanimoto papers \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90}}

Live programming environments typically cannot run incomplete programs, or they simply halt or raise an exception as soon as they encounter a hole during evaluation. This leaves a ``temporal and perceptive gap'' because the results of even those computations that do not depend critically on the hole are not available to the programmer. Our second track of research will address this gap by defining a dynamic semantics for incomplete programs that proceeds as far as possible around holes. 
Every program assigned static meaning by the semantics of Track 1 will also be assigned dynamic meaning by the semantics of Track 2.

For example, in \autoref{fig:intro-example} we saw that mapping the incomplete function \li{weighted_average} over the \li{grades} list produced an incomplete result. If holes simply raised exceptions, then the system would have aborted evaluation as soon as \li{weighted_average} was applied for the first time by \li{map}, and the programmer would not have been able to see that the result was a list of two elements, nor see the value of the sub-expression \li{10.0 *. g.hw1} for each element \li{g} of the \li{grades} list. Even programs with type inconsistencies can be evaluated under our proposed dynamics, because, as discussed in Track 1, non-empty holes operate as membranes around type inconsistencies. Taken together, the results from Tracks 1 and 2 may help to address a common complaint: that a static type discipline makes it difficult to do exploratory programming.

Our proposed dynamics goes beyond staged evaluation (\eg{}~\cite{Taha:1999}), partial evaluation (\eg{}~\cite{Jones:1993uq}) and symbolic evaluation (\eg{}~\cite{King:1976}) in that we track the dynamic environment (\ie{}~the substitutions performed) around each instance of an expression hole in the result. This allows the programmer to see, in the sidebar of \autoref{fig:intro-example}, the actual values that the variables in scope take on everywhere that the hole they are filling ends up during evaluation. This provides a hole-oriented specification of a standard feature of debuggers: inspection of the environment at designated points in the program. Finally, this environmental information is useful (though not required) for the edit action suggestion service of Track 3 and the direct manipulation programming service of Track 4.

### 2018-03-07 calculus.tex
split metatheory into two sections
calculus edit

@@
 
+Talk about weak vs strong conversion vs sharing -- see \url{http://gallium.inria.fr/~maranget/papers/jwk60.pdf} and cited work in the intro.
+See also \url{https://www.lri.fr/~blsk/Docs/Balabonski-FullLaziness-POPL12.pdf}??
+Original Church-Rosser paper -- \url{http://www.ams.org/journals/tran/1936-039-03/S0002-9947-1936-1501858-0/S0002-9947-1936-1501858-0.pdf}.
+
 \begin{comment}

### 2018-03-07 discussion.tex
note in discussion
Section 2, first full rough draft

@@
 \label{sec:discussion}
+
+Weinberg from The Psychology of Computer Programming (1998): ``how truly sad it is that just at the very moment
+when the computer has something important to tell us, it starts
+speaking gibberish.''
 
@@
 
+\rkc{something about visualizing hole environment traces}
+
 \rkc{\citet{Seidel2016} define a partial evaluation with a form of holes. compare.
@@
 trace help students focus on the part that has the error.}
+
+\rkc{quick mention that lazy evaluation with run-time exceptions for holes is
+not enough. this would delay the failure a bit longer, but not indefinitely.}

### 2018-03-07 calculus.tex
"Substitution Typing"

--- 2018-02-28 calculus.tex
@@
 %% \begin{figure}[!ht]
-  \begin{definition}
+  \begin{definition}[Substitution Typing]
     $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Gamma'}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
@@
 
-\subsection{Metatheory: Correctness of Edit-and-Resume}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Metatheory: Correctness of Edit-and-Resume}
+\label{sec:resumption}
 

### 2018-03-08 examples-1.tex
splitting Section 2 subsections into separate files
to help avoid merge conflicts

--- 2018-02-20 examples.tex
@@
-% !TEX root = hazelnut-dynamics.tex
-%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
-%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
-\newcommand{\examplesSec}{\HazelnutLive, By Example}
-\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
-\label{sec:examples}
-
-\newcommand{\overviewExample}[2]{\paragraph{Example {#1}: {#2}}}
-
-We introduce \HazelnutLive{} in this section using several examples,
-demonstrating how running incomplete programs---with missing expressions and/or
-type-inconsistent expressions---can be useful.
-%
-We will also demonstrate how running even complete programs with subexpressions
-marked as holes may be useful in debugging.
-
-\HazelnutLive{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
-alternative Reason syntax \cite{reason-what} extended to support holes.
-%
-As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
-literate program \cite{knuth1984literate}: rich text comments appear at the top
-level and code appears within cells embedded into these comments.
-%
-The reader is encouraged to work through these examples in our prototype
-implementation or watch video demonstrations, both available in
-\suppMaterials{}.\footnote{
-%
-The programs in this section are written using minor syntactic conveniences that
-are not currently supported in our implementation.
-%
-For example, the implementation of recursive types in our current implementation
-requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
-inferred in standard ways (by pairing them with constructors and deconstructors
-for algebraic datatypes) in future work.
-%
-More details about other minor syntactic differences are described in
-\suppMaterials{}.
-%
-}
-
 \subsection{Live Programming with Incomplete Expressions}
@@
 going back and forth between different pieces under development.
-
-%% \subsection{Live Programming with Type Errors}
-\subsection{Live Programming with Static Type Errors}
-
-In the previous section, we described example programs that were incomplete
-because of \emph{missing} expressions.
-%
-Next, we describe example programs that are incomplete because of
-\emph{type-inconsistent} expressions.
-%
-For example, consider the following two definitions.
-
-\begin{lstlisting}
-let bad_bool : bool = ?? 0 ??_bad_bool;
-
-let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
-\end{lstlisting}
-
-\noindent
-%
-These two definitions are ill-typed under standard typing disciplines.
-%
-In contrast, \citet{popl-paper} present a bidirectional type system that assigns
-types to both, by wrapping type-inconsistent expressions (\li{0} on line
-\rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
-%
-Non-empty holes prevent local type inconsistencies from polluting the rest of
-the program surrounding it, which may or may not itself contain additional
-inconsistencies.
-
-Understanding and debugging static type errors is notoriously difficult,
-particularly for novices.
-%
-A variety of approaches have been
-proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
-and explain type errors.
-%
-One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
-that demonstrates a run-time failure, and then displays a compressed execution
-trace to the user as a graph.
-%
-In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
-non-empty holes) as far as possible, in particular, until they
-would go wrong.
-%
-Thus, our operational semantics provides a similar benefit to the approach of
-\citet{Seidel2016}.
-
-\overviewExample{3}{Sum List}
-%
-Consider the following buggy program (observed during an undergraduate
-functional programming course~\cite{Seidel2016}) that attempts to sum a list
-integers.
-%
-The error is that the base case produces a list rather than an integer.
-
-\begin{lstlisting}
-sumList : list(int) -> int
-sumList [] = ?[]?
-sumList (n:ns) = n + sumList ns
-\end{lstlisting}
-
-\noindent
-%
-Because the list expression on line 2 does not have type \li{int} as required,
-it is wrapped in a (non-empty) hole by the bidirectional type
-checker~\cite{popl-paper}.
-%
-Rather than trying to debug the error based on the static error, the programmer
-may wish to trying running the function anyway by calling, say, \li{sumList(2)}.
-%
-\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
-%
-By observing that the hole expression is being added to the integer \li{3}, he
-realizes that it needs to be an integer, specifically, \li{0}.
-%
-Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
-produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
-successfully proceeded to evaluate despite the error elsewhere.
-
-%% TODO fold error from Erwig paper.
-%% %
-%% see that final call on stack does have the right answer, but
-%% it's wrapped in a singleton list when the expected type is not
-%% a list.
-%% %
-%% fix is to remove the list, the rest of the computation remains
-%% the same, but b/c they were all wrapped in holes, need to re-run.
-%% %
-%% (add some mechanism for type-consistent non-empty-holes...)
-
-\overviewExample{4}{Stutter}
-%
-Consider the following function which attempts to produce a
-list where every element is repeated twice (borrowed from \citet{Osera2015}).
-%
-The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
-it produces a \li{list(list(int)} instead.
-
-\begin{lstlisting}
-stutter : list(int) -> list(int)
-stutter xs = List.foldr (\x acc -> ?[x,x]? : acc) [] xs
-\end{lstlisting}
-
-\noindent
-%
-The bidirectional type checker of \citet{popl-paper} wraps the expression
-\li{[x,x]} inside a non-empty hole.
-%
-%% The editor has a choice about which expression to ``blame'' for the error; the
-%% entire application that forms the body of the lambda is analyzed against the
-%% return type \li{list(int)}, so that is a reasonable choice for the editor to
-%% make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
-%% both as intended and that only the function \li{(:)} is type-consistent.
-%% %
-%% Although one could imagine a setting in which a user would perform this
-%% reasoning, let's assume the simplest approach for marking consistencies that
-%% wraps the entire application.
-%
-Running this on \li{stutter [1,2,3]} produces the indeterminate result
-
-\begin{lstlisting}
-?  [1,1] : (? [2,2] : (? [[3:3]] ?) ?) ?,
-\end{lstlisting}
-
-\noindent
-%
-which shows the unfolding of \li{List.foldr}.
-%
-We refer to nested indeterminate computations like this as \rkc{\emph{hole
-environment traces} or \emph{hole environment trees}}.
-%
-The result of the innermost indeterminate expression is \li{[[3,3]]}.
-%
-The user realizes that there are too many levels of nesting, so
-he replaces the \li{(:)} with \li{(@)}, which addresses the type inconsistency
-and, when reevaluated, produces the desired result.
-
-\rkc{STOPPING HERE FOR NOW}
-
-\subsection{Live Programming with Gradual Type Errors}
-
-\overviewExample{5}{\rkc{...}}
-
-\rkc{gradual typing error, cast error, run around}
-
-\begin{lstlisting}
-foo : bool -> ? -> int
-foo b =
-  if b then
-    x + 10
-  else
-    x 10
-
-foo false 1 + 2 + 2
-\end{lstlisting}
-
-this demonstrates how, when no expression holes, our system is like gradual
-typing, but with the ability to run around cast errors.
-%
-just as it may be useful to report multiple type errors, this allows ability to
-report multiple dynamic errors, as well additional dynamic behavior around them.
-%
-could be incorporated into other gradual type systems
-
-\subsection{Live Programming for Debugging}
-
-\overviewExample{6}{\rkc{...}}
-
-\rkc{breakpoints, understanding input/output behavior}
-
-\parasection{Example 6: Quicksort}
-
-\begin{lstlisting}
-quicksort : list('a) -> list('a)
-quicksort [] = []
-quicksort (x:xs) =
-  let (left, right) = span ((>) x) xs in
-  ??
-\end{lstlisting}
-
-programmer has some intended relationship between names and values they bind.
-%
-live results show that left has larger elements and right has smaller.
-%
-right away, can see that need to change filter predicate. (alternatively, the
-names could have been swapped, but that order of names was the natural way
-to think about it.)
-%
-but next, still see that right has some smaller elements.
-%
-should pick partition instead of span.
-
-\begin{lstlisting}
-quicksort : list('a) -> list('a)
-quicksort [] = []
-quicksort (x:xs) =
-  let (left, right) = partition ((<) x) xs in
-  let (left', right') = (quicksort left, quicksort right) in
-  ? (left' @ [x] @ right) ?
-\end{lstlisting}
-
-now, put the lists back together.
-%
-doesn't do what is expected.
-%
-live feedback shows right' is computed correctly, but ah, it's not being used
-in the return list.
-%
-fix, and re-run, keeping the final expression in a hole.
-%
-can visualize that the nested appends are being built up correctly, and that the
-result would be the correct list.
-%
-finalize the hole and resume. everything but the nested appends can be reused.
-%
-\emph{hole environment traces} or \emph{hole environment traces}
-
-\paragraph{Recap}
-%
-In the above four subsections, respectively, we considered programs with:
-%
-(1) empty expression holes,
-%
-(2) non-empty expression holes for ill-typed expressions,
-%
-(3) type holes, and
-%
-(4) non-empty expression holes for well-typed expressions.
-%
-Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
-that allows programs with combinations of these kinds of incompleteness to be
-evaluated.

### 2018-03-08 examples-2.tex
splitting Section 2 subsections into separate files
to help avoid merge conflicts

--- 2018-02-20 examples.tex
@@
-% !TEX root = hazelnut-dynamics.tex
-%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
-%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
-\newcommand{\examplesSec}{\HazelnutLive, By Example}
-\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
-\label{sec:examples}
-
-\newcommand{\overviewExample}[2]{\paragraph{Example {#1}: {#2}}}
-
-We introduce \HazelnutLive{} in this section using several examples,
-demonstrating how running incomplete programs---with missing expressions and/or
-type-inconsistent expressions---can be useful.
-%
-We will also demonstrate how running even complete programs with subexpressions
-marked as holes may be useful in debugging.
-
-\HazelnutLive{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
-alternative Reason syntax \cite{reason-what} extended to support holes.
-%
-As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
-literate program \cite{knuth1984literate}: rich text comments appear at the top
-level and code appears within cells embedded into these comments.
-%
-The reader is encouraged to work through these examples in our prototype
-implementation or watch video demonstrations, both available in
-\suppMaterials{}.\footnote{
-%
-The programs in this section are written using minor syntactic conveniences that
-are not currently supported in our implementation.
-%
-For example, the implementation of recursive types in our current implementation
-requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
-inferred in standard ways (by pairing them with constructors and deconstructors
-for algebraic datatypes) in future work.
-%
-More details about other minor syntactic differences are described in
-\suppMaterials{}.
-%
-}
-
-\subsection{Live Programming with Incomplete Expressions}
-
-The process of writing new program fragments includes many states in which the
-program under construction is, by definition, incomplete.
-%
-It is natural for a programmer to ``jump back and forth'' between different
-places in the program, iteratively and alternately developing the missing
-pieces.
-%
-The primary benefit of our approach is that such programs can be evaluated in
-order to provide useful feedback during this editing workflow.
-
-\input{fig-grades-example}
-
-Consider a teacher who is in the midst of developing a \HazelnutLive{} notebook
-(depicted in \autoref{fig:grades-example}) to compute final student grades at
-the end of a term.
-%
-In the first cell in \autoref{fig:grades-example}, the teacher defines a record
-type, \li{student_data}, for each student's course data.
-%
-In our simplified example, this includes the student's name, of type
-\li{string}, and five grades, each of type \li{float}.
-
-\overviewExample{1}{Computing Weighted Averages}
-%
-The teacher begins writing a \li{weighted_average} function that will
-compute a weighted average, of type \li{float}, for each \li{student_data}
-record.
-%
-The body of \li{weighted_average} is not yet written---marked with the
-\emph{empty hole} on line \rkc{XXX}--- but the teacher jumps ahead to map the
-function over the \li{grades} list.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  ??;
-
-let weighted_averages = List.map weighted_average grades;
-\end{lstlisting}
-
-\noindent
-%
-\HazelnutLive{} evaluates this incomplete program, showing its
-\emph{indeterminate} result in the sidebar on the right half of
-\autoref{fig:grades-example}; this value is a list whose length is the same as
-\li{grades}, where each element is the (indeterminate) result of evaluating the
-hole expression inside \li{weighted_average}.
-%
-Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
-(\ie{}~\li{a}) and that the indeterminate results it produces are assigned
-corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2}, etc.).
-%
-In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
-using placeholder expressions like \li{raise "Not yet implemented."}, would halt
-the program as soon it is evaluated (in this case, inside the definition of
-\li{List.map} when the function is applied to the first element of \li{grades}.
-%
-By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
-least see that the resulting expression does indeed evaluate to a list with the
-correct number of (indeterminate) values.
-
-Now the teacher returns to \li{weighted_average} to work on the missing
-expression, with the goal of computing a weighted sum of each of a student's
-grades.
-%
-The teacher uses a \HazelnutLive{} edit action to create a sum expression,
-filling in the first summand by weighting the \li{hw1} score to account for 10\%
-of the weighted average; the rest of the computation is a new hole expression
-(\li{??_b}), yet to be filled in.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  (10.0 *. g.hw1) +. ??_b;
-\end{lstlisting}
-
-\noindent
-%
-\HazelnutLive{} immediately runs this incomplete program, displaying the updated
-list of indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
-%
-Right away, the teacher recognizes that the values are too large; they should be
-at most \li{100.0}.
-%
-The teacher realizes that representing percentage points as \li{float}s requires
-that the constant on line \rkc{XXX} ought to be \li{0.10} instead.
-%
-Because of this live feedback, the teacher corrects this error right away and
-avoids making similar programming errors in the rest of the calculation.
-%
-The teacher continues to build the rest of the arithmetic expression until it is
-complete (there are no longer any expression holes), and the result of
-immediately running the finished program shows that each of values in the final
-result list is in the range \li{0.0} to \li{100.0}.
-
-\overviewExample{2}{Assigning Letter Grades}
-%
-The teacher's next task is to map the weighted averages to the letter grades A
-through F (we will consider only ``whole'' letter grades, for simplicity).
-%
-The \li{grade_cutoffs} record type describes the minimum cutoff for each of
-these six possible grades.
-%
-Initially, each value in the \li{cutoffs} record is a hole.
-
-\begin{lstlisting}
-type grade_cutoffs =
-  { a: float, b: float, c: float, d: float, f: float };
-
-let cutoffs =
-  { a = ??_a, b = ??_b, c = ??_c, d = ??_d, f = ??_f };
-\end{lstlisting}
-
-\noindent
-%
-%% initially all holes, because it will be different year-to-year based on the %
-%data, differences in course difficulty, and to satisfy fairness criteria.
-%
-Before starting to fill in the cutoffs, the teacher jumps ahead to write a
-function \li{letter_grade} that will make the connection between \li{cutoffs}
-and \li{weighted_averages}.
-%
-Because she intends to look at the data to help select the cutoff values, the
-teacher sorts the \li{weighted_averages} and then maps them to
-\li{letter_grades}.
-
-\begin{lstlisting}
-let letter_grade(n: weighted_average) =
-  if n >= cutoffs.a then "A" else
-  if n >= cutoffs.b then "B" else
-  if n >= cutoffs.c then "C" else
-  if n >= cutoffs.d then "D" else
-  if n >= cutoffs.f then "F" else "Incomplete";
-
-let sorted_weighted_averages = List.sort weighted_averages;
-
-let letter_grades = List.map letter_grade sorted_weighted_averages;
-\end{lstlisting}
-
-\noindent
-%
-When \HazelnutLive{} runs this program, the guard of the outermost conditional
-(\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
-is.
-%
-Therefore, each of the indeterminate expressions in \li{letter_grades} is the
-entire expression body, albeit with different bindings for \li{n}.
-%
-Displaying such ``large'' indeterminate expressions can quickly consume all
-available screen space, overwhelming the user with too much information.
-%
-\autoref{fig:XXX} shows how \HazelnutLive{} renders large indeterminate
-expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
-or hole) or list of small values) simply as \li{..} to save space.
-%
-Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
-indeterminate expression---as well as the evaluation environment that surrounds
-it---as a tooltip.
-%
-\autoref{sec:discussion} discusses this and other user interface concerns when
-trying to display useful live feedback without overwhelming the user.
-%
-These usability factors are beyond the scope of our work, which is to define
-semantic foundations on which such user interfaces can be built.
-
-To start deciding \li{cutoffs}, the teacher looks at the \li{weighted_averages}
-sorted in descending order.
-%
-The data shows a natural gap between \li{92.2} and \li{89.5}.
-%
-So, she chooses to use \li{92.0} as the cutoff for A.
-%
-Resuming the computation from before, \HazelnutLive{} resolves the conditional
-expressions for the first \rkc{XXX} indeterminate expressions, because each of
-those \li{n} values was greater than \li{cutoffs.a}.
-%
-The remaining \rkc{XXX} expressions also proceeded to evaluate the first guard,
-and are now indeterminate at the guard for the second conditional.
-
-Before assigning other cutoff values, the teacher would like to get a sense of
-whether this first choice is a good one.
-%
-She jumps ahead to to write a function that computes the distribution of letter
-grades.
-
-\begin{lstlisting}
-let compute_distribution(list: list(float)) =
-  let n = List.length letter_grades in
-  List.map
-    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list)) /. n))
-    ["A","B","C","D","F","Incomplete"];
-
-let distribution = compute_distribution(letter_grades);
-\end{lstlisting}
-
-\noindent
-%
-Running \li{compute_distribution} shows that the percentage of As is
-\rkc{XXX\%}, which is smaller than what the teacher would like;
-the remaining percentages are all indeterminate.
-%
-Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
-and then another gap between \li{88.2} and {85.5}.
-%
-So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
-%
-Because this edit is outside of a hole expression, \HazelnutLive{} discards the
-previous execution state and reevaluates the entire program.
-%
-Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
-be applied to seek opportunities for reuse even when non-hole expressions are
-modified.
-%
-Because the focus of work is on the novel implications of running programs with
-holes, our calculus and implementation supports caching ``edit-and-resume'' only
-for the novel situation in which evaluation proceeds around holes.
-%
-After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
-matches the teacher's intention.
-
-In this fashion, the teacher continues down the list of sorted averages,
-determining appropriate values for each cutoff.
-%
-Whenever the teacher is only filling in the ``next'' cutoff, the computation
-from before can simply be resumed.
-%
-Overall, throughout the workflow described in these two examples, the programmer
-can continue to evaluate the program, and receive meaningful feedback, while
-going back and forth between different pieces under development.
-
 %% \subsection{Live Programming with Type Errors}
@@
 and, when reevaluated, produces the desired result.
-
-\rkc{STOPPING HERE FOR NOW}
-
-\subsection{Live Programming with Gradual Type Errors}
-
-\overviewExample{5}{\rkc{...}}
-
-\rkc{gradual typing error, cast error, run around}
-
-\begin{lstlisting}
-foo : bool -> ? -> int
-foo b =
-  if b then
-    x + 10
-  else
-    x 10
-
-foo false 1 + 2 + 2
-\end{lstlisting}
-
-this demonstrates how, when no expression holes, our system is like gradual
-typing, but with the ability to run around cast errors.
-%
-just as it may be useful to report multiple type errors, this allows ability to
-report multiple dynamic errors, as well additional dynamic behavior around them.
-%
-could be incorporated into other gradual type systems
-
-\subsection{Live Programming for Debugging}
-
-\overviewExample{6}{\rkc{...}}
-
-\rkc{breakpoints, understanding input/output behavior}
-
-\parasection{Example 6: Quicksort}
-
-\begin{lstlisting}
-quicksort : list('a) -> list('a)
-quicksort [] = []
-quicksort (x:xs) =
-  let (left, right) = span ((>) x) xs in
-  ??
-\end{lstlisting}
-
-programmer has some intended relationship between names and values they bind.
-%
-live results show that left has larger elements and right has smaller.
-%
-right away, can see that need to change filter predicate. (alternatively, the
-names could have been swapped, but that order of names was the natural way
-to think about it.)
-%
-but next, still see that right has some smaller elements.
-%
-should pick partition instead of span.
-
-\begin{lstlisting}
-quicksort : list('a) -> list('a)
-quicksort [] = []
-quicksort (x:xs) =
-  let (left, right) = partition ((<) x) xs in
-  let (left', right') = (quicksort left, quicksort right) in
-  ? (left' @ [x] @ right) ?
-\end{lstlisting}
-
-now, put the lists back together.
-%
-doesn't do what is expected.
-%
-live feedback shows right' is computed correctly, but ah, it's not being used
-in the return list.
-%
-fix, and re-run, keeping the final expression in a hole.
-%
-can visualize that the nested appends are being built up correctly, and that the
-result would be the correct list.
-%
-finalize the hole and resume. everything but the nested appends can be reused.
-%
-\emph{hole environment traces} or \emph{hole environment traces}
-
-\paragraph{Recap}
-%
-In the above four subsections, respectively, we considered programs with:
-%
-(1) empty expression holes,
-%
-(2) non-empty expression holes for ill-typed expressions,
-%
-(3) type holes, and
-%
-(4) non-empty expression holes for well-typed expressions.
-%
-Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
-that allows programs with combinations of these kinds of incompleteness to be
-evaluated.

### 2018-03-08 examples-3.tex
splitting Section 2 subsections into separate files
to help avoid merge conflicts

--- 2018-02-21 examples.tex
@@
-% !TEX root = hazelnut-dynamics.tex
-%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
-%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
-\newcommand{\examplesSec}{\HazelnutLive, By Example}
-\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
-\label{sec:examples}
-
-\newcommand{\overviewExample}[2]{\paragraph{Example {#1}: {#2}}}
-
-We introduce \HazelnutLive{} in this section using several examples,
-demonstrating how running incomplete programs---with missing expressions and/or
-type-inconsistent expressions---can be useful.
-%
-We will also demonstrate how running even complete programs with subexpressions
-marked as holes may be useful in debugging.
-
-\HazelnutLive{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
-alternative Reason syntax \cite{reason-what} extended to support holes.
-%
-As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
-literate program \cite{knuth1984literate}: rich text comments appear at the top
-level and code appears within cells embedded into these comments.
-%
-The reader is encouraged to work through these examples in our prototype
-implementation or watch video demonstrations, both available in
-\suppMaterials{}.\footnote{
-%
-The programs in this section are written using minor syntactic conveniences that
-are not currently supported in our implementation.
-%
-For example, the implementation of recursive types in our current implementation
-requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
-inferred in standard ways (by pairing them with constructors and deconstructors
-for algebraic datatypes) in future work.
-%
-More details about other minor syntactic differences are described in
-\suppMaterials{}.
-%
-}
-
-\subsection{Live Programming with Incomplete Expressions}
-
-The process of writing new program fragments includes many states in which the
-program under construction is, by definition, incomplete.
-%
-It is natural for a programmer to ``jump back and forth'' between different
-places in the program, iteratively and alternately developing the missing
-pieces.
-%
-The primary benefit of our approach is that such programs can be evaluated in
-order to provide useful feedback during this editing workflow.
-
-\input{fig-grades-example}
-
-Consider a teacher who is in the midst of developing a \HazelnutLive{} notebook
-(depicted in \autoref{fig:grades-example}) to compute final student grades at
-the end of a term.
-%
-In the first cell in \autoref{fig:grades-example}, the teacher defines a record
-type, \li{student_data}, for each student's course data.
-%
-In our simplified example, this includes the student's name, of type
-\li{string}, and five grades, each of type \li{float}.
-
-\overviewExample{1}{Computing Weighted Averages}
-%
-The teacher begins writing a \li{weighted_average} function that will
-compute a weighted average, of type \li{float}, for each \li{student_data}
-record.
-%
-The body of \li{weighted_average} is not yet written---marked with the
-\emph{empty hole} on line \rkc{XXX}--- but the teacher jumps ahead to map the
-function over the \li{grades} list.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  ??;
-
-let weighted_averages = List.map weighted_average grades;
-\end{lstlisting}
-
-\noindent
-%
-\HazelnutLive{} evaluates this incomplete program, showing its
-\emph{indeterminate} result in the sidebar on the right half of
-\autoref{fig:grades-example}; this value is a list whose length is the same as
-\li{grades}, where each element is the (indeterminate) result of evaluating the
-hole expression inside \li{weighted_average}.
-%
-Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
-(\ie{}~\li{a}) and that the indeterminate results it produces are assigned
-corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2}, etc.).
-%
-In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
-using placeholder expressions like \li{raise "Not yet implemented."}, would halt
-the program as soon it is evaluated (in this case, inside the definition of
-\li{List.map} when the function is applied to the first element of \li{grades}.
-%
-By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
-least see that the resulting expression does indeed evaluate to a list with the
-correct number of (indeterminate) values.
-
-Now the teacher returns to \li{weighted_average} to work on the missing
-expression, with the goal of computing a weighted sum of each of a student's
-grades.
-%
-The teacher uses a \HazelnutLive{} edit action to create a sum expression,
-filling in the first summand by weighting the \li{hw1} score to account for 10\%
-of the weighted average; the rest of the computation is a new hole expression
-(\li{??_b}), yet to be filled in.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  (10.0 *. g.hw1) +. ??_b;
-\end{lstlisting}
-
-\noindent
-%
-\HazelnutLive{} immediately runs this incomplete program, displaying the updated
-list of indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
-%
-Right away, the teacher recognizes that the values are too large; they should be
-at most \li{100.0}.
-%
-The teacher realizes that representing percentage points as \li{float}s requires
-that the constant on line \rkc{XXX} ought to be \li{0.10} instead.
-%
-Because of this live feedback, the teacher corrects this error right away and
-avoids making similar programming errors in the rest of the calculation.
-%
-The teacher continues to build the rest of the arithmetic expression until it is
-complete (there are no longer any expression holes), and the result of
-immediately running the finished program shows that each of values in the final
-result list is in the range \li{0.0} to \li{100.0}.
-
-\overviewExample{2}{Assigning Letter Grades}
-%
-The teacher's next task is to map the weighted averages to the letter grades A
-through F (we will consider only ``whole'' letter grades, for simplicity).
-%
-The \li{grade_cutoffs} record type describes the minimum cutoff for each of
-these six possible grades.
-%
-Initially, each value in the \li{cutoffs} record is a hole.
-
-\begin{lstlisting}
-type grade_cutoffs =
-  { a: float, b: float, c: float, d: float, f: float };
-
-let cutoffs =
-  { a = ??_a, b = ??_b, c = ??_c, d = ??_d, f = ??_f };
-\end{lstlisting}
-
-\noindent
-%
-%% initially all holes, because it will be different year-to-year based on the %
-%data, differences in course difficulty, and to satisfy fairness criteria.
-%
-Before starting to fill in the cutoffs, the teacher jumps ahead to write a
-function \li{letter_grade} that will make the connection between \li{cutoffs}
-and \li{weighted_averages}.
-%
-Because she intends to look at the data to help select the cutoff values, the
-teacher sorts the \li{weighted_averages} and then maps them to
-\li{letter_grades}.
-
-\begin{lstlisting}
-let letter_grade(n: weighted_average) =
-  if n >= cutoffs.a then "A" else
-  if n >= cutoffs.b then "B" else
-  if n >= cutoffs.c then "C" else
-  if n >= cutoffs.d then "D" else
-  if n >= cutoffs.f then "F" else "Incomplete";
-
-let sorted_weighted_averages = List.sort weighted_averages;
-
-let letter_grades = List.map letter_grade sorted_weighted_averages;
-\end{lstlisting}
-
-\noindent
-%
-When \HazelnutLive{} runs this program, the guard of the outermost conditional
-(\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
-is.
-%
-Therefore, each of the indeterminate expressions in \li{letter_grades} is the
-entire expression body, albeit with different bindings for \li{n}.
-%
-Displaying such ``large'' indeterminate expressions can quickly consume all
-available screen space, overwhelming the user with too much information.
-%
-\autoref{fig:XXX} shows how \HazelnutLive{} renders large indeterminate
-expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
-or hole) or list of small values) simply as \li{..} to save space.
-%
-Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
-indeterminate expression---as well as the evaluation environment that surrounds
-it---as a tooltip.
-%
-\autoref{sec:discussion} discusses this and other user interface concerns when
-trying to display useful live feedback without overwhelming the user.
-%
-These usability factors are beyond the scope of our work, which is to define
-semantic foundations on which such user interfaces can be built.
-
-To start deciding \li{cutoffs}, the teacher looks at the \li{weighted_averages}
-sorted in descending order.
-%
-The data shows a natural gap between \li{92.2} and \li{89.5}.
-%
-So, she chooses to use \li{92.0} as the cutoff for A.
-%
-Resuming the computation from before, \HazelnutLive{} resolves the conditional
-expressions for the first \rkc{XXX} indeterminate expressions, because each of
-those \li{n} values was greater than \li{cutoffs.a}.
-%
-The remaining \rkc{XXX} expressions also proceeded to evaluate the first guard,
-and are now indeterminate at the guard for the second conditional.
-
-Before assigning other cutoff values, the teacher would like to get a sense of
-whether this first choice is a good one.
-%
-She jumps ahead to to write a function that computes the distribution of letter
-grades.
-
-\begin{lstlisting}
-let compute_distribution(list: list(float)) =
-  let n = List.length letter_grades in
-  List.map
-    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list)) /. n))
-    ["A","B","C","D","F","Incomplete"];
-
-let distribution = compute_distribution(letter_grades);
-\end{lstlisting}
-
-\noindent
-%
-Running \li{compute_distribution} shows that the percentage of As is
-\rkc{XXX\%}, which is smaller than what the teacher would like;
-the remaining percentages are all indeterminate.
-%
-Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
-and then another gap between \li{88.2} and {85.5}.
-%
-So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
-%
-Because this edit is outside of a hole expression, \HazelnutLive{} discards the
-previous execution state and reevaluates the entire program.
-%
-Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
-be applied to seek opportunities for reuse even when non-hole expressions are
-modified.
-%
-Because the focus of work is on the novel implications of running programs with
-holes, our calculus and implementation supports caching ``edit-and-resume'' only
-for the novel situation in which evaluation proceeds around holes.
-%
-After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
-matches the teacher's intention.
-
-In this fashion, the teacher continues down the list of sorted averages,
-determining appropriate values for each cutoff.
-%
-Whenever the teacher is only filling in the ``next'' cutoff, the computation
-from before can simply be resumed.
-%
-Overall, throughout the workflow described in these two examples, the programmer
-can continue to evaluate the program, and receive meaningful feedback, while
-going back and forth between different pieces under development.
-
-%% \subsection{Live Programming with Type Errors}
-\subsection{Live Programming with Static Type Errors}
-
-In the previous section, we described example programs that were incomplete
-because of \emph{missing} expressions.
-%
-Next, we describe example programs that are incomplete because of
-\emph{type-inconsistent} expressions.
-%
-For example, consider the following two definitions.
-
-\begin{lstlisting}
-let bad_bool : bool = ?? 0 ??_bad_bool;
-
-let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
-\end{lstlisting}
-
-\noindent
-%
-These two definitions are ill-typed under standard typing disciplines.
-%
-In contrast, \citet{popl-paper} present a bidirectional type system that assigns
-types to both, by wrapping type-inconsistent expressions (\li{0} on line
-\rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
-%
-Non-empty holes prevent local type inconsistencies from polluting the rest of
-the program surrounding it, which may or may not itself contain additional
-inconsistencies.
-
-Understanding and debugging static type errors is notoriously difficult,
-particularly for novices.
-%
-A variety of approaches have been
-proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
-and explain type errors.
-%
-One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
-that demonstrates a run-time failure, and then displays a compressed execution
-trace to the user as a graph.
-%
-In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
-non-empty holes) as far as possible, in particular, until they
-would go wrong.
-%
-Thus, our operational semantics provides a similar benefit to the approach of
-\citet{Seidel2016}.
-
-\overviewExample{3}{Sum List}
-%
-Consider the following buggy program (observed during an undergraduate
-functional programming course~\cite{Seidel2016}) that attempts to sum a list
-integers.
-%
-The error is that the base case produces a list rather than an integer.
-
-\begin{lstlisting}
-sum_list : list(int) -> int
-sum_list [] = ?[]?
-sum_list (n:ns) = n + sum_list ns
-\end{lstlisting}
-
-\noindent
-%
-Because the list expression on line 2 does not have type \li{int} as required,
-it is wrapped in a (non-empty) hole by the bidirectional type
-checker~\cite{popl-paper}.
-%
-Rather than trying to debug the error based on the static error, the programmer
-may wish to trying running the function anyway by calling, say, \li{sum_list(2)}.
-%
-\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
-%
-By observing that the hole expression is being added to the integer \li{3}, he
-realizes that it needs to be an integer, specifically, \li{0}.
-%
-Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
-produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
-successfully proceeded to evaluate despite the error elsewhere.
-
-%% TODO fold error from Erwig paper.
-%% %
-%% see that final call on stack does have the right answer, but
-%% it's wrapped in a singleton list when the expected type is not
-%% a list.
-%% %
-%% fix is to remove the list, the rest of the computation remains
-%% the same, but b/c they were all wrapped in holes, need to re-run.
-%% %
-%% (add some mechanism for type-consistent non-empty-holes...)
-
-\overviewExample{4}{Stutter}
-%
-Consider the following function which attempts to produce a
-list where every element is repeated twice (borrowed from \citet{Osera2015}).
-%
-The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
-it produces a \li{list(list(int)} instead.
-
-\begin{lstlisting}
-stutter : list(int) -> list(int)
-stutter xs = List.foldr (\x acc -> ?[x,x]? : acc) [] xs
-\end{lstlisting}
-
-\noindent
-%
-The bidirectional type checker of \citet{popl-paper} wraps the expression
-\li{[x,x]} inside a non-empty hole.
-%
-%% The editor has a choice about which expression to ``blame'' for the error; the
-%% entire application that forms the body of the lambda is analyzed against the
-%% return type \li{list(int)}, so that is a reasonable choice for the editor to
-%% make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
-%% both as intended and that only the function \li{(:)} is type-consistent.
-%% %
-%% Although one could imagine a setting in which a user would perform this
-%% reasoning, let's assume the simplest approach for marking consistencies that
-%% wraps the entire application.
-%
-Running this on \li{stutter [1,2,3]} produces the indeterminate result
-
-\begin{lstlisting}
-?  [1,1] : (? [2,2] : (? [[3:3]] ?) ?) ?,
-\end{lstlisting}
-
-\noindent
-%
-which shows the unfolding of \li{List.foldr}.
-%
-We refer to nested indeterminate computations like this as \rkc{\emph{hole
-environment traces} or \emph{hole environment trees}}.
-%
-The result of the innermost indeterminate expression is \li{[[3,3]]}.
-%
-The user realizes that there are too many levels of nesting, so
-he replaces the \li{(:)} with \li{(@)}, which addresses the type inconsistency
-and, when reevaluated, produces the desired result.
-
 \subsection{Live Programming with Gradual Type Errors}
@@
 This approach can be incorporated into existing gradually typed languages.
-
-\subsection{Live Programming for Debugging}
-
-Previous examples highlighted how running different kinds of incomplete
-programs---ones with missing expressions, type-inconsistent expressions, or
-missing types---can be useful.
-%
-Our final example demonstrates how expression holes may be useful even for
-complete programs, in order to facilitate debugging and program understanding.
-
-%% \overviewExample{6}{\rkc{...}}
-%%
-%% \rkc{breakpoints, understanding input/output behavior}
-
-\overviewExample{6}{Quicksort}
-
-Consider the following buggy implementation of quicksort, which contains several
-errors.
-%
-The student wants to debug the implementation and so inserts an expression
-around the return expression on line \rkc{XXX}, and runs
-\li{quicksort [X,X,X,X,X,X]}.
-
-\begin{lstlisting}
-quicksort : list('a) -> list('a)
-quicksort [] = []
-quicksort (x:xs) =
-  let (left, right) = span ((>) x) xs in
-  let (left', right') = (quicksort left, quicksort right) in
-  ? (left' @ [x] @ right) ?
-\end{lstlisting}
-
-\noindent
-%
-Based on the live environment for the initial call, the student notices that the
-values in \li{left} are larger than the pivot \li{x}, but \li{left} was intended
-to bind smaller values.
-%
-This suggests that the student got the filtering predicate backwards, so he
-replaces it with \li{(<)} on line \rkc{XXX}.
-%
-After re-running the program and viewing the live output again, \li{left} now
-contains smaller values than \li{x}, but not all of them; \li{right} has some
-values that are larger and some that are smaller.
-%
-The student uses type-based search to find other functions, besides \li{span},
-of type \li{('a -> bool) -> 'a list -> ('a list, 'a list)} and finds
-\li{partition}.
-%
-It may be easier for the student to simply try the function rather than studying
-the documentation, so he edits line \rkc{XXX} to call \li{partition} instead,
-re-runs, and observes that both \li{left} and \li{right} now seem to achieve the
-intended partitioning.
-
-Despite these two fixes, the result expression is not yet sorted.
-%
-In particular, the values greater than the pivot \li{x} are not completely
-sorted.
-%
-Viewing the values of bindings of \li{left'} and \li{right'}, the student sees
-that both seem to have correct prefixes of sorted lists, but that they are
-incorrect after the pivots.
-%
-This suggests that something may be wrong with the handling of \li{right'}.
-%
-Indeed, the student takes a look for uses of \li{right'} and sees that line
-\rkc{XXX} mistakenly uses \li{right} instead.
-%
-Making this change and re-running now produces the desired result, built up as a
-hole environment tree that shows how all of the recursive results will be
-combined by a series of nested appends.
-%
-Happy with the results of the program, the student removes the hole around the
-expression on line \rkc{XXX}, and evaluation can proceed to perform the nested
-appends, producing the final sorted result.
-%
-\autoref{sec:discussion} discusses how debugging tools may be designed to
-systematically insert hole expressions in particular places to achieve certain
-debugging and program understanding strategies.
-%
-%% When run on a larger input, this projection of the full evaluation trace shows a
-%% nesting structure---which could be rendered as visualization to help
-%% understanding, as described in \autoref{sec:discussion}---whose depth is clearly
-%% much smaller than the length of the input list; this suggests that the
-%% algorithm, indeed, achieves some sort of sublinear complexity, as intended.
-%
-
-\begin{lstlisting}
-quicksort : list('a) -> list('a)
-quicksort [] = []
-quicksort (x:xs) =
-  let (left, right) = partition ((<) x) xs in
-  let (left', right') = (quicksort left, quicksort right) in
-  ? (left' @ [x] @ right') ?
-\end{lstlisting}
-
-\paragraph{Recap}
-%
-In the above four subsections, respectively, we considered programs with:
-%
-(1) empty expression holes,
-%
-(2) non-empty expression holes for ill-typed expressions,
-%
-(3) type holes, and
-%
-(4) non-empty expression holes for well-typed expressions.
-%
-Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
-that allows programs with combinations of these kinds of incompleteness to be
-evaluated.

### 2018-03-08 examples-4.tex
splitting Section 2 subsections into separate files
to help avoid merge conflicts

--- 2018-02-21 examples.tex
@@
-% !TEX root = hazelnut-dynamics.tex
-%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
-%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
-\newcommand{\examplesSec}{\HazelnutLive, By Example}
-\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
-\label{sec:examples}
-
-\newcommand{\overviewExample}[2]{\paragraph{Example {#1}: {#2}}}
-
-We introduce \HazelnutLive{} in this section using several examples,
-demonstrating how running incomplete programs---with missing expressions and/or
-type-inconsistent expressions---can be useful.
-%
-We will also demonstrate how running even complete programs with subexpressions
-marked as holes may be useful in debugging.
-
-\HazelnutLive{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
-alternative Reason syntax \cite{reason-what} extended to support holes.
-%
-As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
-literate program \cite{knuth1984literate}: rich text comments appear at the top
-level and code appears within cells embedded into these comments.
-%
-The reader is encouraged to work through these examples in our prototype
-implementation or watch video demonstrations, both available in
-\suppMaterials{}.\footnote{
-%
-The programs in this section are written using minor syntactic conveniences that
-are not currently supported in our implementation.
-%
-For example, the implementation of recursive types in our current implementation
-requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
-inferred in standard ways (by pairing them with constructors and deconstructors
-for algebraic datatypes) in future work.
-%
-More details about other minor syntactic differences are described in
-\suppMaterials{}.
-%
-}
-
-\subsection{Live Programming with Incomplete Expressions}
-
-The process of writing new program fragments includes many states in which the
-program under construction is, by definition, incomplete.
-%
-It is natural for a programmer to ``jump back and forth'' between different
-places in the program, iteratively and alternately developing the missing
-pieces.
-%
-The primary benefit of our approach is that such programs can be evaluated in
-order to provide useful feedback during this editing workflow.
-
-\input{fig-grades-example}
-
-Consider a teacher who is in the midst of developing a \HazelnutLive{} notebook
-(depicted in \autoref{fig:grades-example}) to compute final student grades at
-the end of a term.
-%
-In the first cell in \autoref{fig:grades-example}, the teacher defines a record
-type, \li{student_data}, for each student's course data.
-%
-In our simplified example, this includes the student's name, of type
-\li{string}, and five grades, each of type \li{float}.
-
-\overviewExample{1}{Computing Weighted Averages}
-%
-The teacher begins writing a \li{weighted_average} function that will
-compute a weighted average, of type \li{float}, for each \li{student_data}
-record.
-%
-The body of \li{weighted_average} is not yet written---marked with the
-\emph{empty hole} on line \rkc{XXX}--- but the teacher jumps ahead to map the
-function over the \li{grades} list.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  ??;
-
-let weighted_averages = List.map weighted_average grades;
-\end{lstlisting}
-
-\noindent
-%
-\HazelnutLive{} evaluates this incomplete program, showing its
-\emph{indeterminate} result in the sidebar on the right half of
-\autoref{fig:grades-example}; this value is a list whose length is the same as
-\li{grades}, where each element is the (indeterminate) result of evaluating the
-hole expression inside \li{weighted_average}.
-%
-Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
-(\ie{}~\li{a}) and that the indeterminate results it produces are assigned
-corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2}, etc.).
-%
-In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
-using placeholder expressions like \li{raise "Not yet implemented."}, would halt
-the program as soon it is evaluated (in this case, inside the definition of
-\li{List.map} when the function is applied to the first element of \li{grades}.
-%
-By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
-least see that the resulting expression does indeed evaluate to a list with the
-correct number of (indeterminate) values.
-
-Now the teacher returns to \li{weighted_average} to work on the missing
-expression, with the goal of computing a weighted sum of each of a student's
-grades.
-%
-The teacher uses a \HazelnutLive{} edit action to create a sum expression,
-filling in the first summand by weighting the \li{hw1} score to account for 10\%
-of the weighted average; the rest of the computation is a new hole expression
-(\li{??_b}), yet to be filled in.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  (10.0 *. g.hw1) +. ??_b;
-\end{lstlisting}
-
-\noindent
-%
-\HazelnutLive{} immediately runs this incomplete program, displaying the updated
-list of indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
-%
-Right away, the teacher recognizes that the values are too large; they should be
-at most \li{100.0}.
-%
-The teacher realizes that representing percentage points as \li{float}s requires
-that the constant on line \rkc{XXX} ought to be \li{0.10} instead.
-%
-Because of this live feedback, the teacher corrects this error right away and
-avoids making similar programming errors in the rest of the calculation.
-%
-The teacher continues to build the rest of the arithmetic expression until it is
-complete (there are no longer any expression holes), and the result of
-immediately running the finished program shows that each of values in the final
-result list is in the range \li{0.0} to \li{100.0}.
-
-\overviewExample{2}{Assigning Letter Grades}
-%
-The teacher's next task is to map the weighted averages to the letter grades A
-through F (we will consider only ``whole'' letter grades, for simplicity).
-%
-The \li{grade_cutoffs} record type describes the minimum cutoff for each of
-these six possible grades.
-%
-Initially, each value in the \li{cutoffs} record is a hole.
-
-\begin{lstlisting}
-type grade_cutoffs =
-  { a: float, b: float, c: float, d: float, f: float };
-
-let cutoffs =
-  { a = ??_a, b = ??_b, c = ??_c, d = ??_d, f = ??_f };
-\end{lstlisting}
-
-\noindent
-%
-%% initially all holes, because it will be different year-to-year based on the %
-%data, differences in course difficulty, and to satisfy fairness criteria.
-%
-Before starting to fill in the cutoffs, the teacher jumps ahead to write a
-function \li{letter_grade} that will make the connection between \li{cutoffs}
-and \li{weighted_averages}.
-%
-Because she intends to look at the data to help select the cutoff values, the
-teacher sorts the \li{weighted_averages} and then maps them to
-\li{letter_grades}.
-
-\begin{lstlisting}
-let letter_grade(n: weighted_average) =
-  if n >= cutoffs.a then "A" else
-  if n >= cutoffs.b then "B" else
-  if n >= cutoffs.c then "C" else
-  if n >= cutoffs.d then "D" else
-  if n >= cutoffs.f then "F" else "Incomplete";
-
-let sorted_weighted_averages = List.sort weighted_averages;
-
-let letter_grades = List.map letter_grade sorted_weighted_averages;
-\end{lstlisting}
-
-\noindent
-%
-When \HazelnutLive{} runs this program, the guard of the outermost conditional
-(\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
-is.
-%
-Therefore, each of the indeterminate expressions in \li{letter_grades} is the
-entire expression body, albeit with different bindings for \li{n}.
-%
-Displaying such ``large'' indeterminate expressions can quickly consume all
-available screen space, overwhelming the user with too much information.
-%
-\autoref{fig:XXX} shows how \HazelnutLive{} renders large indeterminate
-expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
-or hole) or list of small values) simply as \li{..} to save space.
-%
-Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
-indeterminate expression---as well as the evaluation environment that surrounds
-it---as a tooltip.
-%
-\autoref{sec:discussion} discusses this and other user interface concerns when
-trying to display useful live feedback without overwhelming the user.
-%
-These usability factors are beyond the scope of our work, which is to define
-semantic foundations on which such user interfaces can be built.
-
-To start deciding \li{cutoffs}, the teacher looks at the \li{weighted_averages}
-sorted in descending order.
-%
-The data shows a natural gap between \li{92.2} and \li{89.5}.
-%
-So, she chooses to use \li{92.0} as the cutoff for A.
-%
-Resuming the computation from before, \HazelnutLive{} resolves the conditional
-expressions for the first \rkc{XXX} indeterminate expressions, because each of
-those \li{n} values was greater than \li{cutoffs.a}.
-%
-The remaining \rkc{XXX} expressions also proceeded to evaluate the first guard,
-and are now indeterminate at the guard for the second conditional.
-
-Before assigning other cutoff values, the teacher would like to get a sense of
-whether this first choice is a good one.
-%
-She jumps ahead to to write a function that computes the distribution of letter
-grades.
-
-\begin{lstlisting}
-let compute_distribution(list: list(float)) =
-  let n = List.length letter_grades in
-  List.map
-    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list)) /. n))
-    ["A","B","C","D","F","Incomplete"];
-
-let distribution = compute_distribution(letter_grades);
-\end{lstlisting}
-
-\noindent
-%
-Running \li{compute_distribution} shows that the percentage of As is
-\rkc{XXX\%}, which is smaller than what the teacher would like;
-the remaining percentages are all indeterminate.
-%
-Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
-and then another gap between \li{88.2} and {85.5}.
-%
-So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
-%
-Because this edit is outside of a hole expression, \HazelnutLive{} discards the
-previous execution state and reevaluates the entire program.
-%
-Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
-be applied to seek opportunities for reuse even when non-hole expressions are
-modified.
-%
-Because the focus of work is on the novel implications of running programs with
-holes, our calculus and implementation supports caching ``edit-and-resume'' only
-for the novel situation in which evaluation proceeds around holes.
-%
-After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
-matches the teacher's intention.
-
-In this fashion, the teacher continues down the list of sorted averages,
-determining appropriate values for each cutoff.
-%
-Whenever the teacher is only filling in the ``next'' cutoff, the computation
-from before can simply be resumed.
-%
-Overall, throughout the workflow described in these two examples, the programmer
-can continue to evaluate the program, and receive meaningful feedback, while
-going back and forth between different pieces under development.
-
-%% \subsection{Live Programming with Type Errors}
-\subsection{Live Programming with Static Type Errors}
-
-In the previous section, we described example programs that were incomplete
-because of \emph{missing} expressions.
-%
-Next, we describe example programs that are incomplete because of
-\emph{type-inconsistent} expressions.
-%
-For example, consider the following two definitions.
-
-\begin{lstlisting}
-let bad_bool : bool = ?? 0 ??_bad_bool;
-
-let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
-\end{lstlisting}
-
-\noindent
-%
-These two definitions are ill-typed under standard typing disciplines.
-%
-In contrast, \citet{popl-paper} present a bidirectional type system that assigns
-types to both, by wrapping type-inconsistent expressions (\li{0} on line
-\rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
-%
-Non-empty holes prevent local type inconsistencies from polluting the rest of
-the program surrounding it, which may or may not itself contain additional
-inconsistencies.
-
-Understanding and debugging static type errors is notoriously difficult,
-particularly for novices.
-%
-A variety of approaches have been
-proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
-and explain type errors.
-%
-One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
-that demonstrates a run-time failure, and then displays a compressed execution
-trace to the user as a graph.
-%
-In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
-non-empty holes) as far as possible, in particular, until they
-would go wrong.
-%
-Thus, our operational semantics provides a similar benefit to the approach of
-\citet{Seidel2016}.
-
-\overviewExample{3}{Sum List}
-%
-Consider the following buggy program (observed during an undergraduate
-functional programming course~\cite{Seidel2016}) that attempts to sum a list
-integers.
-%
-The error is that the base case produces a list rather than an integer.
-
-\begin{lstlisting}
-sum_list : list(int) -> int
-sum_list [] = ?[]?
-sum_list (n:ns) = n + sum_list ns
-\end{lstlisting}
-
-\noindent
-%
-Because the list expression on line 2 does not have type \li{int} as required,
-it is wrapped in a (non-empty) hole by the bidirectional type
-checker~\cite{popl-paper}.
-%
-Rather than trying to debug the error based on the static error, the programmer
-may wish to trying running the function anyway by calling, say, \li{sum_list(2)}.
-%
-\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
-%
-By observing that the hole expression is being added to the integer \li{3}, he
-realizes that it needs to be an integer, specifically, \li{0}.
-%
-Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
-produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
-successfully proceeded to evaluate despite the error elsewhere.
-
-%% TODO fold error from Erwig paper.
-%% %
-%% see that final call on stack does have the right answer, but
-%% it's wrapped in a singleton list when the expected type is not
-%% a list.
-%% %
-%% fix is to remove the list, the rest of the computation remains
-%% the same, but b/c they were all wrapped in holes, need to re-run.
-%% %
-%% (add some mechanism for type-consistent non-empty-holes...)
-
-\overviewExample{4}{Stutter}
-%
-Consider the following function which attempts to produce a
-list where every element is repeated twice (borrowed from \citet{Osera2015}).
-%
-The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
-it produces a \li{list(list(int)} instead.
-
-\begin{lstlisting}
-stutter : list(int) -> list(int)
-stutter xs = List.foldr (\x acc -> ?[x,x]? : acc) [] xs
-\end{lstlisting}
-
-\noindent
-%
-The bidirectional type checker of \citet{popl-paper} wraps the expression
-\li{[x,x]} inside a non-empty hole.
-%
-%% The editor has a choice about which expression to ``blame'' for the error; the
-%% entire application that forms the body of the lambda is analyzed against the
-%% return type \li{list(int)}, so that is a reasonable choice for the editor to
-%% make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
-%% both as intended and that only the function \li{(:)} is type-consistent.
-%% %
-%% Although one could imagine a setting in which a user would perform this
-%% reasoning, let's assume the simplest approach for marking consistencies that
-%% wraps the entire application.
-%
-Running this on \li{stutter [1,2,3]} produces the indeterminate result
-
-\begin{lstlisting}
-?  [1,1] : (? [2,2] : (? [[3:3]] ?) ?) ?,
-\end{lstlisting}
-
-\noindent
-%
-which shows the unfolding of \li{List.foldr}.
-%
-We refer to nested indeterminate computations like this as \rkc{\emph{hole
-environment traces} or \emph{hole environment trees}}.
-%
-The result of the innermost indeterminate expression is \li{[[3,3]]}.
-%
-The user realizes that there are too many levels of nesting, so
-he replaces the \li{(:)} with \li{(@)}, which addresses the type inconsistency
-and, when reevaluated, produces the desired result.
-
-\subsection{Live Programming with Gradual Type Errors}
-
-\emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept
-programs that would otherwise be rejected by static type systems---either
-because type inference cannot reconstruct a valid type assignment, or because
-there may not be a unique valid type assignment at all.
-%
-In either case, gradual type systems allow types to contain \emph{unknown
-types}, and partially unknown types can be used in contexts that require
-different types, so long as they are \emph{consistent} (intuitively, equal
-modulo the unknown parts).
-%
-Dynamic casts are then inserted to ensure that these remaining static
-discrepancies are not violated at run-time.
-
-\HazelnutLive{} inherits the notion of \emph{type holes} from
-\citet{popl-paper}, which serves a similar static purpose as the unknown type in
-gradual type systems.
-%
-In contrast to prior gradually typed languages, however, \HazelnutLive{} can
-evaluate ``around'' cast errors in the same way as it does for expression holes.
-
-\overviewExample{5}{Dynamically Typed Negation}
-
-Consider the \li{negate} function (adapted from \cite{ChughPOPL2012}), which
-cannot be assigned a static type in a conventional type system---either a
-bidirectional one, like in \HazelnutLive{}, or one with ML-style,
-unification-based inference---because the argument \li{x} is used at type
-\li{int} on line \rkc{XXX} and \li{bool} on line \rkc{XXX}.
-%
-Therefore, the declared type of \li{x} is the hole type \li{??}, which allows
-\li{x} to be used at the conflicting types, with each use expanded into an
-expression wrapped in a \emph{cast} that will dynamically check for safety.
-%
-The return type also uses the hole type \li{??}, leading to additional casts in
-the expansion.
-
-\begin{lstlisting}
-negate : bool -> ?? -> ??
-negate b x =
-  if b
-    then 0 - x     // expanded to: (0 - (x<?? => int>))<int => ??>
-    else not x     // expanded to: (not (x<?? => bool>))<bool => ??>
-
-(negate false 1) + 2 + 3 + (negate true 4)
-\end{lstlisting}
-
-\noindent
-%
-As in prior gradually typed languages~\cite{XXX},
-evaluating the expression \li{negate false 1} on line \rkc{XXX} leads to
-\li{(not (1<int => ??><?? => bool>))<bool => ??>}, where the inner casts
-lead to the \emph{cast error} \li{1<?? =/=> bool>} because \li{1} cannot be
-safely treated as having type \li{bool}.
-%
-Unlike in prior gradually typed languages, however, \HazelnutLive{} evaluates
-around the cast error, making progress on the \li{2 + 3 + negate false 4}
-expression that surrounds the error; the final indeterminate result is
-\li{(not (1<?? =/=> bool>))<bool => ??> + 1}.
-%
-Just like it is useful for static type checkers to report multiple errors, our
-approach allows us to report multiple dynamic cast errors, and otherwise make
-progress on expressions that do not depend on failed casts.
-%
-This approach can be incorporated into existing gradually typed languages.
-
 \subsection{Live Programming for Debugging}
@@
 \end{lstlisting}
-
-\paragraph{Recap}
-%
-In the above four subsections, respectively, we considered programs with:
-%
-(1) empty expression holes,
-%
-(2) non-empty expression holes for ill-typed expressions,
-%
-(3) type holes, and
-%
-(4) non-empty expression holes for well-typed expressions.
-%
-Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
-that allows programs with combinations of these kinds of incompleteness to be
-evaluated.

### 2018-03-08 examples.tex
splitting Section 2 subsections into separate files
to help avoid merge conflicts

--- 2018-02-20 examples.tex
@@
 
-\subsection{Live Programming with Incomplete Expressions}
-
-The process of writing new program fragments includes many states in which the
-program under construction is, by definition, incomplete.
-%
-It is natural for a programmer to ``jump back and forth'' between different
-places in the program, iteratively and alternately developing the missing
-pieces.
-%
-The primary benefit of our approach is that such programs can be evaluated in
-order to provide useful feedback during this editing workflow.
-
-\input{fig-grades-example}
-
-Consider a teacher who is in the midst of developing a \HazelnutLive{} notebook
-(depicted in \autoref{fig:grades-example}) to compute final student grades at
-the end of a term.
-%
-In the first cell in \autoref{fig:grades-example}, the teacher defines a record
-type, \li{student_data}, for each student's course data.
-%
-In our simplified example, this includes the student's name, of type
-\li{string}, and five grades, each of type \li{float}.
-
-\overviewExample{1}{Computing Weighted Averages}
-%
-The teacher begins writing a \li{weighted_average} function that will
-compute a weighted average, of type \li{float}, for each \li{student_data}
-record.
-%
-The body of \li{weighted_average} is not yet written---marked with the
-\emph{empty hole} on line \rkc{XXX}--- but the teacher jumps ahead to map the
-function over the \li{grades} list.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  ??;
-
-let weighted_averages = List.map weighted_average grades;
-\end{lstlisting}
-
-\noindent
-%
-\HazelnutLive{} evaluates this incomplete program, showing its
-\emph{indeterminate} result in the sidebar on the right half of
-\autoref{fig:grades-example}; this value is a list whose length is the same as
-\li{grades}, where each element is the (indeterminate) result of evaluating the
-hole expression inside \li{weighted_average}.
-%
-Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
-(\ie{}~\li{a}) and that the indeterminate results it produces are assigned
-corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2}, etc.).
-%
-In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
-using placeholder expressions like \li{raise "Not yet implemented."}, would halt
-the program as soon it is evaluated (in this case, inside the definition of
-\li{List.map} when the function is applied to the first element of \li{grades}.
-%
-By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
-least see that the resulting expression does indeed evaluate to a list with the
-correct number of (indeterminate) values.
-
-Now the teacher returns to \li{weighted_average} to work on the missing
-expression, with the goal of computing a weighted sum of each of a student's
-grades.
-%
-The teacher uses a \HazelnutLive{} edit action to create a sum expression,
-filling in the first summand by weighting the \li{hw1} score to account for 10\%
-of the weighted average; the rest of the computation is a new hole expression
-(\li{??_b}), yet to be filled in.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  (10.0 *. g.hw1) +. ??_b;
-\end{lstlisting}
-
-\noindent
-%
-\HazelnutLive{} immediately runs this incomplete program, displaying the updated
-list of indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
-%
-Right away, the teacher recognizes that the values are too large; they should be
-at most \li{100.0}.
-%
-The teacher realizes that representing percentage points as \li{float}s requires
-that the constant on line \rkc{XXX} ought to be \li{0.10} instead.
-%
-Because of this live feedback, the teacher corrects this error right away and
-avoids making similar programming errors in the rest of the calculation.
-%
-The teacher continues to build the rest of the arithmetic expression until it is
-complete (there are no longer any expression holes), and the result of
-immediately running the finished program shows that each of values in the final
-result list is in the range \li{0.0} to \li{100.0}.
-
-\overviewExample{2}{Assigning Letter Grades}
-%
-The teacher's next task is to map the weighted averages to the letter grades A
-through F (we will consider only ``whole'' letter grades, for simplicity).
-%
-The \li{grade_cutoffs} record type describes the minimum cutoff for each of
-these six possible grades.
-%
-Initially, each value in the \li{cutoffs} record is a hole.
-
-\begin{lstlisting}
-type grade_cutoffs =
-  { a: float, b: float, c: float, d: float, f: float };
-
-let cutoffs =
-  { a = ??_a, b = ??_b, c = ??_c, d = ??_d, f = ??_f };
-\end{lstlisting}
-
-\noindent
-%
-%% initially all holes, because it will be different year-to-year based on the %
-%data, differences in course difficulty, and to satisfy fairness criteria.
-%
-Before starting to fill in the cutoffs, the teacher jumps ahead to write a
-function \li{letter_grade} that will make the connection between \li{cutoffs}
-and \li{weighted_averages}.
-%
-Because she intends to look at the data to help select the cutoff values, the
-teacher sorts the \li{weighted_averages} and then maps them to
-\li{letter_grades}.
-
-\begin{lstlisting}
-let letter_grade(n: weighted_average) =
-  if n >= cutoffs.a then "A" else
-  if n >= cutoffs.b then "B" else
-  if n >= cutoffs.c then "C" else
-  if n >= cutoffs.d then "D" else
-  if n >= cutoffs.f then "F" else "Incomplete";
-
-let sorted_weighted_averages = List.sort weighted_averages;
-
-let letter_grades = List.map letter_grade sorted_weighted_averages;
-\end{lstlisting}
-
-\noindent
-%
-When \HazelnutLive{} runs this program, the guard of the outermost conditional
-(\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
-is.
-%
-Therefore, each of the indeterminate expressions in \li{letter_grades} is the
-entire expression body, albeit with different bindings for \li{n}.
-%
-Displaying such ``large'' indeterminate expressions can quickly consume all
-available screen space, overwhelming the user with too much information.
-%
-\autoref{fig:XXX} shows how \HazelnutLive{} renders large indeterminate
-expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
-or hole) or list of small values) simply as \li{..} to save space.
-%
-Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
-indeterminate expression---as well as the evaluation environment that surrounds
-it---as a tooltip.
-%
-\autoref{sec:discussion} discusses this and other user interface concerns when
-trying to display useful live feedback without overwhelming the user.
-%
-These usability factors are beyond the scope of our work, which is to define
-semantic foundations on which such user interfaces can be built.
-
-To start deciding \li{cutoffs}, the teacher looks at the \li{weighted_averages}
-sorted in descending order.
-%
-The data shows a natural gap between \li{92.2} and \li{89.5}.
-%
-So, she chooses to use \li{92.0} as the cutoff for A.
-%
-Resuming the computation from before, \HazelnutLive{} resolves the conditional
-expressions for the first \rkc{XXX} indeterminate expressions, because each of
-those \li{n} values was greater than \li{cutoffs.a}.
-%
-The remaining \rkc{XXX} expressions also proceeded to evaluate the first guard,
-and are now indeterminate at the guard for the second conditional.
-
-Before assigning other cutoff values, the teacher would like to get a sense of
-whether this first choice is a good one.
-%
-She jumps ahead to to write a function that computes the distribution of letter
-grades.
-
-\begin{lstlisting}
-let compute_distribution(list: list(float)) =
-  let n = List.length letter_grades in
-  List.map
-    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list)) /. n))
-    ["A","B","C","D","F","Incomplete"];
-
-let distribution = compute_distribution(letter_grades);
-\end{lstlisting}
-
-\noindent
-%
-Running \li{compute_distribution} shows that the percentage of As is
-\rkc{XXX\%}, which is smaller than what the teacher would like;
-the remaining percentages are all indeterminate.
-%
-Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
-and then another gap between \li{88.2} and {85.5}.
-%
-So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
-%
-Because this edit is outside of a hole expression, \HazelnutLive{} discards the
-previous execution state and reevaluates the entire program.
-%
-Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
-be applied to seek opportunities for reuse even when non-hole expressions are
-modified.
-%
-Because the focus of work is on the novel implications of running programs with
-holes, our calculus and implementation supports caching ``edit-and-resume'' only
-for the novel situation in which evaluation proceeds around holes.
-%
-After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
-matches the teacher's intention.
-
-In this fashion, the teacher continues down the list of sorted averages,
-determining appropriate values for each cutoff.
-%
-Whenever the teacher is only filling in the ``next'' cutoff, the computation
-from before can simply be resumed.
-%
-Overall, throughout the workflow described in these two examples, the programmer
-can continue to evaluate the program, and receive meaningful feedback, while
-going back and forth between different pieces under development.
-
-%% \subsection{Live Programming with Type Errors}
-\subsection{Live Programming with Static Type Errors}
-
-In the previous section, we described example programs that were incomplete
-because of \emph{missing} expressions.
-%
-Next, we describe example programs that are incomplete because of
-\emph{type-inconsistent} expressions.
-%
-For example, consider the following two definitions.
-
-\begin{lstlisting}
-let bad_bool : bool = ?? 0 ??_bad_bool;
-
-let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
-\end{lstlisting}
-
-\noindent
-%
-These two definitions are ill-typed under standard typing disciplines.
-%
-In contrast, \citet{popl-paper} present a bidirectional type system that assigns
-types to both, by wrapping type-inconsistent expressions (\li{0} on line
-\rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
-%
-Non-empty holes prevent local type inconsistencies from polluting the rest of
-the program surrounding it, which may or may not itself contain additional
-inconsistencies.
-
-Understanding and debugging static type errors is notoriously difficult,
-particularly for novices.
-%
-A variety of approaches have been
-proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
-and explain type errors.
-%
-One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
-that demonstrates a run-time failure, and then displays a compressed execution
-trace to the user as a graph.
-%
-In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
-non-empty holes) as far as possible, in particular, until they
-would go wrong.
-%
-Thus, our operational semantics provides a similar benefit to the approach of
-\citet{Seidel2016}.
-
-\overviewExample{3}{Sum List}
-%
-Consider the following buggy program (observed during an undergraduate
-functional programming course~\cite{Seidel2016}) that attempts to sum a list
-integers.
-%
-The error is that the base case produces a list rather than an integer.
-
-\begin{lstlisting}
-sumList : list(int) -> int
-sumList [] = ?[]?
-sumList (n:ns) = n + sumList ns
-\end{lstlisting}
-
-\noindent
-%
-Because the list expression on line 2 does not have type \li{int} as required,
-it is wrapped in a (non-empty) hole by the bidirectional type
-checker~\cite{popl-paper}.
-%
-Rather than trying to debug the error based on the static error, the programmer
-may wish to trying running the function anyway by calling, say, \li{sumList(2)}.
-%
-\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
-%
-By observing that the hole expression is being added to the integer \li{3}, he
-realizes that it needs to be an integer, specifically, \li{0}.
-%
-Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
-produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
-successfully proceeded to evaluate despite the error elsewhere.
-
-%% TODO fold error from Erwig paper.
-%% %
-%% see that final call on stack does have the right answer, but
-%% it's wrapped in a singleton list when the expected type is not
-%% a list.
-%% %
-%% fix is to remove the list, the rest of the computation remains
-%% the same, but b/c they were all wrapped in holes, need to re-run.
-%% %
-%% (add some mechanism for type-consistent non-empty-holes...)
-
-\overviewExample{4}{Stutter}
-%
-Consider the following function which attempts to produce a
-list where every element is repeated twice (borrowed from \citet{Osera2015}).
-%
-The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
-it produces a \li{list(list(int)} instead.
-
-\begin{lstlisting}
-stutter : list(int) -> list(int)
-stutter xs = List.foldr (\x acc -> ?[x,x]? : acc) [] xs
-\end{lstlisting}
-
-\noindent
-%
-The bidirectional type checker of \citet{popl-paper} wraps the expression
-\li{[x,x]} inside a non-empty hole.
-%
-%% The editor has a choice about which expression to ``blame'' for the error; the
-%% entire application that forms the body of the lambda is analyzed against the
-%% return type \li{list(int)}, so that is a reasonable choice for the editor to
-%% make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
-%% both as intended and that only the function \li{(:)} is type-consistent.
-%% %
-%% Although one could imagine a setting in which a user would perform this
-%% reasoning, let's assume the simplest approach for marking consistencies that
-%% wraps the entire application.
-%
-Running this on \li{stutter [1,2,3]} produces the indeterminate result
-
-\begin{lstlisting}
-?  [1,1] : (? [2,2] : (? [[3:3]] ?) ?) ?,
-\end{lstlisting}
-
-\noindent
-%
-which shows the unfolding of \li{List.foldr}.
-%
-We refer to nested indeterminate computations like this as \rkc{\emph{hole
-environment traces} or \emph{hole environment trees}}.
-%
-The result of the innermost indeterminate expression is \li{[[3,3]]}.
-%
-The user realizes that there are too many levels of nesting, so
-he replaces the \li{(:)} with \li{(@)}, which addresses the type inconsistency
-and, when reevaluated, produces the desired result.
-
-\rkc{STOPPING HERE FOR NOW}
-
-\subsection{Live Programming with Gradual Type Errors}
-
-\overviewExample{5}{\rkc{...}}
-
-\rkc{gradual typing error, cast error, run around}
-
-\begin{lstlisting}
-foo : bool -> ? -> int
-foo b =
-  if b then
-    x + 10
-  else
-    x 10
-
-foo false 1 + 2 + 2
-\end{lstlisting}
-
-this demonstrates how, when no expression holes, our system is like gradual
-typing, but with the ability to run around cast errors.
-%
-just as it may be useful to report multiple type errors, this allows ability to
-report multiple dynamic errors, as well additional dynamic behavior around them.
-%
-could be incorporated into other gradual type systems
-
-\subsection{Live Programming for Debugging}
-
-\overviewExample{6}{\rkc{...}}
-
-\rkc{breakpoints, understanding input/output behavior}
-
-\parasection{Example 6: Quicksort}
-
-\begin{lstlisting}
-quicksort : list('a) -> list('a)
-quicksort [] = []
-quicksort (x:xs) =
-  let (left, right) = span ((>) x) xs in
-  ??
-\end{lstlisting}
-
-programmer has some intended relationship between names and values they bind.
-%
-live results show that left has larger elements and right has smaller.
-%
-right away, can see that need to change filter predicate. (alternatively, the
-names could have been swapped, but that order of names was the natural way
-to think about it.)
-%
-but next, still see that right has some smaller elements.
-%
-should pick partition instead of span.
-
-\begin{lstlisting}
-quicksort : list('a) -> list('a)
-quicksort [] = []
-quicksort (x:xs) =
-  let (left, right) = partition ((<) x) xs in
-  let (left', right') = (quicksort left, quicksort right) in
-  ? (left' @ [x] @ right) ?
-\end{lstlisting}
-
-now, put the lists back together.
-%
-doesn't do what is expected.
-%
-live feedback shows right' is computed correctly, but ah, it's not being used
-in the return list.
-%
-fix, and re-run, keeping the final expression in a hole.
-%
-can visualize that the nested appends are being built up correctly, and that the
-result would be the correct list.
-%
-finalize the hole and resume. everything but the nested appends can be reused.
-%
-\emph{hole environment traces} or \emph{hole environment traces}
+\input{examples-1}
+\input{examples-2}
+\input{examples-3}
+\input{examples-4}
 

### 2018-03-11 examples-1.tex
nuke stutter and re-number examples to match subsection numbers
some work on Sec 2.1, with placeholders for mockup screenshots
I still don't have good sense of how these mockups are going to fake
all of the necessary features for the grades examples, so this text is
still pretty wishy-washy.

@@
 
-Consider a teacher who is in the midst of developing a \HazelnutLive{} notebook
-(depicted in \autoref{fig:grades-example}) to compute final student grades at
-the end of a term.
+\overviewExample{1a}{Computing Weighted Averages}
 %
-In the first cell in \autoref{fig:grades-example}, the teacher defines a record
-type, \li{student_data}, for each student's course data.
+Consider a teacher who is in the midst of developing a \Hazel{} notebook
+(depicted at the top of \autoref{fig:grades-example}) to compute final student
+grades at the end of a term.
+%
+At the top of the program, the teacher defines a
+record type, \li{student_data}, for each student's course data.
 %
@@
 
-\overviewExample{1}{Computing Weighted Averages}
-%
 The teacher begins writing a \li{weighted_average} function that will
@@
 The body of \li{weighted_average} is not yet written---marked with the
-\emph{empty hole} on line \rkc{XXX}--- but the teacher jumps ahead to map the
-function over the \li{grades} list.
-
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  ??;
-
-let weighted_averages = List.map weighted_average grades;
-\end{lstlisting}
-
-\noindent
+\emph{empty hole} on line \rkc{XXX}---but the teacher jumps ahead to map the
+function over the \li{grades} list on line \rkc{XXX}.
 %
-\HazelnutLive{} evaluates this incomplete program, showing its
-\emph{indeterminate} result in the sidebar on the right half of
-\autoref{fig:grades-example}; this value is a list whose length is the same as
+\Hazel{} evaluates this incomplete program, showing its
+\emph{indeterminate} result in \emph{results panel} just below the
+program in the center of the window (\autoref{fig:grades-example});
+this value is a list whose length is the same as
 \li{grades}, where each element is the (indeterminate) result of evaluating the
@@
 %
-Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
-(\ie{}~\li{a}) and that the indeterminate results it produces are assigned
-corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2}, etc.).
+\rkc{The editor shows the first three elements of a result list, by default; the
+... can be clicked and expanded to see additional elements.}
+
+Notice that the hole expression on line \rkc{XXX} is assigned a unique static
+identifier (\ie{}~\li{a}) and that the indeterminate results it produces are
+assigned corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2},
+etc.).
 %
@@
 %
-The teacher uses a \HazelnutLive{} edit action to create a sum expression,
-filling in the first summand by weighting the \li{hw1} score to account for 10\%
-of the weighted average; the rest of the computation is a new hole expression
-(\li{??_b}), yet to be filled in.
+The teacher performs several \Hazel{} edit actions (each of which is a single
+keystroke, a shortcut to selecting one of the edit tools from the left panel,
+followed by a tool-specific argument terminated with the Enter key):
+%
+navigating the cursor to hole \li{a} and typing \li{+}, which creates a
+sum expression \li{??_b + ??_c} with the cursor placed, by default, under
+left operand (the hole labeled \li{b});
+%
+typing \li{*} to create a multiplication expression \li{??_d * ??_e + ??_c},
+with the cursor place under hole \li{d};
+%
+typing \li{\#10} to enter a numeric scaling factor, which then moves
+the cursor to the second operand, hole \li{e}; and
+%
+typing \li{vg.hw1} to use the \li{hw1} field of the function argument
+as the right operand of the multiplication.
+%
+The resulting partial expression, \li{(10.0 *. g.hw1) +. ??_b}---shown
+in the screenshot in the bottom half of \autoref{fig:grades-example}---marks the
+first of five weighted summands in the eventual finished \li{weighted_average}
+calculation.
 
-\begin{lstlisting}
-let weighted_average(g: student_data) =
-  (10.0 *. g.hw1) +. ??_b;
-\end{lstlisting}
-
-\noindent
+Each \Hazel{} edit action transforms an incomplete but well-formed and
+well-typed program into another such program, and, thus, \Hazel{} immediately
+runs the program after each edit.
 %
-\HazelnutLive{} immediately runs this incomplete program, displaying the updated
-list of indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
+After the above edit sequence, \Hazel{} display the updated list of
+indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
 %
@@
 
-\overviewExample{2}{Assigning Letter Grades}
+\input{fig-grades-example-2}
+
+\overviewExample{1b}{Assigning Letter Grades}
 %
 The teacher's next task is to map the weighted averages to the letter grades A
-through F (we will consider only ``whole'' letter grades, for simplicity).
+through F (we consider only ``whole'' letter grades, for simplicity).
 %
-The \li{grade_cutoffs} record type describes the minimum cutoff for each of
+The \li{grade_cutoffs} record type, shown at the top of
+\autoref{fig:grades-example-b}, describes the minimum cutoff for each of
 these six possible grades.
@@
 
-\begin{lstlisting}
-type grade_cutoffs =
-  { a: float, b: float, c: float, d: float, f: float };
-
-let cutoffs =
-  { a = ??_a, b = ??_b, c = ??_c, d = ??_d, f = ??_f };
-\end{lstlisting}
-
-\noindent
 %
@@
 %
-Before starting to fill in the cutoffs, the teacher jumps ahead to write a
+Before starting to fill in the cutoff values, the teacher jumps ahead to write a
 function \li{letter_grade} that will make the connection between \li{cutoffs}
@@
 Because she intends to look at the data to help select the cutoff values, the
-teacher sorts the \li{weighted_averages} and then maps them to
-\li{letter_grades}.
+teacher sorts the \li{weighted_averages} (on line \rkc{XXX}) and then maps
+them to \li{letter_grades} (on line \rkc{XXX}).
 
-\begin{lstlisting}
-let letter_grade(n: weighted_average) =
-  if n >= cutoffs.a then "A" else
-  if n >= cutoffs.b then "B" else
-  if n >= cutoffs.c then "C" else
-  if n >= cutoffs.d then "D" else
-  if n >= cutoffs.f then "F" else "Incomplete";
-
-let sorted_weighted_averages = List.sort weighted_averages;
-
-let letter_grades = List.map letter_grade sorted_weighted_averages;
-\end{lstlisting}
-
-\noindent
-%
-When \HazelnutLive{} runs this program, the guard of the outermost conditional
+When \Hazel{} runs this program, the guard of the outermost conditional
 (\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
@@
 %
-\autoref{fig:XXX} shows how \HazelnutLive{} renders large indeterminate
+\autoref{fig:XXX} shows how \Hazel{} renders large indeterminate
 expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
@@
 %
-Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
+Hovering over this abbreviation (\rkc{???}) displays the full
 indeterminate expression---as well as the evaluation environment that surrounds
@@
 
-To start deciding \li{cutoffs}, the teacher looks at the \li{weighted_averages}
-sorted in descending order.
+To start deciding \li{cutoffs}, the teacher clicks the \li{weighted_averages}
+expression, and views the results panel to see the data sorted in descending
+order.
 %
-The data shows a natural gap between \li{92.2} and \li{89.5}.
+The result shows a natural gap between \li{92.2} and \li{89.5}.
 %
-So, she chooses to use \li{92.0} as the cutoff for A.
+So, she chooses to use \li{92.0} as the cutoff for A, replacing hole \li{XXX} on
+line \rkc{XXX} with that numeric value.
 %
-Resuming the computation from before, \HazelnutLive{} resolves the conditional
+Resuming the computation from before, \Hazel{} resolves the conditional
 expressions for the first \rkc{XXX} indeterminate expressions, because each of
@@
 
-Before assigning other cutoff values, the teacher would like to get a sense of
-whether this first choice is a good one.
+Before assigning subsequent cutoff values, the teacher would like to get a sense
+of whether this first choice is a good one.
 %
-She jumps ahead to to write a function that computes the distribution of letter
-grades.
-
-\begin{lstlisting}
-let compute_distribution(list: list(float)) =
-  let n = List.length letter_grades in
-  List.map
-    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list)) /. n))
-    ["A","B","C","D","F","Incomplete"];
-
-let distribution = compute_distribution(letter_grades);
-\end{lstlisting}
-
-\noindent
+She jumps ahead to to write a function (\li{compute_distribution} on lines
+rkc{XXX}-\rkc{XXX}) that computes the distribution of letter grades based
+on \verb+cutoffs+..
 %
@@
 %
-Because this edit is outside of a hole expression, \HazelnutLive{} discards the
+Because this edit is not just filling a hole expression, \Hazel{} discards the
 previous execution state and reevaluates the entire program.

### 2018-03-11 fig-grades-example.tex
some work on Sec 2.1, with placeholders for mockup screenshots
I still don't have good sense of how these mockups are going to fake
all of the necessary features for the grades examples, so this text is
still pretty wishy-washy.

@@
 
+\rkc{Move all this code into the mockup screenshots below.}
+
+\lstset{basicstyle=\scriptsize\ttfamily}
 \begin{lstlisting}
 type student_data =
-  { name: string, 
-    hw1: float, hw2: float, hw3: float, 
-    midterm: float, final: float };
+  { name: string, hw1: float, hw2: float, hw3: float, midterm: float, final: float };
 
@@
   , { name: "Bob"  , hw1: 88.0, hw2: 75.0, hw3: 80.0, midterm: 89.0, final: 87.0 }
+  , ...
   ];
+
+let weighted_average(g: student_data) =
+  ??_a;          =====>        (10.0 *. g.hw1) +. ??_b;
+
+let weighted_averages = List.map weighted_average grades;
 \end{lstlisting}
+%% restore settings from main.tex
+\lstset{basicstyle=\footnotesize\ttfamily}
 
-\rkc{Place this code fragment (and some pieces of the subsequent fragments) in a
-single mockup or screenshot.}
+%% TODO once the code above is removed, scale up the screenshots
+\includegraphics[scale=0.20]{images/hazel-placeholder-0.png}
 
-\caption{\rkc{...}}
+\rkc{Draw arrows and captions on the top figure to show how to get
+to the bottom figure.
+ser navigates to hole a, types + to create a plus, types * to create a
+multiplication, types \#10 to create 10, types vh1 to create variable use.}
+
+\includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}
+
+\caption{Hazel mockups for Example 1a.}
 \label{fig:grades-example}

### 2018-03-11 implementation.tex
some work on Sec 2.1, with placeholders for mockup screenshots
I still don't have good sense of how these mockups are going to fake
all of the necessary features for the grades examples, so this text is
still pretty wishy-washy.

@@
 well-formed edit states}
+
+\rkc{
+The programs in \autoref{sec:examples}
+are written using minor syntactic conveniences that
+are not currently supported in our implementation.
+%
+For example, the implementation of recursive types in our current implementation
+requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
+inferred in standard ways (by pairing them with constructors and deconstructors
+for algebraic datatypes) in future work.
+%
+More details about other minor syntactic differences are described in
+\suppMaterials{}.
+}

### 2018-03-22 examples.tex

@@
 
-\HazelnutLive{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
+The screenshot at the top of \autoref{fig:grades-example} shows a mockup of
+\Hazel{}, a work-in-progress programming environment that we are building on top
+of our dynamic semantics.
+%
+\Hazel{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
 alternative Reason syntax \cite{reason-what} extended to support holes.
 %
-As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
-literate program \cite{knuth1984literate}: rich text comments appear at the top
-level and code appears within cells embedded into these comments.
+On the left side of the window is the \emph{edit action panel}, which
+provides a user interface for performing structured edits to the program
+displayed in the center of the window.
 %
-The reader is encouraged to work through these examples in our prototype
+On the right side of the window is the \emph{environment panel}, which provides
+static and dynamic information about the expression in the program currently in
+focus (\ie{}~under the cursor).
+
+Our current implementation, a reference implementation of the \HazelnutLive{}
+calculus, provides the edit action, code, and environment panels as shown in
+\autoref{fig:grades-example}, but does not yet provide support for many of the
+programming constructs and syntactic sugar described in the examples in this
+section.
+%
+\autoref{sec:implementation} describes the status of \HazelnutLive{} in more
+detail.
+%
+%
+%% As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
+%% literate program \cite{knuth1984literate}: rich text comments appear at the top
+%% level and code appears within cells embedded into these comments.
+%
+The reader is encouraged \rkc{(maybe not)}
+to work through the following examples in our prototype
 implementation or watch video demonstrations, both available in
-\suppMaterials{}.\footnote{
-%
-The programs in this section are written using minor syntactic conveniences that
-are not currently supported in our implementation.
-%
-For example, the implementation of recursive types in our current implementation
-requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
-inferred in standard ways (by pairing them with constructors and deconstructors
-for algebraic datatypes) in future work.
-%
-More details about other minor syntactic differences are described in
 \suppMaterials{}.
-%
-}
 

### 2018-03-22 fig-grades-example-2.tex

\begin{figure}[t]

\rkc{Move all this code into the mockup screenshots below.}

\lstset{basicstyle=\scriptsize\ttfamily}
\begin{lstlisting}
type grade_cutoffs =
  { a: float, b: float, c: float, d: float, f: float };

let cutoffs =
  { a = ??_a, b = ??_b, c = ??_c, d = ??_d, f = ??_f };

let letter_grade(n: weighted_average) =
  if n >= cutoffs.a then "A" else
  if n >= cutoffs.b then "B" else
  if n >= cutoffs.c then "C" else
  if n >= cutoffs.d then "D" else
  if n >= cutoffs.f then "F" else "Incomplete";

let sorted_weighted_averages = List.sort weighted_averages;

let letter_grades = List.map letter_grade sorted_weighted_averages;

let compute_distribution(list: list(float)) =
  let n = List.length letter_grades in
  List.map
    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list)) /. n))
    ["A","B","C","D","F","Incomplete"];

let distribution = compute_distribution(letter_grades);
\end{lstlisting}
%% restore settings from main.tex
\lstset{basicstyle=\footnotesize\ttfamily}

\includegraphics[scale=0.27]{images/hazel-placeholder-1b.png}

\caption{Hazel mockup for Example 1b.}
\label{fig:grades-example-b}
\end{figure}

### 2018-03-22 misc.tex

@@
 Finally, this environmental information is useful (though not required) for the edit action suggestion service of Track 3 and the direct manipulation programming service of Track 4.
+
+\subsection{Raw Material for Related Work}
+
+\begin{itemize}
+	\item Hazelnut paper
+	\item simply typed underdeterminism paper
+	\item work on partial evaluation and staging (incl. connections to modal logic -- ``modal analysis of staged computation'')
+	\item symbolic evaluation
+	\item full beta reduction
+	\item Agda
+	\item Idris
+	\item GHC holes
+	\item CMTT
+	\item gradual typing (and dynamic typing)
+		\begin{itemize}
+			\item siek and taha paper
+			\item snapl15 paper
+			\item gradualizer paper
+			\item maybe other things, e.g. several papers by ron garcia
+		\end{itemize}
+	\item DuctileJ stuff -- \url{https://homes.cs.washington.edu/~mernst/pubs/ductile-icse2011.pdf}
+	\item OLEG from McBride's thesis
+	\item Visual Studio (and others) support for edit-and-resume
+	\item Scratch lets you just skip over statement holes
+	\item prior work on confluence for the lambda calculus
+	\item work on debuggers that allow you to inspect environments
+		\begin{itemize}
+		\item might be something relevant in the paper ``A Debugger for Standard ML'' 
+		\item "Visualizing the evaluation of functional programs for debugging" by Whitington and Ridge
+		\item "A lightweight interactive debugger for Haskell'' and ``Multiple-View Tracing for Haskell: a New Hat'' might be relevant
+		\item ocamli -- \url{https://github.com/johnwhitington/ocamli}
+		\item Better supporting debugging aids learning a novel programming language. -- Scaffidi at VLHCC 2017
+		\item quote from Wadler in ``Why no one uses functional languages'':
+			\begin{quote}
+			“...there are few debuggers or
+profilers for strict [functional] languages, perhaps because constructing them is not considered
+research. This is a shame, since such tools are sorely needed, and there remains much of
+interest to learn about their construction and use.
+\end{quote}
+		\end{itemize}
+	\item papers that show up in a search for ``typed holes'' -- \url{https://scholar.google.com/scholar?hl=en&as_sdt=0%2C39&q=%22typed+holes%22&btnG=}
+	\item maybe also search for ``partial programs'' 
+	\item roly pererra work on program slicing
+	\item should say something about how holes show up in program synthesis under various names (what?) 
+	\item ``Achieving flexibility in direct-manipulation programming environments by relaxing the edit-time grammar'' -- \url{http://ieeexplore.ieee.org/document/1509511/}
+	\item ``Call-by-value is dual to call-by-name'' might be relevant? \url{http://homepages.inf.ed.ac.uk/wadler/papers/dual/dual.pdf}
+	\item mention how unspecified evaluation order is something that people do when talking about parallelism?
+\end{itemize}

### 2018-03-28 abstract.tex
draft of sec 3.0 through 3.1

--- 2018-02-09 abstract.tex
@@
+% !TEX root = hazelnut-dynamics.tex
+
 \begin{abstract}
 
-\emph{Live programming environments} aim to provide programmers (and, in some
-cases, audiences) with continuous feedback about a program's dynamic behavior
-throughout the editing process.
+\emph{Live programming environments} aim to provide programmers (and sometimes audiences) 
+with continuous feedback about a program's dynamic behavior as it is being edited. 
 %
-Unfortunately, because programming languages typically assign dynamic meaning
-only to programs that are \emph{complete}---syntactically well-formed and free
-of static type errors---live programming features are often severely restricted,
-if available at all, when programs are incomplete.
+The problem is that programming languages typically assign dynamic meaning only 
+ to programs that are \emph{complete}, i.e. syntactically well-formed and free
+of static type errors, whereas 
 %
-Incomplete edit states are frequently encountered when writing programs, so they
-are, arguably, when live programming services would be most valuable.
+during much of the editing process, the program is not yet complete. As a result,  
+live programming services flicker in and out, 
+in many cases going out for substantial lengths of time.
 %
-Defining a semantics for incomplete functional programs would help bridge the
-gap between well-formed states and allow live programming environments to
-provide continuous, uninterrupted, and meaningful feedback.
-
-Recent work by \citet{popl-paper} makes progress toward this goal by developing
-a type system for lambda terms with \emph{holes} in both expression and type
-position, with empty holes standing for missing expressions or types, and
-non-empty holes operating as ``membranes'' around type inconsistencies.
+% Defining a semantics for incomplete functional programs would help bridge the
+% gap between well-formed states and allow live programming environments to
+% provide continuous, uninterrupted, and meaningful feedback.
+This paper confronts this problem from type-theoretic first principles by developing a 
+\emph{dynamic semantics for incomplete functional programs}, 
+based in part on the static semantics for incomplete functional programs
+developed in recent work by \citet{popl-paper}. 
 %
-The contribution of this paper is a corresponding \emph{dynamic semantics for
-functional programs with holes}.
+Incomplete functional programs are modeled as expressions with \emph{holes}, 
+with empty holes standing for missing expressions or types, and  non-empty holes 
+operating as ``membranes'' around static and dynamic type inconsistencies. 
 %
-Rather than aborting with an exception when evaluation encounters a hole (as in
-several existing systems), evaluation proceeds ``around'' expression holes,
-performing as much computation as is possible and tracking the dynamic
-environment around each hole as it flows through the program.  These accumulated
-hole environments can be reported directly to the programmer by the editor as
-they work to fill holes.  They also enable an ``edit-and-resume'' feature that
+Rather than aborting with an exception when evaluation encounters any of these holes (as in
+several existing systems), evaluation proceeds ``around'' the holes,
+performing as much of the remaining computation as is possible and tracking the 
+closure around each hole instance as it flows through the program.  These hole closures
+ can be reported both directly and indirectly, via various editor services, to help the programmer decide how to fill a hole. 
+They also enable an ``edit-and-resume'' feature that 
 avoids the need to restart evaluation after edits that amount to hole filling. 
-
-Formally, our semantics combines features from gradual type theory (to handle
-incomplete types) and contextual modal type theory (CMTT, which provides a
-logical foundation for hole environment tracking) with novel structure necessary
-to prove key metatheoretic properties, including (1) a type safety property that
-accounts for expression holes (i.e. free metavariables, from the perspective of
-CMTT), and (2) commutativity and confluence properties that guarantee that the
-``edit-and-resume'' feature is sound.
+Formally, the semantics combines features from gradual type theory (to handle
+type holes) and contextual modal type theory (which provides a
+logical foundation for hole closures) with novel formal structure necessary
+to continue evaluation past holes and to prove key metatheoretic properties. 
+% , including 
+% (1) a type safety property that accounts for expression holes (i.e. free metavariables, 
+% from the perspective of CMTT) and cast failure holes (which provide a novel), and (2) a commutativity property that establishes that the
+% ``edit-and-resume'' feature is sound.
 %
-We have mechanized our formal development using the Agda proof assistant.
+We have mechanized the core formal development using the Agda proof assistant.
 %
-We additionally describe a simple reference implementation, called Hazelnut
-Live, that exposes a language of structured edit actions that insert holes as
-necessary to guarantee that every edit state has some (possibly incomplete)
-type, based on the edit action calculus in the prior work by \citet{popl-paper}.
+We additionally describe a simple reference implementation, called \HazelnutLive{},
+that inserts holes as
+necessary during the editing process to guarantee that every edit state has some (possibly incomplete)
+type, based in part on the edit action calculus in prior work by \citet{popl-paper}.
 %
-Taken together, the result is a proof-of-concept live functional programming
-environment where dynamic feedback is truly continuous, i.e. it is available for
-every possible edit state.
+Taken together with the type safety property that this paper establishes, the result is 
+a proof-of-concept live typed functional programming environment where dynamic feedback 
+is truly continuous, i.e. it is available for every possible edit state.
 % i.e. \emph{every edit state is guaranteed to have

### 2018-03-28 appendix-defns.tex
draft of sec 3.0 through 3.1

% !TEX root = hazelnut-dynamics.tex

\newcommand{\additionalDefnsSec}{Additional Definitions for Hazelnut Live}
\section{\protect\additionalDefnsSec} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:additional-defns}

\input{fig-complete}

### 2018-03-28 calculus.tex
draft of sec 3.0 through 3.1

@@
 \label{sec:calculus}
-\cy{exposition}
 
-\subsection{Calculus}
+We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory.
+We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}.
+Sec.~\ref{sec:agda-mechanization} provides some additional details on this mechanization and the supplemental material includes the full mechanization.
+% The proofs of the metatheorems are established by this mechanization; the proofs are only very briefly outlined in this section.
 
-\subsubsection{Syntax}
+The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes, which we refer to as H-types and H-expressions, respectively.
+We distinguish between \emph{external} H-expressions, $e$, and \emph{internal} H-expressions, $d$.
+External H-expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec.
+2?}{} for examples of manual, semi-automated and fully automated hole entry methods).
+Each well-typed external H-expression (as specified in Sec.~\ref{sec:external-statics}) expands to a well-typed internal H-expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}).
+We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts.
+This machinery is initialized by the expansion step, rather than entered explicitly by the programmer.
+In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
 
@@
 
-{\color{gray}\blindtext}
+% \rkc{this syntactic sugar is used in four places: ITCastSucceed, ITCastFail,
+% ITGround, and ITExpand. that's not many, and those rules don't look much more
+% cluttered without the sugar, so consider eliminating it. if so, just toggle the
+% definition of the dcastthree macro to the unsugared option.}
 
-Shorthand:
+\subsection{Static Semantics of the External Language}
+\label{sec:external-statics}
 
-$$
-\dcastthree{\dexp}{\htau_1}{\htau_2}{\htau_3} \defeq
-  \dcasttwo{\dcasttwo{\dexp}{\htau_1}{\htau_2}}{\htau_2}{\htau_3}
-$$
+\input{fig-bidirectional-typing}
+\input{fig-type-consistency}
 
-\rkc{this syntactic sugar is used in four places: ITCastSucceed, ITCastFail,
-ITGround, and ITExpand. that's not many, and those rules don't look much more
-cluttered without the sugar, so consider eliminating it. if so, just toggle the
-definition of the dcastthree macro to the unsugared option.}
+Let us start with a brief overview of the static semantics of the \HazelnutLive external language.
+H-types, $\htau$, classify external H-expressions, $e$, according to the type system in Fig.
+\ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up).
+The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
+The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner.
+The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$.
+Algorithmically, the type is an input of type analysis.% Algorithmically, the type is an output of type synthesis but an input of type analysis.
 
-Explain that $c$ is a placeholder constant (i.e., unit), $b$ is a placeholder
-type (i.e., unit).
+The benefit of specifying the \HazelnutLive external language bidirectionally, rather than as a type assignment system, is that the programmer does not need to annotate each hole with a type.
+Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole names are new to \HazelnutLive).
+Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$.
+If an empty hole appears where an expression of some other type is expected, e.g.
+under an explicit ascription $e : \htau$ (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$.
+Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}.
+Two types are consistent if they differ only up to type holes in corresponding positions.
+Note that type consistency is reflexive and symmetric but it is not transitive (quite unlike subtyping, which is anti-symmetric and transitive, though subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).
+The hole type is consistent with every type, and so, by the subsumption rule, empty expression holes can appear where an expression of any type is expected.
 
-\subsubsection{Statics}
+Non-empty expression holes, written $\hhole{\hexp}{u}$, behave similarly.
+Rule {SNEHole} specifies that non-empty expression holes also synthesize hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type.
+Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}
 
-\input{fig-type-consistency}
-\input{fig-bidirectional-typing}
+For the familiar forms of the lambda calculus, the rules again follow the prior work.
+For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e.
+$b$ is the unit type).
+\Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
 
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
+Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.
 
-A typing context, $\hGamma$, is a list of bindings $x : \htau$.
+For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
+Half-annotated lambdas can appear in synthetic position (where an expected type is not known, e.g.
+at the top-level) according to Rule {SLam}, which is standard \cite{Chlipala:2005da}.
+Unannotated lambdas can only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were the type $\tarr{\tehole}{\tehole}$.
+To avoid the need for two separate rules, the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in Fig.~\ref{fig:arrmatch} produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
+This allows us to specify a single rule, Rule {ALam}, governing unannotated lambdas.
+Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might also define a synthetic rule governing unannotated lambdas, e.g.
+as outlined by \citet{DBLP:conf/icfp/DunfieldK13}.
 
-\subsubsection{Expansion}
+The matched arrow type judgement also appears in the rule governing function application, Rule {SAp}, which treats expressions in function position of hole type as if they were of type $\tarr{\tehole}{\tehole}$.
+
+\subsection{Expansion}
+\label{sec:expansion}
 
@@
 
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-
 A hole context, $\hDelta$, is a list of bindings $\Dbinding{u}{\hGamma}{\htau}$.
-
-\subsubsection{Dynamics}
-
-\input{fig-dynamics-aux}
-%% \input{fig-dynamics-steps}
-\input{fig-dynamics-contexts}
-
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-
-\begin{lemma}[Focus Formation] ~
-  If $\selectEvalCtx{\dexp}{\evalctx}{\dexp'}$
-  then $\isevalctx{\evalctx}$.
-\end{lemma}
-
-\begin{lemma}[Matched Ground Type Invariant] ~
-  If $\groundmatch{\htau}{\htau'}$
-  then $\isGround{\htau'}$
-  and $\tconsistent{\htau}{\htau'}$
-  and $\htau\neq\htau'$.
-\end{lemma}
-
-\subsubsection{Examples}
-
-\cy{definition of complete types and terms? do we care about completeness
-  of $\hexp$s or $\dexp$s?}
-
-%% \begin{figure}[!ht]
-%%   \begin{definition}
-%%     $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
-%%   \end{definition}
-%%   \caption{substitution type assignment}
-%%   \label{fig:subassign}
-%% \end{figure}
-
-%% \begin{figure}[!ht]
-  \begin{definition}[Substitution Typing]
-    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Gamma'}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
-  \end{definition}
-%%   \caption{substitution type assignment}
-%% \end{figure}
-
-%%   \halam{x}{\htau}{\evalctx} ~\vert~
-
-\clearpage
-%% \subsection{Metatheory}
-\subsection{Metatheory: Type Soundness}
-
-{\color{gray}\blindtext}
 
@@
 \end{theorem}
+
+\subsection{Dynamic Semantics}
+\label{sec:evaluation}
+
+\input{fig-dynamics-aux}
+%% \input{fig-dynamics-steps}
+\input{fig-dynamics-contexts}
+
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+{\color{gray}\blindtext}
+
+\begin{lemma}[Focus Formation] ~
+  If $\selectEvalCtx{\dexp}{\evalctx}{\dexp'}$
+  then $\isevalctx{\evalctx}$.
+\end{lemma}
+
+\begin{lemma}[Matched Ground Type Invariant] ~
+  If $\groundmatch{\htau}{\htau'}$
+  then $\isGround{\htau'}$
+  and $\tconsistent{\htau}{\htau'}$
+  and $\htau\neq\htau'$.
+\end{lemma}
 
@@
 
-\input{fig-complete}
-
 \autoref{fig:complete} defines \emph{complete} H-types, H-expressions, and
@@
 \end{theorem}
+
+\cy{definition of complete types and terms? do we care about completeness
+  of $\hexp$s or $\dexp$s?}
+
+%% \begin{figure}[!ht]
+%%   \begin{definition}
+%%     $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+%%   \end{definition}
+%%   \caption{substitution type assignment}
+%%   \label{fig:subassign}
+%% \end{figure}
+
+%% \begin{figure}[!ht]
+  \begin{definition}[Substitution Typing]
+    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Gamma'}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+  \end{definition}
+%%   \caption{substitution type assignment}
+%% \end{figure}
+
+\subsection{Agda Mechanization}
+\label{sec:agda-mechanization}
+
+%%   \halam{x}{\htau}{\evalctx} ~\vert~
+
+{\color{gray}\blindtext}
 

### 2018-03-28 calculus.tex
remove the H- prefix and a few minor wording tweaks to 3.1

@@
 
-The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes, which we refer to as H-types and H-expressions, respectively.
-We distinguish between \emph{external} H-expressions, $e$, and \emph{internal} H-expressions, $d$.
-External H-expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec.
+The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes.
+We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$.
+External expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec.
 2?}{} for examples of manual, semi-automated and fully automated hole entry methods).
-Each well-typed external H-expression (as specified in Sec.~\ref{sec:external-statics}) expands to a well-typed internal H-expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}).
+Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics}) expands to a well-typed internal expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}).
 We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts.
-This machinery is initialized by the expansion step, rather than entered explicitly by the programmer.
+This machinery is inserted by the expansion step, rather than entered explicitly by the programmer.
 In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
@@
 Let us start with a brief overview of the static semantics of the \HazelnutLive external language.
-H-types, $\htau$, classify external H-expressions, $e$, according to the type system in Fig.
+Types, $\htau$, classify external expressions, $e$, according to the type system in Fig.
 \ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up).
-The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
+It is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
 The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner.
@@
 
-The matched arrow type judgement also appears in the rule governing function application, Rule {SAp}, which treats expressions in function position of hole type as if they were of type $\tarr{\tehole}{\tehole}$.
+The rule governing function application, Rule {SAp}, similarly treats expressions in function position of hole type as if they were of type $\tarr{\tehole}{\tehole}$ using the matched arrow type judgement.
 
@@
 
-\autoref{fig:complete} defines \emph{complete} H-types, H-expressions, and
-dynamic H-expressions to be those terms without type or expression holes.
+\autoref{fig:complete} defines \emph{complete} types, expressions, and
+dynamic expressions to be those terms without type or expression holes.
 

### 2018-03-29 calculus.tex
put expansion and type assignment on the same page

@@
 External expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec.
-2?}{} for examples of manual, semi-automated and fully automated hole entry methods).
-Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics}) expands to a well-typed internal expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}).
+2?}{} for discussion of manual, semi-automated and fully automated hole entry methods).
+Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}).
 We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts.
@@
 \ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up).
-It is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
-The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner.
+The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
+The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}.
 The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$.
@@
 
-The benefit of specifying the \HazelnutLive external language bidirectionally, rather than as a type assignment system, is that the programmer does not need to annotate each hole with a type.
-Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole names are new to \HazelnutLive).
+The benefit of specifying the \HazelnutLive external language bidirectionally is that the programmer does not need to annotate each hole with a type.
+Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (this notation is taken from \Hazelnut; hole names are new to \HazelnutLive).
 Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$.
@@
 Two types are consistent if they differ only up to type holes in corresponding positions.
-Note that type consistency is reflexive and symmetric but it is not transitive (quite unlike subtyping, which is anti-symmetric and transitive, though subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).
+Note that type consistency is reflexive and symmetric but it is not transitive (quite unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).
 The hole type is consistent with every type, and so, by the subsumption rule, empty expression holes can appear where an expression of any type is expected.
@@
 
-For the familiar forms of the lambda calculus, the rules again follow the prior work.
+For the familiar forms of the lambda calculus, the rules again follow prior work.
 For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e.
@@
 Unannotated lambdas can only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were the type $\tarr{\tehole}{\tehole}$.
-To avoid the need for two separate rules, the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in Fig.~\ref{fig:arrmatch} produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
-This allows us to specify a single rule, Rule {ALam}, governing unannotated lambdas.
-Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might also define a synthetic rule governing unannotated lambdas, e.g.
+To avoid the need for two separate rules, Rule {ALam} uses the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in Fig.~\ref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
+Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might include a synthetic rule governing unannotated lambdas, e.g.
 as outlined by \citet{DBLP:conf/icfp/DunfieldK13}.
 
-The rule governing function application, Rule {SAp}, similarly treats expressions in function position of hole type as if they were of type $\tarr{\tehole}{\tehole}$ using the matched arrow type judgement.
+The rule governing function application, Rule {SAp}, similarly treats expressions of hole type in function position as if they were of type $\tarr{\tehole}{\tehole}$ using the matched arrow type judgement.
 
@@
 \input{fig-typing-dexp}
+
+Well-typed external expressions expand to well-typed internal expressions, $d$, for evaluation.
+The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}.
 

### 2018-03-30 calculus.tex
draft of sec 3.2 through 3.2.1 and a stub starting to talk about cast insertion

@@
 For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
-Half-annotated lambdas can appear in synthetic position (where an expected type is not known, e.g.
-at the top-level) according to Rule {SLam}, which is standard \cite{Chlipala:2005da}.
-Unannotated lambdas can only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were the type $\tarr{\tehole}{\tehole}$.
+Half-annotated lambdas can appear in synthetic position  according to Rule {SLam}, which is standard \cite{Chlipala:2005da}.
+Unannotated lambdas can only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.
 To avoid the need for two separate rules, Rule {ALam} uses the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in Fig.~\ref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
-Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might include a synthetic rule governing unannotated lambdas, e.g.
+Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might include a synthetic rule for unannotated lambdas, e.g.
 as outlined by \citet{DBLP:conf/icfp/DunfieldK13}.
@@
 
-A hole context, $\hDelta$, is a list of bindings $\Dbinding{u}{\hGamma}{\htau}$.
+Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements.
+The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$.
+We say more about hole contexts below.
+The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$.
+The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
+\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
+  \begin{enumerate}[nolistsep]
+    \item
+      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
+    \item
+      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
+  \end{enumerate}
+\end{thm}
+\noindent
+The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, while every internal expression can be assigned at most one type, i.e.
+the following standard unicity property holds of the type assignment system:
+\begin{thm}[Type Assignment Unicity]
+  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
+  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
+  then $\htau=\htau'$.
+\end{thm}
+\noindent
+Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
+For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$, where $\emptyset$ is the empty hole context.
 
-\begin{theorem}[Expandability] ~
-  \begin{enumerate}
+Before describing the rules in detail, let us state a few other useful theorems. The following theorem establishes that an expansion exists for every well-typed external expression.
+ \begin{thm}[Expandability] \label{thm:expandability}~
+  \begin{enumerate}[nolistsep]
     \item
@@
   \end{enumerate}
-\end{theorem}
-
-\begin{theorem}[Correspondence] ~
-  \begin{enumerate}
-    \item
-      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
-      then $\hsyn{\hGamma}{\hexp}{\htau}$.
-    \item
-      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
-      then $\hana{\hGamma}{\hexp}{\htau}$.
-  \end{enumerate}
-\end{theorem}
-
-\begin{theorem}[Typed Expansion] ~
-  \begin{enumerate}
-    \item
-      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
-      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
-    \item
-      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
-      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
-  \end{enumerate}
-\end{theorem}
-
-\begin{theorem}[Expansion Unicity] ~
-  \begin{enumerate}
+\end{thm}
+\noindent
+The following theorem establishes that when an expansion exists, it is unique.
+\begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
+  \begin{enumerate}[nolistsep]
     \item
@@
   \end{enumerate}
-\end{theorem}
+\end{thm}
+\noindent
+The following theorem establishes that expansion generalizes external typing.\todo{rename correspondence to generality}
+\begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
+  \begin{enumerate}[nolistsep]
+    \item
+      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      then $\hsyn{\hGamma}{\hexp}{\htau}$.
+    \item
+      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+      then $\hana{\hGamma}{\hexp}{\htau}$.
+  \end{enumerate}
+\end{thm}
 
-\begin{theorem}[Type Assignment Unicity]
-  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
-  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
-  then $\htau=\htau'$.
-\end{theorem}
+The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} (so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward).
+Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation onto unannotated external lambdas based on the given type.
+The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.
+
+\subsubsection{Hole Expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
+
+The hole name, $u$, on a hole closure identifies the external hole that the hole closure corresponds to.
+Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution (though in the initial expansion, the uniqueness condition happens to hold because evaluation has yet to occur).
+For example, the result from Fig.~\ref{fig:grades-example} showed four closures for the hole named 1\todo{update with actual number (four?) and hole name later}.
+There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation so we omit hole closure numbers from the core calculus.
+
+The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$.
+The purpose of the hole context is to record a typing context, $\Gamma$, and a type, $\tau$, for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with metavariable contexts (we say more about the connection with CMTT below).
+In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded.
+In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules.
+However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position during expansion.
+Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context.
+This is again so that we can use a type assignment system for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign the corresponding type from the hole context to the hole closure.
+
+Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
+The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs.
+Initially, no evaluation has occured, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
+\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
+\end{defn}
+\noindent
+The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
+\begin{defn}[Substitution Typing]
+$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
+\end{defn}
+\noindent
+It is easy to verify that the identity substitution satisfies this requirement, i.e.
+that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.
+We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}.
+
+\subsubsection{Cast Insertion} 
+Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
+The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$ as previously discussed, and the outer argument, $c$, can be checked against type $\tehole$ by subsumption.
+However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}.
+Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is not well-typed ($c$ does not have arrow type).
+The problem is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
+Rather than arbitrarily choosing one of the possible types for $x$ and inserting non-empty holes around uses that are inconsistent with that choice, we could leave the type hole unfilled and rely on the cast insertion machinery described here.
+On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.}\todo{this footnote should probably be integrated into and refer to Sec.
+2?}{} Recall that we can interpret hole types as unknown types from gradual type theory.
+This provides us with the solution to this problem: cast insertion.
 

### 2018-03-30 hazelnut-dynamics.tex
placeholder for implementation section
some layout tweaks and placeholders
metatheory definitions, wip
some work on Sec 2.1, with placeholders for mockup screenshots
I still don't have good sense of how these mockups are going to fake
all of the necessary features for the grades examples, so this text is
still pretty wishy-washy.
draft of sec 3.0 through 3.1
draft of sec 3.2 through 3.2.1 and a stub starting to talk about cast insertion

@@
 %% For double-blind review submission
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
+\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
 %% For single-blind review submission
@@
 
+%% Can remove this eventually
+\usepackage{blindtext}
+
+\usepackage{enumitem}
+
 % A macro for the name of the system being described by ``this paper''
 \newcommand{\HazelnutLive}{\textsf{Hazelnut Live}\xspace}
+\newcommand{\Hazelnut}{\textsf{Hazelnut}\xspace}
+% The mockup, work-in-progress system.
+\newcommand{\Hazel}{\textsf{Hazel}\xspace}
 
@@
 % \newtheorem{condition}[theorem]{Condition}
+
+\newtheoremstyle{slplain}% name
+  {.16\baselineskip\@plus.1\baselineskip\@minus.1\baselineskip}% Space above
+  {.16\baselineskip\@plus.1\baselineskip\@minus.1\baselineskip}% Space below
+  {\slshape}% Body font
+  {\parindent}%Indent amount (empty = no indent, \parindent = para indent)
+  {\bfseries}%  Thm head font
+  {.}%       Punctuation after thm head
+  { }%      Space after thm head: " " = normal interword space;
+        %       \newline = linebreak
+  {}%       Thm head spec
+\theoremstyle{slplain}
+\newtheorem{thm}{Theorem}  % Numbered with the equation counter
+\numberwithin{thm}{section}
+\newtheorem{defn}[thm]{Definition}
+% \newtheorem{cor}[section]{Corollary}     
+% \newtheorem{lem}[section]{Lemma}         
+% \newtheorem{prop}[section]{Proposition}
+
+% \setlength{\abovedisplayskip}{0pt}
+% \setlength{\belowdisplayskip}{0pt}
+% \setlength{\abovedisplayshortskip}{0pt}
+% \setlength{\belowdisplayshortskip}{0pt}
 
@@
 \acmYear{2018}
-\acmMonth{1}
+\acmMonth{3}
 \acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
 \input{macros}
+
+\setlength{\abovecaptionskip}{4pt plus 3pt minus 2pt} % Chosen fairly arbitrarily
+\setlength{\belowcaptionskip}{-4pt plus 3pt minus 2pt} % Chosen fairly arbitrarily
 
@@
 \input{calculus}
+\input{implementation}
 \input{related-work}
@@
 \appendix
-\cy{for the submission, appendix should be separated into a separate supplemental PDF}
-
+\input{appendix-defns}
 \input{extensions}
+\input{misc}
 

### 2018-03-30 calculus.tex
fixed statement of complete preservation

@@
   and $\stepsToD{}{\dexp}{\dexp'}$
-  then $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau}$.
+  then $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau}$
+  and $\isComplete{\dexp'}$.
 \end{theorem}

### 2018-04-01 calculus.tex
finished draft of 3.2

@@
 We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory.
-We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}.
-Sec.~\ref{sec:agda-mechanization} provides some additional details on this mechanization and the supplemental material includes the full mechanization.
+We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} (see Sec.~\ref{sec:agda-mechanization} for some additional details and the supplemental material for the full mechanization).
 % The proofs of the metatheorems are established by this mechanization; the proofs are only very briefly outlined in this section.
@@
 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods).
-Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}).
+Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
 We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts.
@@
 
-Let us start with a brief overview of the static semantics of the \HazelnutLive external language.
+Let us start with an overview of the static semantics of the \HazelnutLive external language.
 Types, $\htau$, classify external expressions, $e$, according to the type system in Fig.
-\ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up).
+\ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up below).
 The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
@@
 The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$.
-Algorithmically, the type is an input of type analysis.% Algorithmically, the type is an output of type synthesis but an input of type analysis.
+Algorithmically, the type is an output of type synthesis but an input of type analysis.
+At the top level, we start with synthesis.
+% Algorithmically, the type is an output of type synthesis but an input of type analysis.
 
 The benefit of specifying the \HazelnutLive external language bidirectionally is that the programmer does not need to annotate each hole with a type.
-Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (this notation is taken from \Hazelnut; hole names are new to \HazelnutLive).
+Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole notation is taken from \Hazelnut; hole names are new to \HazelnutLive).
 Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$.
 If an empty hole appears where an expression of some other type is expected, e.g.
-under an explicit ascription $e : \htau$ (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$.
-Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}.
-Two types are consistent if they differ only up to type holes in corresponding positions.
+under an explicit ascription (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$.
+Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.
+The hole type is consistent with every type, and so, by the subsumption rule, empty expression holes can appear where an expression of any type is expected.
+The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}.
 Note that type consistency is reflexive and symmetric but it is not transitive (quite unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).
-The hole type is consistent with every type, and so, by the subsumption rule, empty expression holes can appear where an expression of any type is expected.
 
@@
 
-The rule governing function application, Rule {SAp}, similarly treats expressions of hole type in function position as if they were of type $\tarr{\tehole}{\tehole}$ using the matched arrow type judgement.
+The rule governing function application, Rule {SAp}, similarly treats expressions of hole type in function position as if they were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.
 
@@
 The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}.
+We will return to evaluation in Sec.~\ref{sec:evaluation}.
 
@@
 \noindent
-The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, while every internal expression can be assigned at most one type, i.e.
+The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e.
 the following standard unicity property holds of the type assignment system:
@@
 Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context.
-This is again so that we can use a type assignment system for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign the corresponding type from the hole context to the hole closure.
+This is again so that we can use a type assignment system for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.
 
@@
 However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}.
-Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is not well-typed ($c$ does not have arrow type).
+Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed ($c$ does not have arrow type).
 The problem is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
-Rather than arbitrarily choosing one of the possible types for $x$ and inserting non-empty holes around uses that are inconsistent with that choice, we could leave the type hole unfilled and rely on the cast insertion machinery described here.
-On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.}\todo{this footnote should probably be integrated into and refer to Sec.
-2?}{} Recall that we can interpret hole types as unknown types from gradual type theory.
-This provides us with the solution to this problem: cast insertion.
+%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
+}\todo{talk about elsewhere?
+maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory.
+This leads us to the solution to this problem: cast insertion.
+
+The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of some consistent type $\htau_2$.\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}
+
+Casts are inserted during the expansion of function applications and ascriptions.
+The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the type assigned to the ascribed expression to the ascribed type.
+Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.
+Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.
+
+Cast insertion by the application expansion rule, Rule~{ESAp}, has much the same intuition, but requires a bit more care. To understand the rule, consider the expansion for the example above:
+\[\hap{\dcasttwo{(\halam{x}{\tehole}{\underline{\hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}}})}{\tarr{\tehole}{\tehole}}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}
+\]
+Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. A cast is inserted on both the function expression, $x$, and the the argument, $c$.
+
+The cast on $x$ allows us to treat the variable $x$, which is of type $\tehole$, as being of the matched arrow type $\tarr{\tehole}{\tehole}$, as described in Sec.~\ref{sec:external-statics}.
+The first three premises of Rule~{ESAp} accomplish this by first synthesizing a type for the function expression, here $\tehole$, then determining the matched arrow type, $\tarr{\tehole}{\tehole}$, and then performing analytic expansion on the function expression with this matched arrow type.
+The resulting expansion will have some type $\tau_1'$ consistent with the matched arrow type.
+In this case, because $x$ is of variable form, analytic expansion goes through subsumption so $\tau_1'$ is simply $\tehole$.
+The conclusion of the rule inserts the corresponding cast.
+We perform type synthesis first and then use analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.
+
+The conclusion of Rule~{ESAp} also inserts a cast on the argument's expansion, from the type it assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type, $\tehole$, of the matched arrow type of the function expression, here $\tehole$.
+
+Observe that together, these casts allow us assign a type to the function body according to the rules in Fig.~\ref{fig:hasType}, where we could not do so under the same context without casts.
+
+The outer application in the example above goes through the same rule.
+In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$.
+We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), but in practice it is straightforward to do so (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).
 
@@
 
+Failed casts next...
+
+\input{fig-ground-types}
 \input{fig-dynamics-aux}

### 2018-04-02 calculus.tex
draft of 3.3

@@
 Rule {SNEHole} specifies that non-empty expression holes also synthesize hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type.
-Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}
+Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}\todo{briefly say something about binding inconsistencies?}
 
@@
 
-\subsubsection{Hole Expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
+\subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
 
@@
 The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$.
-The purpose of the hole context is to record a typing context, $\Gamma$, and a type, $\tau$, for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with metavariable contexts (we say more about the connection with CMTT below).
+The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$,   for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
 In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded.
 In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules.
-However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position during expansion.
+However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position.
 Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context.
-This is again so that we can use a type assignment system for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.
+This is again so that we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.
 
@@
 \noindent
-It is easy to verify that the identity substitution satisfies this requirement, i.e.
-that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.
-We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}.
+It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.
 
-\subsubsection{Cast Insertion} 
+Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a.
+deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}.
+Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT\todo{mention reason in Sec 4}.
+
+\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
 Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
-The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$ as previously discussed, and the outer argument, $c$, can be checked against type $\tehole$ by subsumption.
+The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and the outer argument, $c$, can be checked against type $\tehole$ by subsumption.
 However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}.
@@
 
-\subsection{Dynamic Semantics}
+\subsection{Evaluation}
 \label{sec:evaluation}
+
+To recap, the result of expansion is a well-typed internal expression with hole closures and casts.
+As described in Sec.~\ref{sec:hole-expansion} and Sec.~\ref{sec:cast-insertion} respectively, empty hole closures correspond to metavariable closures from CMTT, with the metavariable free (i.e.
+tracked in $\Delta$) \cite{Nanevski2008} and casts correspond to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}.
+However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.
+
+The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) correspondence.
+The paper therefore proved only a subject reduction property, which corresponds to type preservation for elimination transitions, generalized to open terms\todo{citation?}{}.
+This is not a full dynamic semantics, and in particular, there is no notion of progress, i.e.
+that well-typed terms cannot get ``stuck'' in an undefined state.
+In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need an operational semantics for terms with free metavariables.
+\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion\footnote{Permutation conversions are necessary to encode the commutative reductions of CMTT, which in turn are necessary to prove a strong normalization property.
+These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.}, which has been studied by \citet{DBLP:journals/iandc/Groote02}, but under this interpretation the same problem arises -- metavariables become variables of a function type, so we cannot rely on any standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
+
+Of course, we also need to integrate casts into the dynamic semantics.
+Fortunately, the operational semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery.
+Briefly, the subtlety is that in the cast calculus, the only irreducible terms of hole type are casts, but in $\HazelnutLive$, hole closures induce additional irreducible terms.
+We will return to the technical implications of this fact below.
+The second missing piece in existing work on casts is that, when cast failure occurs, evaluation aborts (i.e.
+blame propagates).
+Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}, is to have cast failure spawn a membrane around the dynamic type error, much like a non-empty hole serves as a membrane around a static type error, allowing evaluation to safely continue past the cast failure when desired.
+
+This section defines a reduction semantics for well-typed internal expressions.
+
+Matthias Felleisen, Robert Hieb,
+The revised report on the syntactic theories of sequential control and state,
 
@@
   (iii) $\isBoxedValue{\dexp}$.
-\end{theorem}
-
-\begin{theorem}[Canonical Value Forms]
-  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isValue{\dexp}$
-  then $\htau\neq\tehole$ and
-  \begin{enumerate}[label=(\alph*)]
-    \item If $\htau=b$ then $\dexp=c$.
-    \item If $\htau=\tarr{\htau_1}{\htau_2}$
-          then $\dexp=\halam{x}{\htau_1}{\dexp'}$
-          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$.
-  \end{enumerate}
 \end{theorem}
@@
 
+\subsection{Evaluating Complete Terms}\label{sec:complete-terms}
+
 \autoref{fig:complete} defines \emph{complete} types, expressions, and
 dynamic expressions to be those terms without type or expression holes.
+\begin{theorem}[Canonical Value Forms]
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isValue{\dexp}$
+  then $\htau\neq\tehole$ and
+  \begin{enumerate}[label=(\alph*)]
+    \item If $\htau=b$ then $\dexp=c$.
+    \item If $\htau=\tarr{\htau_1}{\htau_2}$
+          then $\dexp=\halam{x}{\htau_1}{\dexp'}$
+          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$.
+  \end{enumerate}
+\end{theorem}
 
@@
 %% \begin{figure}[!ht]
-  \begin{definition}[Substitution Typing]
-    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Gamma'}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
-  \end{definition}
 %%   \caption{substitution type assignment}

### 2018-04-04 calculus.tex
a bunch of progress on sec 3.3

@@
 We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory.
-We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} (see Sec.~\ref{sec:agda-mechanization} for some additional details and the supplemental material for the full mechanization).
+We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} (see Sec.~\ref{sec:agda-mechanization} for some additional details, and the supplemental material for the full mechanization).
 % The proofs of the metatheorems are established by this mechanization; the proofs are only very briefly outlined in this section.
@@
 The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
-The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}.
+The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ under typing context $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}.
 The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$.
@@
 Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.
-The hole type is consistent with every type, and so, by the subsumption rule, empty expression holes can appear where an expression of any type is expected.
+The hole type is consistent with every type, and so, by the subsumption rule, expression holes can appear where an expression of any type is expected.
 The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}.
-Note that type consistency is reflexive and symmetric but it is not transitive (quite unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).
+Note that type consistency is reflexive and symmetric but it is not transitive (unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).
 
@@
 Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
-For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$, where $\emptyset$ is the empty hole context.
+For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.
 
@@
 Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
-The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and the outer argument, $c$, can be checked against type $\tehole$ by subsumption.
+The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption.
 However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}.
-Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed ($c$ does not have arrow type).
-The problem is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
+Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed.
+The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
 %On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
@@
 Casts are inserted during the expansion of function applications and ascriptions.
-The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the type assigned to the ascribed expression to the ascribed type.
+The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type.
 Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.
@@
 The conclusion of the rule inserts the corresponding cast.
-We perform type synthesis first and then use analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.
+Note that we go through type synthesis and analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.
 
-The conclusion of Rule~{ESAp} also inserts a cast on the argument's expansion, from the type it assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type, $\tehole$, of the matched arrow type of the function expression, here $\tehole$.
+The conclusion of Rule~{ESAp} also inserts a cast on the argument's expansion, from the type it is assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type, $\tehole$, of the matched arrow type of the function expression, here $\tehole$.
 
@@
 In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$.
-We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), but in practice it is straightforward to do so (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).
+We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), but it is safe to do so (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).
 
-\subsection{Evaluation}
+\subsection{Dynamic Semantics}
 \label{sec:evaluation}
 
-To recap, the result of expansion is a well-typed internal expression with hole closures and casts.
-As described in Sec.~\ref{sec:hole-expansion} and Sec.~\ref{sec:cast-insertion} respectively, empty hole closures correspond to metavariable closures from CMTT, with the metavariable free (i.e.
-tracked in $\Delta$) \cite{Nanevski2008} and casts correspond to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}.
+To recap, the result of expansion is a well-typed internal expression with hole closures and casts, where the former corresponds to metavariable closures from CMTT \cite{Nanevski2008}, and the latter corresponds to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}.
 However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.
 
-The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) correspondence.
-The paper therefore proved only a subject reduction property, which corresponds to type preservation for elimination transitions, generalized to open terms\todo{citation?}{}.
-This is not a full dynamic semantics, and in particular, there is no notion of progress, i.e.
-that well-typed terms cannot get ``stuck'' in an undefined state.
-In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need an operational semantics for terms with free metavariables.
+The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.
+The paper therefore proved only a subject reduction property (which is closely related to type preservation).
+This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e.
+that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}.
+In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables.
 \citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion\footnote{Permutation conversions are necessary to encode the commutative reductions of CMTT, which in turn are necessary to prove a strong normalization property.
-These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.}, which has been studied by \citet{DBLP:journals/iandc/Groote02}, but under this interpretation the same problem arises -- metavariables become variables of a function type, so we cannot rely on any standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
+These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.}, which has been studied by \citet{DBLP:journals/iandc/Groote02}, but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on any standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
 
-Of course, we also need to integrate casts into the dynamic semantics.
-Fortunately, the operational semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery.
-Briefly, the subtlety is that in the cast calculus, the only irreducible terms of hole type are casts, but in $\HazelnutLive$, hole closures induce additional irreducible terms.
-We will return to the technical implications of this fact below.
-The second missing piece in existing work on casts is that, when cast failure occurs, evaluation aborts (i.e.
-blame propagates).
-Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}, is to have cast failure spawn a membrane around the dynamic type error, much like a non-empty hole serves as a membrane around a static type error, allowing evaluation to safely continue past the cast failure when desired.
+Furthermore, we need to integrate casts into the dynamic semantics.
+Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery.
+The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms.
+The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs.
+Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}, is for cast failure to instead insert a membrane around the dynamic type error, much like a non-empty hole serves as a membrane around a static type error, allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.
 
-This section defines a reduction semantics for well-typed internal expressions.
+The dynamic semantics for \HazelnutLive specified in Figures~\ref{fig:isGround}-\ref{fig:contextual-dynamics} addresses the difficulties just outlined, resulting in a system capable of running incomplete programs without aborting when a hole is encountered.
+This semantics is equipped with a meaningful notion of type safety (involving both preservation and progress).
+We also establish that the standard notion of type safety falls out when running complete terms.
+% We also establish that the standard type safety theorem holds when running complete terms.
 
-Matthias Felleisen, Robert Hieb,
-The revised report on the syntactic theories of sequential control and state,
-
-Failed casts next...
+The dynamic semantics is specified as a ``small-step'' transition system.
+The cast-related machinery is based closely on the cast calculus from the ``refined'' account of the gradually typed lambda calculus by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be theoretically well-behaved.
+In particular, Fig.~\ref{fig:isGround} defines the judgement $\isGround{\htau}$, which distinguishes the base type, $b$, and the least specific arrow type, $\tarr{\tehole}{\tehole}$, as \emph{ground types}, which play a role in simplifying the treatment of function casts.
+Fig.~\ref{fig:isFinal} defines the judgement $\isFinal{d}$, which distinguishes the final forms of the transition system.
+There are two classes of final forms: (possibly-)boxed values and indeterminate forms.\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention.
+Our judgemental formulation is more faithful to the mechanization and cleaner for our purposes, because we are distinguishing several classes of final forms.}
 
@@
 \input{fig-dynamics-aux}
-%% \input{fig-dynamics-steps}
+
+The judgement $\isBoxedValue{d}$ distinguishes (possibly-)boxed values, which correspond to the values from the cast calculus and include the classic values from the lambda calculus, distinguished by $\isValue{d}$, as well as two cast forms: casts between function types (except identity casts) and casts from a ground type to the hole type.
+In both cases, the cast must be on a boxed value.
+% (in the prior work, ground types and values were distinguished syntactically, but it is cleaner and more faithful to the mechanization to make these distinctions judgementally).
+
+The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}.
+The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final).
+The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}
+
 \input{fig-dynamics-contexts}
 
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
-{\color{gray}\blindtext}
+The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}.
+Top-level transitions, i.e.
+\emph{steps}, $\stepsToD{}{d}{d'}$, are governed by a single rule, Rule {Step} in Fig.~\ref{fig:step}, which (1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$; (2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in Fig.~\ref{fig:instruction-transitions}; and (3) places $d_0'$ in the selected position, which is marked\footnote{In the literature, the form $\evalhole$ in the grammar of evaluation contexts is referred to as the \emph{hole}, but this hole is a technical device entirely orthogonal to the holes of this paper, so we use the term ``mark'' instead.} in the evaluation context by $\evalhole$, to obtain $d'$.
+This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing.
+Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}.
+It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} compares the two approaches).\todo{Put the search rules in the appendix?}{}
 
-\begin{lemma}[Focus Formation] ~
-  If $\selectEvalCtx{\dexp}{\evalctx}{\dexp'}$
-  then $\isevalctx{\evalctx}$.
-\end{lemma}
+%% \input{fig-dynamics-steps}
 
-\begin{lemma}[Matched Ground Type Invariant] ~
+\subsubsection{Application and Substitution} Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the standard beta reduction transition.
+The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} can be included to specify an eager, left-to-right evaluation strategy, or excluded to leave the evaluation order unspecified.
+We do the latter in our metatheory, both for the sake of generality and for reasons we return to in Sec.~\ref{sec:commutativity}.
+
+Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. The only cases of special interest are when substitution reaches a hole closure:
+\[
+\begin{array}{rcl}
+  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
+  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
+\end{array}
+\]
+In both cases, we write $[d/x]\sigma$ to perform substitution on each expression in the hole environment, $\sigma$.
+For example, $\stepsToD{}{\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}{\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$.
+Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g.
+because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?} Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.
+
+The ITBeta rule is not the only rule we need to handle function application, because lambdas are not the only final forms of arrow type. There are two other classes of possibilities.
+
+The expression in function position might be a cast between arrow types, in which case we apply the arrow cast conversion rule, Rule {ITApCast}, to rewrite the application form, obtaining an equivalent application where the expression under the function cast, $d_1$, is exposed.
+We know from inverting the typing rules that $d_1$ is of type $\tarr{\htau_1}{\htau_2}$ and that $d_2$ is of type $\htau_1'$ where $\tconsistent{\htau_1}{\htau_1'}$, so to maintain type safety, we must place a cast on $d_2$ from  $\htau_1'$ to $\htau_1$.
+The result of this application will be of type $\htau_2$, but the original cast promised that the result would be of consistent type $\htau_2'$, so we also need a cast on the result from $\htau_2$ to $\htau_2'$.
+
+The last possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g.
+$\hap{(\dehole{u}{\sigma}{})}{c}$.
+In this case, no further progress can be made, so the function application as a whole is also deemed indeterminate by Rule {IAp} in Fig.~\ref{fig:isFinal}.
+Note that we are careful to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.\todo{prove this as stated from Ian's disjointness proofs}\todo{do we need a typing premise?}
+\begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
+\end{thm}
+
+\subsubsection{Casts}
+The remaining instruction transition rules handle casts directly.
+Rule {ITCastId} drops identity casts.
+As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term cast to hole type is statically obscure, so we need to wait until it is actually used at some other type, via a cast away from hole type, to be able to determine whether there is a problem.
+Rules {ITCastSucceed} and {ITCastFail} handle this situation when the two types involved are ground types (Fig.~\ref{fig:isGround}).
+If the two ground types are equal, then the cast succeeds and the cast forms can be dropped.
+If they are not equal, then the cast fails and the failed cast form, $\dcastfail{\dexp}{\htau_1}{\htau_2}$, arises.
+Rule {TAFailedCast} specifies as expected from Rule {ITCastFail} that a failed cast is well-typed exactly when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type disequal to $\tau_1$.
+Rule {IFailedCast} specifies that a failed cast operates as an indeterminate form once $d$ is final.
+This allows evaluation to proceed after cast failure, as with holes.
+
+The two rules just described operate at ground type.
+The two remaining instruction transition rules, Rule {ITGround} and {ITExpand}, operate as technical devices that allow us to restrict our interest to ground types exclusively by inserting intermediate casts from (resp.
+to) non-ground types to (resp.
+from) consistent ground types.
+In the core calculus, the only non-ground types are the arrow types, so the grounding judgement, $\groundmatch{\tau_1}{\htau_2}$, defined in Fig.~\ref{fig:groundmatch}, produces the ground arrow type, $\tarr{\tehole}{\tehole}$.
+More generally, this judgement is governed by the following invariant:
+\begin{lem}[Grounding] 
   If $\groundmatch{\htau}{\htau'}$
@@
   and $\htau\neq\htau'$.
-\end{lemma}
+\end{lem}
 
-\begin{theorem}[Preservation]
+\subsubsection{Type Safety} afdasfdas dfasdfs
+\begin{thm}[Preservation]
   If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
@@
   $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$.
-\end{theorem}
+\end{thm}
 
-\begin{theorem}[Progress]
+\begin{thm}[Progress]
   If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
@@
   (iii) $\isBoxedValue{\dexp}$.
-\end{theorem}
+\end{thm}
 
-\begin{theorem}[Canonical Boxed Forms]
+\begin{thm}[Canonical Boxed Forms]
   If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isBoxedValue{\dexp}$
@@
   \end{enumerate}
-\end{theorem}
+\end{thm}
 
-\begin{theorem}[Canonical Indeterminate Forms]
+\begin{thm}[Canonical Indeterminate Forms]
   If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
@@
   \end{enumerate}
-\end{theorem}
+\end{thm}
 
-\subsection{Evaluating Complete Terms}\label{sec:complete-terms}
+\subsubsection{Complete Terms}\label{sec:complete-terms}
 

### 2018-04-04 intro.tex
intro nits; up to and including 'prior approaches'

@@
 program development and code execution''~\cite{burckhardt2013s} with
-fine-grained interleaving of editing and evaluation to varying
-degrees~\cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13}.
+fine-grained interleaving of editing and evaluation~\cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13}.
 
@@
 cannot evaluate \emph{incomplete programs}, \ie{}, programs that are either
-syntactically ill-formed or ill-typed.
+syntactically ill-formed, or ill-typed.
 %
@@
 programs, various editor services may be provided in ad-hoc, restricted
-forms---if available, at all---during the periods of time between complete
+forms---if available at all---during the periods of time between complete
 program states.
@@
 %
-One common approach for emulating expression holes is to use exceptions (such as
-\verb+Debug.crash "not yet implemented"+ or \verb+raise NotYetImplementedException+).
+In lieu of proper language and tool support for expression holes,
+today's programmers commonly emulate them with placeholder exceptions, such as
+\verb+Debug.crash "not yet implemented"+ or
+\verb+raise NotYetImplementedException+.
 %
-Some languages even provide syntax for holes
+Some languages even provide syntax for these kind of ``holes''
 %
@@
 which, semantically, behave similarly to exceptions.
+
+Such exceptions serve to appease the static type checker---because an
+error expression is allowed to have any type required by the surrounding
+context.
 %
-A placeholder exception serves to appease the static type checker---because an
-error expression is allowed to have any type required by the surrounding
-context---allowing the program to run, while preserving dynamic type safety by
+Thus, this mechanism permits the program to run, while preserving dynamic type safety, and
 simply halting execution as soon as the expression appears in evaluation
@@
 
-Although useful, this ad-hoc approach has limitations.
+Although useful, an exception-based mechanism for ``program hole'' has limitations.
 %
-Because the program must halt as soon as a placeholder expression is reached (in
-order to ensure type safety), the programmer cannot benefit from seeing how
-other parts of the program, which do not depend on the missing expression,
+First,
+the program must halt as soon as this expression is reached (ensuring type safety), and consequently, the programmer cannot benefit from seeing how
+\emph{other parts} of the program, which may not depend on the missing expression,
 continue to evaluate.
-%
-Another limitation is that exceptions typically carry nothing more than the
+
+Second, exceptions typically carry nothing more than the
 stack trace that led to the error.
@@
 rather than something nicer like marking them inside non-empty holes?)}
+
+%%%%%%%%%
+%% Matt says: I added the remarks below; My intention is to tie the
+%% last bit into the ``edit and resume'' feature, without making a
+%% huge deal about it.
+
+Finally, exceptions are often thrown with the intent that some calling
+context will possibly ``handle'' them, and will thus resume ordinary
+(non-exception-based) program control flow.
+%
+Meanwhile, it never really makes sense to handle a ``not yet
+implemented'' exception, save for \emph{editing the program to further
+  complete it}, which is beyond the scope and control of ordinary
+exception handlers, and (ordinary) programs' execution behavior.
+%
+Indeed, this is one (of many) ``perceptual gaps'' for live programming
+environments to close.
+%
+In this work, we refer to the programmer-user interaction pattern that
+follows such exceptions as ``edit and resume''.
 

### 2018-04-08 calculus.tex
comments with small suggestions in Section 3

@@
 \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
-
+%
+\rkc{perhaps too much info / unnecessary to say Hazelnut had numbers instead of unit?} 
+%
 Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.
@@
 The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}.
-We will return to evaluation in Sec.~\ref{sec:evaluation}.
+Evaluation of internal expressions will be discussed in Sec.~\ref{sec:evaluation}.
 
@@
 The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
+%
+\rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
+entirely standard, because of the hole context, so perhaps worth mentioning.}
+%
 \begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
@@
 The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.
+%
+\rkc{this list (holes, func app, and ascription) has three items, but there are
+only two subsections. see comment at beginning of Sec 3.2.2.}
 
 \subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
+%
+\rkc{perhaps add ``empty and non-empty'' before hole closures.}
 
@@
 The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs.
-Initially, no evaluation has occured, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
+Initially, no evaluation has occurred, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
 \begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
@@
 \subsubsection{Cast Insertion}\label{sec:cast-insertion} 
+%
+\rkc{consider renaming subsection to ``Function Application and Cast
+Intersion''. and adding a transition sentence like ``The remaining development
+of expansion deals with the delicate interplay between function application and
+cast insertion.''}
+%
 Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
@@
 This leads us to the solution to this problem: cast insertion.
+%
+\rkc{Instead of the last two sentences, maybe something like: ``By analogy
+between hole types in our system and unknown types from gradual type theory, we
+address the problem with the same mechanism used in gradual type theory: cast
+insertion.''}
 
@@
 Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.
+%
+\rkc{move this Note sentence to the same place where half-annotated lambdas were
+noted. This note breaks the flow here.}
 
@@
 
-The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}.
+The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}).
 The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final).
@@
 Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g.
-because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?} Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.
+because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?}.
+Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.
 

### 2018-04-09 calculus.tex
writing suggestions for Cyrus, and Section 3

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Matt's high-level Writing comments for Cyrus (for Section 3 mostly):
+
+% Please, please please: Use line breaks.  Git does't work well without them.
+% Change your editor or editor settings please.
+
+% Cyrus, your writing is generally very good, but is often harder for
+% me to digest than it should be, as a reader.  I think it'd be easier
+% to read on a first pass if we followed my preferences below, which I
+% humbly offer as gentle suggestions:
+
+% I think that some these suggestions are especially important for reading
+% really wide columns of text, as in this new PACMPL format. :)
+
+% My nits boil down to these preferences:
+
+% 0. NO PASSIVE VOICE --- EVER.  Use ``We'' in lots of places below instead of passive voice.
+
+% 1. Shorter sentences.
+
+% 2. Fewer parenthenticals: just use more sentences instead.
+
+% 3. Shorter paragraphs, with more defined 'paragraph structure'. (E.g., topic sentence)
+
+% 4. Shorter sentence subjects (with subject closer to verb).
+
+% 5. Use macros for fig and section references.
+
+% 6. If a sentence introduces a figure, that sentence should be the
+% first sentence of a paragraph, and the figure reference should be
+% the first word of that sentence, ideally.  As a reader, then I look
+% at figures first, and find the paragraphs for them easily by
+% scanning the text.
+
+% 7. For each greek letter, prefix with an english word 'classifier',
+% e.g., 'external expression~$e$', not just '$e$'.  Use way, way fewer
+% commas please.
+
+% 8. Use singlar over plural sentence subjects, whenever possible;
+% e.g., an ``empty hole is written..'' not ``empty holes are
+% written''; or ``an empty hole synthesizes'' not ``empty holes
+% synthesize''
+
+% 9. Fewer commas. Each 'extra' comma is a burden to the reader (it's
+% asking the reader to allocate a stack frame in their brain!)
+
+% *** I'm transforming the next paragraph below as a demonstration,
+% creating smaller, more cohesive paragraphs with shorter sentences.
+% See what you think.
+%
+% I'm not doing all of Section 3, tho, since
+%
+% (1) you may hate what I've done, and
+%
+% (2) I don't want to do a lot of work that will just be
+% reverted/resented :)
+
+% BTW: My version is _shorter_, not longer than the original :)
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \input{fig-bidirectional-typing}
+
+We give an overview of the static semantics of the \HazelnutLive
+external language.
+
+\Figref{fig:bidirectional-typing} defines \emph{bidirectional} typing,
+where each type~$\htau$ classifies an external expression~$e$.
+%
+This bidirectional style uses two mutually defined judgements: The type synthesis
+judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type $\htau$
+for $\hexp$ under typing context $\hGamma$, which tracks typing
+assumptions of the form $x : \htau$ in the usual
+manner \cite{pfpl,tapl}.
+%
+The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
+expression~$\hexp$ against a given type~$\htau$. Algorithmically, this
+type is an output of type synthesis but an input of type analysis. At
+the top level, we start with synthesis.
+
+% Algorithmically, the type is an output of type synthesis but an input of type analysis.
+
+The benefit of specifying the \HazelnutLive external language
+bidirectionally is that the programmer does not need to annotate each
+hole with a type
+%
+\cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}.
+%
+These typing rules closely follow the \Hazelnut type
+system \cite{popl-paper}; we discuss minor differences below.
+
+We supplement each expression hole from \Hazelnut with such a
+hole \emph{name}, which are new to \HazelnutLive; an empty hole is
+written $\hehole{u}$, where $u$ is a hole name assumed unique.
+%
+Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
+%
+If an empty hole appears where an expression of some other type is
+expected, e.g. under an explicit ascription (governed by Rule {SAsc})
+or in the argument position of a function application (governed by
+Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
+Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
+type $\htau$, then it can be checked against any \emph{consistent}
+type, $\htau'$.
+
 \input{fig-type-consistency}
 
-Let us start with an overview of the static semantics of the \HazelnutLive external language.
-Types, $\htau$, classify external expressions, $e$, according to the type system in Fig.
-\ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up below).
-The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
-The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ under typing context $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}.
-The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$.
-Algorithmically, the type is an output of type synthesis but an input of type analysis.
-At the top level, we start with synthesis.
-% Algorithmically, the type is an output of type synthesis but an input of type analysis.
+%%%%%%%%%%%%%%%%%%%% MATT stopped editing Cyrus's text here.
 
-The benefit of specifying the \HazelnutLive external language bidirectionally is that the programmer does not need to annotate each hole with a type.
-Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole notation is taken from \Hazelnut; hole names are new to \HazelnutLive).
-Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$.
-If an empty hole appears where an expression of some other type is expected, e.g.
-under an explicit ascription (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$.
 Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.

### 2018-04-09 calculus.tex
minor: fix awkward wording

@@
 hole \emph{name}, which are new to \HazelnutLive; an empty hole is
-written $\hehole{u}$, where $u$ is a hole name assumed unique.
+written $\hehole{u}$, where $u$ is a unique name for the hole.
 %

### 2018-04-09 calculus.tex
fix semantically-wrong statement that I introduced

@@
 % BTW: My version is _shorter_, not longer than the original :)
+%
+% Also, In my humble opinion, it is way, way easier to read. ;)
 
@@
 
-\Figref{fig:bidirectional-typing} defines \emph{bidirectional} typing,
-where each type~$\htau$ classifies an external expression~$e$.
+\Figref{fig:bidirectional-typing} defines \emph{bidirectional} typing
 %
-This bidirectional style uses two mutually defined judgements: The type synthesis
+as two mutually defined judgements: The type synthesis
 judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type $\htau$

### 2018-04-09 calculus.tex
minor

@@
 as two mutually defined judgements: The type synthesis
-judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type $\htau$
-for $\hexp$ under typing context $\hGamma$, which tracks typing
+judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
+for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
 assumptions of the form $x : \htau$ in the usual

### 2018-04-09 calculus.tex
remove 'extra' statement
minor edits to my edits

@@
 The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
-expression~$\hexp$ against a given type~$\htau$. Algorithmically, this
-type is an output of type synthesis but an input of type analysis. At
-the top level, we start with synthesis.
+expression~$\hexp$ against a given type~$\htau$.
+%
+Algorithmically, analysis accepts a type as input, and synthesis gives
+a type as output.
+%
+We use synthesis for each of the programmer's ``top level'' external
+exressions~$e$, giving the corresponding type as feedback.
 

### 2018-04-09 calculus.tex
write down my cite guideline
typo: remove extra word

@@
 
+% 10. Citations in this Names+Year form break sentence and paragraph
+% flow for the reader, sometimes quite severly.  Put long citation
+% chains at the ends of sentences (or better yet, in their own little
+% paragraph, as I did below).
+
 % *** I'm transforming the next paragraph below as a demonstration,
@@
 
-We supplement each expression hole from \Hazelnut with such a
+We supplement each expression hole from \Hazelnut with a
 hole \emph{name}, which are new to \HazelnutLive; an empty hole is

### 2018-04-10 appendix-defns.tex
finished draft of 3.2
finished draft of sec3 and 4

@@
 
+\subsection{Substitution}
+\label{sec:substitution}
+\judgbox
+  {[\dexp/x]\dexp' = \dexp''}
+  {$\dexp''$ is obtained by substituting $\dexp$ for $x$ in $\dexp'$}
+  %% {$\dexp''$ is the result of substituting $\dexp$ for $u$ in $\dexp'$}
+
+\vspace{5pt}
+\judgbox
+  {[\dexp/x]\sigma = \sigma'}
+  {$\sigma'$ is obtained by substituting $\dexp$ for $x$ in $\sigma$}
+  %% {$\dexp''$ is the result of substituting $\dexp$ for $u$ in $\dexp'$}
+\[
+\begin{array}{lcll}
+\substitute{\dexp}{x}{c}
+&=&
+c\\
+\substitute{\dexp}{x}{x}
+&=&
+\dexp
+\\%
+\substitute{\dexp}{x}{y}
+&=&
+y
+& \text{when $x \neq y$}
+\\
+\substitute{\dexp}{x}{\halam{y}{\htau}{\dexp'}}
+&=&
+\halam{y}{\htau}{\substitute{\dexp}{x}{\dexp'}}
+& \text{when $x \neq y$ and $y \notin \fvof{d}$}
+\\
+\substitute{\dexp}{x}{\dap{\dexp_1}{\dexp_2}}
+&=&
+\dap{(\substitute{\dexp}{x}{\dexp_1})}{\substitute{\dexp}{x}{\dexp_2}}
+\\
+\substitute{\dexp}{x}{\dehole{u}{\subst}{}}
+&=&
+\dehole{u}{\substitute{\dexp}{x}{\subst}}{}
+\\
+\substitute{\dexp}{x}{\dhole{\dexp'}{u}{\subst}{}}
+&=&
+\dhole{\substitute{\dexp}{x}{\dexp'}}{u}{\substitute{\dexp}{x}{\subst}}{}
+\\
+\substitute{\dexp}{x}{\dcasttwo{\dexp'}{\htau_1}{\htau_2}}
+&=&
+\dcasttwo{(\substitute{\dexp}{x}{\dexp'})}{\htau_1}{\htau_2}
+\\
+\substitute{\dexp}{x}{\dcastfail{\dexp'}{\htau_1}{\htau_2}}
+&=&
+\dcastfail{(\substitute{\dexp}{x}{\dexp'})}{\htau_1}{\htau_2}
+\\
+\substitute{\dexp}{x}{\cdot}
+&=&
+\cdot\\
+\substitute{\dexp}{x}{\sigma, d/y}
+&=&
+\substitute{\dexp}{x}{\sigma}, \substitute{\dexp}{x}{d}/y
+%% {[\dexp_1 / x] \dcast{\htau}{\dexp}}
+%% &=&
+%% \dcast{\htau}{[\dexp_1 / x] \dexp}
+\end{array}
+\]
+
+\vspace{5pt}
+\begin{lem}[Substitution] \label{thm:substitution}~
+\begin{enumerate}[nolistsep]
+\item If $\hasType{\hDelta}{\hGamma, x : \htau'}{d}{\htau}$ and $\hasType{\hDelta}{\hGamma}{d'}{\htau'}$ then $\hasType{\hDelta}{\hGamma}{[d'/x]d}{\htau}$.
+\item If $\hasType{\hDelta}{\hGamma, x : \htau'}{\sigma}{\hGamma'}$ and $\hasType{\hDelta}{\hGamma}{d'}{\htau'}$ then $\hasType{\hDelta}{\hGamma}{[d'/x]\sigma}{\hGamma'}$.
+\end{enumerate}
+\end{lem}
+
+\subsection{Canonical Forms} 
+\begin{lem}[Canonical Value Forms]\label{thm:canonincal-value-forms}
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isValue{\dexp}$
+  then $\htau\neq\tehole$ and
+  \begin{enumerate}[nolistsep]
+    \item If $\htau=b$ then $\dexp=c$.
+    \item If $\htau=\tarr{\htau_1}{\htau_2}$
+          then $\dexp=\halam{x}{\htau_1}{\dexp'}$
+          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$.
+  \end{enumerate}
+\end{lem}
+
+\begin{lem}[Canonical Boxed Forms]\label{thm:canonical-boxed-forms}
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isBoxedValue{\dexp}$
+  then
+  \begin{enumerate}[nolistsep]
+    \item If $\htau=b$ then $\dexp=c$.
+    \item If $\htau=\tarr{\htau_1}{\htau_2}$ then either
+      \begin{enumerate}
+        \item[i.]
+          $\dexp=\halam{x}{\htau_1}{\dexp'}$
+          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$, or
+        \item[ii.]
+          $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1'}{\htau_2'}}{\tarr{\htau_1}{\htau_2}}$
+          where $\tarr{\htau_1'}{\htau_2'}\neq\tarr{\htau_1}{\htau_2}$
+          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\tarr{\htau_1'}{\htau_2'}}$.
+      \end{enumerate}
+    \item If $\htau=\tehole$
+          then $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
+          where $\isGround{\htau'}$
+          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$.
+  \end{enumerate}
+\end{lem}
+
+\begin{lem}[Canonical Indeterminate Forms]
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
+  and $\isIndet{\dexp}$
+  then either\todo{expand this to have one case per type}
+  \begin{enumerate}[nolistsep]
+    \item
+      $\dexp=\dehole{u}{\subst}{}$
+      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
+    \item
+      $\dexp=\dhole{\dexp'}{u}{\subst}{}$
+      and $\isFinal{\dexp'}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$
+      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
+    \item
+      $\dexp=\dap{\dexp_1}{\dexp_2}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2}{\htau}}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2}$
+      and $\isIndet{\dexp_1}$
+      and $\isFinal{\dexp_2}$
+      and $\dexp_1\neq\dcasttwo{\dexp_1}{\tarr{\htau_3}{\htau_4}}
+                                        {\tarr{\htau_3'}{\htau_4'}}$, or
+    %% \item
+    %%   \begin{enumerate}
+    %%     \item blah
+    %%     \item blah
+    %%     \item blah
+    %%   \end{enumerate}
+    \item
+      $\htau=b$
+      and $\dexp=\dcasttwo{\dexp'}{\tehole}{b}$
+      and $\isIndet{\dexp'}$
+      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
+    \item
+      $\htau=b$
+      and $\dexp=\dcastfail{\dexp'}{\htau'}{b}$
+      and $\isGround{\htau'}$
+      and $\htau'\neq{b}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
+    \item
+      $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+      and $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1}{\htau_2}}
+                                  {\tarr{\htau_{11}}{\htau_{12}}}$
+      and $\isIndet{\dexp'}$
+      and $\tarr{\htau_1}{\htau_2}\neq\tarr{\htau_{11}}{\htau_{12}}$, or
+    \item
+      $\htau=\tarr{\tehole}{\tehole}$
+      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+      %% and $\htau_{11}=\tehole$
+      %% and $\htau_{12}=\tehole$
+      and $\dexp=\dcastthree{\dexp'}{\tehole}{\tehole}{\tehole}$
+      and $\isIndet{\dexp'}$
+      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
+    \item
+      $\htau=\tarr{\tehole}{\tehole}$
+      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+      %% and $\htau_{11}=\tehole$
+      %% and $\htau_{12}=\tehole$
+      and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\tehole}{\tehole}}$
+      %% and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\htau_{11}}{\htau_{12}}}$
+      and $\htau'\neq\htau$
+      and $\isGround{\htau'}$
+      and $\isIndet{\dexp'}$
+      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
+    \item
+      $\htau=\tehole$
+      and $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
+      and $\isGround{\htau'}$
+      and $\isIndet{\dexp'}$.
+  \end{enumerate}
+\end{lem}
+
+% No weakening for Gammas in Delta: 
+% If $\hasType{\Delta, \Dbinding{u}{\hGamma}{\tau}}{\hGamma'}{d}{\tau'}$ then $\hasType{\Delta, \Dbinding{u}{\hGamma, x : \tau''}{\tau}}{\hGamma'}{d}{\tau'}$.
+
+\clearpage
+\subsection{Complete Programs}
+
 \input{fig-complete}
+
+\clearpage
+\subsection{Hole Filling}\label{sec:hole-filling}
+\input{fig-multi-step}
+\input{fig-evalctx-instantiation}
+\begin{lem}[Filling] ~
+  \begin{enumerate}[nolistsep]
+  \item If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\dexp}{\tau}$
+  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
+  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
+  \item If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\sigma}{\hGamma''}$
+  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
+  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\sigma}}{\hGamma''}$.
+  \end{enumerate}
+\end{lem}
+\begin{proof}
+In each case, we proceed by rule induction on the first assumption, appealing to the Substitution Lemma as necessary.
+\end{proof}
+
+\begin{lem}[Substitution Commutativity]
+  If 
+  \begin{enumerate}[nolistsep]
+  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{x : \htau_2}{\dexp_1}{\tau}$ and 
+  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_2}{\htau_2}$ and 
+  	\item $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
+  \end{enumerate}
+
+  then  $\instantiate{d'}{u}{\substitute{d_2}{x}{d_1}} = \substitute{\instantiate{d'}{u}{d_2}}{x}{\instantiate{d'}{u}{d_1}}$.
+\end{lem}
+
+\begin{lem}[Instruction Commutativity]
+  If 
+  \begin{enumerate}[nolistsep]
+  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$ and 
+  	\item $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and 
+  	\item $\reducesE{}{\dexp_1}{\dexp_2}$
+  \end{enumerate}
+
+  then $\reducesE{}{\instantiate{\dexp'}{u}{\dexp_1}}
+                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
+\end{lem}
+
+\begin{lem}[Filling Totality]
+Either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx}=\evalctx'$ for some $\evalctx'$.
+\end{lem}
+
+\begin{lem}[Discarding] If 
+	\begin{enumerate}[nolistsep]
+	\item $\selectEvalCtx{d_1}{\evalctx}{\dexp_1'}$ and 
+	\item $\selectEvalCtx{d_2}{\evalctx}{\dexp_2'}$ and 
+	\item $\inhole{u}{\evalctx}$
+	\end{enumerate}
+
+	then $\instantiate{d}{u}{d_1} = \instantiate{d}{u}{d_2}$.
+\end{lem}
+
+\begin{lem}[Filling Distribution] If 
+	$\selectEvalCtx{d_1}{\evalctx}{d_1'}$ and $\instantiate{d}{u}{\evalctx}=\evalctx'$ then $\selectEvalCtx{\instantiate{d}{u}{d_1}}{\evalctx'}{\instantiate{d}{u}{d_1'}}$.
+\end{lem}
+
+\begin{thm}[Commutativity]
+  If 
+  \begin{enumerate}[nolistsep]
+  \item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$ and 
+  \item $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and 
+  \item $\multiStepsTo{\dexp_1}{\dexp_2}$
+\end{enumerate}
+
+  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
+                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
+\end{thm}
+\begin{proof}
+By rule induction on (3).
+The reflexive case is immediate.
+In the inductive case, we proceed by rule induction on the stepping premise.
+There is one rule, Rule~{Step}.
+By Filling Totality, either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx} = \evalctx'$.
+In the former case, by Discarding, we can conclude by MultiStepRefl.
+In the latter case, by Instruction Commutativity and Filling Distribution we can take a step, and we can conclude via MultiStepSteps by applying Preservation and then the IH.
+\end{proof}

### 2018-04-10 calculus.tex
note about passive voice; 100 occurences of 'is' in this section\!
finished draft of sec3 and 4

--- 2018-04-09 calculus.tex
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Matt's high-level Writing comments for Cyrus (for Section 3 mostly):
+\input{fig-bidirectional-typing}
+\input{fig-type-consistency}
 
-% Please, please please: Use line breaks.  Git does't work well without them.
-% Change your editor or editor settings please.
-
-% Cyrus, your writing is generally very good, but is often harder for
-% me to digest than it should be, as a reader.  I think it'd be easier
-% to read on a first pass if we followed my preferences below, which I
-% humbly offer as gentle suggestions:
-
-% I think that some these suggestions are especially important for reading
-% really wide columns of text, as in this new PACMPL format. :)
-
-% My nits boil down to these preferences:
-
-% 0. NO PASSIVE VOICE --- EVER.  Use ``We'' in lots of places below instead of passive voice.
-
-% 1. Shorter sentences.
-
-% 2. Fewer parenthenticals: just use more sentences instead.
-
-% 3. Shorter paragraphs, with more defined 'paragraph structure'. (E.g., topic sentence)
-
-% 4. Shorter sentence subjects (with subject closer to verb).
-
-% 5. Use macros for fig and section references.
-
-% 6. If a sentence introduces a figure, that sentence should be the
-% first sentence of a paragraph, and the figure reference should be
-% the first word of that sentence, ideally.  As a reader, then I look
-% at figures first, and find the paragraphs for them easily by
-% scanning the text.
-
-% 7. For each greek letter, prefix with an english word 'classifier',
-% e.g., 'external expression~$e$', not just '$e$'.  Use way, way fewer
-% commas please.
-
-% 8. Use singlar over plural sentence subjects, whenever possible;
-% e.g., an ``empty hole is written..'' not ``empty holes are
-% written''; or ``an empty hole synthesizes'' not ``empty holes
-% synthesize''
-
-% 9. Fewer commas. Each 'extra' comma is a burden to the reader (it's
-% asking the reader to allocate a stack frame in their brain!)
-
-% *** I'm transforming the next paragraph below as a demonstration,
-% creating smaller, more cohesive paragraphs with shorter sentences.
-% See what you think.
-%
-% I'm not doing all of Section 3, tho, since
-%
-% (1) you may hate what I've done, and
-%
-% (2) I don't want to do a lot of work that will just be
-% reverted/resented :)
-
-% BTW: My version is _shorter_, not longer than the original :)
-%
-% Also, In my humble opinion, it is way, way easier to read. ;)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\input{fig-bidirectional-typing}
-
-We give an overview of the static semantics of the \HazelnutLive
-external language.
-
-\Figref{fig:bidirectional-typing} defines \emph{bidirectional} typing
-%
-as two mutually defined judgements: The type synthesis
-judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
-for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
-assumptions of the form $x : \htau$ in the usual
-manner \cite{pfpl,tapl}.
-%
-The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
-expression~$\hexp$ against a given type~$\htau$. Algorithmically, this
-type is an output of type synthesis but an input of type analysis. At
-the top level, we start with synthesis.
-
+Let us start with an overview of the static semantics of the \HazelnutLive external language.
+Types, $\htau$, classify external expressions, $e$, according to the type system in Fig.
+\ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up below).
+The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
+The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ under typing context $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}.
+The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$.
+Algorithmically, the type is an output of type synthesis but an input of type analysis.
+At the top level, we start with synthesis.
 % Algorithmically, the type is an output of type synthesis but an input of type analysis.
 
-The benefit of specifying the \HazelnutLive external language
-bidirectionally is that the programmer does not need to annotate each
-hole with a type
-%
-\cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}.
-%
-These typing rules closely follow the \Hazelnut type
-system \cite{popl-paper}; we discuss minor differences below.
-
-We supplement each expression hole from \Hazelnut with such a
-hole \emph{name}, which are new to \HazelnutLive; an empty hole is
-written $\hehole{u}$, where $u$ is a unique name for the hole.
-%
-Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
-%
-If an empty hole appears where an expression of some other type is
-expected, e.g. under an explicit ascription (governed by Rule {SAsc})
-or in the argument position of a function application (governed by
-Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
-Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
-type $\htau$, then it can be checked against any \emph{consistent}
-type, $\htau'$.
-
-\input{fig-type-consistency}
-
-%%%%%%%%%%%%%%%%%%%% MATT stopped editing Cyrus's text here.
-
+The benefit of specifying the \HazelnutLive external language bidirectionally is that the programmer does not need to annotate each hole with a type.
+Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole notation is taken from \Hazelnut; hole names are new to \HazelnutLive).
+Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$.
+If an empty hole appears where an expression of some other type is expected, e.g.
+under an explicit ascription (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$.
 Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.
@@
 \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
-%
-\rkc{perhaps too much info / unnecessary to say Hazelnut had numbers instead of unit?} 
-%
+
 Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.
@@
 The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}.
-Evaluation of internal expressions will be discussed in Sec.~\ref{sec:evaluation}.
+We will return to evaluation in Sec.~\ref{sec:evaluation}.
 
@@
 The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
-%
-\rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
-entirely standard, because of the hole context, so perhaps worth mentioning.}
-%
 \begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
@@
 The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.
-%
-\rkc{this list (holes, func app, and ascription) has three items, but there are
-only two subsections. see comment at beginning of Sec 3.2.2.}
 
 \subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
-%
-\rkc{perhaps add ``empty and non-empty'' before hole closures.}
 
@@
 The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs.
-Initially, no evaluation has occurred, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
+Initially, no evaluation has occured, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
 \begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
@@
 \subsubsection{Cast Insertion}\label{sec:cast-insertion} 
-%
-\rkc{consider renaming subsection to ``Function Application and Cast
-Intersion''. and adding a transition sentence like ``The remaining development
-of expansion deals with the delicate interplay between function application and
-cast insertion.''}
-%
 Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
@@
 This leads us to the solution to this problem: cast insertion.
-%
-\rkc{Instead of the last two sentences, maybe something like: ``By analogy
-between hole types in our system and unknown types from gradual type theory, we
-address the problem with the same mechanism used in gradual type theory: cast
-insertion.''}
 
@@
 Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.
-%
-\rkc{move this Note sentence to the same place where half-annotated lambdas were
-noted. This note breaks the flow here.}
 
@@
 
-The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}).
+The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}.
 The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final).
@@
 Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g.
-because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?}.
-Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.
+because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?} Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.
 
@@
 
-The last possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g.
+The other possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g.
 $\hap{(\dehole{u}{\sigma}{})}{c}$.
@@
 
-\subsubsection{Type Safety} afdasfdas dfasdfs
+In all other cases, casts are either boxed values or indeterminate according to the remaining rules in Fig.~\ref{fig:isFinal}.
+Note in particular Rule {ICastHoleGround}, which handles casts from hole to ground type that are not of the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$.
+In the cast calculus, there are no such irreducible terms because the only canonical form of type $\tehole$ is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.
+
+\subsubsection{Type Safety} The purpose of establishing type safety is to ensure that the static and dynamic semantics of a language cohere.
+We follow the approach developed by \citet{wright94:_type_soundness}, now standard \cite{pfpl}, which distinguishes two type safety properties, preservation and progress.
+In order to allow for the evaluation of incomplete programs, we establish these properties for terms typed under arbitrary hole context $\Delta$.
+We can assume $\hGamma$ is empty (in order to run programs with free variables, the system can treat them as empty holes with a corresponding name).
+
+The preservation theorem establishes that type assignment is preserved by transitions, i.e. that the type of an expression accurately predicts the type of the result of evaluating that expression.
+
 \begin{thm}[Preservation]
@@
 \end{thm}
+\noindent
+The proof relies on an analagous preservation lemma for instruction transitions and a standard substitution lemma. Hole closures can disappear during evaluation, so we rely on weakening of $\Delta$.
 
+The progress theorem establishes that the dynamic semantics accounts for every well-typed term, i.e. that we have not forgotten some necessary rules or premises.
 \begin{thm}[Progress]
   If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
-  (i) $\stepsToD{}{\dexp}{\dexp'}$ or
-  (ii) $\isIndet{\dexp}$ or
-  (iii) $\isBoxedValue{\dexp}$.
+  (a) $\stepsToD{}{\dexp}{\dexp'}$ or
+  (b) $\isBoxedValue{\dexp}$ or 
+  (c) $\isIndet{\dexp}$.
 \end{thm}
+\noindent
+Explicitly accounting for indeterminate forms, which are rooted at either a hole closure or a failed cast, is the key to being able to establish the progress theorem under a non-empty hole context.
+The proof relies on canonical forms lemmas stated in Appendix~\ref{sec:additional-defns}.
 
-\begin{thm}[Canonical Boxed Forms]
-  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isBoxedValue{\dexp}$
-  then
-  \begin{enumerate}[label=(\alph*)]
-    \item If $\htau=b$ then $\dexp=c$.
-    \item If $\htau=\tarr{\htau_1}{\htau_2}$ then either
-      \begin{enumerate}
-        \item[i.]
-          $\dexp=\halam{x}{\htau_1}{\dexp'}$
-          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$, or
-        \item[ii.]
-          $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1'}{\htau_2'}}{\tarr{\htau_1}{\htau_2}}$
-          where $\tarr{\htau_1'}{\htau_2'}\neq\tarr{\htau_1}{\htau_2}$
-          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\tarr{\htau_1'}{\htau_2'}}$.
-      \end{enumerate}
-    \item If $\htau=\tehole$
-          then $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
-          where $\isGround{\htau'}$
-          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$.
-  \end{enumerate}
-\end{thm}
-
-\begin{thm}[Canonical Indeterminate Forms]
-  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
-  and $\isIndet{\dexp}$
-  then either
-  \begin{enumerate}[label=(\alph*)]
-    \item
-      $\dexp=\dehole{u}{\subst}{}$
-      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
-    \item
-      $\dexp=\dhole{\dexp'}{u}{\subst}{}$
-      and $\isFinal{\dexp'}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$
-      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
-    \item
-      $\dexp=\dap{\dexp_1}{\dexp_2}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2}{\htau}}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2}$
-      and $\isIndet{\dexp_1}$
-      and $\isFinal{\dexp_2}$
-      and $\dexp_1\neq\dcasttwo{\dexp_1}{\tarr{\htau_3}{\htau_4}}
-                                        {\tarr{\htau_3'}{\htau_4'}}$, or
-    %% \item
-    %%   \begin{enumerate}
-    %%     \item blah
-    %%     \item blah
-    %%     \item blah
-    %%   \end{enumerate}
-    \item
-      $\htau=b$
-      and $\dexp=\dcasttwo{\dexp'}{\tehole}{b}$
-      and $\isIndet{\dexp'}$
-      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
-    \item
-      $\htau=b$
-      and $\dexp=\dcastfail{\dexp'}{\htau'}{b}$
-      and $\isGround{\htau'}$
-      and $\htau'\neq{b}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
-    \item
-      $\htau=\tarr{\htau_{11}}{\htau_{12}}$
-      and $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1}{\htau_2}}
-                                  {\tarr{\htau_{11}}{\htau_{12}}}$
-      and $\isIndet{\dexp'}$
-      and $\tarr{\htau_1}{\htau_2}\neq\tarr{\htau_{11}}{\htau_{12}}$, or
-    \item
-      $\htau=\tarr{\tehole}{\tehole}$
-      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
-      %% and $\htau_{11}=\tehole$
-      %% and $\htau_{12}=\tehole$
-      and $\dexp=\dcastthree{\dexp'}{\tehole}{\tehole}{\tehole}$
-      and $\isIndet{\dexp'}$
-      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
-    \item
-      $\htau=\tarr{\tehole}{\tehole}$
-      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
-      %% and $\htau_{11}=\tehole$
-      %% and $\htau_{12}=\tehole$
-      and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\tehole}{\tehole}}$
-      %% and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\htau_{11}}{\htau_{12}}}$
-      and $\htau'\neq\htau$
-      and $\isGround{\htau'}$
-      and $\isIndet{\dexp'}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
-    \item
-      $\htau=\tehole$
-      and $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
-      and $\isGround{\htau'}$
-      and $\isIndet{\dexp'}$.
-  \end{enumerate}
-\end{thm}
-
-\subsubsection{Complete Terms}\label{sec:complete-terms}
-
-\autoref{fig:complete} defines \emph{complete} types, expressions, and
-dynamic expressions to be those terms without type or expression holes.
-\begin{theorem}[Canonical Value Forms]
-  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isValue{\dexp}$
-  then $\htau\neq\tehole$ and
-  \begin{enumerate}[label=(\alph*)]
-    \item If $\htau=b$ then $\dexp=c$.
-    \item If $\htau=\tarr{\htau_1}{\htau_2}$
-          then $\dexp=\halam{x}{\htau_1}{\dexp'}$
-          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$.
-  \end{enumerate}
-\end{theorem}
-
-\begin{theorem}[Complete Progress]
-  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
-  and $\isComplete{\dexp}$
-  then either $\stepsToD{}{\dexp}{\dexp'}$
-  or $\isValue{\dexp}$.
-\end{theorem}
-
-\begin{theorem}[Complete Preservation]
-  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
-  and $\isComplete{\dexp}$
-  and $\stepsToD{}{\dexp}{\dexp'}$
-  then $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau}$
-  and $\isComplete{\dexp'}$.
-\end{theorem}
-
-\begin{theorem}[Complete Expansion] ~
-  \begin{enumerate}
+\subsubsection{Complete Programs} Although our focus in this paper is on running incomplete programs, this machinery does not interfere with reasoning about the behavior of complete programs, i.e.
+those with no holes.
+Appendix~\ref{sec:additional-defns} defines complete programs with the judgements $\isComplete{e}$ and $\isComplete{d}$.
+Failed casts cannot appear in complete internal expressions.
+The following theorem establishes that expansion preserves program completeness.
+\begin{thm}[Complete Expansion] ~
+  \begin{enumerate}[nolistsep]
     \item
@@
   \end{enumerate}
-\end{theorem}
+\end{thm}
 
-\cy{definition of complete types and terms? do we care about completeness
-  of $\hexp$s or $\dexp$s?}
+The following preservation theorem establishes that stepping preserves program completeness.\todo{can we do complete preservation without the typing assumption?}
+\begin{thm}[Complete Preservation]
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
+  and $\isComplete{\dexp}$
+  and $\stepsToD{}{\dexp}{\dexp'}$
+  then $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau}$
+  and $\isComplete{\dexp'}$.
+\end{thm}
+
+The following progress theorem establishes that only classic values result from evaluating a complete program. There are no boxed values or indeterminate forms. 
+\begin{thm}[Complete Progress]
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
+  and $\isComplete{\dexp}$
+  then either $\stepsToD{}{\dexp}{\dexp'}$
+  or $\isValue{\dexp}$.
+\end{thm}
 
@@
 
+The supplemental material includes our mechanization of the semantics and metatheory of \HazelnutLive, 
+including proofs of all of the theorems stated above and any necessary lemmas\todo{make this so ian!}{}. 
+We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis},  in part because it was also used in the mechanization of \Hazelnut \cite{popl-paper}. 
+No proof automation is used because our intention is to communicate the proof
+structure itself. The basic approach is standard: we model judgements as 
+dependent inductive datatypes. 
+We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
+contexts, hole contexts and hole environments using metafunctions.\todo{say something about axioms?}{}
+
+\subsection{Implementation}\label{sec:implementation}
+
+The supplemental material also includes a browser-based implementation of \HazelnutLive.
+The implementation is a functional reactive program \cite{DBLP:conf/pldi/CzaplickiC13} written using the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} together with the OCaml \lismall{React} library \cite{OcamlReact} and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}.
+A screenshot of the user interface, which implements all of the live programming features described in Sec.~\ref{sec:examples}, but for the more austere language of this section, is shown in Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{}.
+
+The editor component of the implementation is derived from \Hazelnut. It exposes a language of edit actions that insert holes  
+automatically to guarantee that every edit state has some (possibly incomplete) type. Formally,
+this corresponds to the Sensibility theorem from the prior work \cite{popl-paper}. 
+By composing this theorem with the Expandability, Typed Expansion and Progress theorems from this paper, we establish a uniquely powerful 
+invariant: that every edit state is both statically and dynamically meaningful.
+In other words, live feedback does not ``flicker in and out'' or reflect past edit states.
+The primary purpose of the included implementation is to serve as a proof-of-concept of this invariant, and to be of use to researchers studying the calculus as presented.% Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.
+
+We are also integrating the mechanisms developed in this paper into the \Hazel programming environment, which is being developed by \citet{HazelnutSNAPL} with the goal of providing a full-scale hole-driven development experience rooted in the type theoretic foundations established by \Hazelnut.
+Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper.
+We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as the \Hazel language evolves.
+The included implementation provides a link to this evolving effort.
+For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed certain syntactic conveniences that are not yet implemented in \Hazel, but that are available in Elm and standard in ML-like languages.
+
+Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}
+
 %%   \halam{x}{\htau}{\evalctx} ~\vert~
-
-{\color{gray}\blindtext}
 
@@
 
-\section{Metatheory: Correctness of Edit-and-Resume}
+\newcommand{\commutativitySec}{A Contextual Modal Interpretation of Fill-and-Resume}
+\section{\protect\commutativitySec}
 \label{sec:resumption}
 
+The result of evaluation is a final internal expression with hole closures, each with an associated hole environment, $\sigma$.
+These hole environments can be reported directly to the programmer, e.g.
+via the sidebar shown in Fig.~X\todo{fig}, to help them as they think about how to fill in the corresponding hole in the external expression.
+Hole environments might also be useful indirectly, e.g.
+by informing an edit action synthesis and suggestion system.
+In any case, once the programmer does perform one or more edit actions to fill in a hole, a new result must be computed.
+Na\"ively, the system would need to compute the result ``from scratch'' on each such edit.
+A more efficient approach supported by the structure of \HazelnutLive is to resume evaluation from where it left off after an edit that amounts to hole filling.
+We call this feature \emph{fill-and-resume}, by analogy to the ``edit-and-resume'' features available in some systems, e.g.
+Visual Studio (see below for a comparison)\todo{cite and compare}{}.
+
+\todo{fix definition of instantiation in the agda}The key is to interpret the hole environment as a delayed substitution, and indeed this is exactly the interpretation suggested for closures in contextual modal type theory by \citet{Nanevski2008}.
+We can derive the hole filling operation, $\instantiate{d}{u}{d'}$, defined in Fig.~\ref{fig:substitution}, from the contextual substitution operation of CMTT.
+Notice that unlike standard substitution, which is capture-avoiding, hole filling imposes no condition on the binder when passing into the body of a lambda expression---the expression that fills a hole can, of course, refer to variables in scope where the hole appears.
+When hole filling encounters an empty closure for the hole being instantiated, $\instantiate{d}{u}{\dehole{u}{\sigma}{}}$, the result is $[\instantiate{d}{u}{\sigma}]d$, i.e.
+the delayed substitution is applied to the fill expression, $d$, after first recursively filling any instances of hole $u$ that may appear in $\sigma$.
+Hole filling for non-empty closures is analagous (the enveloped expression can be discarded).
+Note that this is the reason why we cannot interpret a non-empty hole as an empty hole of arrow type applied to the enveloped expression---the hole filling operation would not operate as expected under this interpretation.
+
 \input{fig-substitution}
-\input{fig-multi-step}
 
-{\color{gray}\blindtext}
-
-\rkc{some of these statements may need additional typing premises}
-
-\begin{theorem}[Instantiation]
-  If $\hasType{\hDelta}{\hGamma}{\dexp}{\tau}$
-  and $\Dbinding{u}{\hGamma'}{\htau'}\in\hDelta$
+The following theorem characterizes the static behavior of hole filling.
+\begin{thm}[Filling]
+  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\dexp}{\tau}$
   and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
   then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
-\end{theorem}
+\end{thm}
+\begin{proof}
+We prove a stronger version of the theorem which accounts for both terms and finite substitutions (see \cite{Nanevski2008}).
+In each case, we proceed by rule induction on the first assumption, appealing to the finite substitution lemma in the corresponding cases.
+\end{proof}
 
-\autoref{fig:multi-step} defines the multi-step relation as the reflexive,
-transitive closure of the single-step relation.
-
-\begin{lemma}[Finality]
-  If $\isFinal{\dexp}$ and $\multiStepsTo{\dexp}{\dexp'}$
-  then $\dexp=\dexp'$.
-\end{lemma}
-
-\begin{theorem}[Commutativity]
-  If $\multiStepsTo{\dexp_0}{\dexp_1}$
-  then $\multiStepsTo{\instantiate{\dexp}{u}{\dexp_0}}
-                     {\instantiate{\dexp}{u}{\dexp_1}}$.
-\end{theorem}
-
-\begin{theorem}[Confluence]
-  If $\multiStepsTo{\dexp}{\dexp_1}$
-  and $\multiStepsTo{\dexp}{\dexp_2}$
-  then there exists $\dexp'$ such that
-  $\multiStepsTo{\dexp_1}{\dexp'}$ and
-  $\multiStepsTo{\dexp_2}{\dexp'}$.
-\end{theorem}
-
-\begin{corollary}[Final Confluence]
-  If $\multiStepsTo{\dexp}{\dexp_1}$
-  and $\isFinal{\dexp_1}$
+For the fill-and-resume feature to be reasonable, we need the following commutativity property, where we write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive, transitive closure of  stepping (see Fig.~\ref{fig:multi-step} in Appendix~\ref{sec:additional-defns}).
+In words, if there is some sequence of steps that go from $d_1$ to $d_2$, then you can fill the hole either at the beginning or at the end of that step sequence because if you do the former, you can take some sequence of steps to the latter.
+\begin{thm}[Commutativity]
+  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$
+  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and $\multiStepsTo{\dexp_1}{\dexp_2}$
+  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
+                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
+\end{thm}
+Note that this property follows because the specification in Sec.~\ref{sec:calculus} left evaluation order unspecified (i.e.
+formally, we omitted the bracketed premises).
+It is not the case in general that resuming from $\instantiate{\dexp'}{u}{d_2}$ will cause sub-expressions to be evaluated in the same order as they would have been if we had restarted from $\instantiate{\dexp'}{u}{d_1}$ and selected sub-expressions in, for example, an eager, left-to-right manner, because the hole that is being filled may have been ``skipped past''.
+In other words, this simple fill-and-resume protocol only works for languages where evaluation order ``does not matter''.
+There are various ways to encode this intuition.
+One way to do so is by establishing a confluence property (which is closely related to the Church-Rosser property \cite{church1936some}).
+The most general confluence property does not hold for the dynamic semantics in Sec.~\ref{sec:calculus} for the usual reason: we do not reduce under binders (\citet{DBLP:conf/birthday/BlancLM05} discuss the standard counterexample).
+We could recover confluence by specifying reduction under binders, either generally or in a more restricted form where only closed sub-expressions are reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit}, but reduction under binders is at odds with the standard approaches used in implementing programming languages \cite{DBLP:conf/birthday/BlancLM05}.
+A more satisfying approach is to consider confluence modulo equality \cite{Huet:1980ng}.
+The simplest approach is to restrict our interest to top-level expressions of base type that result in values, in which case the following special case of confluence does hold (trivially when the only base type has a single value, but also more generally for other base types).
+\begin{lem}[Base Confluence]
+  If $\hasType{\Delta}{\emptyset}{\dexp}{b}$ and 
+  $\multiStepsTo{\dexp}{\dexp_1}$
+  and $\isValue{\dexp_1}$
   and $\multiStepsTo{\dexp}{\dexp_2}$
   then $\multiStepsTo{\dexp_2}{\dexp_1}$.
-  \begin{proof}
-    By Confluence and Finality.
-  \end{proof}
-\end{corollary}
-
-\begin{theorem}[Resumption]
-  If $\multiStepsTo{\dexp_1}{\dexp_2}$
-  and $\isFinal{\dexp_2}$
-  and $\multiStepsTo{\instantiate{\dexp_3}{u}{\dexp_1}}{\dexp_4}$
-  and $\isFinal{\dexp_4}$
-  then $\multiStepsTo{\instantiate{\dexp_3}{u}{\dexp_2}}{\dexp_4}$.
+\end{lem}
+We can then prove the following property, which establishes that fill-and-resume is sound.
+\begin{thm}[Resumption]
+  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp}{b}$
+  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ 
+  and $\multiStepsTo{\dexp}{\dexp_1}$
+  and $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp}}{\dexp_2}$
+  and $\isValue{\dexp_2}$
+  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}{\dexp_2}$.
   \begin{proof}
     By Commutativity,
-    $\multiStepsTo{\instantiate{\dexp_3}{u}{\dexp_1}}
-                  {\instantiate{\dexp_3}{u}{\dexp_2}}$.
-    By Final Confluence, we can conclude.
+    $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp}}
+                  {\instantiate{\dexp'}{u}{\dexp_1}}$.
+    By Base Confluence, we can conclude.
   \end{proof}
-\end{theorem}
+\end{thm}
 
-Talk about weak vs strong conversion vs sharing -- see \url{http://gallium.inria.fr/~maranget/papers/jwk60.pdf} and cited work in the intro.
-See also \url{https://www.lri.fr/~blsk/Docs/Balabonski-FullLaziness-POPL12.pdf}??
-Original Church-Rosser paper -- \url{http://www.ams.org/journals/tran/1936-039-03/S0002-9947-1936-1501858-0/S0002-9947-1936-1501858-0.pdf}.
+The proofs are outlined in Appendix~\ref{sec:hole-filling}, along with a number of other necessary lemmas and definitions.
+In particular, we need additional machinery to properly handle the situation where we have filled a non-empty hole where a step had taken place in the original evaluation.
+These proofs and definitions are not included in the mechanization for two reasons.
+First, fill-and-resume is an optimization, and as just discussed, these properties, unlike those in Sec.~\ref{sec:calculus}, would not be conserved by some reasonable extensions of the core calculus.
+Second, it would require a significantly more complex representation of hole environments to encode the hole filling operation.
+Under the simple representation that suffices for the developments in Sec.~\ref{sec:calculus}, Agda cannot be convinced that the definition is well-founded (\citet{Nanevski2008} establish that it is in fact well-founded).
+
+Hole filling in the external language is understood in terms of hole filling in its expansion.
+Note that in practice, it may be useful to cache results from several previous expansions.
+For example, consider two edits, the first filling a hole $u$ with the number $2$, and the next applying the $+$ operator, resulting in $2 + \dehole{v}{\sigma}{}$.
+This second edit is not a hole filling edit with respect to the immediately preceding edit state, but it can be understood as filling hole $u$ with $2 + \dehole{v}{\sigma}{}$.
+Another related case arises in lab notebook interfaces like that of Jupyter/IPython \cite{PER-GRA:2007}.
+In this case, each cell can be understood as a series of \li{let} bindings ending in a hole, which is filled by the next cell.
+The result from each cell can be cached to avoid recomputing the environment from preceding cells.
+This contextual modal interpretation of lab notebook cells (and read-eval-print loops as a restricted case where edits to previous cells are impossible) is novel.
+
+So far, we have understood hole filling as an extralinguistic operation.
+However, a key feature CMTT that we have not yet touched on is the \emph{internalization} of metavariable binding and contextual substitution via the contextual modal types, $[\hGamma]\htau$, which are introduced by the operation $\mathsf{box}(\hGamma.d)$ and eliminated by the operation $\mathsf{letbox}(d_1, u.d_2)$.
+A hole filling can be interpreted as an expression of contextual modal type, and the act of hole filling followed by evaluation to the next possibly-indeterminate edit state as evaluation under the binder of a suitable $\mathsf{letbox}$ construct, which is enabled by the dynamic semantics in Sec.~\ref{sec:calculus}.
+This interpretation allows us to \emph{compute} hole fillings, rather than simply stating them, by specifying non-trivial expressions of modal type.
+This could serve as the basis for a \emph{live} computational hole refinement system, extending the capabilities of purely static hole refinement systems like those available in some proof assistants, e.g.
+the elaborator reflection system of Idris \cite{brady2013idris,DBLP:conf/icfp/ChristiansenB16} and the tactic system\todo{right?
+read a bit more}{} of Beluga \cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
+Each applied hole filling serves as a boundary between dynamic \emph{edit stages}.
+This contextual modal interpretation of live staged hole refinement nicely mirrors the modal interpretation of staged argument evaluation \cite{Davies:2001op}.
+We leave further development of this interpretation of hole refinement to future work.
 

### 2018-04-10 calculus.tex

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Matt's high-level Writing comments for Cyrus (for Section 3 mostly):
+
+% Please, please please: Use line breaks.  Git does't work well without them.
+% Change your editor or editor settings please.
+
+% Cyrus, your writing is generally very good, but is often harder for
+% me to digest than it should be, as a reader.  I think it'd be easier
+% to read on a first pass if we followed my preferences below, which I
+% humbly offer as gentle suggestions:
+
+% I think that some these suggestions are especially important for reading
+% really wide columns of text, as in this new PACMPL format. :)
+
+% My nits boil down to these preferences:
+
+% 0. NO PASSIVE VOICE --- EVER.  Instead, use ``We'' as the subject,
+% and choose a more interesting verb than 'is' for the sentence.
+% (Emacs says that there are 100 occurences of the word 'is' in the
+% text below.  That's way, way too much passive voice.)
+
+% 1. Shorter sentences.
+
+% 2. Fewer parenthenticals: just use more sentences instead.
+
+% 3. Shorter paragraphs, with more defined 'paragraph structure'. (E.g., topic sentence)
+
+% 4. Shorter sentence subjects (with subject closer to verb).
+
+% 5. Use macros for fig and section references.
+
+% 6. If a sentence introduces a figure, that sentence should be the
+% first sentence of a paragraph, and the figure reference should be
+% the first word of that sentence, ideally.  As a reader, then I look
+% at figures first, and find the paragraphs for them easily by
+% scanning the text.
+
+% 7. For each greek letter, prefix with an english word 'classifier',
+% e.g., 'external expression~$e$', not just '$e$'.  Use way, way fewer
+% commas please.
+
+% 8. Use singlar over plural sentence subjects, whenever possible;
+% e.g., an ``empty hole is written..'' not ``empty holes are
+% written''; or ``an empty hole synthesizes'' not ``empty holes
+% synthesize''
+
+% 9. Fewer commas. Each 'extra' comma is a burden to the reader (it's
+% asking the reader to allocate a stack frame in their brain!)
+
+% 10. Citations in this Names+Year form break sentence and paragraph
+% flow for the reader, sometimes quite severly.  Put long citation
+% chains at the ends of sentences (or better yet, in their own little
+% paragraph, as I did below).
+
+% *** I'm transforming the next paragraph below as a demonstration,
+% creating smaller, more cohesive paragraphs with shorter sentences.
+% See what you think.
+%
+% I'm not doing all of Section 3, tho, since
+%
+% (1) you may hate what I've done, and
+%
+% (2) I don't want to do a lot of work that will just be
+% reverted/resented :)
+
+% BTW: My version is _shorter_, not longer than the original :)
+%
+% Also, In my humble opinion, it is way, way easier to read. ;)
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \input{fig-bidirectional-typing}
+
+We give an overview of the static semantics of the \HazelnutLive
+external language.
+
+\Figref{fig:bidirectional-typing} defines \emph{bidirectional} typing
+%
+as two mutually defined judgements: The type synthesis
+judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
+for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
+assumptions of the form $x : \htau$ in the usual
+manner \cite{pfpl,tapl}.
+%
+The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
+expression~$\hexp$ against a given type~$\htau$.
+%
+Algorithmically, analysis accepts a type as input, and synthesis gives
+a type as output.
+%
+We use synthesis for each of the programmer's ``top level'' external
+exressions~$e$, giving the corresponding type as feedback.
+
+% Algorithmically, the type is an output of type synthesis but an input of type analysis.
+
+The benefit of specifying the \HazelnutLive external language
+bidirectionally is that the programmer does not need to annotate each
+hole with a type
+%
+\cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}.
+%
+These typing rules closely follow the \Hazelnut type
+system \cite{popl-paper}; we discuss minor differences below.
+
+We supplement each expression hole from \Hazelnut with a
+hole \emph{name}, which are new to \HazelnutLive; an empty hole is
+written $\hehole{u}$, where $u$ is a unique name for the hole.
+%
+Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
+%
+If an empty hole appears where an expression of some other type is
+expected, e.g. under an explicit ascription (governed by Rule {SAsc})
+or in the argument position of a function application (governed by
+Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
+Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
+type $\htau$, then it can be checked against any \emph{consistent}
+type, $\htau'$.
+
 \input{fig-type-consistency}
 
-Let us start with an overview of the static semantics of the \HazelnutLive external language.
-Types, $\htau$, classify external expressions, $e$, according to the type system in Fig.
-\ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up below).
-The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements.
-The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ under typing context $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}.
-The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$.
-Algorithmically, the type is an output of type synthesis but an input of type analysis.
-At the top level, we start with synthesis.
-% Algorithmically, the type is an output of type synthesis but an input of type analysis.
+%%%%%%%%%%%%%%%%%%%% MATT stopped editing Cyrus's text here.
 
-The benefit of specifying the \HazelnutLive external language bidirectionally is that the programmer does not need to annotate each hole with a type.
-Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole notation is taken from \Hazelnut; hole names are new to \HazelnutLive).
-Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$.
-If an empty hole appears where an expression of some other type is expected, e.g.
-under an explicit ascription (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$.
 Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.

### 2018-04-11 calculus.tex
added line breaks at sentence boundaries in sec 3 and addressed ravis comments in blue, along with a few other small tweaks. ready for Matt to give it a pass.

@@
 We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory. 
-We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} (see Sec.~\ref{sec:agda-mechanization} for some additional details, and the supplemental material for the full mechanization).
-% The proofs of the metatheorems are established by this mechanization; the proofs are only very briefly outlined in this section.
+We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
+(see Sec.~\ref{sec:agda-mechanization} for some additional details, and the supplemental material for the full mechanization).
 
@@
 We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. 
-External expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec.
-2?}{} for discussion of manual, semi-automated and fully automated hole entry methods).
+External expressions correspond to programs as entered by the programmer 
+(see Sec.~\ref{sec:implementation}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
 Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
-We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts.
+We distinguish the external and internal languages because 
+(1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and
+(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
 This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
@@
 
-We give an overview of the static semantics of the \HazelnutLive
-external language.
+We start with the type system of the \HazelnutLive
+external language, which closely follows the \Hazelnut type
+system \cite{popl-paper}; we discuss the minor differences as they come up below.
 
-\Figref{fig:bidirectional-typing} defines \emph{bidirectional} typing
+\Figref{fig:bidirectional-typing} defines the type system in the \emph{bidirectional} style 
 %
-as two mutually defined judgements: The type synthesis
+with two mutually defined judgements \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}. The type synthesis
 judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
@@
 %
-We use synthesis for each of the programmer's ``top level'' external
-exressions~$e$, giving the corresponding type as feedback.
+We start with synthesis for the programmer's ``top level'' external
+exression.
 
@@
 
-The benefit of specifying the \HazelnutLive external language
-bidirectionally is that the programmer does not need to annotate each
-hole with a type
+The primary benefit of specifying the \HazelnutLive external language 
+bidirectionally is that the programmer does not need to annotate each hole with a type. 
 %
-\cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}.
-%
-These typing rules closely follow the \Hazelnut type
-system \cite{popl-paper}; we discuss minor differences below.
-
-We supplement each expression hole from \Hazelnut with a
-hole \emph{name}, which are new to \HazelnutLive; an empty hole is
-written $\hehole{u}$, where $u$ is a unique name for the hole.
+An empty hole is
+written $\hehole{u}$, where $u$ is a name for the hole, which we tacitly assume is unique. 
+Holes did not have names in \Hazelnut. 
 %
@@
 %
-\rkc{perhaps too much info / unnecessary to say Hazelnut had numbers instead of unit?} 
-%
 Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.
@@
 
-The rule governing function application, Rule {SAp}, similarly treats expressions of hole type in function position as if they were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.
+The rule governing function application, Rule {SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.
 
+\vspace{-4px}
 \subsection{Expansion}
 \label{sec:expansion}
+\vspace{-1px}
 
@@
 The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. 
-We say more about hole contexts below.
+We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below. 
 The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$.
-The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
+The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is necessarily consistent with $\htau$.
 %
-\rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
-entirely standard, because of the hole context, so perhaps worth mentioning.}
+% \rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
+% entirely standard, because of the hole context, so perhaps worth mentioning.}
 %
@@
 The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e.
-the following standard unicity property holds of the type assignment system:
+the following standard unicity property holds of the type assignment system.
 \begin{thm}[Type Assignment Unicity]
@@
 
-Before describing the rules in detail, let us state a few other useful theorems. The following theorem establishes that an expansion exists for every well-typed external expression.
+Before describing the rules in detail, let us state a few other useful theorems. 
+The following theorem establishes that an expansion exists for every well-typed external expression.
  \begin{thm}[Expandability] \label{thm:expandability}~
@@
 
-The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} (so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward).
-Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation onto unannotated external lambdas based on the given type.
-The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.
+The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}
+(so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). 
+Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation when expanding an unannotated external lambda based on the given type.
+The rules governing holes and the rules governing function applications and ascriptions, which both perform \emph{cast insertion}, are more interesting, so let us consider both cases in turn.
+
+\subsubsection{Hole Expansion}\label{sec:hole-expansion} 
+Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$.
 %
-\rkc{this list (holes, func app, and ascription) has three items, but there are
-only two subsections. see comment at beginning of Sec 3.2.2.}
-
-\subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
-%
-\rkc{perhaps add ``empty and non-empty'' before hole closures.}
 
 The hole name, $u$, on a hole closure identifies the external hole that the hole closure corresponds to. 
-Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution (though in the initial expansion, the uniqueness condition happens to hold because evaluation has yet to occur).
+Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution. 
 For example, the result from Fig.~\ref{fig:grades-example} showed four closures for the hole named 1\todo{update with actual number (four?) and hole name later}. 
@@
 The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. 
-The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$,   for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
+The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$, for each hole name, $u$.\footnote{
+We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} 
+This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
 In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. 
@@
 
-Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a.
-deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}.
-Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT\todo{mention reason in Sec 4}.
+Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} 
+We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}. 
+Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT (see Sec.~\ref{sec:resumption}).
 
 \subsubsection{Cast Insertion}\label{sec:cast-insertion} 
-%
-\rkc{consider renaming subsection to ``Function Application and Cast
-Intersion''. and adding a transition sentence like ``The remaining development
-of expansion deals with the delicate interplay between function application and
-cast insertion.''}
 %
@@
 %On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
-}\todo{talk about elsewhere?
-maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory.
-This leads us to the solution to this problem: cast insertion.
+}
+\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} 
+By our  interpretation of hole types as unknown types from gradual type theory, we can address the problem by performing cast insertion. 
 %
-\rkc{Instead of the last two sentences, maybe something like: ``By analogy
-between hole types in our system and unknown types from gradual type theory, we
-address the problem with the same mechanism used in gradual type theory: cast
-insertion.''}
 
-The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of some consistent type $\htau_2$.\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}
+The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of a consistent type $\htau_2$.%
+\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}
 
 Casts are inserted during the expansion of function applications and ascriptions. 
-The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type.
+The latter is more straightforward: 
+Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. 
 Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  
-Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.
-%
-\rkc{move this Note sentence to the same place where half-annotated lambdas were
-noted. This note breaks the flow here.}
+Note that we included ascription only for the sake of exposition---it can be defined using application together with the half-annotated identity, $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$, so application expansion, discussed next, is more general.
 
-Cast insertion by the application expansion rule, Rule~{ESAp}, has much the same intuition, but requires a bit more care. To understand the rule, consider the expansion for the example above:
+Cast insertion by the application expansion rule, Rule~{ESAp}, requires a bit more care. 
+To understand the rule, consider the expansion for the example above:
 \[\hap{\dcasttwo{(\halam{x}{\tehole}{\underline{\hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}}})}{\tarr{\tehole}{\tehole}}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}
 \]
-Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. A cast is inserted on both the function expression, $x$, and the the argument, $c$.
+Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. 
+A cast has been inserted on both the function expression, $x$, and the the argument, $c$.
 
@@
 The conclusion of the rule inserts the corresponding cast. 
-Note that we go through type synthesis and analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.
+We go through type synthesis then analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.
 
-The conclusion of Rule~{ESAp} also inserts a cast on the argument's expansion, from the type it is assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type, $\tehole$, of the matched arrow type of the function expression, here $\tehole$.
+The conclusion of Rule~{ESAp} inserts the cast on the argument's expansion, from the type it is assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type of the matched arrow type of the function expression, here $\tehole$.
 
@@
 In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$. 
-We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), but it is safe to do so (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).
+We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), 
+but it is safe to eliminate these during expansion (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).
 
@@
 
-To recap, the result of expansion is a well-typed internal expression with hole closures and casts, where the former corresponds to metavariable closures from CMTT \cite{Nanevski2008}, and the latter corresponds to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}.
+To recap, the result of expansion is a well-typed internal expression with hole closures and casts, 
+where the former corresponds to metavariable closures from CMTT \cite{Nanevski2008}, 
+and the latter corresponds to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}. 
 However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.
@@
 The paper therefore proved only a subject reduction property (which is closely related to type preservation). 
-This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e.
-that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}.
+This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. 
 In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables.
-\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion\footnote{Permutation conversions are necessary to encode the commutative reductions of CMTT, which in turn are necessary to prove a strong normalization property.
-These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.}, which has been studied by \citet{DBLP:journals/iandc/Groote02}, but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on any standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
+\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion%
+\footnote{Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property.
+These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.},
+which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
+but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
 
@@
 The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. 
-Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}, is for cast failure to instead insert a membrane around the dynamic type error, much like a non-empty hole serves as a membrane around a static type error, allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.
+Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}\todo{where?}{}, is for cast failure to instead insert a membrane around the dynamic type error, 
+much like a non-empty hole serves as a membrane around a static type error, 
+allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.
 
-The dynamic semantics for \HazelnutLive specified in Figures~\ref{fig:isGround}-\ref{fig:contextual-dynamics} addresses the difficulties just outlined, resulting in a system capable of running incomplete programs without aborting when a hole is encountered.
+The dynamic semantics for \HazelnutLive specified in Figures~\ref{fig:isGround}-\ref{fig:step} addresses the difficulties just outlined, 
+resulting in a system capable of running incomplete programs without aborting when a hole is encountered. 
 This semantics is equipped with a meaningful notion of type safety (involving both preservation and progress). 
 We also establish that the standard notion of type safety falls out when running complete terms.
-% We also establish that the standard type safety theorem holds when running complete terms.
 
@@
 Fig.~\ref{fig:isFinal} defines the judgement $\isFinal{d}$, which distinguishes the final forms of the transition system. 
-There are two classes of final forms: (possibly-)boxed values and indeterminate forms.\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention.
+There are two classes of final forms: (possibly-)boxed values and indeterminate forms.%
+\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention. 
 Our judgemental formulation is more faithful to the mechanization and cleaner for our purposes, because we are distinguishing several classes of final forms.}
@@
 
-The judgement $\isBoxedValue{d}$ distinguishes (possibly-)boxed values, which correspond to the values from the cast calculus and include the classic values from the lambda calculus, distinguished by $\isValue{d}$, as well as two cast forms: casts between function types (except identity casts) and casts from a ground type to the hole type.
-In both cases, the cast must be on a boxed value.
-% (in the prior work, ground types and values were distinguished syntactically, but it is cleaner and more faithful to the mechanization to make these distinctions judgementally).
+The judgement $\isBoxedValue{d}$ distinguishes (possibly-)boxed values, which correspond to the values from the cast calculus and include the classic values from the lambda calculus, distinguished by $\isValue{d}$, as well as two cast forms:
+casts between function types (except identity casts) 
+and casts from a ground type to the hole type. In both cases, the cast must be on a boxed value.
 
-The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}).
+The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, 
+so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:resumption}). 
 The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). 
@@
 The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}. 
-Top-level transitions, i.e.
-\emph{steps}, $\stepsToD{}{d}{d'}$, are governed by a single rule, Rule {Step} in Fig.~\ref{fig:step}, which (1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$; (2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in Fig.~\ref{fig:instruction-transitions}; and (3) places $d_0'$ in the selected position, which is marked\footnote{In the literature, the form $\evalhole$ in the grammar of evaluation contexts is referred to as the \emph{hole}, but this hole is a technical device entirely orthogonal to the holes of this paper, so we use the term ``mark'' instead.} in the evaluation context by $\evalhole$, to obtain $d'$.
+Top-level transitions, i.e. \emph{steps}, $\stepsToD{}{d}{d'}$, are governed by a single rule, Rule {Step} in Fig.~\ref{fig:step}, which 
+(1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$; 
+(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in Fig.~\ref{fig:instruction-transitions}; 
+and (3) places $d_0'$ back in the selected position, which is marked%
+\footnote{In the literature, the form $\evalhole$ in the grammar of evaluation contexts is referred to as the \emph{hole}, but this hole is a technical device entirely orthogonal to the holes of this paper, so we use the term ``mark'' instead.}
+in the evaluation context by $\evalhole$, to obtain $d'$.
 This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
 Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
-It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} compares the two approaches).\todo{Put the search rules in the appendix?}{}
+It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
+\todo{Put the search rules in the appendix?}{}
 
@@
 
-\subsubsection{Application and Substitution} Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the standard beta reduction transition.
-The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} can be included to specify an eager, left-to-right evaluation strategy, or excluded to leave the evaluation order unspecified.
-We do the latter in our metatheory, both for the sake of generality and for reasons we return to in Sec.~\ref{sec:commutativity}.
+\subsubsection{Application and Substitution} 
+Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the standard beta reduction transition. 
+The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} can be included to specify an eager, left-to-right evaluation strategy, 
+or excluded to leave the evaluation order unspecified. 
+We do the latter in our metatheory, both for the sake of generality and for reasons we return to in Sec.~\ref{sec:resumption}.
 
-Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. The only cases of special interest are when substitution reaches a hole closure:
+Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. 
+The only cases of special interest arise when substitution reaches a hole closure:
 \[
@@
 Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g.
-because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?}.
-Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.
+because a function with a hole closure in its body is applied multiple times.%
+\todo{show example here or refer back to sec 2?}{} 
+Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}\todo{where?}{}).
 
@@
 
-The other possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g.
-$\hap{(\dehole{u}{\sigma}{})}{c}$.
+The other possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$. 
 In this case, no further progress can be made, so the function application as a whole is also deemed indeterminate by Rule {IAp} in Fig.~\ref{fig:isFinal}. 
-Note that we are careful to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.\todo{prove this as stated from Ian's disjointness proofs}\todo{do we need a typing premise?}
+Note that we are careful to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.%
+\todo{prove this as stated from Ian's disjointness proofs}{}%
+\todo{do we need a typing premise?}{}
 \begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
@@
 \subsubsection{Casts}
-The remaining instruction transition rules handle casts directly.
-Rule {ITCastId} drops identity casts.
-As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term cast to hole type is statically obscure, so we need to wait until it is actually used at some other type, via a cast away from hole type, to be able to determine whether there is a problem.
+Rule {ITCastId} drops identity casts. The remaining instruction transition rules assign meaning to non-trivial casts. 
+As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term cast to hole type is statically obscure, 
+so we need to wait until it is actually used at some other type, via a cast away from hole type, to be able to determine whether there is a problem. 
 Rules {ITCastSucceed} and {ITCastFail} handle this situation when the two types involved are ground types (Fig.~\ref{fig:isGround}). 
-If the two ground types are equal, then the cast succeeds and the cast forms can be dropped.
+If the two ground types are equal, then the cast succeeds and the cast can be dropped. 
 If they are not equal, then the cast fails and the failed cast form, $\dcastfail{\dexp}{\htau_1}{\htau_2}$, arises. 
-Rule {TAFailedCast} specifies as expected from Rule {ITCastFail} that a failed cast is well-typed exactly when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type disequal to $\tau_1$.
+Rule {TAFailedCast} specifies that a failed cast is well-typed exactly when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type disequal to $\tau_1$. 
 Rule {IFailedCast} specifies that a failed cast operates as an indeterminate form once $d$ is final. 
-This allows evaluation to proceed after cast failure, as with holes.
+This allows evaluation to proceed after cast failure, much as with hole closures.
 
 The two rules just described operate at ground type. 
-The two remaining instruction transition rules, Rule {ITGround} and {ITExpand}, operate as technical devices that allow us to restrict our interest to ground types exclusively by inserting intermediate casts from (resp.
-to) non-ground types to (resp.
-from) consistent ground types.
-In the core calculus, the only non-ground types are the arrow types, so the grounding judgement, $\groundmatch{\tau_1}{\htau_2}$, defined in Fig.~\ref{fig:groundmatch}, produces the ground arrow type, $\tarr{\tehole}{\tehole}$.
+The two remaining instruction transition rules, Rule {ITGround} and {ITExpand}, operate as technical devices that allow us to restrict our interest to ground types exclusively by inserting intermediate casts from non-ground type to  a consistent ground type, and \emph{vice versa}.
+In this core calculus, the only non-ground types are the arrow types, so the grounding judgement $\groundmatch{\tau_1}{\htau_2}$ defined in Fig.~\ref{fig:groundmatch}, produces the ground arrow type, $\tarr{\tehole}{\tehole}$.
 More generally, this judgement is governed by the following invariant:
@@
 
-\subsubsection{Type Safety} The purpose of establishing type safety is to ensure that the static and dynamic semantics of a language cohere.
+\subsubsection{Type Safety} 
+The purpose of establishing type safety is to ensure that the static and dynamic semantics of a language cohere. 
 We follow the approach developed by \citet{wright94:_type_soundness}, now standard \cite{pfpl}, which distinguishes two type safety properties, preservation and progress. 
@@
 \noindent
-The proof relies on an analagous preservation lemma for instruction transitions and a standard substitution lemma. Hole closures can disappear during evaluation, so we rely on weakening of $\Delta$.
+The proof relies on an analagous preservation lemma for instruction transitions and a standard substitution lemma (stated in Appendix~\ref{sec:additional-defns}). 
+Hole closures can disappear during evaluation, so we rely on weakening of $\Delta$.
 
@@
 \noindent
-Explicitly accounting for indeterminate forms, which are rooted at either a hole closure or a failed cast, is the key to being able to establish the progress theorem under a non-empty hole context.
+The key to being able to establish the progress theorem under a non-empty hole context lies in the fact that we have explicitly accounted for indeterminate forms, i.e.
+those rooted at either a hole closure or a failed cast.
 The proof relies on canonical forms lemmas stated in Appendix~\ref{sec:additional-defns}.
 
-\subsubsection{Complete Programs} Although our focus in this paper is on running incomplete programs, this machinery does not interfere with reasoning about the behavior of complete programs, i.e.
+\subsubsection{Complete Programs} 
+Although our focus in this paper is on running incomplete programs, it is helpful to know that the necessary machinery does not interfere with reasoning about the behavior of complete programs, i.e.
 those with no holes.
@@
 \label{sec:agda-mechanization}
+\vspace{-3px}
 
 The supplemental material includes our mechanization of the semantics and metatheory of \HazelnutLive, 
-including proofs of all of the theorems stated above and any necessary lemmas\todo{make this so ian!}{}. 
-We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis},  in part because it was also used in the mechanization of \Hazelnut \cite{popl-paper}. 
-No proof automation is used because our intention is to communicate the proof
-structure itself. The basic approach is standard: we model judgements as 
-dependent inductive datatypes. 
+including proofs of all of the theorems stated above and all necessary lemmas\todo{make this so ian!}{}. We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis} 
+(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
+Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation.
+Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf.
+Sec.~\ref{sec:intro}).
+The basic approach is standard: we model judgements as 
+inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite somtehing?}{}. 
 We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
@@
 The supplemental material also includes a browser-based implementation of \HazelnutLive. 
-The implementation is a functional reactive program \cite{DBLP:conf/pldi/CzaplickiC13} written using the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} together with the OCaml \lismall{React} library \cite{OcamlReact} and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}.
-A screenshot of the user interface, which implements all of the live programming features described in Sec.~\ref{sec:examples}, but for the more austere language of this section, is shown in Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{}.
+The implementation is a functional reactive program \cite{DBLP:conf/pldi/CzaplickiC13} 
+written using the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
+together with the OCaml \lismall{React} library \cite{OcamlReact} 
+and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}. 
+Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{} provides a screenshot of the user interface, which implements all of the live programming features described in Sec.~\ref{sec:examples}, but for the more austere language of this section.
 
-The editor component of the implementation is derived from \Hazelnut. It exposes a language of edit actions that insert holes  
-automatically to guarantee that every edit state has some (possibly incomplete) type. Formally,
-this corresponds to the Sensibility theorem from the prior work \cite{popl-paper}. 
-By composing this theorem with the Expandability, Typed Expansion and Progress theorems from this paper, we establish a uniquely powerful 
+The editor component of the implementation is derived from \Hazelnut. 
+It exposes a language of structured edit actions (summarized in the sidebar on the left of Fig.~\ref{fig:screenshot}) that insert holes automatically to guarantee that every edit state has some (possibly incomplete) type.
+Formally,
+this corresponds to the Sensibility property from the prior work \cite{popl-paper}. 
+By composing this property with the Expandability, Typed Expansion and Progress properties from this paper, we establish a uniquely powerful 
 invariant: that every edit state is both statically and dynamically meaningful. 
 In other words, live feedback does not ``flicker in and out'' or reflect past edit states. 
-The primary purpose of the included implementation is to serve as a proof-of-concept of this invariant, and to be of use to researchers studying the calculus as presented.% Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.
+The primary purpose of the included implementation is to serve as a proof-of-concept of this invariant, and to be of use to researchers studying the calculus as presented in this section.
+% Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.
 
-We are also integrating the mechanisms developed in this paper into the \Hazel programming environment, which is being developed by \citet{HazelnutSNAPL} with the goal of providing a full-scale hole-driven development experience rooted in the type theoretic foundations established by \Hazelnut.
+We are also separately integrating the mechanisms developed in this paper into the \Hazel programming environment being developed by \citet{HazelnutSNAPL} with the goal of realizing a full-scale hole-driven development experience rooted in the type theoretic foundations established by \Hazelnut.
 Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper.
-We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as the \Hazel language evolves.
-The included implementation provides a link to this evolving effort.
-For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed certain syntactic conveniences that are not yet implemented in \Hazel, but that are available in Elm and standard in ML-like languages.
+For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
+We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
+\todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
+The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}
 
@@
 
-The result of evaluation is a final internal expression with hole closures, each with an associated hole environment, $\sigma$.
+%The result of evaluation is a final internal expression with hole closures, each with an associated hole environment, $\sigma$.
 These hole environments can be reported directly to the programmer, e.g.
@@
 by informing an edit action synthesis and suggestion system.
-In any case, once the programmer does perform one or more edit actions to fill in a hole, a new result must be computed.
+In any case,
+When the programmer performs one or more edit actions to fill in a hole in the program, a new result must be computed.
 Na\"ively, the system would need to compute the result ``from scratch'' on each such edit.
@@
 
-\todo{fix definition of instantiation in the agda}The key is to interpret the hole environment as a delayed substitution, and indeed this is exactly the interpretation suggested for closures in contextual modal type theory by \citet{Nanevski2008}.
+\todo{fix definition of instantiation in the agda}The key is to interpret the hole environments as delayed substitutions, and indeed this is exactly the interpretation suggested for closures in contextual modal type theory by \citet{Nanevski2008}.
 We can derive the hole filling operation, $\instantiate{d}{u}{d'}$, defined in Fig.~\ref{fig:substitution}, from the contextual substitution operation of CMTT.

### 2018-04-11 hazelnut-dynamics.tex
a bunch of progress on sec 3.3
finished draft of sec3 and 4
added line breaks at sentence boundaries in sec 3 and addressed ravis comments in blue, along with a few other small tweaks. ready for Matt to give it a pass.

@@
 \let\li\lstinline
+\newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
 
@@
 \newtheoremstyle{slplain}% name
-  {.16\baselineskip\@plus.1\baselineskip\@minus.1\baselineskip}% Space above
-  {.16\baselineskip\@plus.1\baselineskip\@minus.1\baselineskip}% Space below
+  {.15\baselineskip\@plus.1\baselineskip\@minus.1\baselineskip}% Space above
+  {.15\baselineskip\@plus.1\baselineskip\@minus.1\baselineskip}% Space below
   {\slshape}% Body font
@@
 \newtheorem{defn}[thm]{Definition}
+\newtheorem{lem}[thm]{Lemma}
+\newtheorem{prop}[thm]{Proposition}
 % \newtheorem{cor}[section]{Corollary}     
@@
 \input{calculus}
-\input{implementation}
+% \input{implementation}
 \input{related-work}
@@
 
+\clearpage
 \bibliography{references,all.short,hazel_NSF}

### 2018-04-11 calculus.tex
partial pass over Sec 3 -- will resume later today

@@
 \label{sec:external-statics}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Matt's high-level Writing comments for Cyrus (for Section 3 mostly):
-
-% Please, please please: Use line breaks.  Git does't work well without them.
-% Change your editor or editor settings please.
-
-% Cyrus, your writing is generally very good, but is often harder for
-% me to digest than it should be, as a reader.  I think it'd be easier
-% to read on a first pass if we followed my preferences below, which I
-% humbly offer as gentle suggestions:
-
-% I think that some these suggestions are especially important for reading
-% really wide columns of text, as in this new PACMPL format. :)
-
-% My nits boil down to these preferences:
-
-% 0. NO PASSIVE VOICE --- EVER.  Instead, use ``We'' as the subject,
-% and choose a more interesting verb than 'is' for the sentence.
-% (Emacs says that there are 100 occurences of the word 'is' in the
-% text below.  That's way, way too much passive voice.)
-
-% 1. Shorter sentences.
-
-% 2. Fewer parenthenticals: just use more sentences instead.
-
-% 3. Shorter paragraphs, with more defined 'paragraph structure'. (E.g., topic sentence)
-
-% 4. Shorter sentence subjects (with subject closer to verb).
-
-% 5. Use macros for fig and section references.
-
-% 6. If a sentence introduces a figure, that sentence should be the
-% first sentence of a paragraph, and the figure reference should be
-% the first word of that sentence, ideally.  As a reader, then I look
-% at figures first, and find the paragraphs for them easily by
-% scanning the text.
-
-% 7. For each greek letter, prefix with an english word 'classifier',
-% e.g., 'external expression~$e$', not just '$e$'.  Use way, way fewer
-% commas please.
-
-% 8. Use singlar over plural sentence subjects, whenever possible;
-% e.g., an ``empty hole is written..'' not ``empty holes are
-% written''; or ``an empty hole synthesizes'' not ``empty holes
-% synthesize''
-
-% 9. Fewer commas. Each 'extra' comma is a burden to the reader (it's
-% asking the reader to allocate a stack frame in their brain!)
-
-% 10. Citations in this Names+Year form break sentence and paragraph
-% flow for the reader, sometimes quite severly.  Put long citation
-% chains at the ends of sentences (or better yet, in their own little
-% paragraph, as I did below).
-
-% *** I'm transforming the next paragraph below as a demonstration,
-% creating smaller, more cohesive paragraphs with shorter sentences.
-% See what you think.
-%
-% I'm not doing all of Section 3, tho, since
-%
-% (1) you may hate what I've done, and
-%
-% (2) I don't want to do a lot of work that will just be
-% reverted/resented :)
-
-% BTW: My version is _shorter_, not longer than the original :)
-%
-% Also, In my humble opinion, it is way, way easier to read. ;)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
-type $\htau$, then it can be checked against any \emph{consistent}
+type $\htau$, then it may be checked against any \emph{consistent}
 type, $\htau'$.
@@
 
-%%%%%%%%%%%%%%%%%%%% MATT stopped editing Cyrus's text here.
+Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.
+%
+The hole type is consistent with every type, and so, by the subsumption rule, expression holes may appear where an expression of any type is expected.
+The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}.
+%
+Type consistency is reflexive and symmetric, but it is \emph{not} transitive.
+%
+This relation stands in contrast to subtyping, which is anti-symmetric and transitive; subtyping may be integrated into a gradual type system following \citet{Siek:2007qy}).
 
-Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.
-The hole type is consistent with every type, and so, by the subsumption rule, expression holes can appear where an expression of any type is expected.
-The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}.
-Note that type consistency is reflexive and symmetric but it is not transitive (unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).
-
-Non-empty expression holes, written $\hhole{\hexp}{u}$, behave similarly.
-Rule {SNEHole} specifies that non-empty expression holes also synthesize hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type.
+Non-empty expression holes, written $\hhole{\hexp}{u}$, behave similarly to empty holes.
+%
+Rule {SNEHole} specifies that a non-empty expression hole also synthesizes hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type.
+%
 Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}\todo{briefly say something about binding inconsistencies?}
@@
 For the familiar forms of the lambda calculus, the rules again follow prior work.
-For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e.
-$b$ is the unit type).
-\Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
+%
+For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule {SConst} (i.e. $b$ is the unit type).
+%
+By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
 %
@@
 For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
-Half-annotated lambdas can appear in synthetic position  according to Rule {SLam}, which is standard \cite{Chlipala:2005da}.
-Unannotated lambdas can only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.
-To avoid the need for two separate rules, Rule {ALam} uses the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in Fig.~\ref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
-Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might include a synthetic rule for unannotated lambdas, e.g.
-as outlined by \citet{DBLP:conf/icfp/DunfieldK13}.
+%
+Half-annotated lambdas may appear in synthetic position according to Rule {SLam}, which is standard \cite{Chlipala:2005da}.
+%
+Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.
+%
+To avoid the need for two separate rules, Rule {ALam} uses the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% To Cyrus from Matt:
+%
+% Why say the following here? --- It's discussing a different design that we didn't pursue here.  We should move such discussion to related work.
+%
+%Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might include a synthetic rule for unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}.
+%%
 
@@
 
-Well-typed external expressions expand to well-typed internal expressions, $d$, for evaluation. 
-The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}. 
-Evaluation of internal expressions will be discussed in Sec.~\ref{sec:evaluation}.
+Each well-typed external expression~$e$ expands to a well-typed internal expression~$d$, for evaluation.
+%
+\Figref{fig:expansion} gives the rules governing expansion, and \Figref{fig:hasType} gives the rules governing type assignment for internal expressions.
+%
+\Figref{sec:evaluation} discusses internal expression evaluation.
 
-Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. 
-The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. 
-We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below. 
-The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$.
+As with the type system for the external language (above), 
+we specify expansion in the bidirectional style, via two mutually defined judgements.
+%
+The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ synthesizes a type~$\htau$ from~$\hexp$.
+%
+Synthetic expansion produces an internal expression~$d$ and a hole context~$\hDelta$.
+%
+%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
+We discuss hole contexts further below, including their role in the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$.
+
+The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks~$\hexp$ against~$\htau$ and produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$.
+%
 The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is necessarily consistent with $\htau$.
@@
 \noindent
-The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e.
-the following standard unicity property holds of the type assignment system.
+
+%The reason analytic expansion produces an expansion of consistent
+%type is because the subsumption rule, as previously discussed, allows
+%us to check an external expression against any type consistent with
+%the type the expression actually synthesizes, whereas every internal
+%expression can be assigned at most one type, i.e. the following
+%standard unicity property holds of the type assignment system.
+%
+Analytic expansion produces an expansion of consistent type:
+%
+Inuitively, the subsumption rule, as previously discussed, allows us
+to check an external expression against any type consistent with the
+type the expression actually synthesizes.
+
 \begin{thm}[Type Assignment Unicity]
@@
 Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
+%
 For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.
@@
 \end{thm}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% <<<<<<<<<<<<< Matt is here (latest pass over everything above is done -- 2018.04.11)
 

### 2018-04-11 calculus.tex
reword cite sentence and reinsert
fix typo

@@
 %
+\Secref{sec:related} dicusses how \HazelnutLive might be enriched with
+with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
+the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
+%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %
-\Figref{sec:evaluation} discusses internal expression evaluation.
+\Secref{sec:evaluation} discusses internal expression evaluation.
 

### 2018-04-11 related-work.tex
draft of sec 3.2 through 3.2.1 and a stub starting to talk about cast insertion
finished draft of sec3 and 4
uncommitted Related Work wip from several weeks ago
- breaking up the big list into smaller sections/lists
- importing some text from the grant

@@
 
+Program editors frequently encounter program text that is not well-formed with
+respect to the concrete syntax of the language \cite{popl-paper}.
+%
+Within the simplest editors, editor services that take well-formed program text
+or an abstract syntax tree as input are simply disabled when this occurs.
+%
+More sophisticated editors, including those generated by modern editor
+generators like Spoofax \cite{DBLP:conf/oopsla/KatsV10}, deploy \emph{error
+recovery} heuristics that silently insert tokens or discard tokens from the
+program so that the editor-internal representation is a standard abstract syntax
+tree
+\cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09}.
+%
+These heuristics are typically defined manually by the grammar designer, though
+certain heuristics can be generated semi-automatically by tools that are given a
+description of the scoping conventions of the language and of secondary
+notational conventions (\eg{}~whitespace conventions)
+\cite{DBLP:conf/oopsla/KatsJNV09,DBLP:conf/sle/JongeNKV09}.
+%
+These error recovery heuristics require guessing at the programmer's intent, and
+result in tokens being inserted or deleted from the program invisibly, so they
+are \emph{ad hoc} and can confuse the programmer
+\cite{DBLP:conf/oopsla/KatsJNV09}.
+
+\parahead{Hole-Driven Development}
+
+A cleaner alternative approach, and the approach that is central to our proposed
+research, is to add \emph{holes} to both the concrete and abstract syntax of the
+language.
+%
+Syntax trees with holes first arose in prior work on \emph{structure editors},
+which are program editors where every edit action directly causes changes to the
+abstract syntax tree of the program, rather than going indirectly through a
+parser.
+%
+Structure editors have a long history: the Cornell Program Synthesizer was
+developed in the early 1980s \cite{teitelbaum_cornell_1981} and there has been a
+steady stream of new designs in this space
+(\eg{}~\cite{ko_barista:_2006,Resnick:2009:SP:1592761.1592779,voelter_mbeddr:_2012,tillmann_touchdevelop:_2011,lamdu,ActiveCodeCompletion,Greenfoot,Deuce}).
+%
+For example, Scratch and other block-based editors have achieved success as a
+tool for teaching children how to program
+\cite{Resnick:2009:SP:1592761.1592779}. \texttt{mbeddr} is an editor for a
+C-like language \cite{voelter_mbeddr:_2012}, built using the commercially
+supported MPS structure editor workbench \cite{voelter2011language}, that is
+presently deployed to a number of customers.
+%
+TouchDevelop is a structure editor for an object-oriented language designed for
+use on mobile devices \cite{tillmann_touchdevelop:_2011}. Lamdu \cite{lamdu},
+Unison \cite{unison} and  isomorf~\cite{isomorf} are open source structure
+editors for functional languages similar to ML.
+
+%% There is reason to be optimistic, however, with recent studies 
+%% suggesting that programmers experienced with a modern keyboard-driven structure editor like \texttt{mbeddr} 
+%% can be highly productive \cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14}.
+
+Nevertheless, text-based program editors continue to dominate in practice, in
+part because text is flexible and familiar and because tools for interacting
+with text, like regular expressions and various differencing techniques used by
+version control systems, are already very well-developed.
+%
+Consequently, researchers have also explored various ``hybrid'' approaches,
+which incorporate holes into an otherwise text-based program editor.
+%
+For example, recent iterations of the Glasgow Haskell Compiler (GHC)
+\cite{GHCWIKI}, as well as the dependently typed proof assistants
+Idris~\cite{brady2013idris} and Agda~\cite{norell2009dependently}, support a
+workflow where the programmer places holes manually at locations in the program
+that remain under construction. Agda and Idris also allow the programmer to put
+terms inside holes, to indicate that the term may not yet be well-typed or
+otherwise finished.
+%
+Recent work on \emph{syntactic placeholders} takes some of the tedium out of
+this approach by providing low-level edit actions that cause these holes (called
+placeholders in their paper) to be inserted automatically \cite{Amorim2016}. 
+%
+This approximates the experience of a structure editor in common usage, while
+allowing the programmer to perform arbitrary text edits when necessary.
+%
+Expression holes also appear in the context of program synthesis, serving as
+placeholders in \emph{templates}~\cite{srivastava2013template} or
+\emph{sketches}~\cite{solar2009sketching} to be filled in by an expression
+synthesis engine.
+
+\begin{itemize}
+	%% \item Agda
+	%% \item Idris
+	%% \item GHC holes
+	%% \item Visual Studio (and others) support for edit-and-resume
+	\item Scratch lets you just skip over statement holes
+	\item papers that show up in a search for ``typed holes'' -- \url{https://scholar.google.com/scholar?hl=en&as_sdt=0%2C39&q=%22typed+holes%22&btnG=}
+\end{itemize}
+
+\parahead{Structure Editor Calculus}
+
+Most work on structure editors has focused on the user interfaces that they
+present. This is important work---presenting a fluid user interface involving
+exclusively structural edit actions is a non-trivial problem that has not yet
+been fully resolved, though recent studies have started to show productivity
+gains for keyboard-driven structure
+editors~\cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14}.
+
+We emphasize
+that our proposed contributions will be relevant no matter how holes come to be
+inserted into the program. % We will say more about all of these systems later.
+
+two main remaining questions (beyond scaling up the calculus) are how to run
+programs with type and expression holes.
+
 \begin{itemize}
 	\item Hazelnut paper
-	\item simply typed underdeterminism paper
-	\item work on partial evaluation and staging (incl. connections to modal logic -- ``modal analysis of staged computation'')
-	\item symbolic evaluation
-	\item full beta reduction
-	\item Agda
-	\item Idris
-	\item GHC holes
-	\item CMTT
+\end{itemize}
+
+\parahead{Gradual Typing}
+
+As already discussed in Track 1, type holes are closely related to the unknown
+types of gradual type theory \cite{Siek06a}. This has implications for our
+dynamic semantics as well---we need to be able to run programs where type holes
+have obscured a type inconsistency, \eg{}~\li{let f(g : }\ \li{) => g 3 in f 4}.
+This term is well-typed according to the static semantics of Track 1 because the
+argument \li{g} of \li{f} has an unknown type, and so it can be applied to
+\li{3}, but when \li{f} is applied to \li{4}, there is a problem because \li{4 3}
+is meaningless. There are two ways to address the problem. The first is to
+apply ML-style type inference to discover a more specific type schema for
+\li{g}, \ie{}~that it must actually be an arrow type, and then mark the argument
+\li{4} to \li{f} as a type inconsistency using a non-empty hole. We will
+investigate this in Track 1. The second approach is to adopt run-time cast
+insertions from research on gradual type theory. We will investigate this second
+approach in Track 2.
+
+\begin{itemize}
 	\item gradual typing (and dynamic typing)
-		\begin{itemize}
   \item siek and taha paper
@@
 \end{itemize}
-	\item DuctileJ stuff -- \url{https://homes.cs.washington.edu/~mernst/pubs/ductile-icse2011.pdf}
-	\item OLEG from McBride's thesis
-	\item Visual Studio (and others) support for edit-and-resume
-	\item Scratch lets you just skip over statement holes
+
+\parahead{Contextual Modal Type Theory}
+
+compared to the dynamics of type holes (which are based on Grad Typing), the
+dynamics of expression holes are less clear.
+
+Formally, expression holes in our proposed dynamics are closely related to the
+metavariables of contextual modal type theory (CMTT) \cite{Nanevski2008}.
+Metavariables in CMTT are associated with both a type and a typing context.
+Expression holes can similarly be assigned a most specific type and typing
+context. The environment tracking service can then be understood as being
+related to the concept of a \emph{metavariable closure} in CMTT, which consists
+of a metavariable paired with a substitution for all of the variables in the
+typing context associated with that metavariable. Indeed, this neatly explains
+the difference between our proposed dynamics and existing work on staging and
+partial evaluation---that existing work is rooted in modal type theory
+\cite{Davies:2001op} rather than the derivative contextual modal type theory
+(which, to date, has been studied far less extensively.) In other words,
+existing systems are focused on partial evaluation with respect to an input to a
+function, whereas holes can appear in context, \eg{}~directly within the body of
+a function. Recent work on the Beluga theorem prover is also based on CMTT,
+albeit for the purpose of proof refinement and contextual reasoning rather than
+program evaluation \cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
+McBride's OLEG is another system for reasoning contextually about metavariables
+\cite{DBLP:phd/ethos/McBride00}, and it is the basis of certain analagous
+features of Idris \cite{brady2013idris}.
+
+\begin{itemize}
+	\item CMTT
+\end{itemize}
+
+\parahead{Evaluation Order}
+
+\begin{itemize}
+	\item full beta reduction
 	\item prior work on confluence for the lambda calculus
+	\item ``Call-by-value is dual to call-by-name'' might be relevant? \url{http://homepages.inf.ed.ac.uk/wadler/papers/dual/dual.pdf}
+	\item mention how unspecified evaluation order is something that people do when talking about parallelism?
+\end{itemize}
+
+\parahead{Partial Evaluation and Symbolic Execution}
+
+\begin{itemize}
+	\item work on partial evaluation and staging (incl. connections to modal logic -- ``modal analysis of staged computation'')
+	\item symbolic evaluation
+	\item maybe also search for ``partial programs'' 
+\end{itemize}
+
+put Seidel here
+
+\parahead{Debugging}
+
+\begin{itemize}
 	\item work on debuggers that allow you to inspect environments
-		\begin{itemize}
   \item might be something relevant in the paper ``A Debugger for Standard ML'' 
@@
     \end{quote}
+   \item edit-and-resume in visual studio, others
 \end{itemize}
-	\item papers that show up in a search for ``typed holes'' -- \url{https://scholar.google.com/scholar?hl=en&as_sdt=0%2C39&q=%22typed+holes%22&btnG=}
-	\item maybe also search for ``partial programs'' 
-	\item roly pererra work on program slicing
-	\item should say something about how holes show up in program synthesis under various names (what?) 
+
+Visual Studio edit-and-continue \url{https://docs.microsoft.com/en-us/visualstudio/debugger/edit-and-continue}
+
+Hole-like constructs also appear in work on program slicing
+\cite{DBLP:conf/icfp/PereraACL12}, where holes are \rkc{dynamically inserted...}
+
+\parahead{Others?}
+
+\begin{itemize}
+	\item simply typed underdeterminism paper
+	\item DuctileJ stuff -- \url{https://homes.cs.washington.edu/~mernst/pubs/ductile-icse2011.pdf}
 	\item ``Achieving flexibility in direct-manipulation programming environments by relaxing the edit-time grammar'' -- \url{http://ieeexplore.ieee.org/document/1509511/}
-	\item ``Call-by-value is dual to call-by-name'' might be relevant? \url{http://homepages.inf.ed.ac.uk/wadler/papers/dual/dual.pdf}
-	\item mention how unspecified evaluation order is something that people do when talking about parallelism?
+	\item talk about closures in environmental semantics
 \end{itemize}

### 2018-04-11 calculus.tex
added citation to intro to sec3 and reorganized it slightly

@@
 \label{sec:calculus}
+\input{fig-syntax}
+\input{fig-bidirectional-typing}
 
 We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory.
-We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
-(see Sec.~\ref{sec:agda-mechanization} for some additional details, and the supplemental material for the full mechanization).
 
@@
 External expressions correspond to programs as entered by the programmer 
-(see Sec.~\ref{sec:implementation}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
+(see Sec.~\ref{sec:intro}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
 Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
-We distinguish the external and internal languages because 
-(1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and
+We take this approach, notably also taken in the specification of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and
 (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
@@
 In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
-
-\input{fig-syntax}
+We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}
+(see Sec.~\ref{sec:agda-mechanization}). We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).
 
@@
 \label{sec:external-statics}
-
-\input{fig-bidirectional-typing}
 

### 2018-04-11 calculus.tex
small edits to sec 3

@@
 As with the type system for the external language (above), 
-we specify expansion in the bidirectional style, via two mutually defined judgements.
+we specify expansion bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
 %
@@
 %We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
-We discuss hole contexts further below, including their role in the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$.
-
+We describe hole contexts, which serve as ``inputs'' to the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below. 
+%
 The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks~$\hexp$ against~$\htau$ and produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$.
 %
-The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is necessarily consistent with $\htau$.
-%
-% \rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
-% entirely standard, because of the hole context, so perhaps worth mentioning.}
+The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is consistent with $\htau$.
 %
@@
 Formally,
-this corresponds to the Sensibility property from the prior work \cite{popl-paper}. 
+this corresponds to the Sensibility property from the prior work \cite{popl-paper}.\todo{put statement in the appendix}{} 
 By composing this property with the Expandability, Typed Expansion and Progress properties from this paper, we establish a uniquely powerful 

### 2018-04-11 abstract.tex
three abstract pars
final(??) version of abstract

@@
 \begin{abstract}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% PAR 1: Live programming environments --- Problem statement
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
  to programs that are \emph{complete}, i.e. syntactically well-formed and free
-of static type errors, whereas 
+of type errors, whereas 
 %
 during much of the editing process, the program is not yet complete. As a result,  
-live programming services flicker in and out, 
-in many cases going out for substantial lengths of time.
+live feedback flickers in and out or goes stale, 
+in many cases for substantial lengths of time.
 %
@@
 % provide continuous, uninterrupted, and meaningful feedback.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% PAR 2: This paper's approach: New (dynamic) semantics for incomplete programs
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 This paper confronts this problem from type-theoretic first principles by developing a 
@@
  can be reported both directly and indirectly, via various editor services, to help the programmer decide how to fill a hole. 
-They also enable an ``edit-and-resume'' feature that 
+They also enable a ``fill-and-resume'' feature that 
 avoids the need to restart evaluation after edits that amount to hole filling. 
-Formally, the semantics combines features from gradual type theory (to handle
+Formally, the semantics draws from both gradual type theory (to handle
 type holes) and contextual modal type theory (which provides a
-logical foundation for hole closures) with novel formal structure necessary
-to continue evaluation past holes and to prove key metatheoretic properties. 
+logical foundation for hole closures), and develops additional technical machinery necessary 
+to continue evaluation past the various hole forms and to prove important metatheoretic properties.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% PAR 3: We have an implementation, and have proofs
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 % , including 
@@
 %
-We additionally describe a simple reference implementation, called \HazelnutLive{},
+We also describe a simple implementation, called \HazelnutLive{},
 that inserts holes as
 necessary during the editing process to guarantee that every edit state has some (possibly incomplete)
-type, based in part on the edit action calculus in prior work by \citet{popl-paper}.
+type, based in part on the \Hazelnut edit action calculus in prior work by \citet{popl-paper}.
 %

### 2018-04-12 calculus.tex
a bit more progress through Section 3; more to come later today

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% <<<<<<<<<<<<< Matt is here (latest pass over everything above is done -- 2018.04.11)
-
-The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}
-(so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). 
-Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation when expanding an unannotated external lambda based on the given type.
-The rules governing holes and the rules governing function applications and ascriptions, which both perform \emph{cast insertion}, are more interesting, so let us consider both cases in turn.
+The rules governing expansion of constants, variables and lambda expressions
+--- Rules {ESConst}, {ESVar}, {ESLam} and {EALam}
+--- and the corresponding type assignment rules
+--- Rules {TAConst}, {TAVar} and {TALam}
+--- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}.
+%
+Consequently, the corresponding cases of
+Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
+Theorem~\ref{thm:expansion-generality} are straightforward.
+%
+To support type assignment, all lambdas in the internal language are
+half-annotated---Rule {EALam} inserts the annotation when expanding an
+unannotated external lambda based on the given type.
+%
+Some rules perform \emph{cast insertion}, making them more interesting
+than the others; these rules include those for holes, and the those for
+function applications and ascriptions.  We consider each below.
 
 \subsubsection{Hole Expansion}\label{sec:hole-expansion} 
-Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$.
+Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
 %
+The hole name~$u$ on a hole closure identifies the external hole to which the hole closure corresponds.
+%
+%Note that
+While we assume each hole name to be unique in the external language, due to substitution, evaluated (internal) terms may consist of multiple hole closures with the same name.
+%
+For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}.
+%
+There, we number each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.
 
-The hole name, $u$, on a hole closure identifies the external hole that the hole closure corresponds to. 
-Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution. 
-For example, the result from Fig.~\ref{fig:grades-example} showed four closures for the hole named 1\todo{update with actual number (four?) and hole name later}. 
-There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation so we omit hole closure numbers from the core calculus.
+The hole expansion rules introduce hypotheses of the form~$\Dbinding{u}{\hGamma}{\htau}$ into the hole context~$\Delta$.
+%
+For each hole name~$u$, the hole context records a type~$\tau$, and a typing context~$\Gamma$.\footnote{
+We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.}
+%
+We borrow this hole context notation from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
+%
+Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
+%
+In the synthetic hole expansion rules~{ESEHole} and~{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the typing rules.
+%
+However, the first two premises of the expansion subsumption rule~{EASubsume} disallow the use of subsumption for holes in analytic position.
+%
+Instead, we employ separate analytic rules~{EAEHole} and~{EANEHole}, which each record the checked type~$\tau$ in the hole context.
+%
+Consequently, we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.
 
-The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. 
-The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$, for each hole name, $u$.\footnote{
-We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} 
-This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
-In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. 
-In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. 
-However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position. 
-Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. 
-This is again so that we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.
-
-Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$. 
-The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. 
-Initially, no evaluation has occurred, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
+Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
+%
+The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
+%
+Initially, when no evaluation has yet occurred, the hole expansion rules generate an initial environment that consists of the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
+%
 \begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
@@
 \noindent
-The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
+The type assignment rules for hole closures~{TAEHole} and {TANEHole} each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
+%
+Formally, we define this relation as follows:\todo{check that the definition in the Agda corresponds}
 \begin{defn}[Substitution Typing]
@@
 Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{}
-We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}. 
-Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT (see Sec.~\ref{sec:resumption}).
+%
+\Secref{sec:evaluation} defines how these closure environments evolve during evaluation.
+%
+Non-empty hole closures~$\dhole{d}{u}{\sigma}{}$ have no direct correspondance with a notion from CMTT (see Sec.~\ref{sec:resumption}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Matt is here (2018.04.12, ~8am MT)
 

### 2018-04-14 calculus.tex
sec 3: cast insertion

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Matt is here (2018.04.12, ~8am MT)
-
 \subsubsection{Cast Insertion}\label{sec:cast-insertion}
 %
-Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. 
-The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption.
-However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. 
-Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed.
-The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
-%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
+% USE A TOPIC SENTENCE SO THAT THE READER IS GUIDED A LITTLE MORE, e.g.,
+%
+Cast insertion mediates between the type systems of the external and
+internal languages, where the former permits subsumption, and the
+latter does not.
+%-----------------------------------------------------------------------------------------------
+%
+To see why this mediation is necessary, consider the following
+example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
+%
+Viewed as an external expression, this example synthesizes type
+$\tehole$, since the hole type annotation on variable~$x$ permits
+applying~$x$ as a function of type~$\tarr{\tehole}{\tehole}$, and base
+constant~$c$ may be checked against type~$\tehole$, by subsumption.
+%
+However, viewed as an internal expression, this example is not
+well-typed---the type assignment system defined in
+\Figref{fig:hasType} lacks subsumption.
+%
+Indeed, it would violate type safety if we could assign a type to this
+example in the internal language, because beta reduction of this
+example viewed as an internal expression would result in $c(c)$, which
+is clearly not well-typed.
+%
+The difficulty arises because leaving the argument type unknown also leaves unknown how
+the argument is being used (in this case, as a function).\footnote{In a system where type reconstruction is first used
+to try to fill in type holes, we could express a similar example by
+using $x$ at two or more different types, thereby causing type
+reconstruction to fail.
+%
+% On the other hand, if it is acceptable to arbitrarily choose one of
+%the possible types, and type reconstruction is complete, then type
+%holes will never appear in the internal language and the cast
+%insertion machinery described in this section can be omitted
+%entirely, leaving only the hole closure machinery described
+%previously.
 }
 \todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} 
-By our  interpretation of hole types as unknown types from gradual type theory, we can address the problem by performing cast insertion. 
+By our interpretation of hole types as unknown types from gradual type
+theory, we can address the problem by performing cast insertion.
 %
 
-The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of a consistent type $\htau_2$.%
-\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}
+The cast form in Hazelnut Live consists of $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
+%
+This form serves to ``box'' an expression of type $\htau_1$ for
+treatment as an expression of a consistent type $\htau_2$
+(Rule TACast in \Figref{fig:hasType}).
+%
+\footnote{
+In the earliest work on gradual type theory, the cast form only gave
+the target type~$\htau_2$ \cite{Siek06a}, but it simplifies matters
+to include the assigned type~$\htau_1$ in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.
+}
 
-Casts are inserted during the expansion of function applications and ascriptions. 
-The latter is more straightforward: 
-Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. 
-Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  
-Note that we included ascription only for the sake of exposition---it can be defined using application together with the half-annotated identity, $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$, so application expansion, discussed next, is more general.
+Expansion inserts casts for function applications and ascriptions.
+%
+The latter is more straightforward: Rule~{ESAsc}
+in \Figref{fig:expandSyn} inserts a cast from the assigned type to the
+ascribed type.
+%
+Theorem~\ref{thm:typed-expansion} inductively ensures that the two
+types are consistent.
+%
+We include ascription for exposition purposes---this form is derivable
+by using application together with the half-annotated identity, $e
+: \tau = \hap{(\halam{x}{\htau}{x})}{e}$; as such, application
+expansion, discussed below, is more general.
 
-Cast insertion by the application expansion rule, Rule~{ESAp}, requires a bit more care. 
-To understand the rule, consider the expansion for the example above:
-\[\hap{\dcasttwo{(\halam{x}{\tehole}{\underline{\hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}}})}{\tarr{\tehole}{\tehole}}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}
+Rule~{ESAp} expands function applications.
+%
+To understand the rule, consider the expansion of external
+expression~$\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$, the example
+discussed above:
+\[
+        \hap{\dcasttwo{
+        (\halam{x}{\tehole}{\underbrace{
+                \hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}
+                {\dcasttwo{c}{b}{\tehole}}
+                }_{\textrm{expansion of function body}~x(c)}
+        }
+        )}{\tarr{\tehole}{\tehole}}{
+           \tarr{\tehole}{\tehole}}
+           }
+           {\dcasttwo{c}{b}{\tehole}}
 \]
-Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. 
-A cast has been inserted on both the function expression, $x$, and the the argument, $c$.
+Consider the (indicated) function body,
+%
+where expansion inserts a cast on both the function expression~$x$ and its argument~$c$.
+%
+Together, these casts for~$x$ and~$c$ permit assigning a type to the
+function body according to the rules in \Figref{fig:hasType}, where we
+could not do so under the same context without casts.
+%
+We separately consider the expansions of~$x$ and of~$c$.
 
-The cast on $x$ allows us to treat the variable $x$, which is of type $\tehole$, as being of the matched arrow type $\tarr{\tehole}{\tehole}$, as described in Sec.~\ref{sec:external-statics}. 
-The first three premises of Rule~{ESAp} accomplish this by first synthesizing a type for the function expression, here $\tehole$, then determining the matched arrow type, $\tarr{\tehole}{\tehole}$, and then performing analytic expansion on the function expression with this matched arrow type.
-The resulting expansion will have some type $\tau_1'$ consistent with the matched arrow type. 
-In this case, because $x$ is of variable form, analytic expansion goes through subsumption so $\tau_1'$ is simply $\tehole$. 
+First, consider the function position of this application, here variable~$x$.
+%
+Without any cast, the type of variable~$x$ is hole type~$\tehole$;
+however, the inserted cast on~$x$ permits treating it as though it has
+arrow type $\tarr{\tehole}{\tehole}$, as described
+in \Secref{sec:external-statics}.
+%
+The first three premises of Rule~{ESAp} accomplish this
+%
+by first synthesizing a type for the function expression, here
+$\tehole$, then
+%
+by determining the matched arrow type~$\tarr{\tehole}{\tehole}$, and
+finally,
+%
+by performing analytic expansion on the function expression with this
+matched arrow type.
+%
+The resulting expansion has some type~$\tau_1'$ consistent with the matched arrow type.
+%
+In this case, because the subexpression~$x$ is a variable, analytic
+expansion goes through subsumption so that type~$\tau_1'$ is
+simply~$\tehole$.
+%
 The conclusion of the rule inserts the corresponding cast.
-We go through type synthesis then analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.
+%
+We go through type synthesis, \emph{then} analytic expansion, so that the hole
+context records the matched arrow type for holes in function position,
+rather than the type~$\tehole$ for all such holes, as would be the case
+in a variant of this rule using synthetic expansion for the function
+expression\todo{put this in the appendix?}{}.
 
-The conclusion of Rule~{ESAp} inserts the cast on the argument's expansion, from the type it is assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type of the matched arrow type of the function expression, here $\tehole$.
+Next, consider the application's argument, here constant~$c$.
+%
+The conclusion of Rule~{ESAp} inserts the cast on the argument's
+expansion, ufrom the type it is assigned by the final premise of the
+rule~(type~$b$ here), to the argument type of the matched arrow type
+of the function expression~(type~$\tehole$ here).
 
-Observe that together, these casts allow us assign a type to the function body according to the rules in Fig.~\ref{fig:hasType}, where we could not do so under the same context without casts.
+The the example's second, outermost application uses through the same
+application expansion rule.
+%
+In this case, the cast on the function is the identity cast for
+$\tarr{\tehole}{\tehole}$.
+%
+For simplicity, we do not attempt to avoid the insertion of identity
+casts in the core calculus; these will simply never fail during
+evaluation.
+%
+However, it is safe to eliminate such identity casts during expansion,
+and some formal accounts of gradual typing do so by defining three
+application rules, including the original account of \cite{Siek06a}).
 
-The outer application in the example above goes through the same rule. 
-In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$. 
-We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), 
-but it is safe to eliminate these during expansion (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Matt is here (2018.04.14, ~6:15am MT)
 

### 2018-04-14 calculus.tex
move related work discussion to related work

@@
 
-To recap, the result of expansion is a well-typed internal expression with hole closures and casts, 
-where the former corresponds to metavariable closures from CMTT \cite{Nanevski2008}, 
-and the latter corresponds to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}. 
-However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.
+Figures~\ref{fig:isGround}-\ref{fig:step} give dynamic semantics for \HazelnutLive,
+%
+resulting in a system capable of running incomplete programs without aborting when a hole is encountered.
+%
+We equip this semantics with a meaningful notion of type safety (involving both preservation and progress).
+%
+We also establish that the standard notion of type safety falls out when running complete terms.
 
-The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.
-The paper therefore proved only a subject reduction property (which is closely related to type preservation). 
-This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. 
-In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables.
-\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion%
-\footnote{Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property.
-These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.},
-which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
-but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
-
-Furthermore, we need to integrate casts into the dynamic semantics. 
-Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. 
-The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms.
-The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. 
-Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}\todo{where?}{}, is for cast failure to instead insert a membrane around the dynamic type error, 
-much like a non-empty hole serves as a membrane around a static type error, 
-allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.
-
-The dynamic semantics for \HazelnutLive specified in Figures~\ref{fig:isGround}-\ref{fig:step} addresses the difficulties just outlined, 
-resulting in a system capable of running incomplete programs without aborting when a hole is encountered. 
-This semantics is equipped with a meaningful notion of type safety (involving both preservation and progress). 
-We also establish that the standard notion of type safety falls out when running complete terms.
+%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- 6:40am MT)
 

### 2018-04-14 related-work.tex
move related work discussion to related work

@@
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% 2018.04.14
+%
+% MATT MOVED THIS MATERIAL HERE FROM calculus.tex
+%
+% We should discuss this stuff, but IMO, it should not be necessary to
+% read or appreciate these connections to understand calculus.tex on a
+% first read.  These details are very interesting, but shouldn't be in
+% the main technical section, IMO.
+%
+\subsection{MOVED FROM: Dynamic Semantics}
+
+To recap, the result of expansion is a well-typed internal expression with hole closures and casts, 
+where the former corresponds to metavariable closures from CMTT \cite{Nanevski2008}, 
+and the latter corresponds to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}. 
+However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.
+%
+For further discussion, see \Secref{sec:related}
+
+The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.
+The paper therefore proved only a subject reduction property (which is closely related to type preservation). 
+This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. 
+In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables.
+\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion%
+\footnote{Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property.
+These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.},
+which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
+but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
+
+Furthermore, we need to integrate casts into the dynamic semantics. 
+Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. 
+The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms.
+The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. 
+Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}\todo{where?}{}, is for cast failure to instead insert a membrane around the dynamic type error, 
+much like a non-empty hole serves as a membrane around a static type error, 
+allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.

### 2018-04-14 calculus.tex
another page down

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Matt is here (2018.04.14, ~6:15am MT)
-
 \subsection{Dynamic Semantics}
@@
 
-%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- 6:40am MT)
-
-The dynamic semantics is specified as a ``small-step'' transition system. 
-The cast-related machinery is based closely on the cast calculus from the ``refined'' account of the gradually typed lambda calculus by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be theoretically well-behaved.
-In particular, Fig.~\ref{fig:isGround} defines the judgement $\isGround{\htau}$, which distinguishes the base type, $b$, and the least specific arrow type, $\tarr{\tehole}{\tehole}$, as \emph{ground types}, which play a role in simplifying the treatment of function casts.
-Fig.~\ref{fig:isFinal} defines the judgement $\isFinal{d}$, which distinguishes the final forms of the transition system. 
-There are two classes of final forms: (possibly-)boxed values and indeterminate forms.%
-\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention. 
-Our judgemental formulation is more faithful to the mechanization and cleaner for our purposes, because we are distinguishing several classes of final forms.}
+The dynamic semantics consists of a ``small-step'' transition system,
+%
+where cast-related machinery closely follows the cast calculus from
+the ``refined'' account of the gradually typed lambda calculus
+by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be
+theoretically well-behaved.
+%
+In particular, \Figref{fig:isGround} defines the judgement
+$\isGround{\htau}$, which distinguishes the base type~$b$ and the
+least specific arrow type~$\tarr{\tehole}{\tehole}$, as \emph{ground
+types}; this judgement helps simplify the treatment of function casts.
+%
+\Figref{fig:isFinal} defines the judgement $\isFinal{d}$, which
+distinguishes the final forms of the transition system.
+%
+There are two classes of final forms: (possibly-)boxed values and
+indeterminate forms.%
+%
+\footnote{
+        In most accounts of the cast calculus, values and ground types
+        are distinguished with separate grammars together with an
+        implicit identification convention.
+        %
+        Our judgemental formulation is more faithful to the mechanization and
+        cleaner for our purposes, because we are distinguishing several
+        classes of final forms.
+}
 
@@
 
-The judgement $\isBoxedValue{d}$ distinguishes (possibly-)boxed values, which correspond to the values from the cast calculus and include the classic values from the lambda calculus, distinguished by $\isValue{d}$, as well as two cast forms:
-casts between function types (except identity casts) 
-and casts from a ground type to the hole type. In both cases, the cast must be on a boxed value.
-
-The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, 
-so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:resumption}). 
-The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). 
-The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}
+The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as
+either ordinary values~(Rule~{BVVal}), or one of two cast forms: casts
+between function types (except identity casts) and casts from a ground
+type to the hole type. In each case, the cast operates on a boxed
+value.
+%
+%, which correspond to the values from the cast calculus and include
+%the classic values from the lambda calculus, distingui%shed by
+%$\isValue{d}$,
+%
+The judgement $\isIndet{d}$ defines \emph{indeterminate} forms, so
+named because they arise from the presence of expression holes and
+failed casts in a program, and so, conceptually, their ultimate value
+awaits programmer action (see Sec.~\ref{sec:resumption}).
+%
+The first two rules specify that \emph{empty} hole closures are always
+indeterminate, and that \emph{non}-empty hole closures are indeterminate when
+they consist of a \emph{final} inner expression.
+%
+Below, we discuss the the remaining indeterminate forms simultaneously
+with the corresponding transition rules, also discussed
+below.\todo{relationship to weak head normal forms}
 
@@
 The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}.
-Top-level transitions, i.e. \emph{steps}, $\stepsToD{}{d}{d'}$, are governed by a single rule, Rule {Step} in Fig.~\ref{fig:step}, which 
+%
+Top-level transitions are \emph{steps}~$\stepsToD{}{d}{d'}$, governed by Rule~{Step} in \Figref{fig:step}, which
+%
 (1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$;
-(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in Fig.~\ref{fig:instruction-transitions}; 
-and (3) places $d_0'$ back in the selected position, which is marked%
-\footnote{In the literature, the form $\evalhole$ in the grammar of evaluation contexts is referred to as the \emph{hole}, but this hole is a technical device entirely orthogonal to the holes of this paper, so we use the term ``mark'' instead.}
-in the evaluation context by $\evalhole$, to obtain $d'$.
+%
+(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in \Figref{fig:instruction-transitions};
+%
+and (3) places $d_0'$ back in the selected position, indicated in
+the evaluation context by the \emph{mark}~$\evalhole$, to obtain $d'$.%
+\footnote{
+        We say ``mark'' here to avoid confusion with the (orthogonal) ``holes'' of \HazelnutLive.
+        %
+        %the form $\evalhole$ in the grammar of
+        %evaluation contexts is referred to as the \emph{hole}, but
+        %this hole is a technical device entirely orthogonal to the
+        %holes of this paper, so we use the term ``mark'' instead.
+        }
+%
+
+\matt{Next paragraph can be dropped entirely for space}
+%
 This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
@@
 \todo{Put the search rules in the appendix?}{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- ~7am MT)
 

### 2018-04-14 calculus.tex
fix typo
more copy editing for Sec 3 --- through 3.3.3

@@
 
-The the example's second, outermost application uses through the same
+The example's second, outermost application uses through the same
 application expansion rule.
@@
 \subsubsection{Application and Substitution} 
-Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the standard beta reduction transition. 
-The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} can be included to specify an eager, left-to-right evaluation strategy, 
-or excluded to leave the evaluation order unspecified. 
-We do the latter in our metatheory, both for the sake of generality and for reasons we return to in Sec.~\ref{sec:resumption}.
+%
+Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the
+standard beta reduction transition.
+%
+The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in
+Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} may be \emph{included}
+to specify an eager, left-to-right evaluation strategy, or \emph{excluded} to
+leave the evaluation order unspecified.
+%
+In our metatheory, we use the latter version of the rules, both for
+the sake of generality, and to support fill-and-resume~(\Secref{sec:resumption}).
 
-Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. 
+A substitution~$[d/x]d'$ operates in the standard capture-avoiding manner~\cite{pfpl}. 
+%
 The only cases of special interest arise when substitution reaches a hole closure:
@@
 \]
-In both cases, we write $[d/x]\sigma$ to perform substitution on each expression in the hole environment, $\sigma$. 
-For example, $\stepsToD{}{\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}{\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. 
-Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g.
-because a function with a hole closure in its body is applied multiple times.%
+In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$. 
+%
+For example, $\stepsToD{}
+    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
+    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. 
+%
+Reduction may duplicate hole closures, just as with any other substituted term;
+%
+consequently, the environments of different closures for the same hole name may differ, 
+e.g., when an reduction applies a function with a hole closure body multiple times.%
 \todo{show example here or refer back to sec 2?}{} 
-Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}\todo{where?}{}).
+%
+Hole closures may also appear within the environments of other hole
+closures, giving rise to the closure paths described in
+Sec.~\ref{sec:paths}\todo{where?}{}.
 
-The ITBeta rule is not the only rule we need to handle function application, because lambdas are not the only final forms of arrow type. There are two other classes of possibilities.
+The ITBeta rule is not the only rule we need to handle function
+application, because lambdas merely one (final) form of arrow type.
+%
+Two other situations may also arise.
 
-The expression in function position might be a cast between arrow types, in which case we apply the arrow cast conversion rule, Rule {ITApCast}, to rewrite the application form, obtaining an equivalent application where the expression under the function cast, $d_1$, is exposed.
-We know from inverting the typing rules that $d_1$ is of type $\tarr{\htau_1}{\htau_2}$ and that $d_2$ is of type $\htau_1'$ where $\tconsistent{\htau_1}{\htau_1'}$, so to maintain type safety, we must place a cast on $d_2$ from  $\htau_1'$ to $\htau_1$.
-The result of this application will be of type $\htau_2$, but the original cast promised that the result would be of consistent type $\htau_2'$, so we also need a cast on the result from $\htau_2$ to $\htau_2'$.
+First, the expression in function position might be a cast between
+arrow types, in which case we apply the arrow cast conversion rule,
+Rule {ITApCast}, to rewrite the application form, obtaining an
+equivalent application where the expression under the function
+cast~$d_1$ is exposed.
+%
+We know from inverting the typing rules that~$d_1$ has type
+$\tarr{\htau_1}{\htau_2}$, and that~$d_2$ has type~$\htau_1'$, where
+$\tconsistent{\htau_1}{\htau_1'}$; consequently, we maintain type
+safety by placing a cast on~$d_2$ from~$\htau_1'$ to~$\htau_1$.
+%
+The result of this application will has type $\htau_2$, but the
+original cast promised that the result would have consistent type
+$\htau_2'$, so we also need a cast on the result from $\htau_2$ to
+$\htau_2'$.
 
-The other possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$. 
-In this case, no further progress can be made, so the function application as a whole is also deemed indeterminate by Rule {IAp} in Fig.~\ref{fig:isFinal}. 
-Note that we are careful to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.%
+Second, the expression in function position may be indeterminate,
+where arrow cast conversion is not applicable,
+e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$.
+%
+In this case, the application is indeterminate (Rule~{IAp}
+in \Figref{fig:isFinal}), and the application's reduction reduces no
+further.
+
+The rules take care to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.%
 \todo{prove this as stated from Ian's disjointness proofs}{}%
@@
 \subsubsection{Casts}
-Rule {ITCastId} drops identity casts. The remaining instruction transition rules assign meaning to non-trivial casts. 
-As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term cast to hole type is statically obscure, 
-so we need to wait until it is actually used at some other type, via a cast away from hole type, to be able to determine whether there is a problem. 
-Rules {ITCastSucceed} and {ITCastFail} handle this situation when the two types involved are ground types (Fig.~\ref{fig:isGround}). 
-If the two ground types are equal, then the cast succeeds and the cast can be dropped. 
-If they are not equal, then the cast fails and the failed cast form, $\dcastfail{\dexp}{\htau_1}{\htau_2}$, arises. 
-Rule {TAFailedCast} specifies that a failed cast is well-typed exactly when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type disequal to $\tau_1$. 
-Rule {IFailedCast} specifies that a failed cast operates as an indeterminate form once $d$ is final. 
-This allows evaluation to proceed after cast failure, much as with hole closures.
+Rule {ITCastId} drops identity casts. The remaining instruction
+transition rules assign meaning to non-trivial casts.
+%
+As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term
+cast to hole type is statically obscure,
+%
+so we must wait for \emph{use} of the term at some other type, via a
+cast away from hole type, to detect dynamic type errors.
+%
+Rules {ITCastSucceed} and {ITCastFail} handle this situation when the
+two types involved are ground types (Fig.~\ref{fig:isGround}).
+%
+If the two ground types are equal, then the cast succeeds and the cast
+may be dropped.
+%
+If they are not equal, then the cast fails and the failed cast form,
+$\dcastfail{\dexp}{\htau_1}{\htau_2}$, arises.
+%
+Rule {TAFailedCast} specifies that a failed cast is well-typed exactly
+when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type
+disequal to $\tau_1$.
+%
+Rule {IFailedCast} specifies that a failed cast operates as an
+indeterminate form once $d$ is final.
+%
+This rule allows evaluation to proceed after cast failure, much as
+with hole closures.
 
-The two rules just described operate at ground type. 
-The two remaining instruction transition rules, Rule {ITGround} and {ITExpand}, operate as technical devices that allow us to restrict our interest to ground types exclusively by inserting intermediate casts from non-ground type to  a consistent ground type, and \emph{vice versa}.
-In this core calculus, the only non-ground types are the arrow types, so the grounding judgement $\groundmatch{\tau_1}{\htau_2}$ defined in Fig.~\ref{fig:groundmatch}, produces the ground arrow type, $\tarr{\tehole}{\tehole}$.
+Rules {TAFailedCast} and {IFailedCast} each operate at ground type. 
+%
+The two remaining instruction transition rules, Rule~{ITGround}
+and~{ITExpand}, insert intermediate casts from non-ground type to a
+consistent ground type, and \emph{vice versa}.
+%
+These rules serve as technical devices, permitting us to restrict our
+interest exclusively to ground types.
+%
+Here, the only non-ground types are the arrow types, so the grounding
+judgement~$\groundmatch{\tau_1}{\htau_2}$ (\Figref{fig:groundmatch}),
+produces the ground arrow type~$\tarr{\tehole}{\tehole}$.
+%
 More generally, this judgement is governed by the following invariant:
@@
 
-In all other cases, casts are either boxed values or indeterminate according to the remaining rules in Fig.~\ref{fig:isFinal}. 
-Note in particular Rule {ICastHoleGround}, which handles casts from hole to ground type that are not of the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$. 
-In the cast calculus, there are no such irreducible terms because the only canonical form of type $\tehole$ is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.
+In all other cases, casts either consist of boxed values or
+indeterminate terms according to the remaining rules
+in \Figref{fig:isFinal}.
+%
+In particular, Rule {ICastHoleGround} handles casts that are not of
+the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$, from hole to
+ground type.
+%
+In the cast calculus, since the only canonical form of type $\tehole$
+is $\dcasttwo{\dexp}{\htau_1}{\tehole}$, there exist no such
+irreducible terms.
 
 \subsubsection{Type Safety} 
-The purpose of establishing type safety is to ensure that the static and dynamic semantics of a language cohere. 
-We follow the approach developed by \citet{wright94:_type_soundness}, now standard \cite{pfpl}, which distinguishes two type safety properties, preservation and progress. 
-In order to allow for the evaluation of incomplete programs, we establish these properties for terms typed under arbitrary hole context $\Delta$. 
-We can assume $\hGamma$ is empty (in order to run programs with free variables, the system can treat them as empty holes with a corresponding name).
+%
+Type safety establishes that the static and dynamic semantics of a
+language cohere.
+%
+We follow the approach developed by \citet{wright94:_type_soundness},
+now standard \cite{pfpl}, which distinguishes two type safety
+properties, preservation and progress.
+%
+To permit the evaluation of incomplete programs, we establish these
+properties for terms typed under arbitrary hole context $\Delta$.
+%
+We assume an empty context for~$\hGamma$; to run open programs, the
+system may treat free variables as empty holes with a corresponding
+name.
 
-The preservation theorem establishes that type assignment is preserved by transitions, i.e. that the type of an expression accurately predicts the type of the result of evaluating that expression.
+The preservation theorem establishes that transitions perserve type
+assignment, i.e. that the type of an expression accurately predicts
+the type of the result of reducing that expression.
 
@@
 \noindent
-The proof relies on an analagous preservation lemma for instruction transitions and a standard substitution lemma (stated in Appendix~\ref{sec:additional-defns}). 
-Hole closures can disappear during evaluation, so we rely on weakening of $\Delta$.
+%
+The proof relies on an analagous preservation lemma for instruction
+transitions and a standard substitution lemma (stated in
+Appendix~\ref{sec:additional-defns}).
+%
+Hole closures can disappear during evaluation, so we rely on weakening
+of $\Delta$.
 
-The progress theorem establishes that the dynamic semantics accounts for every well-typed term, i.e. that we have not forgotten some necessary rules or premises.
+The progress theorem establishes that the dynamic semantics accounts
+for every well-typed term, i.e. that we have not forgotten some
+necessary rules or premises.
+%
 \begin{thm}[Progress]
@@
 \noindent
-The key to being able to establish the progress theorem under a non-empty hole context lies in the fact that we have explicitly accounted for indeterminate forms, i.e.
-those rooted at either a hole closure or a failed cast.
-The proof relies on canonical forms lemmas stated in Appendix~\ref{sec:additional-defns}.
+The key to establishing the progress theorem under a non-empty hole
+context consists of explicitly accounting for indeterminate forms,
+i.e. those rooted at either a hole closure or a failed cast.
+%
+The proof relies on canonical forms lemmas stated in
+Appendix~\ref{sec:additional-defns}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- ~noon MT)
 
 \subsubsection{Complete Programs} 
-Although our focus in this paper is on running incomplete programs, it is helpful to know that the necessary machinery does not interfere with reasoning about the behavior of complete programs, i.e.
-those with no holes.
+%
+Although this paper focuses running incomplete programs, it helps
+to know that its machinery does not interfere
+with reasoning about the behavior of complete programs, i.e. those
+with no holes.
+%
 Appendix~\ref{sec:additional-defns} defines complete programs with the judgements $\isComplete{e}$ and $\isComplete{d}$. 

### 2018-04-14 calculus.tex
done with pass over Section 3

@@
 
-\matt{Next paragraph can be dropped entirely for space}
-%
-This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
-Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
-It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
-\todo{Put the search rules in the appendix?}{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- ~7am MT)
+%% \matt{Next paragraph can be dropped entirely for space}
+%% %
+%% This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing.
+%% Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
+%% It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
+%% \todo{Put the search rules in the appendix?}{}
 
@@
 Rule {ITApCast}, to rewrite the application form, obtaining an
-equivalent application where the expression under the function
-cast~$d_1$ is exposed.
+equivalent application where the expression~$d_1$ under the function
+cast is exposed.
 %
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- ~noon MT)
-
 \subsubsection{Complete Programs} 
 %
-Although this paper focuses running incomplete programs, it helps
+Although this paper focuses running \emph{incomplete} programs, it helps
 to know that its machinery does not interfere
-with reasoning about the behavior of complete programs, i.e. those
+with reasoning about the behavior of \emph{complete} programs, i.e. those
 with no holes.
 %
-Appendix~\ref{sec:additional-defns} defines complete programs with the judgements $\isComplete{e}$ and $\isComplete{d}$. 
+Appendix~\ref{sec:additional-defns} defines complete programs via judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
+%
 Failed casts cannot appear in complete internal expressions. 
+%
 The following theorem establishes that expansion preserves program completeness.
@@
 
-The following progress theorem establishes that only classic values result from evaluating a complete program. There are no boxed values or indeterminate forms. 
+The following progress theorem establishes that evaluating a complete
+program always results in classic values, not boxed values nor
+indeterminate forms.
+%
 \begin{thm}[Complete Progress]
@@
 The supplemental material includes our mechanization of the semantics and metatheory of \HazelnutLive, 
-including proofs of all of the theorems stated above and all necessary lemmas\todo{make this so ian!}{}. We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis} 
+including proofs of all of the theorems stated above and all necessary lemmas\todo{make this so ian!}{}. 
+%
+We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis} 
 (as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
+%
 Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation. 
-Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf.
-Sec.~\ref{sec:intro}).
+%
+Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
+%
 The basic approach is standard: we model judgements as 
 inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite somtehing?}{}. 
+%
 We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
@@
 
-The supplemental material also includes a browser-based implementation of \HazelnutLive. 
-The implementation is a functional reactive program \cite{DBLP:conf/pldi/CzaplickiC13} 
-written using the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
+The supplemental material includes a browser-based implementation
+of \HazelnutLive, which consists of a functional reactive program
+%
+% Functional Reactive Animation (Hudak 1997) is the canonical FRP cite, not Elm, if you want one; we don't need one, I think. --Matt
+%
+%\cite{DBLP:conf/pldi/CzaplickiC13} 
+%
+written with the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
 together with the OCaml \lismall{React} library \cite{OcamlReact} 
 and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}. 
-Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{} provides a screenshot of the user interface, which implements all of the live programming features described in Sec.~\ref{sec:examples}, but for the more austere language of this section.
+%
+Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{} provides a screenshot of the user interface, 
+which implements the core calculus of this section, 
+along with the corresponding live programming features from \Secref{sec:examples}.
 
-The editor component of the implementation is derived from \Hazelnut. 
-It exposes a language of structured edit actions (summarized in the sidebar on the left of Fig.~\ref{fig:screenshot}) that insert holes automatically to guarantee that every edit state has some (possibly incomplete) type.
-Formally,
-this corresponds to the Sensibility property from the prior work \cite{popl-paper}.\todo{put statement in the appendix}{} 
-By composing this property with the Expandability, Typed Expansion and Progress properties from this paper, we establish a uniquely powerful 
-invariant: that every edit state is both statically and dynamically meaningful. 
-In other words, live feedback does not ``flicker in and out'' or reflect past edit states. 
-The primary purpose of the included implementation is to serve as a proof-of-concept of this invariant, and to be of use to researchers studying the calculus as presented in this section.
+To facilitate interactively editing terms, the implementation employs
+the editor semantics of~\Hazelnut, exposing a language of structured
+edit actions (summarized in the sidebar on the left of
+Fig.~\ref{fig:screenshot})
+%
+As with prior work, this editor semantics enjoys
+the \Property{Sensibility} property by automatically inserting holes
+to guarantee that every edit state has some (possibly incomplete) type
+.\todo{put statement in the appendix}{}
+
+By composing this meta-theoretic property with
+the \Property{Expandability}, \Property{Typed Expansion}
+and \Property{Progress} properties from this paper, we establish a
+uniquely powerful invariant: \emph{Every} edit state has a \emph{static
+meaning} (a la~\Hazelnut) and a \emph{dynamic meaning} (a la~\HazelnutLive).
+%
+In other words, live feedback never ``flickers in and out'' nor reflects ``stale'' edit states.
+%
+The current implementation serves as a proof-of-concept of this invariant, 
+by providing a working prototype for this section's definitions.
+
+%, and provides to be of use to researchers studying the calculus as presented in this section.
 % Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.
 
-We are also separately integrating the mechanisms developed in this paper into the \Hazel programming environment being developed by \citet{HazelnutSNAPL} with the goal of realizing a full-scale hole-driven development experience rooted in the type theoretic foundations established by \Hazelnut.
-Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper.
-For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
-We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
-\todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
-The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}
+%% \matt{I'd drop everything else in this sub-section; 
+%% it leaves us open to attack from a hostile reviewer 
+%% who is cranky that the implementation is not ``complete'' yet; also, we'd save the space}
+
+%% Separately, we are integrating the mechanisms developed in this paper
+%% into the \Hazel programming environment being developed
+%% by \citet{HazelnutSNAPL} with the goal of realizing a full-scale
+%% hole-driven development experience rooted in the type theoretic
+%% foundations established by \Hazelnut.
+%% %
+%% Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper.
+%% For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
+%% We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
+%% \todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
+%% The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}
+
+%% Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}
 

### 2018-04-15 calculus.tex
minor: typo

@@
 %
-The result of this application will has type $\htau_2$, but the
+The result of this application has type $\htau_2$, but the
 original cast promised that the result would have consistent type

### 2018-04-15 calculus.tex
pass over Section 4; copy edits and reorganization

@@
 In any case,
-When the programmer performs one or more edit actions to fill in a hole in the program, a new result must be computed.
-Na\"ively, the system would need to compute the result ``from scratch'' on each such edit.
-A more efficient approach supported by the structure of \HazelnutLive is to resume evaluation from where it left off after an edit that amounts to hole filling.
-We call this feature \emph{fill-and-resume}, by analogy to the ``edit-and-resume'' features available in some systems, e.g.
+
+When the programmer interactively fills a program hole, that program's
+reduction and result are generally each affected, and the system
+should respond by (quickly) updating the program's output.
+%
+Na\"ively, this response requires recomputing the program output
+``from scratch'' after each such edit.
+%
+However, by virtue of its design principles (based on
+CMTT), \HazelnutLive offers a more efficient approach: To \emph{resume
+evaluation} of previously indeterminate expressions, which are
+(generally) more determined after the programmer fills a hole than
+beforehand.
+%
+We call this feature \emph{fill-and-resume}, by analogy to the
+``edit-and-resume'' features available in some systems today, such as
 Visual Studio (see below for a comparison)\todo{cite and compare}{}.
 
-\todo{fix definition of instantiation in the agda}The key is to interpret the hole environments as delayed substitutions, and indeed this is exactly the interpretation suggested for closures in contextual modal type theory by \citet{Nanevski2008}.
-We can derive the hole filling operation, $\instantiate{d}{u}{d'}$, defined in Fig.~\ref{fig:substitution}, from the contextual substitution operation of CMTT.
-Notice that unlike standard substitution, which is capture-avoiding, hole filling imposes no condition on the binder when passing into the body of a lambda expression---the expression that fills a hole can, of course, refer to variables in scope where the hole appears.
-When hole filling encounters an empty closure for the hole being instantiated, $\instantiate{d}{u}{\dehole{u}{\sigma}{}}$, the result is $[\instantiate{d}{u}{\sigma}]d$, i.e.
-the delayed substitution is applied to the fill expression, $d$, after first recursively filling any instances of hole $u$ that may appear in $\sigma$.
-Hole filling for non-empty closures is analagous (the enveloped expression can be discarded).
-Note that this is the reason why we cannot interpret a non-empty hole as an empty hole of arrow type applied to the enveloped expression---the hole filling operation would not operate as expected under this interpretation.
+%%%%%%%%% New par
+Formally, 
+%
+we interpret the hole environments as \emph{delayed substitutions},
+the same interpretation suggested for closures in contextual modal
+type theory (CMTT) by \citet{Nanevski2008}.
+%
+%In practice, it may be useful to cache results from several previous
+%expansions, e.g., by employing off-the-shelf programming language
+%abstractions for incremental computation~\cite{Hammer14,Hammer15}.
+%
+We define hole filling in terms of the expanded (internal) program,
+and define filling a hole in an external language term as filling the
+corresponding hole in its expansion.
+
+%%%%%%%%% New par
+\Figref{fig:substitution} defines the hole filling operation~$\instantiate{d}{u}{d'}$,
+mirroring the contextual substitution operation of~CMTT.
+%
+Unlike usual notions of capture-avoiding substitution, 
+hole filling imposes no condition on the binder when passing into the
+body of a lambda expression---the expression that fills a hole can, of
+course, refer to variables in scope where the hole appears.
+%
+When hole filling encounters an empty closure for the hole being
+instantiated, $\instantiate{d}{u}{\dehole{u}{\sigma}{}}$, the result
+is $[\instantiate{d}{u}{\sigma}]d$.
+%
+That is, we apply the delayed substitution to the fill expression~$d$
+after first recursively filling any instances of hole~$u$ in~$\sigma$.
+%
+Hole filling for non-empty closures is analogous, where it discards
+the previously-enveloped expression.
+%
+%
+% \matt{Will any reader actually wonder this? Does this thought connect to any other statement elsewhere in the paper?}
+%
+%This case shows why we cannot interpret a non-empty hole as an empty
+%hole of arrow type applied to the enveloped expression---the hole
+%filling operation would not operate as expected under this
+%interpretation.
 
@@
 
+%%%%%%%%% New par
 The following theorem characterizes the static behavior of hole filling.
@@
   then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
-\end{thm}
 \begin{proof}
-We prove a stronger version of the theorem which accounts for both terms and finite substitutions (see \cite{Nanevski2008}).
-In each case, we proceed by rule induction on the first assumption, appealing to the finite substitution lemma in the corresponding cases.
+We prove a stronger version of the theorem which accounts for both
+terms and finite substitutions (see \cite{Nanevski2008}).
+%
+Each case proceeds by rule induction on the first assumption, applying
+the finite substitution lemma.
 \end{proof}
-
-For the fill-and-resume feature to be reasonable, we need the following commutativity property, where we write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive, transitive closure of  stepping (see Fig.~\ref{fig:multi-step} in Appendix~\ref{sec:additional-defns}).
-In words, if there is some sequence of steps that go from $d_1$ to $d_2$, then you can fill the hole either at the beginning or at the end of that step sequence because if you do the former, you can take some sequence of steps to the latter.
+\end{thm}
+
+Dynamically, the fill-and-resume enjoys an
+important \emph{commutativity} property: If there is some sequence of
+steps that go from $d_1$ to $d_2$, then one can fill a hole in these
+terms at \emph{either} the beginning or at the end of that step
+sequence.
+%
+We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
+transitive closure of stepping (see \Figref{fig:multi-step} in
+Appendix~\ref{sec:additional-defns}).
+%
 \begin{thm}[Commutativity]
@@
 \end{thm}
-Note that this property follows because the specification in Sec.~\ref{sec:calculus} left evaluation order unspecified (i.e.
-formally, we omitted the bracketed premises).
-It is not the case in general that resuming from $\instantiate{\dexp'}{u}{d_2}$ will cause sub-expressions to be evaluated in the same order as they would have been if we had restarted from $\instantiate{\dexp'}{u}{d_1}$ and selected sub-expressions in, for example, an eager, left-to-right manner, because the hole that is being filled may have been ``skipped past''.
-In other words, this simple fill-and-resume protocol only works for languages where evaluation order ``does not matter''.
-There are various ways to encode this intuition.
-One way to do so is by establishing a confluence property (which is closely related to the Church-Rosser property \cite{church1936some}).
-The most general confluence property does not hold for the dynamic semantics in Sec.~\ref{sec:calculus} for the usual reason: we do not reduce under binders (\citet{DBLP:conf/birthday/BlancLM05} discuss the standard counterexample).
-We could recover confluence by specifying reduction under binders, either generally or in a more restricted form where only closed sub-expressions are reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit}, but reduction under binders is at odds with the standard approaches used in implementing programming languages \cite{DBLP:conf/birthday/BlancLM05}.
+%
+Critically, this property relies on the more permissive version of
+reduction from \Secref{sec:calculus}, where evaluation order is
+unspecified (i.e. formally, we omitted the bracketed premises):
+%
+In general, resuming from $\instantiate{\dexp'}{u}{d_2}$ will not
+reduce sub-expressions in the same order as a ``fresh'' left-to-right
+reduction sequence starting from $\instantiate{\dexp'}{u}{d_1}$.
+%
+In other words, this notion of fill-and-resume only works for
+languages where evaluation order always results in \emph{confluent}
+reductions; viz., those that \emph{eventually} yield the same
+resulting term.
+
+%% Everyone knows what confluence means, or they should.  We don't need to cite Church :)
+%%
+%There are various ways to encode this intuition. One way to do so is
+%by establishing a confluence property (which is closely related to
+%the Church-Rosser property \cite{church1936some}).
+
+The most general confluence property does not hold for the dynamic
+semantics in Sec.~\ref{sec:calculus} for the usual reason: We do not
+reduce under binders (\citet{DBLP:conf/birthday/BlancLM05} discuss the
+standard counterexample).
+%
+We could recover confluence by specifying reduction under binders,
+either generally or in a more restricted form where only closed
+sub-expressions are
+reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit},
+but reduction under binders is at odds with the standard approaches
+used in implementing programming
+languages \cite{DBLP:conf/birthday/BlancLM05}.
+%
 A more satisfying approach is to consider confluence modulo equality \cite{Huet:1980ng}. 
-The simplest approach is to restrict our interest to top-level expressions of base type that result in values, in which case the following special case of confluence does hold (trivially when the only base type has a single value, but also more generally for other base types).
+%
+The simplest approach restricts our interest to top-level expressions
+of base type that result in values, in which case the following
+special case of confluence does hold (trivially when the only base
+type has a single value, but also more generally for other base
+types).
 \begin{lem}[Base Confluence]
@@
 
-The proofs are outlined in Appendix~\ref{sec:hole-filling}, along with a number of other necessary lemmas and definitions.
-In particular, we need additional machinery to properly handle the situation where we have filled a non-empty hole where a step had taken place in the original evaluation.
-These proofs and definitions are not included in the mechanization for two reasons.
-First, fill-and-resume is an optimization, and as just discussed, these properties, unlike those in Sec.~\ref{sec:calculus}, would not be conserved by some reasonable extensions of the core calculus.
-Second, it would require a significantly more complex representation of hole environments to encode the hole filling operation.
-Under the simple representation that suffices for the developments in Sec.~\ref{sec:calculus}, Agda cannot be convinced that the definition is well-founded (\citet{Nanevski2008} establish that it is in fact well-founded).
-
-Hole filling in the external language is understood in terms of hole filling in its expansion.
-Note that in practice, it may be useful to cache results from several previous expansions.
-For example, consider two edits, the first filling a hole $u$ with the number $2$, and the next applying the $+$ operator, resulting in $2 + \dehole{v}{\sigma}{}$.
-This second edit is not a hole filling edit with respect to the immediately preceding edit state, but it can be understood as filling hole $u$ with $2 + \dehole{v}{\sigma}{}$.
-Another related case arises in lab notebook interfaces like that of Jupyter/IPython \cite{PER-GRA:2007}.
-In this case, each cell can be understood as a series of \li{let} bindings ending in a hole, which is filled by the next cell.
-The result from each cell can be cached to avoid recomputing the environment from preceding cells.
-This contextual modal interpretation of lab notebook cells (and read-eval-print loops as a restricted case where edits to previous cells are impossible) is novel.
-
-So far, we have understood hole filling as an extralinguistic operation.
-However, a key feature CMTT that we have not yet touched on is the \emph{internalization} of metavariable binding and contextual substitution via the contextual modal types, $[\hGamma]\htau$, which are introduced by the operation $\mathsf{box}(\hGamma.d)$ and eliminated by the operation $\mathsf{letbox}(d_1, u.d_2)$.
-A hole filling can be interpreted as an expression of contextual modal type, and the act of hole filling followed by evaluation to the next possibly-indeterminate edit state as evaluation under the binder of a suitable $\mathsf{letbox}$ construct, which is enabled by the dynamic semantics in Sec.~\ref{sec:calculus}.
-This interpretation allows us to \emph{compute} hole fillings, rather than simply stating them, by specifying non-trivial expressions of modal type.
-This could serve as the basis for a \emph{live} computational hole refinement system, extending the capabilities of purely static hole refinement systems like those available in some proof assistants, e.g.
-the elaborator reflection system of Idris \cite{brady2013idris,DBLP:conf/icfp/ChristiansenB16} and the tactic system\todo{right?
-read a bit more}{} of Beluga \cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
-Each applied hole filling serves as a boundary between dynamic \emph{edit stages}.
-This contextual modal interpretation of live staged hole refinement nicely mirrors the modal interpretation of staged argument evaluation \cite{Davies:2001op}.
-We leave further development of this interpretation of hole refinement to future work.
+We outline the proofs in Appendix~\ref{sec:hole-filling}, along with a
+number of other necessary lemmas and definitions.
+%
+In particular, additional machinery handles the situation wherein a
+now-filled non-empty hole had taken a step in the original evaluation.
+
+\matt{We could (should?) move the next paragraph to the appendix, 
+before/after the proofs mentioned above; 
+its not really adding much to the discussion above, 
+but may be interesting to the very studious reader/prover}
+
+We exclude these proofs and definitions from the Agda mechanization
+for two reasons.
+%
+First, fill-and-resume is merely an optimization, and unlike the meta
+theory of \Secref{sec:calculus}, these properties are generally not
+conserved by certain reasonable extensions of the core
+calculus~(e.g., \matt{such as what?}).
+%
+Second, to properly encode the hole filling operation, such a
+mechanization requires a significantly more complex representation of
+hole environments; unfortunately, Agda cannot be easily convinced that
+the corresponding definition is well-founded (\citet{Nanevski2008}
+establish that it is in fact well-founded).
+%
+By contrast, the developments in \Secref{sec:calculus} do not require
+these more complex (and somewhat problematic) representations.
 

### 2018-04-15 related-work.tex
pass over Section 4; copy edits and reorganization

@@
 
+\matt{This discussion relates our design to CMTT; if we leave it in
+  the paper, we should leave it in this section, whose point is to be
+  comparative; the technical sections lack a sufficent ``attention
+  budget'' for the reader to wade through this discussion as they are
+  still learning the details of our paper.  Even for a really studious
+  reader that already knows CMTT, I still think it makes more sense
+  (organizationally) to give a detailed comparison here, and not
+  there.}
+
 To recap, the result of expansion is a well-typed internal expression with hole closures and casts, 
@@
 which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
-but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
+but under this interpretation an analogous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
 
@@
 allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% 2018.04.15
+%
+% MATT MOVED THIS MATERIAL HERE FROM calculus.tex
+%
+\subsection{MOVED FROM: Fill-and-resume}
+
+\matt{This discussion relies on a more developed exposition of CMTT
+  that we have/will give in this paper; we should leave it in this
+  section, or remove it}
+
+So far, we have understood hole filling as an extralinguistic operation. 
+%
+However, a key feature CMTT that we have not yet touched on is the
+\emph{internalization} of metavariable binding and contextual
+substitution via the contextual modal types, $[\hGamma]\htau$, which
+are introduced by the operation $\mathsf{box}(\hGamma.d)$ and
+eliminated by the operation $\mathsf{letbox}(d_1, u.d_2)$.
+%
+A hole filling can be interpreted as an expression of contextual modal
+type, and the act of hole filling followed by evaluation to the next
+possibly-indeterminate edit state as evaluation under the binder of a
+suitable $\mathsf{letbox}$ construct, which is enabled by the dynamic
+semantics in Sec.~\ref{sec:calculus}.
+%
+This interpretation allows us to \emph{compute} hole fillings, rather
+than simply stating them, by specifying non-trivial expressions of
+modal type.
+%
+This could serve as the basis for a \emph{live} computational hole
+refinement system, extending the capabilities of purely static hole
+refinement systems like those available in some proof assistants,
+e.g. the elaborator reflection system of Idris
+\cite{brady2013idris,DBLP:conf/icfp/ChristiansenB16} and the tactic
+system\todo{right? read a bit more}{} of Beluga
+\cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
+%
+Each applied hole filling serves as a boundary between dynamic
+\emph{edit stages}.
+%
+This contextual modal interpretation of live staged hole refinement
+nicely mirrors the modal interpretation of staged argument evaluation
+\cite{Davies:2001op}.
+%
+We leave further development of this interpretation of hole refinement
+to future work.
+
+\subsubsection{Fill-and-resume example}
+
+\matt{I think this example is intended to illustrate some point, but I
+  could not discern what that point was}
+
+As an example, consider two edits, the first filling a hole~$u$ with
+the number~$2$, and the next applying operator~$+$, resulting in $2
++ \dehole{v}{\sigma}{}$.
+%
+This second edit is not a hole filling edit with respect to the
+immediately preceding edit state, but it can be understood as filling
+hole $u$ with $2 + \dehole{v}{\sigma}{}$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+
+\matt{This thought seems to rely on a more developed exposition of
+  CMTT that we have/will give in this paper; we should leave it in
+  this section, or remove it}
+
+Another related case arises in lab notebook interfaces like that of
+Jupyter/IPython \cite{PER-GRA:2007}.
+%
+In this case, each cell can be understood as a series of \li{let}
+bindings ending in a hole, which is filled by the next cell.
+%
+The result from each cell can be cached to avoid recomputing the
+environment from preceding cells. This contextual modal interpretation
+of lab notebook cells (and read-eval-print loops as a restricted case
+where edits to previous cells are impossible) is novel.
+
+\subsubsection{Unmoored notes}
+
+ \matt{Will any reader actually wonder this? Other than the definition
+   of substitution in Section 4, does this thought connect to any
+   other statement elsewhere in the paper?}
+
+This case shows why we cannot interpret a non-empty hole as an empty
+hole of arrow type applied to the enveloped expression---the hole
+filling operation would not operate as expected under this
+interpretation.

### 2018-04-15 calculus.tex
spell check sections 3 and 4

@@
 This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
-In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
+In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
 We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}
@@
 We start with synthesis for the programmer's ``top level'' external
-exression.
+expression.
 
@@
 %
-Inuitively, the subsumption rule, as previously discussed, allows us
+Intuitively, the subsumption rule, as previously discussed, allows us
 to check an external expression against any type consistent with the
@@
 %
-Non-empty hole closures~$\dhole{d}{u}{\sigma}{}$ have no direct correspondance with a notion from CMTT (see Sec.~\ref{sec:resumption}).
+Non-empty hole closures~$\dhole{d}{u}{\sigma}{}$ have no direct correspondence with a notion from CMTT (see Sec.~\ref{sec:resumption}).
 
@@
 The conclusion of Rule~{ESAp} inserts the cast on the argument's
-expansion, ufrom the type it is assigned by the final premise of the
+expansion, from the type it is assigned by the final premise of the
 rule~(type~$b$ here), to the argument type of the matched arrow type
@@
         %
-        Our judgemental formulation is more faithful to the mechanization and
+        Our judgmental formulation is more faithful to the mechanization and
         cleaner for our purposes, because we are distinguishing several
@@
 
-The preservation theorem establishes that transitions perserve type
+The preservation theorem establishes that transitions preserve type
 assignment, i.e. that the type of an expression accurately predicts
@@
 %
-The proof relies on an analagous preservation lemma for instruction
+The proof relies on an analogous preservation lemma for instruction
 transitions and a standard substitution lemma (stated in
@@
 The basic approach is standard: we model judgements as 
-inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite somtehing?}{}. 
+inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite something?}{}. 
 %

### 2018-04-15 calculus.tex
present tense
nit

@@
 reduction from \Secref{sec:calculus}, where evaluation order is
-unspecified (i.e. formally, we omitted the bracketed premises):
+unspecified (i.e. formally, we omit the bracketed premises).
 %
@@
 sub-expressions are
-reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit},
-but reduction under binders is at odds with the standard approaches
-used in implementing programming
-languages \cite{DBLP:conf/birthday/BlancLM05}.
+reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit}.
+%
+However, reduction under binders conflicts with the standard implementation approaches
+for most programming languages \cite{DBLP:conf/birthday/BlancLM05}.
 %

### 2018-04-15 calculus.tex
nit

@@
 %
-A more satisfying approach is to consider confluence modulo equality \cite{Huet:1980ng}. 
-%
-The simplest approach restricts our interest to top-level expressions
+A more satisfying approach considers confluence modulo equality \cite{Huet:1980ng}. 
+%
+The simplest such approach restricts our interest to top-level expressions
 of base type that result in values, in which case the following

### 2018-04-15 calculus.tex
note to Cyrus
nit

@@
 now-filled non-empty hole had taken a step in the original evaluation.
+\matt{Cyrus --- I tried to simplify and shorten the earlier version of the prior sentence, but I \emph{may} have unintentionally goofed up what it was trying to say; please double check the prior sentence}
 
@@
 its not really adding much to the discussion above, 
-but may be interesting to the very studious reader/prover}
+but may be interesting to the very studious reader/prover; also, leaving it here also gives a cranky reviewer a nice weapon to hit us with}
 

### 2018-04-15 calculus.tex
nit

@@
 its not really adding much to the discussion above, 
-but may be interesting to the very studious reader/prover; also, leaving it here also gives a cranky reviewer a nice weapon to hit us with}
+but may be interesting to the very studious reader/prover; also, leaving it here also gives a lazy/cranky reviewer a nice weapon to hit us with}
 

### 2018-04-15 calculus.tex
cut extraneous discussion

@@
 
-The supplemental material includes our mechanization of the semantics and metatheory of \HazelnutLive, 
+The supplemental material includes our Agda-based
+mechanization  \cite{norell2009dependently,norell:thesis} 
+of the semantics and metatheory of \HazelnutLive, 
 including proofs of all of the theorems stated above and all necessary lemmas\todo{make this so ian!}{}. 
 %
-We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis} 
-(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
-%
-Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation. 
-%
-Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
+%We choose the 
+%(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
+%
+%Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation. 
+%
+A%gda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
 %

### 2018-04-15 calculus.tex
fewer words

@@
 of the semantics and metatheory of \HazelnutLive, 
-including proofs of all of the theorems stated above and all necessary lemmas\todo{make this so ian!}{}. 
+%including proofs of all of the 
+for the theorems stated above, and all necessary lemmas\todo{make this so ian!}{}. 
 %
@@
 %
-A%gda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
-%
-The basic approach is standard: we model judgements as 
+%Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
+%
+%
+Our approach is standard: We model judgements as 
 inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite something?}{}. 

### 2018-04-15 calculus.tex
nit

@@
 The primary benefit of specifying the \HazelnutLive external language 
-bidirectionally is that the programmer does not need to annotate each hole with a type. 
+bidirectionally is that the programmer need not annotate each hole with a type. 
 %
@@
 %including proofs of all of the 
-for the theorems stated above, and all necessary lemmas\todo{make this so ian!}{}. 
+including the theorems stated above, and all necessary lemmas\todo{make this so ian!}{}. 
 %

### 2018-04-15 calculus.tex
nit
organize diffs with Hazelnut into the first paragraph of Sec 3, with a pointer to appendix

@@
 
-We start with the type system of the \HazelnutLive
-external language, which closely follows the \Hazelnut type
-system \cite{popl-paper}; we discuss the minor differences as they come up below.
+We start with the type system of the \HazelnutLive external language,
+which closely follows the \Hazelnut type system \cite{popl-paper},
+except that (1) holes in \HazelnutLive each have a \emph{unqiue name},
+and (2) for brievity of exposition, we consider a more austere core
+language here that lacks numbers and sum types, which we discuss
+(along with other systematic extensions) in
+Appendix~\ref{sec:extensions}.
 
@@
 written $\hehole{u}$, where $u$ is a name for the hole, which we tacitly assume is unique.
-Holes did not have names in \Hazelnut. 
+%
+%(Holes in \Hazelnut are \emph{unnamed}).
 %
@@
 %
-By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
+%\matt{Extraneous and uninteresting:}
+%By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
 %

### 2018-04-15 calculus.tex
fewer words

@@
 %
-Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.
+Rule {SVar} synthesizes the corresponding type from $\hGamma$.
 
@@
 %
-\Figref{fig:expansion} gives the rules governing expansion, and \Figref{fig:hasType} gives the rules governing type assignment for internal expressions.
+\Figref{fig:expansion} defines this expansion, and \Figref{fig:hasType} defines type assignment for internal expressions.
 %

### 2018-04-15 calculus.tex
nit

@@
 %
-The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is consistent with $\htau$.
+The following theorem establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is consistent with $\htau$.
 %
@@
 \noindent
-
+%
 %The reason analytic expansion produces an expansion of consistent
@@
 %
-Intuitively, the subsumption rule, as previously discussed, allows us
+Intuitively, the subsumption rule permits us
 to check an external expression against any type consistent with the
-type the expression actually synthesizes.
-
+type that the expression \emph{actually} synthesizes:
+%
 \begin{thm}[Type Assignment Unicity]
@@
 %
-We call this feature \emph{fill-and-resume}, by analogy to the
-``edit-and-resume'' features available in some systems today, such as
-Visual Studio (see below for a comparison)\todo{cite and compare}{}.
+This \emph{fill-and-resume} behavior formalizes the (ad hoc)
+``edit-and-resume'' behavior permitted by some current systems,
+e.g.,~Microsoft Visual Studio.\todo{cite}
+
+%(see below for a comparison)\todo{cite and compare}{}. ---NO TIME/SPACE for this.
 
@@
 %
-we interpret the hole environments as \emph{delayed substitutions},
+we interpret hole environments as \emph{delayed substitutions},
 the same interpretation suggested for closures in contextual modal

### 2018-04-16 calculus.tex
near-final intro

--- 2018-04-15 calculus.tex
@@
 
-The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes. 
+The syntax of the core calculus specified in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
 We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. 
@@
 This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
-In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
+In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
 We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}
@@
 
-We start with the type system of the \HazelnutLive external language,
-which closely follows the \Hazelnut type system \cite{popl-paper},
-except that (1) holes in \HazelnutLive each have a \emph{unqiue name},
-and (2) for brievity of exposition, we consider a more austere core
-language here that lacks numbers and sum types, which we discuss
-(along with other systematic extensions) in
-Appendix~\ref{sec:extensions}.
+We start with the type system of the \HazelnutLive
+external language, which closely follows the \Hazelnut type
+system \cite{popl-paper}; we discuss the minor differences as they come up below.
 
@@
 We start with synthesis for the programmer's ``top level'' external
-expression.
+exression.
 
@@
 The primary benefit of specifying the \HazelnutLive external language 
-bidirectionally is that the programmer need not annotate each hole with a type. 
+bidirectionally is that the programmer does not need to annotate each hole with a type. 
 %
@@
 written $\hehole{u}$, where $u$ is a name for the hole, which we tacitly assume is unique. 
-%
-%(Holes in \Hazelnut are \emph{unnamed}).
+Holes did not have names in \Hazelnut. 
 %
@@
 %
-%\matt{Extraneous and uninteresting:}
-%By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
+By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
 %
@@
 %
-Intuitively, the subsumption rule, as previously discussed, allows us
+Inuitively, the subsumption rule, as previously discussed, allows us
 to check an external expression against any type consistent with the
@@
 
-The rules governing expansion of constants, variables and lambda expressions
---- Rules {ESConst}, {ESVar}, {ESLam} and {EALam}
---- and the corresponding type assignment rules
---- Rules {TAConst}, {TAVar} and {TALam}
---- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}.
-%
-Consequently, the corresponding cases of
-Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
-Theorem~\ref{thm:expansion-generality} are straightforward.
-%
-To support type assignment, all lambdas in the internal language are
-half-annotated---Rule {EALam} inserts the annotation when expanding an
-unannotated external lambda based on the given type.
-%
-Some rules perform \emph{cast insertion}, making them more interesting
-than the others; these rules include those for holes, and the those for
-function applications and ascriptions.  We consider each below.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% <<<<<<<<<<<<< Matt is here (latest pass over everything above is done -- 2018.04.11)
+
+The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}
+(so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). 
+Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation when expanding an unannotated external lambda based on the given type.
+The rules governing holes and the rules governing function applications and ascriptions, which both perform \emph{cast insertion}, are more interesting, so let us consider both cases in turn.
 
 \subsubsection{Hole Expansion}\label{sec:hole-expansion} 
-Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
+Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$.
 %
-The hole name~$u$ on a hole closure identifies the external hole to which the hole closure corresponds.
-%
-%Note that
-While we assume each hole name to be unique in the external language, due to substitution, evaluated (internal) terms may consist of multiple hole closures with the same name.
-%
-For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}.
-%
-There, we number each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.
 
-The hole expansion rules introduce hypotheses of the form~$\Dbinding{u}{\hGamma}{\htau}$ into the hole context~$\Delta$.
-%
-For each hole name~$u$, the hole context records a type~$\tau$, and a typing context~$\Gamma$.\footnote{
+The hole name, $u$, on a hole closure identifies the external hole that the hole closure corresponds to. 
+Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution. 
+For example, the result from Fig.~\ref{fig:grades-example} showed four closures for the hole named 1\todo{update with actual number (four?) and hole name later}. 
+There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation so we omit hole closure numbers from the core calculus.
+
+The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. 
+The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$, for each hole name, $u$.\footnote{
 We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} 
-%
-We borrow this hole context notation from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
-%
-Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
-%
-In the synthetic hole expansion rules~{ESEHole} and~{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the typing rules.
-%
-However, the first two premises of the expansion subsumption rule~{EASubsume} disallow the use of subsumption for holes in analytic position.
-%
-Instead, we employ separate analytic rules~{EAEHole} and~{EANEHole}, which each record the checked type~$\tau$ in the hole context.
-%
-Consequently, we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.
+This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
+In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. 
+In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. 
+However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position. 
+Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. 
+This is again so that we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.
 
-Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
-%
-The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
-%
-Initially, when no evaluation has yet occurred, the hole expansion rules generate an initial environment that consists of the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
-%
+Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$. 
+The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. 
+Initially, no evaluation has occurred, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
 \begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
@@
 \noindent
-The type assignment rules for hole closures~{TAEHole} and {TANEHole} each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
-%
-Formally, we define this relation as follows:\todo{check that the definition in the Agda corresponds}
+The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
 \begin{defn}[Substitution Typing]
@@
 Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} 
-%
-\Secref{sec:evaluation} defines how these closure environments evolve during evaluation.
-%
-Non-empty hole closures~$\dhole{d}{u}{\sigma}{}$ have no direct correspondence with a notion from CMTT (see Sec.~\ref{sec:resumption}).
+We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}. 
+Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT (see Sec.~\ref{sec:resumption}).
 
@@
 %
-% USE A TOPIC SENTENCE SO THAT THE READER IS GUIDED A LITTLE MORE, e.g.,
-%
-Cast insertion mediates between the type systems of the external and
-internal languages, where the former permits subsumption, and the
-latter does not.
-%-----------------------------------------------------------------------------------------------
-%
-To see why this mediation is necessary, consider the following
-example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
-%
-Viewed as an external expression, this example synthesizes type
-$\tehole$, since the hole type annotation on variable~$x$ permits
-applying~$x$ as a function of type~$\tarr{\tehole}{\tehole}$, and base
-constant~$c$ may be checked against type~$\tehole$, by subsumption.
-%
-However, viewed as an internal expression, this example is not
-well-typed---the type assignment system defined in
-\Figref{fig:hasType} lacks subsumption.
-%
-Indeed, it would violate type safety if we could assign a type to this
-example in the internal language, because beta reduction of this
-example viewed as an internal expression would result in $c(c)$, which
-is clearly not well-typed.
-%
-The difficulty arises because leaving the argument type unknown also leaves unknown how
-the argument is being used (in this case, as a function).\footnote{In a system where type reconstruction is first used
-to try to fill in type holes, we could express a similar example by
-using $x$ at two or more different types, thereby causing type
-reconstruction to fail.
-%
-% On the other hand, if it is acceptable to arbitrarily choose one of
-%the possible types, and type reconstruction is complete, then type
-%holes will never appear in the internal language and the cast
-%insertion machinery described in this section can be omitted
-%entirely, leaving only the hole closure machinery described
-%previously.
+Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. 
+The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption.
+However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. 
+Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed.
+The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
+%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
 }
 \todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} 
-By our interpretation of hole types as unknown types from gradual type
-theory, we can address the problem by performing cast insertion.
+By our  interpretation of hole types as unknown types from gradual type theory, we can address the problem by performing cast insertion. 
 %
 
-The cast form in Hazelnut Live consists of $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
-%
-This form serves to ``box'' an expression of type $\htau_1$ for
-treatment as an expression of a consistent type $\htau_2$
-(Rule TACast in \Figref{fig:hasType}).
-%
-\footnote{
-In the earliest work on gradual type theory, the cast form only gave
-the target type~$\htau_2$ \cite{Siek06a}, but it simplifies matters
-to include the assigned type~$\htau_1$ in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.
-}
+The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of a consistent type $\htau_2$.%
+\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}
 
-Expansion inserts casts for function applications and ascriptions.
-%
-The latter is more straightforward: Rule~{ESAsc}
-in \Figref{fig:expandSyn} inserts a cast from the assigned type to the
-ascribed type.
-%
-Theorem~\ref{thm:typed-expansion} inductively ensures that the two
-types are consistent.
-%
-We include ascription for exposition purposes---this form is derivable
-by using application together with the half-annotated identity, $e
-: \tau = \hap{(\halam{x}{\htau}{x})}{e}$; as such, application
-expansion, discussed below, is more general.
+Casts are inserted during the expansion of function applications and ascriptions. 
+The latter is more straightforward: 
+Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. 
+Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  
+Note that we included ascription only for the sake of exposition---it can be defined using application together with the half-annotated identity, $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$, so application expansion, discussed next, is more general.
 
-Rule~{ESAp} expands function applications.
-%
-To understand the rule, consider the expansion of external
-expression~$\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$, the example
-discussed above:
-\[
-        \hap{\dcasttwo{
-        (\halam{x}{\tehole}{\underbrace{
-                \hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}
-                {\dcasttwo{c}{b}{\tehole}}
-                }_{\textrm{expansion of function body}~x(c)}
-        }
-        )}{\tarr{\tehole}{\tehole}}{
-           \tarr{\tehole}{\tehole}}
-           }
-           {\dcasttwo{c}{b}{\tehole}}
+Cast insertion by the application expansion rule, Rule~{ESAp}, requires a bit more care. 
+To understand the rule, consider the expansion for the example above:
+\[\hap{\dcasttwo{(\halam{x}{\tehole}{\underline{\hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}}})}{\tarr{\tehole}{\tehole}}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}
 \]
-Consider the (indicated) function body,
-%
-where expansion inserts a cast on both the function expression~$x$ and its argument~$c$.
-%
-Together, these casts for~$x$ and~$c$ permit assigning a type to the
-function body according to the rules in \Figref{fig:hasType}, where we
-could not do so under the same context without casts.
-%
-We separately consider the expansions of~$x$ and of~$c$.
+Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. 
+A cast has been inserted on both the function expression, $x$, and the the argument, $c$.
 
-First, consider the function position of this application, here variable~$x$.
-%
-Without any cast, the type of variable~$x$ is hole type~$\tehole$;
-however, the inserted cast on~$x$ permits treating it as though it has
-arrow type $\tarr{\tehole}{\tehole}$, as described
-in \Secref{sec:external-statics}.
-%
-The first three premises of Rule~{ESAp} accomplish this
-%
-by first synthesizing a type for the function expression, here
-$\tehole$, then
-%
-by determining the matched arrow type~$\tarr{\tehole}{\tehole}$, and
-finally,
-%
-by performing analytic expansion on the function expression with this
-matched arrow type.
-%
-The resulting expansion has some type~$\tau_1'$ consistent with the matched arrow type.
-%
-In this case, because the subexpression~$x$ is a variable, analytic
-expansion goes through subsumption so that type~$\tau_1'$ is
-simply~$\tehole$.
-%
+The cast on $x$ allows us to treat the variable $x$, which is of type $\tehole$, as being of the matched arrow type $\tarr{\tehole}{\tehole}$, as described in Sec.~\ref{sec:external-statics}. 
+The first three premises of Rule~{ESAp} accomplish this by first synthesizing a type for the function expression, here $\tehole$, then determining the matched arrow type, $\tarr{\tehole}{\tehole}$, and then performing analytic expansion on the function expression with this matched arrow type.
+The resulting expansion will have some type $\tau_1'$ consistent with the matched arrow type. 
+In this case, because $x$ is of variable form, analytic expansion goes through subsumption so $\tau_1'$ is simply $\tehole$. 
 The conclusion of the rule inserts the corresponding cast. 
-%
-We go through type synthesis, \emph{then} analytic expansion, so that the hole
-context records the matched arrow type for holes in function position,
-rather than the type~$\tehole$ for all such holes, as would be the case
-in a variant of this rule using synthetic expansion for the function
-expression\todo{put this in the appendix?}{}.
+We go through type synthesis then analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.
 
-Next, consider the application's argument, here constant~$c$.
-%
-The conclusion of Rule~{ESAp} inserts the cast on the argument's
-expansion, from the type it is assigned by the final premise of the
-rule~(type~$b$ here), to the argument type of the matched arrow type
-of the function expression~(type~$\tehole$ here).
+The conclusion of Rule~{ESAp} inserts the cast on the argument's expansion, from the type it is assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type of the matched arrow type of the function expression, here $\tehole$.
 
-The example's second, outermost application uses through the same
-application expansion rule.
-%
-In this case, the cast on the function is the identity cast for
-$\tarr{\tehole}{\tehole}$.
-%
-For simplicity, we do not attempt to avoid the insertion of identity
-casts in the core calculus; these will simply never fail during
-evaluation.
-%
-However, it is safe to eliminate such identity casts during expansion,
-and some formal accounts of gradual typing do so by defining three
-application rules, including the original account of \cite{Siek06a}).
+Observe that together, these casts allow us assign a type to the function body according to the rules in Fig.~\ref{fig:hasType}, where we could not do so under the same context without casts.
+
+The outer application in the example above goes through the same rule. 
+In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$. 
+We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), 
+but it is safe to eliminate these during expansion (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).
 
@@
 
-Figures~\ref{fig:isGround}-\ref{fig:step} give dynamic semantics for \HazelnutLive,
-%
+To recap, the result of expansion is a well-typed internal expression with hole closures and casts, 
+where the former corresponds to metavariable closures from CMTT \cite{Nanevski2008}, 
+and the latter corresponds to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}. 
+However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.
+
+The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.
+The paper therefore proved only a subject reduction property (which is closely related to type preservation). 
+This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. 
+In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables.
+\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion%
+\footnote{Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property.
+These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.},
+which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
+but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
+
+Furthermore, we need to integrate casts into the dynamic semantics. 
+Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. 
+The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms.
+The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. 
+Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}\todo{where?}{}, is for cast failure to instead insert a membrane around the dynamic type error, 
+much like a non-empty hole serves as a membrane around a static type error, 
+allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.
+
+The dynamic semantics for \HazelnutLive specified in Figures~\ref{fig:isGround}-\ref{fig:step} addresses the difficulties just outlined, 
 resulting in a system capable of running incomplete programs without aborting when a hole is encountered. 
-%
-We equip this semantics with a meaningful notion of type safety (involving both preservation and progress).
-%
+This semantics is equipped with a meaningful notion of type safety (involving both preservation and progress). 
 We also establish that the standard notion of type safety falls out when running complete terms.
 
-The dynamic semantics consists of a ``small-step'' transition system,
-%
-where cast-related machinery closely follows the cast calculus from
-the ``refined'' account of the gradually typed lambda calculus
-by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be
-theoretically well-behaved.
-%
-In particular, \Figref{fig:isGround} defines the judgement
-$\isGround{\htau}$, which distinguishes the base type~$b$ and the
-least specific arrow type~$\tarr{\tehole}{\tehole}$, as \emph{ground
-types}; this judgement helps simplify the treatment of function casts.
-%
-\Figref{fig:isFinal} defines the judgement $\isFinal{d}$, which
-distinguishes the final forms of the transition system.
-%
-There are two classes of final forms: (possibly-)boxed values and
-indeterminate forms.%
-%
-\footnote{
-        In most accounts of the cast calculus, values and ground types
-        are distinguished with separate grammars together with an
-        implicit identification convention.
-        %
-        Our judgmental formulation is more faithful to the mechanization and
-        cleaner for our purposes, because we are distinguishing several
-        classes of final forms.
-}
+The dynamic semantics is specified as a ``small-step'' transition system. 
+The cast-related machinery is based closely on the cast calculus from the ``refined'' account of the gradually typed lambda calculus by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be theoretically well-behaved.
+In particular, Fig.~\ref{fig:isGround} defines the judgement $\isGround{\htau}$, which distinguishes the base type, $b$, and the least specific arrow type, $\tarr{\tehole}{\tehole}$, as \emph{ground types}, which play a role in simplifying the treatment of function casts.
+Fig.~\ref{fig:isFinal} defines the judgement $\isFinal{d}$, which distinguishes the final forms of the transition system. 
+There are two classes of final forms: (possibly-)boxed values and indeterminate forms.%
+\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention. 
+Our judgemental formulation is more faithful to the mechanization and cleaner for our purposes, because we are distinguishing several classes of final forms.}
 
@@
 
-The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as
-either ordinary values~(Rule~{BVVal}), or one of two cast forms: casts
-between function types (except identity casts) and casts from a ground
-type to the hole type. In each case, the cast operates on a boxed
-value.
-%
-%, which correspond to the values from the cast calculus and include
-%the classic values from the lambda calculus, distingui%shed by
-%$\isValue{d}$,
-%
-The judgement $\isIndet{d}$ defines \emph{indeterminate} forms, so
-named because they arise from the presence of expression holes and
-failed casts in a program, and so, conceptually, their ultimate value
-awaits programmer action (see Sec.~\ref{sec:resumption}).
-%
-The first two rules specify that \emph{empty} hole closures are always
-indeterminate, and that \emph{non}-empty hole closures are indeterminate when
-they consist of a \emph{final} inner expression.
-%
-Below, we discuss the the remaining indeterminate forms simultaneously
-with the corresponding transition rules, also discussed
-below.\todo{relationship to weak head normal forms}
+The judgement $\isBoxedValue{d}$ distinguishes (possibly-)boxed values, which correspond to the values from the cast calculus and include the classic values from the lambda calculus, distinguished by $\isValue{d}$, as well as two cast forms:
+casts between function types (except identity casts) 
+and casts from a ground type to the hole type. In both cases, the cast must be on a boxed value.
+
+The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, 
+so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:resumption}). 
+The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). 
+The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}
 
@@
 The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}. 
-%
-Top-level transitions are \emph{steps}~$\stepsToD{}{d}{d'}$, governed by Rule~{Step} in \Figref{fig:step}, which
-%
+Top-level transitions, i.e. \emph{steps}, $\stepsToD{}{d}{d'}$, are governed by a single rule, Rule {Step} in Fig.~\ref{fig:step}, which 
 (1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$; 
-%
-(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in \Figref{fig:instruction-transitions};
-%
-and (3) places $d_0'$ back in the selected position, indicated in
-the evaluation context by the \emph{mark}~$\evalhole$, to obtain $d'$.%
-\footnote{
-        We say ``mark'' here to avoid confusion with the (orthogonal) ``holes'' of \HazelnutLive.
-        %
-        %the form $\evalhole$ in the grammar of
-        %evaluation contexts is referred to as the \emph{hole}, but
-        %this hole is a technical device entirely orthogonal to the
-        %holes of this paper, so we use the term ``mark'' instead.
-        }
-%
-
-%% \matt{Next paragraph can be dropped entirely for space}
-%% %
-%% This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing.
-%% Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
-%% It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
-%% \todo{Put the search rules in the appendix?}{}
+(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in Fig.~\ref{fig:instruction-transitions}; 
+and (3) places $d_0'$ back in the selected position, which is marked%
+\footnote{In the literature, the form $\evalhole$ in the grammar of evaluation contexts is referred to as the \emph{hole}, but this hole is a technical device entirely orthogonal to the holes of this paper, so we use the term ``mark'' instead.}
+in the evaluation context by $\evalhole$, to obtain $d'$.
+This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
+Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
+It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
+\todo{Put the search rules in the appendix?}{}
 
@@
 \subsubsection{Application and Substitution} 
-%
-Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the
-standard beta reduction transition.
-%
-The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in
-Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} may be \emph{included}
-to specify an eager, left-to-right evaluation strategy, or \emph{excluded} to
-leave the evaluation order unspecified.
-%
-In our metatheory, we use the latter version of the rules, both for
-the sake of generality, and to support fill-and-resume~(\Secref{sec:resumption}).
+Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the standard beta reduction transition. 
+The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} can be included to specify an eager, left-to-right evaluation strategy, 
+or excluded to leave the evaluation order unspecified. 
+We do the latter in our metatheory, both for the sake of generality and for reasons we return to in Sec.~\ref{sec:resumption}.
 
-A substitution~$[d/x]d'$ operates in the standard capture-avoiding manner~\cite{pfpl}. 
-%
+Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. 
 The only cases of special interest arise when substitution reaches a hole closure:
@@
 \]
-In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$. 
-%
-For example, $\stepsToD{}
-    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
-    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. 
-%
-Reduction may duplicate hole closures, just as with any other substituted term;
-%
-consequently, the environments of different closures for the same hole name may differ, 
-e.g., when an reduction applies a function with a hole closure body multiple times.%
+In both cases, we write $[d/x]\sigma$ to perform substitution on each expression in the hole environment, $\sigma$. 
+For example, $\stepsToD{}{\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}{\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. 
+Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g.
+because a function with a hole closure in its body is applied multiple times.%
 \todo{show example here or refer back to sec 2?}{} 
-%
-Hole closures may also appear within the environments of other hole
-closures, giving rise to the closure paths described in
-Sec.~\ref{sec:paths}\todo{where?}{}.
+Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}\todo{where?}{}).
 
-The ITBeta rule is not the only rule we need to handle function
-application, because lambdas merely one (final) form of arrow type.
-%
-Two other situations may also arise.
+The ITBeta rule is not the only rule we need to handle function application, because lambdas are not the only final forms of arrow type. There are two other classes of possibilities.
 
-First, the expression in function position might be a cast between
-arrow types, in which case we apply the arrow cast conversion rule,
-Rule {ITApCast}, to rewrite the application form, obtaining an
-equivalent application where the expression~$d_1$ under the function
-cast is exposed.
-%
-We know from inverting the typing rules that~$d_1$ has type
-$\tarr{\htau_1}{\htau_2}$, and that~$d_2$ has type~$\htau_1'$, where
-$\tconsistent{\htau_1}{\htau_1'}$; consequently, we maintain type
-safety by placing a cast on~$d_2$ from~$\htau_1'$ to~$\htau_1$.
-%
-The result of this application has type $\htau_2$, but the
-original cast promised that the result would have consistent type
-$\htau_2'$, so we also need a cast on the result from $\htau_2$ to
-$\htau_2'$.
+The expression in function position might be a cast between arrow types, in which case we apply the arrow cast conversion rule, Rule {ITApCast}, to rewrite the application form, obtaining an equivalent application where the expression under the function cast, $d_1$, is exposed.
+We know from inverting the typing rules that $d_1$ is of type $\tarr{\htau_1}{\htau_2}$ and that $d_2$ is of type $\htau_1'$ where $\tconsistent{\htau_1}{\htau_1'}$, so to maintain type safety, we must place a cast on $d_2$ from  $\htau_1'$ to $\htau_1$.
+The result of this application will be of type $\htau_2$, but the original cast promised that the result would be of consistent type $\htau_2'$, so we also need a cast on the result from $\htau_2$ to $\htau_2'$.
 
-Second, the expression in function position may be indeterminate,
-where arrow cast conversion is not applicable,
-e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$.
-%
-In this case, the application is indeterminate (Rule~{IAp}
-in \Figref{fig:isFinal}), and the application's reduction reduces no
-further.
-
-The rules take care to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.%
+The other possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$. 
+In this case, no further progress can be made, so the function application as a whole is also deemed indeterminate by Rule {IAp} in Fig.~\ref{fig:isFinal}. 
+Note that we are careful to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.%
 \todo{prove this as stated from Ian's disjointness proofs}{}%
@@
 \subsubsection{Casts}
-Rule {ITCastId} drops identity casts. The remaining instruction
-transition rules assign meaning to non-trivial casts.
-%
-As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term
-cast to hole type is statically obscure,
-%
-so we must wait for \emph{use} of the term at some other type, via a
-cast away from hole type, to detect dynamic type errors.
-%
-Rules {ITCastSucceed} and {ITCastFail} handle this situation when the
-two types involved are ground types (Fig.~\ref{fig:isGround}).
-%
-If the two ground types are equal, then the cast succeeds and the cast
-may be dropped.
-%
-If they are not equal, then the cast fails and the failed cast form,
-$\dcastfail{\dexp}{\htau_1}{\htau_2}$, arises.
-%
-Rule {TAFailedCast} specifies that a failed cast is well-typed exactly
-when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type
-disequal to $\tau_1$.
-%
-Rule {IFailedCast} specifies that a failed cast operates as an
-indeterminate form once $d$ is final.
-%
-This rule allows evaluation to proceed after cast failure, much as
-with hole closures.
+Rule {ITCastId} drops identity casts. The remaining instruction transition rules assign meaning to non-trivial casts. 
+As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term cast to hole type is statically obscure, 
+so we need to wait until it is actually used at some other type, via a cast away from hole type, to be able to determine whether there is a problem. 
+Rules {ITCastSucceed} and {ITCastFail} handle this situation when the two types involved are ground types (Fig.~\ref{fig:isGround}). 
+If the two ground types are equal, then the cast succeeds and the cast can be dropped. 
+If they are not equal, then the cast fails and the failed cast form, $\dcastfail{\dexp}{\htau_1}{\htau_2}$, arises. 
+Rule {TAFailedCast} specifies that a failed cast is well-typed exactly when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type disequal to $\tau_1$. 
+Rule {IFailedCast} specifies that a failed cast operates as an indeterminate form once $d$ is final. 
+This allows evaluation to proceed after cast failure, much as with hole closures.
 
-Rules {TAFailedCast} and {IFailedCast} each operate at ground type. 
-%
-The two remaining instruction transition rules, Rule~{ITGround}
-and~{ITExpand}, insert intermediate casts from non-ground type to a
-consistent ground type, and \emph{vice versa}.
-%
-These rules serve as technical devices, permitting us to restrict our
-interest exclusively to ground types.
-%
-Here, the only non-ground types are the arrow types, so the grounding
-judgement~$\groundmatch{\tau_1}{\htau_2}$ (\Figref{fig:groundmatch}),
-produces the ground arrow type~$\tarr{\tehole}{\tehole}$.
-%
+The two rules just described operate at ground type. 
+The two remaining instruction transition rules, Rule {ITGround} and {ITExpand}, operate as technical devices that allow us to restrict our interest to ground types exclusively by inserting intermediate casts from non-ground type to  a consistent ground type, and \emph{vice versa}.
+In this core calculus, the only non-ground types are the arrow types, so the grounding judgement $\groundmatch{\tau_1}{\htau_2}$ defined in Fig.~\ref{fig:groundmatch}, produces the ground arrow type, $\tarr{\tehole}{\tehole}$.
 More generally, this judgement is governed by the following invariant:
@@
 
-In all other cases, casts either consist of boxed values or
-indeterminate terms according to the remaining rules
-in \Figref{fig:isFinal}.
-%
-In particular, Rule {ICastHoleGround} handles casts that are not of
-the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$, from hole to
-ground type.
-%
-In the cast calculus, since the only canonical form of type $\tehole$
-is $\dcasttwo{\dexp}{\htau_1}{\tehole}$, there exist no such
-irreducible terms.
+In all other cases, casts are either boxed values or indeterminate according to the remaining rules in Fig.~\ref{fig:isFinal}. 
+Note in particular Rule {ICastHoleGround}, which handles casts from hole to ground type that are not of the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$. 
+In the cast calculus, there are no such irreducible terms because the only canonical form of type $\tehole$ is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.
 
 \subsubsection{Type Safety} 
-%
-Type safety establishes that the static and dynamic semantics of a
-language cohere.
-%
-We follow the approach developed by \citet{wright94:_type_soundness},
-now standard \cite{pfpl}, which distinguishes two type safety
-properties, preservation and progress.
-%
-To permit the evaluation of incomplete programs, we establish these
-properties for terms typed under arbitrary hole context $\Delta$.
-%
-We assume an empty context for~$\hGamma$; to run open programs, the
-system may treat free variables as empty holes with a corresponding
-name.
+The purpose of establishing type safety is to ensure that the static and dynamic semantics of a language cohere. 
+We follow the approach developed by \citet{wright94:_type_soundness}, now standard \cite{pfpl}, which distinguishes two type safety properties, preservation and progress. 
+In order to allow for the evaluation of incomplete programs, we establish these properties for terms typed under arbitrary hole context $\Delta$. 
+We can assume $\hGamma$ is empty (in order to run programs with free variables, the system can treat them as empty holes with a corresponding name).
 
-The preservation theorem establishes that transitions preserve type
-assignment, i.e. that the type of an expression accurately predicts
-the type of the result of reducing that expression.
+The preservation theorem establishes that type assignment is preserved by transitions, i.e. that the type of an expression accurately predicts the type of the result of evaluating that expression.
 
@@
 \noindent
-%
-The proof relies on an analogous preservation lemma for instruction
-transitions and a standard substitution lemma (stated in
-Appendix~\ref{sec:additional-defns}).
-%
-Hole closures can disappear during evaluation, so we rely on weakening
-of $\Delta$.
+The proof relies on an analagous preservation lemma for instruction transitions and a standard substitution lemma (stated in Appendix~\ref{sec:additional-defns}). 
+Hole closures can disappear during evaluation, so we rely on weakening of $\Delta$.
 
-The progress theorem establishes that the dynamic semantics accounts
-for every well-typed term, i.e. that we have not forgotten some
-necessary rules or premises.
-%
+The progress theorem establishes that the dynamic semantics accounts for every well-typed term, i.e. that we have not forgotten some necessary rules or premises.
 \begin{thm}[Progress]
@@
 \noindent
-The key to establishing the progress theorem under a non-empty hole
-context consists of explicitly accounting for indeterminate forms,
-i.e. those rooted at either a hole closure or a failed cast.
-%
-The proof relies on canonical forms lemmas stated in
-Appendix~\ref{sec:additional-defns}.
+The key to being able to establish the progress theorem under a non-empty hole context lies in the fact that we have explicitly accounted for indeterminate forms, i.e.
+those rooted at either a hole closure or a failed cast.
+The proof relies on canonical forms lemmas stated in Appendix~\ref{sec:additional-defns}.
 
 \subsubsection{Complete Programs} 
-%
-Although this paper focuses running \emph{incomplete} programs, it helps
-to know that its machinery does not interfere
-with reasoning about the behavior of \emph{complete} programs, i.e. those
-with no holes.
-%
-Appendix~\ref{sec:additional-defns} defines complete programs via judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
-%
+Although our focus in this paper is on running incomplete programs, it is helpful to know that the necessary machinery does not interfere with reasoning about the behavior of complete programs, i.e.
+those with no holes.
+Appendix~\ref{sec:additional-defns} defines complete programs with the judgements $\isComplete{e}$ and $\isComplete{d}$. 
 Failed casts cannot appear in complete internal expressions. 
-%
 The following theorem establishes that expansion preserves program completeness.
@@
 
-The following progress theorem establishes that evaluating a complete
-program always results in classic values, not boxed values nor
-indeterminate forms.
-%
+The following progress theorem establishes that only classic values result from evaluating a complete program. There are no boxed values or indeterminate forms. 
 \begin{thm}[Complete Progress]
@@
 
-The supplemental material includes our Agda-based
-mechanization  \cite{norell2009dependently,norell:thesis} 
-of the semantics and metatheory of \HazelnutLive, 
-%including proofs of all of the 
-including the theorems stated above, and all necessary lemmas\todo{make this so ian!}{}. 
-%
-%We choose the 
-%(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
-%
-%Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation. 
-%
-%Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
-%
-%
-Our approach is standard: We model judgements as 
-inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite something?}{}. 
-%
+The supplemental material includes our mechanization of the semantics and metatheory of \HazelnutLive, 
+including proofs of all of the theorems stated above and all necessary lemmas\todo{make this so ian!}{}. We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis} 
+(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
+Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation.
+Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf.
+Sec.~\ref{sec:intro}).
+The basic approach is standard: we model judgements as 
+inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite somtehing?}{}. 
 We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
@@
 
-The supplemental material includes a browser-based implementation
-of \HazelnutLive, which consists of a functional reactive program
-%
-% Functional Reactive Animation (Hudak 1997) is the canonical FRP cite, not Elm, if you want one; we don't need one, I think. --Matt
-%
-%\cite{DBLP:conf/pldi/CzaplickiC13} 
-%
-written with the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
+The supplemental material also includes a browser-based implementation of \HazelnutLive. 
+The implementation is a functional reactive program \cite{DBLP:conf/pldi/CzaplickiC13} 
+written using the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
 together with the OCaml \lismall{React} library \cite{OcamlReact} 
 and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}. 
-%
-Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{} provides a screenshot of the user interface, 
-which implements the core calculus of this section, 
-along with the corresponding live programming features from \Secref{sec:examples}.
+Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{} provides a screenshot of the user interface, which implements all of the live programming features described in Sec.~\ref{sec:examples}, but for the more austere language of this section.
 
-To facilitate interactively editing terms, the implementation employs
-the editor semantics of~\Hazelnut, exposing a language of structured
-edit actions (summarized in the sidebar on the left of
-Fig.~\ref{fig:screenshot})
-%
-As with prior work, this editor semantics enjoys
-the \Property{Sensibility} property by automatically inserting holes
-to guarantee that every edit state has some (possibly incomplete) type
-.\todo{put statement in the appendix}{}
-
-By composing this meta-theoretic property with
-the \Property{Expandability}, \Property{Typed Expansion}
-and \Property{Progress} properties from this paper, we establish a
-uniquely powerful invariant: \emph{Every} edit state has a \emph{static
-meaning} (a la~\Hazelnut) and a \emph{dynamic meaning} (a la~\HazelnutLive).
-%
-In other words, live feedback never ``flickers in and out'' nor reflects ``stale'' edit states.
-%
-The current implementation serves as a proof-of-concept of this invariant, 
-by providing a working prototype for this section's definitions.
-
-%, and provides to be of use to researchers studying the calculus as presented in this section.
+The editor component of the implementation is derived from \Hazelnut. 
+It exposes a language of structured edit actions (summarized in the sidebar on the left of Fig.~\ref{fig:screenshot}) that insert holes automatically to guarantee that every edit state has some (possibly incomplete) type.
+Formally,
+this corresponds to the Sensibility property from the prior work \cite{popl-paper}.\todo{put statement in the appendix}{} 
+By composing this property with the Expandability, Typed Expansion and Progress properties from this paper, we establish a uniquely powerful 
+invariant: that every edit state is both statically and dynamically meaningful. 
+In other words, live feedback does not ``flicker in and out'' or reflect past edit states. 
+The primary purpose of the included implementation is to serve as a proof-of-concept of this invariant, and to be of use to researchers studying the calculus as presented in this section.
 % Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+We are also separately integrating the mechanisms developed in this paper into the \Hazel programming environment being developed by \citet{HazelnutSNAPL} with the goal of realizing a full-scale hole-driven development experience rooted in the type theoretic foundations established by \Hazelnut.
+Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper.
+For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
+We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
+\todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
+The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}
 
-%% \matt{I'd drop everything else in this sub-section; 
-%% it leaves us open to attack from a hostile reviewer 
-%% who is cranky that the implementation is not ``complete'' yet; also, we'd save the space}
+Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}
 
-%% Separately, we are integrating the mechanisms developed in this paper
-%% into the \Hazel programming environment being developed
-%% by \citet{HazelnutSNAPL} with the goal of realizing a full-scale
-%% hole-driven development experience rooted in the type theoretic
-%% foundations established by \Hazelnut.
-%% %
-%% Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper.
-%% For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
-%% We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
-%% \todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
-%% The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}
-
-%% Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}
+\todo{integrate this}{}
+In the more common situation of a text editor, there are several alternative approaches,  
+reviewed in more detail by \citet{HazelnutSNAPL}. 
+Briefly, 
+(1) holes can be inserted implicitly using error recovery heuristics, like those used in production-grade parsers today \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09}; or
+(2) holes can be explicit as part of the surface syntax of the language as supported 
+by an extension to the Glasgow Haskell Compiler \cite{GHCWIKI}, and the dependently typed proof assistants Idris~\cite{brady2013idris} and Agda~\cite{norell2009dependently}. 
+When holes are explicitly part of the surface syntax, recent work by \cite{Amorim2016} shows how an editor can insert them in a semi-automatic manner as a form of code completion.
 
@@
 In any case,
-
-When the programmer interactively fills a program hole, that program's
-reduction and result are generally each affected, and the system
-should respond by (quickly) updating the program's output.
-%
-Na\"ively, this response requires recomputing the program output
-``from scratch'' after each such edit.
-%
-However, by virtue of its design principles (based on
-CMTT), \HazelnutLive offers a more efficient approach: To \emph{resume
-evaluation} of previously indeterminate expressions, which are
-(generally) more determined after the programmer fills a hole than
-beforehand.
-%
-We call this feature \emph{fill-and-resume}, by analogy to the
-``edit-and-resume'' features available in some systems today, such as
+\todo{cite tanimoto paper, mention Level 4 liveness property}When the programmer performs one or more edit actions to fill in a hole in the program, a new result must be computed.
+Na\"ively, the system would need to compute the result ``from scratch'' on each such edit.
+A more efficient approach supported by the structure of \HazelnutLive is to resume evaluation from where it left off after an edit that amounts to hole filling.
+We call this feature \emph{fill-and-resume}, by analogy to the ``edit-and-resume'' features available in some systems, e.g.
 Visual Studio (see below for a comparison)\todo{cite and compare}{}.
 
-%%%%%%%%% New par
-Formally, 
-%
-we interpret the hole environments as \emph{delayed substitutions},
-the same interpretation suggested for closures in contextual modal
-type theory (CMTT) by \citet{Nanevski2008}.
-%
-%In practice, it may be useful to cache results from several previous
-%expansions, e.g., by employing off-the-shelf programming language
-%abstractions for incremental computation~\cite{Hammer14,Hammer15}.
-%
-We define hole filling in terms of the expanded (internal) program,
-and define filling a hole in an external language term as filling the
-corresponding hole in its expansion.
-
-%%%%%%%%% New par
-\Figref{fig:substitution} defines the hole filling operation~$\instantiate{d}{u}{d'}$,
-mirroring the contextual substitution operation of~CMTT.
-%
-Unlike usual notions of capture-avoiding substitution, 
-hole filling imposes no condition on the binder when passing into the
-body of a lambda expression---the expression that fills a hole can, of
-course, refer to variables in scope where the hole appears.
-%
-When hole filling encounters an empty closure for the hole being
-instantiated, $\instantiate{d}{u}{\dehole{u}{\sigma}{}}$, the result
-is $[\instantiate{d}{u}{\sigma}]d$.
-%
-That is, we apply the delayed substitution to the fill expression~$d$
-after first recursively filling any instances of hole~$u$ in~$\sigma$.
-%
-Hole filling for non-empty closures is analogous, where it discards
-the previously-enveloped expression.
-%
-%
-% \matt{Will any reader actually wonder this? Does this thought connect to any other statement elsewhere in the paper?}
-%
-%This case shows why we cannot interpret a non-empty hole as an empty
-%hole of arrow type applied to the enveloped expression---the hole
-%filling operation would not operate as expected under this
-%interpretation.
+\todo{fix definition of instantiation in the agda}The key is to interpret the hole environments as delayed substitutions, and indeed this is exactly the interpretation suggested for closures in contextual modal type theory by \citet{Nanevski2008}.
+We can derive the hole filling operation, $\instantiate{d}{u}{d'}$, defined in Fig.~\ref{fig:substitution}, from the contextual substitution operation of CMTT.
+Notice that unlike standard substitution, which is capture-avoiding, hole filling imposes no condition on the binder when passing into the body of a lambda expression---the expression that fills a hole can, of course, refer to variables in scope where the hole appears.
+When hole filling encounters an empty closure for the hole being instantiated, $\instantiate{d}{u}{\dehole{u}{\sigma}{}}$, the result is $[\instantiate{d}{u}{\sigma}]d$, i.e.
+the delayed substitution is applied to the fill expression, $d$, after first recursively filling any instances of hole $u$ that may appear in $\sigma$.
+Hole filling for non-empty closures is analagous (the enveloped expression can be discarded).
+Note that this is the reason why we cannot interpret a non-empty hole as an empty hole of arrow type applied to the enveloped expression---the hole filling operation would not operate as expected under this interpretation.
 
@@
 
-%%%%%%%%% New par
 The following theorem characterizes the static behavior of hole filling.
@@
   then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
+\end{thm}
 \begin{proof}
-We prove a stronger version of the theorem which accounts for both
-terms and finite substitutions (see \cite{Nanevski2008}).
-%
-Each case proceeds by rule induction on the first assumption, applying
-the finite substitution lemma.
+We prove a stronger version of the theorem which accounts for both terms and finite substitutions (see \cite{Nanevski2008}).
+In each case, we proceed by rule induction on the first assumption, appealing to the finite substitution lemma in the corresponding cases.
 \end{proof}
-\end{thm}
 
-Dynamically, the fill-and-resume enjoys an
-important \emph{commutativity} property: If there is some sequence of
-steps that go from $d_1$ to $d_2$, then one can fill a hole in these
-terms at \emph{either} the beginning or at the end of that step
-sequence.
-%
-We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
-transitive closure of stepping (see \Figref{fig:multi-step} in
-Appendix~\ref{sec:additional-defns}).
-%
+For the fill-and-resume feature to be reasonable, we need the following commutativity property, where we write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive, transitive closure of  stepping (see Fig.~\ref{fig:multi-step} in Appendix~\ref{sec:additional-defns}).
+In words, if there is some sequence of steps that go from $d_1$ to $d_2$, then you can fill the hole either at the beginning or at the end of that step sequence because if you do the former, you can take some sequence of steps to the latter.
 \begin{thm}[Commutativity]
@@
 \end{thm}
-%
-Critically, this property relies on the more permissive version of
-reduction from \Secref{sec:calculus}, where evaluation order is
-unspecified (i.e. formally, we omit the bracketed premises).
-%
-In general, resuming from $\instantiate{\dexp'}{u}{d_2}$ will not
-reduce sub-expressions in the same order as a ``fresh'' left-to-right
-reduction sequence starting from $\instantiate{\dexp'}{u}{d_1}$.
-%
-In other words, this notion of fill-and-resume only works for
-languages where evaluation order always results in \emph{confluent}
-reductions; viz., those that \emph{eventually} yield the same
-resulting term.
-
-%% Everyone knows what confluence means, or they should.  We don't need to cite Church :)
-%%
-%There are various ways to encode this intuition. One way to do so is
-%by establishing a confluence property (which is closely related to
-%the Church-Rosser property \cite{church1936some}).
-
-The most general confluence property does not hold for the dynamic
-semantics in Sec.~\ref{sec:calculus} for the usual reason: We do not
-reduce under binders (\citet{DBLP:conf/birthday/BlancLM05} discuss the
-standard counterexample).
-%
-We could recover confluence by specifying reduction under binders,
-either generally or in a more restricted form where only closed
-sub-expressions are
-reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit}.
-%
-However, reduction under binders conflicts with the standard implementation approaches
-for most programming languages \cite{DBLP:conf/birthday/BlancLM05}.
-%
-A more satisfying approach considers confluence modulo equality \cite{Huet:1980ng}. 
-%
-The simplest such approach restricts our interest to top-level expressions
-of base type that result in values, in which case the following
-special case of confluence does hold (trivially when the only base
-type has a single value, but also more generally for other base
-types).
+Note that this property follows because the specification in Sec.~\ref{sec:calculus} left evaluation order unspecified (i.e.
+formally, we omitted the bracketed premises).
+It is not the case in general that resuming from $\instantiate{\dexp'}{u}{d_2}$ will cause sub-expressions to be evaluated in the same order as they would have been if we had restarted from $\instantiate{\dexp'}{u}{d_1}$ and selected sub-expressions in, for example, an eager, left-to-right manner, because the hole that is being filled may have been ``skipped past''.
+In other words, this simple fill-and-resume protocol only works for languages where evaluation order ``does not matter''.
+There are various ways to encode this intuition.
+One way to do so is by establishing a confluence property (which is closely related to the Church-Rosser property \cite{church1936some}).
+The most general confluence property does not hold for the dynamic semantics in Sec.~\ref{sec:calculus} for the usual reason: we do not reduce under binders (\citet{DBLP:conf/birthday/BlancLM05} discuss the standard counterexample).
+We could recover confluence by specifying reduction under binders, either generally or in a more restricted form where only closed sub-expressions are reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit}, but reduction under binders is at odds with the standard approaches used in implementing programming languages \cite{DBLP:conf/birthday/BlancLM05}.
+A more satisfying approach is to consider confluence modulo equality \cite{Huet:1980ng}.
+The simplest approach is to restrict our interest to top-level expressions of base type that result in values, in which case the following special case of confluence does hold (trivially when the only base type has a single value, but also more generally for other base types).
 \begin{lem}[Base Confluence]
@@
 
-We outline the proofs in Appendix~\ref{sec:hole-filling}, along with a
-number of other necessary lemmas and definitions.
-%
-In particular, additional machinery handles the situation wherein a
-now-filled non-empty hole had taken a step in the original evaluation.
-\matt{Cyrus --- I tried to simplify and shorten the earlier version of the prior sentence, but I \emph{may} have unintentionally goofed up what it was trying to say; please double check the prior sentence}
+The proofs are outlined in Appendix~\ref{sec:hole-filling}, along with a number of other necessary lemmas and definitions.
+In particular, we need additional machinery to properly handle the situation where we have filled a non-empty hole where a step had taken place in the original evaluation.
+These proofs and definitions are not included in the mechanization for two reasons.
+First, fill-and-resume is an optimization, and as just discussed, these properties, unlike those in Sec.~\ref{sec:calculus}, would not be conserved by some reasonable extensions of the core calculus.
+Second, it would require a significantly more complex representation of hole environments to encode the hole filling operation.
+Under the simple representation that suffices for the developments in Sec.~\ref{sec:calculus}, Agda cannot be convinced that the definition is well-founded (\citet{Nanevski2008} establish that it is in fact well-founded).
 
-\matt{We could (should?) move the next paragraph to the appendix, 
-before/after the proofs mentioned above; 
-its not really adding much to the discussion above, 
-but may be interesting to the very studious reader/prover; also, leaving it here also gives a lazy/cranky reviewer a nice weapon to hit us with}
+Hole filling in the external language is understood in terms of hole filling in its expansion.
+Note that in practice, it may be useful to cache results from several previous expansions.
+For example, consider two edits, the first filling a hole $u$ with the number $2$, and the next applying the $+$ operator, resulting in $2 + \dehole{v}{\sigma}{}$.
+This second edit is not a hole filling edit with respect to the immediately preceding edit state, but it can be understood as filling hole $u$ with $2 + \dehole{v}{\sigma}{}$.
+Another related case arises in lab notebook interfaces like that of Jupyter/IPython \cite{PER-GRA:2007}.
+In this case, each cell can be understood as a series of \li{let} bindings ending in a hole, which is filled by the next cell.
+The result from each cell can be cached to avoid recomputing the environment from preceding cells.
+This contextual modal interpretation of lab notebook cells (and read-eval-print loops as a restricted case where edits to previous cells are impossible) is novel.
 
-We exclude these proofs and definitions from the Agda mechanization
-for two reasons.
-%
-First, fill-and-resume is merely an optimization, and unlike the meta
-theory of \Secref{sec:calculus}, these properties are generally not
-conserved by certain reasonable extensions of the core
-calculus~(e.g., \matt{such as what?}).
-%
-Second, to properly encode the hole filling operation, such a
-mechanization requires a significantly more complex representation of
-hole environments; unfortunately, Agda cannot be easily convinced that
-the corresponding definition is well-founded (\citet{Nanevski2008}
-establish that it is in fact well-founded).
-%
-By contrast, the developments in \Secref{sec:calculus} do not require
-these more complex (and somewhat problematic) representations.
+So far, we have understood hole filling as an extralinguistic operation.
+However, a key feature CMTT that we have not yet touched on is the \emph{internalization} of metavariable binding and contextual substitution via the contextual modal types, $[\hGamma]\htau$, which are introduced by the operation $\mathsf{box}(\hGamma.d)$ and eliminated by the operation $\mathsf{letbox}(d_1, u.d_2)$.
+A hole filling can be interpreted as an expression of contextual modal type, and the act of hole filling followed by evaluation to the next possibly-indeterminate edit state as evaluation under the binder of a suitable $\mathsf{letbox}$ construct, which is enabled by the dynamic semantics in Sec.~\ref{sec:calculus}.
+This interpretation allows us to \emph{compute} hole fillings, rather than simply stating them, by specifying non-trivial expressions of modal type.
+This could serve as the basis for a \emph{live} computational hole refinement system, extending the capabilities of purely static hole refinement systems like those available in some proof assistants, e.g.
+the elaborator reflection system of Idris \cite{brady2013idris,DBLP:conf/icfp/ChristiansenB16} and the tactic system\todo{right?
+read a bit more}{} of Beluga \cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
+Each applied hole filling serves as a boundary between dynamic \emph{edit stages}.
+This contextual modal interpretation of live staged hole refinement nicely mirrors the modal interpretation of staged argument evaluation \cite{Davies:2001op}.
+We leave further development of this interpretation of hole refinement to future work.
 

### 2018-04-16 intro.tex
near-final intro

--- 2018-02-08 intro.tex
@@
 % !TEX root = hazelnut-dynamics.tex
+\vspace{-7px}
 \newcommand{\introSec}{Introduction}
@@
 \label{sec:intro}
-\cy{intro from LIVE 2017 paper is below}
+% \vspace{-2px}
 
-Broadly speaking, live programming environments are those that granularly interleave editing and evaluation \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90,McDirmid:2007:LUL:1297027.1297073,burckhardt2013s}.
-In the words of \citet{burckhardt2013s}, live programming environments 
-``promise to narrow the temporal and perceptive gap 
-between program development and code execution''. Examples of live programming environments include {lab notebook environments},
-e.g. the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
-programmer to interactively edit and evaluate program fragments organized into a
-sequence of cells (an extension of the read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007:LUL:1297027.1297073}, Sketch-n-Sketch \cite{DBLP:conf/pldi/ChughHSA16} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples.
+% Programming environments often operate in ``batch mode'', assuming that the programmer will spend a substantial amount 
+% of time editing the program text blindly before evaluating (i.e. running) the program.
 
-The problem that has motivated much of our recent work is that most  
-programming environments, live programming environments included,  provide feedback via various editor services only once the program being edited is syntactically well-formed and, when relevant, well-typed.
-This leaves a ``temporal and perceptive gap'', because programmers sometimes leave a program malformed or ill-typed for extended periods of time, e.g.
-as they think about what to enter at the cursor, or as they work on a different part of the program.
+Programmers typically shift back and forth between program editing and evaluation many times before converging upon a program that behaves correctly. 
+Live programming environments support this workflow by interleaving editing and evaluation so as to   
+narrow what \citet{burckhardt2013s} call ``the temporal and perceptive gap'' between these activities.
+% In other words, the goal is to provide continuous feedback about the dynamic behavior of the program, in whole or in part,
+% directly alongside the program text itself.
 
-In view of this general problem, we recently developed a \emph{structure editor calculus} called Hazelnut where every edit state consists of a well-formed and statically meaningful, i.e.
-well-typed, incomplete expression, which we take to mean an expression with holes \cite{popl-paper}.
-This calculus addressed fundamental questions relevant to editor services that operate statically, but there was no solution in that paper to the problems faced by editor services that also require knowledge of the dynamic meaning of an incomplete program, as would be relevant to a live programming environment.
-For example, consider a \emph{stepper} (a.k.a.
-a \emph{stepwise debugger}), like that available to Haskell programmers in the GHCi system \cite{GHC-stepper} and other systems \cite{DBLP:conf/haskell/MarlowIPG07,kar13566}, to OCaml programmers with recent work on \texttt{ocamli} by \citet{ocaml-stepper} and to Standard ML programmers \cite{DBLP:journals/jfp/TolmachA95}.
-A stepper requires that the expression being stepped be assigned dynamic meaning according to a small-step operational semantics \cite{DBLP:journals/jlp/Plotkin04a,pfpl}, but no such semantics was defined for incomplete expressions that arise when using Hazelnut, or any of these other systems.
-Defining such a semantics was left as future work in the Hazelnut paper, and in a subsequent ``vision paper'' \cite{snapl17-paper}.
-The purpose of this paper is to sketch out our progress toward a theoretically well-grounded solution.
+For example, read-evaluate-print loops (REPLs) and derivatives thereof, like the IPython/Jupyter lab notebooks popular in data science~\cite{PER-GRA:2007}, allow the programmer to edit and immediately execute program fragments organized into a sequence of cells.
+Spreadsheets are live functional dataflow environments, with cells organized into a grid \cite{DBLP:journals/jfp/Wakeling07}. 
+More specialized examples include live direct manipulation programming environments like SuperGlue
+\cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist}, and the tools
+demonstrated by \citet{victor2012inventing} in his lectures;
+%
+live user interface frameworks \cite{burckhardt2013s};
+%
+live image processing languages~\cite{DBLP:journals/vlc/Tanimoto90};
+%
+and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding in an artistic performance.
+Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support editing the state of a running program, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
+Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13}, has been and continues to be an active area of research and development.
 
-\matt{below are misc points that may be worth mentioning in section 1
-  (somehow, somewhere), and weaving into the continuation of the story
-  above:}
+The problem that specifically motivates this paper is that programming languages typically assign dynamic meaning only to {complete programs}, i.e.
+programs that are syntactically well-formed and free of static type and binding errors.
+A program editor, however, frequently encounters incomplete, and therefore meaningless, editor states.
+As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e.
+it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer is actually seeing and writing in the editor.\todo{Could put the sentence about ``this is when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}
 
-\textbf{Type holes vs expression holes:}
+In some cases these gaps are momentary, like while the programmer is entering a short expression.
+In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, or when the programmer makes a mistake (so they arise somewhat more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}).
+The problem is particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
+Throughout the process of addressing these errors, the program text remains formally meaningless.
+Overall, an analysis of edits performed by Java programmers using Eclipse found that about 40\% of edit states were malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.
 
-The dynamics of type holes is gradual typing, including run-time cast
-insertion and execution.
-%
-The dynamics of expression holes consists of defining
-\emph{indeterminate} expressions, as placeholders for subterms that
-traditionally would be deemed ``stuck'' .
+ % From the perspective of the language definition, these edit states are wholly meaningless, so editor services cannot rely on the same operations and reasoning principles that would be available to, for example, the compiler.
 
-In live programming, we expect that programs will generally have holes
-in both terms and types (CITE?); however, to understand the mechanisms
-of \HazelnutLive, it is instructive to imagine situations where the
-only holes are in terms, or are in types.
-%
-When we only use expression holes and no type holes, we do not need
-type casts, or the other dynamic mechanisms of gradual typing.
-% 
-When we only use type holes, we recover gradual typing and do not use
-indeterminate forms until and unless cast errors arise.
+% Despite these advances, there is a fundamental limitation which hinders
+% the full realization of live programming: traditional programming languages
+% cannot evaluate \emph{incomplete programs}, \ie{}, programs that are either
+% syntactically ill-formed, or ill-typed.
+% %
+% Without a well-defined semantics to dictate the treatment of incomplete
+% programs, various editor services may be provided in ad-hoc, restricted
+% forms---if available at all---during the periods of time between complete
+% program states.
+% %
+% %% ``flickering in and out''
+% %
+% %% In other words, the ``Edit-Compile-Run'' cycle routinely fails (either in the
+% %% first or second steps) during the program editing process.
+% %
+% The resulting editor experience includes services ``flickering in and out,''
+% thus introducing ``temporal and perceptive gaps'' because programmers often
+% leave a program malformed or ill-typed for extended periods of time, \eg{}~as
+% they think about what to enter at the cursor, or as they work on a different
+% part of the program.
+% %
+% These times are, arguably, when rich, live editor services would be most helpful
+% to the user.
 
-\textbf{Cast errors as indeterminate forms:}
+In recognition of this problem---that incomplete programs lack meaning---%
+\citet{popl-paper} developed a static semantics (i.e. a type system) for incomplete 
+functional programs, modeled formally as typed expressions with \emph{holes} in 
+both expression and type position. 
+Empty holes stand for missing expressions or types,
+and non-empty holes operate as ``membranes'' around static type inconsistencies 
+(i.e. they internalize the ``red underline'' that editors commonly display under a type inconsistency).
+% For editor states into this language of incomplete programs, 
+% editor services can reason about types and binding in many more situations than previously possible.
+\citet{popl-paper,HazelnutSNAPL} discuss several ways to determine an incomplete expression from an editor state. 
+Briefly, error recovery mechanisms can insert the necessary holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09,DBLP:conf/oopsla/KatsV10},
+the programmer can insert explicitly holes manually (see below) 
+or semi-automatically via code completion techniques \cite{Amorim2016}, 
+or a structure editor can insert holes fully automatically \cite{popl-paper} (we say more about this technique in Sec.~\ref{sec:implementation}).
 
-Instead of being ``stuck'', cast errors in \HazelnutLive are
-fully-reduced indeterminate terms.
+For the purposes of live programming, however, a static semantics for incomplete programs does not suffice---%
+we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes. 
+Our goal in this paper is to develop just such a dynamic semantics for incomplete functional programs, 
+based on the static semantics just mentioned.
 
-\textbf{Indeterminate forms as exceptions:}
+One approach would be for evaluation to simply stop with an error when it encounters an expression hole.
+This mirrors a workaround that programmers commonly deploy: 
+raising an exception when the intended expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
+Some languages provide more concise notation for this interpretation of expression holes, 
+e.g. Scala uses \lismall{???} for missing method bodies 
+and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u} where \lismall{u} is an optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
+\footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
+This is also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}
 
-TODO --- transfer thoughts from 2016 grant writing
+Although better than nothing, this approach of stopping with an exception upon reaching an expression hole has limitations 
+as the basis for a live programming environment because 
+(1)~it provides limited information about the dynamic state of the program when the exception occurs
+(typically only a stack trace);  
+(2)~it provides no information about the behavior of the remainder of program, 
+parts of which may not depend at all on the missing or erroneous expression; and 
+(3)~it provides no means by which to resume evaluation after filling the hole.
 
-\textbf{Holes as breakpoints:}
+Moreover, we need to be able to run programs containing not only expression holes but also type holes.
+In the static semantics developed by \citet{popl-paper}, the machinery for 
+reasoning statically about type holes is derived from gradual type theory, 
+identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}. 
+To specify our intended dynamic semantics for incomplete programs, then, we can look to the dynamic semantics from
+gradual type theory, which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
+However, this is again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
+an exception. As a result, the live programming environment is left unable to provide feedback about even parts 
+of the program that do not depend on the problematic sub-expression (e.g. later cells in a lab notebook).
 
-\begin{enumerate}
+\parahead{Contributions}
 
-\item 
-  \texttt{map somefun list}
+This paper develops a theoretically well-grounded dynamic semantics for incomplete functional programs that addresses the limitations of the ``exceptional approach'' just described. 
+In particular, rather than stopping with an exception when evaluation encounters an expression hole, evaluation continues ``around'' the hole, performing as much of the remaining computation as possible.
+Evaluation also proceeds past failed casts in much the same way. 
+The system tracks the environment around each hole instance as evaluation proceeds. 
+The live programming environment can feed relevant information from these hole environments to the programmer as they work to fill the holes in the program. 
+Then, when the programmer performs an edit that fills an empty expression hole or that replaces a non-empty hole with a type-correct expression, evaluation can resume from the paused, i.e.
+\emph{indeterminate}, evaluation state.
 
-Suppose the programmer wishes to map with \texttt{somefun} that they
-either did not author themselves, or they forgot how the function
-works:
+\parahead{Paper Outline}
 
-\item 
-  \textt{map ($\lambda \texttt{x}. \hhole{\texttt{x}}{a}$) list}
+\newcommand{\contribution}[2]{\paragraph{#1. #2}}
 
-First, the programmer can intercept the function's inputs.  These
-hole~$a$ above functions like a ``breakpoint'', and shows the
-programmer me all of the inputs to \texttt{somefun}.
+We are integrating this approach into the \Hazel programming environment being developed independently of this paper by \citet{HazelnutSNAPL}. 
+We begin in Sec.~\ref{sec:examples} by detailing the approach informally, with several examples, in the setting of this specific design.
 
-TODO -- show the result of running above.
+Sec.~\ref{sec:calculus} then abstracts away orthogonal details and makes these intuitions formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions as just outlined, in a manner that admits clean type safety theorems (in particular, a Progress theorem).
+The semantics of \HazelnutLive borrows machinery from gradual type theory~\cite{DBLP:conf/snapl/SiekVCB15,Siek06a}
+(to handle incomplete types) and contextual modal type theory~\cite{Nanevski2008} (CMTT,
+which provides a logical foundation for hole environment tracking). 
+These non-obvious connections to well-established existing systems help support our claim that this approach is theoretically well-grounded\todo{note that our exposition is self-contained here?
+or later?}.
 
-\item
-  \textt{map ($\lambda \texttt{x}. \hhole{\texttt{somefun x}}{a}$) list}
+Sec.~\ref{sec:agda-mechanization} outlines our mechanization of the core calculus using the Agda proof assistant. 
+Sec.~\ref{sec:implementation} outlines our simple implementation of the core calculus, which implements the live programming features from Sec.~\ref{sec:examples},
+but for the more austere language of Sec.~\ref{sec:calculus}. 
+The editor component of this implementation provides a language of structured edit actions, 
+based on the \Hazelnut structure editor calculus developed by \cite{popl-paper}, that guarantees that
+every editor state has some, possibly incomplete, type. 
+Together with the theorems established earlier in Sec.~\ref{sec:calculus}, the result is a proof-of-concept
+live functional programming environment where every possible editor state has both non-trivial static and dynamic 
+meaning, i.e. live feedback never ``flickers'' or ``goes out''. 
+The mechanization and the implementation are both available in the supplemental material.
 
-  Next, suppose the programmer wishes to see all of the the outputs of
-  \texttt{somefun}.  To do so, they put that (single static)
-  expression into a hole, as above.
+Sec.~\ref{sec:resumption} defines the fill-and-resume feature, which is rooted in the contextual substitution operation from CMTT.
+We establish the correctness of fill-and-resume with a novel commutativity property and discuss how it provides a semantic interpretation for cells in a live lab notebook environment.
 
-\item
-  \textt{map ($\lambda \texttt{x}. \hhole{(x, \texttt{somefun x})}{a}$) list}
-
-  Finally, the programmer can see input-output pairs without breaking
-  the typing derivation of their program globally, since holes are
-  permitted to go \emph{anywhere}.  (They may have also exploited this
-  fact to show the inputs, in the first step).
-
-\end{enumerate}
-
-\cy{intro from grant section is below}
-
-Live programming environments granularly interleave editing and evaluation,
-``promis[ing] to narrow the temporal and perceptive gap 
-between program development and code execution''~\cite{burckhardt2013s}.
-Examples of live programming environments include {lab notebook environments},
-like the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
-programmer to interactively edit and evaluate program fragments organized into a
-sequence of cells (an extension of the ubiquitous read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples.
-\cy{find somewhere to cite these tanimoto papers \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90}}
-
-Live programming environments typically cannot run incomplete programs, or they simply halt or raise an exception as soon as they encounter a hole during evaluation.
-This leaves a ``temporal and perceptive gap'' because the results of even those computations that do not depend critically on the hole are not available to the programmer.
-Our second track of research will address this gap by defining a dynamic semantics for incomplete programs that proceeds as far as possible around holes.
-Every program assigned static meaning by the semantics of Track 1 will also be assigned dynamic meaning by the semantics of Track 2.
-
-For example, in \autoref{fig:intro-example} we saw that mapping the incomplete function \li{weighted_average} over the \li{grades} list produced an incomplete result.
-If holes simply raised exceptions, then the system would have aborted evaluation as soon as \li{weighted_average} was applied for the first time by \li{map}, and the programmer would not have been able to see that the result was a list of two elements, nor see the value of the sub-expression \li{10.0 *.
-g.hw1} for each element \li{g} of the \li{grades} list.
-Even programs with type inconsistencies can be evaluated under our proposed dynamics, because, as discussed in Track 1, non-empty holes operate as membranes around type inconsistencies.
-Taken together, the results from Tracks 1 and 2 may help to address a common complaint: that a static type discipline makes it difficult to do exploratory programming.
-
-Our proposed dynamics goes beyond staged evaluation (\eg{}~\cite{Taha:1999}), partial evaluation (\eg{}~\cite{Jones:1993uq}) and symbolic evaluation (\eg{}~\cite{King:1976}) in that we track the dynamic environment (\ie{}~the substitutions performed) around each instance of an expression hole in the result.
-This allows the programmer to see, in the sidebar of \autoref{fig:intro-example}, the actual values that the variables in scope take on everywhere that the hole they are filling ends up during evaluation.
-This provides a hole-oriented specification of a standard feature of debuggers: inspection of the environment at designated points in the program.
-Finally, this environmental information is useful (though not required) for the edit action suggestion service of Track 3 and the direct manipulation programming service of Track 4.
+Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneous discuss various directions for future work.
+Sec.~\ref{sec:discussion} briefly concludes.
+The appendix includes some straightforward auxiliary definitions and proofs mentioned in the paper, some extensions to the core calculus, and some screenshots and additional details on the implementation.

### 2018-04-16 calculus.tex

--- 2018-04-15 calculus.tex
@@
 
-%% Separately, we are integrating the mechanisms developed in this paper
-%% into the \Hazel programming environment being developed
-%% by \citet{HazelnutSNAPL} with the goal of realizing a full-scale
-%% hole-driven development experience rooted in the type theoretic
-%% foundations established by \Hazelnut.
-%% %
-%% Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper.
-%% For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
-%% We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
-%% \todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
-%% The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}
-
-%% Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}
-
 %%   \halam{x}{\htau}{\evalctx} ~\vert~
@@
 In any case,
-
-When the programmer interactively fills a program hole, that program's
+% When the programmer performs one or more edit actions to fill in a hole in the program, a new result must be computed.
+Na\"ively, the system would need to compute the result ``from scratch'' on each such edit.
+A more efficient approach supported by the structure of \HazelnutLive is to resume evaluation from where it left off after an edit that amounts to hole filling.
+We call this feature \emph{fill-and-resume}, by analogy to the ``edit-and-resume'' features available in some systems, e.g.
+Visual Studio (see below for a comparison)\todo{cite and compare}{}.
+
+\todo{cite tanimoto paper, mention Level 4 liveness property}When the programmer interactively fills a program hole, that program's
 reduction and result are generally each affected, and the system

### 2018-04-16 calculus.tex
revisions through the first 2 paragraphs of 3.3

@@
 \label{sec:calculus}
+
+% \def \TirNameStyle \Vtexttt{#1}{{#1}}
+
 \input{fig-syntax}
@@
 
-We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory.
-
-The syntax of the core calculus specified in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
-We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. 
+We will now make the intuitions developed in the previous section formally precise by specifying a core calculus, which we call \HazelnutLive, and its accompanying metatheory.
+
+\parahead{Overview} The syntax of the core calculus given in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
+We distinguish between {external} expressions, $e$, and {internal} expressions, $d$. 
 External expressions correspond to programs as entered by the programmer 
-(see Sec.~\ref{sec:intro}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
+(see Sec.~\ref{sec:intro} for discussion of implicit, manual, semi-automated and fully automated hole entry methods). 
 Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
-We take this approach, notably also taken in the specification of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and
+We take this approach, notably also taken in the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and
 (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
@@
 We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}
-(see Sec.~\ref{sec:agda-mechanization}). We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).
+(see Sec.~\ref{sec:agda-mechanization}).
+Rule names, written \rulename{SVar}, correspond to the names used in the mechanization.
+We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).
 
@@
 We start with the type system of the \HazelnutLive external language,
-which closely follows the \Hazelnut type system \cite{popl-paper},
-except that (1) holes in \HazelnutLive each have a \emph{unqiue name},
-and (2) for brievity of exposition, we consider a more austere core
-language here that lacks numbers and sum types, which we discuss
-(along with other systematic extensions) in
-Appendix~\ref{sec:extensions}.
+which closely follows the \Hazelnut type system \cite{popl-paper}; we summarize the minor differences as they come up.
+
+% except that (1) holes in \HazelnutLive each have a \emph{unique name};
+% (2) for brevity of exposition, we removed numbers in favor of a simpler base type, $b$, with one value, $c$ (i.e. $b$ is the unit type); and 
+% (3) we include both unannotated lambdas, $\hlam{x}{\hexp}$, and half-annotated lambdas, $\halam{x}{\htau}{\hexp}$.  which we discuss
+% (along with other systematic extensions) in
+% Appendix~\ref{sec:extensions}.
 
@@
 An empty hole is
-written $\hehole{u}$, where $u$ is a name for the hole, which we tacitly assume is unique.
-%
-%(Holes in \Hazelnut are \emph{unnamed}).
-%
-Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
+written simply $\hehole{u}$, where $u$ is the hole name, which we tacitly assume is unique 
+(holes in \Hazelnut were not named).
+Rule \rulename{SEHole} specifies that an empty hole synthesizes hole type, written $\tehole$.
 %
 If an empty hole appears where an expression of some other type is
-expected, e.g. under an explicit ascription (governed by Rule {SAsc})
+expected, e.g. under an explicit ascription (governed by Rule \rulename{SAsc})
 or in the argument position of a function application (governed by
-Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
-Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
+Rule \rulename{SAp}, discussed below), we apply the \emph{subsumption rule},
+Rule \rulename{ASubsume}, which specifies that if an expression $e$ synthesizes
 type $\htau$, then it may be checked against any \emph{consistent}
@@
 %
-This relation stands in contrast to subtyping, which is anti-symmetric and transitive; subtyping may be integrated into a gradual type system following \citet{Siek:2007qy}).
+This stands in contrast to subtyping, which is anti-symmetric and transitive; subtyping may be integrated into a gradual type system following \citet{Siek:2007qy}.
 
@@
 %
-Rule {SNEHole} specifies that a non-empty expression hole also synthesizes hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type.
-%
-Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}\todo{briefly say something about binding inconsistencies?}
+Rule \rulename{SNEHole} specifies that a non-empty expression hole also synthesizes hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type.
+%
+Non-empty expression holes therefore internalize the ``red underline'' that many editors display under or around type inconsistencies in a program.
 
@@
 %
-For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule {SConst} (i.e. $b$ is the unit type).
+For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule \rulename{SConst} (i.e. $b$ is the unit type).
 %
 %\matt{Extraneous and uninteresting:}
-%By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.
-%
-Rule {SVar} synthesizes the corresponding type from $\hGamma$.
-
+By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other simple extensions to the calculus\todo{do this, say which ones}.
+%
+
+Rule \rulename{SVar} synthesizes the corresponding type from $\hGamma$. 
 For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
 %
-Half-annotated lambdas may appear in synthetic position according to Rule {SLam}, which is standard \cite{Chlipala:2005da}.
+Half-annotated lambdas may appear in synthetic position according to Rule \rulename{SLam}, which is standard \cite{Chlipala:2005da}.
 %
@@
 %
-To avoid the need for two separate rules, Rule {ALam} uses the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
-%
-\Secref{sec:related} dicusses how \HazelnutLive might be enriched with
-with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
-the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
+To avoid the need for two separate rules, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.\footnote{A system supporting ML-style type reconstruction \cite{damas1982principal} might also include a synthetic rule for unannotated lambdas, e.g.
+as outlined by \citet{DBLP:conf/icfp/DunfieldK13}, but we stick to this simpler ``Scala-style'' local type inference scheme in this paper \cite{Pierce:2000ve,Odersky:2001lb}.}
+%
+% \Secref{sec:related} dicusses how \HazelnutLive might be enriched with
+% with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
+% the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
 %
@@
 %
-%Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might include a synthetic rule for unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}.
+%Note that 
 %%
 
-The rule governing function application, Rule {SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.
+The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.
 
@@
 %
-\Figref{fig:expansion} defines this expansion, and \Figref{fig:hasType} defines type assignment for internal expressions.
-%
-\Secref{sec:evaluation} discusses internal expression evaluation.
+\Figref{fig:expansion} specifies expansion, and \Figref{fig:hasType} specifies type assignment for internal expressions.
+%
+% \Secref{sec:evaluation} discusses internal expression evaluation.
 
@@
 %
-The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ synthesizes a type~$\htau$ from~$\hexp$.
-%
-Synthetic expansion produces an internal expression~$d$ and a hole context~$\hDelta$.
+The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$  
+%
+produces an expansion~$d$ and a hole context~$\hDelta$ when synthesizing type $\htau$ for $\hexp$. 
 %
@@
 %
-The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks~$\hexp$ against~$\htau$ and produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$.
-%
-The following theorem establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is consistent with $\htau$.
+The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$, when checking~$\hexp$ against~$\htau$.
+%
+The following theorem establishes that expansions are well-typed and in the analytic case that the assigned type, $\htau'$, is consistent with provided type, $\htau$.
 %
@@
 %
-Analytic expansion produces an expansion of consistent type:
-%
-Intuitively, the subsumption rule permits us
+The reason that $\htau'$ is only consistent with the provided type $\htau$ is because 
+%
+ the subsumption rule permits us
 to check an external expression against any type consistent with the
-type that the expression \emph{actually} synthesizes:
+type that the expression \emph{actually} synthesizes, whereas every internal
+expression can be assigned at most one type, i.e. the following
+standard unicity property holds of the type assignment system.
 %
@@
 
-Before describing the rules in detail, let us state a few other useful theorems. 
+Before describing the rules in detail, let us state two other guiding theorems. 
 The following theorem establishes that an expansion exists for every well-typed external expression. 
+The mechanization also establishes when an expansion exists, it is unique (not shown).
  \begin{thm}[Expandability] \label{thm:expandability}~
@@
 \end{thm}
-\noindent
-The following theorem establishes that when an expansion exists, it is unique.
-\begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
-  \begin{enumerate}[nolistsep]
-    \item
-      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
-      and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
-      then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
-    \item
-      If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
-      and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
-      then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
-  \end{enumerate}
-\end{thm}
-\noindent
-The following theorem establishes that expansion generalizes external typing.\todo{rename correspondence to generality}
+% \noindent
+% The following theorem establishes that when an expansion exists, it is unique.
+% \begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
+%   % \begin{enumerate}[nolistsep]
+%   %   \item
+%       If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+%       and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
+%       then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
+%   %   \item
+%   %     If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
+%   %     and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
+%   %     then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
+%   % \end{enumerate}
+% \end{thm}
+% \noindent
+The following theorem establishes that expansion generalizes external typing.
 \begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
@@
 The rules governing expansion of constants, variables and lambda expressions
---- Rules {ESConst}, {ESVar}, {ESLam} and {EALam}
---- and the corresponding type assignment rules
---- Rules {TAConst}, {TAVar} and {TALam}
---- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}.
-%
-Consequently, the corresponding cases of
-Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
-Theorem~\ref{thm:expansion-generality} are straightforward.
+--- Rules \rulename{ESConst}, \rulename{ESVar}, \rulename{ESLam} and \rulename{EALam}
+--- mirror the corresponding type assignment rules
+--- Rules \rulename{TAConst}, \rulename{TAVar} and \rulename{TALam}
+--- mirror the bidirectional typing rules from Fig.~\ref{fig:bidirectional-typing}.
+%
+% Consequently, the corresponding cases of
+% Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
+% Theorem~\ref{thm:expansion-generality} are straightforward.
 %
 To support type assignment, all lambdas in the internal language are
-half-annotated---Rule {EALam} inserts the annotation when expanding an
+half-annotated---Rule \rulename{EALam} inserts the annotation when expanding an
 unannotated external lambda based on the given type.
 %
-Some rules perform \emph{cast insertion}, making them more interesting
-than the others; these rules include those for holes, and the those for
-function applications and ascriptions.  We consider each below.
+The rules governing hole expansion, and the rules that perform \emph{cast insertion}---those governing
+function application and type ascription---are more interesting. We consider each of these two groups of rules, in turn, below.
 
 \subsubsection{Hole Expansion}\label{sec:hole-expansion} 
-Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.
+Rules \rulename{ESEHole}, \rulename{ESNEHole}, \rulename{EAEHole} and \rulename{EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$.
 %
@@
 %
-%Note that
-While we assume each hole name to be unique in the external language, due to substitution, evaluated (internal) terms may consist of multiple hole closures with the same name.
-%
-For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}.
+While we assume each hole name to be unique in the external language, once evaluation begins, there may be multiple hole closures with the same name due to substitution.
+%
+For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}{}.
 %
@@
 
-The hole expansion rules introduce hypotheses of the form~$\Dbinding{u}{\hGamma}{\htau}$ into the hole context~$\Delta$.
-%
-For each hole name~$u$, the hole context records a type~$\tau$, and a typing context~$\Gamma$.\footnote{
+%
+For each hole, $u$, in an external expression, the hole context generated by expansion, $\Delta$, contains a hypothesis of the form~$\Dbinding{u}{\hGamma}{\htau}$, which records the hole's type, $\tau$, and the typing context, $\Gamma$, from where it appears in the original expression.\footnote{
 We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.}
@@
 %
-Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
-%
-In the synthetic hole expansion rules~{ESEHole} and~{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the typing rules.
-%
-However, the first two premises of the expansion subsumption rule~{EASubsume} disallow the use of subsumption for holes in analytic position.
-%
-Instead, we employ separate analytic rules~{EAEHole} and~{EANEHole}, which each record the checked type~$\tau$ in the hole context.
-%
-Consequently, we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.
+% Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
+%
+In the synthetic hole expansion rules~\rulename{ESEHole} and~\rulename{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the external typing rules.
+%
+However, the first two premises of the expansion subsumption rule~\rulename{EASubsume} disallow the use of subsumption for holes in analytic position.
+%
+Instead, we employ separate analytic rules~\rulename{EAEHole} and~\rulename{EANEHole}, which each record the checked type~$\tau$ in the hole context.
+%
+Consequently, we can use type assignment for the internal language --- the type assignment rules \rulename{TAEHole} and \rulename{TANEHole} in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.
 
@@
 %
-Initially, when no evaluation has yet occurred, the hole expansion rules generate an initial environment that consists of the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
+Initially, when no evaluation has yet occurred, the hole expansion rules generate the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
 %
@@
 %
-Formally, we define this relation as follows:\todo{check that the definition in the Agda corresponds}
+Formally, we define this relation in terms of type assignment as follows:
 \begin{defn}[Substitution Typing]
@@
 
-Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{}
+Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.
 %
@@
 %
-Cast insertion mediates between the type systems of the external and
-internal languages, where the former permits subsumption, and the
-latter does not.
+Holes in types require us to defer certain structural checks to run time. 
 %-----------------------------------------------------------------------------------------------
 %
-To see why this mediation is necessary, consider the following
+To see why this is necessary, consider the following
 example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
@@
 }
-\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} 
 By our interpretation of hole types as unknown types from gradual type
@@
 
-The cast form in Hazelnut Live consists of $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
+The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
 %
@@
 treatment as an expression of a consistent type $\htau_2$
-(Rule TACast in \Figref{fig:hasType}).
-%
+(Rule TACast in \Figref{fig:hasType}).%
 \footnote{
 In the earliest work on gradual type theory, the cast form only gave
-the target type~$\htau_2$ \cite{Siek06a}, but it simplifies matters
+the target type~$\htau_2$ \cite{Siek06a}, but it simplifies the dynamic semantics substantially
 to include the assigned type~$\htau_1$ in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.
@@
 
-Expansion inserts casts for function applications and ascriptions.
-%
-The latter is more straightforward: Rule~{ESAsc}
+Expansion inserts casts at function applications and ascriptions.
+%
+The latter is more straightforward: Rule~\rulename{ESAsc}
 in \Figref{fig:expandSyn} inserts a cast from the assigned type to the
@@
 
-Rule~{ESAp} expands function applications.
+Rule~\rulename{ESAp} expands function applications.
 %
@@
 %
-The first three premises of Rule~{ESAp} accomplish this
+The first three premises of Rule~\rulename{ESAp} accomplish this
 %
@@
 in a variant of this rule using synthetic expansion for the function
-expression\todo{put this in the appendix?}{}.
+expression.
 
@@
 %
-The conclusion of Rule~{ESAp} inserts the cast on the argument's
+The conclusion of Rule~\rulename{ESAp} inserts the cast on the argument's
 expansion, from the type it is assigned by the final premise of the
@@
 
-The example's second, outermost application uses through the same
+The example's second, outermost application goes through the same
 application expansion rule.
@@
 %
-However, it is safe to eliminate such identity casts during expansion,
+However, it is safe in practice to eliminate such identity casts during expansion,
 and some formal accounts of gradual typing do so by defining three
-application rules, including the original account of \cite{Siek06a}).
+application expansion rules, including the original account of \citet{Siek06a}.
 
@@
 
-Figures~\ref{fig:isGround}-\ref{fig:step} give dynamic semantics for \HazelnutLive,
-%
-resulting in a system capable of running incomplete programs without aborting when a hole is encountered.
-%
-We equip this semantics with a meaningful notion of type safety (involving both preservation and progress).
-%
-We also establish that the standard notion of type safety falls out when running complete terms.
-
-The dynamic semantics consists of a ``small-step'' transition system,
+To recap, the result of expansion is a well-typed internal expression with appropriately initialized hole closures and casts. 
+We can now specify, in Figures~\ref{fig:isGround}-\ref{fig:step}, the dynamic semantics of \HazelnutLive as a ``small-step'' transition system equipped with a meaningful notion of type safety even for incomplete programs, i.e.
+expressions typed under a non-empty hole context, $\Delta$.
+We establish that evaluation does not stop immediately when it encounters a hole, nor when a cast fails, by precisely characterizing when evaluation \emph{does} stop. 
+We also establish that an essentially standard notion of type safety holds when running complete programs.
+
+It is perhaps worth stating at the outset that a dynamic semantics equipped with these properties does not simply ``fall out'' from the observations made above that (1) empty hole closures correspond to metavariable closures from CMTT \cite{Nanevski2008} and (2) casts also arise in gradual type theory \cite{DBLP:conf/snapl/SiekVCB15}.
+%
+We say more after discussing type safety below.\todo{do this, or change to related work}{}
+%
+
+%  specify the dynamic semantics of \HazelnutLive. 
+% %
+% The dynamic semantics is capable of running incomplete programs, i.e. those with hole closures, without aborting at holes, or when a cast fails.
+% %
+
+The dynamic semantics consists of ,
 %
@@
 
-\input{fig-ground-types}
-\input{fig-dynamics-aux}
-
 The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as
-either ordinary values~(Rule~{BVVal}), or one of two cast forms: casts
+either ordinary values~(Rule~\rulename{BVVal}), or one of two cast forms: casts
 between function types (except identity casts) and casts from a ground
@@
 
-\input{fig-dynamics-contexts}
-
 The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}.
 %
-Top-level transitions are \emph{steps}~$\stepsToD{}{d}{d'}$, governed by Rule~{Step} in \Figref{fig:step}, which
+Top-level transitions are \emph{steps}~$\stepsToD{}{d}{d'}$, governed by Rule~\rulename{Step} in \Figref{fig:step}, which
 %
@@
 %
-Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the
+Rule \rulename{ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the
 standard beta reduction transition.
@@
 
+\input{fig-ground-types}
+\input{fig-dynamics-aux}
+
 The ITBeta rule is not the only rule we need to handle function
@@
 arrow types, in which case we apply the arrow cast conversion rule,
-Rule {ITApCast}, to rewrite the application form, obtaining an
+Rule \rulename{ITApCast}, to rewrite the application form, obtaining an
 equivalent application where the expression~$d_1$ under the function
@@
 %
-In this case, the application is indeterminate (Rule~{IAp}
+In this case, the application is indeterminate (Rule~\rulename{IAp}
 in \Figref{fig:isFinal}), and the application's reduction reduces no
 further.
+
+\input{fig-dynamics-contexts}
 
@@
 \subsubsection{Casts}
-Rule {ITCastId} drops identity casts. The remaining instruction
+Rule \rulename{ITCastId} drops identity casts. The remaining instruction
 transition rules assign meaning to non-trivial casts.
@@
 %
-Rule {TAFailedCast} specifies that a failed cast is well-typed exactly
+Rule \rulename{TAFailedCast} specifies that a failed cast is well-typed exactly
 when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type
@@
 %
-Rule {IFailedCast} specifies that a failed cast operates as an
+Rule \rulename{IFailedCast} specifies that a failed cast operates as an
 indeterminate form once $d$ is final.
@@
 %
-The two remaining instruction transition rules, Rule~{ITGround}
+The two remaining instruction transition rules, Rule~\rulename{ITGround}
 and~{ITExpand}, insert intermediate casts from non-ground type to a
@@
 %
-In particular, Rule {ICastHoleGround} handles casts that are not of
+In particular, Rule \rulename{ICastHoleGround} handles casts that are not of
 the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$, from hole to

### 2018-04-16 related-work.tex
revisions through the first 2 paragraphs of 3.3

@@
 
-To recap, the result of expansion is a well-typed internal expression with hole closures and casts, 
-where the former corresponds to metavariable closures from CMTT \cite{Nanevski2008}, 
-and the latter corresponds to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}. 
-However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.
-%
-For further discussion, see \Secref{sec:related}
-
 The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.

### 2018-04-16 intro.tex
typo
take this red pill, and see what I see

@@
 narrow what \citet{burckhardt2013s} call ``the temporal and perceptive gap'' between these activities.
-% In other words, the goal is to provide continuous feedback about the dynamic behavior of the program, in whole or in part,
+% In other words, the goal \IS to provide continuous feedback about the dynamic behavior of the program, in whole or in part,
 % directly alongside the program text itself.
@@
 
-The problem that specifically motivates this paper is that programming languages typically assign dynamic meaning only to {complete programs}, i.e.
+The problem that specifically motivates this paper \IS that programming languages typically assign dynamic meaning only to {complete programs}, i.e.
 programs that are syntactically well-formed and free of static type and binding errors.
@@
 As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e.
-it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer is actually seeing and writing in the editor.\todo{Could put the sentence about ``this is when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}
+it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer \IS actually seeing and writing in the editor.\todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}
 
-In some cases these gaps are momentary, like while the programmer is entering a short expression.
+In some cases these gaps are momentary, like while the programmer \IS entering a short expression.
 In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, or when the programmer makes a mistake (so they arise somewhat more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}).
-The problem is particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
+The problem \IS particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
 Throughout the process of addressing these errors, the program text remains formally meaningless.
@@
 
-% Despite these advances, there is a fundamental limitation which hinders
+% Despite these advances, there \IS a fundamental limitation which hinders
 % the full realization of live programming: traditional programming languages
@@
 we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes. 
-Our goal in this paper is to develop just such a dynamic semantics for incomplete functional programs, 
+Our goal in this paper \IS to develop just such a dynamic semantics for incomplete functional programs, 
 based on the static semantics just mentioned.
@@
 e.g. Scala uses \lismall{???} for missing method bodies 
-and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u} where \lismall{u} is an optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
+and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u} where \lismall{u} \IS an optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
 \footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
-This is also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}
+This \IS also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}
 
@@
 In the static semantics developed by \citet{popl-paper}, the machinery for 
-reasoning statically about type holes is derived from gradual type theory, 
+reasoning statically about type holes \IS derived from gradual type theory, 
 identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}. 
@@
 gradual type theory, which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
-However, this is again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
-an exception. As a result, the live programming environment is left unable to provide feedback about even parts 
+However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
+an exception. As a result, the live programming environment \IS left unable to provide feedback about even parts 
 of the program that do not depend on the problematic sub-expression (e.g. later cells in a lab notebook).
@@
 which provides a logical foundation for hole environment tracking). 
-These non-obvious connections to well-established existing systems help support our claim that this approach is theoretically well-grounded\todo{note that our exposition is self-contained here?
+These non-obvious connections to well-established existing systems help support our claim that this approach \IS theoretically well-grounded\todo{note that our exposition \IS self-contained here?
 or later?}.
@@
 every editor state has some, possibly incomplete, type. 
-Together with the theorems established earlier in Sec.~\ref{sec:calculus}, the result is a proof-of-concept
+Together with the theorems established earlier in Sec.~\ref{sec:calculus}, the result \IS a proof-of-concept
 live functional programming environment where every possible editor state has both non-trivial static and dynamic 
@@
 
-Sec.~\ref{sec:resumption} defines the fill-and-resume feature, which is rooted in the contextual substitution operation from CMTT.
+Sec.~\ref{sec:resumption} defines the fill-and-resume feature, which \IS rooted in the contextual substitution operation from CMTT.
 We establish the correctness of fill-and-resume with a novel commutativity property and discuss how it provides a semantic interpretation for cells in a live lab notebook environment.
 
-Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneous discuss various directions for future work.
+Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneous discusses directions for future work.
 Sec.~\ref{sec:discussion} briefly concludes.

### 2018-04-16 intro.tex
Section 1: reduce the 'is' count; use fewer words

@@
 
-The problem that specifically motivates this paper \IS that programming languages typically assign dynamic meaning only to {complete programs}, i.e.
+%The problem that specifically motivates this paper \IS that
+Meanwhile,
+programming languages typically assign dynamic meaning only to {complete programs}, i.e.
 programs that are syntactically well-formed and free of static type and binding errors.
@@
 
-In some cases these gaps are momentary, like while the programmer \IS entering a short expression.
+In some cases these gaps are momentary, like while the programmer
+%\IS entering
+enters
+a short expression.
 In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, or when the programmer makes a mistake (so they arise somewhat more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}).
+%
 The problem \IS particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
@@
 we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes.
-Our goal in this paper \IS to develop just such a dynamic semantics for incomplete functional programs, 
+%
+%Our goal in this paper \IS to develop
+This paper develops
+just such a dynamic semantics for incomplete functional programs, 
 based on the static semantics just mentioned.

### 2018-04-16 intro.tex
unbury the phrase 'this paper develops a dynamic semantics for incomplete functional programs'; rephrase more occurences of 'is'

@@
 Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support editing the state of a running program, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
-Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13}, has been and continues to be an active area of research and development.
+Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13},
+%has been and continues to be an active area of research and development.
+has been, and continues to be, an active area of research and development.
 
+% \matt{ ``The problem that specifically motivates this paper'' is a
+% really long noun phrase; the sentence is more interesting, IMO, when
+% the subject is 'programming languages' and not that noun phrase.  }
+%
 %The problem that specifically motivates this paper \IS that
-Meanwhile,
+Problematically,
 programming languages typically assign dynamic meaning only to {complete programs}, i.e.
@@
 As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e.
-it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer \IS actually seeing and writing in the editor.\todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}
+it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer
+%
+%\IS actually seeing and writing in the editor.
+actually sees and writes in the editor.
+\todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}
 
@@
 a short expression.
-In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, or when the programmer makes a mistake (so they arise somewhat more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}).
+In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, when the programmer changes a type definition, or when the programmer makes a mistake;
+Consequently, they arise somewhat more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}.
 %
-The problem \IS particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
+%\matt{we are not really modeling a language with user type definitions -- what does the following sentence really add here? also, I just added this scenerio to the list above}
+%The problem \IS particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
 Throughout the process of addressing these errors, the program text remains formally meaningless.
@@
 
+%\matt{ Below, notice the broadening from ``the problem that motivates
+%this paper is that incomplete lack dynamic meaing'' to just the
+%general quesiton of ``incomplete program meaning''--- another reason
+%to do my edit above, and remove that specific phrase.
+%}
 In recognition of this problem---that incomplete programs lack meaning---%
@@
 or a structure editor can insert holes fully automatically \cite{popl-paper} (we say more about this technique in Sec.~\ref{sec:implementation}).
-
+%
 For the purposes of live programming, however, a static semantics for incomplete programs does not suffice---%
@@
 %Our goal in this paper \IS to develop
-This paper develops
-just such a dynamic semantics for incomplete functional programs, 
-based on the static semantics just mentioned.
 
-One approach would be for evaluation to simply stop with an error when it encounters an expression hole.
-This mirrors a workaround that programmers commonly deploy: 
-raising an exception when the intended expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
+%%%%%%%%%%%% New par:
+To address this need,
+this paper develops
+a \emph{dynamic semantics} for incomplete functional programs.
+%
+To understand this problem, consider a (common) approach in practice, where evaluation simply stops with an error when it encounters an ``expression hole,'' encoded by the programmer with a raised exception.
+%This mirrors a workaround that programmers commonly deploy: 
+%raising an exception when the intended expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
 Some languages provide more concise notation for this interpretation of expression holes, 
 e.g. Scala uses \lismall{???} for missing method bodies 
-and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u} where \lismall{u} \IS an optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
+and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u}
+%where \lismall{u} \IS an
+where \lismall{u} serves as an
+optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
 \footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
-This \IS also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}
+%This \IS also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}
+proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave this way.}
 
@@
 Moreover, we need to be able to run programs containing not only expression holes but also type holes.
-In the static semantics developed by \citet{popl-paper}, the machinery for 
-reasoning statically about type holes \IS derived from gradual type theory, 
+In the static semantics developed by \citet{popl-paper}, we derive the machinery for 
+reasoning statically about type holes from gradual type theory, 
 identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}. 
@@
 gradual type theory, which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
-However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
-an exception. As a result, the live programming environment \IS left unable to provide feedback about even parts 
+%However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
+However, when a cast fails, evaluation again stops with 
+an exception.
+%
+As a result, traditional gradual typing approaches still leave the live programming environment unable to provide feedback about the parts 
 of the program that do not depend on the problematic sub-expression (e.g. later cells in a lab notebook).

### 2018-04-16 intro.tex
minor nit

@@
 a short expression.
-In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, when the programmer changes a type definition, or when the programmer makes a mistake;
-Consequently, they arise somewhat more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}.
+In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, when the programmer changes a type definition, or when the programmer makes a mistake.
+%
+Further, such mistakes arise more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}.
 %

### 2018-04-16 intro.tex
reorder sentence for flow

@@
 Moreover, we need to be able to run programs containing not only expression holes but also type holes.
-In the static semantics developed by \citet{popl-paper}, we derive the machinery for 
+The static semantics developed by \citet{popl-paper} derives the machinery for 
 reasoning statically about type holes from gradual type theory, 
 identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}.
-To specify our intended dynamic semantics for incomplete programs, then, we can look to the dynamic semantics from
-gradual type theory, which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
+%
+As such, we can look to the dynamic semantics from
+gradual type theory to specify the dynamic semantics for incomplete programs,
+which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
 %However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 

### 2018-04-16 calculus.tex
intro edits

@@
 %
-Although this paper focuses running \emph{incomplete} programs, it helps
+Although this paper focuses on running \emph{incomplete} programs, it helps
 to know that its machinery does not interfere

### 2018-04-16 intro.tex
intro edits

@@
 %
-%The problem that specifically motivates this paper \IS that
-Problematically,
-programming languages typically assign dynamic meaning only to {complete programs}, i.e.
+The problem at the heart of this paper is that
+programming languages typically assign meaning only to {complete programs}, i.e.
 programs that are syntactically well-formed and free of static type and binding errors.
@@
 As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e.
-it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer
-%
-%\IS actually seeing and writing in the editor.
-actually sees and writes in the editor.
+it relies on the most recent complete editor state, creating a perceptive gap because the feedback may not accurately reflect what the programmer is actually seeing and writing in the editor.
 \todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}
@@
 a short expression.
-In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, when the programmer changes a type definition, or when the programmer makes a mistake.
+In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank or when the programmer makes a mistake.
 %
-Further, such mistakes arise more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}.
+Novice programmers, of course, make more mistakes \cite{mccauley2008debugging,fitzgerald2008debugging}.
 %
 %\matt{we are not really modeling a language with user type definitions -- what does the following sentence really add here? also, I just added this scenerio to the list above}
-%The problem \IS particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
+The problem is particularly pronounced for languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
 Throughout the process of addressing these errors, the program text remains formally meaningless.
@@
 In recognition of this problem---that incomplete programs lack meaning---%
-\citet{popl-paper} developed a static semantics (i.e. a type system) for incomplete 
-functional programs, modeled formally as typed expressions with \emph{holes} in 
+\citet{popl-paper} develop a static semantics (i.e. a type system) for incomplete 
+functional programs, modeling them formally as typed expressions with \emph{holes} in 
 both expression and type position. 
@@
 Briefly, error recovery mechanisms can insert the necessary holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09,DBLP:conf/oopsla/KatsV10},
-the programmer can insert explicitly holes manually (see below) 
+the programmer can insert explicit holes manually (see below) 
 or semi-automatically via code completion techniques \cite{Amorim2016}, 
@@
 For the purposes of live programming, however, a static semantics for incomplete programs does not suffice---%
-we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes.
+we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes. This paper aims to address this need by developing
+a {dynamic semantics} for incomplete functional programs, 
+starting from the static semantics of \citet{popl-paper}.
 %
@@
 
+One approach would be to define a dynamic semantics that simply aborts with an error when evaluation reaches a hole. 
 %%%%%%%%%%%% New par:
-To address this need,
-this paper develops
-a \emph{dynamic semantics} for incomplete functional programs.
 %
-To understand this problem, consider a (common) approach in practice, where evaluation simply stops with an error when it encounters an ``expression hole,'' encoded by the programmer with a raised exception.
-%This mirrors a workaround that programmers commonly deploy: 
-%raising an exception when the intended expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
-Some languages provide more concise notation for this interpretation of expression holes, 
-e.g. Scala uses \lismall{???} for missing method bodies 
+This mirrors a workaround to the problem that programmers commonly deploy: 
+raising an exception where an expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
+Some languages provide more concise notation for this ``encoding'' of expression holes, 
+e.g. Scala provides \lismall{???}  
 and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u}
-%where \lismall{u} \IS an
 where \lismall{u} serves as an
-optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
+optional hole name \cite{GHCHoles}.%
 \footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
-%This \IS also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}
-proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave this way.}
+Proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} also respond to holes in this way.}
 
@@
 (2)~it provides no information about the behavior of the remainder of program, 
-parts of which may not depend at all on the missing or erroneous expression; and 
-(3)~it provides no means by which to resume evaluation after filling the hole.
+parts of which may not depend on the missing or erroneous expression (e.g. later cells in a lab notebook); and 
+(3)~it provides no means by which to resume evaluation after filling a hole.
 
@@
 As a result, traditional gradual typing approaches still leave the live programming environment unable to provide feedback about the parts 
-of the program that do not depend on the problematic sub-expression (e.g. later cells in a lab notebook).
+of the program that do not depend on the problematic sub-expression.
 
@@
 
-We are integrating this approach into the \Hazel programming environment being developed independently of this paper by \citet{HazelnutSNAPL}. 
+We are integrating this approach into the \Hazel programming environment being developed (separately from this paper) by \citet{HazelnutSNAPL}. 
 We begin in Sec.~\ref{sec:examples} by detailing the approach informally, with several examples, in the setting of this specific design.
 
-Sec.~\ref{sec:calculus} then abstracts away orthogonal details and makes these intuitions formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions as just outlined, in a manner that admits clean type safety theorems (in particular, a Progress theorem).
-The semantics of \HazelnutLive borrows machinery from gradual type theory~\cite{DBLP:conf/snapl/SiekVCB15,Siek06a}
+Sec.~\ref{sec:calculus} then abstracts away orthogonal details of the language and user interface and makes these intuitions formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions as just outlined, in a manner that admits clean type safety theorems (in particular, a Progress theorem).
+The semantics of \HazelnutLive borrows machinery from gradual type theory~\cite{DBLP:conf/snapl/SiekVCB15}
 (to handle incomplete types) and contextual modal type theory~\cite{Nanevski2008} (CMTT,
 which provides a logical foundation for hole environment tracking). 
-These non-obvious connections to well-established existing systems help support our claim that this approach \IS theoretically well-grounded\todo{note that our exposition \IS self-contained here?
+These non-obvious connections to well-established existing systems help support our claim that this approach is theoretically well-grounded\todo{note that our exposition \IS self-contained here?
 or later?}.
@@
 every editor state has some, possibly incomplete, type. 
-Together with the theorems established earlier in Sec.~\ref{sec:calculus}, the result \IS a proof-of-concept
+Together with the theorems established earlier in Sec.~\ref{sec:calculus}, the result is a proof-of-concept
 live functional programming environment where every possible editor state has both non-trivial static and dynamic 
 meaning, i.e. live feedback never ``flickers'' or ``goes out''. 
-The mechanization and the implementation are both available in the supplemental material.
+The supplemental material includes both the mechanization and the implementation.
 
-Sec.~\ref{sec:resumption} defines the fill-and-resume feature, which \IS rooted in the contextual substitution operation from CMTT.
-We establish the correctness of fill-and-resume with a novel commutativity property and discuss how it provides a semantic interpretation for cells in a live lab notebook environment.
+Sec.~\ref{sec:resumption} defines the fill-and-resume feature, which is rooted in the contextual substitution operation from CMTT.
+We establish the correctness of fill-and-resume with a novel commutativity property and discuss how it provides a semantic interpretation of cells in a live lab notebook environment.
 
-Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneous discusses directions for future work.
+Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneously discusses directions for future work.
 Sec.~\ref{sec:discussion} briefly concludes.

### 2018-04-16 appendix-defns.tex
revisions of sec3 through 3.4 and additional proofs and definitions in the appendix

@@
 \end{lem}
+\begin{proof} By rule induction on the first assumption in each case. The conclusion follows from the definition of substitution in each case. \end{proof}
 
@@
 
+The proofs for all three of these theorems follow by straightforward rule induction.
+
 % No weakening for Gammas in Delta: 
@@
 
+\subsection{Multiple Steps}
+\input{fig-multi-step}
+
 \clearpage
 \subsection{Hole Filling}\label{sec:hole-filling}
-\input{fig-multi-step}
-\input{fig-evalctx-instantiation}
 \begin{lem}[Filling] ~
@@
 
+We need the following auxiliary definitions, which lift hole filling to evaluation contexts taking care to consider the special situation where the mark is inside the hole that is being filled, to prove the Commutativity theorem.
+\input{fig-evalctx-instantiation}
 \begin{lem}[Substitution Commutativity]
@@
 \end{lem}
+\begin{proof}
+We proceed by structural induction on $d_1$ and rule induction on the typing premises, which serve to ensure that the free
+variables in $d'$ are accounted for by every closure for $u$.
+\end{proof}
 
@@
 \end{lem}
+\begin{proof}
+We proceed by cases on the instruction transition assumption (no induction is needed).
+For Rule \rulename{ITBeta}, we defer to the Substitution Commutativity lemma above.
+For the remaining cases, the conclusion follows from the definition of hole filling.
+\end{proof}
 
@@
 \end{lem}
+\begin{proof} We proceed by structural induction on $\evalctx$. Every case is handled by one of the two judgements. \end{proof}
 
@@
 \end{lem}
+\begin{proof} We proceed by structural induction on $\evalctx$ and rule induction on all three assumptions. Each case follows from the definition of instruction selection and hole filling. \end{proof}
 
@@
 \end{lem}
+\begin{proof} We proceed by rule induction on both assumptions. Each case follows from the definition of instruction selection and hole filling. \end{proof}
 
@@
 \begin{proof}
-By rule induction on (3).
+By rule induction on assumption (3).
 The reflexive case is immediate.
 In the inductive case, we proceed by rule induction on the stepping premise.
-There is one rule, Rule~{Step}.
+There is one rule, Rule~\rulename{Step}.
 By Filling Totality, either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx} = \evalctx'$.
-In the former case, by Discarding, we can conclude by MultiStepRefl.
-In the latter case, by Instruction Commutativity and Filling Distribution we can take a step, and we can conclude via MultiStepSteps by applying Preservation and then the IH.
+In the former case, by Discarding, we can conclude by \rulename{MultiStepRefl}.
+In the latter case, by Instruction Commutativity and Filling Distribution we can take a \rulename{Step}, and we can conclude via \rulename{MultiStepSteps} by applying Filling, Preservation and then the induction hypothesis.
 \end{proof}

### 2018-04-16 calculus.tex
revisions of sec3 through 3.4 and additional proofs and definitions in the appendix

@@
 (see Sec.~\ref{sec:agda-mechanization}).
-Rule names, written \rulename{SVar}, correspond to the names used in the mechanization.
+Rule names, e.g.
+\rulename{SVar}, correspond to the names used in the mechanization.
 We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).
@@
 The following theorem establishes that an expansion exists for every well-typed external expression. 
-The mechanization also establishes when an expansion exists, it is unique (not shown).
+The mechanization also establishes that ywhen an expansion exists, it is unique (not shown).
  \begin{thm}[Expandability] \label{thm:expandability}~
@@
 To recap, the result of expansion is a well-typed internal expression with appropriately initialized hole closures and casts. 
-We can now specify, in Figures~\ref{fig:isGround}-\ref{fig:step}, the dynamic semantics of \HazelnutLive as a ``small-step'' transition system equipped with a meaningful notion of type safety even for incomplete programs, i.e.
+This section specifies the dynamic semantics of \HazelnutLive as a ``small-step'' transition system over internal expressions equipped with a meaningful notion of type safety even for incomplete programs, i.e.
 expressions typed under a non-empty hole context, $\Delta$.
@@
 
-The dynamic semantics consists of ,
-%
-where cast-related machinery closely follows the cast calculus from
+Figures~\ref{fig:isGround}-\ref{fig:step} specify the dynamic semantics. 
+%
+Most of the cast-related machinery closely follows the cast calculus from
 the ``refined'' account of the gradually typed lambda calculus
@@
 $\isGround{\htau}$, which distinguishes the base type~$b$ and the
-least specific arrow type~$\tarr{\tehole}{\tehole}$, as \emph{ground
-types}; this judgement helps simplify the treatment of function casts.
+least specific arrow type~$\tarr{\tehole}{\tehole}$ as \emph{ground
+types}; this judgement helps simplify the treatment of function casts, discussed below.
+
 %
 \Figref{fig:isFinal} defines the judgement $\isFinal{d}$, which
-distinguishes the final forms of the transition system.
-%
-There are two classes of final forms: (possibly-)boxed values and
+distinguishes the final, i.e. irreducible, forms of the transition system.
+%
+The two rules distinguish two classes of final forms: (possibly-)boxed values and
 indeterminate forms.%
@@
         %
-        Our judgmental formulation is more faithful to the mechanization and
+        Our judgemental formulation is more faithful to the mechanization and
         cleaner for our purposes, because we are distinguishing several
@@
 }
-
 The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as
 either ordinary values~(Rule~\rulename{BVVal}), or one of two cast forms: casts
-between function types (except identity casts) and casts from a ground
-type to the hole type. In each case, the cast operates on a boxed
+between disequal function types and casts from a ground
+type to the hole type. In each case, the cast must appear inductively on a boxed
 value.
+These forms are irreducible because they represent values that have been boxed 
+but have never flowed into a corresponding ``unboxing'' operation, discussed below. 
 %
@@
 The judgement $\isIndet{d}$ defines \emph{indeterminate} forms, so
-named because they arise from the presence of expression holes and
-failed casts in a program, and so, conceptually, their ultimate value
+named because they are rooted at expression holes and
+failed casts, and so, conceptually, their ultimate value
 awaits programmer action (see Sec.~\ref{sec:resumption}).
 %
-The first two rules specify that \emph{empty} hole closures are always
-indeterminate, and that \emph{non}-empty hole closures are indeterminate when
-they consist of a \emph{final} inner expression.
-%
-Below, we discuss the the remaining indeterminate forms simultaneously
-with the corresponding transition rules, also discussed
-below.\todo{relationship to weak head normal forms}
-
-The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}.
+The first two rules specify that {empty} hole closures are always
+indeterminate, and that {non}-empty hole closures are indeterminate when
+they consist of a {final} inner expression.
+%
+Below, we describe failed casts and the remaining indeterminate forms simultaneously
+with the corresponding transition rules.\todo{relationship to weak head normal forms}
+
+Figures~\ref{fig:instruction-transitions}-\ref{fig:step} define\todo{use define instead of specify everywhere?}{} the transition rules. 
 %
@@
 %
-and (3) places $d_0'$ back in the selected position, indicated in
-the evaluation context by the \emph{mark}~$\evalhole$, to obtain $d'$.%
+and (3) places $d_0'$ back at the selected position, indicated in
+the evaluation context by the \emph{mark}, $\evalhole$, to obtain $d'$.%
 \footnote{
-        We say ``mark'' here to avoid confusion with the (orthogonal) ``holes'' of \HazelnutLive.
+        We say ``mark'', rather than the more conventional ``hole'', to avoid confusion with the (orthogonal) holes of \HazelnutLive.
         %
@@
 %
-
 %% \matt{Next paragraph can be dropped entirely for space}
 %% %
-%% This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing.
-%% Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
-%% It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
-%% \todo{Put the search rules in the appendix?}{}
+This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
+Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
+It would be straightforward to construct an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
+
+The rules take care to maintain the property that final expressions cannot take a step.%
+\todo{prove this as stated from Ian's disjointness proofs}{}%
+\begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
+\end{thm}
 
@@
 to specify an eager, left-to-right evaluation strategy, or \emph{excluded} to
-leave the evaluation order unspecified.
-%
-In our metatheory, we use the latter version of the rules, both for
-the sake of generality, and to support fill-and-resume~(\Secref{sec:resumption}).
-
-A substitution~$[d/x]d'$ operates in the standard capture-avoiding manner~\cite{pfpl}. 
+leave the evaluation strategy and order unspecified.
+%
+In our metatheory, we exclude these premises, both for
+the sake of generality, and to support the specification of the fill-and-resume operation~(see \Secref{sec:resumption}).
+
+\input{fig-ground-types}
+\input{fig-dynamics-aux}
+
+Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see Appendix~\ref{sec:additional-defns}).
 %
@@
 \]
-In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$. 
+In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$, i.e. the environment ``records'' the substitution. 
 %
@@
 %
-Reduction may duplicate hole closures, just as with any other substituted term;
-%
-consequently, the environments of different closures for the same hole name may differ, 
-e.g., when an reduction applies a function with a hole closure body multiple times.%
+Beta reduction can duplicate hole closures.
+%
+Consequently, the environments of different closures with the same hole name may differ, 
+e.g., when an reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.%
 \todo{show example here or refer back to sec 2?}{} 
@@
 
-\input{fig-ground-types}
-\input{fig-dynamics-aux}
-
 The ITBeta rule is not the only rule we need to handle function
-application, because lambdas merely one (final) form of arrow type.
+application, because lambdas are not the only final form of arrow type.
 %
@@
 In this case, the application is indeterminate (Rule~\rulename{IAp}
-in \Figref{fig:isFinal}), and the application's reduction reduces no
-further.
+in \Figref{fig:isFinal}), and the application reduces no
+further.\todo{mention WHNF here?}
 
 \input{fig-dynamics-contexts}
-
-The rules take care to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.%
-\todo{prove this as stated from Ian's disjointness proofs}{}%
-\todo{do we need a typing premise?}{}
-\begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
-\end{thm}
 
@@
 Rule \rulename{ITCastId} drops identity casts. The remaining instruction
-transition rules assign meaning to non-trivial casts.
+transition rules assign meaning to non-identity casts. 
 %
@@
 so we must wait for \emph{use} of the term at some other type, via a
-cast away from hole type, to detect dynamic type errors.
-%
-Rules {ITCastSucceed} and {ITCastFail} handle this situation when the
+cast away from hole type, to detect the type error dynamically.
+%
+Rules \rulename{ITCastSucceed} and \rulename{ITCastFail} handle this situation when the
 two types involved are ground types (Fig.~\ref{fig:isGround}).
@@
 Rule \rulename{IFailedCast} specifies that a failed cast operates as an
-indeterminate form once $d$ is final.
-%
-This rule allows evaluation to proceed after cast failure, much as
-with hole closures.
-
-Rules {TAFailedCast} and {IFailedCast} each operate at ground type. 
+indeterminate form (once $d$ is final), i.e. evaluation does not stop.\todo{talk about blame somewhere}{}
+
+Rules \rulename{TAFailedCast} and \rulename{IFailedCast} each operate at ground type. 
 %
 The two remaining instruction transition rules, Rule~\rulename{ITGround}
-and~{ITExpand}, insert intermediate casts from non-ground type to a
+and~\rulename{ITExpand}, insert intermediate casts from non-ground type to a
 consistent ground type, and \emph{vice versa}.
@@
 %
-More generally, this judgement is governed by the following invariant:
+More generally, the following invariant governs this judgement.
 \begin{lem}[Grounding] 
-  If $\groundmatch{\htau}{\htau'}$
-  then $\isGround{\htau'}$
-  and $\tconsistent{\htau}{\htau'}$
-  and $\htau\neq\htau'$.
+  If $\groundmatch{\htau_1}{\htau_2}$
+  then $\isGround{\htau_2}$
+  and $\tconsistent{\htau_1}{\htau_2}$
+  and $\htau_1\neq\htau_2$.
 \end{lem}
 
-In all other cases, casts either consist of boxed values or
-indeterminate terms according to the remaining rules
+In all other cases, casts evaluate either to boxed values or to 
+indeterminate forms according to the remaining rules
 in \Figref{fig:isFinal}.
@@
 %
-In the cast calculus, since the only canonical form of type $\tehole$
-is $\dcasttwo{\dexp}{\htau_1}{\tehole}$, there exist no such
-irreducible terms.
+(In the cast calculus, there exist no such
+irreducible terms; the only canonical form of type $\tehole$
+is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.)
 
@@
 %
-Type safety establishes that the static and dynamic semantics of a
+The purpose of establishing type safety is to ensure that the static and dynamic semantics of a
 language cohere.
@@
 %
-We assume an empty context for~$\hGamma$; to run open programs, the
+We assume an empty typing context, $\hGamma$; to run open programs, the
 system may treat free variables as empty holes with a corresponding
@@
 The key to establishing the progress theorem under a non-empty hole
-context consists of explicitly accounting for indeterminate forms,
+context is to explicitly account for indeterminate forms,
 i.e. those rooted at either a hole closure or a failed cast.
@@
 %
-Appendix~\ref{sec:additional-defns} defines complete programs via judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
+Appendix~\ref{sec:additional-defns} defines the judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
 %
@@
 
-The following preservation theorem establishes that stepping preserves program completeness.\todo{can we do complete preservation without the typing assumption?}
+The following preservation theorem establishes that stepping preserves program completeness.
 \begin{thm}[Complete Preservation]
@@
 %including proofs of all of the 
-including the theorems stated above, and all necessary lemmas\todo{make this so ian!}{}. 
+including the theorems stated above and necessary lemmas.
+We take as assumptions only a few standard properties of general hypothetical judgements and finite maps.\todo{what about hole disjointness}
 %
@@
 %
-Our approach is standard: We model judgements as 
+Our approach is standard: we model judgements as 
 inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite something?}{}. 
@@
 We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
-contexts, hole contexts and hole environments using metafunctions.\todo{say something about axioms?}{}
+contexts, hole contexts and hole environments using metafunctions.\todo{say something about funext?}{}
 

### 2018-04-16 fig-complete.tex
fig-complete
draft of sec 3.0 through 3.1
finished draft of sec3 and 4
revisions of sec3 through 3.4 and additional proofs and definitions in the appendix

@@
-\begin{figure}[t]
-\rkc{TODO}
-\caption{Complete H-types, H-expressions, and dynamic H-expressions}
+% !TEX root = hazelnut-dynamics.tex
+\begin{figure}[h]
+\judgbox{\isComplete{\htau}}{$\htau$ is complete}
+\begin{mathpar}
+\inferrule[BComplete]{ }{
+  \isComplete{b}
+}
+
+\inferrule[ArrComplete]{
+  \isComplete{\htau_1}\\
+  \isComplete{\htau_2}
+}{
+  \isComplete{\tarr{\htau_1}{\htau_2}}
+}
+\end{mathpar}
+
+\vsepRule
+
+\judgbox{\isComplete{\hexp}}{$\hexp$ is complete}
+\begin{mathpar}
+\inferrule[EVarComplete]{ }{
+  \isComplete{x}
+}
+
+\inferrule[EConstComplete]{ }{
+  \isComplete{c}
+}
+
+\inferrule[ESynLamComplete]{
+  \isComplete{\htau}\\
+  \isComplete{\hexp}
+}{
+  \isComplete{\halam{x}{\htau}{\hexp}}
+}
+
+\inferrule[EAnaLamComplete]{
+  \isComplete{\hexp}
+}{
+  \isComplete{\hlam{x}{\hexp}}
+}
+
+\inferrule[EApComplete]{
+  \isComplete{\hexp_1}\\
+  \isComplete{\hexp_2}
+}{
+  \isComplete{\hap{\hexp_1}{\hexp_2}}
+}
+
+\inferrule[EAscComplete]{
+  \isComplete{\hexp}\\
+  \isComplete{\htau}
+}{
+  \isComplete{\hexp : \htau}
+}
+\end{mathpar}
+
+\vsepRule
+
+\judgbox{\isComplete{\dexp}}{$\dexp$ is complete}
+\begin{mathpar}
+\inferrule[DVarComplete]{ }{
+  \isComplete{x}
+}
+
+\inferrule[DConstComplete]{ }{
+  \isComplete{c}
+}
+
+\inferrule[DLamComplete]{
+  \isComplete{\htau}\\
+  \isComplete{\dexp}
+}{
+  \isComplete{\dlam{x}{\htau}{\dexp}}
+}
+
+\inferrule[DApComplete]{
+  \isComplete{\dexp_1}\\
+  \isComplete{\dexp_2}
+}{
+  \isComplete{\dap{\dexp_1}{\dexp_2}}
+}
+
+\inferrule[DCastComplete]{
+  \isComplete{\dexp}\\
+  \isComplete{\htau_1}\\
+  \isComplete{\htau_2}
+}{
+  \isComplete{\dcasttwo{\dexp}{\htau_1}{\htau_2}}
+}
+\end{mathpar}
+
+\caption{Complete types, external expressions, and internal expressions}
 \label{fig:complete}
 \end{figure}
+
+When two types are complete and consistent, they are equal.
+
+\begin{lem}[Complete Consistency] If $\tconsistent{\htau_1}{\htau_2}$ an $\isComplete{\htau_1}$ and $\isComplete{\htau_2}$ then $\htau_1 = \htau_2$. 
+\end{lem}
+\begin{proof} By straightforward rule induction. \end{proof}
+
+This implies that in a well-typed internal expression, every cast is an identity cast.

### 2018-04-17 fig-grades-example.tex
added mockup

--- 2018-02-20 fig-grades-example.tex
@@
+% !TEX root = hazelnut-dynamics.tex
+
 \begin{figure}[t]
 
-\begin{lstlisting}
-type student_data =
-  { name: string, 
-    hw1: float, hw2: float, hw3: float, 
-    midterm: float, final: float };
+\includegraphics[width=1.01\textwidth,interpolate=false]{images/grades-cell-mockup.png}
 
-let grades : list(student_data) =
-  [ { name: "Alice", hw1: 76.0, hw2: 93.0, hw3: 91.0, midterm: 77.0, final: 95.0 }
-  , { name: "Bob"  , hw1: 88.0, hw2: 75.0, hw3: 80.0, midterm: 89.0, final: 87.0 }
-  ];
-\end{lstlisting}
+% %% TODO once the code above is removed, scale up the screenshots
+% \includegraphics[scale=0.20]{images/hazel-placeholder-0.png}
 
-\rkc{Place this code fragment (and some pieces of the subsequent fragments) in a
-single mockup or screenshot.}
+% \rkc{Draw arrows and captions on the top figure to show how to get
+% to the bottom figure.
+% ser navigates to hole a, types + to create a plus, types * to create a
+% multiplication, types \#10 to create 10, types vh1 to create variable use.}
 
-\caption{\rkc{...}}
+% \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}
+
+\caption{Hazel mockups for Example 1a.}
 \label{fig:grades-example}

### 2018-04-17 examples-1.tex
sec2!

@@
-\subsection{Live Programming with Incomplete Expressions}
+% !TEX root = hazelnut-dynamics.tex
 
-The process of writing new program fragments includes many states in which the
-program under construction is, by definition, incomplete.
+\subsection{Example 1: Evaluating Past Holes and Hole Closures}
+
+Consider the perspective of a teacher in the midst of developing a \Hazel{} notebook
+to compute final student grades at the end of a course.
+The top of \autoref{fig:grades-example} depicts the cell containing the incomplete program that 
+the teacher has written so far (we omit irrelevant parts of the UI). 
 %
-It is natural for a programmer to ``jump back and forth'' between different
-places in the program, iteratively and alternately developing the missing
-pieces.
-%
-The primary benefit of our approach is that such programs can be evaluated in
-order to provide useful feedback during this editing workflow.
 
-\input{fig-grades-example}
+At the top of this program, the teacher defines a
+record type, \lismall{Student}, for recording a student's course data---here,
+the student's name, of type
+\lismall{string}, and, for simplicity, three grades, each of type \lismall{float}.
+The variable \lismall{students} binds a list of such student records.
+For simplicity, we include only three example students.
+At the bottom of the program, the teacher maps a function \li{weighted_average} over this student data (\li{map} is the standard map function over lists, not shown), intending to compute a final weighted average for each student.
+However, the program is incomplete because the teacher has not yet completed the body of the \lismall{weighted_average} function.
+This pattern is well-established: programmers often consume a function before implementing it.
 
-\overviewExample{1a}{Computing Weighted Averages}
-%
-Consider a teacher who is in the midst of developing a \Hazel{} notebook
-(depicted at the top of \autoref{fig:grades-example}) to compute final student
-grades at the end of a term.
-%
-At the top of the program, the teacher defines a
-record type, \li{student_data}, for each student's course data.
-%
-In our simplified example, this includes the student's name, of type
-\li{string}, and five grades, each of type \li{float}.
+Thusfar in Fig.~\ref{fig:grades-example}, the teacher has decomposed the function argument into variables by record destructuring, then multiplied the homework grade, \li{hw}, by \li{30.0} and finally inserted the \li{+} operator.
+The cursor now sits at an \emph{empty hole}, as indicated by the vertical bar and the green background.
+Each hole has a unique name (generated automatically in \Hazel), here simply \li{1}.
+%In this case, there is only one hole in the program, but in general, there may be many such holes.
 
-The teacher begins writing a \li{weighted_average} function that will
-compute a weighted average, of type \li{float}, for each \li{student_data}
-record.
-%
-The body of \li{weighted_average} is not yet written---marked with the
-\emph{empty hole} on line \rkc{XXX}---but the teacher jumps ahead to map the
-function over the \li{grades} list on line \rkc{XXX}.
-%
-\Hazel{} evaluates this incomplete program, showing its
-\emph{indeterminate} result in \emph{results panel} just below the
-program in the center of the window (\autoref{fig:grades-example});
-this value is a list whose length is the same as
-\li{grades}, where each element is the (indeterminate) result of evaluating the
-hole expression inside \li{weighted_average}.
-%
-\rkc{The editor shows the first three elements of a result list, by default; the
-... can be clicked and expanded to see additional elements.}
+Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback available to the programmer.
+In response, the programmer might insert a temporary universal placeholder, e.g.
+by raising an exception as suggested in Sec.~\ref{sec:intro} (note however that Elm does not support exceptions, by design).
+This would cause typechecking to succeed.
+However, when the programming environment attempts to evaluate the program, 
+evaluation would proceed only as far as the first \li{map} iteration, 
+which would call into \li{weighted_average} and then fail when attempting to evaluate the hole.%
+\footnote{In a lazy language, like Haskell, the result would be much the same because the environment forces the result for printing.}
+While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.
 
-Notice that the hole expression on line \rkc{XXX} is assigned a unique static
-identifier (\ie{}~\li{a}) and that the indeterminate results it produces are
-assigned corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2},
-etc.).
-%
-In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
-using placeholder expressions like \li{raise "Not yet implemented."}, would halt
-the program as soon it is evaluated (in this case, inside the definition of
-\li{List.map} when the function is applied to the first element of \li{grades}.
-%
-By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
-least see that the resulting expression does indeed evaluate to a list with the
-correct number of (indeterminate) values.
+\Hazel{} does not take this ``exceptional'' interpretation of holes.
+Instead, evaluation continues past the hole, treating it as an opaque expression of the appropriate type.
+The result, shown at the bottom of Fig.~\ref{fig:grades-cell-mockup}, is a list of length $3$, confirming that \li{map} does indeed behave as expected in this regard despite the teacher having provided an incomplete argument.
+Furthermore, each element of the resulting list has been evaluated as far as possible, i.e.
+the arithmetic expression \li{30.0*hw} has been evaluated for each corresponding value of \li{hw}, as expected.
+Evaluation cannot proceed any further because holes appear as addends.
+We say that each of these addition expressions is an \emph{indeterminate} sub-expression, and the result as a hole is therefore also indeterminate, because it is not yet a value, nor can it take a step due to holes in critical positions.
 
-Now the teacher returns to \li{weighted_average} to work on the missing
+Given this indeterminate result, the teacher might take notice of the magnitude of the numbers being computed, e.g. \li{2640.0} and \li{2280.0}, and
+decide immediately that a mistake has been made: the teacher wants to compute a 
+weighted average between \li{0.0} and \li{100.0}, and so the correct
+constant is \li{0.30} rather than \li{30.0}. (There are of course other possible fixes.)
+
+Although these observations might 
+save only a small amount of time in this case, 
+% relative to the 
+% situation where the teacher notices this mistake only once  
+% the \li{weighted_average} function has been completed, 
+it demonstrates
+the broader motivations of live programming: continuous feedback about the dynamic behavior of the 
+program can help confirm the mental model 
+that the programmer has developed (in this case, regarding the behavior of \li{map}), or else help quickly dispel 
+inconsistencies between the programmer's expectations and the 
+actual behavior of the program (in this case, the arithmetic expression).
+
+% The result itself is not the only live feedback that \Hazel reports to the programmer.
+% Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
+\Hazel also helps programmers reason about the dynamic behavior of bound variables in scope at a hole via the \emph{live context inspector}, normally displays it as a sidebar but shown disembodied in three states in Fig.~\ref{fig:grades-sidebar}.
+In all three states, the live context inspector displays the names and types of the variables that are in scope, as is a common feature of programming environments.
+
+New to our approach are the values associated with each binding.
+These values come from the environments associated with each hole instance in the result.
+In this case, there are three instances of hole \li{1} in the result, numbered sequentially: \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages}.
+An environment is a partial mapping of each variable in scope to a value.
+We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical foundations detailed in Sec.~\ref{sec:calculus}).
+The programmer can select different closures by clicking on a hole instance (here, instance \li{1:1} was selected by default, as indicated by the purple outline), or cycle rapidly through all of the closures for the hole at the cursor by clicking the arrows, e.g.
+$\blacktriangleright$, at the bottom of the context inspector, or corresponding hotkeys.
+
+\vspace{-4px}
+\subsection{Example 2: Recursive Functions}\label{sec:qsort1}
+\vspace{-2px}
+
+\input{qsort-example}
+
+The first example was purposely simple.
+Let us briefly consider a second more sophisticated example of these concepts: an incomplete implementation of quicksort as a recursive function, shown in Fig.~\ref{fig:qsort-cell-mockup}.
+Here, the programmer has selected a pivot, split the remainder of the list relative to this pivot and made the appropriate recursive calls.
+A hole appears in the return position of the function body as the programmer contemplates how to combine the results from the recursive calls.
+
+The programmer has called \li{qsort} on an example list. However, 
+the indeterminate result in this case is rather uninteresting: it simply confirms that for this example, evaluation went through the recursive case of \li{qsort}, which ends with hole \li{1}.
+The live context inspector, on the other hand, provides a wealth of feedback about the dynamic behavior of this incomplete program.
+For example, the programmer can confirm that
+the lists \li{smaller} and \li{bigger} are appropriate named.
+
+The results from the recursive calls, \li{r_smaller} and \li{r_bigger}, are again hole instances, \li{1:2} and \li{1:3}.
+The programmer can click on these hole instances to reveal the environments at the corresponding recursive calls.
+For example, clicking on \li{1:2} reveals that the pivot in that case is \li{7} (not shown).
+In exploring these paths rooted at the result, the programmer can develop intuitions about the recursive structure of the computation before the program is complete.
+(We leave to future work the pedagogical question of whether such early feedback might help students with recursion.)
+
+\begin{comment}
+Now the teacher returns to \lismall{weighted_average} to work on the missing
 expression, with the goal of computing a weighted sum of each of a student's
@@
 %
-navigating the cursor to hole \li{a} and typing \li{+}, which creates a
-sum expression \li{??_b + ??_c} with the cursor placed, by default, under
-left operand (the hole labeled \li{b});
+navigating the cursor to hole \lismall{a} and typing \lismall{+}, which creates a
+sum expression \lismall{??_b + ??_c} with the cursor placed, by default, under
+left operand (the hole labeled \lismall{b});
 %
-typing \li{*} to create a multiplication expression \li{??_d * ??_e + ??_c},
-with the cursor place under hole \li{d};
+typing \lismall{*} to create a multiplication expression \lismall{??_d * ??_e + ??_c},
+with the cursor place under hole \lismall{d};
 %
-typing \li{\#10} to enter a numeric scaling factor, which then moves
-the cursor to the second operand, hole \li{e}; and
+typing \lismall{\#10} to enter a numeric scaling factor, which then moves
+the cursor to the second operand, hole \lismall{e}; and
 %
-typing \li{vg.hw1} to use the \li{hw1} field of the function argument
+typing \lismall{vg.hw1} to use the \lismall{hw1} field of the function argument
 as the right operand of the multiplication.
 %
-The resulting partial expression, \li{(10.0 *. g.hw1) +. ??_b}---shown
+The resulting partial expression, \lismall{(10.0 *. g.hw1) +. ??_b}---shown
 in the screenshot in the bottom half of \autoref{fig:grades-example}---marks the
-first of five weighted summands in the eventual finished \li{weighted_average}
+first of five weighted summands in the eventual finished \lismall{weighted_average}
 calculation.
+\end{comment}
 
+\begin{comment}
 Each \Hazel{} edit action transforms an incomplete but well-formed and
@@
 After the above edit sequence, \Hazel{} display the updated list of
-indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
+indeterminate expressions \lismall{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
 %
 Right away, the teacher recognizes that the values are too large; they should be
-at most \li{100.0}.
+at most \lismall{100.0}.
 %
-The teacher realizes that representing percentage points as \li{float}s requires
-that the constant on line \rkc{XXX} ought to be \li{0.10} instead.
+The teacher realizes that representing percentage points as \lismall{float}s requires
+that the constant on line \rkc{XXX} ought to be \lismall{0.10} instead.
 %
@@
 immediately running the finished program shows that each of values in the final
-result list is in the range \li{0.0} to \li{100.0}.
+result list is in the range \lismall{0.0} to \lismall{100.0}.
+\end{comment}
 
+\begin{comment}
 \input{fig-grades-example-2}
@@
 %
-The \li{grade_cutoffs} record type, shown at the top of
+The \lismall{grade_cutoffs} record type, shown at the top of
 \autoref{fig:grades-example-b}, describes the minimum cutoff for each of
@@
 %
-Initially, each value in the \li{cutoffs} record is a hole.
+Initially, each value in the \lismall{cutoffs} record is a hole.
 
@@
 Before starting to fill in the cutoff values, the teacher jumps ahead to write a
-function \li{letter_grade} that will make the connection between \li{cutoffs}
-and \li{weighted_averages}.
+function \lismall{letter_grade} that will make the connection between \lismall{cutoffs}
+and \lismall{weighted_averages}.
 %
 Because she intends to look at the data to help select the cutoff values, the
-teacher sorts the \li{weighted_averages} (on line \rkc{XXX}) and then maps
-them to \li{letter_grades} (on line \rkc{XXX}).
+teacher sorts the \lismall{weighted_averages} (on line \rkc{XXX}) and then maps
+them to \lismall{letter_grades} (on line \rkc{XXX}).
 
 When \Hazel{} runs this program, the guard of the outermost conditional
-(\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
+(\lismall{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \lismall{cutoffs.a}
 is.
 %
-Therefore, each of the indeterminate expressions in \li{letter_grades} is the
-entire expression body, albeit with different bindings for \li{n}.
+Therefore, each of the indeterminate expressions in \lismall{letter_grades} is the
+entire expression body, albeit with different bindings for \lismall{n}.
 %
@@
 expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
-or hole) or list of small values) simply as \li{..} to save space.
+or hole) or list of small values) simply as \lismall{..} to save space.
 %
@@
 
-To start deciding \li{cutoffs}, the teacher clicks the \li{weighted_averages}
+To start deciding \lismall{cutoffs}, the teacher clicks the \lismall{weighted_averages}
 expression, and views the results panel to see the data sorted in descending
@@
 %
-The result shows a natural gap between \li{92.2} and \li{89.5}.
+The result shows a natural gap between \lismall{92.2} and \lismall{89.5}.
 %
-So, she chooses to use \li{92.0} as the cutoff for A, replacing hole \li{XXX} on
+So, she chooses to use \lismall{92.0} as the cutoff for A, replacing hole \lismall{XXX} on
 line \rkc{XXX} with that numeric value.
@@
 expressions for the first \rkc{XXX} indeterminate expressions, because each of
-those \li{n} values was greater than \li{cutoffs.a}.
+those \lismall{n} values was greater than \lismall{cutoffs.a}.
 %
@@
 %
-She jumps ahead to to write a function (\li{compute_distribution} on lines
+She jumps ahead to to write a function (\lismall{compute_distribution} on lines
 rkc{XXX}-\rkc{XXX}) that computes the distribution of letter grades based
@@
 %
-Running \li{compute_distribution} shows that the percentage of As is
+Running \lismall{compute_distribution} shows that the percentage of As is
 \rkc{XXX\%}, which is smaller than what the teacher would like;
@@
 %
-Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
-and then another gap between \li{88.2} and {85.5}.
+Returning to the value of \lismall{sorted_data}, she sees a cluster around \lismall{89.0}
+and then another gap between \lismall{88.2} and {85.5}.
 %
-So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
+So, the teacher adjusts \lismall{cutoffs.a} to be \lismall{88.0}.
 %
@@
 going back and forth between different pieces under development.
+\end{comment}

### 2018-04-17 examples-2.tex
sec2!

@@
+% !TEX root = hazelnut-dynamics.tex
+
 %% \subsection{Live Programming with Type Errors}
-\subsection{Live Programming with Static Type Errors}
 
-In the previous section, we described example programs that were incomplete
+\subsection{Example 3: Live Programming with Static Type Errors}
+
+% \begin{subfigure}[t]{\textwidth}
+\begin{figure}
+\centering
+\includegraphics[width=0.7\textwidth,interpolate=false,valign=t]{images/qsort-type-error.png}
+\caption{Example 3: Ill-Typed Quicksort}
+\label{fig:qsort-type-error}
+\vspace{-8px}
+\end{figure}
+% \end{subfigure}
+
+The previous examples were incomplete 
 because of \emph{missing} expressions.
 %
-Next, we describe example programs that are incomplete because of
-\emph{type-inconsistent} expressions.
+Now, we discuss programs that are incomplete, 
+and therefore conventionally meaningless, because of
+\emph{type inconsistencies}. 
 %
-For example, consider the following two definitions.
+Let us return to the quicksort example just described, 
+but assume that the programmer has filled in the previous hole
+as shown in Fig.~\ref{fig:qsort-type-error}. (In Sec.~\ref{sec:resumption}, we discuss
+how the programming environment might avoid restarting evaluation after such edits, but for small examples like this, restarting works fine.)
 
-\begin{lstlisting}
-let bad_bool : bool = ?? 0 ??_bad_bool;
+The programmer appears to be on the right track conceptually
+in recognizing that the pivot needs to appear between the 
+smaller and bigger elements. 
+However, the types do not quite work out: the \li{@} operator here
+performs list concatenation, but the pivot is an integer. 
+Most compilers and editors will report a static error message, 
+of varying utility, to the programmer in this case (and \Hazel 
+follows suit, not shown). 
+Our argument is that this should not cause all feedback about 
+the dynamic behavior of the program to ``flicker out'' or ``go stale''. After all,
+the error is localized and there is perfectly good code elsewhere 
+in the program (if not nearby, then perhaps far away).
 
-let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
-\end{lstlisting}
+Our approach, following the prior work of \citet{popl-paper}, 
+is to semantically internalize the ``red outline'' around
+type inconsistencies as a \emph{non-empty hole}.
+From the outside, a non-empty hole behaves statically much like an empty hole.
+Our approach takes the same approach for the dynamic semantics,
+i.e. evaluation safely proceeds past a non-empty hole just as if it were an empty hole.
+The semantics also associates an environment with each instance of a non-empty hole (not shown). 
+The difference is that evaluation also proceeds inside the hole, so that 
+feedback about an expression that might ``almost'' be correct is available. 
+In this case, the result at the bottom of Fig.~\ref{fig:qsort-type-error}
+reveals that the programmer is on the right track: the list elements 
+(here, for a shorter example for the sake of exposition) are in the right order.
+They simply have not been combined correctly.
 
-\noindent
-%
-These two definitions are ill-typed under standard typing disciplines.
-%
-In contrast, \citet{popl-paper} present a bidirectional type system that assigns
-types to both, by wrapping type-inconsistent expressions (\li{0} on line
-\rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
-%
-Non-empty holes prevent local type inconsistencies from polluting the rest of
-the program surrounding it, which may or may not itself contain additional
-inconsistencies.
+% For example, consider the following two definitions.
 
-Understanding and debugging static type errors is notoriously difficult,
-particularly for novices.
-%
-A variety of approaches have been
-proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
-and explain type errors.
-%
-One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
-that demonstrates a run-time failure, and then displays a compressed execution
-trace to the user as a graph.
-%
-In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
-non-empty holes) as far as possible, in particular, until they
-would go wrong.
-%
-Thus, our operational semantics provides a similar benefit to the approach of
-\citet{Seidel2016}.
+% \begin{lstlisting}
+% let bad_bool : bool = ?? 0 ??_bad_bool;
 
-\overviewExample{3}{Sum List}
+% let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
+% \end{lstlisting}
+
+% \noindent
+% %
+% These two definitions are ill-typed under standard typing disciplines.
+% %
+% In contrast, \citet{popl-paper} present a bidirectional type system that assigns
+% types to both, by wrapping type-inconsistent expressions (\li{0} on line
+% \rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
+% %
+% Non-empty holes prevent local type inconsistencies from polluting the rest of
+% the program surrounding it, which may or may not itself contain additional
+% inconsistencies.
+
+\begin{comment}
+\overviewExample{2}{Sum List}
 %
@@
 successfully proceeded to evaluate despite the error elsewhere.
+\end{comment}
 
@@
 %% (add some mechanism for type-consistent non-empty-holes...)
+
+\begin{comment}
 
@@
 and, when reevaluated, produces the desired result.
+
+\end{comment}

### 2018-04-17 examples-3.tex
sec2!

@@
-\subsection{Live Programming with Gradual Type Errors}
+% !TEX root = hazelnut-dynamics.tex
 
+\vspace{-4px}
+\subsection{Example 4: Gradual Type Errors}
+\vspace{-2px}
+So far, we have only discussed programs where a hole appears within an expression.
+\Hazel also allows holes to appear in types.
+Fortunately, \citet{popl-paper} confirms that the substantial literature on \emph{gradual type systems} \cite{Siek06a,DBLP:conf/snapl/SiekVCB15} is directly relevant to the problem of reasoning statically around type holes.
+Unsurprisingly, it is also relevant to the problem of running 
+programs with incomplete types. Indeed, that is the very purpose of gradual typing. As such, let us consider only a small synthetic example to demonstrate what is different in \Hazel.
+\begin{lstlisting}
+f : ? -> int
+f x = if x then 1 else 2
+(f true, f 3, f false)
+\end{lstlisting}
+Evaluation produces the following result:
+\begin{lstlisting}[numbers=none]
+(1, if SSTR3<int => ? =/> bool>ESTR then 1 else 2, 2)
+\end{lstlisting}
+Here, we have left a hole, \li{?}, in the type of \li{f}.
+On Line 3, we call \li{f} three times, with two different types of arguments.
+The static semantics of gradual typing permits all of these calls (see next section) and indeed, the first call, with \li{true} as the argument, evaluates unproblematically to \li{1}.
+
+The second call, however, is problematic: \li{3} is of type \li{int} but it flows into boolean guard position.
+The usual semantics for gradual typing would simply abort evaluation here with a dynamic type error.
+However, this deprives the programmer of live feedback about other parts of the program that might not depend at all on this problematic function call---in this case, the third call to \li{f}.
+Instead, we follow the approach we took above for static type inconsistencies: we treat a failed cast like a hole.
+A failed cast, notated textually and in red \li{SSTR3<int => ?
+=/> bool>ESTR} (see next section), reports both the expected type, here \li{bool}, and the assigned type, here \li{int}.
+Consequently, the second call to \li{f} stops as if the guard were an empty hole, but the third call can progress safely.
+
+\begin{comment}
 \emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept
@@
 
-\overviewExample{5}{Dynamically Typed Negation}
+\overviewExample{3}{Dynamically Typed Negation}
 
@@
 This approach can be incorporated into existing gradually typed languages.
+\end{comment}

### 2018-04-17 examples.tex
near-final intro
sec2!

--- 2018-03-08 examples.tex
@@
 %% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
-\newcommand{\examplesSec}{\HazelnutLive, By Example}
+\newcommand{\examplesSec}{Live Functional Programming with Typed Holes in Hazel}
 \section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
@@
 
+\input{fig-grades-example}
+
 \newcommand{\overviewExample}[2]{\paragraph{Example {#1}: {#2}}}
 
-We introduce \HazelnutLive{} in this section using several examples,
-demonstrating how running incomplete programs---with missing expressions and/or
-type-inconsistent expressions---can be useful.
-%
-We will also demonstrate how running even complete programs with subexpressions
-marked as holes may be useful in debugging.
+We begin with an example-driven overview of our proposed approach  within the \Hazel programming environment \citet{HazelnutSNAPL}.
+\Hazel is an open source web lab notebook environment based roughly on IPython/Jupyter \cite{Perez:2007:ISI:1251563.1251831}.
+The language of \Hazel is tracking toward parity with \Elm, a popular pure functional programming language similar to ``core ML'' \cite{czaplicki2012elm,Elm}.
+We assume familiarity with the basic concepts of ML/Elm (see \url{http://elm-lang.org/}).
 
-\HazelnutLive{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
-alternative Reason syntax \cite{reason-what} extended to support holes.
-%
-As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
-literate program \cite{knuth1984literate}: rich text comments appear at the top
-level and code appears within cells embedded into these comments.
-%
-The reader is encouraged to work through these examples in our prototype
-implementation or watch video demonstrations, both available in
-\suppMaterials{}.\footnote{
-%
-The programs in this section are written using minor syntactic conveniences that
-are not currently supported in our implementation.
-%
-For example, the implementation of recursive types in our current implementation
-requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
-inferred in standard ways (by pairing them with constructors and deconstructors
-for algebraic datatypes) in future work.
-%
-More details about other minor syntactic differences are described in
-\suppMaterials{}.
-%
-}
+Both the user interface and the language of \Hazel are under active development. For the sake of 
+exposition, we have ``post-processed'' the examples in this section to make use of several convenient language mechanisms from \Elm~that are not yet implemented in \Hazel as of this writing.
+These mechanisms are orthogonal to the contributions of this paper.
+We have implemented all of the live programming features described in this section, both in a branch of \Hazel and in the implementation of the core calculus that we describe in Sec.~\ref{sec:implementation} and include in the supplement.
 
@@
 
-\paragraph{Recap}
-%
-In the above four subsections, respectively, we considered programs with:
-%
-(1) empty expression holes,
-%
-(2) non-empty expression holes for ill-typed expressions,
-%
-(3) type holes, and
-%
-(4) non-empty expression holes for well-typed expressions.
-%
-Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
-that allows programs with combinations of these kinds of incompleteness to be
-evaluated.
+% \paragraph{Recap}
+% %
+% In the above four subsections, respectively, we considered programs with:
+% %
+% (1) empty expression holes,
+% %
+% (2) non-empty expression holes for ill-typed expressions,
+% %
+% (3) type holes, and
+% %
+% (4) non-empty expression holes for well-typed expressions.
+% %
+% Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
+% that allows programs with combinations of these kinds of incompleteness to be
+% evaluated.

### 2018-04-17 fig-grades-example.tex
sec2!

@@
 \begin{figure}[t]
+\begin{subfigure}[t]{\textwidth}
+\centering
+\includegraphics[width=0.85\textwidth,interpolate=false]{images/grades-cell-mockup.png}
+\caption{Evaluating an incomplete functional program past the first hole}
+\label{fig:grades-cell-mockup}
+\end{subfigure}
 
-\includegraphics[width=1.01\textwidth,interpolate=false]{images/grades-cell-mockup.png}
+\vspace{10px}
 
+\begin{subfigure}[t]{\textwidth}
+\centering
+\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-1.png}
+~${}^\blacktriangleright$
+\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-2.png}
+~${}^\blacktriangleright$
+\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-3.png}
+\caption{The programming environment communicates relevant static \emph{and} dynamic information in a sidebar.}
+\label{fig:grades-sidebar}
+\end{subfigure}
 % %% TODO once the code above is removed, scale up the screenshots
@@
 % \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}
+\vspace{5px}
 
-\caption{Hazel mockups for Example 1a.}
+\caption{Example 1: Grades}
 \label{fig:grades-example}
+\vspace{-8px}
 \end{figure}

### 2018-04-17 hazelnut-dynamics.tex
pass over Section 4; copy edits and reorganization
sec2!

@@
 \usepackage{xspace}
+\usepackage{wrapfig}
+\usepackage[export]{adjustbox}% http://ctan.org/pkg/adjustbox
 
@@
 morekeywords={string, float, int},
-classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount
+classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,
+moredelim=**[is][\color{red}]{SSTR}{ESTR}
 }
-\let\li\lstinline
+\newcommand{\li}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
 \newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}

### 2018-04-17 related-work.tex
sec2!

@@
 dynamic semantics as well---we need to be able to run programs where type holes
-have obscured a type inconsistency, \eg{}~\li{let f(g : }\ \li{) => g 3 in f 4}.
+have obscured a type inconsistency, \eg{}~\lstinline{let f(g : }\ \lstinline{) => g 3 in f 4}.
 This term is well-typed according to the static semantics of Track 1 because the
@@
 
-put Seidel here
+Understanding and debugging static type errors is notoriously difficult,
+particularly for novices.
+%
+A variety of approaches have been
+proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
+and explain type errors.
+%
+One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
+that demonstrates a run-time failure, and then displays a compressed execution
+trace to the user as a graph.
+%
+In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
+non-empty holes) as far as possible, in particular, until they
+would go wrong.
+%
+Thus, our operational semantics provides a similar benefit to the approach of
+\citet{Seidel2016}.
 
 \parahead{Debugging}
+This approach is reminiscent of the workflow that debuggers make available using breakpoints \cite{fitzgerald2008debugging,DBLP:journals/jfp/TolmachA95}. 
+The difference here is twofold.
+First, evaluation does not stop at each breakpoint and so it is straightforward to explore the \emph{space} of values that a variable takes.
+Second, breakpoints make available the values of a variable at a position in the evaluation trace.
+Hole closures, on the other hand, convey information from a syntactic position in the result of evaluation.
+The result is, by nature, a simpler object than the full evaluation trace that produced the result.
+These approaches are fundamentally complementary, rather than opposed.\todo{move to related work?}{}\todo{talk about tracing/recording}{}
 

### 2018-04-17 appendix-defns.tex
submission

@@
   and $\isIndet{\dexp}$
-  then either\todo{expand this to have one case per type}
+  then either
   \begin{enumerate}[nolistsep]
@@
 \end{proof}
+
+We exclude these proofs and definitions from the Agda mechanization
+for two reasons.
+%
+First, fill-and-resume is merely an optimization, and unlike the meta
+theory of \Secref{sec:calculus}, these properties are generally not
+conserved by certain reasonable extensions of the core
+calculus~(e.g., reference cells and other non-commuting effects).
+%
+Second, to properly encode the hole filling operation, such a
+mechanization requires a significantly more complex representation of
+hole environments; unfortunately, Agda cannot be easily convinced that
+the definition of hole filling is well-founded (\citet{Nanevski2008}
+establish that it is in fact well-founded).
+%
+By contrast, the developments in \Secref{sec:calculus} do not require
+these more complex (and somewhat problematic) representations.
+
+\subsection{Confluence, and Friends}\label{sec:confluence}
+There are various ways to encode the intuition that ``evaluation order does not matter''. One way to do so is
+by establishing a confluence property (which is closely related to
+the Church-Rosser property \cite{church1936some}).
+
+The most general confluence property does not hold for the dynamic
+semantics in Sec.~\ref{sec:calculus} for the usual reason: We do not
+reduce under binders (\citet{DBLP:conf/birthday/BlancLM05} discuss the
+standard counterexample).
+%
+We could recover confluence by specifying reduction under binders,
+either generally or in a more restricted form where only closed
+sub-expressions are
+reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit}.
+%
+However, reduction under binders conflicts with the standard implementation approaches
+for most programming languages \cite{DBLP:conf/birthday/BlancLM05}.
+%
+A more satisfying approach considers confluence modulo equality \cite{Huet:1980ng}. 
+%
+The simplest such approach restricts our interest to top-level expressions
+of base type that result in values, in which case the following
+special case of confluence does hold (trivially when the only base
+type has a single value, but also more generally for other base
+types).
+\begin{lem}[Base Confluence]
+  If $\hasType{\Delta}{\emptyset}{\dexp}{b}$ and 
+  $\multiStepsTo{\dexp}{\dexp_1}$
+  and $\isValue{\dexp_1}$
+  and $\multiStepsTo{\dexp}{\dexp_2}$
+  then $\multiStepsTo{\dexp_2}{\dexp_1}$.
+\end{lem}
+We can then prove the following property, which establishes that fill-and-resume is sound.
+\begin{thm}[Resumption]
+  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp}{b}$
+  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ 
+  and $\multiStepsTo{\dexp}{\dexp_1}$
+  and $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp}}{\dexp_2}$
+  and $\isValue{\dexp_2}$
+  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}{\dexp_2}$.
+  \begin{proof}
+    By Commutativity,
+    $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp}}
+                  {\instantiate{\dexp'}{u}{\dexp_1}}$.
+    By Base Confluence, we can conclude.
+  \end{proof}
+\end{thm}

### 2018-04-17 calculus.tex
sec2!
submission

@@
 % !TEX root = hazelnut-dynamics.tex
-
-\clearpage
+% \clearpage
 \newcommand{\calculusSec}{Hazelnut Live, Formally}
@@
 \label{sec:calculus}
-
 % \def \TirNameStyle \Vtexttt{#1}{{#1}}
@@
 
-We will now make the intuitions developed in the previous section formally precise by specifying a core calculus, which we call \HazelnutLive, and its accompanying metatheory.
+We will now make the intuitions developed in the previous section formally precise by specifying a core calculus, which we call \HazelnutLive, and developing its metatheory.
 
@@
 (see Sec.~\ref{sec:agda-mechanization}).
-Rule names, e.g.
-\rulename{SVar}, correspond to the names used in the mechanization.
+Rule names in this section, e.g.
+\rulename{SVar}, correspond to variable names used in the mechanization.
 We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).
@@
 %
-Non-empty expression holes therefore internalize the ``red underline'' that many editors display under or around type inconsistencies in a program.
+Non-empty expression holes therefore internalize the ``red underline/outline'' that many editors display around type inconsistencies in a program.
 
@@
 %\matt{Extraneous and uninteresting:}
-By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other simple extensions to the calculus\todo{do this, say which ones}.
+By contrast, \citet{popl-paper} instead defined a number type with a single operation.
+That paper also defined sum types as an extension to the core calculus.
+We follow suit on both counts in Appendix~\ref{sec:extensions}.
 %
@@
 The following theorem establishes that an expansion exists for every well-typed external expression. 
-The mechanization also establishes that ywhen an expansion exists, it is unique (not shown).
+The mechanization also establishes that when an expansion exists, it is unique (not shown).
  \begin{thm}[Expandability] \label{thm:expandability}~
@@
 --- Rules \rulename{TAConst}, \rulename{TAVar} and \rulename{TALam}
---- mirror the bidirectional typing rules from Fig.~\ref{fig:bidirectional-typing}.
+--- and in turn, the corresponding bidirectional typing rules from Fig.~\ref{fig:bidirectional-typing}.
 %
@@
 The rules governing hole expansion, and the rules that perform \emph{cast insertion}---those governing
-function application and type ascription---are more interesting. We consider each of these two groups of rules, in turn, below.
+function application and type ascription---are more interesting.
+Let us consider each of these two groups of rules in turn in Sec.~\ref{sec:hole-expansion} and Sec.~\ref{sec:cast-insertion}, respectively.
 
@@
 %
-For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}{}.
-%
-There, we number each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.
+For example, the result from Fig.~\ref{fig:grades-example} shows three closures for the hole named 1.
+%
+There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and \li{1:3}, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.
 
@@
 %
-Without any cast, the type of variable~$x$ is hole type~$\tehole$;
+Without any cast, the type of variable~$x$ is the hole type~$\tehole$;
 however, the inserted cast on~$x$ permits treating it as though it has
-arrow type $\tarr{\tehole}{\tehole}$, as described
-in \Secref{sec:external-statics}.
+arrow type $\tarr{\tehole}{\tehole}$.
 %
@@
 %
-We say more after discussing type safety below.\todo{do this, or change to related work}{}
+We say more in Sec.~\ref{sec:relatedWork}.
 %
@@
 
-Figures~\ref{fig:isGround}-\ref{fig:step} specify the dynamic semantics. 
+Figures~\ref{fig:isGround}-\ref{fig:step} define the dynamic semantics. 
 %
@@
 \footnote{
-        In most accounts of the cast calculus, values and ground types
-        are distinguished with separate grammars together with an
+        Most accounts of the cast calculus distinguish ground types and values
+        with separate grammars together with an
         implicit identification convention.
@@
 These forms are irreducible because they represent values that have been boxed 
-but have never flowed into a corresponding ``unboxing'' operation, discussed below. 
+but have never flowed into a corresponding ``unboxing'' cast, discussed below. 
 %
@@
 Below, we describe failed casts and the remaining indeterminate forms simultaneously
-with the corresponding transition rules.\todo{relationship to weak head normal forms}
-
-Figures~\ref{fig:instruction-transitions}-\ref{fig:step} define\todo{use define instead of specify everywhere?}{} the transition rules. 
-%
-Top-level transitions are \emph{steps}~$\stepsToD{}{d}{d'}$, governed by Rule~\rulename{Step} in \Figref{fig:step}, which
+with the corresponding transition rules.
+
+Figures~\ref{fig:instruction-transitions}-\ref{fig:step} define the transition rules. 
+%
+Top-level transitions are \emph{steps}, $\stepsToD{}{d}{d'}$, governed by Rule~\rulename{Step} in \Figref{fig:step}, which
 %
@@
 
-The rules take care to maintain the property that final expressions cannot take a step.%
-\todo{prove this as stated from Ian's disjointness proofs}{}%
+The rules maintain the property that final expressions truly cannot take a step.%
 \begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
@@
 %
-Rule \rulename{ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the
+Rule \rulename{ITBeta} in Fig.~\ref{fig:instruction-transitions} defines the
 standard beta reduction transition.
@@
 Consequently, the environments of different closures with the same hole name may differ, 
-e.g., when an reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.%
-\todo{show example here or refer back to sec 2?}{} 
-%
+e.g., when an reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}. 
 Hole closures may also appear within the environments of other hole
 closures, giving rise to the closure paths described in
-Sec.~\ref{sec:paths}\todo{where?}{}.
+Sec.~\ref{sec:paths}.
 
@@
 $\tarr{\htau_1}{\htau_2}$, and that~$d_2$ has type~$\htau_1'$, where
-$\tconsistent{\htau_1}{\htau_1'}$; consequently, we maintain type
+$\tconsistent{\htau_1}{\htau_1'}$. 
+Consequently, we maintain type
 safety by placing a cast on~$d_2$ from~$\htau_1'$ to~$\htau_1$.
@@
 in \Figref{fig:isFinal}), and the application reduces no
-further.\todo{mention WHNF here?}
+further.
 
@@
 \subsubsection{Casts}
-Rule \rulename{ITCastId} drops identity casts. The remaining instruction
+Rule \rulename{ITCastId} strips identity casts. The remaining instruction
 transition rules assign meaning to non-identity casts. 
@@
 As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term
-cast to hole type is statically obscure,
-%
-so we must wait for \emph{use} of the term at some other type, via a
-cast away from hole type, to detect the type error dynamically.
+cast \emph{to} hole type is statically obscure,
+%
+so we must await a \emph{use} of the term at some other type, via a
+cast \emph{away} from hole type, to detect the type error dynamically.
 %
@@
 Rule \rulename{IFailedCast} specifies that a failed cast operates as an
-indeterminate form (once $d$ is final), i.e. evaluation does not stop.\todo{talk about blame somewhere}{}
-
-Rules \rulename{TAFailedCast} and \rulename{IFailedCast} each operate at ground type. 
+indeterminate form (once $d$ is final), i.e.
+evaluation does not stop.
+For simplicity, we do not include blame labels as found in some accounts of gradual typing \cite{DBLP:conf/snapl/SiekVCB15}, but it would be straightforward to do so by recording the blame labels from the two constituent casts on the two arrows of the failed cast form.
+
+Rules \rulename{ITCastSucceed} and \rulename{ITCastFail} rules just described only operate at ground type. 
 %
@@
 These rules serve as technical devices, permitting us to restrict our
-interest exclusively to ground types.
+interest exclusively to casts involving ground types and type holes elsewhere. 
 %
@@
 %
-In particular, Rule \rulename{ICastHoleGround} handles casts that are not of
-the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$, from hole to
-ground type.
-%
-(In the cast calculus, there exist no such
-irreducible terms; the only canonical form of type $\tehole$
-is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.)
+Of note, Rule \rulename{ICastHoleGround} handles casts from hole to
+ground type that are not of
+the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$.
+%
+The need for this case is perhaps not obvious because in the cast calculus,  the only irreducible term of type $\tehole$
+is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.
 
@@
 The proof relies on an analogous preservation lemma for instruction
-transitions and a standard substitution lemma (stated in
-Appendix~\ref{sec:additional-defns}).
-%
-Hole closures can disappear during evaluation, so we rely on weakening
+transitions and a standard substitution lemma stated in
+Appendix~\ref{sec:additional-defns}.
+%
+Hole closures can disappear during evaluation, so we rely on simple weakening
 of $\Delta$.
@@
 Although this paper focuses on running \emph{incomplete} programs, it helps
-to know that its machinery does not interfere
-with reasoning about the behavior of \emph{complete} programs, i.e. those
-with no holes.
-%
-Appendix~\ref{sec:additional-defns} defines the judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
-%
-Failed casts cannot appear in complete internal expressions. 
+to know that the necessary machinery does not interfere
+with running \emph{complete} programs, i.e. those
+with no type or expression holes.
+%
+Appendix~\ref{sec:additional-defns} defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$. 
+%
+Of note, failed casts cannot appear in complete internal expressions. 
 %
 The following theorem establishes that expansion preserves program completeness.
+
+% TODO Delta is always empty too!
 \begin{thm}[Complete Expansion] ~
@@
     \item
-      If $\isComplete{\hexp}$
+      If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
       and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
-      then $\isComplete{\dexp}$.
+      then $\isComplete{\htau}$ and $\isComplete{\dexp}$.
     \item
-      If $\isComplete{\hexp}$
+      If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$ 
+and $\isComplete{\htau}$
       and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
@@
 
-The supplemental material includes our Agda-based
-mechanization  \cite{norell2009dependently,norell:thesis} 
+The supplemental material includes our Agda 
+mechanization  \cite{norell2009dependently,norell:thesis,Aydemir:2005fk} 
 of the semantics and metatheory of \HazelnutLive, 
 %including proofs of all of the 
-including the theorems stated above and necessary lemmas.
-We take as assumptions only a few standard properties of general hypothetical judgements and finite maps.\todo{what about hole disjointness}
+including all of the theorems stated above and necessary lemmas.
+We take as assumptions only a few standard properties of general hypothetical judgements and finite maps, and, to avoid excessive machinery related to hole name tracking in the bidirectional typing rules, we axiomatize some hole name uniqueness assumptions made implicitly in this paper.
 %
@@
 %
-Our approach is standard: we model judgements as 
-inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite something?}{}. 
-%
-We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
-contexts, hole contexts and hole environments using metafunctions.\todo{say something about funext?}{}
+Our approach is otherwise standard: we model judgements as 
+inductive datatypes, and rules as dependently typed constructors of these judgements. 
+%
+We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names and encode typing 
+contexts, hole contexts and finite substitutions using metafunctions.
+To support this encoding choice, we postulate function extensionality (which is independent of Agda's axioms) \cite{awodey2012inductive}. 
+The documentation provided with the mechanization has more technical details.
 
@@
 The supplemental material includes a browser-based implementation
-of \HazelnutLive, which consists of a functional reactive program
+of \HazelnutLive. 
+% The implementation supports the core calculus of this section but with the notional base type $b$ replaced by the $\tnum$ type from Appendix~\ref{sec:extensions}.
+The implementation also includes the sum types extension from Appendix~\ref{sec:extensions} and a few minor  conveniences, e.g.
+\li{let} binding.
+All of the live programming features from \Secref{sec:examples} are available in the implementation essentially as shown, but for this more austere language (with some minor conveniences, notably let binding).
+Appendix~\ref{sec:impl-screenshots} provides screenshots of the full user interface.
+
+ % which is a functional reactive program \cite{Elliott:1997jh,DBLP:conf/pldi/CzaplickiC13}
 %
@@
 %\cite{DBLP:conf/pldi/CzaplickiC13}
-%
-written with the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
+
+The implementation is written with the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
 together with the OCaml \lismall{React} library \cite{OcamlReact} 
 and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}.
-%
-Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{} provides a screenshot of the user interface, 
-which implements the core calculus of this section, 
-along with the corresponding live programming features from \Secref{sec:examples}.
-
-To facilitate interactively editing terms, the implementation employs
-the editor semantics of~\Hazelnut, exposing a language of structured
+This follows the implementation of \Hazel, which, although tracking toward an Elm-like semantics, is implemented in OCaml because the Elm compiler is not yet self-hosted.
+The implementation of the dynamic semantics consists of a simple evaluator that closely follows the rules specified in this section.
+
+The editor component of the implementation is derived
+from the structure editor calculus of~\Hazelnut, but with support for more natural cursor-based movement and infix operator sequences (the details are beyond the scope of this paper).
+It exposes a language of structured
 edit actions (summarized in the sidebar on the left of
-Fig.~\ref{fig:screenshot})
-%
-As with prior work, this editor semantics enjoys
-the \Property{Sensibility} property by automatically inserting holes
+the screenshots in Appendix~\ref{sec:impl-screenshots}) that automatically inserts holes as necessary 
 to guarantee that every edit state has some (possibly incomplete) type
-.\todo{put statement in the appendix}{}
-
-By composing this meta-theoretic property with
+. This corresponds to the Sensibility invariant established by \citet{popl-paper}.  
+By composing this metatheoretic property with
 the \Property{Expandability}, \Property{Typed Expansion}
 and \Property{Progress} properties from this paper, we establish a
-uniquely powerful invariant: \emph{Every} edit state has a \emph{static
+uniquely powerful invariant: \emph{every} edit state has a \emph{static
 meaning} (a la~\Hazelnut) and a \emph{dynamic meaning} (a la~\HazelnutLive).
@@
 %
-The current implementation serves as a proof-of-concept of this invariant, 
-by providing a working prototype for this section's definitions.
+The implementation is intended as a proof-of-concept of this invariant, and as a companion to this paper.
+
+% We make no further claims about the usability or practicality of the implementation (and indeed, there remain many important and decidedly unresolved questions along these lines, which we leave beyond the scope of this paper).
+% We also reiterate that the essential ideas developed from type-theoretic first principles in this paper do not require that
+% the editor component of the programming environment be implemented as a structure editor (see Sec.~\ref{sec:intro}).
+ %,  to scale up these ideas to a ``real-world'' language and programming environment.
 
@@
 In any case,
-% When the programmer performs one or more edit actions to fill in a hole in the program, a new result must be computed.
+When the programmer performs one or more edit actions to fill in a hole in the program, a new result must be computed, ideally quickly \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90}.
 Na\"ively, the system would need to compute the result ``from scratch'' on each such edit.
-A more efficient approach supported by the structure of \HazelnutLive is to resume evaluation from where it left off after an edit that amounts to hole filling.
-We call this feature \emph{fill-and-resume}, by analogy to the ``edit-and-resume'' features available in some systems, e.g.
-Visual Studio (see below for a comparison)\todo{cite and compare}{}.
-
-\todo{cite tanimoto paper, mention Level 4 liveness property}When the programmer interactively fills a program hole, that program's
-reduction and result are generally each affected, and the system
-should respond by (quickly) updating the program's output.
-%
-Na\"ively, this response requires recomputing the program output
-``from scratch'' after each such edit.
-%
-However, by virtue of its design principles (based on
-CMTT), \HazelnutLive offers a more efficient approach: To \emph{resume
-evaluation} of previously indeterminate expressions, which are
-(generally) more determined after the programmer fills a hole than
-beforehand.
-%
-This \emph{fill-and-resume} behavior formalizes the (ad hoc)
-``edit-and-resume'' behavior permitted by some current systems,
-e.g.,~Microsoft Visual Studio.\todo{cite}
-
-%(see below for a comparison)\todo{cite and compare}{}. ---NO TIME/SPACE for this.
+For small exploratory programming tasks, recomputation is acceptable, but in cases where a large amount of computation might occur, e.g.
+in data science tasks, a more efficient approach is to resume evaluation from where it left off after an edit that amounts to hole filling.
+This section briefly discusses the foundations of this feature, which we call \emph{fill-and-resume}.
+This approach is complementary to incremental computing techniques~\cite{Hammer2014}.
 
@@
 %
-we interpret hole environments as \emph{delayed substitutions},
-the same interpretation suggested for closures in contextual modal
+the key idea is to interpret hole environments as \emph{delayed substitutions}. This is the same interpretation suggested for closures in contextual modal
 type theory (CMTT) by \citet{Nanevski2008}.
@@
 %
-We define hole filling in terms of the expanded (internal) program,
-and define filling a hole in an external language term as filling the
-corresponding hole in its expansion.
-
 %%%%%%%%% New par
-\Figref{fig:substitution} defines the hole filling operation~$\instantiate{d}{u}{d'}$,
-mirroring the contextual substitution operation of~CMTT.
+\Figref{fig:substitution} defines the hole filling operation~$\instantiate{d}{u}{d'}$ 
+based on the contextual substitution operation of~CMTT.
 %
@@
 %
-%This case shows why we cannot interpret a non-empty hole as an empty
-%hole of arrow type applied to the enveloped expression---the hole
-%filling operation would not operate as expected under this
-%interpretation.
+This case shows why we cannot interpret a non-empty hole as an empty
+hole of arrow type applied to the enveloped expression---the hole
+filling operation would not operate as expected under this
+interpretation.
 
@@
   then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
-\begin{proof}
-We prove a stronger version of the theorem which accounts for both
-terms and finite substitutions (see \cite{Nanevski2008}).
-%
-Each case proceeds by rule induction on the first assumption, applying
-the finite substitution lemma.
-\end{proof}
-\end{thm}
-
-Dynamically, the fill-and-resume enjoys an
-important \emph{commutativity} property: If there is some sequence of
+\end{thm}
+
+Dynamically, the correctness of fill-and-resume depends on
+the following \emph{commutativity} property: if there is some sequence of
 steps that go from $d_1$ to $d_2$, then one can fill a hole in these
@@
 %
-Critically, this property relies on the more permissive version of
-reduction from \Secref{sec:calculus}, where evaluation order is
+Critically, this property relies on the more version of
+the semantics from \Secref{sec:calculus} where evaluation order is
 unspecified (i.e. formally, we omit the bracketed premises).
@@
 In other words, this notion of fill-and-resume only works for
-languages where evaluation order always results in \emph{confluent}
-reductions; viz., those that \emph{eventually} yield the same
-resulting term.
+languages where evaluation order ``does not matter''. 
+(There are various standard ways to formalize this intuition.
+For the sake of space, we review these in Appendix~\ref{sec:confluence}.)
 
@@
 %%
-%There are various ways to encode this intuition. One way to do so is
-%by establishing a confluence property (which is closely related to
-%the Church-Rosser property \cite{church1936some}).
-
-The most general confluence property does not hold for the dynamic
-semantics in Sec.~\ref{sec:calculus} for the usual reason: We do not
-reduce under binders (\citet{DBLP:conf/birthday/BlancLM05} discuss the
-standard counterexample).
-%
-We could recover confluence by specifying reduction under binders,
-either generally or in a more restricted form where only closed
-sub-expressions are
-reduced \cite{DBLP:journals/tcs/CagmanH98,DBLP:conf/birthday/BlancLM05,levy1999explicit}.
-%
-However, reduction under binders conflicts with the standard implementation approaches
-for most programming languages \cite{DBLP:conf/birthday/BlancLM05}.
-%
-A more satisfying approach considers confluence modulo equality \cite{Huet:1980ng}. 
-%
-The simplest such approach restricts our interest to top-level expressions
-of base type that result in values, in which case the following
-special case of confluence does hold (trivially when the only base
-type has a single value, but also more generally for other base
-types).
-\begin{lem}[Base Confluence]
-  If $\hasType{\Delta}{\emptyset}{\dexp}{b}$ and 
-  $\multiStepsTo{\dexp}{\dexp_1}$
-  and $\isValue{\dexp_1}$
-  and $\multiStepsTo{\dexp}{\dexp_2}$
-  then $\multiStepsTo{\dexp_2}{\dexp_1}$.
-\end{lem}
-We can then prove the following property, which establishes that fill-and-resume is sound.
-\begin{thm}[Resumption]
-  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp}{b}$
-  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ 
-  and $\multiStepsTo{\dexp}{\dexp_1}$
-  and $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp}}{\dexp_2}$
-  and $\isValue{\dexp_2}$
-  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}{\dexp_2}$.
-  \begin{proof}
-    By Commutativity,
-    $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp}}
-                  {\instantiate{\dexp'}{u}{\dexp_1}}$.
-    By Base Confluence, we can conclude.
-  \end{proof}
-\end{thm}
-
-We outline the proofs in Appendix~\ref{sec:hole-filling}, along with a
-number of other necessary lemmas and definitions.
-%
-In particular, additional machinery handles the situation wherein a
+
+We describe the proof, which involves a 
+number of lemmas and definitions, in Appendix~\ref{sec:hole-filling}. 
+%
+In particular, care is needed to handle the situation where a
 now-filled non-empty hole had taken a step in the original evaluation.
-\matt{Cyrus --- I tried to simplify and shorten the earlier version of the prior sentence, but I \emph{may} have unintentionally goofed up what it was trying to say; please double check the prior sentence}
-
-\matt{We could (should?) move the next paragraph to the appendix, 
-before/after the proofs mentioned above; 
-its not really adding much to the discussion above, 
-but may be interesting to the very studious reader/prover; also, leaving it here also gives a lazy/cranky reviewer a nice weapon to hit us with}
-
-We exclude these proofs and definitions from the Agda mechanization
-for two reasons.
-%
-First, fill-and-resume is merely an optimization, and unlike the meta
-theory of \Secref{sec:calculus}, these properties are generally not
-conserved by certain reasonable extensions of the core
-calculus~(e.g., \matt{such as what?}).
-%
-Second, to properly encode the hole filling operation, such a
-mechanization requires a significantly more complex representation of
-hole environments; unfortunately, Agda cannot be easily convinced that
-the corresponding definition is well-founded (\citet{Nanevski2008}
-establish that it is in fact well-founded).
-%
-By contrast, the developments in \Secref{sec:calculus} do not require
-these more complex (and somewhat problematic) representations.
+
+% \matt{We could (should?) move the next paragraph to the appendix, 
+% before/after the proofs mentioned above; 
+% its not really adding much to the discussion above, 
+% but may be interesting to the very studious reader/prover; also, leaving it here also gives a lazy/cranky reviewer a nice weapon to hit us with}
+
+We do not separately define hole filling in the external language 
+(i.e. we consider a change to an external expression to be a hole filling if the new expansion differs from 
+the previous expansion up to hole filling). 
+In practice, it may be useful to cache more than one recent edit state
+to take full advantage of hole filling. 
+As an example, consider two edits, the first filling a hole~$u$ with
+the number~$2$, and the next applying operator~$+$, resulting in $2
++ \dehole{v}{\sigma}{}$.
+%
+This second edit is not a hole filling edit with respect to the
+immediately preceding edit state, $2$, but it can be understood as filling
+hole $u$ from two states back with $2 + \dehole{v}{\sigma}{}$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+
+Hole filling also allows us to give a contextual modal interpretation
+to lab notebook cells  like those of
+Jupyter/IPython \cite{PER-GRA:2007} (and read-eval-print loops as a restricted case
+where edits to previous cells are impossible). 
+%
+Each cell can be understood as a series of \li{let}
+bindings ending implicitly in a hole, which is filled by the next cell.
+The live environment in the subsequent cell is exactly the hole environment of this implicit trailing hole. 
+%
+Hole filling when a subsequent cell changes avoids recomputing the
+environment from preceding cells, without relying on mutable state. Commutativity provides a reproducibility guarantee 
+missing from Jupyter/IPython, where editing and executing previous cells can cause the state to differ substantially from the state that would result when attempting to run the notebook from the top.
+% TODO citation
 

### 2018-04-17 examples.tex
submission

@@
 exposition, we have ``post-processed'' the examples in this section to make use of several convenient language mechanisms from \Elm~that are not yet implemented in \Hazel as of this writing.
-These mechanisms are orthogonal to the contributions of this paper.
-We have implemented all of the live programming features described in this section, both in a branch of \Hazel and in the implementation of the core calculus that we describe in Sec.~\ref{sec:implementation} and include in the supplement.
+These mechanisms are orthogonal to the contributions of this paper---we have implemented all of the live programming features described in this section, both in \Hazel and in the implementation of the core calculus included in the supplement (see Sec.~\ref{sec:implementation}).
 

### 2018-04-17 extensions.tex
draft of sec 3.0 through 3.1
start of sum types extension
first draft of numbers extension; fix some typos in sumtypes extension
'isIndet' ~> 'isFinal'; start on product types
submission

@@
 % !TEX root = hazelnut-dynamics.tex
+\clearpage
 \newcommand{\extensionsSec}{Extensions to Hazelnut Live}
@@
 \label{sec:extensions}
-\subsection{Product Types}
 
-\subsection{Sum Types}
+We give two extensions here, numbers and sum types, to maintain parity with \Hazelnut as specified by \citet{popl-paper}.
 
-\subsection{Recursive Types}
+It is worth observing that these extensions do not make explicit mention of expression holes.
+The ``non-obvious'' machinery is almost entirely related to casts.
+Fortunately, there has been excellent work of late on generating ``gradualized'' specifications from standard specifications \cite{DBLP:conf/popl/CiminiS16}.
+The extensions below, and other extensions of interest, closely follow the output of the gradualizer: \url{http://cimini.info/gradualizerDynamicSemantics/} (which, like our work, is based on the refined account of gradual typing by \cite{DBLP:conf/snapl/SiekVCB15}).
+The rules for indeterminate forms are mainly where the gradualizer is not necessarily sufficient.
 
-\subsection{Polymorphism}
+\include{numbers} %(from Hazelnut)
 
-\subsection{Mutable References}
-\cy{if we add references, would have to talk about how commutativity isn't conserved}
+%\subsection{Product Types}
+%\include{prodtypes}
+
+\include{sumtypes}
+
+%\subsection{Recursive Types}
+%\subsection{Polymorphism}
+%\subsection{Mutable References}
+%\cy{if we add references, would have to talk about how commutativity isn't conserved}

### 2018-04-17 intro.tex
sec2!
submission

@@
 it relies on the most recent complete editor state, creating a perceptive gap because the feedback may not accurately reflect what the programmer is actually seeing and writing in the editor.
-\todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}
 
@@
 which provides a logical foundation for hole environment tracking). 
-These non-obvious connections to well-established existing systems help support our claim that this approach is theoretically well-grounded\todo{note that our exposition \IS self-contained here?
-or later?}.
+These non-obvious connections to well-established existing systems help support our claim that this approach is theoretically well-grounded.
 

### 2018-04-17 related-work.tex
submission

--- 2018-04-16 related-work.tex
@@
 % !TEX root = hazelnut-dynamics.tex
-\newcommand{\relatedWorkSection}{Related Work}
+\newcommand{\relatedWorkSection}{Related and Future Work}
 \section{\protect\relatedWorkSection} % don't like the all-caps thing that the template does, so protecting it from that
@@
 
-Program editors frequently encounter program text that is not well-formed with
-respect to the concrete syntax of the language \cite{popl-paper}.
-%
-Within the simplest editors, editor services that take well-formed program text
-or an abstract syntax tree as input are simply disabled when this occurs.
-%
-More sophisticated editors, including those generated by modern editor
-generators like Spoofax \cite{DBLP:conf/oopsla/KatsV10}, deploy \emph{error
-recovery} heuristics that silently insert tokens or discard tokens from the
-program so that the editor-internal representation is a standard abstract syntax
-tree
-\cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09}.
-%
-These heuristics are typically defined manually by the grammar designer, though
-certain heuristics can be generated semi-automatically by tools that are given a
-description of the scoping conventions of the language and of secondary
-notational conventions (\eg{}~whitespace conventions)
-\cite{DBLP:conf/oopsla/KatsJNV09,DBLP:conf/sle/JongeNKV09}.
-%
-These error recovery heuristics require guessing at the programmer's intent, and
-result in tokens being inserted or deleted from the program invisibly, so they
-are \emph{ad hoc} and can confuse the programmer
-\cite{DBLP:conf/oopsla/KatsJNV09}.
+This paper defined a dynamic semantics for incomplete functional programs
+based directly on the static semantics developed by \citet{popl-paper}.
+A subsequent ``vision paper'' introducing \Hazel suggested the need for 
+a corresponding dynamic semantics for the purposes of live programming \cite{HazelnutSNAPL}. 
+This paper    
+contributes the type-theoretic first steps necessary to achieve this long-term vision.
 
-\parahead{Hole-Driven Development}
+The semantics borrows substantially
+from the theory of type consistency and casts developed in gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}, as discussed
+at length in Sec.~\ref{sec:calculus}. The main innovation relative to this
+prior work is the treatment of cast failures as holes, rather than errors.
+Many of the methods developed to make gradual typing more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein.
+For example, there has been substantial work on the problem of implementing casts efficiently, e.g.
+by \citet{herman2010space}, and on integrating gradual typing with polymorphism, e.g.
+by \cite{garcia2015principal}.
+Another interesting direction that we leave to future work is the integration of these ideas into dependently typed proof assistants,
+where evaluation is at the service of equational reasoning.
 
-A cleaner alternative approach, and the approach that is central to our proposed
-research, is to add \emph{holes} to both the concrete and abstract syntax of the
-language.
-%
-Syntax trees with holes first arose in prior work on \emph{structure editors},
-which are program editors where every edit action directly causes changes to the
-abstract syntax tree of the program, rather than going indirectly through a
-parser.
-%
-Structure editors have a long history: the Cornell Program Synthesizer was
-developed in the early 1980s \cite{teitelbaum_cornell_1981} and there has been a
-steady stream of new designs in this space
-(\eg{}~\cite{ko_barista:_2006,Resnick:2009:SP:1592761.1592779,voelter_mbeddr:_2012,tillmann_touchdevelop:_2011,lamdu,ActiveCodeCompletion,Greenfoot,Deuce}).
-%
-For example, Scratch and other block-based editors have achieved success as a
-tool for teaching children how to program
-\cite{Resnick:2009:SP:1592761.1592779}. \texttt{mbeddr} is an editor for a
-C-like language \cite{voelter_mbeddr:_2012}, built using the commercially
-supported MPS structure editor workbench \cite{voelter2011language}, that is
-presently deployed to a number of customers.
-%
-TouchDevelop is a structure editor for an object-oriented language designed for
-use on mobile devices \cite{tillmann_touchdevelop:_2011}. Lamdu \cite{lamdu},
-Unison \cite{unison} and  isomorf~\cite{isomorf} are open source structure
-editors for functional languages similar to ML.
+Another important future direction is to move beyond
+pure functional programming a la \Elm~ and carefully integrate imperative features, e.g.
+ML-style references.
+These require particular care, as recognized by \cite{Siek06a}, though we see no reason why the type safety properties  established in this paper would not be conserved in an effectful setting.
+Going beyond references to incorporate general effects, e.g.
+network and IO effects, raises some important practical concerns, however---we do not want to continue past a hole or error and in so doing haphazardly trigger an unintended effect.
+Developing language features and user interface features that give the programmer fine-grained control over evaluation when it encounters a hole or cast failure are likely to be particularly helpful in these settings.
 
-%% There is reason to be optimistic, however, with recent studies 
-%% suggesting that programmers experienced with a modern keyboard-driven structure editor like \texttt{mbeddr} 
-%% can be highly productive \cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14}.
+The other major pillar of related work is the work on contextual modal type theory (CMTT) \cite{Nanevski2008}, which we also discussed at length throughout the paper.
+To reiterate, there is a close relationship between holes in this paper and metavariables in CMTT.
+Hole contexts correspond to modal contexts.
+Empty hole closures 
+relate to the concept of a \emph{metavariable closure} in CMTT, which consists
+of a metavariable paired with a substitution for all of the variables in the
+typing context associated with that metavariable. Hole filling relates to contextual substitution.
 
-Nevertheless, text-based program editors continue to dominate in practice, in
-part because text is flexible and familiar and because tools for interacting
-with text, like regular expressions and various differencing techniques used by
-version control systems, are already very well-developed.
-%
-Consequently, researchers have also explored various ``hybrid'' approaches,
-which incorporate holes into an otherwise text-based program editor.
-%
-For example, recent iterations of the Glasgow Haskell Compiler (GHC)
-\cite{GHCWIKI}, as well as the dependently typed proof assistants
-Idris~\cite{brady2013idris} and Agda~\cite{norell2009dependently}, support a
-workflow where the programmer places holes manually at locations in the program
-that remain under construction. Agda and Idris also allow the programmer to put
-terms inside holes, to indicate that the term may not yet be well-typed or
-otherwise finished.
-%
-Recent work on \emph{syntactic placeholders} takes some of the tedium out of
-this approach by providing low-level edit actions that cause these holes (called
-placeholders in their paper) to be inserted automatically \cite{Amorim2016}. 
-%
-This approximates the experience of a structure editor in common usage, while
-allowing the programmer to perform arbitrary text edits when necessary.
-%
-Expression holes also appear in the context of program synthesis, serving as
-placeholders in \emph{templates}~\cite{srivastava2013template} or
-\emph{sketches}~\cite{solar2009sketching} to be filled in by an expression
-synthesis engine.
-
-\begin{itemize}
-	%% \item Agda
-	%% \item Idris
-	%% \item GHC holes
-	%% \item Visual Studio (and others) support for edit-and-resume
-	\item Scratch lets you just skip over statement holes
-	\item papers that show up in a search for ``typed holes'' -- \url{https://scholar.google.com/scholar?hl=en&as_sdt=0%2C39&q=%22typed+holes%22&btnG=}
-\end{itemize}
-
-\parahead{Structure Editor Calculus}
-
-Most work on structure editors has focused on the user interfaces that they
-present. This is important work---presenting a fluid user interface involving
-exclusively structural edit actions is a non-trivial problem that has not yet
-been fully resolved, though recent studies have started to show productivity
-gains for keyboard-driven structure
-editors~\cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14}.
-
-We emphasize
-that our proposed contributions will be relevant no matter how holes come to be
-inserted into the program. % We will say more about all of these systems later.
-
-two main remaining questions (beyond scaling up the calculus) are how to run
-programs with type and expression holes.
-
-\begin{itemize}
-	\item Hazelnut paper
-\end{itemize}
-
-\parahead{Gradual Typing}
-
-As already discussed in Track 1, type holes are closely related to the unknown
-types of gradual type theory \cite{Siek06a}. This has implications for our
-dynamic semantics as well---we need to be able to run programs where type holes
-have obscured a type inconsistency, \eg{}~\li{let f(g : }\ \li{) => g 3 in f 4}.
-This term is well-typed according to the static semantics of Track 1 because the
-argument \li{g} of \li{f} has an unknown type, and so it can be applied to
-\li{3}, but when \li{f} is applied to \li{4}, there is a problem because \li{4 3}
-is meaningless. There are two ways to address the problem. The first is to
-apply ML-style type inference to discover a more specific type schema for
-\li{g}, \ie{}~that it must actually be an arrow type, and then mark the argument
-\li{4} to \li{f} as a type inconsistency using a non-empty hole. We will
-investigate this in Track 1. The second approach is to adopt run-time cast
-insertions from research on gradual type theory. We will investigate this second
-approach in Track 2.
-
-\begin{itemize}
-	\item gradual typing (and dynamic typing)
-  \item siek and taha paper
-  \item snapl15 paper
-  \item gradualizer paper
-  \item maybe other things, e.g. several papers by ron garcia
-\end{itemize}
-
-\parahead{Contextual Modal Type Theory}
-
-compared to the dynamics of type holes (which are based on Grad Typing), the
-dynamics of expression holes are less clear.
-
-Formally, expression holes in our proposed dynamics are closely related to the
-metavariables of contextual modal type theory (CMTT) \cite{Nanevski2008}.
-Metavariables in CMTT are associated with both a type and a typing context.
-Expression holes can similarly be assigned a most specific type and typing
-context. The environment tracking service can then be understood as being
-related to the concept of a \emph{metavariable closure} in CMTT, which consists
-of a metavariable paired with a substitution for all of the variables in the
-typing context associated with that metavariable. Indeed, this neatly explains
-the difference between our proposed dynamics and existing work on staging and
-partial evaluation---that existing work is rooted in modal type theory
-\cite{Davies:2001op} rather than the derivative contextual modal type theory
-(which, to date, has been studied far less extensively.) In other words,
-existing systems are focused on partial evaluation with respect to an input to a
-function, whereas holes can appear in context, \eg{}~directly within the body of
-a function. Recent work on the Beluga theorem prover is also based on CMTT,
-albeit for the purpose of proof refinement and contextual reasoning rather than
-program evaluation \cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
-McBride's OLEG is another system for reasoning contextually about metavariables
-\cite{DBLP:phd/ethos/McBride00}, and it is the basis of certain analagous
-features of Idris \cite{brady2013idris}.
-
-\begin{itemize}
-	\item CMTT
-\end{itemize}
-
-\parahead{Evaluation Order}
-
-\begin{itemize}
-	\item full beta reduction
-	\item prior work on confluence for the lambda calculus
-	\item ``Call-by-value is dual to call-by-name'' might be relevant? \url{http://homepages.inf.ed.ac.uk/wadler/papers/dual/dual.pdf}
-	\item mention how unspecified evaluation order is something that people do when talking about parallelism?
-\end{itemize}
-
-\parahead{Partial Evaluation and Symbolic Execution}
-
-\begin{itemize}
-	\item work on partial evaluation and staging (incl. connections to modal logic -- ``modal analysis of staged computation'')
-	\item symbolic evaluation
-	\item maybe also search for ``partial programs'' 
-\end{itemize}
-
-put Seidel here
-
-\parahead{Debugging}
-
-\begin{itemize}
-	\item work on debuggers that allow you to inspect environments
-  \item might be something relevant in the paper ``A Debugger for Standard ML'' 
-  \item "Visualizing the evaluation of functional programs for debugging" by Whitington and Ridge
-  \item "A lightweight interactive debugger for Haskell'' and ``Multiple-View Tracing for Haskell: a New Hat'' might be relevant
-  \item ocamli -- \url{https://github.com/johnwhitington/ocamli}
-  \item Better supporting debugging aids learning a novel programming language. -- Scaffidi at VLHCC 2017
-  \item quote from Wadler in ``Why no one uses functional languages'':
-    \begin{quote}
-    “...there are few debuggers or
-profilers for strict [functional] languages, perhaps because constructing them is not considered
-research. This is a shame, since such tools are sorely needed, and there remains much of
-interest to learn about their construction and use.
-    \end{quote}
-   \item edit-and-resume in visual studio, others
-\end{itemize}
-
-Visual Studio edit-and-continue \url{https://docs.microsoft.com/en-us/visualstudio/debugger/edit-and-continue}
-
-Hole-like constructs also appear in work on program slicing
-\cite{DBLP:conf/icfp/PereraACL12}, where holes are \rkc{dynamically inserted...}
-
-\parahead{Others?}
-
-\begin{itemize}
-	\item simply typed underdeterminism paper
-	\item DuctileJ stuff -- \url{https://homes.cs.washington.edu/~mernst/pubs/ductile-icse2011.pdf}
-	\item ``Achieving flexibility in direct-manipulation programming environments by relaxing the edit-time grammar'' -- \url{http://ieeexplore.ieee.org/document/1509511/}
-	\item talk about closures in environmental semantics
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 2018.04.14
-%
-% MATT MOVED THIS MATERIAL HERE FROM calculus.tex
-%
-% We should discuss this stuff, but IMO, it should not be necessary to
-% read or appreciate these connections to understand calculus.tex on a
-% first read.  These details are very interesting, but shouldn't be in
-% the main technical section, IMO.
-%
-\subsection{MOVED FROM: Dynamic Semantics}
-
-\matt{This discussion relates our design to CMTT; if we leave it in
-  the paper, we should leave it in this section, whose point is to be
-  comparative; the technical sections lack a sufficent ``attention
-  budget'' for the reader to wade through this discussion as they are
-  still learning the details of our paper.  Even for a really studious
-  reader that already knows CMTT, I still think it makes more sense
-  (organizationally) to give a detailed comparison here, and not
-  there.}
-
+Although these connections are encouraging, our contributions do not neatly fall out from the prior work.
 The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.
@@
 which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
-but under this interpretation an analogous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
+but under this interpretation an analogous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.
 
-Furthermore, we need to integrate casts into the dynamic semantics. 
-Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. 
-The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms.
-The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. 
-Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}\todo{where?}{}, is for cast failure to instead insert a membrane around the dynamic type error, 
-much like a non-empty hole serves as a membrane around a static type error, 
-allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 2018.04.15
-%
-% MATT MOVED THIS MATERIAL HERE FROM calculus.tex
-%
-\subsection{MOVED FROM: Fill-and-resume}
-
-\matt{This discussion relies on a more developed exposition of CMTT
-  that we have/will give in this paper; we should leave it in this
-  section, or remove it}
-
-So far, we have understood hole filling as an extralinguistic operation. 
+It is also worth emphasizing that we use the machinery 
+borrowed from CMTT extralinguistically.
 %
@@
 e.g. the elaborator reflection system of Idris
-\cite{brady2013idris,DBLP:conf/icfp/ChristiansenB16} and the tactic
-system\todo{right? read a bit more}{} of Beluga
+\cite{brady2013idris,DBLP:conf/icfp/ChristiansenB16} and the refinement system of Beluga
 \cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
@@
 %
-We leave further development of this interpretation of hole refinement
-to future work.
 
-\subsubsection{Fill-and-resume example}
+Indeed, the connection to CMTT neatly explains
+the difference between our proposed dynamics and existing work on staging and
+partial evaluation---that existing work is rooted in modal type theory
+\cite{Davies:2001op} rather than the derivative contextual modal type theory
+(which, to date, has been studied far less extensively.) In other words,
+existing systems are focused on partial evaluation with respect to an input that sits outside of a
+function, whereas holes can appear in context, directly within the body of
+a function.
 
-\matt{I think this example is intended to illustrate some point, but I
-  could not discern what that point was}
+There are various other systems similar in various ways to CMTT in 
+that they consider the problem of reasoning about metavariables. 
+For example, McBride's OLEG is another system for reasoning contextually about metavariables
+\cite{DBLP:phd/ethos/McBride00}, and it is the basis of certain analagous
+features of Idris \cite{brady2013idris}.
+\citet{DBLP:conf/csl/GeuversJ02} discuss similar ideas.
+Work on explicit substitutions also confronts many related technical subtleties \cite{Abadi:1991fr,levy1999explicit,Abadi:1990ys}.
+CMTT is unique relative to prior approaches in that it was designed with commutativity properties in mind (though the theory was not fully developed).
 
-As an example, consider two edits, the first filling a hole~$u$ with
-the number~$2$, and the next applying operator~$+$, resulting in $2
-+ \dehole{v}{\sigma}{}$.
+Note that the commutativity property we establish will not hold for a language that supports non-commutative effects.
+We leave to future work the task of defining more restricted special cases of commutativity in effectful settings.
+
+% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}
+
+% Furthermore, we need to integrate casts into the dynamic semantics. 
+% Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. 
+% The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms.
+% The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. 
+% Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}\todo{where?}{}, is for cast failure to instead insert a membrane around the dynamic type error, 
+% much like a non-empty hole serves as a membrane around a static type error, 
+% allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.
+
+% \begin{itemize}
+	%% \item Agda
+	%% \item Idris
+	%% \item GHC holes
+	%% \item Visual Studio (and others) support for edit-and-resume
+	% \item Scratch lets you just skip over statement holes
+	% \item papers that show up in a search for ``typed holes'' -- \url{https://scholar.google.com/scholar?hl=en&as_sdt=0%2C39&q=%22typed+holes%22&btnG=}
+% \end{itemize}
+
+% \parahead{Structure Editor Calculus}
+
+Holes play a prominent role in structure editors, and indeed the prior work on \Hazelnut was primarily motivated by this application. 
+Most work on structure editors has focused on the user interfaces that they
+present. This is important work---presenting a fluid user interface involving
+exclusively structural edit actions is a non-trivial problem that has not yet
+been fully resolved, though recent studies have started to show productivity
+gains for keyboard-driven structure
+editors~\cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14}.
+This work, together with the end-to-end well-definedness guarantee described in Sec.~\ref{sec:implementation}, suggests that structure editing together with live programming might be the foundation for a ``next-generation'' programming experience.
+We emphasize, however,
+that our proposed contributions will be relevant no matter how holes come to be
+inserted into the program. % We will say more about all of these systems later.
+
+% two main remaining questions (beyond scaling up the calculus) are how to run
+% programs with type and expression holes.
+
+% \begin{itemize}
+% 	\item Hazelnut paper
+% \end{itemize}
+
+% \parahead{Gradual Typing}
+
+% As already discussed in Track 1, type holes are closely related to the unknown
+% types of gradual type theory \cite{Siek06a}. This has implications for our
+% dynamic semantics as well---we need to be able to run programs where type holes
+% have obscured a type inconsistency, \eg{}~\lstinline{let f(g : }\ \lstinline{) => g 3 in f 4}.
+% This term is well-typed according to the static semantics of Track 1 because the
+% argument \li{g} of \li{f} has an unknown type, and so it can be applied to
+% \li{3}, but when \li{f} is applied to \li{4}, there is a problem because \li{4 3}
+% is meaningless. There are two ways to address the problem. The first is to
+% apply ML-style type inference to discover a more specific type schema for
+% \li{g}, \ie{}~that it must actually be an arrow type, and then mark the argument
+% \li{4} to \li{f} as a type inconsistency using a non-empty hole. We will
+% investigate this in Track 1. The second approach is to adopt run-time cast
+% insertions from research on gradual type theory. We will investigate this second
+% approach in Track 2.
+
+% \begin{itemize}
+% 	\item gradual typing (and dynamic typing)
+%   \item siek and taha paper
+%   \item snapl15 paper
+%   \item gradualizer paper
+%   \item maybe other things, e.g. several papers by ron garcia
+% \end{itemize}
+
+Understanding and debugging static type errors is notoriously difficult,
+particularly for novices.
 %
-This second edit is not a hole filling edit with respect to the
-immediately preceding edit state, but it can be understood as filling
-hole $u$ with $2 + \dehole{v}{\sigma}{}$.
+A variety of approaches have been
+proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
+and explain type errors.
+%
+One of these approaches~\cite{Seidel2016} proposes to use program synthesis techniques to generate a dynamic witness
+that demonstrates a run-time failure, and then display a compressed execution
+trace to the user as a graph.
+%
+\HazelnutLive{} has similar motivations in that it can run programs with type errors (\ie{}, programs with
+non-empty holes) as far as possible, in particular, until they
+would go wrong, and provide meaningful feedback. However, no attempt is made to generate examples that do not already appear in the program. 
+%
+Combining the strengths of these approaches may be fruitful in the future.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+More generally, expression holes also often appear in the context of program synthesis, serving as
+placeholders in \emph{templates}~\cite{srivastava2013template} or
+\emph{sketches}~\cite{solar2009sketching} to be filled in by an expression
+synthesis engine.
+We leave to future work the possibility of combining these approaches, i.e.
+using the information generated by running an incomplete program to inform the generation of program and edit action suggestions.
 
-\matt{This thought seems to rely on a more developed exposition of
-  CMTT that we have/will give in this paper; we should leave it in
-  this section, or remove it}
+This approach is reminiscent of the workflow that debuggers make available using breakpoints \cite{fitzgerald2008debugging,DBLP:journals/jfp/TolmachA95} and various logging and tracing capabilities. 
+The difference here is twofold.
+First, evaluation does not stop at each breakpoint and so it is straightforward to explore the \emph{space} of values that a variable takes.
+Second, breakpoints make available the values of a variable at a position in the evaluation trace.
+Hole closures, on the other hand, convey information from a syntactic position in the result of evaluation.
+The result is, by nature, a simpler object than the full evaluation trace that produced the result.
+These approaches are fundamentally complementary, rather than opposed.
 
-Another related case arises in lab notebook interfaces like that of
-Jupyter/IPython \cite{PER-GRA:2007}.
-%
-In this case, each cell can be understood as a series of \li{let}
-bindings ending in a hole, which is filled by the next cell.
-%
-The result from each cell can be cached to avoid recomputing the
-environment from preceding cells. This contextual modal interpretation
-of lab notebook cells (and read-eval-print loops as a restricted case
-where edits to previous cells are impossible) is novel.
+% \begin{itemize}
+% 	\item work on debuggers that allow you to inspect environments
+%   \item might be something relevant in the paper ``A Debugger for Standard ML'' 
+%   \item "Visualizing the evaluation of functional programs for debugging" by Whitington and Ridge
+%   \item "A lightweight interactive debugger for Haskell'' and ``Multiple-View Tracing for Haskell: a New Hat'' might be relevant
+%   \item ocamli -- \url{https://github.com/johnwhitington/ocamli}
+%   \item Better supporting debugging aids learning a novel programming language. -- Scaffidi at VLHCC 2017
+%   \item quote from Wadler in ``Why no one uses functional languages'':
+%     \begin{quote}
+%     “...there are few debuggers or
+% profilers for strict [functional] languages, perhaps because constructing them is not considered
+% research. This is a shame, since such tools are sorely needed, and there remains much of
+% interest to learn about their construction and use.
+%     \end{quote}
+%    \item edit-and-resume in visual studio, others
+% \end{itemize}
 
-\subsubsection{Unmoored notes}
+% Visual Studio edit-and-continue \url{https://docs.microsoft.com/en-us/visualstudio/debugger/edit-and-continue}
 
- \matt{Will any reader actually wonder this? Other than the definition
-   of substitution in Section 4, does this thought connect to any
-   other statement elsewhere in the paper?}
+Hole-like constructs also appear in work on program slicing
+\cite{DBLP:conf/icfp/PereraACL12}, where holes are used as a technical device to determine which parts of the program that do not impact a selected part of the result.
 
-This case shows why we cannot interpret a non-empty hole as an empty
-hole of arrow type applied to the enveloped expression---the hole
-filling operation would not operate as expected under this
-interpretation.
+% \begin{itemize}
+% 	% \item simply typed underdeterminism paper
+% 	% \item DuctileJ stuff -- \url{https://homes.cs.washington.edu/~mernst/pubs/ductile-icse2011.pdf}
+% 	% \item ``Achieving flexibility in direct-manipulation programming environments by relaxing the edit-time grammar'' -- \url{http://ieeexplore.ieee.org/document/1509511/}
+% 	\item 
+% \end{itemize}

### 2018-04-24 hazelnut-dynamics.tex
submission
address latex error (clipbox already defined); add watermarked draft for Ben
added authors and small tweaks to intro

@@
 %% For double-blind review submission
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
 %% For single-blind review submission
@@
 \usepackage{wrapfig}
-\usepackage[export]{adjustbox}% http://ctan.org/pkg/adjustbox
 
@@
 \usepackage{enumitem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Matt says: Cyrus, this package `adjustbox` seems directly related
+%% to the `clipbox` error; To get rid of the error, I moved it last
+%% (after other usepackages) and I added the line just above it, which
+%% permits it to redefine `clipbox` (apparently also defined in
+%% `pstricks`, and due to latex's complete lack of namespace
+%% management, these would otherwise names clash).
+\let\clipbox\relax
+\usepackage[export]{adjustbox}% http://ctan.org/pkg/adjustbox
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\usepackage{draftwatermark}
+%\SetWatermarkText{DRAFT}
+%\SetWatermarkScale{1}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %% Author with single affiliation.
-\author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
+\author{Cyrus Omar}
+% \authornote{with author1 note}          %% \authornote is optional;
                                         %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}
+  % \position{Position1}
+  % \department{Department1}              %% \department is recommended
+  \institution{University of Chicago}            %% \institution is required
+  % \streetaddress{Street1 Address1}
+  % \city{City1}
+  % \state{State1}
+  % \postcode{Post-Code1}
+  % \country{Country1}
 }
-\email{first1.last1@inst1.edu}          %% \email is recommended
+\email{comar@cs.uchicago.edu}          %% \email is recommended
 
-%% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
+\author{Ian Voysey}
+% \authornote{with author1 note}          %% \authornote is optional;
                                         %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}
+  % \position{Position1}
+  % \department{Department1}              %% \department is recommended
+  \institution{Carnegie Mellon University}            %% \institution is required
+  % \streetaddress{Street1 Address1}
+  % \city{City1}
+  % \state{State1}
+  % \postcode{Post-Code1}
+  % \country{Country1}
 }
-\email{first2.last2@inst2a.com}         %% \email is recommended
+\email{iev@cs.cmu.edu}          %% \email is recommended
+
+\author{Ravi Chugh}
+% \authornote{with author1 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}
+  % \position{Position1}
+  % \department{Department1}              %% \department is recommended
+  \institution{University of Chicago}            %% \institution is required
+  % \streetaddress{Street1 Address1}
+  % \city{City1}
+  % \state{State1}
+  % \postcode{Post-Code1}
+  % \country{Country1}
 }
-\email{first2.last2@inst2b.org}         %% \email is recommended
+\email{rchugh@cs.uchicago.edu}          %% \email is recommended
+
+\author{Matthew A. Hammer}
+% \authornote{with author1 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  % \position{Position1}
+  % \department{Department1}              %% \department is recommended
+  \institution{University of Colorado Boulder}            %% \institution is required
+  % \streetaddress{Street1 Address1}
+  % \city{City1}
+  % \state{State1}
+  % \postcode{Post-Code1}
+  % \country{Country1}
+}
+\email{matthew.hammer@colorado.edu}          %% \email is recommended
+
+% %% Author with two affiliations and emails.
+% \author{First2 Last2}
+% \authornote{with author2 note}          %% \authornote is optional;
+%                                         %% can be repeated if necessary
+% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+% \affiliation{
+%   \position{Position2a}
+%   \department{Department2a}             %% \department is recommended
+%   \institution{Institution2a}           %% \institution is required
+%   \streetaddress{Street2a Address2a}
+%   \city{City2a}
+%   \state{State2a}
+%   \postcode{Post-Code2a}
+%   \country{Country2a}
+% }
+% \email{first2.last2@inst2a.com}         %% \email is recommended
+% \affiliation{
+%   \position{Position2b}
+%   \department{Department2b}             %% \department is recommended
+%   \institution{Institution2b}           %% \institution is required
+%   \streetaddress{Street3b Address2b}
+%   \city{City2b}
+%   \state{State2b}
+%   \postcode{Post-Code2b}
+%   \country{Country2b}
+% }
+% \email{first2.last2@inst2b.org}         %% \email is recommended
 
@@
 \input{extensions}
-\input{misc}
+% \input{misc}
+\input{implementation-appendix}
 

### 2018-04-24 intro.tex
added authors and small tweaks to intro

@@
 % !TEX root = hazelnut-dynamics.tex
-\vspace{-7px}
+\vspace{-5px}
 \newcommand{\introSec}{Introduction}
@@
 \label{sec:intro}
-% \vspace{-2px}
+\vspace{-2px}
 
@@
 and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding in an artistic performance.
-Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support editing the state of a running program, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
-Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13},
+Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support editing run-time state, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
+% Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13},
 %has been and continues to be an active area of research and development.
-has been, and continues to be, an active area of research and development.
+% has been, and continues to be, an active area of research and development.
 
@@
 A program editor, however, frequently encounters incomplete, and therefore meaningless, editor states.
-As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e.
-it relies on the most recent complete editor state, creating a perceptive gap because the feedback may not accurately reflect what the programmer is actually seeing and writing in the editor.
+As a result, live feedback either ``flickers out'', creating a temporal gap, or it ``goes stale'', i.e.
+it relies on the most recent complete editor state, creating a perceptive gap because the feedback may not accurately reflect what the programmer is seeing in the editor.
 
@@
 Throughout the process of addressing these errors, the program text remains formally meaningless.
-Overall, an analysis of edits performed by Java programmers using Eclipse found that about 40\% of edit states were malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.
+Overall, about 40\% of edits performed by Java programmers using Eclipse left the program text malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.
 
@@
 e.g. Scala provides \lismall{???}  
-and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u}
+and GHC Haskell with the \lismall{-fdefer-typed-holes} flag 
+%or \lismall{-fdefer-ty}\lismall{pe-errors} flags 
+uses \lismall{_u}
 where \lismall{u} serves as an
-optional hole name \cite{GHCHoles}.%
-\footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
-Proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} also respond to holes in this way.}
+optional hole name \cite{GHCHoles}. %
+% \footnote{
+Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
+Proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} also respond to holes in this way.
+% }
 

### 2018-04-30 calculus.tex
substantially expanded the related work

@@
 evaluation does not stop.
-For simplicity, we do not include blame labels as found in some accounts of gradual typing \cite{DBLP:conf/snapl/SiekVCB15}, but it would be straightforward to do so by recording the blame labels from the two constituent casts on the two arrows of the failed cast form.
+For simplicity, we do not include blame labels as found in some accounts of gradual typing \cite{DBLP:conf/esop/WadlerF09,DBLP:conf/snapl/SiekVCB15}, but it would be straightforward to do so by recording the blame labels from the two constituent casts on the two arrows of the failed cast.
 
@@
 including all of the theorems stated above and necessary lemmas.
-We take as assumptions only a few standard properties of general hypothetical judgements and finite maps, and, to avoid excessive machinery related to hole name tracking in the bidirectional typing rules, we axiomatize some hole name uniqueness assumptions made implicitly in this paper.
+We take as assumptions only a few standard properties of general hypothetical judgements and finite maps.
 %

### 2018-04-30 discussion.tex
near-final intro
submission
substantially expanded the related work

--- 2018-02-07 discussion.tex
@@
 % !TEX root = hazelnut-dynamics.tex
-\newcommand{\discussionSection}{Discussion}
+\newcommand{\discussionSection}{Conclusion}
 \section{\protect\discussionSection} % don't like the all-caps thing that the template does, so protecting it from that
 \label{sec:discussion}
+To conclude, we quote Weinberg from The Psychology of Computer Programming (1998): ``how truly sad it is that just at the very moment
+when the computer has something important to tell us, it starts
+speaking gibberish.''
 
-\todo{discussion}
+\newcommand{\acksSection}{Acknowledgments}
+\section*{\protect\acksSection} 
+\label{sec:acks}
+
+The authors would like to thank the participants of the PL reading group at UChicago, Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton and Claire LeGoues for their insights and feedback on various iterations of this work.
+
+% \todo{discussion}
+
+% \rkc{something about UI choices for displaying large indeterminate expressions}
+
+% \rkc{something about UI choices for displaying hole environment traces}
+
+% \rkc{something about patterns of use for debugging, such as inserting holes
+% around all expressions in tail position}
+
+% \rkc{something about visualizing hole environment traces}
+
+% \rkc{\citet{Seidel2016} define a partial evaluation with a form of holes. compare.
+% compare ``flattened'' sumList indeterminate result to their trace. could be
+% useful to study whether evaluating the type-correct parts of the evaluation
+% trace help students focus on the part that has the error.}
+
+% \rkc{quick mention that lazy evaluation with run-time exceptions for holes is
+% not enough. this would delay the failure a bit longer, but not indefinitely.}
+
+% live mobile application development systems like Flutter \cite{flutter};

### 2018-04-30 extensions.tex
substantially expanded the related work

@@
 The ``non-obvious'' machinery is almost entirely related to casts.
-Fortunately, there has been excellent work of late on generating ``gradualized'' specifications from standard specifications \cite{DBLP:conf/popl/CiminiS16}.
+Fortunately, there has been excellent work of late on generating ``gradualized'' specifications from standard specifications \cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/CiminiS17}.
 The extensions below, and other extensions of interest, closely follow the output of the gradualizer: \url{http://cimini.info/gradualizerDynamicSemantics/} (which, like our work, is based on the refined account of gradual typing by \cite{DBLP:conf/snapl/SiekVCB15}).
-The rules for indeterminate forms are mainly where the gradualizer is not necessarily sufficient.
+The rules for indeterminate forms are mainly where the gradualizer is not sufficient.
+We leave to future work the related question of automatically generating a hole-aware static and dynamic semantics from a standard language specification.
 

### 2018-04-30 related-work.tex
substantially expanded the related work

@@
 % !TEX root = hazelnut-dynamics.tex
+\vspace{-5px}
 \newcommand{\relatedWorkSection}{Related and Future Work}
@@
 \label{sec:relatedWork}
+\vspace{-2px}
 
 This paper defined a dynamic semantics for incomplete functional programs
-based directly on the static semantics developed by \citet{popl-paper}.
-A subsequent ``vision paper'' introducing \Hazel suggested the need for 
+based directly on the static semantics developed by \citet{popl-paper}. That paper, as well as a subsequent ``vision paper'' introducing \Hazel, suggested as future work  
 a corresponding dynamic semantics for the purposes of live programming \cite{HazelnutSNAPL}. 
 This paper    
-contributes the type-theoretic first steps necessary to achieve this long-term vision.
+develops the conceptual and type-theoretic details necessary to achieve this broader vision.
 
-The semantics borrows substantially
-from the theory of type consistency and casts developed in gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}, as discussed
+The semantics borrows machinery related to type holes from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}, as discussed
 at length in Sec.~\ref{sec:calculus}. The main innovation relative to this
 prior work is the treatment of cast failures as holes, rather than errors.
-Many of the methods developed to make gradual typing more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein.
+Many of the methods developed to make gradual typing more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein \cite{takikawa_et_al:LIPIcs:2015:5215}.
 For example, there has been substantial work on the problem of implementing casts efficiently, e.g.
-by \citet{herman2010space}, and on integrating gradual typing with polymorphism, e.g.
-by \cite{garcia2015principal}.
-Another interesting direction that we leave to future work is the integration of these ideas into dependently typed proof assistants,
-where evaluation is at the service of equational reasoning.
+by \citet{herman2010space}, and on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.
+Another direction that we leave to future work is the integration of these ideas into dependently typed proof assistants,
+where evaluation is at the service of equational reasoning \cite{DBLP:journals/corr/abs-1009-2789}.
+Several modern proof assistants, e.g.
+Agda \cite{norell2009dependently,norell:thesis}, Idris \cite{brady2013idris} and Beluga \cite{DBLP:conf/flops/Pientka10} already provide initial syntactic and static support for holes (see Sec.~\ref{sec:intro}).
 
-Another important future direction is to move beyond
-pure functional programming a la \Elm~ and carefully integrate imperative features, e.g.
+Although our expectation is that \Hazel will remain a pure functional language \emph{a la} \Elm~or Haskell, another interesting future direction would be to move beyond
+pure functional programming and carefully integrate imperative features, e.g.
 ML-style references.
-These require particular care, as recognized by \cite{Siek06a}, though we see no reason why the type safety properties  established in this paper would not be conserved in an effectful setting.
-Going beyond references to incorporate general effects, e.g.
-network and IO effects, raises some important practical concerns, however---we do not want to continue past a hole or error and in so doing haphazardly trigger an unintended effect.
-Developing language features and user interface features that give the programmer fine-grained control over evaluation when it encounters a hole or cast failure are likely to be particularly helpful in these settings.
+These require particular care, as recognized by \citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15}, but the techniques are known and we see no reason why the type safety properties  established in this paper would not be conserved, with suitable modifications to account for a store.
+Going beyond references to incorporate external effects, e.g.
+IO effects, raises some important practical concerns, however---we do not want to continue past a hole or error and in so doing haphazardly trigger an unintended effect.
+In this setting, it is likely better to explicitly ask the programmer whether to continue evaluation past a hole or cast failure.
+More generally, the small step specification in this paper is suitable as the basis for a single stepper, which generalizes evaluation.
+\citet{ocaml-stepper} discuss some unresolved usability issues relevant to single steppers for functional languages.
+Note also that the commutativity property we establish in Sec.~\ref{sec:resumption} will not hold for a language that supports non-commutative effects.
+We leave to future work the task of defining more restricted special cases of commutativity in effectful settings.
 
-The other major pillar of related work is the work on contextual modal type theory (CMTT) \cite{Nanevski2008}, which we also discussed at length throughout the paper.
-To reiterate, there is a close relationship between holes in this paper and metavariables in CMTT.
+The other major pillar of related work is the work on contextual modal type theory (CMTT) by \citet{Nanevski2008}, which we also discussed at length throughout the paper.
+To reiterate, there is a close relationship between expression holes in this paper and metavariables in CMTT.
 Hole contexts correspond to modal contexts.
 Empty hole closures 
-relate to the concept of a \emph{metavariable closure} in CMTT, which consists
+relate to the concept of a {metavariable closure} in CMTT, which consists
 of a metavariable paired with a substitution for all of the variables in the
@@
 
-Although these connections are encouraging, our contributions do not neatly fall out from the prior work.
+Although these connections with gradual typing and CMTT are encouraging, our contributions do not neatly fall out from this prior work.
 The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.
@@
 This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. 
-In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables.
-\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion%
-\footnote{Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property.
-These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.},
-which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
-but under this interpretation an analogous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.
+In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables (and, of course, with casts, i.e.
+\emph{gradual contextual modal type theory}).
+
+\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion, %, 
+which has been studied by \citet{DBLP:journals/iandc/Groote02}. 
+Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property. 
+These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.
+In any case, under this interpretation an analogous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.
+
+While there has been some formal work on reduction of open terms, leading to programs in a \emph{weak head normal form} \cite{barendregt84:_lambda_calculus,DBLP:journals/corr/abs-1009-2789}, there is no progress theorem in this setting nor is there environment tracking.
+Moreover, in the setting of an evaluator for a programming language, we do not want to evaluate under arbitrary binders but rather only around hole closures \cite{DBLP:conf/birthday/BlancLM05}.
+CMTT makes the appropriate distinctions between variables and metavariables, and our notion of an indeterminate form allows us to prove a progress theorem of familiar form.
+Weak head normal forms are more useful when using evaluation to perform optimizations throughout a program, e.g.
+when using supercompilation-by-evaluation \cite{DBLP:conf/haskell/BolingbrokeJ10} or symbolic evaluation \cite{King:1976}, or when using evaluation in the service of equational reasoning as in proof assistants (e.g.
+Beluga \cite{DBLP:journals/corr/abs-1009-2789}).
 
 It is also worth emphasizing that we use the machinery 
-borrowed from CMTT extralinguistically.
+borrowed from CMTT only extralinguistically.
 %
-However, a key feature CMTT that we have not yet touched on is the
+A key feature of CMTT that we have not yet touched on is the
 \emph{internalization} of metavariable binding and contextual
@@
 %
-A hole filling can be interpreted as an expression of contextual modal
+A hole filling can be interpreted as a value of contextual modal
 type, and the act of hole filling followed by evaluation to the next
@@
 %
-This interpretation allows us to \emph{compute} hole fillings, rather
-than simply stating them, by specifying non-trivial expressions of
+This interpretation could be generalized to allow us to \emph{compute} a hole filling, rather
+than simply stating it, by specifying a non-trivial expression, rather than just a value, of
 modal type.
 %
-This could serve as the basis for a \emph{live} computational hole
+This could, in turn, serve as the basis for a \emph{live} computational hole
 refinement system, extending the capabilities of purely static hole
@@
 %
+This parallel neatly explains
+the difference between our dynamics and existing work on staging and
+partial evaluation---
+existing staging and partial evaluation systems are focused on partial evaluation with respect to an input that sits outside of a
+function \cite{Jones:1993uq}, whereas holes have non-trivial contexts.
 
-Indeed, the connection to CMTT neatly explains
-the difference between our proposed dynamics and existing work on staging and
-partial evaluation---that existing work is rooted in modal type theory
-\cite{Davies:2001op} rather than the derivative contextual modal type theory
-(which, to date, has been studied far less extensively.) In other words,
-existing systems are focused on partial evaluation with respect to an input that sits outside of a
-function, whereas holes can appear in context, directly within the body of
-a function.
-
-There are various other systems similar in various ways to CMTT in 
+There are various other systems similar in certain ways to CMTT in 
 that they consider the problem of reasoning about metavariables. 
 For example, McBride's OLEG is another system for reasoning contextually about metavariables
-\cite{DBLP:phd/ethos/McBride00}, and it is the basis of certain analagous
-features of Idris \cite{brady2013idris}.
+\cite{DBLP:phd/ethos/McBride00}, and it is the conceptual basis of certain hole refinement features in Idris \cite{brady2013idris}.
 \citet{DBLP:conf/csl/GeuversJ02} discuss similar ideas.
-Work on explicit substitutions also confronts many related technical subtleties \cite{Abadi:1991fr,levy1999explicit,Abadi:1990ys}.
-CMTT is unique relative to prior approaches in that it was designed with commutativity properties in mind (though the theory was not fully developed).
+CMTT is unique relative to prior approaches in that it has a clear Curry-Howard interpretation and that it was designed with commutativity properties in mind.
+This paper bears out that it is nicely suited for the purpose of live functional programming with typed expression holes.
 
-Note that the commutativity property we establish will not hold for a language that supports non-commutative effects.
-We leave to future work the task of defining more restricted special cases of commutativity in effectful settings.
+Work on explicit substitutions confronts problems similar to the problem of tracking the environment around a closure using delayed substitutions \cite{Abadi:1991fr,levy1999explicit,Abadi:1990ys}.
+\citet{DBLP:journals/corr/abs-1009-2789} have developed a theory of explicit substitutions for CMTT, which, following other work on explicit substitutions and environmental abstract machines \cite{DBLP:journals/tcs/Curien91}, would be useful in implementing \HazelnutLive more efficiently by delaying both standard and contextual substitutions until needed during evaluation.
+We leave this and other questions of fast implementation (e.g.
+using thunks to encapsulate indeterminate sub-expressions) to future work.
 
@@
 
-Holes play a prominent role in structure editors, and indeed the prior work on \Hazelnut was primarily motivated by this application. 
+Our approach is reminiscent of the workflow that debuggers make available using breakpoints \cite{fitzgerald2008debugging,DBLP:journals/jfp/TolmachA95} and a variety of logging and tracing capabilities.
+Debuggers do not directly support incomplete programs, so the programmer first needs to insert suitable dummy exceptions as discussed in Sec.~\ref{sec:intro}.
+Beyond that, there are two main distinctions.
+First, evaluation does not stop at each breakpoint and so it is straightforward to explore the \emph{space} of values that a variable takes.
+Second, breakpoints, logs and tracing tools convey the values of a variable at a position in the evaluation trace.
+Hole closures, on the other hand, convey information from a syntactic position in the \emph{result} of evaluation.
+The result is, by nature, a simpler object.%  These approaches are fundamentally complementary in that standard debugging facilities are useful even when there are no holes in the program whereas our focus has been on incomplete programs.
+Some debuggers support ``edit-and-continue'', e.g.
+Visual Studio \cite{VSEditAndContinue} and Flutter \cite{flutter}, based on the dynamic software update (DSU) capabilities of the underlying run-time system \cite{DBLP:journals/toplas/StoyleHBSN07,DBLP:conf/vstte/HaydenMHFF12}.
+These features do not come with any guarantee that rerunning the program is  guaranteed to produce the same result.
+The ``fill-and-resume'' feature described in Sec.~\ref{sec:resumption} is a special case that comes with a stronger formal guarantee.
+
+Holes play a prominent role in structure editors, and indeed the prior work on \Hazelnut was primarily motivated by this application \cite{popl-paper}. 
 Most work on structure editors has focused on the user interfaces that they
@@
 been fully resolved, though recent studies have started to show productivity
-gains for keyboard-driven structure
-editors~\cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14}.
-This work, together with the end-to-end well-definedness guarantee described in Sec.~\ref{sec:implementation}, suggests that structure editing together with live programming might be the foundation for a ``next-generation'' programming experience.
+gains for blocks-based structure editors like Scratch for novice programmers \cite{Resnick:2009:SP:1592761.1592779,DBLP:conf/chi/WeintropASFLSF18,DBLP:conf/acmidc/WeintropW15}, and for keyboard-driven structure
+editors like \li{mbeddr} in professional programming settings~\cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14,voelter_mbeddr:_2012}. 
+Some structure editors support live programming in various ways, e.g.
+Lamdu is a structure editor for a typed functional language that displays variable values inline based on recorded execution traces (see above) \cite{lamdu}.
+However, they do not generally support execution of incomplete programs, which was our focus in this paper.
+Scratch will execute a program with holes by simply skipping over incomplete commands, but this is a limited protocol because rerunning the program after filling the hole may produce a result unrelated to initial result in any way.
+Though in some situations, skipping over problematic commands has been observed to work surprisingly well \cite{DBLP:conf/dac/Rinard12}, this paper contributes a semantically sound approach (Sec.~\ref{sec:resumption}).
+Combined with the aforementioned UI advances, there is reason to believe that structure editing could be the foundation for a ``next-generation'' live programming experience.
 We emphasize, however,
-that our proposed contributions will be relevant no matter how holes come to be
-inserted into the program. % We will say more about all of these systems later.
+that our proposed contributions will be relevant no matter how holes come to appear (see Sec.~\ref{sec:intro}). % We will say more about all of these systems later.
 
@@
 
+A key feature of our semantics is that it permits evaluation not only of terms with empty holes, but also terms with non-empty holes, i.e. reified static type inconsistencies. 
 Understanding and debugging static type errors is notoriously difficult,
@@
 A variety of approaches have been
-proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
-and explain type errors.
+proposed to better localize
+and explain type errors \cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc}.
 %
-One of these approaches~\cite{Seidel2016} proposes to use program synthesis techniques to generate a dynamic witness
-that demonstrates a run-time failure, and then display a compressed execution
+One of these approaches, by \citet{Seidel2016},  uses symbolic execution and program synthesis to generate a dynamic witness
+that demonstrates a run-time failure that could be caused by the static type error, and then displays a compressed evaluation
 trace to the user as a graph.
 %
-\HazelnutLive{} has similar motivations in that it can run programs with type errors (\ie{}, programs with
-non-empty holes) as far as possible, in particular, until they
-would go wrong, and provide meaningful feedback. However, no attempt is made to generate examples that do not already appear in the program. 
+\HazelnutLive{} has similar motivations in that it can run programs with static or dynamic type errors  and provide meaningful feedback about where ill-typed terms end up in the program, even after the first problem is encountered (Sec.~\ref{sec:static-errors}-\ref{sec:dynamic-type-errors}).
+However, no attempt is made to synthesize examples that do not already appear in the program.
 %
@@
 
-More generally, expression holes also often appear in the context of program synthesis, serving as
+Expression holes also often appear in the context of program synthesis, serving as
 placeholders in \emph{templates}~\cite{srivastava2013template} or
@@
 We leave to future work the possibility of combining these approaches, i.e.
-using the information generated by running an incomplete program to inform the generation of program and edit action suggestions.
-
-This approach is reminiscent of the workflow that debuggers make available using breakpoints \cite{fitzgerald2008debugging,DBLP:journals/jfp/TolmachA95} and various logging and tracing capabilities. 
-The difference here is twofold.
-First, evaluation does not stop at each breakpoint and so it is straightforward to explore the \emph{space} of values that a variable takes.
-Second, breakpoints make available the values of a variable at a position in the evaluation trace.
-Hole closures, on the other hand, convey information from a syntactic position in the result of evaluation.
-The result is, by nature, a simpler object than the full evaluation trace that produced the result.
-These approaches are fundamentally complementary, rather than opposed.
+using the information generated by running an incomplete program as described in this paper to inform the generation of program and edit action suggestions, e.g.
+via tests or assertions that serve to generate constraints dynamically for type-and-example-driven synthesis \cite{DBLP:conf/popl/FrankleOWZ16}.
 
@@
 Hole-like constructs also appear in work on program slicing
-\cite{DBLP:conf/icfp/PereraACL12}, where holes are used as a technical device to determine which parts of the program that do not impact a selected part of the result.
+\cite{DBLP:conf/icfp/PereraACL12,DBLP:journals/pacmpl/RicciottiSPC17}, where empty expressions holes arise as a technical device to determine which parts of a complete program do not impact a selected part of the result of evaluating that program.
+In other words, the focus is on explaining a result that is presumed to exist, whereas the work in this paper is focused on producing results where they would not have otherwise been available.
+Combining the strengths of these approaches is another fruitful avenue for future research.
 

### 2018-04-30 hazelnut-dynamics.tex
suppress ACM information

@@
 %% For double-blind review submission
-\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\documentclass[acmsmall,review]{acmart}
+\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\setcopyright{none}
+\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
+
 %% For single-blind review submission
@@
 
-%% Listings
 \makeatletter\if@ACM@journal\makeatother
@@
 %% Supplied to authors by publisher for camera-ready submission
-\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
+% \acmConference[]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
+
 \acmYear{2018}
@@
 %\copyrightyear{2017}           %% If different from \acmYear
+
+\fancyfoot{} % suppresses the footer (also need \thispagestyle{empty} after \maketitle below)
 
@@
 \maketitle
+\thispagestyle{empty} % suppresses the footer
 

### 2018-04-30 related-work.tex
suppress ACM information

@@
 Some debuggers support ``edit-and-continue'', e.g.
-Visual Studio \cite{VSEditAndContinue} and Flutter \cite{flutter}, based on the dynamic software update (DSU) capabilities of the underlying run-time system \cite{DBLP:journals/toplas/StoyleHBSN07,DBLP:conf/vstte/HaydenMHFF12}.
-These features do not come with any guarantee that rerunning the program is  guaranteed to produce the same result.
-The ``fill-and-resume'' feature described in Sec.~\ref{sec:resumption} is a special case that comes with a stronger formal guarantee.
+Visual Studio \cite{VSEditAndContinue} and Flutter \cite{flutter}, based on the dynamic software update (DSU) capabilities of the underlying run-time system \cite{DBLP:journals/toplas/StoyleHBSN07,DBLP:conf/vstte/HaydenMHFF12,DBLP:journals/toplas/HicksN05}.
+These features do not come with any guarantee that rerunning the program will produce the same result.
+The ``fill-and-resume'' feature described in Sec.~\ref{sec:resumption} is a special case that comes with this stronger guarantee.
 

### 2018-04-30 discussion.tex
added acknowledgment of LIVE and OBT referees

@@
 
-The authors would like to thank the participants of the PL reading group at UChicago, Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton and Claire LeGoues for their insights and feedback on various iterations of this work.
+The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton and Claire LeGoues, the participants of the PL reading group at UChicago, and referees for talk proposals at the LIVE 2017 and Off the Beaten Track (OBT) 2017 workshops for their insights and feedback on various iterations of this work.
 

### 2018-04-30 discussion.tex
better wording in acks

--- 2018-04-30 discussion.tex
@@
 
-The authors would like to thank the participants of the PL reading group at UChicago, Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton and Claire LeGoues for their insights and feedback on various iterations of this work.
+The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, the participants of the PL reading group at UChicago, and referees for talk proposals as well as the audience members at the LIVE 2017 and Off the Beaten Track (OBT) 2017 workshops for their insights and feedback on various iterations of this work.
 

### 2018-04-30 discussion.tex
better wording in acks

--- 2018-04-30 discussion.tex
@@
 
-The authors would like to thank the participants of the PL reading group at UChicago, Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton and Claire LeGoues for their insights and feedback on various iterations of this work.
+The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, the participants of the PL reading group at UChicago, and referees and participants of the LIVE 2017 and Off the Beaten Track (OBT) 2017 workshops for their insights and feedback on various iterations of this work.
 

### 2018-04-30 discussion.tex
better wording in acks

@@
 
-The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, the participants of the PL reading group at UChicago, and referees and participants of the LIVE 2017 and Off the Beaten Track (OBT) 2017 workshops for their insights and feedback on various iterations of this work.
+The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, the participants of the PL reading group at UChicago, and the referees and participants of the LIVE 2017 and Off the Beaten Track (OBT) 2017 workshops for their insights and feedback on various iterations of this work.
 

### 2018-04-30 discussion.tex
add to acks

@@
 
-The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, the participants of the PL reading group at UChicago, and the referees and participants of the LIVE 2017 and Off the Beaten Track (OBT) 2017 workshops for their insights and feedback on various iterations of this work.
+The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues,
+the participants of the PL reading group at UChicago,
+and the referees and participants of the LIVE 2017 and Off the Beaten Track (OBT)
+2017 workshops for their insights and feedback on various iterations of this work.
+%
+%%%%%%% BEGIN: Matt's Acks
+%
+This material is based in part upon work supported by a gift from
+Facebook, and support from the National Science Foundation under grant
+number CCF-1619282.
+%
+Any opinions, findings, and conclusions or recommendations expressed
+in this material are those of the author(s) and do not necessarily
+reflect the views of Mozilla, Facebook or the National Science Foundation.
+%
+%%%%%%% END: Matt's Acks
 

### 2018-04-30 related-work.tex
add paragraph citing Python tutor, and Andy Ko's work

@@
 Our approach is reminiscent of the workflow that debuggers make available using breakpoints \cite{fitzgerald2008debugging,DBLP:journals/jfp/TolmachA95} and a variety of logging and tracing capabilities.
+%
 Debuggers do not directly support incomplete programs, so the programmer first needs to insert suitable dummy exceptions as discussed in Sec.~\ref{sec:intro}.
+%
 Beyond that, there are two main distinctions.
@@
 editors like \li{mbeddr} in professional programming settings~\cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14,voelter_mbeddr:_2012}.
+%
 Some structure editors support live programming in various ways, e.g.
@@
 that our proposed contributions will be relevant no matter how holes come to appear (see Sec.~\ref{sec:intro}). % We will say more about all of these systems later.
+
+In addition to offering semantic foundations for live programming, we
+believe that \HazelnutLive also holds promise as a formal basis for
+educational programming environments that visualize program execution
+for novice programmers (e.g., students).
+%
+For instance, \citet{Nelson2017} and \citet{Guo13} each give
+approaches for visualizing the internal machine states of imperative
+programs, including mutable program variables, activation records
+(stack frames), pointers, and heap records.
+%
+By contrast, \HazelnutLive offers an alternative basis for visualizing
+the internal machine states of typed, functional programs.
+%
+As outlined above, we believe that this basis can be extended to
+richer programming models, including those with imperative state.
 

### 2018-04-30 abstract.tex
added visualizer stuff from Matt to first sentence as described in Slack

@@
 %
-Incomplete functional programs are modeled as expressions with \emph{holes}, 
+We model incomplete functional programs as expressions with \emph{holes}, 
 with empty holes standing for missing expressions or types, and  non-empty holes 
@@
 performing as much of the remaining computation as is possible and tracking the 
-closure around each hole instance as it flows through the program.  These hole closures
- can be reported both directly and indirectly, via various editor services, to help the programmer decide how to fill a hole. 
+closure around each hole instance as it flows through the program. Various editor services can report information from these hole closures 
+to help the programmer decide how to fill a hole. 
 They also enable a ``fill-and-resume'' feature that 

### 2018-04-30 related-work.tex
added visualizer stuff from Matt to first sentence as described in Slack

--- 2018-04-30 related-work.tex
@@
 
-Our approach is reminiscent of the workflow that debuggers make available using breakpoints \cite{fitzgerald2008debugging,DBLP:journals/jfp/TolmachA95} and a variety of logging and tracing capabilities.
+Our approach is reminiscent of the workflow that debuggers make available using breakpoints \cite{fitzgerald2008debugging,DBLP:journals/jfp/TolmachA95}, visualizers of program state \cite{Nelson2017,Guo13}, and a variety of logging and tracing capabilities.
+%
 Debuggers do not directly support incomplete programs, so the programmer first needs to insert suitable dummy exceptions as discussed in Sec.~\ref{sec:intro}.
+%
 Beyond that, there are two main distinctions.
@@
 Hole closures, on the other hand, convey information from a syntactic position in the \emph{result} of evaluation.
-The result is, by nature, a simpler object.%  These approaches are fundamentally complementary in that standard debugging facilities are useful even when there are no holes in the program whereas our focus has been on incomplete programs.
+The result is, by nature, a simpler object.
+%  These approaches are fundamentally complementary in that standard debugging facilities are useful even when there are no holes in the program whereas our focus has been on incomplete programs.
 Some debuggers support ``edit-and-continue'', e.g.

### 2018-05-01 discussion.tex
jonathan's funding info

--- 2018-04-30 discussion.tex
@@
 
-The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, the participants of the PL reading group at UChicago, and the referees and participants of the LIVE 2017 and Off the Beaten Track (OBT) 2017 workshops for their insights and feedback on various iterations of this work.
+The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues,
+the participants of the PL reading group at UChicago,
+and the referees and participants of the LIVE 2017 and Off the Beaten Track (OBT)
+2017 workshops for their insights and feedback on various iterations of this work.
+%
+This material was supported by a gift from
+Facebook, from the National Science Foundation under grant
+number CCF-1619282, and from AFRL and DARPA under agreement \#FA8750-16-2-0042. 
+%
+Any opinions, findings, and conclusions or recommendations expressed
+in this material are those of the authors and do not necessarily
+reflect the views of Facebook, NSF, DARPA or AFRL.
+%
 

### 2018-05-02 calculus.tex
added additional conclusions to complete expansion and fixed statement in appendix based on discussion earlier today with ian

@@
       and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
-      then $\isComplete{\htau}$ and $\isComplete{\dexp}$.
+      then $\isComplete{\htau}$ and $\isComplete{\dexp}$ and $\Delta = \emptyset$.
     \item
@@
       and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
-      then $\isComplete{\dexp}$.
+      then $\isComplete{\dexp}$ and $\isComplete{\htau'}$ and $\Delta=\emptyset$
   \end{enumerate}
@@
 including all of the theorems stated above and necessary lemmas.
-We take as assumptions only a few standard properties of general hypothetical judgements and finite maps.
+We take as assumptions only a few standard properties of general hypothetical judgements, substitution and finite maps.
 %
@@
 To support this encoding choice, we postulate function extensionality (which is independent of Agda's axioms) \cite{awodey2012inductive}. 
-The documentation provided with the mechanization has more technical details.
+The documentation provided with the mechanization has more details.
 

### 2018-05-02 fig-complete.tex
fixing completeness rule names to match agda
submission
added additional conclusions to complete expansion and fixed statement in appendix based on discussion earlier today with ian

@@
 \begin{mathpar}
-\inferrule[BComplete]{ }{
+\inferrule[TCBase]{ }{
   \isComplete{b}
@@
 
-\inferrule[ArrComplete]{
+\inferrule[TCArr]{
   \isComplete{\htau_1}\\
@@
 \begin{mathpar}
-\inferrule[EVarComplete]{ }{
+\inferrule[ECVar]{ }{
   \isComplete{x}
@@
 
-\inferrule[EConstComplete]{ }{
+\inferrule[ECConst]{ }{
   \isComplete{c}
@@
 
-\inferrule[ESynLamComplete]{
+\inferrule[ECLam1]{
   \isComplete{\htau}\\
@@
 
-\inferrule[EAnaLamComplete]{
+\inferrule[ECLam2]{
   \isComplete{\hexp}
@@
 
-\inferrule[EApComplete]{
+\inferrule[ECAp]{
   \isComplete{\hexp_1}\\
@@
 
-\inferrule[EAscComplete]{
+\inferrule[ECAsc]{
   \isComplete{\hexp}\\
@@
 \begin{mathpar}
-\inferrule[DVarComplete]{ }{
+\inferrule[DCVar]{ }{
   \isComplete{x}
@@
 
-\inferrule[DConstComplete]{ }{
+\inferrule[DCConst]{ }{
   \isComplete{c}
@@
 
-\inferrule[DLamComplete]{
+\inferrule[DCLam]{
   \isComplete{\htau}\\
@@
 
-\inferrule[DApComplete]{
+\inferrule[DCAp]{
   \isComplete{\dexp_1}\\
@@
 
-\inferrule[DCastComplete]{
+\inferrule[DCCast]{
   \isComplete{\dexp}\\
@@
 
+We define $\isComplete{\hGamma}$ as follows.
+\begin{defn}[Typing Context Completeness]
+$\isComplete{\hGamma}$ iff for each $x : \htau \in \hGamma$, we have $\isComplete{\htau}$.
+\end{defn}
+
 When two types are complete and consistent, they are equal.
 
-\begin{lem}[Complete Consistency] If $\tconsistent{\htau_1}{\htau_2}$ an $\isComplete{\htau_1}$ and $\isComplete{\htau_2}$ then $\htau_1 = \htau_2$. 
+\begin{lem}[Complete Consistency] If $\tconsistent{\htau_1}{\htau_2}$ and $\isComplete{\htau_1}$ and $\isComplete{\htau_2}$ then $\htau_1 = \htau_2$.
 \end{lem}
@@
 
-This implies that in a well-typed internal expression, every cast is an identity cast.
+This implies that in a well-typed and complete internal expression, every cast is 
+an identity cast.

### 2018-06-06 calculus.tex

@@
 %
-Rule \rulename{ITBeta} in Fig.~\ref{fig:instruction-transitions} defines the
+Rule \rulename{ITLam} in Fig.~\ref{fig:instruction-transitions} defines the
 standard beta reduction transition.
@@
 
-The ITBeta rule is not the only rule we need to handle function
+The \rulename{ITLam} rule is not the only rule we need to handle function
 application, because lambdas are not the only final form of arrow type.

### 2018-06-07 rebuttal.txt
starting to work on the rebuttal

Thank you for your thoughtful reviews. We believe...

# Review A

> The HazelnutLive calculus in the paper is really basic, it doesn't even have 
> fixpoint.

Type holes actually allow you to express general recursion (e.g. via the Y 
combinator). See Siek and Taha [2006] for more on embedding the untyped lambda
calculus into a gradually typed lambda calculus like Hazelnut Live.

That said, we agree that an explicit fix operator in the style of Plotkin's PCF
might be more clear and there is nothing that would complicate its inclusion, 
so if requested, we would be happy to include it in the second phase submission, 
either in the main calculus or as another extension.

> I'm still confused about which version of the language is implemented.

The language of the submitted implementation is the language of Section 3 + 
the extensions in the appendix. We plan to archive a frozen version of that 
implementation as a simple reference implementation of the core calculus of 
this paper.

The first two paragraphs of Section 2 were, we believe, direct in stating that 
the figures in Section 2 were post-processed to use certain language features 
that were not implemented at the time of submission. None of these language 
features are particularly complex or unusual (records, simple pattern matching, 
and syntactic sugar for function definitions and lists), and they are 
orthogonal to the live programming features that are the topic of the paper, 
which are all implemented as described in Section 2 (see the implementation 
screenshots in Fig. 17-18 in the appendix).

We plan to submit a more recent snapshot of the Hazel implementation for 
artifact evaluation that either substantially narrows or eliminates this 
``convenience gap''.

Our approach has been to...

# Review B

(nothing to say, maybe just acknowledge the typos? remove this section entirely?)

# Review C

> the calculus lacks recursion and control-flow. Dealing with these efficiently 
> and providing a good user interface for inspecting the resulting reified values 
> is the hard problem in making such an environment.

## Recursion

See the response to Review A above regarding recursion in the core calculus.

We also considered the topic of recursion in Sec. 2.2 (Example 2: Recursive 
Functions). The UI was designed with recursion in mind in that it allows the 
user to explore the recursive structure of hole closures iteratively as 
described in that section, rather than displaying nested hole closures all at
once, as is the case in the notation used for the core calculus.

The UI displays a "breadcrumb" path from the result to reflect the exploration 
that the user has done (the basic UI can be seen at the bottom of Fig. 17). We 
would be happy to include this more prominently in the second phase submission 
(it requires showing a second, more deeply nested hole environment, so we will
need the additional space available in the second phase submission).

## Control Flow

The sum types extension in the appendix also adds
primitive control flow. With the additional page budget, we will include this
extension in the main body of the second phase submission.

## Novelty

> While, the technical development is sound, it is mostly based prior work (Hazelnut type system), or a variation on gradually typed calculi, and not very surprising. In fact, evaluation of expression with holes is akin to evaluation of expressions with free variables extended with type casts from gradual typing.

...

### 2018-06-07 rebuttal.txt
edits to rebuttal A

@@
-Thank you for your thoughtful reviews. We believe...
+Thank you for your thoughtful comments, all of which will be incorporated in our
+revision.
 
 # Review A
+
+> I'm still confused about which version of the language is implemented.
+> Are the screenshots in section 2 real, or only the ones in the appendix?
+
+The submitted implementation supports the language of Section 3 plus the
+extensions in the appendix.
+
+The screenshots in Section 2 are mockups, depicting language features that were
+not implemented at the time of submission. To create the screenshots in Section
+2, we first ran simpler versions of the examples (using encodings based on the
+core features), and then we manually post-processed the resulting HTML/CSS of
+the editor to obtain the mockups. The screenshots in the Appendix show the
+implementation at the time of submission. We will make this explanation even
+clearer in our revision.
+
+For artifact evaluation, we plan to submit a more recent snapshot of the Hazel
+implementation that either substantially narrows or eliminates the
+``convenience gap'' between the core language and the features depicted in
+Section 2.
 
@@
 
-That said, we agree that an explicit fix operator in the style of Plotkin's PCF
-might be more clear and there is nothing that would complicate its inclusion, 
-so if requested, we would be happy to include it in the second phase submission, 
-either in the main calculus or as another extension.
-
-> I'm still confused about which version of the language is implemented.
-
-The language of the submitted implementation is the language of Section 3 + 
-the extensions in the appendix. We plan to archive a frozen version of that 
-implementation as a simple reference implementation of the core calculus of 
-this paper.
-
-The first two paragraphs of Section 2 were, we believe, direct in stating that 
-the figures in Section 2 were post-processed to use certain language features 
-that were not implemented at the time of submission. None of these language 
-features are particularly complex or unusual (records, simple pattern matching, 
-and syntactic sugar for function definitions and lists), and they are 
-orthogonal to the live programming features that are the topic of the paper, 
-which are all implemented as described in Section 2 (see the implementation 
-screenshots in Fig. 17-18 in the appendix).
-
-We plan to submit a more recent snapshot of the Hazel implementation for 
-artifact evaluation that either substantially narrows or eliminates this 
-``convenience gap''.
-
-Our approach has been to...
+That said, we agree that an explicit fix operator might be more clear; there is
+nothing that would complicate its inclusion, so if requested in our revision, we
+would be happy to include either in the main calculus or as an extension.
 

### 2018-06-07 rebuttal.txt
added initial text about novelty

--- 2018-06-07 rebuttal.txt
@@
 
-# Review B
+# Review C
 
-(nothing to say, maybe just acknowledge the typos? remove this section entirely?)
+> While, the technical development is sound, it is mostly based prior work 
+> (Hazelnut type system), or a variation on gradually typed calculi, and 
+> not very surprising. In fact, evaluation of expression with holes is akin 
+> to evaluation of expressions with free variables extended with type casts 
+> from gradual typing.
 
-# Review C
+We certainly build directly upon the Hazelnut paper, as reviewed in Sec. 3.1, 
+but that paper did not define a dynamic semantics at all.
+
+Gradually typed lambda calculi consider only type holes, and cast failure 
+immediately aborts evaluation. We do not abort immediately (Sec. 2.4). Moreover,
+we consider not just type holes but also expression holes, which operate not 
+merely as free variables, as suggested, but as closures. Closure tracking is 
+critical to the user experience in Sec. 2 and to the fill-and-resume feature 
+described in Sec. 4.
+
+Closure tracking has a conceptual relationship with prior work on CMTT 
+(a non-obvious insight, we believe!), but that work also did not define an
+operational semantics nor consider the commutativity property in Sec. 4.
+
+So the novelty is in carefully combining all three of these pieces of prior work 
+and then defining the necessary technical infrastructure (e.g. indeterminate 
+forms) to be able to prove preservation, progress and commutativity theorems. 
+We are not aware of any comparable related work.
 
@@
 ## Novelty
-
-> While, the technical development is sound, it is mostly based prior work (Hazelnut type system), or a variation on gradually typed calculi, and not very surprising.
-In fact, evaluation of expression with holes is akin to evaluation of expressions with free variables extended with type casts from gradual typing.
-
-...

### 2018-06-07 rebuttal.txt
revision to review A response

@@
 The submitted implementation supports the language of Section 3 plus the
-extensions in the appendix.
+extensions in the appendix. The screenshots in the Appendix show the 
+implementation at the time of submission.
 
-The screenshots in Section 2 are mockups, depicting language features that were
-not implemented at the time of submission. To create the screenshots in Section
-2, we first ran simpler versions of the examples (using encodings based on the
-core features), and then we manually post-processed the resulting HTML/CSS of
-the editor to obtain the mockups. The screenshots in the Appendix show the
-implementation at the time of submission. We will make this explanation even
-clearer in our revision.
+All live programming features described in Section 2 are implemented. However,
+the figures in Section 2 do depict certain language conveniences that were not 
+implemented at the time of submission (per the intro of Section 2). To create 
+the figures, we first ran simpler versions of the examples (using encodings 
+based on the core calculus), and then we manually post-processed the resulting 
+HTML/CSS of the editor to depict standard conveniences common to languages like 
+Elm and ML, e.g. records, basic pattern matching and syntactic sugar for 
+function definitions and lists.
 
 For artifact evaluation, we plan to submit a more recent snapshot of the Hazel
-implementation that either substantially narrows or eliminates the
-``convenience gap'' between the core language and the features depicted in
-Section 2.
+implementation that either substantially narrows or eliminates this 
+"convenience gap".
 
@@
 nothing that would complicate its inclusion, so if requested in our revision, we
-would be happy to include either in the main calculus or as an extension.
+would be happy to include it either in the main calculus or as an extension.
 

### 2018-06-07 rebuttal.txt
edits to rebuttal C

@@
-Thank you for your thoughtful comments, all of which will be incorporated in our
-revision.
+Thank you for your thoughtful comments.
 
@@
 
-We certainly build directly upon the Hazelnut paper, as reviewed in Sec. 3.1, 
-but that paper did not define a dynamic semantics at all.
+Our focus is the language from the Hazelnut paper, but that paper did not define
+a dynamic semantics at all. The solution we propose in this paper certainly
+builds on prior work (specifically, gradually typed calculi and CMTT), but we
+believe the differences are noteworthy.
 
-Gradually typed lambda calculi consider only type holes, and cast failure 
+1. Gradually typed lambda calculi consider only type holes, and cast failure
 immediately aborts evaluation. We do not abort immediately (Sec. 2.4). Moreover,
@@
 
-Closure tracking has a conceptual relationship with prior work on CMTT 
+2. Closure tracking has a conceptual relationship with prior work on CMTT
 (a non-obvious insight, we believe!), but that work also did not define an
@@
 
-So the novelty is in carefully combining all three of these pieces of prior work 
-and then defining the necessary technical infrastructure (e.g. indeterminate 
-forms) to be able to prove preservation, progress and commutativity theorems. 
+Combining these pieces of prior work with the necessary technical infrastructure
+(e.g. indeterminate forms) to be able to prove preservation, progress and
+commutativity theorems was not (for us, at least) completely straightforward; we
+iterated through several designs before settling on the proposal in the paper.
 We are not aware of any comparable related work.
@@
 
-## Recursion
+## Semantics for Recursion and Control Flow
 
-See the response to Review A above regarding recursion in the core calculus.
+Please see the discussion above regarding recursion.
 
-We also considered the topic of recursion in Sec. 2.2 (Example 2: Recursive 
-Functions). The UI was designed with recursion in mind in that it allows the 
-user to explore the recursive structure of hole closures iteratively as 
-described in that section, rather than displaying nested hole closures all at
-once, as is the case in the notation used for the core calculus.
+The sum types extension in the appendix adds primitive control flow. With the
+additional page budget, we will include this extension in the main body of the
+second phase submission.
 
-The UI displays a "breadcrumb" path from the result to reflect the exploration 
-that the user has done (the basic UI can be seen at the bottom of Fig. 17). We 
-would be happy to include this more prominently in the second phase submission 
-(it requires showing a second, more deeply nested hole environment, so we will
-need the additional space available in the second phase submission).
+## UI for Recursion and Control Flow
 
-## Control Flow
+We describe our initial UI design considerations for recursion in Sec. 2.2. The
+UI allows the user to explore the recursive structure of hole closures
+iteratively as described in that section, rather than displaying nested hole
+closures all at once, as is the case in the notation used for the core calculus.
 
-The sum types extension in the appendix also adds
-primitive control flow. With the additional page budget, we will include this
-extension in the main body of the second phase submission.
+Regarding control flow, the UI displays a "breadcrumb" path from the result to
+reflect the exploration that the user has done (the basic UI can be seen at the
+bottom of Fig. 17). We would be happy to include this more prominently in the
+second phase submission (it requires showing a second, more deeply nested hole
+environment, so we will need the additional space available in the second phase
+submission).
 
-## Novelty
+## Effects and Mutation
+
+TODO

### 2018-06-07 rebuttal.txt
a few more edits from meeting with ravi

@@
 
-Our focus is the language from the Hazelnut paper, but that paper did not define
+We start from the Hazelnut paper, but that paper did not define
 a dynamic semantics at all. The solution we propose in this paper certainly
-builds on prior work (specifically, gradually typed calculi and CMTT), but we
+builds on prior work (specifically, gradual typing and CMTT), but we
 believe the differences are noteworthy.
 
-1. Gradually typed lambda calculi consider only type holes, and cast failure
+Gradually typed lambda calculi consider only type holes, and cast failure
 immediately aborts evaluation. We do not abort immediately (Sec. 2.4). Moreover,
@@
 
-2. Closure tracking has a conceptual relationship with prior work on CMTT
+Closure tracking has a conceptual relationship with prior work on CMTT
 (a non-obvious insight, we believe!), but that work also did not define an
@@
 (e.g. indeterminate forms) to be able to prove preservation, progress and
-commutativity theorems was not (for us, at least) completely straightforward; we
-iterated through several designs before settling on the proposal in the paper.
-We are not aware of any comparable related work.
+commutativity theorems was not completely straightforward. We are not aware of 
+any comparable related work.
 
@@
 
-## UI for Recursion and Control Flow
+## UI
 
@@
 
-Regarding control flow, the UI displays a "breadcrumb" path from the result to
+Regarding , the UI displays a "breadcrumb" path from the result to
 reflect the exploration that the user has done (the basic UI can be seen at the
@@
 TODO
+
+Finish on

### 2018-06-07 rebuttal.txt
more edits to implementation and novelty discussions in rebuttal -- will get to the rest of Review C soon

@@
-Thank you for your thoughtful comments.
+Thank you for your thoughtful feedback!
 
 # Review A
+
+> it doesn't even have fixpoint
+
+Type holes actually allow you to express general recursion (e.g. via the Y 
+combinator) - see footnote 9 and Siek and Taha [2006].
+
+That said, we agree that an explicit fix operator might be more clear; there is
+nothing that would complicate its inclusion, so if requested, we would be happy 
+to include it either in the main calculus or as an extension.
 
@@
 
-The submitted implementation supports the language of Section 3 plus the
-extensions in the appendix. The screenshots in the Appendix show the 
-implementation at the time of submission.
+The screenshots in the Appendix show the submitted implementation, which 
+supports all of the live programming features described in Section 2 for 
+the language of Section 3 plus the extensions in the appendix.
 
-All live programming features described in Section 2 are implemented. However,
-the figures in Section 2 do depict certain language conveniences that were not 
-implemented at the time of submission (per the intro of Section 2). To create 
-the figures, we first ran simpler versions of the examples (using encodings 
-based on the core calculus), and then we manually post-processed the resulting
-HTML/CSS of the editor to depict standard conveniences common to languages like 
-Elm and ML, e.g. records, basic pattern matching and syntactic sugar for 
+For the sake of exposition, the figures in Section 2 do depict certain language
+features that were not implemented at submission (as acknowledged in the 
+intro of Section 2). To create these figures, we first ran simpler variants of 
+the examples (using the usual lambda encoding tricks), and then we manually
+edited the resulting HTML/CSS to add in conveniences common to languages like 
+ML/Elm, namely record labels, basic pattern matching and syntactic sugar for 
 function definitions and lists.
@@
 "convenience gap".
-
-> The HazelnutLive calculus in the paper is really basic, it doesn't even have 
-> fixpoint.
-
-Type holes actually allow you to express general recursion (e.g. via the Y 
-combinator). See Siek and Taha [2006] for more on embedding the untyped lambda
-calculus into a gradually typed lambda calculus like Hazelnut Live.
-
-That said, we agree that an explicit fix operator might be more clear; there is
-nothing that would complicate its inclusion, so if requested in our revision, we
-would be happy to include it either in the main calculus or as an extension.
 
@@
 
-We start from the Hazelnut paper, but that paper did not define
-a dynamic semantics at all. The solution we propose in this paper certainly
-builds on prior work (specifically, gradual typing and CMTT), but we
-believe the differences are noteworthy.
+Only the external type system comes from the Hazelnut paper (reviewed in 3.1). 
+That paper did not define a dynamic semantics at all.
 
-Gradually typed lambda calculi consider only type holes, and cast failure
-immediately aborts evaluation. We do not abort immediately (Sec. 2.4). Moreover,
-we consider not just type holes but also expression holes, which operate not
-merely as free variables, as suggested, but as closures. Closure tracking is
-critical to the user experience in Sec. 2 and to the fill-and-resume feature
-described in Sec. 4.
+Unlike gradually typed lambda calculi, we do not abort immediately on cast
+failure (Sec. 2.4). Moreover, we consider not just type holes but also 
+expression holes, which operate not merely as free variables, as suggested, 
+but as a sort of closure. Closure tracking is critical to the user experience 
+throughout Sec. 2 and to the fill-and-resume feature described in Sec. 4.
 
-Closure tracking has a conceptual relationship with prior work on CMTT
-(a non-obvious insight, we believe!), but that work also did not define an
-operational semantics nor consider the commutativity property in Sec. 4.
+Closure tracking has a conceptual relationship with the prior work on CMTT (a 
+non-obvious and satisfying logical connection, we believe!), but that work 
+also did not define a dynamic semantics nor consider non-empty holes nor the 
+commutativity property in Sec. 4.
 
-Combining these pieces of prior work with the necessary technical infrastructure
-(e.g. indeterminate forms) to be able to prove preservation, progress and
-commutativity theorems was not completely straightforward. We are not aware of 
-any comparable related work.
+We carefully combined mechanisms from these prior systems and define new 
+technical infrastructure (e.g. indeterminate forms) to be able to prove 
+preservation, progress and commutativity theorems mechanically. We are not 
+aware of an operational semantics with comparable features and metatheory.
 

### 2018-06-07 rebuttal.txt
more on UI stuff -- just need to talk about effects now

@@
 Type holes actually allow you to express general recursion (e.g. via the Y 
-combinator) - see footnote 9 and Siek and Taha [2006].
+combinator) - see footnote 9 and [Siek and Taha, 2006] on embedding the 
+untyped lambda calculus.
 
@@
 supports all of the live programming features described in Section 2 for 
-the language of Section 3 plus the extensions in the appendix.
+the language of Section 3 plus the extensions in Appendix B.
 
@@
 features that were not implemented at submission (as acknowledged in the 
-intro of Section 2). To create these figures, we first ran simpler variants of 
-the examples (using the usual lambda encoding tricks), and then we manually
-edited the resulting HTML/CSS to add in conveniences common to languages like 
-ML/Elm, namely record labels, basic pattern matching and syntactic sugar for 
-function definitions and lists.
+intro to Section 2). To create these figures, we first ran simpler variants of 
+the examples (using typical lambda encoding tricks) in our implementation, and 
+then we manually edited the resulting HTML/CSS only to add in conveniences 
+common to languages like ML/Elm, namely record labels, basic pattern matching 
+and syntactic sugar for function definitions and lists.
 
 For artifact evaluation, we plan to submit a more recent snapshot of the Hazel
-implementation that either substantially narrows or eliminates this 
+implementation that either substantially narrows or eliminates this small 
 "convenience gap".
@@
 > to evaluation of expressions with free variables extended with type casts
-> from gradual typing.
 
@@
 
-Unlike gradually typed lambda calculi, we do not abort immediately on cast
-failure (Sec. 2.4). Moreover, we consider not just type holes but also 
-expression holes, which operate not merely as free variables, as suggested, 
-but as a sort of closure. Closure tracking is critical to the user experience 
-throughout Sec. 2 and to the fill-and-resume feature described in Sec. 4.
+Unlike gradually typed calculi, we do not abort immediately on cast failure 
+(Sec. 2.4). Moreover, we consider not only type holes but also expression 
+holes, which operate not merely as free variables (which, as suggested, would
+be rather uninteresting), but as metavariable closures. Closure tracking is 
+critical to the user experience throughout Sec. 2 and to the fill-and-resume 
+feature in Sec. 4.
 
-Closure tracking has a conceptual relationship with the prior work on CMTT (a 
-non-obvious and satisfying logical connection, we believe!), but that work 
-also did not define a dynamic semantics nor consider non-empty holes nor the 
+Closure tracking has a relationship with the prior work on CMTT (a non-obvious 
+and satisfying logical connection, we believe!), but that work also did not 
+define a dynamic semantics, nor did it consider non-empty holes nor the 
 commutativity property in Sec. 4.
 
-We carefully combined mechanisms from these prior systems and define new 
-technical infrastructure (e.g. indeterminate forms) to be able to prove 
+We carefully combined mechanisms from these prior systems and defined new 
+technical infrastructure (e.g. indeterminate forms) to be able to prove clean 
 preservation, progress and commutativity theorems mechanically. We are not 
-aware of an operational semantics with comparable features and metatheory.
+aware of a dynamic semantics with comparable features and metatheory.
 
-> the calculus lacks recursion and control-flow. Dealing with these efficiently 
-> and providing a good user interface for inspecting the resulting reified values 
-> is the hard problem in making such an environment.
+> the calculus lacks recursion and control-flow
 
-## Semantics for Recursion and Control Flow
+Please see the response above regarding recursion. The sum types extension 
+in Appendix B adds primitive control flow. With the additional page budget, 
+we could move this extension into the main body.
 
-Please see the discussion above regarding recursion.
+> providing a good user interface for inspecting the resulting reified values 
+> is the hard problem in making such an environment
 
-The sum types extension in the appendix adds primitive control flow. With the
-additional page budget, we will include this extension in the main body of the
-second phase submission.
+We agree that both theoretical and UI considerations are worthy of substantial 
+investigation. This paper focused on laying detailed type-theoretic foundations
+motivated by one proof-of-concept UI design and several small but non-trivial 
+examples.
 
-## UI
+This UI was designed explicitly with recursion in mind in that it allows 
+the user to explore the recursive structure of hole closures iteratively (see 
+Sec 2.2). It does not display nested hole closures all at once, as in the 
+notation used in the core calculus. We say a little more below the fold.
 
-We describe our initial UI design considerations for recursion in Sec. 2.2. The
-UI allows the user to explore the recursive structure of hole closures
-iteratively as described in that section, rather than displaying nested hole
-closures all at once, as is the case in the notation used for the core calculus.
+Branches blocked on a hole can certainly "blow up" in size. A simple, partial 
+solution is to visually collapse such sub-expressions by default. Because we 
+define a small step semantics, another approach is to pause evaluation when 
+this situation occurs and continue only if explicitly requested. Because we 
+leave evaluation order undefined, the user could also explicitly shift the 
+"evaluation cursor" elsewhere in such a situation.
+
+## Effects and Mutation
+
+TODO
+
+-------
 
@@
 submission).
-
-## Effects and Mutation
-
-TODO
-
-Finish on

### 2018-06-08 rebuttal.txt
space compression part 1

@@
 nothing that would complicate its inclusion, so if requested, we would be happy
-to include it either in the main calculus or as an extension.
+to make this small addition, either in the main calculus or as an extension.
 
-> I'm still confused about which version of the language is implemented.
 > Are the screenshots in section 2 real, or only the ones in the appendix?
@@
 
-For the sake of exposition, the figures in Section 2 do depict certain language
-features that were not implemented at submission (as acknowledged in the 
-intro to Section 2). To create these figures, we first ran simpler variants of 
-the examples (using typical lambda encoding tricks) in our implementation, and 
-then we manually edited the resulting HTML/CSS only to add in conveniences 
-common to languages like ML/Elm, namely record labels, basic pattern matching 
-and syntactic sugar for function definitions and lists.
+For the sake of exposition, the figures in Section 2 do go beyond the language
+of Section 3 (as acknowledged in the intro to Section 2). To create these figures, 
+we first ran simpler variants of the examples (using typical lambda encoding 
+tricks) in our implementation, and then we manually edited the resulting 
+HTML/CSS only to add in conveniences common to languages like ML/Elm, namely 
+record labels, basic pattern matching and syntactic sugar for function 
+definitions and lists.
 
@@
 > to evaluation of expressions with free variables extended with type casts
+
+We disagree with this characterization.
 
@@
 
-We carefully combined mechanisms from these prior systems and defined new 
-technical infrastructure (e.g. indeterminate forms) to be able to prove clean 
-preservation, progress and commutativity theorems mechanically. We are not 
-aware of a dynamic semantics with comparable features and metatheory.
-
 > the calculus lacks recursion and control-flow
 
-Please see the response above regarding recursion. The sum types extension 
-in Appendix B adds primitive control flow. With the additional page budget, 
-we could move this extension into the main body.
+Please see above regarding recursion. The sum types extension in Appendix B 
+adds primitive control flow. With the additional page budget, we could move 
+this extension into the main body.
 
 > providing a good user interface for inspecting the resulting reified values 
-> is the hard problem in making such an environment
+> is the hard problemin making such an environment
 
 We agree that both theoretical and UI considerations are worthy of substantial 
-investigation. This paper focused on laying detailed type-theoretic foundations
-motivated by one proof-of-concept UI design and several small but non-trivial 
-examples.
+investigation. This paper focused on the former, laying detailed type-theoretic
+foundations motivated by one proof-of-concept UI design and several small but 
+non-trivial examples.
 
 This UI was designed explicitly with recursion in mind in that it allows 
-the user to explore the recursive structure of hole closures iteratively (see 
-Sec 2.2). It does not display nested hole closures all at once, as in the 
+the user to explore the recursive structure of hole closures iteratively (Sec 2.2). 
+It does not display nested hole closures all at once, as in the 
 notation used in the core calculus. We say a little more below the fold.

### 2018-06-08 rebuttal.txt
missing space
exactly 600 words before the fold

@@
 
-Type holes actually allow you to express general recursion (e.g. via the Y 
+Type holes actually give you general recursion (e.g. via the Y 
 combinator) - see footnote 9 and [Siek and Taha, 2006] on embedding the 
@@
 
-That said, we agree that an explicit fix operator might be more clear; there is
-nothing that would complicate its inclusion, so if requested, we would be happy
-to make this small addition, either in the main calculus or as an extension.
+That said, we agree that an explicit fix operator might be more clear; if 
+requested, we would be happy to make this small + simple addition.
 
@@
 
-The screenshots in the Appendix show the submitted implementation, which 
-supports all of the live programming features described in Section 2 for 
-the language of Section 3 plus the extensions in Appendix B.
+The appendix shows the submitted implementation, which 
+supports all of the live programming features in Section 2 for 
+the language of Section 3 plus Appendix B.
 
 For the sake of exposition, the figures in Section 2 do go beyond the language
-of Section 3 (as acknowledged in the intro to Section 2). To create these figures, 
-we first ran simpler variants of the examples (using typical lambda encoding 
-tricks) in our implementation, and then we manually edited the resulting 
-HTML/CSS only to add in conveniences common to languages like ML/Elm, namely 
-record labels, basic pattern matching and syntactic sugar for function 
-definitions and lists.
+of Section 3 (acknowledged in the intro to Section 2). To create these figures, 
+we ran simpler variants of the examples (using typical lambda encodings), then 
+manually edited the HTML/CSS only to add conveniences from Elm, namely record 
+labels, pattern matching and sugar for functions and lists.
 
-For artifact evaluation, we plan to submit a more recent snapshot of the Hazel
-implementation that either substantially narrows or eliminates this small 
-"convenience gap".
+(We plan to submit a more recent snapshot of Hazel that substantially narrows 
+or eliminates this small "convenience gap" for artifact evaluation.)
 
@@
 
-> While, the technical development is sound, it is mostly based prior work 
-> (Hazelnut type system), or a variation on gradually typed calculi, and 
-> not very surprising. In fact, evaluation of expression with holes is akin 
-> to evaluation of expressions with free variables extended with type casts
+> mostly based prior work (Hazelnut type system), or a variation on gradually 
+> typed calculi, and not very surprising. In fact, evaluation of expression 
+> with holes is akin to evaluation of expressions with free variables extended 
+> with type casts
 
@@
 
-Only the external type system comes from the Hazelnut paper (reviewed in 3.1). 
-That paper did not define a dynamic semantics at all.
+Only the external type system comes from Hazelnut (Sec 3.1). That paper did 
+not define a dynamic semantics at all.
 
-Unlike gradually typed calculi, we do not abort immediately on cast failure 
-(Sec. 2.4). Moreover, we consider not only type holes but also expression 
-holes, which operate not merely as free variables (which, as suggested, would
-be rather uninteresting), but as metavariable closures. Closure tracking is 
-critical to the user experience throughout Sec. 2 and to the fill-and-resume 
-feature in Sec. 4.
+Unlike gradually typed calculi, we do not abort on cast failure. Furthermore, 
+we consider not only type holes but also expression holes, which operate not 
+merely as free variables (which, as suggested, would be uninteresting), but 
+as metavariable closures. Closure tracking is critical to the features in  
+Sections 2 and 4.
 
-Closure tracking has a relationship with the prior work on CMTT (a non-obvious 
+Closure tracking has a relationship with CMTT (a non-obvious 
 and satisfying logical connection, we believe!), but that work also did not 
-define a dynamic semantics, nor did it consider non-empty holes nor the 
-commutativity property in Sec. 4.
+define a dynamic semantics, nor did it consider non-empty holes or the 
+commutativity property.
 
-> the calculus lacks recursion and control-flow
+> calculus lacks recursion and control-flow
 
 Please see above regarding recursion. The sum types extension in Appendix B 
-adds primitive control flow. With the additional page budget, we could move 
-this extension into the main body.
+adds control flow. We plan to move this into the main body.
 
 > providing a good user interface for inspecting the resulting reified values 
-> is the hard problemin making such an environment
+> is the hard problem
 
-We agree that both theoretical and UI considerations are worthy of substantial 
-investigation. This paper focused on the former, laying detailed type-theoretic
-foundations motivated by one proof-of-concept UI design and several small but 
-non-trivial examples.
+Both theoretical and UI considerations are worthy of substantial 
+investigation. This paper focuses on the former, motivated by a  
+proof-of-concept UI design and several small but non-trivial examples.
 
-This UI was designed explicitly with recursion in mind in that it allows 
-the user to explore the recursive structure of hole closures iteratively (Sec 2.2). 
-It does not display nested hole closures all at once, as in the 
-notation used in the core calculus. We say a little more below the fold.
+This UI was designed with recursion in mind -- it allows you to  
+to explore the recursive structure of closures iteratively (Sec 2.2). 
+It does not display nested closures all at once, as in the 
+formal notation. [more below the fold]
 
-Branches blocked on a hole can certainly "blow up" in size. A simple, partial 
-solution is to visually collapse such sub-expressions by default. Because we 
+Branches blocked on a hole can certainly "blow up" in size. A partial 
+solution is to visually collapse such subexpressions by default. Because we 
 define a small step semantics, another approach is to pause evaluation when 
-this situation occurs and continue only if explicitly requested. Because we 
-leave evaluation order undefined, the user could also explicitly shift the 
-"evaluation cursor" elsewhere in such a situation.
+this occurs, continuing only if requested or, because we leave evaluation order
+undefined, allowing the user to shift the "evaluation cursor" elsewhere.
 
-## Effects and Mutation
+>> effects and mutation
 
-TODO
+The interactions between gradual typing and mutation are well-studied and 
+the standard approach (treating refs as invariant wrt consistency) should 
+be applicable to recover the properties of Section 3. However, the 
+commutativity theorem in Section 4 would not hold.
+
+We anticipate that future work will explore variants of fill-and-resume
+that support a weaker commutativity property (e.g. via checkpointing).
+
+However, in order to provide the simplest and strongest theoretical
+guarantees, we did not include mutation. (Note that Hazel is 
+based on Elm, which is pure and quite practical.)
+Thank you for your thoughtful feedback!
+
+# Review A
+
+> it doesn't even have fixpoint
+
+Type holes actually give you general recursion (e.g. via the Y 
+combinator) - see footnote 9 and [Siek and Taha, 2006] on embedding the 
+untyped lambda calculus.
+
+That said, we agree that an explicit fix operator might be more clear; if 
+requested, we would be happy to make this small + simple addition.
+
+> Are the screenshots in section 2 real, or only the ones in the appendix?
+
+The appendix shows the submitted implementation, which 
+supports all of the live programming features in Section 2 for 
+the language of Section 3 plus Appendix B.
+
+For the sake of exposition, the figures in Section 2 do go beyond the language
+of Section 3 (acknowledged in the intro to Section 2). To create these figures, 
+we ran simpler variants of the examples (using typical lambda encodings), then 
+manually edited the HTML/CSS only to add conveniences from Elm, namely record 
+labels, pattern matching and sugar for functions and lists.
+
+(We plan to submit a more recent snapshot of Hazel that substantially narrows 
+or eliminates this small "convenience gap" for artifact evaluation.)
+
+# Review C
+
+> mostly based prior work (Hazelnut type system), or a variation on gradually 
+> typed calculi, and not very surprising. In fact, evaluation of expression 
+> with holes is akin to evaluation of expressions with free variables extended 
+> with type casts
+
+We disagree with this characterization.
+
+Only the external type system comes from Hazelnut (Sec 3.1). That paper did 
+not define a dynamic semantics at all.
+
+Unlike gradually typed calculi, we do not abort on cast failure. Furthermore, 
+we consider not only type holes but also expression holes, which operate not 
+merely as free variables (which, as suggested, would be uninteresting), but 
+as metavariable closures. Closure tracking is critical to the features in  
+Sections 2 and 4.
+
+Closure tracking has a relationship with CMTT (a non-obvious 
+and satisfying logical connection, we believe!), but that work also did not 
+define a dynamic semantics, nor did it consider non-empty holes or the 
+commutativity property.
+
+> calculus lacks recursion and control-flow
+
+Please see above regarding recursion. The sum types extension in Appendix B 
+adds control flow. We plan to move this into the main body.
+
+> providing a good user interface for inspecting the resulting reified values 
+> is the hard problem
+
+Both theoretical and UI considerations are worthy of substantial 
+investigation. This paper focuses on the former, motivated by a  
+proof-of-concept UI design and several small but non-trivial examples.
+
+This UI was designed with recursion in mind -- it allows you to  
+to explore the recursive structure of closures iteratively (Sec 2.2). 
+It does not display nested closures all at once, as in the 
+formal notation. [more below the fold]
+
+Branches blocked on a hole can certainly "blow up" in size. A partial 
+solution is to visually collapse such subexpressions by default. Because we 
+define a small step semantics, another approach is to pause evaluation when 
+this occurs, continuing only if requested or, because we leave evaluation order
+undefined, allowing the user to shift the "evaluation cursor" elsewhere.
+
+>> effects and mutation
+
+The interactions between gradual typing and mutation are well-studied and 
+the standard approach (treating refs as invariant wrt consistency) should 
+be applicable to recover the properties of Section 3. However, the 
+commutativity theorem in Section 4 would not hold.
+
+We anticipate that future work will explore variants of fill-and-resume
+that support a weaker commutativity property (e.g. via checkpointing).
+
+However, in order to provide the simplest and strongest theoretical
+guarantees, we did not include mutation. (Note that Hazel is 
+based on Elm, which is pure and quite practical.)
 

### 2018-06-30 calculus.tex
fixed typos identified by OOPSLA reviews

@@
 \noindent
-The type assignment rules for hole closures~{TAEHole} and {TANEHole} each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
+The type assignment rules for hole closures,~\rulename{TAEHole} and \rulename{TANEHole}, each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
 %
@@
 Consequently, the environments of different closures with the same hole name may differ,
-e.g., when an reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.
+e.g., when a reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.
 Hole closures may also appear within the environments of other hole
@@
 %
-Critically, this property relies on the more version of
+Critically, this property relies on the version of
 the semantics from \Secref{sec:calculus} where evaluation order is
-unspecified (i.e. formally, we omit the bracketed premises).
+unspecified (i.e. where we omit the bracketed premises).
 %

### 2018-07-04 abstract.tex
tightened abstract

@@
  to programs that are \emph{complete}, i.e. syntactically well-formed and free
-of type errors, whereas 
-%
-during much of the editing process, the program is not yet complete. As a result,  
-live feedback flickers in and out or goes stale, 
+of type errors. As a result,  
+live feedback flickers out or goes stale, 
 in many cases for substantial lengths of time.
@@
 
-This paper confronts this problem from type-theoretic first principles by developing a 
-\emph{dynamic semantics for incomplete functional programs}, 
+This paper confronts this problem from type-theoretic first principles by developing 
+\emph{a dynamic semantics for incomplete functional programs}, 
 based in part on the static semantics for incomplete functional programs
@@
 Rather than aborting with an exception when evaluation encounters any of these holes (as in
-several existing systems), evaluation proceeds ``around'' the holes,
+several existing systems with holes), evaluation proceeds ``around'' the holes,
 performing as much of the remaining computation as is possible and tracking the 
@@
 type holes) and contextual modal type theory (which provides a
-logical foundation for hole closures), and develops additional technical machinery necessary 
-to continue evaluation past the various hole forms and to prove important metatheoretic properties.
+logical basis for hole closures), and develops additional technical machinery necessary 
+to continue evaluation past the various hole forms and to prove several important metatheoretic properties.
 
@@
 %
-We have mechanized the core formal development using the Agda proof assistant.
+We have mechanized this foundational calculus, called \HazelnutLive{}, using the Agda proof assistant.
 %
-We also describe a simple implementation, called \HazelnutLive{},
-that inserts holes as
-necessary during the editing process to guarantee that every edit state has some (possibly incomplete)
-type, based in part on the \Hazelnut edit action calculus in prior work by \citet{popl-paper}.
-%
-Taken together with the type safety property that this paper establishes, the result is 
-a proof-of-concept live typed functional programming environment where dynamic feedback 
-is truly continuous, i.e. it is available for every possible edit state.
+We also describe a simple implementation 
+that inserts holes automatically to guarantee that every editor state has some (possibly incomplete)
+type, based on the \Hazelnut edit action calculus of \citet{popl-paper}.
+Taken together with the type safety property that we establish, the 
+% The 
+result is  
+a proof-of-concept live programming environment where dynamic feedback 
+is truly continuous, i.e. it is available for every possible editor state.
 % i.e. \emph{every edit state is guaranteed to have

### 2018-07-04 calculus.tex
tightened abstract

@@
 treatment as an expression of a consistent type $\htau_2$
-(Rule TACast in \Figref{fig:hasType}).%
+(Rule~\rulename{TACast} in \Figref{fig:hasType}).%
 \footnote{

### 2018-07-06 abstract.tex
substantial revisions to abstract and introduction to add 'gap problem' development and cleaner text

@@
 
-\emph{Live programming environments} aim to provide programmers (and sometimes audiences) 
+Live programming environments aim to provide programmers (and sometimes audiences) 
 with continuous feedback about a program's dynamic behavior as it is being edited. 
@@
  to programs that are \emph{complete}, i.e. syntactically well-formed and free
-of type errors. As a result,  
-live feedback flickers out or goes stale, 
-in many cases for substantial lengths of time.
+of type errors. Consequently,    
+live feedback presented to the programmer has gaps, i.e. it flickers in and out or it goes stale whenever the program becomes incomplete. 
+% In some cases, these gaps persist over a substantial length of time.
 %
@@
 
-This paper confronts this problem from type-theoretic first principles by developing 
+This paper confronts this \emph{gap problem} from type-theoretic first principles by developing 
 \emph{a dynamic semantics for incomplete functional programs}, 
-based in part on the static semantics for incomplete functional programs
-developed in recent work by \citet{popl-paper}. 
+starting from the static semantics for incomplete functional programs
+developed by \citet{popl-paper}. 
 %
@@
 with empty holes standing for missing expressions or types, and  non-empty holes 
-operating as ``membranes'' around static and dynamic type inconsistencies. 
+operating as membranes around static and dynamic type inconsistencies. 
 %
-Rather than aborting with an exception when evaluation encounters any of these holes (as in
-several existing systems with holes), evaluation proceeds ``around'' the holes,
+Rather than aborting with an exception when evaluation encounters any of these holes as in
+several existing systems, evaluation proceeds around the holes,
 performing as much of the remaining computation as is possible and tracking the 
-closure around each hole instance as it flows through the program. Various editor services can report information from these hole closures 
-to help the programmer decide how to fill a hole. 
-They also enable a ``fill-and-resume'' feature that 
+closure around each hole instance as it flows through the program. Editor services can report information from these hole closures 
+to help the programmer decide how to fill the holes in the program. 
+Hole closures also enable a \emph{fill-and-resume} feature that 
 avoids the need to restart evaluation after edits that amount to hole filling. 
-Formally, the semantics draws from both gradual type theory (to handle
+Formally, the semantics borrows machinery from both gradual type theory (to handle
 type holes) and contextual modal type theory (which provides a
-logical basis for hole closures), and develops additional technical machinery necessary 
-to continue evaluation past the various hole forms and to prove several important metatheoretic properties.
+logical basis for tracking hole closures), combining these and developing additional machinery necessary 
+to continue evaluation past the hole forms in a manner that admits important metatheoretic properties.
 
@@
 %
-We have mechanized this foundational calculus, called \HazelnutLive{}, using the Agda proof assistant.
+We have mechanized this core calculus, called \HazelnutLive{}, using the Agda proof assistant.
 %
-We also describe a simple implementation 
+We also describe a companion implementation 
 that inserts holes automatically to guarantee that every editor state has some (possibly incomplete)
 type, based on the \Hazelnut edit action calculus of \citet{popl-paper}.
-Taken together with the type safety property that we establish, the 
+Taken together with this paper's type safety property, the 
 % The 
 result is  
-a proof-of-concept live programming environment where dynamic feedback 
-is truly continuous, i.e. it is available for every possible editor state.
+a proof-of-concept live programming environment where rich dynamic feedback 
+is truly available without gaps, i.e. it is available for every possible editor state.
 % i.e. \emph{every edit state is guaranteed to have

### 2018-07-06 intro.tex
substantial revisions to abstract and introduction to add 'gap problem' development and cleaner text

@@
 \label{sec:intro}
-\vspace{-2px}
 
@@
 
-Programmers typically shift back and forth between program editing and evaluation many times before converging upon a program that behaves correctly. 
-Live programming environments support this workflow by interleaving editing and evaluation so as to   
+Programmers typically shift back and forth between program editing and program evaluation many times before converging upon a program that behaves as intended. 
+Live programming environments aim to support this workflow by interleaving editing and evaluation so as to   
 narrow what \citet{burckhardt2013s} call ``the temporal and perceptive gap'' between these activities.
@@
 %
-and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding in an artistic performance.
+and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding as a performance art.
 Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support editing run-time state, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
@@
 %\IS entering
-enters
+is entering  
 a short expression.
@@
 %\matt{we are not really modeling a language with user type definitions -- what does the following sentence really add here? also, I just added this scenerio to the list above}
-The problem is particularly pronounced for languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
+The problem is particularly pronounced for languages with rich static type systems where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
 Throughout the process of addressing these errors, the program text remains formally meaningless.
@@
 %}
-In recognition of this problem---that incomplete programs lack meaning---%
-\citet{popl-paper} develop a static semantics (i.e. a type system) for incomplete 
+In recognition of this \emph{gap problem}---that incomplete programs are formally meaningless---\citet{popl-paper} develop a static semantics (i.e. a type system) for incomplete 
 functional programs, modeling them formally as typed expressions with \emph{holes} in 
@@
 % editor services can reason about types and binding in many more situations than previously possible.
-\citet{popl-paper,HazelnutSNAPL} discuss several ways to determine an incomplete expression from an editor state. 
-Briefly, error recovery mechanisms can insert the necessary holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09,DBLP:conf/oopsla/KatsV10},
-the programmer can insert explicit holes manually (see below) 
-or semi-automatically via code completion techniques \cite{Amorim2016}, 
-or a structure editor can insert holes fully automatically \cite{popl-paper} (we say more about this technique in Sec.~\ref{sec:implementation}).
+\citet{popl-paper,HazelnutSNAPL} discuss several ways to determine an incomplete expression from the editor state.
+When the editor state is a text buffer, error recovery mechanisms can insert holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09}.
+Alternatively, the language might provide explicit syntax for holes, so that the programmer can insert them either manually
+or semi-automatically via a code completion service \cite{Amorim2016}.
+For example, GHC Haskell supports the notation \li{_u} for empty holes, where \li{u} is an optional hole name \cite{GHCHoles}.
+Structure editors insert explicitly represented holes fully automatically \cite{popl-paper}; we say more about structure editors in Sec.~\ref{sec:implementation}.
+
 %
-For the purposes of live programming, however, a static semantics for incomplete programs does not suffice---%
-we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes. This paper aims to address this need by developing
-a {dynamic semantics} for incomplete functional programs, 
-starting from the static semantics of \citet{popl-paper}.
+For the purposes of live programming, however, a static semantics does not suffice---%
+we also need a corresponding dynamic semantics that specifies how to evaluate expressions with holes. %This paper addresses this need by developing
+% a {dynamic semantics} for incomplete functional programs, 
+% starting from the static semantics developed by \citet{popl-paper}.
 %
@@
 
-One approach would be to define a dynamic semantics that simply aborts with an error when evaluation reaches a hole. 
+The simplest approach would be to define a dynamic semantics that aborts with an error when evaluation reaches a hole. 
 %%%%%%%%%%%% New par:
 %
-This mirrors a workaround to the problem that programmers commonly deploy: 
-raising an exception where an expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
-Some languages provide more concise notation for this ``encoding'' of expression holes, 
-e.g. Scala provides \lismall{???}  
-and GHC Haskell with the \lismall{-fdefer-typed-holes} flag 
-%or \lismall{-fdefer-ty}\lismall{pe-errors} flags 
-uses \lismall{_u}
-where \lismall{u} serves as an
-optional hole name \cite{GHCHoles}. %
-% \footnote{
-Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
+This mirrors a workaround that programmers commonly deploy: 
+raising an exception as a placeholder, e.g. \lismall{raise Unimplemented}. 
+GHC Haskell supports this mode of evaluation for programs with holes using the \lismall{-fdefer-typed-holes} flag.\footnote{
+Without this flag, holes cause compilation to fail with an error message that reports information about each hole's type and typing context. 
 Proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} also respond to holes in this way.
-% }
-
-Although better than nothing, this approach of stopping with an exception upon reaching an expression hole has limitations 
-as the basis for a live programming environment because 
-(1)~it provides limited information about the dynamic state of the program when the exception occurs
+} 
+Although better than nothing, this exceptional approach to expression holes has limitations 
+within a live programming environment because 
+(1)~it provides limited information about the dynamic state of the program where the exception occurs
 (typically only a stack trace);  
 (2)~it provides no information about the behavior of the remainder of program, 
-parts of which may not depend on the missing or erroneous expression (e.g. later cells in a lab notebook); and 
+parts of which may not depend on the missing or erroneous expression (e.g. later cells in a lab notebook, or tests of those components of the program that are already complete); and 
 (3)~it provides no means by which to resume evaluation after filling a hole.
 
-Moreover, we need to be able to run programs containing not only expression holes but also type holes.
+Furthermore, exceptions can appear only in expressions, but we might also like to be able to evaluate programs that have type holes.
+Again, existing approaches do not support this situation well---GHC supports type holes, but compilation fails if type inference cannot automatically fill them \cite{GHCHoles}.
 The static semantics developed by \citet{popl-paper} derives the machinery for 
@@
 %
-As such, we can look to the dynamic semantics from
-gradual type theory to specify the dynamic semantics for incomplete programs,
+As such, we might look to the dynamic semantics from
+gradual type theory, 
 which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
@@
 However, when a cast fails, evaluation again stops with 
-an exception.
-%
-As a result, traditional gradual typing approaches still leave the live programming environment unable to provide feedback about the parts 
-of the program that do not depend on the problematic sub-expression.
+an exception and so the traditional gradual typing approaches still leave the live programming environment unable to provide rich, continuous feedback for the three reasons just enumerated.
 
@@
 
-This paper develops a theoretically well-grounded dynamic semantics for incomplete functional programs that addresses the limitations of the ``exceptional approach'' just described. 
+This paper develops a theoretically well-grounded dynamic semantics for incomplete functional programs, starting from the static semantics developed by \citet{popl-paper},  that addresses the limitations of the ``exceptional approach'' just enumerated.
 In particular, rather than stopping with an exception when evaluation encounters an expression hole, evaluation continues ``around'' the hole, performing as much of the remaining computation as possible.
-Evaluation also proceeds past failed casts in much the same way. 
-The system tracks the environment around each hole instance as evaluation proceeds. 
-The live programming environment can feed relevant information from these hole environments to the programmer as they work to fill the holes in the program. 
+For programs with unfillable type holes, casts are inserted and evaluation also proceeds past failed casts in much the same way. 
+The system tracks the closure around each hole instance as evaluation proceeds so that 
+the live programming environment can feed relevant information from the {hole closures} to the programmer as they work to fill the holes in the program. 
 Then, when the programmer performs an edit that fills an empty expression hole or that replaces a non-empty hole with a type-correct expression, evaluation can resume from the paused, i.e.
 \emph{indeterminate}, evaluation state.
+We call this feature \emph{fill-and-resume}.
+We are integrating this approach into the \Hazel programming environment being developed by \citet{HazelnutSNAPL}.
 
@@
 
-We are integrating this approach into the \Hazel programming environment being developed (separately from this paper) by \citet{HazelnutSNAPL}. 
-We begin in Sec.~\ref{sec:examples} by detailing the approach informally, with several examples, in the setting of this specific design.
+We begin in Sec.~\ref{sec:examples} by detailing the approach informally, with several example programming tasks, in the setting of the \Hazel design, which is based closely on well-established existing designs (in particular, on the Elm programming language and the Jupyter user interface).
+% We also qualitatively describe some user interface features that may help manage visual complexity when working with larger programs.
 
-Sec.~\ref{sec:calculus} then abstracts away orthogonal details of the language and user interface and makes these intuitions formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions as just outlined, in a manner that admits clean type safety theorems (in particular, a Progress theorem).
-The semantics of \HazelnutLive borrows machinery from gradual type theory~\cite{DBLP:conf/snapl/SiekVCB15}
-(to handle incomplete types) and contextual modal type theory~\cite{Nanevski2008} (CMTT,
-which provides a logical foundation for hole environment tracking). 
-These non-obvious connections to well-established existing systems help support our claim that this approach is theoretically well-grounded.
+Sec.~\ref{sec:calculus} then abstracts away the inessential details of the language and user interface and makes the  intuitions developed in Sec.~\ref{sec:examples} formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions and tracking hole closures as just outlined.
+The dynamic semantics of \HazelnutLive is a small-step reduction semantics equipped with clean type safety theorems (including, notably, a Progress theorem for incomplete expressions).
+The semantics of \HazelnutLive borrows some technical machinery from gradual type theory~\cite{DBLP:conf/snapl/SiekVCB15}
+(to handle incomplete types) and contextual modal type theory~(CMTT)~\cite{Nanevski2008} (which, by its Curry-Howard interpretation, provides a logical basis for reasoning about and operating on hole closures), combining these and adding additional machinery necessary to continue evaluation past holes and cast failures.
+Sec.~\ref{sec:agda-mechanization} outlines our Agda-based mechanization of \HazelnutLive, which is included in the anonymized supplement.  
+These connections to well-established existing systems, together with the mechanized proofs, serve to support our claim that this approach is theoretically well-grounded.
 
-Sec.~\ref{sec:agda-mechanization} outlines our mechanization of the core calculus using the Agda proof assistant. 
-Sec.~\ref{sec:implementation} outlines our simple implementation of the core calculus, which implements the live programming features from Sec.~\ref{sec:examples},
-but for the more austere language of Sec.~\ref{sec:calculus}. 
-The editor component of this implementation provides a language of structured edit actions, 
-based on the \Hazelnut structure editor calculus developed by \cite{popl-paper}, that guarantees that
+Sec.~\ref{sec:implementation} outlines our companion implementation of \HazelnutLive, which is also available in the anonymized supplement.
+This implementation is a slimmed-down version of \Hazel that implements the live programming features as shown in Sec.~\ref{sec:examples},
+but for the more austere language of Sec.~\ref{sec:calculus} extended with numbers, sum types, and product types (which, for the sake of simplicity, are specified separately in the anonymized supplement).
+The editor component of this implementation defines a language of structured edit actions, 
+based on the \Hazelnut structure editor calculus developed by \citet{popl-paper}, that inserts holes automatically to guarantee that
 every editor state has some, possibly incomplete, type. 
-Together with the theorems established earlier in Sec.~\ref{sec:calculus}, the result is a proof-of-concept
-live functional programming environment where every possible editor state has both non-trivial static and dynamic 
-meaning, i.e. live feedback never ``flickers'' or ``goes out''. 
+The type safety theorems established earlier in Sec.~\ref{sec:calculus} then guarantee that every editor state has dynamic meaning.
+Taken together, the result is a full solution to the gap problem, i.e.
+a proof-of-concept
+live functional programming environment where the static and dynamic feedback never exhibits gaps. We formally state this continuity corrollary.
+
+Sec.~\ref{sec:resumption} defines the fill-and-resume operation, which is rooted in the contextual substitution operation from CMTT.
+We establish the correctness of fill-and-resume with a commutativity theorem.
+We also discuss how the fill-and-resume operation allows us to semantically interpret the act of editing and evaluating cells in a REPL or Jupyter-like live lab notebook environment.
+
+Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneously discusses directions for future work. Sec.~\ref{sec:discussion} briefly concludes.
+
+%The appendix includes some straightforward auxiliary definitions and proofs mentioned in the paper, some extensions to the core calculus, and some screenshots and additional details on the implementation.
 The supplemental material includes both the mechanization and the implementation.
-
-Sec.~\ref{sec:resumption} defines the fill-and-resume feature, which is rooted in the contextual substitution operation from CMTT.
-We establish the correctness of fill-and-resume with a novel commutativity property and discuss how it provides a semantic interpretation of cells in a live lab notebook environment.
-
-Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneously discusses directions for future work.
-Sec.~\ref{sec:discussion} briefly concludes.
-The appendix includes some straightforward auxiliary definitions and proofs mentioned in the paper, some extensions to the core calculus, and some screenshots and additional details on the implementation.

### 2018-07-07 examples.tex
substantially expanded the related work
updated 2.0

@@
 
-We begin with an example-driven overview of our proposed approach  within the \Hazel programming environment \citet{HazelnutSNAPL}.
-\Hazel is an open source web lab notebook environment based roughly on IPython/Jupyter \cite{Perez:2007:ISI:1251563.1251831}.
-The language of \Hazel is tracking toward parity with \Elm, a popular pure functional programming language similar to ``core ML'' \cite{czaplicki2012elm,Elm}.
-We assume familiarity with the basic concepts of ML/Elm (see \url{http://elm-lang.org/}).
-
-Both the user interface and the language of \Hazel are under active development. For the sake of 
-exposition, we have ``post-processed'' the examples in this section to make use of several convenient language mechanisms from \Elm~that are not yet implemented in \Hazel as of this writing.
-These mechanisms are orthogonal to the contributions of this paper---we have implemented all of the live programming features described in this section, both in \Hazel and in the implementation of the core calculus included in the supplement (see Sec.~\ref{sec:implementation}).
+This section gives an example-driven overview of our  approach as implemented in  \Hazel, a browser-based lab notebook environment being developed by \cite{HazelnutSNAPL}.
+The \Hazel user interface is  based roughly on IPython/Jupyter \cite{PER-GRA:2007} and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}), a popular pure functional programming language similar to ``core ML'' \cite{czaplicki2012elm,Elm}, with which we assume basic  familiarity.
+For the sake of 
+exposition, we have ``post-processed'' the screenshots in this section to make use of a few language conveniences from \Elm~that were not available in \Hazel as of this writing, e.g.
+pattern matching in function arguments, list  notation, and record labels (currently there are only tuples).
+These conveniences are entirely orthogonal to the contributions of this paper: all of the live programming features demonstrated in this section have been implemented, both in \Hazel itself and in the simplified companion implementation of the core calculus in the supplement (see Sec.~\ref{sec:implementation}).
 

### 2018-07-07 examples-1.tex
fix spelling typo
submission
small edits to 2.1

@@
 to compute final student grades at the end of a course.
-The top of \autoref{fig:grades-example} depicts the cell containing the incomplete program that 
+\autoref{fig:grades-cell-mockup} depicts the cell containing the incomplete program that 
 the teacher has written so far (we omit irrelevant parts of the UI). 
@@
 \lismall{string}, and, for simplicity, three grades, each of type \lismall{float}.
-The variable \lismall{students} binds a list of such student records.
+Next, the teacher constructs a list of student records, binding it to the variable \lismall{students}.
 For simplicity, we include only three example students.
@@
 However, the program is incomplete because the teacher has not yet completed the body of the \lismall{weighted_average} function.
-This pattern is well-established: programmers often consume a function before implementing it.
+This pattern is quite common: programmers often consume a function before implementing it.
 
-Thusfar in Fig.~\ref{fig:grades-example}, the teacher has decomposed the function argument into variables by record destructuring, then multiplied the homework grade, \li{hw}, by \li{30.0} and finally inserted the \li{+} operator.
+Thusfar in the body of \lismall{weighted_average}, the teacher has decomposed the function argument into variables by record destructuring, then multiplied the homework grade, \li{hw}, by \li{30.0} and finally inserted the \li{+} operator.
 The cursor now sits at an \emph{empty hole}, as indicated by the vertical bar and the green background.
@@
 
-Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback available to the programmer.
+Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback available to the programmer at all.
 In response, the programmer might insert a temporary universal placeholder, e.g.
@@
 Evaluation cannot proceed any further because holes appear as addends.
-We say that each of these addition expressions is an \emph{indeterminate} sub-expression, and the result as a hole is therefore also indeterminate, because it is not yet a value, nor can it take a step due to holes in critical positions.
+We say that each of these addition expressions is an \emph{indeterminate} sub-expression, and the result as a hole is therefore also indeterminate, because it is not yet a value, nor can it take a step due to holes in elimination positions.
 
-Given this indeterminate result, the teacher might take notice of the magnitude of the numbers being computed, e.g. \li{2640.0} and \li{2280.0}, and
-decide immediately that a mistake has been made: the teacher wants to compute a 
+At this point, the teacher might take notice of the magnitude of the numbers being computed, e.g. \li{2640.0} and \li{2280.0}, and
+realize immediately that a mistake has been made: the teacher wants to compute a 
 weighted average between \li{0.0} and \li{100.0}, and so the correct
@@
 program can help confirm the mental model 
-that the programmer has developed (in this case, regarding the behavior of \li{map}), or else help quickly dispel 
+that the programmer has developed (in this case, regarding the behavior of \li{map}), and also help quickly dispel 
 inconsistencies between the programmer's expectations and the 
-actual behavior of the program (in this case, the arithmetic expression).
+actual behavior of the program (in this case, the magnitude of the arithmetic expressions being computed).
 
@@
 % Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
-\Hazel also helps programmers reason about the dynamic behavior of bound variables in scope at a hole via the \emph{live context inspector}, normally displays it as a sidebar but shown disembodied in three states in Fig.~\ref{fig:grades-sidebar}.
+Going further, \Hazel helps programmers reason about the dynamic behavior of bound variables in scope at a hole via the \emph{live context inspector}, normally displayed  as a sidebar but shown disembodied in three different states in Fig.~\ref{fig:grades-sidebar}.
 In all three states, the live context inspector displays the names and types of the variables that are in scope, as is a common feature of programming environments.
@@
 We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical foundations detailed in Sec.~\ref{sec:calculus}).
-The programmer can select different closures by clicking on a hole instance (here, instance \li{1:1} was selected by default, as indicated by the purple outline), or cycle rapidly through all of the closures for the hole at the cursor by clicking the arrows, e.g.
-$\blacktriangleright$, at the bottom of the context inspector, or corresponding hotkeys.
+The programmer can select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the solid purple outline), or cycle rapidly through all of the closures for the hole at the cursor by clicking the arrows, e.g.
+$\blacktriangleright$, at the bottom of the context inspector, or by using hotkeys.
 
 \vspace{-4px}
-\subsection{Example 2: Recursive Functions}\label{sec:qsort1}
+\subsection{Example 2: Recursive Functions}\label{sec:qsort1}\label{sec:paths}
 \vspace{-2px}
@@
 For example, the programmer can confirm that
-the lists \li{smaller} and \li{bigger} are appropriate named.
+the lists \li{smaller} and \li{bigger} are appropriately named.
 

### 2018-07-07 examples.tex
small edits to 2.1

@@
 
-This section gives an example-driven overview of our  approach as implemented in  \Hazel, a browser-based lab notebook environment being developed by \cite{HazelnutSNAPL}.
-The \Hazel user interface is  based roughly on IPython/Jupyter \cite{PER-GRA:2007} and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}), a popular pure functional programming language similar to ``core ML'' \cite{czaplicki2012elm,Elm}, with which we assume basic  familiarity.
+This section gives an example-driven overview of our  approach as implemented in  \Hazel, a browser-based lab notebook environment being developed by \citet{HazelnutSNAPL}.
+The \Hazel user interface is  based roughly on IPython/Jupyter \cite{PER-GRA:2007} and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume basic  familiarity.
 For the sake of 
@@
 pattern matching in function arguments, list  notation, and record labels (currently there are only tuples).
-These conveniences are entirely orthogonal to the contributions of this paper: all of the live programming features demonstrated in this section have been implemented, both in \Hazel itself and in the simplified companion implementation of the core calculus in the supplement (see Sec.~\ref{sec:implementation}).
+These conveniences are entirely orthogonal to the contributions of this paper: all of the live programming features shown in this section have been implemented, both in \Hazel itself and in the simplified companion implementation of the core calculus included in the supplement (see Sec.~\ref{sec:implementation}).
 

### 2018-07-09 examples-1.tex
updated example 2 to show three different hole closures and edits to 2.2

@@
 to compute final student grades at the end of a course.
-\autoref{fig:grades-cell-mockup} depicts the cell containing the incomplete program that 
+Fig.~\ref{fig:grades-cell-mockup} depicts the cell containing the incomplete program that 
 the teacher has written so far (we omit irrelevant parts of the UI). 
@@
 % Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
-Going further, \Hazel helps programmers reason about the dynamic behavior of bound variables in scope at a hole via the \emph{live context inspector}, normally displayed  as a sidebar but shown disembodied in three different states in Fig.~\ref{fig:grades-sidebar}.
+Going further, \Hazel helps programmers reason about the dynamic behavior of expressions bound to variables in scope at a hole via the \emph{live context inspector}, normally displayed  as a sidebar but shown disembodied in three different states in Fig.~\ref{fig:grades-sidebar}.
 In all three states, the live context inspector displays the names and types of the variables that are in scope, as is a common feature of programming environments. 
-
-New to our approach are the values associated with each binding.
-These values come from the environments associated with each hole instance in the result.
+New to our approach are the values associated with bindings, which come from the environment associated with the selected hole instance in the result.
 In this case, there are three instances of hole \li{1} in the result, numbered sequentially: \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages}.
-An environment is a partial mapping of each variable in scope to a value.
-We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical foundations detailed in Sec.~\ref{sec:calculus}).
-The programmer can select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the solid purple outline), or cycle rapidly through all of the closures for the hole at the cursor by clicking the arrows, e.g.
-$\blacktriangleright$, at the bottom of the context inspector, or by using hotkeys.
+% An environment is a partial mapping of variables in scope to values. 
+We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical connection detailed in Sec.~\ref{sec:calculus}). 
+The programmer can select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the purple outline), or cycle through all of the closures for the hole at the cursor by clicking the arrows, e.g.
+$\blacktriangleright$, at the bottom of the context inspector or using hotkeys.
 
@@
 
-The first example was purposely simple.
-Let us briefly consider a second more sophisticated example of these concepts: an incomplete implementation of quicksort as a recursive function, shown in Fig.~\ref{fig:qsort-cell-mockup}.
-Here, the programmer has selected a pivot, split the remainder of the list relative to this pivot and made the appropriate recursive calls.
-A hole appears in the return position of the function body as the programmer contemplates how to combine the results from the recursive calls.
+Let us now consider a second more sophisticated example: an incomplete implementation of the recursive quicksort function, shown in Fig.~\ref{fig:qsort-example-code}.
+So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid during class) has filled in the base case, and in the recursive case, has selected the head of the list as the pivot, partitioned the remainder of the list relative to the pivot and made the two recursive calls.
+A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables.
 
-The programmer has called \li{qsort} on an example list. However, 
-the indeterminate result in this case is rather uninteresting: it simply confirms that for this example, evaluation went through the recursive case of \li{qsort}, which ends with hole \li{1}.
-The live context inspector, on the other hand, provides a wealth of feedback about the dynamic behavior of this incomplete program.
-For example, the programmer can confirm that
-the lists \li{smaller} and \li{bigger} are appropriately named.
+At the bottom of the cell in Fig.~\ref{fig:qsort-example-code}, the programmer has applied \li{qsort} to an example list. However, 
+the indeterminate result of this function application appears rather uninteresting: it is simply an instance of hole \li{1}, which serves only to confirm that evaluation went through the recursive case of \li{qsort}.
+More interesting is the live context inspector, shown in three states in Fig.~\ref{fig:qsort-sidebars}, which provides feedback about the values of the variables in scope at hole \li{1} from the the various instances of hole \li{1} that appear in the result.
+For example, the initial state of the context inspector (Fig.~\ref{fig:qsort-sidebars}, left) shows the closure at the instance of hole \li{1} that appears immediately in the result due to the outermost application of \li{qsort}.
+From this, the programmer can confirm (or the lecturer can visually point out) that
+the lists \li{smaller} and \li{bigger} computed by the call to \li{partition} are appropriately named, and observe that they are not yet themselves sorted\todo{are there studies to cite here about how concrete examples help people in programming?
+math?}.
 
-The results from the recursive calls, \li{r_smaller} and \li{r_bigger}, are again hole instances, \li{1:2} and \li{1:3}.
-The programmer can click on these hole instances to reveal the environments at the corresponding recursive calls.
-For example, clicking on \li{1:2} reveals that the pivot in that case is \li{7} (not shown).
-In exploring these paths rooted at the result, the programmer can develop intuitions about the recursive structure of the computation before the program is complete.
-(We leave to future work the pedagogical question of whether such early feedback might help students with recursion.)
+The results from the subsequent recursive calls, \li{r_smaller} and \li{r_bigger}, are again hole instances, \li{1:2} and \li{1:3}. 
+The programmer can click on either of these hole instances to reveal the associated closures from the corresponding recursive calls. 
+For example, clicking on \li{1:3} reveals the hole closure from the \li{r_bigger} recursive call as shown in Fig.~\ref{fig:qsort-sidebars} (middle). 
+From there, the programmer can click another hole closure, e.g.
+\li{1:6} to reveal the hole closure from the subsequent \li{r_smaller} recursive call as shown in Fig.~\ref{fig:qsort-sidebars} (right).
+Notice that the path from the root to the selected hole closure is reported as shown at the bottom of the context inspector in Fig.~\ref{fig:qsort-sidebars}. 
+In exploring these paths rooted at the result, the programmer can develop concrete intuitions about the recursive structure of the computation (e.g.
+by following through to the base case as shown in Fig.~\ref{fig:qsort-sidebars}) even before the program is complete.
+% (We leave to future work the pedagogical question of whether such early feedback might help students understand recursion.)
 

### 2018-07-09 qsort-example.tex
missing files
updated example 2 to show three different hole closures and edits to 2.2

--- 2018-04-17 fig-grades-example.tex
@@
 \begin{figure}[t]
+\begin{subfigure}[t]{\textwidth}
+\centering
+\includegraphics[width=0.55\textwidth,interpolate=false,valign=t]{images/qsort-code.png}
+\vspace{-3px}
+\caption{The result of evaluating this incomplete quicksort function applied to a test value is a hole closure.}
+\label{fig:qsort-example-code}
+\end{subfigure}
 
-\includegraphics[width=1.01\textwidth,interpolate=false]{images/grades-cell-mockup.png}
+\vspace{10px}
 
+\begin{subfigure}[t]{\textwidth}
+\centering
+\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/qsort-new-sidebar-1.png}
+~$\xrightarrow[\text{click}]{}$
+\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/qsort-new-sidebar-2.png}
+~$\xrightarrow[\text{click}]{}$
+\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/qsort-new-sidebar-3.png}
+\caption{The programmer can explore the recursive structure of the computation by clicking on hole instances.}
+\label{fig:qsort-sidebars}
+\end{subfigure}
+
+\vspace{3px}
+
+\caption{Example 2: Incomplete Quicksort}
+\label{fig:qsort-cell-mockup}
+% \end{subfigure}
+\end{figure}
+
+% \begin{subfigure}[t]{\textwidth}
+% \centering
+% \includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-1.png}
+% ~${}^\blacktriangleright$
+% \includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-2.png}
+% ~${}^\blacktriangleright$
+% \includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-3.png}
+% \caption{Typing context view with live hole closure information}
+% \label{sec:grades-sidebar}
+% \end{subfigure}
 % %% TODO once the code above is removed, scale up the screenshots
@@
 % \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}
-
-\caption{Hazel mockups for Example 1a.}
-\label{fig:grades-example}
-\end{figure}

### 2018-07-10 examples-1.tex
updates to sec 2

@@
 \footnote{In a lazy language, like Haskell, the result would be much the same because the environment forces the result for printing.}
-While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.
+% While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.
 
@@
 weighted average between \li{0.0} and \li{100.0}, and so the correct
-constant is \li{0.30} rather than \li{30.0}. (There are of course other possible fixes.)
+constant is \li{0.30}, not \li{30.0}. %(There are of course other possible fixes.)
 
@@
 that the programmer has developed (in this case, regarding the behavior of \li{map}), and also help quickly dispel 
-inconsistencies between the programmer's expectations and the 
+misconceptions about the 
 actual behavior of the program (in this case, the magnitude of the arithmetic expressions being computed).
@@
 New to our approach are the values associated with bindings, which come from the environment associated with the selected hole instance in the result. 
+We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical connection detailed in Sec.~\ref{sec:calculus}). 
 In this case, there are three instances of hole \li{1} in the result, numbered sequentially: \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages}. 
 % An environment is a partial mapping of variables in scope to values. 
-We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical connection detailed in Sec.~\ref{sec:calculus}). 
 The programmer can select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the purple outline), or cycle through all of the closures for the hole at the cursor by clicking the arrows, e.g.
-$\blacktriangleright$, at the bottom of the context inspector or using hotkeys.
+$\blacktriangleright$, or using hotkeys.
 

### 2018-07-10 fig-grades-example.tex
updates to sec 2

@@
 \centering
-\includegraphics[width=0.85\textwidth,interpolate=false]{images/grades-cell-mockup.png}
+\includegraphics[width=0.75\textwidth,interpolate=false]{images/grades-cell-mockup.png}
+\vspace{-3px}
 \caption{Evaluating an incomplete functional program past the first hole}
@@
 \centering
-\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-1.png}
+\includegraphics[width=0.29\textwidth,interpolate=false]{images/grades-sidebar-1.png}
 ~${}^\blacktriangleright$
-\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-2.png}
+\includegraphics[width=0.29\textwidth,interpolate=false]{images/grades-sidebar-2.png}
 ~${}^\blacktriangleright$
-\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-3.png}
+\includegraphics[width=0.29\textwidth,interpolate=false]{images/grades-sidebar-3.png}
 \caption{The programming environment communicates relevant static \emph{and} dynamic information in a sidebar.}
@@
 % \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}
-\vspace{5px}
+\vspace{3px}
 
@@
 \label{fig:grades-example}
-\vspace{-8px}
+\vspace{-5px}
 \end{figure}

### 2018-07-10 examples-1.tex
added type inspector to fig2

@@
 So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid during class) has filled in the base case, and in the recursive case, has selected the head of the list as the pivot, partitioned the remainder of the list relative to the pivot and made the two recursive calls.
-A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables.
+A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables with an appropriate list expression, as indicated by the type inspector in Fig.~\ref{fig:qsort-type-inspector}.
 

### 2018-07-10 qsort-example.tex
updates to sec 2
added type inspector to fig2

@@
 \begin{figure}[t]
-\begin{subfigure}[t]{\textwidth}
+\begin{subfigure}[t]{0.65\textwidth}
 \centering
-\includegraphics[width=0.55\textwidth,interpolate=false,valign=t]{images/qsort-code.png}
-\vspace{-3px}
-\caption{The result of evaluating this incomplete quicksort function applied to a test value is a hole closure.}
+\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/qsort-code.png}
+\caption{The result of evaluation is a hole closure.}
 \label{fig:qsort-example-code}
+\end{subfigure}
+~
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
+\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/type-inspector.png}
+\vspace{-2px}
+\caption{The type inspector
+continuously provides static information about the expression at the cursor. Here, it tells us that the hole at the cursor should be filled by a list expression. Holes have hole (i.e. unknown) type.
+}
+\label{fig:qsort-type-inspector}
 \end{subfigure}
@@
 % \end{subfigure}
+
+\vspace{-6px}
 \end{figure}

### 2018-07-10 examples-1.tex
tweaks to the figures and text

--- 2018-07-10 examples-1.tex
@@
 So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid during class) has filled in the base case, and in the recursive case, has selected the head of the list as the pivot, partitioned the remainder of the list relative to the pivot and made the two recursive calls.
-A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables.
+A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables to fill the hole with an appropriate list expression, as indicated by the type inspector in Fig.~\ref{fig:qsort-type-inspector}.
 
 At the bottom of the cell in Fig.~\ref{fig:qsort-example-code}, the programmer has applied \li{qsort} to an example list. However, 
-the indeterminate result of this function application appears rather uninteresting: it is simply an instance of hole \li{1}, which serves only to confirm that evaluation went through the recursive case of \li{qsort}.
+the indeterminate result of this function application is simply an instance of hole \li{1}, which serves only to confirm that evaluation went through the recursive case of \li{qsort}. 
 More interesting is the live context inspector, shown in three states in Fig.~\ref{fig:qsort-sidebars}, which provides feedback about the values of the variables in scope at hole \li{1} from the the various instances of hole \li{1} that appear in the result.
-For example, the initial state of the context inspector (Fig.~\ref{fig:qsort-sidebars}, left) shows the closure at the instance of hole \li{1} that appears immediately in the result due to the outermost application of \li{qsort}.
+For example, in its initial state (Fig.~\ref{fig:qsort-sidebars}, left) it shows the closure at the instance of hole \li{1} that appears immediately in the result due to the outermost application of \li{qsort}.
 From this, the programmer can confirm (or the lecturer can visually point out) that
@@
 \li{1:6} to reveal the hole closure from the subsequent \li{r_smaller} recursive call as shown in Fig.~\ref{fig:qsort-sidebars} (right).
-Notice that the path from the root to the selected hole closure is reported as shown at the bottom of the context inspector in Fig.~\ref{fig:qsort-sidebars}. 
+Notice in each case that the path from the result to the selected hole closure is reported as shown at the bottom of the context inspector in Fig.~\ref{fig:qsort-sidebars}. 
 In exploring these paths rooted at the result, the programmer can develop concrete intuitions about the recursive structure of the computation (e.g.

### 2018-07-10 qsort-example.tex
tweaks to the figures and text

@@
 \begin{figure}[t]
-\begin{subfigure}[t]{0.65\textwidth}
+\begin{subfigure}[t]{0.70\textwidth}
 \centering
 \includegraphics[width=\textwidth,interpolate=false,valign=t]{images/qsort-code.png}
+\vspace{-3px}
 \caption{The result of evaluation is a hole closure.}
@@
 ~
-\begin{subfigure}[t]{0.33\textwidth}
+\begin{subfigure}[t]{0.29\textwidth}
 \centering
 \includegraphics[width=\textwidth,interpolate=false,valign=t]{images/type-inspector.png}
-\vspace{-2px}
-\caption{The type inspector
-continuously provides static information about the expression at the cursor. Here, it tells us that the hole at the cursor should be filled by a list expression. Holes have hole (i.e. unknown) type.
+\vspace{-3px}
+\caption{The type inspector provides static information about the term at the cursor.
+Here, it tells us that the hole should be filled by a list expression.
+Holes have hole (i.e.
+unknown) type, which  is universally consistent (see Sec.~\ref{sec:calculus} and \cite{popl-paper}).
 }
@@
 
-\vspace{10px}
+\vspace{8px}
 
@@
 
-\vspace{-6px}
+\vspace{-8px}
 \end{figure}

### 2018-07-12 related-work.tex
minor edits to related work

@@
 
-This paper defined a dynamic semantics for incomplete functional programs
-based directly on the static semantics developed by \citet{popl-paper}. That paper, as well as a subsequent ``vision paper'' introducing \Hazel, suggested as future work  
+This paper defined a dynamic semantics for incomplete functional programs,
+building on the static semantics developed by \citet{popl-paper}. That paper, as well as a subsequent ``vision paper'' introducing \Hazel, suggested as future work  
 a corresponding dynamic semantics for the purposes of live programming \cite{HazelnutSNAPL}. 
@@
 develops the conceptual and type-theoretic details necessary to achieve this broader vision.
+
+\parahead{Gradual Type Theory}
 
@@
 Many of the methods developed to make gradual typing more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein \cite{takikawa_et_al:LIPIcs:2015:5215}.
-For example, there has been substantial work on the problem of implementing casts efficiently, e.g.
-by \citet{herman2010space}, and on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.
+For example, there has been substantial work on the problem of implementing casts efficiently \cite{herman2010space,siek2010threesomes,garcia2013threesomes}, and on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.
 Another direction that we leave to future work is the integration of these ideas into dependently typed proof assistants,
@@
 
-Although our expectation is that \Hazel will remain a pure functional language \emph{a la} \Elm~or Haskell, another interesting future direction would be to move beyond
-pure functional programming and carefully integrate imperative features, e.g.
-ML-style references.
-These require particular care, as recognized by \citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15}, but the techniques are known and we see no reason why the type safety properties  established in this paper would not be conserved, with suitable modifications to account for a store.
+%% Although our expectation is that \Hazel will remain a pure functional language \emph{a la} \Elm~or Haskell,
+%
+Another interesting future direction would be to move beyond
+pure functional programming and carefully integrate imperative features, e.g. ML-style references.
+%
+%% These require particular care, as recognized by \citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15}, but the techniques are known and we see no reason why the type safety properties  established in this paper would not be conserved, with suitable modifications to account for a store.
+%
+\citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15} have shown how to incorporate such features into gradual type theory; we expect that the type safety properties established in this paper should also be conserved, with suitable modifications to account for a store.
+%
 Going beyond references to incorporate external effects, e.g.
@@
 We leave to future work the task of defining more restricted special cases of commutativity in effectful settings.
+
+\parahead{Contextual Modal Type Theory}
 
@@
 
-Although these connections with gradual typing and CMTT are encouraging, our contributions do not neatly fall out from this prior work.
+%% Although these connections with gradual typing and CMTT are encouraging, our contributions do not neatly fall out from this prior work.
+%
+These connections with gradual typing and CMTT are instructive, but our contributions do not neatly fall out from this prior work.
+%
 The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.
@@
 This contextual modal interpretation of live staged hole refinement
-nicely mirrors the modal interpretation of staged argument evaluation
+mirrors the modal interpretation of staged argument evaluation
 \cite{Davies:2001op}. 
@@
 CMTT is unique relative to prior approaches in that it has a clear Curry-Howard interpretation and that it was designed with commutativity properties in mind.
-This paper bears out that it is nicely suited for the purpose of live functional programming with typed expression holes.
+%
+This paper bears out that it is also well-suited for the purpose of live functional programming with typed expression holes.
 
@@
 
-% \parahead{Structure Editor Calculus}
+\parahead{Debuggers}
 
@@
 The ``fill-and-resume'' feature described in Sec.~\ref{sec:resumption} is a special case that comes with this stronger guarantee.
+
+\parahead{Structure Editors}
 
@@
 
+\parahead{Type Error Messages}
+
 A key feature of our semantics is that it permits evaluation not only of terms with empty holes, but also terms with non-empty holes, i.e. reified static type inconsistencies. 
@@
 Combining the strengths of these approaches may be fruitful in the future.
+
+\parahead{Program Synthesis and Slicing}
 

### 2018-07-12 abstract.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

@@
 of type errors. Consequently,    
-live feedback presented to the programmer has gaps, i.e. it flickers in and out or it goes stale whenever the program becomes incomplete. 
+live feedback presented to the programmer has temporal or perceptive gaps, i.e. it flickers in and out or it goes stale whenever the program becomes incomplete. 
 % In some cases, these gaps persist over a substantial length of time.
@@
 
-This paper confronts this \emph{gap problem} from type-theoretic first principles by developing 
+This paper confronts this ``{gap problem}'' from type-theoretic first principles by developing 
 \emph{a dynamic semantics for incomplete functional programs}, 
@@
 logical basis for tracking hole closures), combining these and developing additional machinery necessary 
-to continue evaluation past the hole forms in a manner that admits important metatheoretic properties.
+to continue evaluation past holes while maintaining important metatheoretic properties.
 

### 2018-07-12 examples-1.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

@@
 
-Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback available to the programmer at all.
-In response, the programmer might insert a temporary universal placeholder, e.g.
-by raising an exception as suggested in Sec.~\ref{sec:intro} (note however that Elm does not support exceptions, by design).
-This would cause typechecking to succeed.
-However, when the programming environment attempts to evaluate the program, 
+Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback.
+In response, the programmer might encode a hole by raising an exception. This would cause typechecking to succeed. 
+However, 
 evaluation would proceed only as far as the first \li{map} iteration, 
@@
 % Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
-Going further, \Hazel helps programmers reason about the dynamic behavior of expressions bound to variables in scope at a hole via the \emph{live context inspector}, normally displayed  as a sidebar but shown disembodied in three different states in Fig.~\ref{fig:grades-sidebar}.
-In all three states, the live context inspector displays the names and types of the variables that are in scope, as is a common feature of programming environments. 
+Going further, \Hazel helps programmers reason about the dynamic behavior of expressions bound to variables in scope at a hole via the \emph{live context inspector}, normally displayed  as a sidebar but shown disembodied in three states in Fig.~\ref{fig:grades-sidebar}.
+In all three states, the live context inspector displays the names and types of the variables that are in scope. 
 New to our approach are the values associated with bindings, which come from the environment associated with the selected hole instance in the result. 
 We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical connection detailed in Sec.~\ref{sec:calculus}). 
-In this case, there are three instances of hole \li{1} in the result, numbered sequentially: \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages}. 
+In this case, there are three instances of hole \li{1} in the result, numbered sequentially \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages} by \li{map}. 
 % An environment is a partial mapping of variables in scope to values. 
-The programmer can select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the purple outline), or cycle through all of the closures for the hole at the cursor by clicking the arrows, e.g.
+The programmer can directly select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the purple outline in Fig.~\ref{fig:grades-cell-mockup}), or cycle through all of the closures for the hole at the cursor by clicking the arrows at the bottom of the context inspector, e.g.
 $\blacktriangleright$, or using hotkeys.
@@
 Let us now consider a second more sophisticated example: an incomplete implementation of the recursive quicksort function, shown in Fig.~\ref{fig:qsort-example-code}.
-So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid during class) has filled in the base case, and in the recursive case, has selected the head of the list as the pivot, partitioned the remainder of the list relative to the pivot and made the two recursive calls.
-A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables to fill the hole with an appropriate list expression, as indicated by the type inspector in Fig.~\ref{fig:qsort-type-inspector}.
+So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid) has filled in the base case, and in the recursive case, partitioned the remainder of the list relative to the head, and made the two recursive calls.
+A hole appears in return position as the programmer contemplates how to fill the hole with an appropriate expression of list type, as indicated by the \emph{type inspector} in Fig.~\ref{fig:qsort-type-inspector}.
 
@@
 the indeterminate result of this function application is simply an instance of hole \li{1}, which serves only to confirm that evaluation went through the recursive case of \li{qsort}. 
-More interesting is the live context inspector, shown in three states in Fig.~\ref{fig:qsort-sidebars}, which provides feedback about the values of the variables in scope at hole \li{1} from the the various instances of hole \li{1} that appear in the result.
+More interesting is the live context inspector, shown in three states in Fig.~\ref{fig:qsort-sidebars}, which provides feedback about the values of the variables in scope at hole \li{1} from the the various instances of hole \li{1} that appear in the result, either immediately or within an outer closure.
 For example, in its initial state (Fig.~\ref{fig:qsort-sidebars}, left) it shows the closure at the instance of hole \li{1} that appears immediately in the result due to the outermost application of \li{qsort}.

### 2018-07-12 examples-2.tex
substantially expanded the related work
updates to sec 2
updated ill-typed quicksort example to use the same test as fig 2 and fixed some formatting errors
tweaks to the figures and text
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

@@
 \subsection{Example 3: Live Programming with Static Type Errors}
+\label{sec:static-errors}
 
@@
 \centering
-\includegraphics[width=0.7\textwidth,interpolate=false,valign=t]{images/qsort-type-error.png}
+\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/qsort-type-error-inset.png}
+% \includegraphics[width=0.28\textwidth,interpolate=false,valign=t]{images/type-inconsistency.png}
+\vspace{2px}
 \caption{Example 3: Ill-Typed Quicksort}
 \label{fig:qsort-type-error}
-\vspace{-8px}
+\vspace{-4px}
 \end{figure}
@@
 but assume that the programmer has filled in the previous hole
-as shown in Fig.~\ref{fig:qsort-type-error}. (In Sec.~\ref{sec:resumption}, we discuss
-how the programming environment might avoid restarting evaluation after such edits, but for small examples like this, restarting works fine.)
+as shown in Fig.~\ref{fig:qsort-type-error}. In Sec.~\ref{sec:resumption}, we discuss
+how the programming environment might avoid restarting evaluation after such edits, but for small examples like this, restarting works fine.
 
@@
 performs list concatenation, but the pivot is an integer. 
-Most compilers and editors will report a static error message, 
-of varying utility, to the programmer in this case (and \Hazel 
-follows suit, not shown). 
-Our argument is that this should not cause all feedback about 
-the dynamic behavior of the program to ``flicker out'' or ``go stale''. After all,
-the error is localized and there is perfectly good code elsewhere 
-in the program (if not nearby, then perhaps far away).
+Most compilers and editors will report a static error message
+to the programmer in this case, and \Hazel 
+follows suit in the type inspector (shown inset in Fig.~\ref{fig:qsort-type-error}). 
+However, our argument is that the presence of a static type error should not cause all feedback about 
+the dynamic behavior of the program to ``flicker out'' or ``go stale'' --
+after all, there are perfectly meaningful parts of the program (both nearby
+and far away from the error) 
+whose dynamic behavior may be of interest, or even assist the programmer in fixing the type error.
+% After all,
+% the error is localized and there is perfectly good code elsewhere 
+% in the program (if not nearby, then perhaps far away).
 
@@
 is to semantically internalize the ``red outline'' around
-type inconsistencies as a \emph{non-empty hole}.
-From the outside, a non-empty hole behaves statically much like an empty hole.
-Our approach takes the same approach for the dynamic semantics,
-i.e. evaluation safely proceeds past a non-empty hole just as if it were an empty hole.
-The semantics also associates an environment with each instance of a non-empty hole (not shown). 
+type inconsistencies, representing it as a \emph{non-empty hole}.
+Evaluation safely proceeds past a non-empty hole just as if it were an empty hole.
+The semantics also associates an environment with each instance of a non-empty hole,
+so we can use the live context inspector essentially as in Fig.~\ref{fig:qsort-sidebars} (not shown). 
 The difference is that evaluation also proceeds inside the hole, so that 
-feedback about an expression that might ``almost'' be correct is available. 
+feedback about the type-inconsistent expression, which might ``almost'' be correct, is available. 
 In this case, the result at the bottom of Fig.~\ref{fig:qsort-type-error}
 reveals that the programmer is on the right track: the list elements 
-(here, for a shorter example for the sake of exposition) are in the right order.
+appear in the correct order.
 They simply have not been combined correctly.

### 2018-07-12 examples-3.tex
substantially expanded the related work
tweaks to the figures and text
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

--- 2018-03-08 examples-3.tex
@@
-\subsection{Live Programming with Gradual Type Errors}
+% !TEX root = hazelnut-dynamics.tex
 
+\subsection{Example 4: Type Holes and Dynamic Type Errors}
+\label{sec:dynamic-type-errors}
+
+% \begin{subfigure}[t]{\textwidth}
+\begin{figure}
+\centering
+\includegraphics[width=0.73\textwidth,interpolate=false,valign=t]{images/cast-errors.png}
+% \includegraphics[width=0.28\textwidth,interpolate=false,valign=t]{images/type-inconsistency.png}
+\caption{Example 4: Type Holes and Dynamic Type Errors}
+\label{fig:cast-errors}
+\vspace{-6px}
+\end{figure}
+% \end{subfigure}
+
+% So far, we have only discussed incomplete programs where a hole appears within an expression.
+In \Hazel, the program can also be incomplete because holes appear in types. 
+\citet{popl-paper} confirmed that the literature on \emph{gradual type systems} \cite{Siek06a,DBLP:conf/snapl/SiekVCB15} is directly relevant to the problem of reasoning statically with type holes, by identifying the type hole with the unknown type.
+Unsurprisingly, then, it is also relevant to the problem of running 
+programs with type holes.
+Indeed, that is the purpose of gradual typing: to be able to run programs that are not yet sufficiently annotated with types by inserting \emph{casts} only where necessary.
+As such, let us consider only a small synthetic example to demonstrate what is unique to our approach.
+
+Fig.~\ref{fig:cast-errors} defines a simple function, \li{f}, of two arguments. 
+The type annotation on the first line leaves the type of those arguments unknown. 
+As such, the \Hazel type system, following the gradual typing approach,
+allows the body of the function to use those two
+arguments at any type at all (that is, the hole type is universally consistent). 
+In this case, the first argument, \li{simple}, is used at one type, \li{bool}, 
+and the second argument, \li{x}, is used at two different types in the two branches (perhaps because the programmer made a mistake), 
+first as an \li{int} and second as a \li{string}  (we use \li{^} for string concatenation).
+Although \Hazel supports only local type inference as of this writing, 
+a system that uses ML-style type reconstruction to fill type holes statically, like GHC Haskell, would only be able to fill the first hole. Leaving the second hole unfilled is 
+a parsimonious alternative to arbitrarily or heuristically\todo{cite error localization papers?}{} choosing one of the possibilities and marking the
+other uses of \li{x} as ill-typed.
+
+At the bottom of the cell in Fig.~\ref{fig:cast-errors}, we have three 
+example applications of \li{f}. In this case, we simply tuple the results, but we could also
+have made them into tests or put them into separate cells. All three are statically
+well-typed, again because the hole type is universally consistent.
+The result at the bottom of Fig.~\ref{fig:cast-errors} demonstrates that the first application
+of \li{f} is dynamically unproblematic. This might help the programmer confirm that at least
+the first branch of the function body has been implemented as intended without the need to comprehensively
+address all of the problems in the program. This flexibility is a common motivation for dynamic languages in programming folklore.
+
+The second application of \li{f}, in contrast, causes a dynamic type error because the second argument, \li{2}, is an \li{int} but evaluation takes the branch where \li{x} is used as a \li{string}. 
+Rather than aborting evaluation immediately when this occurs, as in existing gradual type systems, the problematic sub-term becomes a \emph{failed cast} term, shown shaded in red, which can be read ``\li{2} is an \li{int} that was used through a variable of hole type (\li{?}) as a \li{string}''.
+A failed cast behaves much like a non-empty hole in that it simply becomes an opaque value of the target type, \li{string}, that causes the surrounding concatenation operation to become indeterminate.
+Evaluation can continue on to the third application of \li{f}, which again is problematic, this time because the first argument is not a \li{bool} (perhaps because the programmer had an incorrect understanding of the argument order).
+Again, this causes a failed cast to appear, this time in guard position.
+Like a hole in guard position, evaluation cannot determine which branch to take so the whole guard expression becomes indeterminate.
+The pretty printer
+hides the two branches behind ellipses.
+
+In this synthetic example, it might have only been a small burden for the programmer to provide the intended types in the signature of \li{f}, but there are situations (e.g.
+during rapid prototyping or a live performance) where the programmer may consider the burden more substantial.
+This approach ensures that dynamic feedback does not have gaps even when there is a dynamic type error.
+
+\subsection{Design Variations}
+
+\begin{comment}
 \emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept

### 2018-07-12 examples.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

--- 2018-07-07 examples.tex
@@
 
-This section gives an example-driven overview of our  approach as implemented in  \Hazel, a browser-based lab notebook environment being developed by \cite{HazelnutSNAPL}.
-The \Hazel user interface is  based roughly on IPython/Jupyter \cite{PER-GRA:2007} and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}), a popular pure functional programming language similar to ``core ML'' \cite{czaplicki2012elm,Elm}, with which we assume basic  familiarity.
+This section gives an example-driven overview of our  approach as implemented in  \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL} (initially for use by students and instructors in introductory functional programming courses).
+The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with code organized into cells and results appearing below each cell, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity.
 For the sake of 
-exposition, we have ``post-processed'' the screenshots in this section to make use of a few language conveniences from \Elm~that were not available in \Hazel as of this writing, e.g.
+exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of some ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel as of this writing, e.g.
 pattern matching in function arguments, list  notation, and record labels (currently there are only tuples).
-These conveniences are entirely orthogonal to the contributions of this paper: all of the live programming features demonstrated in this section have been implemented, both in \Hazel itself and in the simplified companion implementation of the core calculus in the supplement (see Sec.~\ref{sec:implementation}).
+These conveniences are orthogonal to the contributions of this paper.
+All of the live programming features demonstrated in the figures in this section have been implemented (see Sec.~\ref{sec:implementation}).
 

### 2018-07-12 fig-grades-example.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

--- 2018-04-17 fig-grades-example.tex
@@
 \begin{figure}[t]
+\begin{subfigure}[t]{\textwidth}
+\centering
+\includegraphics[width=0.8\textwidth,interpolate=false]{images/grades-cell-mockup.png}
+\vspace{-3px}
+\caption{Evaluating an incomplete functional program past the first hole}
+\label{fig:grades-cell-mockup}
+\end{subfigure}
 
-\includegraphics[width=1.01\textwidth,interpolate=false]{images/grades-cell-mockup.png}
+\vspace{10px}
 
+\begin{subfigure}[t]{\textwidth}
+\centering
+\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-1.png}
+~${}^\blacktriangleright$
+\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-2.png}
+~${}^\blacktriangleright$
+\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-3.png}
+\caption{The live context inspector communicates relevant static \emph{and} dynamic information about variables in scope.}
+\label{fig:grades-sidebar}
+\end{subfigure}
 % %% TODO once the code above is removed, scale up the screenshots
@@
 % \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}
+\vspace{3px}
 
-\caption{Hazel mockups for Example 1a.}
+\caption{Example 1: Grades}
 \label{fig:grades-example}
+\vspace{-5px}
 \end{figure}

### 2018-07-12 hazelnut-dynamics.tex
minor formatting for arxiv
fixed page numbers
arxiv version
fixed typos identified by OOPSLA reviews
tightened abstract
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

@@
 %% For double-blind review submission
-\documentclass[acmsmall,review]{acmart}
+%\documentclass[acmsmall,review]{acmart}
+%%% Note: arxiv does not want line numbers (they are detected somehow, and are not allowed)
+\documentclass[acmsmall,review,anonymous,nonacm]{acmart}
 \settopmatter{printfolios=false,printccs=false,printacmref=false}
-\setcopyright{none}
-\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
+% \settopmatter{printccs=false,printacmref=false}
 
@@
 \usepackage{wrapfig}
+% \usepackage[subtle]{savetrees}
 
@@
 numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
-morekeywords={string, float, int},
+morekeywords={string, float, int, bool},
 classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,
@@
 
-\fancyfoot{} % suppresses the footer (also need \thispagestyle{empty} after \maketitle below)
+% \fancyfoot{} % suppresses the footer (also need \thispagestyle{empty} after \maketitle below)
 
@@
 \maketitle
-\thispagestyle{empty} % suppresses the footer
+% \thispagestyle{empty} % suppresses the footer
 
@@
 
-\clearpage
+%\clearpage
 \bibliography{references,all.short,hazel_NSF}

### 2018-07-12 intro.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

@@
 %}
-In recognition of this \emph{gap problem}---that incomplete programs are formally meaningless---\citet{popl-paper} develop a static semantics (i.e. a type system) for incomplete 
+In recognition of this ``{gap problem}''---that incomplete programs are formally meaningless---\citet{popl-paper} develop a static semantics (i.e. a type system) for incomplete 
 functional programs, modeling them formally as typed expressions with \emph{holes} in 
@@
 } 
-Although better than nothing, this exceptional approach to expression holes has limitations 
+Although better than nothing, this ``exceptional approach'' to expression holes has limitations 
 within a live programming environment because 
@@
 
-This paper develops a theoretically well-grounded dynamic semantics for incomplete functional programs, starting from the static semantics developed by \citet{popl-paper},  that addresses the limitations of the ``exceptional approach'' just enumerated.
+This paper develops a theoretically well-grounded dynamic semantics for incomplete functional programs, starting from the static semantics developed by \citet{popl-paper},  that addresses these limitations of the exceptional approach.
 In particular, rather than stopping with an exception when evaluation encounters an expression hole, evaluation continues ``around'' the hole, performing as much of the remaining computation as possible.
 For programs with unfillable type holes, casts are inserted and evaluation also proceeds past failed casts in much the same way. 
-The system tracks the closure around each hole instance as evaluation proceeds so that 
+The system tracks the closure around each expression hole instance as evaluation proceeds so that 
 the live programming environment can feed relevant information from the {hole closures} to the programmer as they work to fill the holes in the program.
@@
 
-Sec.~\ref{sec:calculus} then abstracts away the inessential details of the language and user interface and makes the  intuitions developed in Sec.~\ref{sec:examples} formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions and tracking hole closures as just outlined.
+Sec.~\ref{sec:calculus} then abstracts away the inessential details of the language and user interface and makes the  intuitions developed in Sec.~\ref{sec:examples} formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions and tracking hole closures.
 The dynamic semantics of \HazelnutLive is a small-step reduction semantics equipped with clean type safety theorems (including, notably, a Progress theorem for incomplete expressions).
@@
 every editor state has some, possibly incomplete, type. 
-The type safety theorems established earlier in Sec.~\ref{sec:calculus} then guarantee that every editor state has dynamic meaning.
-Taken together, the result is a full solution to the gap problem, i.e.
-a proof-of-concept
+The type safety theorems then guarantee that every editor state has dynamic meaning. Taken together, the result is a full solution to the gap problem, i.e. a proof-of-concept
 live functional programming environment where the static and dynamic feedback never exhibits gaps. We formally state this continuity corrollary.
@@
 
-Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneously discusses directions for future work. Sec.~\ref{sec:discussion} briefly concludes.
+Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneously discusses directions for future work and Sec.~\ref{sec:discussion} concludes.
 

### 2018-07-12 discussion.tex
fixed typos identified by OOPSLA reviews
tightened abstract
a bit of concluding text around Weinberg quote

@@
 \label{sec:discussion}
+
+\begin{comment}
 To conclude, we quote Weinberg from The Psychology of Computer Programming (1998): ``how truly sad it is that just at the very moment
@@
 speaking gibberish.''
+\end{comment}
 
+\begin{quote}
+``[H]ow truly sad it is that just at the very moment
+when the computer has something important to tell us, it starts
+speaking gibberish.''
+
+\vspace{3pt}
+
+\hfill{}---Gerald Weinberg, The Psychology of Computer Programming (1998)
+\end{quote}
+
+\vspace{3pt}
+
+\noindent
+%
+Weinberg's sentiment applies to countless situations that programmers routinely face---deciphering static error messages, deciphering run-time error messages, mentally simulating large parts of the evaluation process, and so on.
+%
+Efforts to remediate this state of affairs have been increasingly common among programming languages, software engineering, and human-computer interaction researchers.
+%
+Yet essentially all of this work towards next-generation programming environments assumes that the programmer has already written a complete program.
+%
+It is our hope that, by providing a foundation for how to reason about and run incomplete programs, future live programming environments can provide meaningful feedback to users, without temporal and perceptive gaps, throughout the program editing process.
+
+\begin{comment}
 \newcommand{\acksSection}{Acknowledgments}
@@
 %
+\end{comment}
 
+\clearpage
 % \todo{discussion}

### 2018-07-12 abstract.tex
final POPL19 submission

@@
 \emph{a dynamic semantics for incomplete functional programs}, 
-starting from the static semantics for incomplete functional programs
+starting from the static semantics recently 
 developed by \citet{popl-paper}. 
@@
 %
-Rather than aborting with an exception when evaluation encounters any of these holes as in
-several existing systems, evaluation proceeds around the holes,
-performing as much of the remaining computation as is possible and tracking the 
-closure around each hole instance as it flows through the program. Editor services can report information from these hole closures 
+Rather than aborting when evaluation encounters any of these holes as in
+several existing systems, evaluation proceeds around holes,
+tracking the 
+closure around each hole instance as it flows through the remainder of the program. Editor services can report information from these hole closures 
 to help the programmer decide how to fill the holes in the program. 
-Hole closures also enable a \emph{fill-and-resume} feature that 
+Hole closures also enable a \emph{fill-and-resume} operation that 
 avoids the need to restart evaluation after edits that amount to hole filling. 
@@
 type holes) and contextual modal type theory (which provides a
-logical basis for tracking hole closures), combining these and developing additional machinery necessary 
+logical basis for hole closures), combining these and developing additional machinery necessary 
 to continue evaluation past holes while maintaining important metatheoretic properties.
@@
 %
-We have mechanized this core calculus, called \HazelnutLive{}, using the Agda proof assistant.
+We have mechanized the metatheory of this core calculus, called \HazelnutLive{}, using the Agda proof assistant.
 %
-We also describe a companion implementation 
-that inserts holes automatically to guarantee that every editor state has some (possibly incomplete)
-type, based on the \Hazelnut edit action calculus of \citet{popl-paper}.
+We have also implemented these ideas into the \Hazel programming environment. The implementation inserts holes automatically to guarantee that every editor state has some (possibly incomplete)
+type, following the \Hazelnut edit action calculus of \citet{popl-paper}.
 Taken together with this paper's type safety property, the 

### 2018-07-12 calculus.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4
final POPL19 submission

@@
 % !TEX root = hazelnut-dynamics.tex
-% \clearpage
-\newcommand{\calculusSec}{Hazelnut Live, Formally}
+\newcommand{\calculusSec}{Hazelnut Live}
 \section{\protect\calculusSec}
@@
 (see Sec.~\ref{sec:intro} for discussion of implicit, manual, semi-automated and fully automated hole entry methods).
-Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
+Each well-typed external expression (see Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
 We take this approach, notably also taken in the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and
@@
 \rulename{SVar}, correspond to variable names used in the mechanization.
-We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).
+The \Hazel implementation substantially follows the formal specification of \Hazelnut (for the editor component) and \HazelnutLive (for the semantics), and we can formally state a continuity invariant for a putative combined calculus (see Sec.~\ref{sec:implementation}).
 
@@
 That paper also defined sum types as an extension to the core calculus.
-We follow suit on both counts in Appendix~\ref{sec:extensions}.
+We follow suit on both counts in the supplemental appendix.%Appendix~\ref{sec:extensions}.
 %
@@
 The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.
+
+We do not formally need an explicit fixpoint operator because this calculus supports general recursion due to type holes, e.g.
+we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.
+More generally, the untyped lambda calculus can be embedded as described by \citet{Siek06a}.
+%As such, we omit an explicit fixpoint operator for concision.\todo{fixpoint?}{}
 
@@
 
-Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see Appendix~\ref{sec:additional-defns}).
+Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the supplemental appendix for the full definition).
+% Appendix~\ref{sec:additional-defns}).
 %
@@
 The need for this case is perhaps not obvious because in the cast calculus,  the only irreducible term of type $\tehole$
-is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.
+is $\dcasttwo{\dexp}{\htau_1}{\tehole}$, but here there are holes as well.
 
@@
 transitions and a standard substitution lemma stated in
-Appendix~\ref{sec:additional-defns}.
+the supplemental appendix.
+% Appendix~\ref{sec:additional-defns}.
 %
@@
 %
-The proof relies on canonical forms lemmas stated in
-Appendix~\ref{sec:additional-defns}.
+The proof relies on canonical forms lemmas stated in the supplemental appendix.
+% Appendix~\ref{sec:additional-defns}.
 
@@
 %
-Appendix~\ref{sec:additional-defns} defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
+% Appendix~\ref{sec:additional-defns}
+The supplemental appendix defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
 %
@@
 
-% TODO Delta is always empty too!
-\begin{thm}[Complete Expansion] ~
-  \begin{enumerate}[nolistsep]
-    \item
+\begin{thm}[Complete Expansion] 
+% ~
+  % \begin{enumerate}[nolistsep]
       If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
@@
       then $\isComplete{\htau}$ and $\isComplete{\dexp}$ and $\Delta = \emptyset$.
-    \item
-      If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
-and $\isComplete{\htau}$
-      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
-      then $\isComplete{\dexp}$ and $\isComplete{\htau'}$ and $\Delta=\emptyset$
-  \end{enumerate}
+%     \item
+%       If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
+% and $\isComplete{\htau}$
+%       and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+%       then $\isComplete{\dexp}$ and $\isComplete{\htau'}$ and $\Delta=\emptyset$
+  % \end{enumerate}
 \end{thm}
@@
 
+\vspace{-3px}
 \subsection{Agda Mechanization}
 \label{sec:agda-mechanization}
-\vspace{-3px}
+\vspace{-2px}
 
@@
 
-\subsection{Implementation}\label{sec:implementation}
-
-The supplemental material includes a browser-based implementation
-of \HazelnutLive.
-% The implementation supports the core calculus of this section but with the notional base type $b$ replaced by the $\tnum$ type from Appendix~\ref{sec:extensions}.
+\vspace{-3px}
+\subsection{Implementation and Continuity}\label{sec:implementation}
+\vspace{-2px}
+
+The \Hazel implementation demonstrated in Sec.~\ref{sec:examples} 
+includes an unoptimized interpreter, written in OCaml, that implements the semantics as described
+in this section, albeit for an \Elm-like language that goes somewhat beyond the simple
+calculus here. As with many full-scale systems, there is not currently a formal
+specification for the full \Elm{} or \Hazel language, but the supplemental appendix
+discusses how the standard approach for deriving a ``gradualized'' version of a 
+language construct provides most of the necessary scaffolding \cite{DBLP:conf/popl/CiminiS16}, and provides some examples.
+
+% The supplemental material includes a browser-based implementation
+% of \HazelnutLive.
+% % The implementation supports the core calculus of this section but with the notional base type $b$ replaced by the $\tnum$ type from Appendix~\ref{sec:extensions}.
 The implementation also includes the sum types extension from Appendix~\ref{sec:extensions} and a few minor  conveniences, e.g.
 \li{let} binding.
-All of the live programming features from \Secref{sec:examples} are available in the implementation essentially as shown, but for this more austere language (with some minor conveniences, notably let binding).
+% All of the live programming features from \Secref{sec:examples} are available in the implementation essentially as shown, but for this more austere language (with some minor conveniences, notably let binding).
 Appendix~\ref{sec:impl-screenshots} provides screenshots of the full user interface.
 
- % which is a functional reactive program \cite{Elliott:1997jh,DBLP:conf/pldi/CzaplickiC13}
-%
-% Functional Reactive Animation (Hudak 1997) is the canonical FRP cite, not Elm, if you want one; we don't need one, I think. --Matt
-%
-%\cite{DBLP:conf/pldi/CzaplickiC13}
-
-The implementation is written with the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml}
-together with the OCaml \lismall{React} library \cite{OcamlReact}
-and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}.
+%  % which is a functional reactive program \cite{Elliott:1997jh,DBLP:conf/pldi/CzaplickiC13}
+% %
+% % Functional Reactive Animation (Hudak 1997) is the canonical FRP cite, not Elm, if you want one; we don't need one, I think. --Matt
+% %
+% %\cite{DBLP:conf/pldi/CzaplickiC13}
+
+% The implementation is written with the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml}
+% together with the OCaml \lismall{React} library \cite{OcamlReact}
+% and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}.
 This follows the implementation of \Hazel, which, although tracking toward an Elm-like semantics, is implemented in OCaml because the Elm compiler is not yet self-hosted.
@@
 
-The editor component of the implementation is derived
-from the structure editor calculus of~\Hazelnut, but with support for more natural cursor-based movement and infix operator sequences (the details are beyond the scope of this paper).
+The editor component of the \Hazel implementation is derived
+from the structure editor calculus of~\Hazelnut, but with support for more natural cursor-based movement and infix operator sequences (the details of which are beyond the scope of this paper).
 It exposes a language of structured
-edit actions (summarized in the sidebar on the left of
-the screenshots in Appendix~\ref{sec:impl-screenshots}) that automatically inserts holes as necessary
-to guarantee that every edit state has some (possibly incomplete) type
-. This corresponds to the Sensibility invariant established by \citet{popl-paper}.
-By composing this metatheoretic property with
-the \Property{Expandability}, \Property{Typed Expansion}
-and \Property{Progress} properties from this paper, we establish a
-uniquely powerful invariant: \emph{every} edit state has a \emph{static
-meaning} (a la~\Hazelnut) and a \emph{dynamic meaning} (a la~\HazelnutLive).
-%
-In other words, live feedback never ``flickers in and out'' nor reflects ``stale'' edit states.
-%
-The implementation is intended as a proof-of-concept of this invariant, and as a companion to this paper.
-
+edit actions that automatically insert empty and non-empty holes as necessary
+to guarantee that every edit state has some (possibly incomplete) type.
+This corresponds to the top-level Sensibility invariant established for the \Hazelnut calculus by \citet{popl-paper}, reproduced below:
+\begin{prop}[Sensibility]
+  \label{thrm:actsafe} 
+  If $\hsyn{\Gamma}{\removeSel{\zexp}}{\htau}$ and
+    $\performSyn{\Gamma}{\zexp}{\htau}{\alpha}{\zexp'}{\tau'}$ then
+    $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
+  % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+  %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
+  %   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
+\end{prop}
+\noindent 
+Here, $\zexp$ is an editor state (an expression with a cursor), and $\removeSel{\zexp}$ drops the cursor, producing an expression ($e$ in this paper).
+So in words, ``if, ignoring the cursor, the editor state, $\removeSel{\zexp}$, initially has type $\htau$ and we perform an edit action $\alpha$ on it, then the resulting editor state, $\removeSel{\zexp'}$, will have type $\htau'$''.
+
+By composing this Sensibility property with
+the \Property{Expandability}, \Property{Typed Expansion}, \Property{Progress} and \Property{Preservation} properties from this section, we establish a
+uniquely powerful Continuity invariant:
+\begin{corol}[Continuity]
+  \label{thrm:actsafe} 
+  If $\hsyn{\emptyset}{\removeSel{\zexp}}{\htau}$ and
+    $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\tau'}$ then
+    $\expandSyn{\emptyset}{\removeSel{\zexp'}}{\htau'}{\dexp}{\Delta}$
+      for some $\dexp$ and $\Delta$ such that 
+$\hasType{\Delta}{\emptyset}{\dexp}{\htau'}$
+and either
+  (a) $\stepsToD{}{\dexp}{\dexp'}$ for some $d'$ such that $\hasType{\Delta}{\emptyset}{\dexp'}{\htau'}$; or
+  (b) $\isBoxedValue{\dexp}$ or
+  (c) $\isIndet{\dexp}$.
+  % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+  %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
+  %   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
+\end{corol}
+
+This addresses the gap problem: \emph{every} editor state has a {static
+meaning} (so editor services like the type inspector from Fig.~\ref{fig:qsort-type-inspector} are always available) and a non-trivial {dynamic meaning} (a result is always available, evaluation does not stop when a hole or cast failure is encountered, and editor services that rely on hole closures, like the live context inspector from Fig.~\ref{fig:grades-sidebar}, are always available).
+
+In settings where the editor does not maintain this Sensibility invariant, our approach still helps to reduce the severity of the gap problem, i.e.
+\emph{more} editor states are meaningful, even if not \emph{all} are.% We can formally state this end-to-end continuity corollary as follows:
+%
 % We make no further claims about the usability or practicality of the implementation (and indeed, there remain many important and decidedly unresolved questions along these lines, which we leave beyond the scope of this paper).
@@
 in data science tasks, a more efficient approach is to resume evaluation from where it left off after an edit that amounts to hole filling.
-This section briefly discusses the foundations of this feature, which we call \emph{fill-and-resume}.
-This approach is complementary to incremental computing techniques~\cite{Hammer2014}.
+This section develops a foundational account of this feature, which we call \emph{fill-and-resume}.
+This approach is complementary to, but distinct from, incremental computing (which is concerned with changes in input, not code insertions)~\cite{Hammer2014}.
 
@@
 %
-the key idea is to interpret hole environments as \emph{delayed substitutions}. This is the same interpretation suggested for closures in contextual modal
+the key idea is to interpret hole environments as \emph{delayed substitutions}. This is the same interpretation suggested for metavariable closures in contextual modal
 type theory (CMTT) by \citet{Nanevski2008}.
@@
 We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
-transitive closure of stepping (see \Figref{fig:multi-step} in
-Appendix~\ref{sec:additional-defns}).
+transitive closure of stepping (see the supplemental appendix).
+% Appendix~\ref{sec:additional-defns}).
 %
@@
 (There are various standard ways to formalize this intuition.
-For the sake of space, we review these in Appendix~\ref{sec:confluence}.)
+For the sake of space, we review these in the supplemental appendix.)
+% Appendix~\ref{sec:confluence}.)
 
@@
 
-We describe the proof, which involves a
-number of lemmas and definitions, in Appendix~\ref{sec:hole-filling}.
+We describe the proof, which is straightforward but involves a
+number of lemmas and definitions, in the supplemental appendix.
+%Appendix~\ref{sec:hole-filling}.
 %

### 2018-07-12 discussion.tex
final POPL19 submission

--- 2018-05-01 discussion.tex
@@
 speaking gibberish.''
+\end{comment}
+\vspace{3pt}
+\begin{quote}
+\textit{``[H]ow truly sad it is that just at the very moment
+when the computer has something important to tell us, it starts
+speaking gibberish.''}
 
+\vspace{3pt}
+
+\hfill{}--- Gerald Weinberg, The Psychology of Computer Programming (1971)
+\end{quote}
+\vspace{3pt}
+
+\noindent
+%
+Weinberg's sentiment applies to countless situations that programmers face as they work to develop and critically investigate the mental model of the program they are writing---at the very moment where rich feedback might be most helpful, e.g.
+when there is an error in the program or when the programmer is unsure how to fill a hole, the computer often has comparatively little feedback to offer (perhaps just a parser error message, or an explanation of one type error).
+%
+% Efforts to develop tools that help programmers explore the behavior of parts of their program interactively, i.e.
+live programming tools, have been increasingly common in the literature on programming languages, software engineering, and human-computer interaction.
+%
+% Yet much of this work has assumed that the programmer has already written a formally complete program, or relies on \emph{ad hoc} heuristics to handle some incomplete states. 
+%
+It is our hope that the well-behaved  type-theoretic foundations developed here will enable not only \Hazel but live programming tools of a wide variety of other designs to further narrow the temporal and perceptive gap and provide meaningful feedback to programmers at these important   moments.
+
+\begin{comment}
 \newcommand{\acksSection}{Acknowledgments}

### 2018-07-12 examples-1.tex
final POPL19 submission

@@
 From this, the programmer can confirm (or the lecturer can visually point out) that
-the lists \li{smaller} and \li{bigger} computed by the call to \li{partition} are appropriately named, and observe that they are not yet themselves sorted\todo{are there studies to cite here about how concrete examples help people in programming?
-math?}.
+the lists \li{smaller} and \li{bigger} computed by the call to \li{partition} are appropriately named, and observe that they are not yet themselves sorted.
 

### 2018-07-12 examples-2.tex
final POPL19 submission

@@
 and far away from the error) 
-whose dynamic behavior may be of interest, or even assist the programmer in fixing the type error.
+whose dynamic behavior may be of interest. Evaluation can also assist the programmer in understanding the type error by presenting concrete values \cite{Seidel2016}.
 % After all,
@@
 so we can use the live context inspector essentially as in Fig.~\ref{fig:qsort-sidebars} (not shown). 
-The difference is that evaluation also proceeds inside the hole, so that 
+Evaluation proceeds inside the hole, so that 
 feedback about the type-inconsistent expression, which might ``almost'' be correct, is available. 

### 2018-07-12 examples-3.tex
final POPL19 submission

--- 2018-03-08 examples-3.tex
@@
-\subsection{Live Programming with Gradual Type Errors}
+% !TEX root = hazelnut-dynamics.tex
 
+\subsection{Example 4: Type Holes and Dynamic Type Errors}
+\label{sec:dynamic-type-errors}
+
+% \begin{subfigure}[t]{\textwidth}
+\begin{figure}
+\centering
+\includegraphics[width=0.71\textwidth,interpolate=false,valign=t]{images/cast-errors.png}
+% \includegraphics[width=0.28\textwidth,interpolate=false,valign=t]{images/type-inconsistency.png}
+\caption{Example 4: Type Holes and Dynamic Type Errors}
+\label{fig:cast-errors}
+\vspace{-6px}
+\end{figure}
+% \end{subfigure}
+
+% So far, we have only discussed incomplete programs where a hole appears within an expression.
+In \Hazel, the program can also be incomplete because holes appear in types. 
+\citet{popl-paper} confirmed that the literature on \emph{gradual type systems} \cite{Siek06a,DBLP:conf/snapl/SiekVCB15} is directly relevant to the problem of reasoning with type holes, by identifying the type hole with the unknown type.
+% Unsurprisingly, then, it is also relevant to the problem of running 
+% programs with type holes. 
+Indeed, the purpose of gradual typing is to be able to run programs that are not yet sufficiently annotated with types by inserting \emph{casts} only where necessary.
+As such, let us consider only a small synthetic example to demonstrate what is unique to our approach.
+
+Fig.~\ref{fig:cast-errors} defines a simple function, \li{f}, of two arguments. 
+The type annotation on the first line leaves the type of those arguments unknown. 
+As such, the \Hazel type system, following the gradual typing approach,
+allows the body of the function to use those two
+arguments at any type (that is, the hole type is universally consistent). 
+Here, the first argument, \li{simple}, is used at one type, \li{bool}, 
+and the second argument, \li{x}, is used at two different types in the two branches (perhaps because the programmer made a mistake), 
+ \li{int} and  \li{string}  ( \li{^} is string concatenation).
+Although \Hazel supports only local type inference as of this writing, 
+a system that uses ML-style type reconstruction to fill type holes statically, like GHC Haskell, would only be able to fill the first hole. Leaving the second hole unfilled is 
+a parsimonious alternative to arbitrarily or heuristically choosing one of the possibilities and marking the
+other uses of \li{x} as ill-typed (see \cite{DBLP:journals/jfp/ChenE18}).
+
+At the bottom of the cell in Fig.~\ref{fig:cast-errors}, we have three 
+example applications of \li{f}, tupled together for concision. All three are statically
+well-typed, again because the hole type is universally consistent. 
+The result at the bottom of Fig.~\ref{fig:cast-errors} demonstrates that the first application
+of \li{f} is dynamically unproblematic. This allows the programmer confirm that 
+the first branch of the conditional operates as intended without the need to 
+address the typing problems in the other branch. 
+This flexibility is a common motivation for dynamic and gradual languages \cite{Bayne:2011:ASD:1985793.1985864}.
+
+The second application of \li{f}, in contrast, causes a dynamic type error because the second argument, \li{2}, is an \li{int} but evaluation takes the branch where it is used as a \li{string}. 
+Rather than aborting evaluation when this occurs, as in existing gradual type systems, the problematic term becomes a \emph{failed cast} term, shown shaded in red, which can be read ``\li{2} is an \li{int} that was used through a variable of hole type (\li{?}) as a \li{string}''.
+A failed cast acts much like a non-empty hole as a membrane around a problematic term. The surrounding concatenation operation becomes indeterminate, but  
+evaluation can continue on to the third application of \li{f}, which is also problematic, this time because the first argument is not a \li{bool} (perhaps because the programmer had an incorrect understanding of the argument order).
+Again, this causes a failed cast to appear, this time in guard position.
+Like a hole in guard position, evaluation cannot determine which branch to take so the whole conditional becomes indeterminate.
+The pretty printer
+hides the two branches behind ellipses for concision.
+
+In this small example, it might have only been a small burden for the programmer to provide the intended types in the signature for \li{f}, but there are situations (e.g.
+during rapid prototyping or a live performance) where the programmer might consider the burden more substantial.
+This approach ensures that dynamic feedback does not exhibit gaps even when there is a dynamic type error.
+
+\begin{comment}
 \emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept

### 2018-07-12 examples.tex
final POPL19 submission

@@
 
-This section gives an example-driven overview of our  approach as implemented in  \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL} (initially for use by students and instructors in introductory functional programming courses).
-The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with code organized into cells and results appearing below each cell, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity.
+This section gives an example-driven overview of our  approach as implemented in  \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL}.
+The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with results appearing below cells containing code, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity.
+\Hazel is intended initially for use by students and instructors in introductory functional programming courses (where \Elm~ has been successful \cite{DBLP:journals/corr/abs-1805-05125}).
 For the sake of 
@@
 pattern matching in function arguments, list  notation, and record labels (currently there are only tuples).
-These conveniences are orthogonal to the contributions of this paper.
-All of the live programming features demonstrated in the figures in this section have been implemented (see Sec.~\ref{sec:implementation}).
+These conveniences are orthogonal to the contributions of this paper; all of the live programming and user interface features demonstrated in the figures in this section have been implemented.
 

### 2018-07-12 intro.tex
final POPL19 submission

@@
 Live programming environments aim to support this workflow by interleaving editing and evaluation so as to   
-narrow what \citet{burckhardt2013s} call ``the temporal and perceptive gap'' between these activities.
+narrow what \citet{burckhardt2013s} call the ``temporal and perceptive gap'' between these activities.
 % In other words, the goal \IS to provide continuous feedback about the dynamic behavior of the program, in whole or in part,
@@
 
-This paper develops a theoretically well-grounded dynamic semantics for incomplete functional programs, starting from the static semantics developed by \citet{popl-paper},  that addresses these limitations of the exceptional approach.
-In particular, rather than stopping with an exception when evaluation encounters an expression hole, evaluation continues ``around'' the hole, performing as much of the remaining computation as possible.
-For programs with unfillable type holes, casts are inserted and evaluation also proceeds past failed casts in much the same way. 
+This paper develops a dynamic semantics for incomplete functional programs, starting from the static semantics developed by \citet{popl-paper},  that addresses these limitations of the exceptional approach.
+In particular, rather than stopping when evaluation encounters an expression hole, evaluation continues ``around'' the hole so that dynamic feedback about other parts of the program remains available.
+For programs with unfillable type holes, casts are inserted as in gradual type theory \cite{DBLP:conf/snapl/SiekVCB15}. Evaluation proceeds around failed casts in the same way.
+
 The system tracks the closure around each expression hole instance as evaluation proceeds so that 
@@
 \emph{indeterminate}, evaluation state.
-We call this feature \emph{fill-and-resume}.
+We call this operation \emph{fill-and-resume}.
+We observe that empty hole closures are closely related to metavariable closures from contextual modal type theory~(CMTT)~\cite{Nanevski2008} (which, by its Curry-Howard interpretation, provides a logical basis for reasoning about and operating on hole closures).
+These connections to well-established systems, together with our mechanized proofs, serve to support our claim that this approach is theoretically well-grounded.
+
 We are integrating this approach into the \Hazel programming environment being developed by \citet{HazelnutSNAPL}.
+\Hazel features a live context inspector that integrates static information with hole closure information and interactively presents recursive hole closures, which arise from incomplete recursive functions.
+The editor component of this implementation defines a language of structured edit actions,
+based on the \Hazelnut structure editor calculus developed by \citet{popl-paper}, that inserts holes automatically to guarantee that
+every editor state has some, possibly incomplete, type. 
+The type safety invariant that we establish then guarantees that every editor state has dynamic meaning. Taken together, the result is a full solution to the gap problem, i.e. a proof-of-concept
+live functional programming environment that provides rich static and dynamic feedback that never suffers from temporal or perceptive gaps.
 
+\vspace{-2px}
 \parahead{Paper Outline}
@@
 Sec.~\ref{sec:calculus} then abstracts away the inessential details of the language and user interface and makes the  intuitions developed in Sec.~\ref{sec:examples} formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions and tracking hole closures.
-The dynamic semantics of \HazelnutLive is a small-step reduction semantics equipped with clean type safety theorems (including, notably, a Progress theorem for incomplete expressions).
-The semantics of \HazelnutLive borrows some technical machinery from gradual type theory~\cite{DBLP:conf/snapl/SiekVCB15}
-(to handle incomplete types) and contextual modal type theory~(CMTT)~\cite{Nanevski2008} (which, by its Curry-Howard interpretation, provides a logical basis for reasoning about and operating on hole closures), combining these and adding additional machinery necessary to continue evaluation past holes and cast failures.
+% The dynamic semantics of \HazelnutLive is a small-step reduction semantics equipped with clean type safety theorems (including, notably, a Progress theorem for incomplete expressions). 
 Sec.~\ref{sec:agda-mechanization} outlines our Agda-based mechanization of \HazelnutLive, which is included in the anonymized supplement.  
-These connections to well-established existing systems, together with the mechanized proofs, serve to support our claim that this approach is theoretically well-grounded.
-
-Sec.~\ref{sec:implementation} outlines our companion implementation of \HazelnutLive, which is also available in the anonymized supplement.
-This implementation is a slimmed-down version of \Hazel that implements the live programming features as shown in Sec.~\ref{sec:examples},
-but for the more austere language of Sec.~\ref{sec:calculus} extended with numbers, sum types, and product types (which, for the sake of simplicity, are specified separately in the anonymized supplement).
-The editor component of this implementation defines a language of structured edit actions, 
-based on the \Hazelnut structure editor calculus developed by \citet{popl-paper}, that inserts holes automatically to guarantee that
-every editor state has some, possibly incomplete, type. 
-The type safety theorems then guarantee that every editor state has dynamic meaning. Taken together, the result is a full solution to the gap problem, i.e. a proof-of-concept
-live functional programming environment where the static and dynamic feedback never exhibits gaps. We formally state this continuity corrollary.
+Sec.~\ref{sec:implementation} provides some additional details on the implementation of these ideas into \Hazel, which is based on the \Hazelnut edit action calculus \cite{popl-paper}, and formally states the continuity invariant, i.e.
+the solution to the gap problem, as a corollary of the primary theorems of \Hazelnut and \HazelnutLive.
 
@@
 
-Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneously discusses directions for future work and Sec.~\ref{sec:discussion} concludes.
+Sec.~\ref{sec:relatedWork} describes related work in detail and simultaneously discusses limitations and directions for future work. Sec.~\ref{sec:discussion} briefly concludes.
 
-%The appendix includes some straightforward auxiliary definitions and proofs mentioned in the paper, some extensions to the core calculus, and some screenshots and additional details on the implementation.
-The supplemental material includes both the mechanization and the implementation.
+The anonymous supplement includes, in addition to the Agda mechanization, an appendix that (1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for concision; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions.
+% These are not necessary to communicate the fundamental ideas of the paper, but may be of interest to some readers.

### 2018-07-12 related-work.tex
final POPL19 submission

@@
 building on the static semantics developed by \citet{popl-paper}. That paper, as well as a subsequent ``vision paper'' introducing \Hazel, suggested as future work  
-a corresponding dynamic semantics for the purposes of live programming \cite{HazelnutSNAPL}. 
-This paper    
-develops the conceptual and type-theoretic details necessary to achieve this broader vision.
+a corresponding dynamic semantics for the purposes of live programming without semantic gaps \cite{HazelnutSNAPL}.
+Others have extensively argued the value of assigning meaning even to incomplete and erroneous programs \cite{Bayne:2011:ASD:1985793.1985864}.
+This paper delivers conceptual and type-theoretic details necessary to further advance this  vision.
 
@@
 at length in Sec.~\ref{sec:calculus}. The main innovation relative to this
-prior work is the treatment of cast failures as holes, rather than errors.
+prior work is the treatment of cast failures like holes, rather than errors.
 Many of the methods developed to make gradual typing more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein \cite{takikawa_et_al:LIPIcs:2015:5215}.
-For example, there has been substantial work on the problem of implementing casts efficiently \cite{herman2010space,siek2010threesomes,garcia2013threesomes}, and on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.
+For example, there has been substantial work on the problem of implementing casts efficiently \cite{herman2010space,siek2010threesomes,garcia2013threesomes}, and on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,DBLP:journals/pacmpl/DevriesePP18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.
 Another direction that we leave to future work is the integration of these ideas into dependently typed proof assistants,
-where evaluation is at the service of equational reasoning \cite{DBLP:journals/corr/abs-1009-2789}.
+where evaluation is primarily at the service of equational reasoning \cite{DBLP:journals/corr/abs-1009-2789}.
 Several modern proof assistants, e.g.
-Agda \cite{norell2009dependently,norell:thesis}, Idris \cite{brady2013idris} and Beluga \cite{DBLP:conf/flops/Pientka10} already provide initial syntactic and static support for holes (see Sec.~\ref{sec:intro}).
+Agda \cite{norell2009dependently,norell:thesis}, Idris \cite{brady2013idris} and Beluga \cite{DBLP:conf/flops/Pientka10} already provide  syntactic and static support for holes (see Sec.~\ref{sec:intro}) and interactive, hole-driven development is standard practice.
 
@@
 %
-\citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15} have shown how to incorporate such features into gradual type theory; we expect that the type safety properties established in this paper should also be conserved, with suitable modifications to account for a store.
+\citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15} show how to incorporate such features into gradual type theory (by treating $\mathtt{ref}(\tau)$ as invariant with respect to consistency); we expect that this approach would also work for the semantics of Sec.~\ref{sec:calculus}, i.e.
+it would conserve the type safety properties established there, with suitable modifications to account for a store.
+However, the commutativity property we establish in Sec.~\ref{sec:resumption} will not hold for a language that supports non-commutative effects.
+We leave to future work the task of defining more restricted special cases of the fill-and-resume operation that respects a suitable commutativity property in effectful settings (perhaps by checkpointing).
+
 %
 Going beyond references to incorporate external effects, e.g.
-IO effects, raises some important practical concerns, however---we do not want to continue past a hole or error and in so doing haphazardly trigger an unintended effect.
+IO effects, raises some additional practical concerns, however---we do not want to continue past a hole or error and in so doing haphazardly trigger an unintended effect.
 In this setting, it is likely better to explicitly ask the programmer whether to continue evaluation past a hole or cast failure.
-More generally, the small step specification in this paper is suitable as the basis for a single stepper, which generalizes evaluation.
+The small step specification in this paper is suitable as the basis for a  step-based evaluator like this.
 \citet{ocaml-stepper} discuss some unresolved usability issues relevant to single steppers for functional languages.
-Note also that the commutativity property we establish in Sec.~\ref{sec:resumption} will not hold for a language that supports non-commutative effects.
-We leave to future work the task of defining more restricted special cases of commutativity in effectful settings.
 
@@
 Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property. 
-These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.
-In any case, under this interpretation an analogous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.
+These issues are not relevant in \HazelnutLive because, as in gradual type theory, type holes admit non-termination \cite{Siek06a}.  
+In any case, under this interpretation an analogous problem arises---metavariables become variables of a function type, so we cannot rely on the standard notion of progress on closed terms.
 
-While there has been some formal work on reduction of open terms, leading to programs in a \emph{weak head normal form} \cite{barendregt84:_lambda_calculus,DBLP:journals/corr/abs-1009-2789}, there is no progress theorem in this setting nor is there environment tracking.
-Moreover, in the setting of an evaluator for a programming language, we do not want to evaluate under arbitrary binders but rather only around hole closures \cite{DBLP:conf/birthday/BlancLM05}.
-CMTT makes the appropriate distinctions between variables and metavariables, and our notion of an indeterminate form allows us to prove a progress theorem of familiar form.
+While there has been some formal work on reduction of open terms, leading to programs in a \emph{weak head normal form} \cite{barendregt84:_lambda_calculus,DBLP:journals/corr/abs-1009-2789,Abramsky:1990vv}, there is no clear progress theorem in this setting: in the setting of an evaluator for a programming language, we do not want to evaluate under arbitrary binders but rather only around holes when they would otherwise have been evaluated \cite{DBLP:conf/birthday/BlancLM05}.
+Another important distinction is that simple evaluation of open terms will not support hole closure tracking, which is an essential component of our approach.
+CMTT makes the appropriate distinctions between variables and metavariables, and our notion of an indeterminate form allows us to prove a progress theorem of familiar form (where there are no free variables).
 Weak head normal forms are more useful when using evaluation to perform optimizations throughout a program, e.g.
-when using supercompilation-by-evaluation \cite{DBLP:conf/haskell/BolingbrokeJ10} or symbolic evaluation \cite{King:1976}, or when using evaluation in the service of equational reasoning as in proof assistants (e.g.
-Beluga \cite{DBLP:journals/corr/abs-1009-2789}).
+when using supercompilation-by-evaluation \cite{DBLP:conf/haskell/BolingbrokeJ10} or symbolic evaluation \cite{King:1976,SurveySymExec-CSUR18}, or when using evaluation in the service of equational reasoning (e.g.
+in Beluga \cite{DBLP:journals/corr/abs-1009-2789}).
 
@@
 existing staging and partial evaluation systems are focused on partial evaluation with respect to an input that sits outside of a
-function \cite{Jones:1993uq}, whereas holes have non-trivial contexts.
+function \cite{Jones:1993uq}, whereas holes have  contexts.
 
@@
 %
-This paper bears out that it is also well-suited for the purpose of live functional programming with typed expression holes.
+This paper bears out that the basic constructs of CMTT are ideally suited for a purpose previously not considered: live functional programming with typed expression holes, with commutative edits.
 
@@
 
+\parahead{Type Error Messages}
+
+A key feature of our semantics is that it permits evaluation not only of terms with empty holes, but also terms with non-empty holes, i.e.
+reified static type inconsistencies.
+DuctileJ \cite{Bayne:2011:ASD:1985793.1985864} and GHC \cite{DBLP:conf/icfp/VytiniotisJM12}
+have also considered this problem, but have taken
+an ``exceptional approach'' --- these systems can defer static type errors until run-time, but do not continue further once the term containing the error
+ has been evaluated.
+
+Understanding and debugging static type errors is notoriously difficult,
+particularly for novices.
+%
+A variety of approaches have been
+proposed to better localize
+and explain type errors \cite{Seminal,ChenErwig2014,DBLP:journals/jfp/ChenE18,Pavlinovic2015,sherrloc}.
+%
+One of these approaches, by \citet{Seidel2016},  uses symbolic execution and program synthesis to generate a dynamic witness
+that demonstrates a run-time failure that could be caused by the static type error.
+%
+\HazelnutLive{} has similar motivations in that it can run programs with type errors  and provide concrete feedback about the values that erroneous terms take during evaluation (Sec.~\ref{sec:static-errors}-\ref{sec:dynamic-type-errors}).
+However, no attempt is made to synthesize examples that do not already appear in the program.
+%
+% Combining the strengths of these approaches may be fruitful in the future.
+
+\parahead{Coroutines} 
+The fill-and-resume interaction is reminiscent of the interactions that occur when using coroutines \cite{DBLP:conf/ifip/KahnM77} and related mechanisms (e.g.
+algebraic effects) --  a coroutine might yield to the caller until a value is sent back in and then continue in the suspended environment.
+The difference is that
+the hole filling can make use of the context
+around the hole.
+
+\parahead{Dynamic Error Propagation} 
+\citet{DBLP:conf/sp/HritcuGKPM13} consider the problem of dynamic violations of information-flow control (IFC) policies.
+An exceptional approach here is problematic in practice, because it would lead critical systems to shutdown entirely.
+Instead, the authors develop several mechanisms for
+propagating errors through subsequent computations (in a manner that preserves non-interference properties).
+The most closely related is the NaV-lax approach, which turns errors into special ``not-a-value'' (NaV) terms that consume other terms that they interact with (like floating point NaN).
+Our approach differs in that terms are not consumed.
+Furthermore, we track hole closures and consider static and dynamic type errors, but not exception propagation.
+
 \parahead{Debuggers}
@@
 Beyond that, there are two main distinctions.
-First, evaluation does not stop at each breakpoint and so it is straightforward to explore the \emph{space} of values that a variable takes.
+First, evaluation does not stop at each hole and so it is straightforward to explore the \emph{space} of values that a variable takes.
 Second, breakpoints, logs and tracing tools convey the values of a variable at a position in the evaluation trace.
 Hole closures, on the other hand, convey information from a syntactic position in the \emph{result} of evaluation.
-The result is, by nature, a simpler object.
+The result is, by nature, a simpler object than the trace.
 %  These approaches are fundamentally complementary in that standard debugging facilities are useful even when there are no holes in the program whereas our focus has been on incomplete programs.
+
 Some debuggers support ``edit-and-continue'', e.g.
@@
 These features do not come with any guarantee that rerunning the program will produce the same result.
-The ``fill-and-resume'' feature described in Sec.~\ref{sec:resumption} is a special case that comes with this stronger guarantee.
 
@@
 However, they do not generally support execution of incomplete programs, which was our focus in this paper.
-Scratch will execute a program with holes by simply skipping over incomplete commands, but this is a limited protocol because rerunning the program after filling the hole may produce a result unrelated to initial result in any way.
+Scratch will execute a program with holes by simply skipping over incomplete commands, but this is a limited protocol because rerunning the program after filling the hole may produce a result unrelated to initial result.
 Though in some situations, skipping over problematic commands has been observed to work surprisingly well \cite{DBLP:conf/dac/Rinard12}, this paper contributes a semantically sound approach (Sec.~\ref{sec:resumption}).
-Combined with the aforementioned UI advances, there is reason to believe that structure editing could be the foundation for a ``next-generation'' live programming experience.
+
+We make no empirical claims regarding the usability of the particular user interface presented in this paper. The \Hazel user interface as presented is a proof of concept
+that demonstrates (1) a comprehensive solution to the gap problem, as described in Sec.~\ref{sec:implementation}; and 
+(2) one possible user interface for presenting hole closures, including recursive hole closures, to the programmer (Fig.~\ref{fig:qsort-sidebars}).
+
+For larger programs, there are some limitations that call for further UI refinement. 
+Indeterminate results can get to be quite large, particularly when a hole or failed cast
+appears in guard position as in Fig.~\ref{fig:cast-errors}. 
+Although the pretty printer elides irrelevant branches, 
+the semantics in Sec.~\ref{sec:calculus} would also allow for evaluation to pause when such
+a form is produced, continuing only if requested.
+
+In larger programs, there can be many dozens of variables in scope, so search and sorting tools in the live context inspector would be useful. It would also be useful
+to support the evaluation of arbitrary ``watched'' expressions in the selected 
+closure. There are other ways to display the variable values, e.g. intercalated
+into the code \cite{DBLP:journals/corr/abs-1806-07449,lamdu}.
+
+Finally, it is not always helpful to continue evaluation as far as possible.
+In particular, when there is a hole or cast failure in the argument position
+of a function application, it may often be more helpful to stop before 
+beta reduction, to avoid exposing the internals of the applied function to 
+the client. The semantics in Sec.~\ref{sec:calculus} leaves evaluation order
+undefined, so various heuristic approaches are possible (e.g. we might wish avoid beta reduction  for functions from imported libraries).
+
+% With sufficient UI advances, there is reason to believe that structure editing could be the foundation for a powerful ``next-generation'' live programming experience, and the \Hazel effort, with the contributions of our dynamic semantics, is actively exploring this direction.
 We emphasize, however,
-that our proposed contributions will be relevant no matter how holes come to appear (see Sec.~\ref{sec:intro}). % We will say more about all of these systems later.
+% that our proposed contributions will be relevant no matter how holes come to appear in a program, including when they are inserted manually (see Sec.~\ref{sec:implementation}).
+% We will say more about all of these systems later.
 
@@
 
-\parahead{Type Error Messages}
+\parahead{Program Slicing}
 
-A key feature of our semantics is that it permits evaluation not only of terms with empty holes, but also terms with non-empty holes, i.e. reified static type inconsistencies. 
-Understanding and debugging static type errors is notoriously difficult,
-particularly for novices.
-%
-A variety of approaches have been
-proposed to better localize
-and explain type errors \cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc}.
-%
-One of these approaches, by \citet{Seidel2016},  uses symbolic execution and program synthesis to generate a dynamic witness
-that demonstrates a run-time failure that could be caused by the static type error, and then displays a compressed evaluation
-trace to the user as a graph.
-%
-\HazelnutLive{} has similar motivations in that it can run programs with static or dynamic type errors  and provide meaningful feedback about where ill-typed terms end up in the program, even after the first problem is encountered (Sec.~\ref{sec:static-errors}-\ref{sec:dynamic-type-errors}).
-However, no attempt is made to synthesize examples that do not already appear in the program.
-%
-Combining the strengths of these approaches may be fruitful in the future.
+Hole-like constructs also appear in work on program slicing
+\cite{DBLP:conf/icfp/PereraACL12,DBLP:journals/pacmpl/RicciottiSPC17}, where empty expression holes arise as a technical device to determine which parts of a complete program do not impact a selected part of the result of evaluating that program.
+In other words, the focus is on explaining a result that is presumed to exist, whereas the work in this paper is focused on producing results where they would not have otherwise been available.
+Combining the strengths of these approaches is another fruitful avenue for future research.
 
-\parahead{Program Synthesis and Slicing}
+\parahead{Program Synthesis}
 
@@
 synthesis engine.
-We leave to future work the possibility of combining these approaches, i.e.
-using the information generated by running an incomplete program as described in this paper to inform the generation of program and edit action suggestions, e.g.
-via tests or assertions that serve to generate constraints dynamically for type-and-example-driven synthesis \cite{DBLP:conf/popl/FrankleOWZ16}.
+We leave to future work the exciting possibility of combining these approaches.
+For example, one can imagine running an incomplete program as described in this paper and then adding tests or assertions about the value that a variable should take on in the different hole closures where it appears, via the live context inspector.
+These could serve as  constraints for use by a type-and-example-driven synthesis engine \cite{DBLP:conf/popl/FrankleOWZ16}.
 
@@
 
-Hole-like constructs also appear in work on program slicing
-\cite{DBLP:conf/icfp/PereraACL12,DBLP:journals/pacmpl/RicciottiSPC17}, where empty expressions holes arise as a technical device to determine which parts of a complete program do not impact a selected part of the result of evaluating that program.
-In other words, the focus is on explaining a result that is presumed to exist, whereas the work in this paper is focused on producing results where they would not have otherwise been available.
-Combining the strengths of these approaches is another fruitful avenue for future research.
-
 % \begin{itemize}

### 2018-07-17 calculus.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)

@@
 That paper also defined sum types as an extension to the core calculus.
-We follow suit on both counts in the supplemental appendix.%Appendix~\ref{sec:extensions}.
+We follow suit on both counts in the \appendixName.%Appendix~\ref{sec:extensions}.
 %
@@
 
-Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the supplemental appendix for the full definition).
+Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the \appendixName for the full definition).
 % Appendix~\ref{sec:additional-defns}).
@@
 transitions and a standard substitution lemma stated in
-the supplemental appendix.
+the \appendixName.
 % Appendix~\ref{sec:additional-defns}.
@@
 %
-The proof relies on canonical forms lemmas stated in the supplemental appendix.
+The proof relies on canonical forms lemmas stated in the \appendixName.
 % Appendix~\ref{sec:additional-defns}.
@@
 % Appendix~\ref{sec:additional-defns}
-The supplemental appendix defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
+The \appendixName defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
 %
@@
 calculus here. As with many full-scale systems, there is not currently a formal
-specification for the full \Elm{} or \Hazel language, but the supplemental appendix
+specification for the full \Elm{} or \Hazel language, but the \appendixName
 discusses how the standard approach for deriving a ``gradualized'' version of a 
@@
 We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
-transitive closure of stepping (see the supplemental appendix).
+transitive closure of stepping (see the \appendixName).
 % Appendix~\ref{sec:additional-defns}).
@@
 (There are various standard ways to formalize this intuition.
-For the sake of space, we review these in the supplemental appendix.)
+For the sake of space, we review these in the \appendixName.)
 % Appendix~\ref{sec:confluence}.)
@@
 We describe the proof, which is straightforward but involves a
-number of lemmas and definitions, in the supplemental appendix.
+number of lemmas and definitions, in the \appendixName.
 %Appendix~\ref{sec:hole-filling}.

### 2018-07-17 examples-3.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)

@@
 The result at the bottom of Fig.~\ref{fig:cast-errors} demonstrates that the first application
-of \li{f} is dynamically unproblematic. This allows the programmer confirm that 
-the first branch of the conditional operates as intended without the need to 
-address the typing problems in the other branch. 
-This flexibility is a common motivation for dynamic and gradual languages \cite{Bayne:2011:ASD:1985793.1985864}.
+of \li{f} is dynamically unproblematic. This allows the programmer to confirm that 
+the first branch operates as intended without the need to 
+address the typing problems in the other branch \cite{Bayne:2011:ASD:1985793.1985864}. 
+% This flexibility is a common motivation for dynamic and gradual languages .
 
@@
 
+\ifarxiv
+
+\else
 In this small example, it might have only been a small burden for the programmer to provide the intended types in the signature for \li{f}, but there are situations (e.g.
@@
 This approach ensures that dynamic feedback does not exhibit gaps even when there is a dynamic type error.
+\fi
 

### 2018-07-17 examples.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)

@@
 pattern matching in function arguments, list  notation, and record labels (currently there are only tuples).
-These conveniences are orthogonal to the contributions of this paper; all of the live programming and user interface features demonstrated in the figures in this section have been implemented.
+These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented.
 

### 2018-07-17 intro.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)

@@
 
-The anonymous supplement includes, in addition to the Agda mechanization, an appendix that (1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for concision; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions.
+\ifarxiv
+The appendix 
+\else
+The anonymous supplement includes, in addition to the Agda mechanization, an appendix that 
+\fi
+(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for concision; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions.
 % These are not necessary to communicate the fundamental ideas of the paper, but may be of interest to some readers.

### 2018-10-23 calculus.tex
close #7

@@
 
-We will now make the intuitions developed in the previous section formally precise by specifying a core calculus, which we call \HazelnutLive, and developing its metatheory.
-
-\parahead{Overview} The syntax of the core calculus given in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes.
-We distinguish between {external} expressions, $e$, and {internal} expressions, $d$.
-External expressions correspond to programs as entered by the programmer
-(see Sec.~\ref{sec:intro} for discussion of implicit, manual, semi-automated and fully automated hole entry methods).
-Each well-typed external expression (see Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
-We take this approach, notably also taken in the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and
-(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts.
-This machinery is inserted by the expansion step, rather than entered explicitly by the programmer.
-In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
-We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}
-(see Sec.~\ref{sec:agda-mechanization}).
-Rule names in this section, e.g.
-\rulename{SVar}, correspond to variable names used in the mechanization.
-The \Hazel implementation substantially follows the formal specification of \Hazelnut (for the editor component) and \HazelnutLive (for the semantics), and we can formally state a continuity invariant for a putative combined calculus (see Sec.~\ref{sec:implementation}).
+We will now make the intuitions developed in the previous section formally
+precise by specifying a core calculus, which we call \HazelnutLive, and
+developing its metatheory.
+
+\parahead{Overview} The syntax of the core calculus given in
+Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with
+holes.  We distinguish between {external} expressions, $e$, and {internal}
+expressions, $d$.  External expressions correspond to programs as entered
+by the programmer (see Sec.~\ref{sec:intro} for discussion of implicit,
+manual, semi-automated and fully automated hole entry methods).  Each
+well-typed external expression (see Sec.~\ref{sec:external-statics} below)
+expands to a well-typed internal expression (see Sec.~\ref{sec:expansion})
+before it is evaluated (see Sec.~\ref{sec:evaluation}).  We take this
+approach, notably also taken in the ``redefinition'' of Standard ML by
+\citet{Harper00atype-theoretic}, because (1) the external language supports
+type inference and explicit type ascriptions, $\hexp : \htau$, but it is
+formally simpler to eliminate ascriptions and specify a type assignment
+system when defining the dynamic semantics; and (2) we need additional
+syntactic machinery during evaluation for tracking hole closures and
+dynamic type casts.  This machinery is inserted by the expansion step,
+rather than entered explicitly by the programmer.  In this regard, the
+internal language is analogous to the cast calculus in the gradually typed
+lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will
+see the \HazelnutLive internal language goes beyond the cast calculus in
+several respects. We have mechanized these formal developments using the
+Agda proof assistant \cite{norell:thesis,norell2009dependently} (see
+Sec.~\ref{sec:agda-mechanization}). Rule names in this section,
+e.g. \rulename{SVar}, correspond to variable names used in the
+mechanization. The \Hazel implementation substantially follows the formal
+specification of \Hazelnut (for the editor component) and \HazelnutLive
+(for the semantics), and we can formally state a continuity invariant for a
+putative combined calculus (see Sec.~\ref{sec:implementation}).
 
@@
 
-Before describing the rules in detail, let us state two other guiding theorems.
-The following theorem establishes that an expansion exists for every well-typed external expression.
-The mechanization also establishes that when an expansion exists, it is unique (not shown).
+Before describing the rules in detail, let us state two other guiding
+theorems.  The following theorem establishes that an expansion exists for
+every well-typed external expression.
  \begin{thm}[Expandability] \label{thm:expandability}~
@@
  \end{thm}
+
 % \noindent
@@
 % \end{thm}
-% \noindent
-The following theorem establishes that expansion generalizes external typing.
+\noindent
+The following theorem establishes that expansion generalizes external
+typing.
 \begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
@@
       then $\hana{\hGamma}{\hexp}{\htau}$.
+  \end{enumerate}
+\end{thm}
+
+We also establish that the elaboration produces unique results.
+\begin{thm}[Elaboration Unicity] \label{thm:expansion-unicity}~
+  \begin{enumerate}[nolistsep]
+  \item If $\expandSyn{\hGamma}{\hexp}{\htau_1}{\dexp{}_1}{\Delta_1}$ and
+    $\expandSyn{\hGamma}{\hexp}{\htau_2}{\dexp{}_2}{\Delta_2}$, then
+    $\htau_1 = \htau_2$ and $\dexp{}_1 = \dexp{}_2$ and $\Delta_1 =
+    \Delta_2$.
+  \item If
+    $\expandAna{\hGamma}{\hexp}{\htau}{\dexp{}_1}{\htau_1}{\Delta_1}$ and
+    $\expandAna{\hGamma}{\hexp}{\htau}{\dexp{}_2}{\htau_2}{\Delta_2}$, then
+    $\htau_1 = \htau_2$ and $\dexp{}_1 = \dexp{}_2$ and $\Delta_1 =
+    \Delta_2$
   \end{enumerate}
@@
 % Appendix~\ref{sec:additional-defns}
-The \appendixName defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
+The \appendixName{} defines the predicates~$\isComplete{e}$,
+$\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
 %
@@
 %
-The following theorem establishes that expansion preserves program completeness.
+The following theorem establishes that expansion preserves program
+completeness.
 

### 2018-10-23 calculus.tex
close #6

--- 2018-07-17 calculus.tex
@@
 
-We will now make the intuitions developed in the previous section formally precise by specifying a core calculus, which we call \HazelnutLive, and developing its metatheory.
-
-\parahead{Overview} The syntax of the core calculus given in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes.
-We distinguish between {external} expressions, $e$, and {internal} expressions, $d$.
-External expressions correspond to programs as entered by the programmer
-(see Sec.~\ref{sec:intro} for discussion of implicit, manual, semi-automated and fully automated hole entry methods).
-Each well-typed external expression (see Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
-We take this approach, notably also taken in the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and
-(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts.
-This machinery is inserted by the expansion step, rather than entered explicitly by the programmer.
-In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.
-We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}
-(see Sec.~\ref{sec:agda-mechanization}).
-Rule names in this section, e.g.
-\rulename{SVar}, correspond to variable names used in the mechanization.
-The \Hazel implementation substantially follows the formal specification of \Hazelnut (for the editor component) and \HazelnutLive (for the semantics), and we can formally state a continuity invariant for a putative combined calculus (see Sec.~\ref{sec:implementation}).
+We will now make the intuitions developed in the previous section formally
+precise by specifying a core calculus, which we call \HazelnutLive, and
+developing its metatheory.
+
+\parahead{Overview} The syntax of the core calculus given in
+Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with
+holes.  We distinguish between {external} expressions, $e$, and {internal}
+expressions, $d$.  External expressions correspond to programs as entered
+by the programmer (see Sec.~\ref{sec:intro} for discussion of implicit,
+manual, semi-automated and fully automated hole entry methods).  Each
+well-typed external expression (see Sec.~\ref{sec:external-statics} below)
+elaborates to a well-typed internal expression (see
+Sec.~\ref{sec:elaboration}) before it is evaluated (see
+Sec.~\ref{sec:evaluation}).  We take this approach, notably also taken in
+the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic},
+because (1) the external language supports type inference and explicit type
+ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate
+ascriptions and specify a type assignment system when defining the dynamic
+semantics; and (2) we need additional syntactic machinery during evaluation
+for tracking hole closures and dynamic type casts.  This machinery is
+inserted by elaboration, rather than entered explicitly by the programmer.
+In this regard, the internal language is analogous to the cast calculus in
+the gradually typed lambda calculus
+\cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the
+\HazelnutLive internal language goes beyond the cast calculus in several
+respects. We have mechanized these formal developments using the Agda proof
+assistant \cite{norell:thesis,norell2009dependently} (see
+Sec.~\ref{sec:agda-mechanization}). Rule names in this section,
+e.g. \rulename{SVar}, correspond to variable names used in the
+mechanization. The \Hazel implementation substantially follows the formal
+specification of \Hazelnut (for the editor component) and \HazelnutLive
+(for the semantics), and we can formally state a continuity invariant for a
+putative combined calculus (see Sec.~\ref{sec:implementation}).
 
@@
 \vspace{-4px}
-\subsection{Expansion}
-\label{sec:expansion}
+\subsection{Elaboration}
+\label{sec:elaboration}
 \vspace{-1px}
 
-\input{fig-expansion}
+\input{fig-elab}
 \input{fig-typing-dexp}
 
-Each well-typed external expression~$e$ expands to a well-typed internal expression~$d$, for evaluation.
-%
-\Figref{fig:expansion} specifies expansion, and \Figref{fig:hasType} specifies type assignment for internal expressions.
+Each well-typed external expression~$e$ elaborates to a well-typed internal
+expression~$d$, for evaluation.
+%
+\Figref{fig:elaboration} specifies elaboration, and \Figref{fig:hasType}
+specifies type assignment for internal expressions.
 %
@@
 
-As with the type system for the external language (above),
-we specify expansion bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
-%
-The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
-%
-produces an expansion~$d$ and a hole context~$\hDelta$ when synthesizing type $\htau$ for $\hexp$.
+As with the type system for the external language (above), we specify
+elaboration bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
+%
+The synthetic elaboration
+judgement~$\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+%
+produces an elaboration~$d$ and a hole context~$\hDelta$ when synthesizing
+type $\htau$ for $\hexp$.
 %
 %We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
-We describe hole contexts, which serve as ``inputs'' to the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below.
-%
-The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$, when checking~$\hexp$ against~$\htau$.
-%
-The following theorem establishes that expansions are well-typed and in the analytic case that the assigned type, $\htau'$, is consistent with provided type, $\htau$.
-%
-\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
+We describe hole contexts, which serve as ``inputs'' to the type assignment
+judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below.
+%
+The analytic elaboration
+judgement~$\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$,
+produces an elaboration~$d$ of type~$\htau'$, and a hole context~$\hDelta$,
+when checking~$\hexp$ against~$\htau$.
+%
+The following theorem establishes that elaborations are well-typed and in
+the analytic case that the assigned type, $\htau'$, is consistent with
+provided type, $\htau$.
+%
+\begin{thm}[Typed Elaboration]\label{thm:typed-elaboration} ~
   \begin{enumerate}[nolistsep]
     \item
-      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      If $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
       then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
     \item
-      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
-      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
+      If $\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then
+      $\tconsistent{\htau}{\htau'}$ and
+      $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
   \end{enumerate}
@@
 %
- the subsumption rule permits us
-to check an external expression against any type consistent with the
-type that the expression \emph{actually} synthesizes, whereas every internal
-expression can be assigned at most one type, i.e. the following
-standard unicity property holds of the type assignment system.
+the subsumption rule permits us to check an external expression against any
+type consistent with the type that the expression \emph{actually}
+synthesizes, whereas every internal expression can be assigned at most one
+type, i.e. the following standard unicity property holds of the type
+assignment system.
 %
@@
 \noindent
-Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
-%
-For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.
-
-Before describing the rules in detail, let us state two other guiding theorems.
-The following theorem establishes that an expansion exists for every well-typed external expression.
-The mechanization also establishes that when an expansion exists, it is unique (not shown).
- \begin{thm}[Expandability] \label{thm:expandability}~
+Consequently, analytic elaboration reports the type actually assigned to
+the elaboration it produces.
+%
+For example, we can derive that
+$\elabAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.
+% where $\emptyset$ is the empty hole context.
+
+Before describing the rules in detail, let us state two other guiding
+theorems.  The following theorem establishes that every well-typed external
+expression can be elaborated.
+ \begin{thm}[Elaborability] \label{thm:elaborability}~
   \begin{enumerate}[nolistsep]
@@
       If $\hsyn{\hGamma}{\hexp}{\htau}$
-      then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      then $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
       for some $\dexp$ and $\Delta$.
@@
       If $\hana{\hGamma}{\hexp}{\htau}$
-      then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+      then $\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
       for some $\dexp$ and $\htau'$ and $\Delta$.
@@
 %   %   \item
-%       If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
-%       and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
+%       If $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+%       and $\elabSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
 %       then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
 %   %   \item
-%   %     If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
-%   %     and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
+%   %     If $\elabAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
+%   %     and $\elabAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
 %   %     then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
@@
 % \end{thm}
-% \noindent
-The following theorem establishes that expansion generalizes external typing.
-\begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
+\noindent
+The following theorem establishes that elaboration generalizes external
+typing.
+\begin{thm}[Elaboration Generality] \label{thm:elaboration-generality}~
   \begin{enumerate}[nolistsep]
     \item
-      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      If $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
       then $\hsyn{\hGamma}{\hexp}{\htau}$.
     \item
-      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+      If $\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
       then $\hana{\hGamma}{\hexp}{\htau}$.
@@
 
-The rules governing expansion of constants, variables and lambda expressions
---- Rules \rulename{ESConst}, \rulename{ESVar}, \rulename{ESLam} and \rulename{EALam}
---- mirror the corresponding type assignment rules
---- Rules \rulename{TAConst}, \rulename{TAVar} and \rulename{TALam}
---- and in turn, the corresponding bidirectional typing rules from Fig.~\ref{fig:bidirectional-typing}.
+We also establish that the elaboration produces unique results.
+\begin{thm}[Elaboration Unicity] \label{thm:expansion-unicity}~
+  \begin{enumerate}[nolistsep]
+  \item If $\elabSyn{\hGamma}{\hexp}{\htau_1}{\dexp{}_1}{\Delta_1}$ and
+    $\elabSyn{\hGamma}{\hexp}{\htau_2}{\dexp{}_2}{\Delta_2}$, then
+    $\htau_1 = \htau_2$ and $\dexp{}_1 = \dexp{}_2$ and $\Delta_1 =
+    \Delta_2$.
+  \item If
+    $\elabAna{\hGamma}{\hexp}{\htau}{\dexp{}_1}{\htau_1}{\Delta_1}$ and
+    $\elabAna{\hGamma}{\hexp}{\htau}{\dexp{}_2}{\htau_2}{\Delta_2}$, then
+    $\htau_1 = \htau_2$ and $\dexp{}_1 = \dexp{}_2$ and $\Delta_1 =
+    \Delta_2$
+  \end{enumerate}
+\end{thm}
+
+The rules governing elaboration of constants, variables and lambda
+expressions---Rules \rulename{ESConst}, \rulename{ESVar}, \rulename{ESLam}
+and \rulename{EALam}---mirror the corresponding type assignment rules---
+Rules \rulename{TAConst}, \rulename{TAVar} and \rulename{TALam}---and in
+turn, the corresponding bidirectional typing rules from
+Fig.~\ref{fig:bidirectional-typing}.
 %
@@
 To support type assignment, all lambdas in the internal language are
-half-annotated---Rule \rulename{EALam} inserts the annotation when expanding an
-unannotated external lambda based on the given type.
-%
-The rules governing hole expansion, and the rules that perform \emph{cast insertion}---those governing
-function application and type ascription---are more interesting.
-Let us consider each of these two groups of rules in turn in Sec.~\ref{sec:hole-expansion} and Sec.~\ref{sec:cast-insertion}, respectively.
-
-\subsubsection{Hole Expansion}\label{sec:hole-expansion}
-Rules \rulename{ESEHole}, \rulename{ESNEHole}, \rulename{EAEHole} and \rulename{EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$.
-%
-The hole name~$u$ on a hole closure identifies the external hole to which the hole closure corresponds.
-%
-While we assume each hole name to be unique in the external language, once evaluation begins, there may be multiple hole closures with the same name due to substitution.
-%
-For example, the result from Fig.~\ref{fig:grades-example} shows three closures for the hole named 1.
-%
-There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and \li{1:3}, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.
-
-%
-For each hole, $u$, in an external expression, the hole context generated by expansion, $\Delta$, contains a hypothesis of the form~$\Dbinding{u}{\hGamma}{\htau}$, which records the hole's type, $\tau$, and the typing context, $\Gamma$, from where it appears in the original expression.\footnote{
-We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.}
-%
-We borrow this hole context notation from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
+half-annotated---Rule \rulename{EALam} inserts the annotation when
+elaborating an unannotated external lambda based on the given type.
+%
+The rules governing hole elaboration, and the rules that perform \emph{cast
+  insertion}---those governing function application and type
+ascription---are more interesting. Let us consider each of these two groups
+of rules in turn in Sec.~\ref{sec:hole-elaboration} and
+Sec.~\ref{sec:cast-insertion}, respectively.
+
+\subsubsection{Hole Elaboration}\label{sec:hole-elaboration}
+Rules \rulename{ESEHole}, \rulename{ESNEHole}, \rulename{EAEHole} and
+\rulename{EANEHole} govern the elaboration of empty and non-empty
+expression holes to empty and non-empty \emph{hole closures},
+$\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$.
+%
+The hole name~$u$ on a hole closure identifies the external hole to which
+the hole closure corresponds.
+%
+While we assume each hole name to be unique in the external language, once
+evaluation begins, there may be multiple hole closures with the same name
+due to substitution.
+%
+For example, the result from Fig.~\ref{fig:grades-example} shows three
+closures for the hole named 1.
+%
+There, we numbered each hole closure for a given hole sequentially,
+\li{1:1}, \li{1:2} and \li{1:3}, but this is strictly for the sake of
+presentation, so we omit hole closure numbers from the core calculus.
+
+%
+For each hole, $u$, in an external expression, the hole context generated
+by elaboration, $\Delta$, contains a hypothesis of the
+form~$\Dbinding{u}{\hGamma}{\htau}$, which records the hole's type, $\tau$,
+and the typing context, $\Gamma$, from where it appears in the original
+expression.\footnote{ We use a hole context, rather than recording the
+  typing context and type directly on each hole closure, to ensure that all
+  closures for a hole name have the same typing context and type.}
+%
+We borrow this hole context notation from contextual modal type theory
+(CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and
+hole contexts with modal contexts (we say more about the connection with
+CMTT below).
 %
@@
 %
-In the synthetic hole expansion rules~\rulename{ESEHole} and~\rulename{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the external typing rules.
-%
-However, the first two premises of the expansion subsumption rule~\rulename{EASubsume} disallow the use of subsumption for holes in analytic position.
-%
-Instead, we employ separate analytic rules~\rulename{EAEHole} and~\rulename{EANEHole}, which each record the checked type~$\tau$ in the hole context.
-%
-Consequently, we can use type assignment for the internal language --- the type assignment rules \rulename{TAEHole} and \rulename{TANEHole} in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.
-
-Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
+In the synthetic hole elaboration rules~\rulename{ESEHole}
+and~\rulename{ESNEHole}, the generated hole context assigns the hole
+type~$\tehole$ to hole name~$u$, as in the external typing rules.
+%
+However, the first two premises of the elaboration subsumption
+rule~\rulename{EASubsume} disallow the use of subsumption for holes in
+analytic position.
+%
+Instead, we employ separate analytic rules~\rulename{EAEHole}
+and~\rulename{EANEHole}, which each record the checked type~$\tau$ in the
+hole context.
+%
+Consequently, we can use type assignment for the internal language --- the
+type assignment rules \rulename{TAEHole} and \rulename{TANEHole} in
+Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the
+corresponding type from the hole context.
+
+Each hole closure also has an associated environment~$\sigma$ which
+consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$
+for $n \geq 0$.
 %
@@
 %
-Initially, when no evaluation has yet occurred, the hole expansion rules generate the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
+Initially, when no evaluation has yet occurred, the hole elaboration rules
+generate the identity substitution for the typing context associated with
+hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$,
+and define as follows.
 %
@@
 
-Expansion inserts casts at function applications and ascriptions.
+Elaboration inserts casts at function applications and ascriptions.
 %
 The latter is more straightforward: Rule~\rulename{ESAsc}
-in \Figref{fig:expandSyn} inserts a cast from the assigned type to the
+in \Figref{fig:elabSyn} inserts a cast from the assigned type to the
 ascribed type.
 %
-Theorem~\ref{thm:typed-expansion} inductively ensures that the two
-types are consistent.
+Theorem~\ref{thm:typed-elaboration} inductively ensures that the two types
+are consistent.
 %
@@
 : \tau = \hap{(\halam{x}{\htau}{x})}{e}$; as such, application
-expansion, discussed below, is more general.
-
-Rule~\rulename{ESAp} expands function applications.
-%
-To understand the rule, consider the expansion of external
+elaboration, discussed below, is more general.
+
+Rule~\rulename{ESAp} elaborates function applications.
+%
+To understand the rule, consider the elaboration of external
 expression~$\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$, the example
@@
                 {\dcasttwo{c}{b}{\tehole}}
-                }_{\textrm{expansion of function body}~x(c)}
+                }_{\textrm{elaboration of function body}~x(c)}
         }
@@
 %
-where expansion inserts a cast on both the function expression~$x$ and its argument~$c$.
+where elaboration inserts a cast on both the function expression~$x$ and
+its argument~$c$.
 %
@@
 %
-We separately consider the expansions of~$x$ and of~$c$.
+We separately consider the elaborations of~$x$ and of~$c$.
 
@@
 %
-by performing analytic expansion on the function expression with this
+by performing analytic elaboration on the function expression with this
 matched arrow type.
 %
-The resulting expansion has some type~$\tau_1'$ consistent with the matched arrow type.
+The resulting elaboration has some type~$\tau_1'$ consistent with the
+matched arrow type.
 %
 In this case, because the subexpression~$x$ is a variable, analytic
-expansion goes through subsumption so that type~$\tau_1'$ is
+elaboration goes through subsumption so that type~$\tau_1'$ is
 simply~$\tehole$.
@@
 %
-We go through type synthesis, \emph{then} analytic expansion, so that the hole
-context records the matched arrow type for holes in function position,
-rather than the type~$\tehole$ for all such holes, as would be the case
-in a variant of this rule using synthetic expansion for the function
+We go through type synthesis, \emph{then} analytic elaboration, so that the
+hole context records the matched arrow type for holes in function position,
+rather than the type~$\tehole$ for all such holes, as would be the case in
+a variant of this rule using synthetic elaboration for the function
 expression.
@@
 The conclusion of Rule~\rulename{ESAp} inserts the cast on the argument's
-expansion, from the type it is assigned by the final premise of the
-rule~(type~$b$ here), to the argument type of the matched arrow type
-of the function expression~(type~$\tehole$ here).
+elaboration, from the type it is assigned by the final premise of the
+rule~(type~$b$ here), to the argument type of the matched arrow type of the
+function expression~(type~$\tehole$ here).
 
 The example's second, outermost application goes through the same
-application expansion rule.
+application elaboration rule.
 %
@@
 %
-However, it is safe in practice to eliminate such identity casts during expansion,
-and some formal accounts of gradual typing do so by defining three
-application expansion rules, including the original account of \citet{Siek06a}.
+However, it is safe in practice to eliminate such identity casts during
+elaboration, and some formal accounts of gradual typing do so by defining
+three application elaboration rules, including the original account of
+\citet{Siek06a}.
 
@@
 
-To recap, the result of expansion is a well-typed internal expression with appropriately initialized hole closures and casts.
-This section specifies the dynamic semantics of \HazelnutLive as a ``small-step'' transition system over internal expressions equipped with a meaningful notion of type safety even for incomplete programs, i.e.
-expressions typed under a non-empty hole context, $\Delta$.
-We establish that evaluation does not stop immediately when it encounters a hole, nor when a cast fails, by precisely characterizing when evaluation \emph{does} stop.
-We also establish that an essentially standard notion of type safety holds when running complete programs.
+To recap, the result of elaboration is a well-typed internal expression
+with appropriately initialized hole closures and casts.  This section
+specifies the dynamic semantics of \HazelnutLive as a ``small-step''
+transition system over internal expressions equipped with a meaningful
+notion of type safety even for incomplete programs, i.e. expressions typed
+under a non-empty hole context, $\Delta$.  We establish that evaluation
+does not stop immediately when it encounters a hole, nor when a cast fails,
+by precisely characterizing when evaluation \emph{does} stop.  We also
+establish that an essentially standard notion of type safety holds when
+running complete programs.
 
@@
 %
-The following theorem establishes that expansion preserves program completeness.
-
-\begin{thm}[Complete Expansion] 
+The following theorem establishes that elaboration preserves program
+completeness.
+
+\begin{thm}[Complete Elaboration]
 % ~
@@
       If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
-      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
+      and $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
       then $\isComplete{\htau}$ and $\isComplete{\dexp}$ and $\Delta = \emptyset$.
@@
 % and $\isComplete{\htau}$
-%       and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
+%       and $\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
 %       then $\isComplete{\dexp}$ and $\isComplete{\htau'}$ and $\Delta=\emptyset$
@@
 
-The following preservation theorem establishes that stepping preserves program completeness.
+The following preservation theorem establishes that stepping preserves
+program completeness.
 \begin{thm}[Complete Preservation]
@@
 
-By composing this Sensibility property with
-the \Property{Expandability}, \Property{Typed Expansion}, \Property{Progress} and \Property{Preservation} properties from this section, we establish a
+By composing this Sensibility property with the \Property{Elaborability},
+\Property{Typed Elaboration}, \Property{Progress} and
+\Property{Preservation} properties from this section, we establish a
 uniquely powerful Continuity invariant:
@@
     $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\tau'}$ then
-    $\expandSyn{\emptyset}{\removeSel{\zexp'}}{\htau'}{\dexp}{\Delta}$
+    $\elabSyn{\emptyset}{\removeSel{\zexp'}}{\htau'}{\dexp}{\Delta}$
       for some $\dexp$ and $\Delta$ such that
@@
 
-We do not separately define hole filling in the external language
-(i.e. we consider a change to an external expression to be a hole filling if the new expansion differs from
-the previous expansion up to hole filling).
-In practice, it may be useful to cache more than one recent edit state
-to take full advantage of hole filling.
-As an example, consider two edits, the first filling a hole~$u$ with
-the number~$2$, and the next applying operator~$+$, resulting in $2
-+ \dehole{v}{\sigma}{}$.
+We do not separately define hole filling in the external language (i.e. we
+consider a change to an external expression to be a hole filling if the new
+elaboration differs from the previous elaboration up to hole filling).  In
+practice, it may be useful to cache more than one recent edit state to take
+full advantage of hole filling.  As an example, consider two edits, the
+first filling a hole~$u$ with the number~$2$, and the next applying
+operator~$+$, resulting in $2 + \dehole{v}{\sigma}{}$.
 %
@@
 \begin{theorem}[Maximum Informativity]
-If the expansion produces $t1$, and there exists another possible type choice
-$t2$, then $t1 \sim t2$ and $t1 JOIN t2 = t1$
+If the elaboration produces $t1$, and there exists another possible type
+choice $t2$, then $t1 \sim t2$ and $t1 JOIN t2 = t1$
 \end{theorem}\footnote{idea is that special casing the holes in EANEHole gives you ``the

### 2018-10-23 calculus.tex
close #17

@@
   If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
-  (a) $\stepsToD{}{\dexp}{\dexp'}$ or
+  (a) there exists $\dexp'$ such that $\stepsToD{}{\dexp}{\dexp'}$ or
   (b) $\isBoxedValue{\dexp}$ or
@@
 \begin{thm}[Complete Progress]
-  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
-  and $\isComplete{\dexp}$
-  then either $\stepsToD{}{\dexp}{\dexp'}$
-  or $\isValue{\dexp}$.
+  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isComplete{\dexp}$
+  then either there exists a $\dexp'$ such that
+  $\stepsToD{}{\dexp}{\dexp'}$, or $\isValue{\dexp}$.
 \end{thm}

### 2018-10-23 calculus.tex
close #16

@@
 }
-The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as
-either ordinary values~(Rule~\rulename{BVVal}), or one of two cast forms: casts
-between disequal function types and casts from a ground
-type to the hole type. In each case, the cast must appear inductively on a boxed
-value.
-These forms are irreducible because they represent values that have been boxed
-but have never flowed into a corresponding ``unboxing'' cast, discussed below.
+The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as either
+ordinary values~(Rule~\rulename{BVVal}), or one of two cast forms: casts
+between unequal function types and casts from a ground type to the hole
+type. In each case, the cast must appear inductively on a boxed value.
+These forms are irreducible because they represent values that have been
+boxed but have never flowed into a corresponding ``unboxing'' cast,
+discussed below.
 %
@@
 when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type
-disequal to $\tau_1$.
+not equal to $\tau_1$.
 %

### 2018-10-23 calculus.tex
close #11
close #12

@@
 does not stop immediately when it encounters a hole, nor when a cast fails,
-by precisely characterizing when evaluation \emph{does} stop.  We also
-establish that an essentially standard notion of type safety holds when
-running complete programs.
-
-It is perhaps worth stating at the outset that a dynamic semantics equipped with these properties does not simply ``fall out'' from the observations made above that (1) empty hole closures correspond to metavariable closures from CMTT \cite{Nanevski2008} and (2) casts also arise in gradual type theory \cite{DBLP:conf/snapl/SiekVCB15}.
+by precisely characterizing when evaluation \emph{does} stop. In the case
+of complete programs, we also recover the familiar statements of progress
+and preservation for the simply typed lambda calculus.
+
+It is perhaps worth stating at the outset that a dynamic semantics equipped
+with these properties does not simply ``fall out'' from the observations
+made above that (1) empty hole closures correspond to metavariable closures
+from CMTT \cite{Nanevski2008} and (2) casts also arise in gradual type
+theory \cite{DBLP:conf/snapl/SiekVCB15}.
 %
@@
 %
-The judgement $\isIndet{d}$ defines \emph{indeterminate} forms, so
-named because they are rooted at expression holes and
-failed casts, and so, conceptually, their ultimate value
-awaits programmer action (see Sec.~\ref{sec:resumption}).
+The judgement $\isIndet{d}$ defines \emph{indeterminate} forms, so named
+because they are rooted at expression holes and failed casts, and so,
+conceptually, their ultimate value awaits programmer action (see
+Sec.~\ref{sec:resumption}). Note that no term is both complete, i.e. has no
+holes, and indeterminate.
 %
@@
 Although this paper focuses on running \emph{incomplete} programs, it helps
-to know that the necessary machinery does not interfere
-with running \emph{complete} programs, i.e. those
-with no type or expression holes.
+to know that the necessary machinery does not interfere with running
+\emph{complete} programs, i.e. those with no type or expression holes.
 %

### 2018-10-23 abstract.tex
added ACM formatting stuff + tweaked abstract + Aadded various zip files from previous submissions

@@
 of type errors. Consequently,    
-live feedback presented to the programmer has temporal or perceptive gaps, i.e. it flickers in and out or it goes stale whenever the program becomes incomplete. 
+live feedback presented to the programmer exhibits temporal or perceptive gaps, i.e. it flickers out or it goes stale whenever the program becomes incomplete. 
 % In some cases, these gaps persist over a substantial length of time.
@@
 \emph{a dynamic semantics for incomplete functional programs}, 
-starting from the static semantics recently 
-developed by \citet{popl-paper}. 
+starting from the static semantics of \citet{popl-paper}. 
 %
@@
 Rather than aborting when evaluation encounters any of these holes as in
-several existing systems, evaluation proceeds around holes,
+some existing systems, evaluation proceeds around holes,
 tracking the 
-closure around each hole instance as it flows through the remainder of the program. Editor services can report information from these hole closures 
-to help the programmer decide how to fill the holes in the program. 
+closure around each hole instance as it flows through the remainder of the program. Editor services can use the information in these hole closures 
+to help the programmer develop and confirm their mental model of the behavior of the complete portions of the program as they decide how to fill the remaining holes. 
 Hole closures also enable a \emph{fill-and-resume} operation that 
 avoids the need to restart evaluation after edits that amount to hole filling. 
-Formally, the semantics borrows machinery from both gradual type theory (to handle
-type holes) and contextual modal type theory (which provides a
+Formally, the semantics borrows machinery from both gradual type theory (to handle unfillable 
+type holes) and contextual modal type theory (which supplies the
 logical basis for hole closures), combining these and developing additional machinery necessary 
@@
 %
-We have mechanized the metatheory of this core calculus, called \HazelnutLive{}, using the Agda proof assistant.
+We have mechanized the metatheory of the core calculus, called \HazelnutLive{}, using the Agda proof assistant.
 %

### 2018-10-23 supplement-popl19/agda-snapshot/postulates.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+grep -n --color=auto -A 1 'postulate' *.agda

### 2018-10-23 hazelnut-dynamics.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)
added ACM formatting stuff + tweaked abstract + Aadded various zip files from previous submissions
added CCS concepts + keywords to front matter

@@
 \PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
-%% For double-blind review submission
-%\documentclass[acmsmall,review]{acmart}
+\newif\ifarxiv
+\arxivfalse
+\ifarxiv
+\documentclass[acmsmall,nonacm]{acmart}
+\else
 %%% Note: arxiv does not want line numbers (they are detected somehow, and are not allowed)
-\documentclass[acmsmall,review,anonymous,nonacm]{acmart}
-\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\documentclass[acmsmall]{acmart}
+% \settopmatter{printfolios=false,printccs=false,printacmref=false}
 % \settopmatter{printccs=false,printacmref=false}
+\fi
+
+\ifarxiv
+\newcommand{\appendixName}{appendix}
+\else
+\newcommand{\appendixName}{supplemental appendix}
+\fi
 
@@
 \usepackage{wrapfig}
+\usepackage{extarrows}
 % \usepackage[subtle]{savetrees}
@@
 \newtheorem{prop}[thm]{Proposition}
+\newtheorem{corol}[thm]{Corollary}
 % \newtheorem{cor}[section]{Corollary}     
@@
 \acmArticle{1}
-\acmYear{2018}
-\acmMonth{3}
+\acmYear{2019}
+\acmMonth{1}
 \acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
 
-\acmYear{2018}
+\acmYear{2019}
 \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
@@
 % \end{CCSXML}
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+<concept_desc>Software and its engineering~Functional languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+% <concept>
+% <concept_id>10003752.10010124.10010131.10010134</concept_id>
+% <concept_desc>Theory of computation~Operational semantics</concept_desc>
+% <concept_significance>500</concept_significance>
+% </concept>
+</ccs2012>
+\end{CCSXML}
 
-% \ccsdesc[500]{Software and its engineering~General programming languages}
-% \ccsdesc[300]{Social and professional topics~History of programming languages}
+\ccsdesc[500]{Software and its engineering~Functional languages}
+% \ccsdesc[500]{Theory of computation~Operational semantics}
 %% End of generated code
@@
 %% comma separated list
-% \keywords{keyword1, keyword2, keyword3}  %% \keywords is optional
+\keywords{live programming, gradual typing, contextual modal type theory, dynamic software updating, typed holes, structured editing}  %% \keywords is optional
 
@@
 
+\ifarxiv
 \clearpage
@@
 \input{extensions}
+\else
+
+\fi
 % \input{misc}
-\input{implementation-appendix}
+% \input{implementation-appendix}
 

### 2018-10-26 abstract.tex
a few more clarifying edits to the abstract

@@
 logical basis for hole closures), combining these and developing additional machinery necessary 
-to continue evaluation past holes while maintaining important metatheoretic properties.
+to continue evaluation past holes while maintaining type safety. We have mechanized the metatheory of the core calculus, called \HazelnutLive{}, using the Agda proof assistant.
 
@@
 %
-We have mechanized the metatheory of the core calculus, called \HazelnutLive{}, using the Agda proof assistant.
 %
-We have also implemented these ideas into the \Hazel programming environment. The implementation inserts holes automatically to guarantee that every editor state has some (possibly incomplete)
-type, following the \Hazelnut edit action calculus of \citet{popl-paper}.
+We have also implemented these ideas into the \Hazel programming environment.
+The implementation inserts holes automatically, following the \Hazelnut edit action calculus of \citet{popl-paper}, to guarantee that every editor state has some (possibly incomplete)
+type.
 Taken together with this paper's type safety property, the 
@@
 a proof-of-concept live programming environment where rich dynamic feedback 
-is truly available without gaps, i.e. it is available for every possible editor state.
+is truly available without gaps, i.e. for every reachable editor state.
 % i.e. \emph{every edit state is guaranteed to have

### 2018-10-26 intro.tex
revised the intro for general clarity + to make our specific contributions more clear + to include URLs to the various artifacts (Agda stuff, implementation and extended version of the paper on ArXiV)

@@
 
-Programmers typically shift back and forth between program editing and program evaluation many times before converging upon a program that behaves as intended. 
-Live programming environments aim to support this workflow by interleaving editing and evaluation so as to   
+Programmers typically shift between program editing and program evaluation many times before converging upon a program that behaves as intended. 
+Live programming environments aim to support this workflow by granularly interleaving editing and evaluation so as to   
 narrow what \citet{burckhardt2013s} call the ``temporal and perceptive gap'' between these activities.
@@
 programming languages typically assign meaning only to {complete programs}, i.e.
-programs that are syntactically well-formed and free of static type and binding errors.
-A program editor, however, frequently encounters incomplete, and therefore meaningless, editor states.
+programs that are syntactically well-formed and free of type and binding errors.
+A program editor, however, frequently encounters incomplete, and therefore conventionally meaningless, editor states.
 As a result, live feedback either ``flickers out'', creating a temporal gap, or it ``goes stale'', i.e.
-it relies on the most recent complete editor state, creating a perceptive gap because the feedback may not accurately reflect what the programmer is seeing in the editor.
+it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not accurately reflect what the programmer is editing.
 
-In some cases these gaps are momentary, like while the programmer
-%\IS entering
-is entering  
+In some cases these gaps are brief, like while the programmer
+is in the process of entering  
 a short expression.
-In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank or when the programmer makes a mistake.
+In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank or when the programmer is puzzling over a mistake.
 %
@@
 The problem is particularly pronounced for languages with rich static type systems where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
-Throughout the process of addressing these errors, the program text remains formally meaningless.
+Throughout the process of addressing these errors, which can sometimes span many days, the program text remains formally meaningless.
 Overall, about 40\% of edits performed by Java programmers using Eclipse left the program text malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.
@@
 %}
-In recognition of this ``{gap problem}''---that incomplete programs are formally meaningless---\citet{popl-paper} develop a static semantics (i.e. a type system) for incomplete 
+In recognition of this ``{gap problem}''---that incomplete programs are formally meaningless---\citet{popl-paper} developed a static semantics (i.e. a type system) for incomplete 
 functional programs, modeling them formally as typed expressions with \emph{holes} in 
@@
 % editor services can reason about types and binding in many more situations than previously possible.
-\citet{popl-paper,HazelnutSNAPL} discuss several ways to determine an incomplete expression from the editor state.
-When the editor state is a text buffer, error recovery mechanisms can insert holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09}.
+\citet{HazelnutSNAPL} discuss several ways to determine an incomplete expression from the editor state.
+When the editor state is a text buffer, error recovery mechanisms might insert holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09}.
 Alternatively, the language might provide explicit syntax for holes, so that the programmer can insert them either manually
@@
 For example, GHC Haskell supports the notation \li{_u} for empty holes, where \li{u} is an optional hole name \cite{GHCHoles}.
-Structure editors insert explicitly represented holes fully automatically \cite{popl-paper}; we say more about structure editors in Sec.~\ref{sec:implementation}.
+When the editor state is a tree or graph structure, i.e.
+in a structure editor, the editor typically inserts explicitly represented holes fully automatically \cite{popl-paper}; we say more about structure editors in Sec.~\ref{sec:implementation}.
 
@@
 raising an exception as a placeholder, e.g. \lismall{raise Unimplemented}. 
-GHC Haskell supports this mode of evaluation for programs with holes using the \lismall{-fdefer-typed-holes} flag.\footnote{
+GHC Haskell supports this ``exceptional approach'' using the \lismall{-fdefer-typed-holes} flag.\footnote{
 Without this flag, holes cause compilation to fail with an error message that reports information about each hole's type and typing context. 
@@
 } 
-Although better than nothing, this ``exceptional approach'' to expression holes has limitations 
+Although better than nothing, the exceptional approach to expression holes has limitations 
 within a live programming environment because 
-(1)~it provides limited information about the dynamic state of the program where the exception occurs
-(typically only a stack trace);  
-(2)~it provides no information about the behavior of the remainder of program, 
-parts of which may not depend on the missing or erroneous expression (e.g. later cells in a lab notebook, or tests of those components of the program that are already complete); and 
+(1)~it provides no information about the behavior of the remainder of program, 
+parts of which may not depend on the missing or erroneous expression (e.g. subsequent cells in a lab notebook, or tests involving other components of the program);  
+(2)~it provides limited information about the dynamic state of the program where the hole appears 
+(typically only a stack trace);  and
 (3)~it provides no means by which to resume evaluation after filling a hole.
@@
 Again, existing approaches do not support this situation well---GHC supports type holes, but compilation fails if type inference cannot automatically fill them \cite{GHCHoles}.
-The static semantics developed by \citet{popl-paper} derives the machinery for 
+The static semantics developed by \citet{popl-paper} provides more hope because it derives the machinery for 
 reasoning statically about type holes from gradual type theory, 
@@
 gradual type theory, 
-which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
+which inserts dynamic casts as necessary to maintain type safety. 
 %However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
 However, when a cast fails, evaluation again stops with 
-an exception and so the traditional gradual typing approaches still leave the live programming environment unable to provide rich, continuous feedback for the three reasons just enumerated.
+an exception, again leaving the live programming environment unable to provide rich, continuous feedback about the behavior of the remainder of the program.
 
@@
 
-This paper develops a dynamic semantics for incomplete functional programs, starting from the static semantics developed by \citet{popl-paper},  that addresses these limitations of the exceptional approach.
-In particular, rather than stopping when evaluation encounters an expression hole, evaluation continues ``around'' the hole so that dynamic feedback about other parts of the program remains available.
-For programs with unfillable type holes, casts are inserted as in gradual type theory \cite{DBLP:conf/snapl/SiekVCB15}. Evaluation proceeds around failed casts in the same way.
+This paper develops a dynamic semantics for incomplete functional programs, starting from the static semantics developed by \citet{popl-paper},  that addresses the three limitations of the exceptional approach enumerated above.
 
-The system tracks the closure around each expression hole instance as evaluation proceeds so that 
-the live programming environment can feed relevant information from the {hole closures} to the programmer as they work to fill the holes in the program.
-Then, when the programmer performs an edit that fills an empty expression hole or that replaces a non-empty hole with a type-correct expression, evaluation can resume from the paused, i.e.
-\emph{indeterminate}, evaluation state.
+In particular, rather than stopping when evaluation encounters an expression hole instance, evaluation continues ``around'' it.   
+The system tracks the closure around each expression hole instance as evaluation proceeds.
+The live programming environment can feed the incomplete result and relevant information from these {hole closures} to the programmer to help them develop and confirm their mental model of the portions of the program that are complete as they work to fill the remaining holes.
+Then, when the programmer performs an edit that fills an empty expression hole or that replaces a non-empty hole with a type-correct expression, evaluation can resume from the previous evaluation state.
 We call this operation \emph{fill-and-resume}.
-We observe that empty hole closures are closely related to metavariable closures from contextual modal type theory~(CMTT)~\cite{Nanevski2008} (which, by its Curry-Howard interpretation, provides a logical basis for reasoning about and operating on hole closures).
-These connections to well-established systems, together with our mechanized proofs, serve to support our claim that this approach is theoretically well-grounded.
+For programs with unfilled type holes, casts are inserted as in gradual type theory (GTT) \cite{DBLP:conf/snapl/SiekVCB15} and, uniquely, evaluation proceeds around failed casts as it does around expression holes.
 
-We are integrating this approach into the \Hazel programming environment being developed by \citet{HazelnutSNAPL}.
-\Hazel features a live context inspector that integrates static information with hole closure information and interactively presents recursive hole closures, which arise from incomplete recursive functions.
-The editor component of this implementation defines a language of structured edit actions,
-based on the \Hazelnut structure editor calculus developed by \citet{popl-paper}, that inserts holes automatically to guarantee that
+The primary contribution of this paper is a type-theoretic account of these ideas in the form of a core calculus, \HazelnutLive.
+We observe that expression hole closures are closely related to metavariable closures from contextual modal type theory~(CMTT)~\cite{Nanevski2008}, which, by its Curry-Howard correspondence with contextual modal logic, provides a logical basis for reasoning about and operating on hole closures.
+These connections to well-established systems (GTT and CMTT), together with mechanized proofs of the metatheory of \HazelnutLive, serve to support our main claim: that this approach is theoretically well-grounded.
+
+There are many possible ways to present incomplete results and hole closure information to programmers.
+A secondary contribution of this paper is one  proof-of-concept user interface, which has been implemented into the \Hazel programming environment being developed by \citet{HazelnutSNAPL}.
+In particular, we describe \Hazel's novel live context inspector, which  combines static type information with hole closure information and interactively presents nested hole closures.
+We make no strong empirical claims about this particular user interface; we evaluate it with several suggestive example programming tasks where this interface presents information that would not otherwise be available and that we conjecture would be useful when teaching functional programming.
+
+The editor component of \Hazel is organized around a language of structured edit actions, 
+based on the \Hazelnut structure editor calculus developed by \citet{popl-paper}, that insert holes automatically to guarantee that
 every editor state has some, possibly incomplete, type. 
 The type safety invariant that we establish then guarantees that every editor state has dynamic meaning. Taken together, the result is a full solution to the gap problem, i.e. a proof-of-concept
-live functional programming environment that provides rich static and dynamic feedback that never suffers from temporal or perceptive gaps.
+live functional programming environment that continuously provides rich static and dynamic feedback.
 
@@
 
-We begin in Sec.~\ref{sec:examples} by detailing the approach informally, with several example programming tasks, in the setting of the \Hazel design, which is based closely on well-established existing designs (in particular, on the Elm programming language and the Jupyter user interface).
+We begin in Sec.~\ref{sec:examples} by detailing the approach informally, with several example programming tasks, in the setting of the \Hazel design. 
 % We also qualitatively describe some user interface features that may help manage visual complexity when working with larger programs.
@@
 % The dynamic semantics of \HazelnutLive is a small-step reduction semantics equipped with clean type safety theorems (including, notably, a Progress theorem for incomplete expressions). 
-Sec.~\ref{sec:agda-mechanization} outlines our Agda-based mechanization of \HazelnutLive, which is included in the anonymized supplement.  
-Sec.~\ref{sec:implementation} provides some additional details on the implementation of these ideas into \Hazel, which is based on the \Hazelnut edit action calculus \cite{popl-paper}, and formally states the continuity invariant, i.e.
-the solution to the gap problem, as a corollary of the primary theorems of \Hazelnut and \HazelnutLive.
+Sec.~\ref{sec:agda-mechanization} outlines our Agda-based mechanization of \HazelnutLive, which is included in the supplemental material and is also available from the following URL:
+
+\begin{center}
+\url{https://github.com/hazelgrove/hazelnut-dynamics-agda}
+\end{center}
+
+\noindent
+Sec.~\ref{sec:implementation} states the continuity invariant, which formally solves the gap problem, as a corollary of the primary theorems of \Hazelnut and \HazelnutLive.
+It also provides some additional details on the implementation of \Hazel.
+A snapshot of the implementation is included in the supplemental material.
+An online version of the ongoing implementation of \Hazel is available from \url{hazel.org}, and the source code is available from the following URL:
+
+\begin{center}
+\url{https://github.com/hazelgrove/hazel}
+\end{center}
 
@@
 
-Sec.~\ref{sec:relatedWork} describes related work in detail and simultaneously discusses limitations and directions for future work. Sec.~\ref{sec:discussion} briefly concludes.
+Sec.~\ref{sec:relatedWork} describes related work in detail and simultaneously discusses limitations and a number of directions for future work. Sec.~\ref{sec:discussion} briefly concludes.
 
@@
 \else
-The anonymous supplement includes, in addition to the Agda mechanization, an appendix that 
+The extended version of the paper, which is included in the supplemental material and available from \url{https://arxiv.org/abs/1805.00155}, includes an appendix that 
 \fi
-(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for concision; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions.
+(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for the sake of space; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions.
 % These are not necessary to communicate the fundamental ideas of the paper, but may be of interest to some readers.

### 2018-10-26 intro.tex
replaced arxiv url in text with an arxiv citation + added cimini and siek 2016 citation to intro

@@
 \else
-The extended version of the paper, which is included in the supplemental material and available from \url{https://arxiv.org/abs/1805.00155}, includes an appendix that 
+The extended version of the paper, which is included in the supplemental material and available online \cite{2018arXiv180500155O}, includes an appendix that 
 \fi
-(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for the sake of space; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions.
+(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for the sake of space; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief discussion on  defining other extensions by following the approach of \citet{DBLP:conf/popl/CiminiS16}.
 % These are not necessary to communicate the fundamental ideas of the paper, but may be of interest to some readers.

### 2018-10-30 abstract.tex
some more wording tweaks to abstract

@@
 of type errors. Consequently,    
-live feedback presented to the programmer exhibits temporal or perceptive gaps, i.e. it flickers out or it goes stale whenever the program becomes incomplete. 
+live feedback presented to the programmer exhibits temporal or perceptive gaps. 
 % In some cases, these gaps persist over a substantial length of time.
@@
 avoids the need to restart evaluation after edits that amount to hole filling. 
-Formally, the semantics borrows machinery from both gradual type theory (to handle unfillable 
-type holes) and contextual modal type theory (which supplies the
+Formally, the semantics borrows machinery from both gradual type theory (which supplies the basis for handling unfilled 
+type holes) and contextual modal type theory (which supplies a
 logical basis for hole closures), combining these and developing additional machinery necessary 

### 2018-10-30 intro.tex
tweaks to wording in intro + make claims more clear -- Fixes #13

--- 2018-10-26 intro.tex
@@
 Programmers typically shift between program editing and program evaluation many times before converging upon a program that behaves as intended. 
-Live programming environments aim to support this workflow by granularly interleaving editing and evaluation so as to   
+Live programming environments aim to granularly interleave editing and evaluation so as to   
 narrow what \citet{burckhardt2013s} call the ``temporal and perceptive gap'' between these activities.
@@
 and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding as a performance art.
-Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support editing run-time state, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
+Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support inspecting run-time state in the editor, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
 % Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13},
@@
 programs that are syntactically well-formed and free of type and binding errors.
-A program editor, however, frequently encounters incomplete, and therefore conventionally meaningless, editor states.
+Programming environments, however, often encounter incomplete, and therefore conventionally meaningless, editor states.
 As a result, live feedback either ``flickers out'', creating a temporal gap, or it ``goes stale'', i.e.
-it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not accurately reflect what the programmer is editing.
+it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not accurately reflect the editor state.
 
@@
 %
-Novice programmers, of course, make more mistakes \cite{mccauley2008debugging,fitzgerald2008debugging}.
+This is particularly problematic for novice programmers, who make more mistakes \cite{mccauley2008debugging,fitzgerald2008debugging}.
 %
@@
 Throughout the process of addressing these errors, which can sometimes span many days, the program text remains formally meaningless.
-Overall, about 40\% of edits performed by Java programmers using Eclipse left the program text malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.
+About 40\% of edits performed by Java programmers using Eclipse left the program text malformed \cite{popl-paper} and some additional number, which could not be determined from the data gathered by \citet{6883030}, were well-formed but ill-typed.
 
@@
 %}
-In recognition of this ``{gap problem}''---that incomplete programs are formally meaningless---\citet{popl-paper} developed a static semantics (i.e. a type system) for incomplete 
+In recognition of this ``{gap problem}''---that incomplete programs are formally meaningless---\citet{popl-paper} describe a static semantics (i.e. a type system) for incomplete 
 functional programs, modeling them formally as typed expressions with \emph{holes} in 
@@
 For example, GHC Haskell supports the notation \li{_u} for empty holes, where \li{u} is an optional hole name \cite{GHCHoles}.
-When the editor state is a tree or graph structure, i.e.
+When the editor state is instead a tree or graph structure, i.e.
 in a structure editor, the editor typically inserts explicitly represented holes fully automatically \cite{popl-paper}; we say more about structure editors in Sec.~\ref{sec:implementation}.
@@
 %
-For the purposes of live programming, however, a static semantics does not suffice---%
-we also need a corresponding dynamic semantics that specifies how to evaluate expressions with holes. %This paper addresses this need by developing
+A static semantics for incomplete programs is useful, but for the purposes of live programming, it does not suffice---%
+we also need a corresponding dynamic semantics that specifies how to evaluate expressions with holes. That is the focus of this paper. %This paper addresses this need by developing
 % a {dynamic semantics} for incomplete functional programs, 
@@
 within a live programming environment because 
-(1)~it provides no information about the behavior of the remainder of program, 
+(1)~it provides no information about the behavior of the remainder of the program, 
 parts of which may not depend on the missing or erroneous expression (e.g. subsequent cells in a lab notebook, or tests involving other components of the program);  
@@
 (typically only a stack trace);  and
-(3)~it provides no means by which to resume evaluation after filling a hole.
+(3)~it provides no means by which to resume evaluation after hole filling.
 
 Furthermore, exceptions can appear only in expressions, but we might also like to be able to evaluate programs that have type holes.
-Again, existing approaches do not support this situation well---GHC supports type holes, but compilation fails if type inference cannot automatically fill them \cite{GHCHoles}.
+Again, existing approaches do not support this situation well---GHC supports type holes, but compilation fails if type inference cannot automatically fill them (such as when a variable is used at multiple types) \cite{GHCHoles}.
 The static semantics developed by \citet{popl-paper} provides more hope because it derives the machinery for 
-reasoning statically about type holes from gradual type theory, 
+reasoning about type holes from gradual type theory, 
 identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}.
@@
 gradual type theory, 
-which inserts dynamic casts as necessary to maintain type safety. 
+which inserts dynamic casts as necessary. 
 %However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
-However, when a cast fails, evaluation again stops with 
+The only problem is that when a cast fails, evaluation stops with 
 an exception, again leaving the live programming environment unable to provide rich, continuous feedback about the behavior of the remainder of the program.
@@
 
-The primary contribution of this paper is a type-theoretic account of these ideas in the form of a core calculus, \HazelnutLive.
+The primary contribution of this paper is a simple type-theoretic account of this approach in the form of a core calculus, \HazelnutLive.
 We observe that expression hole closures are closely related to metavariable closures from contextual modal type theory~(CMTT)~\cite{Nanevski2008}, which, by its Curry-Howard correspondence with contextual modal logic, provides a logical basis for reasoning about and operating on hole closures.
-These connections to well-established systems (GTT and CMTT), together with mechanized proofs of the metatheory of \HazelnutLive, serve to support our main claim: that this approach is theoretically well-grounded.
+These connections to well-established systems (GTT and CMTT), together with mechanized proofs of the metatheory of \HazelnutLive, serve to support our main claim: that this approach to live programming is theoretically well-grounded.
 
@@
 In particular, we describe \Hazel's novel live context inspector, which  combines static type information with hole closure information and interactively presents nested hole closures.
-We make no strong empirical claims about this particular user interface; we evaluate it with several suggestive example programming tasks where this interface presents information that would not otherwise be available and that we conjecture would be useful when teaching functional programming.
+We make no strong claims about this particular user interface; we evaluate it only with several suggestive example programming tasks where this interface presents information that would not otherwise be available and that we conjecture would be useful when teaching functional programming.
 
@@
 every editor state has some, possibly incomplete, type. 
-The type safety invariant that we establish then guarantees that every editor state has dynamic meaning. Taken together, the result is a full solution to the gap problem, i.e. a proof-of-concept
+The type safety invariant that we establish then guarantees that every editor state has dynamic meaning.
+Taken together, the result is an end-to-end solution to the gap problem, i.e.
+a proof-of-concept
 live functional programming environment that continuously provides rich static and dynamic feedback.
@@
 \else
-The extended version of the paper, which is included in the supplemental material and available from \url{https://arxiv.org/abs/1805.00155}, includes an appendix that 
+The extended version of the paper, which is available in the ArXiV \cite{2018arXiv180500155O}, includes an appendix that 
 \fi
-(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for the sake of space; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions.
+(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for the sake of space; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief discussion on defining other extensions (in part by by following the ``gradualization'' approach of \citet{DBLP:conf/popl/CiminiS16}).
 % These are not necessary to communicate the fundamental ideas of the paper, but may be of interest to some readers.

### 2018-10-30 examples-1.tex
Minor clarifications to wording in Example 1, typo fixes

@@
 
-Thusfar in the body of \lismall{weighted_average}, the teacher has decomposed the function argument into variables by record destructuring, then multiplied the homework grade, \li{hw}, by \li{30.0} and finally inserted the \li{+} operator.
-The cursor now sits at an \emph{empty hole}, as indicated by the vertical bar and the green background.
-Each hole has a unique name (generated automatically in \Hazel), here simply \li{1}.
-%In this case, there is only one hole in the program, but in general, there may be many such holes.
-
-Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback.
-In response, the programmer might encode a hole by raising an exception. This would cause typechecking to succeed. 
+Thus far in the body of \lismall{weighted_average}, the teacher has decomposed the function argument into variables by record destructuring, then multiplied the homework grade, \li{hw}, by \li{30.0} and finally inserted the \li{+} operator.
+In a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback.
+In response, the programmer might temporarily raise an exception. This would cause typechecking to succeed. 
 However, 
 evaluation would proceed only as far as the first \li{map} iteration, 
-which would call into \li{weighted_average} and then fail when attempting to evaluate the hole.%
+which would call into \li{weighted_average} and then fail when the exception is raised.%
 \footnote{In a lazy language, like Haskell, the result would be much the same because the environment forces the result for printing.}
 % While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.
+Hazel instead inserts an \emph{empty hole} at the cursor, as indicated by the vertical bar and the green background.
+Each hole has a unique name (generated automatically in \Hazel), here simply \li{1}.
 
-\Hazel{} does not take this ``exceptional'' interpretation of holes.
+When running the program, \Hazel{} does not take an ``exceptional'' approach to holes.
 Instead, evaluation continues past the hole, treating it as an opaque expression of the appropriate type.
@@
 Evaluation cannot proceed any further because holes appear as addends.
-We say that each of these addition expressions is an \emph{indeterminate} sub-expression, and the result as a hole is therefore also indeterminate, because it is not yet a value, nor can it take a step due to holes in elimination positions.
+We say that each of these addition expressions is an \emph{indeterminate} sub-expression, and the result as a whole is therefore also indeterminate, because it is not yet a value, nor can it take a step due to holes in elimination positions.
 
@@
 The programmer can directly select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the purple outline in Fig.~\ref{fig:grades-cell-mockup}), or cycle through all of the closures for the hole at the cursor by clicking the arrows at the bottom of the context inspector, e.g.
-$\blacktriangleright$, or using hotkeys.
+$\blacktriangleright$, or using corresponding hotkeys.
 

### 2018-10-30 examples-2.tex
Minor clarifications to wording in Example 3 and added a paragraph about how non-empty holes around type-correct expressions might have the same benefits

@@
 as shown in Fig.~\ref{fig:qsort-type-error}. In Sec.~\ref{sec:resumption}, we discuss
-how the programming environment might avoid restarting evaluation after such edits, but for small examples like this, restarting works fine.
+how the programming environment could avoid restarting evaluation after such edits.
 
@@
 and far away from the error) 
-whose dynamic behavior may be of interest. Evaluation can also assist the programmer in understanding the type error by presenting concrete values \cite{Seidel2016}.
+whose dynamic behavior may be of interest. Concrete results can also help the programmer understand the implications of the type error \cite{Seidel2016}.
 % After all,
@@
 is to semantically internalize the ``red outline'' around
-type inconsistencies, representing it as a \emph{non-empty hole}.
+a type-inconsistent expression as a \emph{non-empty hole} around that expression.
 Evaluation safely proceeds past a non-empty hole just as if it were an empty hole.
@@
 They simply have not been combined correctly.
+
+Although \Hazel inserts non-empty holes automatically, the earlier work on \Hazelnut allowed the programmer to explicitly insert non-empty holes.
+It may be that these are useful even when there is not a type inconsistency because the non-empty hole defers elimination of the enclosed expression and causes hole closure information to be tracked.
+For example, if we repaired the program in Fig.~\ref{fig:qsort-type-error} by replacing the erroneous variable \li{pivot} with \li{[pivot]} and then inserted a non-empty hole around this expression, the result would show all of the list concatenation operations that would be performed without actually performing them, producing a result much like that in Fig.~\ref{fig:qsort-type-error}.
+This provides another way to investigate the recursive structure of the \li{qsort} function.
 

### 2018-10-30 examples-2.tex
Added explanation of why list concat was not further reduced

@@
 \centering
-\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/qsort-type-error-inset.png}
+\includegraphics[width=0.95\textwidth,interpolate=false,valign=t]{images/qsort-type-error-inset.png}
 % \includegraphics[width=0.28\textwidth,interpolate=false,valign=t]{images/type-inconsistency.png}
@@
 \label{fig:qsort-type-error}
-\vspace{-4px}
+\vspace{-3px}
 \end{figure}
@@
 as shown in Fig.~\ref{fig:qsort-type-error}. In Sec.~\ref{sec:resumption}, we discuss
-how the programming environment could avoid restarting evaluation after such edits.
+how the programming environment could avoid restarting evaluation after such edits by using the values stored in the hole closures.
 
@@
 
+We are treating \li{@} as a primitive operator in this example. If it were defined as a function, then
+it would be possible to further reduce the example by proceeding into the function body. This is likely unhelpful for
+functions other than those that programmer is actively working on. Our semantics \emph{allows} beta reduction when the argument is a hole, but it does not \emph{require} it.
+
 Although \Hazel inserts non-empty holes automatically, the earlier work on \Hazelnut allowed the programmer to explicitly insert non-empty holes.
@@
 For example, if we repaired the program in Fig.~\ref{fig:qsort-type-error} by replacing the erroneous variable \li{pivot} with \li{[pivot]} and then inserted a non-empty hole around this expression, the result would show all of the list concatenation operations that would be performed without actually performing them, producing a result much like that in Fig.~\ref{fig:qsort-type-error}.
-This provides another way to investigate the recursive structure of the \li{qsort} function.
+This provides another way to explore the recursive structure of the \li{qsort} function.
 

### 2018-10-30 examples-3.tex
Minor clarifications to wording in Example 4

@@
 \label{sec:dynamic-type-errors}
-
-% \begin{subfigure}[t]{\textwidth}
-\begin{figure}
-\centering
-\includegraphics[width=0.71\textwidth,interpolate=false,valign=t]{images/cast-errors.png}
-% \includegraphics[width=0.28\textwidth,interpolate=false,valign=t]{images/type-inconsistency.png}
-\caption{Example 4: Type Holes and Dynamic Type Errors}
-\label{fig:cast-errors}
-\vspace{-6px}
-\end{figure}
-% \end{subfigure}
 
@@
 % programs with type holes. 
-Indeed, the purpose of gradual typing is to be able to run programs that are not yet sufficiently annotated with types by inserting \emph{casts} only where necessary.
+Gradual type systems can run programs that are not yet sufficiently annotated with types by inserting \emph{casts}  where necessary.
+We take the same well-studied approach in \Hazel.
 As such, let us consider only a small synthetic example to demonstrate what is unique to our approach.
@@
 other uses of \li{x} as ill-typed (see \cite{DBLP:journals/jfp/ChenE18}).
+
+% \begin{subfigure}[t]{\textwidth}
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,interpolate=false,valign=t]{images/cast-errors.png}
+% \includegraphics[width=0.28\textwidth,interpolate=false,valign=t]{images/type-inconsistency.png}
+\caption{Example 4: Type Holes and Dynamic Type Errors}
+\label{fig:cast-errors}
+\vspace{-6px}
+\end{figure}
+% \end{subfigure}
 

### 2018-10-30 examples.tex
Better disclaimer about status of the implementation at the beginning of Sec 3

@@
 %% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
-\newcommand{\examplesSec}{Live Functional Programming with Typed Holes in Hazel}
+\newcommand{\examplesSec}{Live Programming in Hazel}
 \section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
@@
 
-This section gives an example-driven overview of our  approach as implemented in  \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL}.
-The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with results appearing below cells containing code, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity.
+Let us start with an example-driven overview of this paper's approach in \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL}.
+The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with a result appearing below each cell that contains an expression, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity.
 \Hazel is intended initially for use by students and instructors in introductory functional programming courses (where \Elm~ has been successful \cite{DBLP:journals/corr/abs-1805-05125}).
 For the sake of 
-exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of some ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel as of this writing, e.g.
-pattern matching in function arguments, list  notation, and record labels (currently there are only tuples).
-These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented.
+exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of  ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel (which, as of this writing, implements little more than the language features described in Sec.~\ref{sec:calculus} and the appendix).
+These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented and all of the computations can be expressed using standard ``encoding tricks'' -- the implementations are included in the supplement.
 

### 2018-10-30 calculus.tex
minor edit to 3.0

@@
 precise by specifying a core calculus, which we call \HazelnutLive, and
-developing its metatheory.
+characterizing its metatheory.
 
@@
 specification of \Hazelnut (for the editor component) and \HazelnutLive
-(for the semantics), and we can formally state a continuity invariant for a
+(for the evaluator), and we can formally state a continuity invariant for a
 putative combined calculus (see Sec.~\ref{sec:implementation}).

### 2018-10-31 calculus.tex
small clarifications through sec 3, addressing R2's confusion about matching relation, and more editor states being meaningful; also changed the description of the mechanization to be consistent with our final artifact

@@
 %
-We start with synthesis for the programmer's ``top level'' external
+We start with synthesis for the programmer's top level external
 expression.
@@
 %
-Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.
-%
-To avoid the need for two separate rules, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.\footnote{A system supporting ML-style type reconstruction \cite{damas1982principal} might also include a synthetic rule for unannotated lambdas, e.g.
+Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.\footnote{A system supporting ML-style type reconstruction \cite{damas1982principal} might also include a synthetic rule for unannotated lambdas, e.g.
 as outlined by \citet{DBLP:conf/icfp/DunfieldK13}, but we stick to this simpler ``Scala-style'' local type inference scheme in this paper \cite{Pierce:2000ve,Odersky:2001lb}.}
+%
+To avoid the need for separate rules for these two cases, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
 %
@@
 
-The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.
+The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matching relation.
 
@@
 
-The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
+The cast form in \HazelnutLive is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
 %
@@
 The latter is more straightforward: Rule~\rulename{ESAsc}
-in \Figref{fig:elabSyn} inserts a cast from the assigned type to the
+in \Figref{fig:expandSyn} inserts a cast from the assigned type to the
 ascribed type.
@@
 %
-We include ascription for exposition purposes---this form is derivable
+We include ascription for expository purposes---this form is derivable
 by using application together with the half-annotated identity, $e
@@
 %
-To understand the rule, consider the elaboration of external
-expression~$\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$, the example
-discussed above:
+To understand the rule, consider the elaboration of the example
+discussed above, $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$:
 \[
@@
 elaboration, from the type it is assigned by the final premise of the
-rule~(type~$b$ here), to the argument type of the matched arrow type of the
-function expression~(type~$\tehole$ here).
+rule~(type~$b$), to the argument type of the matched arrow type of the
+function expression~(type~$\tehole$).
 
@@
 
-Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the \appendixName for the full definition).
+Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the \appendixName~for the full definition).
 % Appendix~\ref{sec:additional-defns}).
@@
 
-Rules \rulename{ITCastSucceed} and \rulename{ITCastFail} rules just described only operate at ground type.
+% Rules \rulename{ITCastSucceed} and \rulename{ITCastFail} only operate at ground type.
 %
@@
 %
-The need for this case is perhaps not obvious because in the cast calculus,  the only irreducible term of type $\tehole$
-is $\dcasttwo{\dexp}{\htau_1}{\tehole}$, but here there are holes as well.
 
@@
 %
-Hole closures can disappear during evaluation, so we rely on simple weakening
+Hole closures can disappear during evaluation, so we must have structural weakening 
 of $\Delta$.
@@
 including all of the theorems stated above and necessary lemmas.
-We take as assumptions only a few standard properties of general hypothetical judgements, substitution and finite maps.
 %
@@
 %
-Our approach is otherwise standard: we model judgements as
+Our approach is standard: we model judgements as
 inductive datatypes, and rules as dependently typed constructors of these judgements.
 %
-We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names and encode typing
-contexts, hole contexts and finite substitutions using metafunctions.
+We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names, and avoid certain other complications relted to substitution by enforcing the requirement that all bound variables in a term are unique when convenient (this requirement can always be discharged by alpha-variation).
+We encode typing
+contexts and hole contexts using metafunctions.
 To support this encoding choice, we postulate function extensionality (which is independent of Agda's axioms) \cite{awodey2012inductive}.
+We encode finite substitutions as an inductive datatype with a base case representing the identity substitution and an inductive case that records a single substitution.
+Every finite substitution can be represented in this scheme.
+This makes it easier for Agda to see why certain inductions that we perform are well-founded.
 The documentation provided with the mechanization has more details.
@@
 
-The \Hazel implementation demonstrated in Sec.~\ref{sec:examples}
+The \Hazel implementation described in Sec.~\ref{sec:examples}
 includes an unoptimized interpreter, written in OCaml, that implements the semantics as described
-in this section, albeit for an \Elm-like language that goes somewhat beyond the simple
-calculus here. As with many full-scale systems, there is not currently a formal
-specification for the full \Elm{} or \Hazel language, but the \appendixName
+in this section, with some simple extensions. As with many full-scale systems, there is not currently a formal
+specification for the full \Hazel language, but the \appendixName
 discusses how the standard approach for deriving a ``gradualized'' version of a
-language construct provides most of the necessary scaffolding \cite{DBLP:conf/popl/CiminiS16}, and provides some examples.
+language construct provides most of the necessary scaffolding \cite{DBLP:conf/popl/CiminiS16}, and provides some examples (sum types and numbers).
 
@@
 \begin{prop}[Sensibility]
-  \label{thrm:actsafe}
+  \label{thrm:sensibility}
   If $\hsyn{\Gamma}{\removeSel{\zexp}}{\htau}$ and
@@
 \begin{corol}[Continuity]
-  \label{thrm:actsafe}
+  \label{thrm:continuity}
   If $\hsyn{\emptyset}{\removeSel{\zexp}}{\htau}$ and
@@
 
-In settings where the editor does not maintain this Sensibility invariant, our approach still helps to reduce the severity of the gap problem, i.e.
-\emph{more} editor states are meaningful, even if not \emph{all} are.% We can formally state this end-to-end continuity corollary as follows:
+In settings where the editor does not maintain this Sensibility invariant, but where programmers can manually insert holes, our approach still helps to reduce the severity of the gap problem, i.e.
+\emph{more} editor states are dynamically meaningful, even if not \emph{all} of them are.% We can formally state this end-to-end continuity corollary as follows:
 %

### 2018-10-31 calculus.tex
typos

@@
 %
-We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names, and avoid certain other complications relted to substitution by enforcing the requirement that all bound variables in a term are unique when convenient (this requirement can always be discharged by alpha-variation).
+We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names, and avoid certain other complications related to substitution by enforcing the requirement that all bound variables in a term are unique when convenient (this requirement can always be discharged by alpha-variation).
 We encode typing
@@
 We encode finite substitutions as an inductive datatype with a base case representing the identity substitution and an inductive case that records a single substitution.
-Every finite substitution can be represented in this scheme.
+Every finite substitution can be represented this way.
 This makes it easier for Agda to see why certain inductions that we perform are well-founded.
@@
 in this section, with some simple extensions. As with many full-scale systems, there is not currently a formal
-specification for the full \Hazel language, but the \appendixName
+specification for the full \Hazel language, but the \appendixName~
 discusses how the standard approach for deriving a ``gradualized'' version of a

### 2018-10-31 calculus.tex
tweaks to wording in Sec 4, addressing confusion around evaluation order does not matter

@@
 %
-Critically, this property relies on the version of
-the semantics from \Secref{sec:calculus} where evaluation order is
-unspecified (i.e. where we omit the bracketed premises).
-%
-In general, resuming from $\instantiate{\dexp'}{u}{d_2}$ will not
-reduce sub-expressions in the same order as a ``fresh'' left-to-right
-reduction sequence starting from $\instantiate{\dexp'}{u}{d_1}$.
-%
-In other words, this notion of fill-and-resume only works for
-languages where evaluation order ``does not matter''.
-(There are various standard ways to formalize this intuition.
-For the sake of space, we review these in the \appendixName.)
+The key idea is that we can resume evaluation by replaying the substitutions that were recorded in the closures and then taking the necessary ``catch up'' steps that evaluate the hole fillings that now appear.
+%
+The caveat is that resuming from $\instantiate{\dexp'}{u}{d_2}$ will not
+reduce sub-expressions in the same order as a ``fresh'' eager left-to-right
+reduction sequence starting from $\instantiate{\dexp'}{u}{d_1}$ so 
+filling commutes with reduction only for 
+languages where evaluation order ``does not matter'', e.g.
+pure functional languages like \HazelnutLive.\footnote{There are various standard ways to formalize this intuition, e.g.
+by stating a suitable confluence property.
+For the sake of space, we review confluence in the \appendixName.} Languages with non-commutative effects do not enjoy this property.
 % Appendix~\ref{sec:confluence}.)
@@
 In particular, care is needed to handle the situation where a
-now-filled non-empty hole had taken a step in the original evaluation.
+now-filled non-empty hole had taken a step in the original evaluation trace.
 

### 2018-10-31 calculus.tex
added type hole filling as future work

@@
 In any case,
-When the programmer performs one or more edit actions to fill in a hole in the program, a new result must be computed, ideally quickly \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90}.
+When the programmer performs one or more edit actions to fill in an expression hole in the program, a new result must be computed, ideally quickly \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90}.
 Na\"ively, the system would need to compute the result ``from scratch'' on each such edit.

### 2018-10-31 related-work.tex
added type hole filling as future work

@@
 
-This paper defined a dynamic semantics for incomplete functional programs,
-building on the static semantics developed by \citet{popl-paper}. That paper, as well as a subsequent ``vision paper'' introducing \Hazel, suggested as future work  
-a corresponding dynamic semantics for the purposes of live programming without semantic gaps \cite{HazelnutSNAPL}.
-Others have extensively argued the value of assigning meaning even to incomplete and erroneous programs \cite{Bayne:2011:ASD:1985793.1985864}.
-This paper delivers conceptual and type-theoretic details necessary to further advance this  vision.
+This paper builds directly on the static semantics developed by \citet{popl-paper}.
+That paper, as well as a subsequent ``vision paper'' introducing \Hazel \cite{HazelnutSNAPL}, suggested as future work
+a corresponding dynamic semantics for the purposes of live programming without gaps.
+\citet{Bayne:2011:ASD:1985793.1985864} have also extensively argued the value of assigning meaning even to incomplete and erroneous programs.
+This paper delivers conceptual and type-theoretic details necessary to advance these visions.
 
@@
 prior work is the treatment of cast failures like holes, rather than errors.
-Many of the methods developed to make gradual typing more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein \cite{takikawa_et_al:LIPIcs:2015:5215}.
-For example, there has been substantial work on the problem of implementing casts efficiently \cite{herman2010space,siek2010threesomes,garcia2013threesomes}, and on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,DBLP:journals/pacmpl/DevriesePP18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.
+Many of the methods developed to make gradual type systems more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein \cite{takikawa_et_al:LIPIcs:2015:5215}.
+For example, there has been substantial work on the problem of implementing casts efficiently \cite{herman2010space,siek2010threesomes,garcia2013threesomes}.
+There has also been substantial work on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,DBLP:journals/pacmpl/DevriesePP18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.
 Another direction that we leave to future work is the integration of these ideas into dependently typed proof assistants,
@@
 \citet{ocaml-stepper} discuss some unresolved usability issues relevant to single steppers for functional languages.
+
+We did not give type holes unique names nor did we define a commutative type hole filling operation, but we conjecture that it is possible to do so by converting casts that go from the filled hole to an inconsistent type into non-empty expression holes.
+We leave the details to future work.
 

### 2018-10-31 related-work.tex
fixed incorrect characterization of ref(tau) + invariance

@@
 There has also been substantial work on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,DBLP:journals/pacmpl/DevriesePP18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.
-Another direction that we leave to future work is the integration of these ideas into dependently typed proof assistants,
-where evaluation is primarily at the service of equational reasoning \cite{DBLP:journals/corr/abs-1009-2789}.
-Several modern proof assistants, e.g.
-Agda \cite{norell2009dependently,norell:thesis}, Idris \cite{brady2013idris} and Beluga \cite{DBLP:conf/flops/Pientka10} already provide syntactic and static support for holes (see Sec.~\ref{sec:intro}) and interactive, hole-driven development is standard practice.
 
@@
 %
-\citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15} show how to incorporate such features into gradual type theory (by treating $\mathtt{ref}(\tau)$ as invariant with respect to consistency); we expect that this approach would also work for the semantics of Sec.~\ref{sec:calculus}, i.e.
+\citet{DBLP:conf/esop/SiekVCTG15} show how to incorporate such features into gradual type theory; we expect that this approach would also work for the semantics of Sec.~\ref{sec:calculus}, i.e.
 it would conserve the type safety properties established there, with suitable modifications to account for a store.
@@
 
-We did not give type holes unique names nor did we define a commutative type hole filling operation, but we conjecture that it is possible to do so by converting casts that go from the filled hole to an inconsistent type into non-empty expression holes.
+We did not give type holes unique names nor did we define a commutative type hole filling operation, but we conjecture that it is possible to do so by replacing casts that go from the filled hole to an inconsistent type with non-empty expression holes.
 We leave the details to future work.
@@
 using thunks to encapsulate indeterminate sub-expressions) to future work.
+
+\todo{integrate me}Another direction that we leave to future work is the integration of type holes into dependently typed proof assistants, 
+where evaluation is primarily at the service of equational reasoning \cite{DBLP:journals/corr/abs-1009-2789}.
+Several modern proof assistants, e.g.
+Agda \cite{norell2009dependently,norell:thesis}, Idris \cite{brady2013idris} and Beluga \cite{DBLP:conf/flops/Pientka10} already provide syntactic and static support for holes (see Sec.~\ref{sec:intro}) and interactive, expression-hole-driven development is standard practice, but type holes are less well-supported.
 

### 2018-10-31 intro.tex
a few more intro tweaks

@@
 %
-live user interface frameworks \cite{burckhardt2013s};
+live GUI frameworks \cite{burckhardt2013s};
 %
@@
 %
-and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding as a performance art.
-Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support inspecting run-time state in the editor, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
+and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding as performance art \cite{DBLP:journals/programming/ReinRLHP19}.
+Editor-integrated debuggers \cite{mccauley2008debugging} and editors that support inspecting run-time state, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments.
 % Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13},
@@
 %\matt{we are not really modeling a language with user type definitions -- what does the following sentence really add here? also, I just added this scenerio to the list above}
-The problem is particularly pronounced for languages with rich static type systems where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
+The problem is also particularly pronounced for languages with rich static type systems where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program.
 Throughout the process of addressing these errors, which can sometimes span many days, the program text remains formally meaningless.
@@
 When the editor state is instead a tree or graph structure, i.e.
-in a structure editor, the editor typically inserts explicitly represented holes fully automatically \cite{popl-paper}; we say more about structure editors in Sec.~\ref{sec:implementation}.
+in a structure editor, the editor inserts explicitly represented holes fully automatically \cite{popl-paper}; we say more about structure editors in Sec.~\ref{sec:implementation}.
 

### 2018-10-31 discussion.tex
clarified discussion of related work, discussed some additional related work, misc clarifications

@@
 
-\hfill{}--- Gerald Weinberg, The Psychology of Computer Programming (1971)
+\hfill{}--- Gerald Weinberg, The Psychology of Computer Programming \cite{weinberg1971psychology}
 \end{quote}
@@
 
-\begin{comment}
 \newcommand{\acksSection}{Acknowledgments}
@@
 
-The authors would like to thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues,
+We thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, Conor McBride, 
 the participants of the PL reading group at UChicago,
-and the referees and participants of the LIVE 2017 and Off the Beaten Track (OBT)
-2017 workshops for their insights and feedback on various iterations of this work.
+the participants of the LIVE 2017 and Off the Beaten Track (OBT)
+2017 workshops, and the anonymous referees for their insights and feedback on various iterations of this work.
 %
@@
 Facebook, from the National Science Foundation under grant
-number CCF-1619282, and from AFRL and DARPA under agreement \#FA8750-16-2-0042. 
+numbers CCF-1619282, SHF-1814900 and SHF-1817145, and from AFRL and DARPA under agreement \#FA8750-16-2-0042. 
+The U.S. Government is authorized to reproduce and distribute reprints for 
+Governmental purposes notwithstanding any copyright notation thereon.
 %
@@
 in this material are those of the authors and do not necessarily
-reflect the views of Facebook, NSF, DARPA or AFRL.
+reflect the views of Facebook, NSF, DARPA, AFRL or the U.S. Government. 
 %
-\end{comment}
-
 \clearpage

### 2018-10-31 related-work.tex
clarified discussion of related work, discussed some additional related work, misc clarifications

@@
 However, the commutativity property we establish in Sec.~\ref{sec:resumption} will not hold for a language that supports non-commutative effects.
-We leave to future work the task of defining more restricted special cases of the fill-and-resume operation that respects a suitable commutativity property in effectful settings (perhaps by checkpointing).
+We leave to future work the task of defining more restricted special cases of the fill-and-resume operation that respects a suitable commutativity property in effectful settings (perhaps by checkpointing or by using a type and effect system to granularly determine where re-evaluation is needed \cite{burckhardt2013s}).
 
@@
 
-We did not give type holes unique names nor did we define a commutative type hole filling operation, but we conjecture that it is possible to do so by replacing casts that go from the filled hole to an inconsistent type with non-empty expression holes.
-We leave the details to future work.
+We did not give type holes unique names nor did we define a commutative type hole filling operation, but we conjecture that it is possible to do so by using non-empty expression holes to replace casts that go from the filled hole to an inconsistent type.
 
-\parahead{Contextual Modal Type Theory}
+\parahead{Metavariables in Type Theory}
 
 The other major pillar of related work is the work on contextual modal type theory (CMTT) by \citet{Nanevski2008}, which we also discussed at length throughout the paper.
-To reiterate, there is a close relationship between expression holes in this paper and metavariables in CMTT.
-Hole contexts correspond to modal contexts.
+To reiterate, there is a close relationship between expression holes in this paper and metavariables in CMTT: both are associated with a type and a typing context.
 Empty hole closures 
-relate to the concept of a {metavariable closure} in CMTT, which consists
+relate to the concept of a {metavariable closure} (delayed substitution) in CMTT, which consists
 of a metavariable paired with a substitution for all of the variables in the
-typing context associated with that metavariable. Hole filling relates to contextual substitution.
+typing context associated with that metavariable. 
+Empty expression hole filling relates to contextual substitution.
 
@@
 %
-These connections with gradual typing and CMTT are instructive, but our contributions do not neatly fall out from this prior work.
+These connections with gradual typing and CMTT are encouraging, but our contributions do not immediately fall out from this prior work.
 %
 The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle.
-The paper therefore proved only a subject reduction property (which is closely related to type preservation). 
-This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. 
-In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables (and, of course, with casts, i.e.
-\emph{gradual contextual modal type theory}).
+The paper therefore proved only a subject reduction property (which is closely related to type preservation) and sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion, %,
+which has been studied by \citet{DBLP:journals/iandc/Groote02}. 
+% Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property.
+\citet{DBLP:conf/ppdp/PientkaD08} more directly defines a dynamic semantics for an extension of CMTT.
+In both cases, the issue is that these systems can evaluate only closed terms, while we need to consider terms with free metavariables.
 
-\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion, %, 
-which has been studied by \citet{DBLP:journals/iandc/Groote02}. 
-Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property. 
-These issues are not relevant in \HazelnutLive because, as in gradual type theory, type holes admit non-termination \cite{Siek06a}.  
-In any case, under this interpretation an analogous problem arises---metavariables become variables of a function type, so we cannot rely on the standard notion of progress on closed terms.
+There has been much formal work on reduction of open terms.
+In particular, the typical approach is to define the \emph{weak head normal forms} \cite{barendregt84:_lambda_calculus,DBLP:journals/corr/abs-1009-2789,Abramsky:1990vv}, and a weak head normalization procedure.
+This is useful when using reduction to perform optimizations throughout a program, e.g.
+when using supercompilation-by-evaluation \cite{DBLP:conf/haskell/BolingbrokeJ10} or symbolic evaluation \cite{King:1976,SurveySymExec-CSUR18}, or when using evaluation in the service of equational reasoning as in dependently typed proof assistants.
+Notably, \citet{DBLP:journals/corr/abs-1009-2789} considered weak head normalization for CMTT, which forms the basis for equational reasoning in the Beluga proof assistant \cite{DBLP:conf/flops/Pientka10}.
+There are two points of note here.
+First, the addition of type holes allow us to express general recursion \cite{Siek06a}, so we cannot rely on normalization arguments and instead need a more conventional dynamic semantics with a progress theorem.
+Second, we do not want to evaluate under arbitrary binders but rather only around holes when they would otherwise have been evaluated \cite{DBLP:conf/birthday/BlancLM05}.
+As such, we do not need to consider terms with free variables, and can restrict our interest to only those with free metavariables.
+These considerations together lead us to the indeterminate forms and to the progress theorem that we established.
 
-While there has been some formal work on reduction of open terms, leading to programs in a \emph{weak head normal form} \cite{barendregt84:_lambda_calculus,DBLP:journals/corr/abs-1009-2789,Abramsky:1990vv}, there is no clear progress theorem in this setting: in the setting of an evaluator for a programming language, we do not want to evaluate under arbitrary binders but rather only around holes when they would otherwise have been evaluated \cite{DBLP:conf/birthday/BlancLM05}.
-Another important distinction is that simple evaluation of open terms will not support hole closure tracking, which is an essential component of our approach.
-CMTT makes the appropriate distinctions between variables and metavariables, and our notion of an indeterminate form allows us to prove a progress theorem of familiar form (where there are no free variables).
-Weak head normal forms are more useful when using evaluation to perform optimizations throughout a program, e.g.
-when using supercompilation-by-evaluation \cite{DBLP:conf/haskell/BolingbrokeJ10} or symbolic evaluation \cite{King:1976,SurveySymExec-CSUR18}, or when using evaluation in the service of equational reasoning (e.g.
-in Beluga \cite{DBLP:journals/corr/abs-1009-2789}).
+There are various other systems similar in many ways to CMTT in 
+that they consider the problem of reasoning about metavariables. 
+For example, McBride's OLEG is another system for reasoning contextually about metavariables
+\cite{DBLP:phd/ethos/McBride00}, and it is the conceptual basis of certain hole refinement features in Idris \cite{brady2013idris}.
+\citet{DBLP:conf/csl/GeuversJ02} discuss similar ideas, again in the setting of hole refinement in proof assistants.
+However, these systems do not account for substitutions around metavariables.
+The systems underlying TypeLab \cite{Strecker:98a} and Alf \cite{magnusson1994implementation} do record substitutions around metavariables.
+These can be considered predecessors of CMTT, which is unique in that it has a clear Curry-Howard interpretation.
+The discussion above applies similarly to these systems.
 
-It is also worth emphasizing that we use the machinery 
+We use the machinery 
 borrowed from CMTT only extralinguistically.
@@
 %
-A hole filling can be interpreted as a value of contextual modal
-type, and the act of hole filling followed by evaluation to the next
-possibly-indeterminate edit state as evaluation under the binder of a
-suitable $\mathsf{letbox}$ construct, which is enabled by the dynamic
-semantics in Sec.~\ref{sec:calculus}.
+A program with expression holes can be interpretted as being bound under a number of these $\mathsf{letbox}$ constructs, i.e.
+CMTT serves as a ``{development calculus}'' \cite{DBLP:phd/ethos/McBride00}.
+Live programming corresponds to reduction under the metavariable binders interleaved with elimination steps.
 %
-This interpretation could be generalized to allow us to \emph{compute} a hole filling, rather
-than simply stating it, by specifying a non-trivial expression, rather than just a value, of
-modal type.
+This suggests the possibility of \emph{computing} hole fillings by specifying a non-trivial expression of the appropriate contextual 
+modal type in this development calculus.
 %
-This could, in turn, serve as the basis for a \emph{live} computational hole
-refinement system, extending the capabilities of purely static hole
-refinement systems like those available in some proof assistants,
-e.g. the elaborator reflection system of Idris
-\cite{brady2013idris,DBLP:conf/icfp/ChristiansenB16} and the refinement system of Beluga
+This could, in turn, serve as the basis for a computational hole
+refinement system that supports efficient live programming, extending the capabilities of purely static hole
+refinement systems like those available in many languages,
+e.g. the editor-integrated system of Idris
+\cite{brady2013idris,Korkut:2018:ETE:3240719.3241791} and the hole refinement system of Beluga
 \cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
 %
-Each applied hole filling serves as a boundary between dynamic
-\emph{edit stages}.
+Each applied hole filling can be interpretted as inducing a new dynamic
+\emph{edit stage}.
 %
-This contextual modal interpretation of live staged hole refinement
-mirrors the modal interpretation of staged argument evaluation
-\cite{Davies:2001op}. 
+This contextual modal interpretation of live hole refinement
+nicely mirrors the modal interpretation of staging and partial evaluation
+\cite{Davies:2001op}, with indeterminate forms corresponding to the residual programs that arise when performing partial evaluation. 
 %
-This parallel neatly explains
-the difference between our dynamics and existing work on staging and
-partial evaluation---
-existing staging and partial evaluation systems are focused on partial evaluation with respect to an input that sits outside of a
-function \cite{Jones:1993uq}, whereas holes have  contexts.
+The difference is that staging and partial evaluation systems evaluate around an input that sits outside of a
+function \cite{Jones:1993uq}, whereas holes are contextual, i.e. they are located inside the program.
 
-There are various other systems similar in certain ways to CMTT in 
-that they consider the problem of reasoning about metavariables. 
-For example, McBride's OLEG is another system for reasoning contextually about metavariables
-\cite{DBLP:phd/ethos/McBride00}, and it is the conceptual basis of certain hole refinement features in Idris \cite{brady2013idris}.
-\citet{DBLP:conf/csl/GeuversJ02} discuss similar ideas.
-CMTT is unique relative to prior approaches in that it has a clear Curry-Howard interpretation and that it was designed with commutativity properties in mind.
 %
-This paper bears out that the basic constructs of CMTT are ideally suited for a purpose previously not considered: live functional programming with typed expression holes, with commutative edits.
-
-Work on explicit substitutions confronts problems similar to the problem of tracking the environment around a closure using delayed substitutions \cite{Abadi:1991fr,levy1999explicit,Abadi:1990ys}.
-\citet{DBLP:journals/corr/abs-1009-2789} have developed a theory of explicit substitutions for CMTT, which, following other work on explicit substitutions and environmental abstract machines \cite{DBLP:journals/tcs/Curien91}, would be useful in implementing \HazelnutLive more efficiently by delaying both standard and contextual substitutions until needed during evaluation.
+There is also more general work on explicit substitutions, which records all  substitutions, not just substitutions around metavariables, explicitly \cite{Abadi:1991fr,levy1999explicit,Abadi:1990ys}.
+\citet{DBLP:journals/corr/abs-1009-2789} have developed a theory of explicit substitutions for CMTT, which, following other work on explicit substitutions and environmental abstract machines \cite{DBLP:journals/tcs/Curien91}, could be useful in implementing \HazelnutLive more efficiently by delaying both standard and contextual substitutions until needed during evaluation.
 We leave this and other questions of fast implementation (e.g.
 using thunks to encapsulate indeterminate sub-expressions) to future work.
-
-\todo{integrate me}Another direction that we leave to future work is the integration of type holes into dependently typed proof assistants, 
-where evaluation is primarily at the service of equational reasoning \cite{DBLP:journals/corr/abs-1009-2789}.
-Several modern proof assistants, e.g.
-Agda \cite{norell2009dependently,norell:thesis}, Idris \cite{brady2013idris} and Beluga \cite{DBLP:conf/flops/Pientka10} already provide syntactic and static support for holes (see Sec.~\ref{sec:intro}) and interactive, expression-hole-driven development is standard practice, but type holes are less well-supported.
 
@@
 Lamdu is a structure editor for a typed functional language that displays variable values inline based on recorded execution traces (see above) \cite{lamdu}.
-However, they do not generally support execution of incomplete programs, which was our focus in this paper.
+However, Lamdu takes the exceptional approach to holes.
 Scratch will execute a program with holes by simply skipping over incomplete commands, but this is a limited protocol because rerunning the program after filling the hole may produce a result unrelated to initial result.
-Though in some situations, skipping over problematic commands has been observed to work surprisingly well \cite{DBLP:conf/dac/Rinard12}, this paper contributes a semantically sound approach (Sec.~\ref{sec:resumption}).
+Though in some situations, skipping over problematic commands has been observed to work surprisingly well \cite{DBLP:conf/dac/Rinard12}, this paper focuses on a sound approach.
+
+\parahead{Scaling Challenges}
 
@@
 that demonstrates (1) a comprehensive solution to the gap problem, as described in Sec.~\ref{sec:implementation}; and 
-(2) one possible user interface for presenting hole closures, including recursive hole closures, to the programmer (Fig.~\ref{fig:qsort-sidebars}).
+(2) one possible user interface for presenting hole closures, including deeply nested hole closures, to the programmer.
+We leave detailed empirical evaluation to future work, and note that the purpose of defining a formal semantics is to provide a foundation for a variety of user interface experiments.
 
-For larger programs, there are some limitations that call for further UI refinement. 
+For larger programs, there are some limitations that will certainly require further UI refinement. 
 Indeterminate results can get to be quite large, particularly when a hole or failed cast
 appears in guard position as in Fig.~\ref{fig:cast-errors}. 
-Although the pretty printer elides irrelevant branches, 
-the semantics in Sec.~\ref{sec:calculus} would also allow for evaluation to pause when such
+One approach is for the pretty printer can elide irrelevant branches. The semantics in Sec.~\ref{sec:calculus} would also allow for evaluation to pause when such
 a form is produced, continuing only if requested.
@@
 
-Finally, it is not always helpful to continue evaluation as far as possible.
-In particular, when there is a hole or cast failure in the argument position
-of a function application, it may often be more helpful to stop before 
-beta reduction, to avoid exposing the internals of the applied function to 
-the client. The semantics in Sec.~\ref{sec:calculus} leaves evaluation order
-undefined, so various heuristic approaches are possible (e.g. we might wish avoid beta reduction  for functions from imported libraries).
+Finally, we have not yet investigated nor attempted to optimize the memory and performance overhead of tracking hole closures. We note simply that it is often the case that 
+programmers use small example inputs during initial development, providing larger inputs
+only after the program is complete. There is no run-time overhead when there are no holes remaining.
+
+% Finally, it is not always helpful to continue evaluation as far as possible.
+% In particular, when there is a hole or cast failure in the argument position
+% of a function application, it may often be more helpful to stop before 
+% beta reduction, to avoid exposing the internals of the applied function to 
+% the client. The semantics in Sec.~\ref{sec:calculus} leaves evaluation order
+% undefined, so various heuristic approaches are possible (e.g. we might wish to avoid beta reduction  for functions from imported libraries).
 
@@
 We leave to future work the exciting possibility of combining these approaches.
-For example, one can imagine running an incomplete program as described in this paper and then adding tests or assertions about the value that a variable should take on in the different hole closures where it appears, via the live context inspector.
+For example, one can imagine running an incomplete program as described in this paper and then adding tests or assertions about the value that a hole should take on under each of the different associated hole closures, via the live context inspector.
 These could serve as  constraints for use by a type-and-example-driven synthesis engine \cite{DBLP:conf/popl/FrankleOWZ16}.
+Relatedly, \emph{prorogued programming} proposes soliciting an external source, e.g.
+the programmer, for a suitable value when a hole is encountered \cite{DBLP:conf/oopsla/AfshariBS12}.
 

### 2018-11-09 examples.tex
formatting tweaks after inserting the ACM copyright stuff

@@
 exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of  ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel (which, as of this writing, implements little more than the language features described in Sec.~\ref{sec:calculus} and the appendix).
-These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented and all of the computations can be expressed using standard ``encoding tricks'' -- the implementations are included in the supplement.
+These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented and all of the computations can be expressed using standard ``encoding tricks''.
 

### 2018-11-09 abstract.tex
tweak abstract to remove citations per the ACM formatting requirements

@@
 \emph{a dynamic semantics for incomplete functional programs}, 
-starting from the static semantics of \citet{popl-paper}. 
+starting from the static semantics for incomplete functional programs developed in our recent work on \Hazelnut. 
 %
@@
 We have also implemented these ideas into the \Hazel programming environment.
-The implementation inserts holes automatically, following the \Hazelnut edit action calculus of \citet{popl-paper}, to guarantee that every editor state has some (possibly incomplete)
+The implementation inserts holes automatically, following the \Hazelnut edit action calculus, to guarantee that every editor state has some (possibly incomplete)
 type.

### 2018-11-10 abstract.tex
format tweaking after adding copyright information

@@
 \begin{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% PAR 1: Live programming environments --- Problem statement
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 Live programming environments aim to provide programmers (and sometimes audiences) 
 with continuous feedback about a program's dynamic behavior as it is being edited. 
-%
 The problem is that programming languages typically assign dynamic meaning only 
@@
 live feedback presented to the programmer exhibits temporal or perceptive gaps.
-% In some cases, these gaps persist over a substantial length of time.
-%
-% Defining a semantics for incomplete functional programs would help bridge the
-% gap between well-formed states and allow live programming environments to
-% provide continuous, uninterrupted, and meaningful feedback.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% PAR 2: This paper's approach: New (dynamic) semantics for incomplete programs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \emph{a dynamic semantics for incomplete functional programs}, 
-starting from the static semantics for incomplete functional programs developed in our recent work on \Hazelnut. 
-%
+starting from the static semantics for incomplete functional programs developed in recent work on \Hazelnut. 
 We model incomplete functional programs as expressions with \emph{holes}, 
@@
 operating as membranes around static and dynamic type inconsistencies. 
-%
 Rather than aborting when evaluation encounters any of these holes as in
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% PAR 3: We have an implementation, and have proofs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% , including 
-% (1) a type safety property that accounts for expression holes (i.e. free metavariables, 
-% from the perspective of CMTT) and cast failure holes (which provide a novel), and (2) a commutativity property that establishes that the
-% ``edit-and-resume'' feature is sound.
-%
-%
 We have also implemented these ideas into the \Hazel programming environment.
@@
 Taken together with this paper's type safety property, the 
-% The 
 result is  
@@
 is truly available without gaps, i.e. for every reachable editor state.
-% i.e. \emph{every edit state is guaranteed to have
-% non-trivial static and dynamic meaning}.
 

### 2018-11-10 calculus.tex
formatting tweaks after inserting the ACM copyright stuff
tweaks to spacing
make section titles all caps again (blah)
most recent acmart and bibliography files
format tweaking after adding copyright information

@@
 \newcommand{\calculusSec}{Hazelnut Live}
-\section{\protect\calculusSec}
+\section{\calculusSec}
 \label{sec:calculus}
@@
 \input{fig-syntax}
-\input{fig-bidirectional-typing}
 
@@
 
+\noindent
 \parahead{Overview} The syntax of the core calculus given in
@@
 Sec.~\ref{sec:agda-mechanization}). Rule names in this section,
-e.g. \rulename{SVar}, correspond to variable names used in the
+e.g. \rulename{SVar}, correspond to variables from the
 mechanization. The \Hazel implementation substantially follows the formal
-specification of \Hazelnut (for the editor component) and \HazelnutLive
-(for the evaluator), and we can formally state a continuity invariant for a
+specification of \Hazelnut (for the editor) and \HazelnutLive
+(for the evaluator). We can formally state a continuity invariant for a
 putative combined calculus (see Sec.~\ref{sec:implementation}).
@@
 \label{sec:external-statics}
+
+\input{fig-bidirectional-typing}
 
@@
 To avoid the need for separate rules for these two cases, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
+The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matching relation.
 %
@@
 %%
-
-The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matching relation.
 
@@
 by precisely characterizing when evaluation \emph{does} stop. In the case
-of complete programs, we also recover the familiar statements of progress
-and preservation for the simply typed lambda calculus.
-
-It is perhaps worth stating at the outset that a dynamic semantics equipped
-with these properties does not simply ``fall out'' from the observations
-made above that (1) empty hole closures correspond to metavariable closures
-from CMTT \cite{Nanevski2008} and (2) casts also arise in gradual type
-theory \cite{DBLP:conf/snapl/SiekVCB15}.
-%
-We say more in Sec.~\ref{sec:relatedWork}.
+of complete programs, we also recover the familiar statements of preservation
+and progress for the simply typed lambda calculus.
+
+% It is perhaps worth stating at the outset that a dynamic semantics equipped
+% with these properties does not simply ``fall out'' from the observations
+% made above that (1) empty hole closures correspond to metavariable closures
+% from CMTT \cite{Nanevski2008} and (2) casts also arise in gradual type
+% theory \cite{DBLP:conf/snapl/SiekVCB15}.
+%
+% We say more in Sec.~\ref{sec:relatedWork}.
 %
@@
 % %
+
+\input{fig-ground-types}
+\input{fig-dynamics-aux}
 
@@
 \subsubsection{Application and Substitution}
+\input{fig-dynamics-contexts}
 %
@@
 the sake of generality, and to support the specification of the fill-and-resume operation~(see \Secref{sec:resumption}).
-
-\input{fig-ground-types}
-\input{fig-dynamics-aux}
 
@@
 further.
-
-\input{fig-dynamics-contexts}
 
@@
 
-The supplemental material includes our Agda
+The archived artifact includes our Agda
 mechanization  \cite{norell2009dependently,norell:thesis,Aydemir:2005fk}
@@
 \newcommand{\commutativitySec}{A Contextual Modal Interpretation of Fill-and-Resume}
-\section{\protect\commutativitySec}
+\section{\commutativitySec}
 \label{sec:resumption}

### 2018-11-10 discussion.tex
make section titles all caps again (blah)
format tweaking after adding copyright information

@@
 \newcommand{\discussionSection}{Conclusion}
-\section{\protect\discussionSection} % don't like the all-caps thing that the template does, so protecting it from that
+\section{\discussionSection} % don't like the all-caps thing that the template does, so protecting it from that
 \label{sec:discussion}
@@
 Weinberg's sentiment applies to countless situations that programmers face as they work to develop and critically investigate the mental model of the program they are writing---at the very moment where rich feedback might be most helpful, e.g.
-when there is an error in the program or when the programmer is unsure how to fill a hole, the computer often has comparatively little feedback to offer (perhaps just a parser error message, or an explanation of one type error).
+when there is an error in the program or when the programmer is unsure how to fill a hole, the computer often has comparatively little feedback to offer (perhaps just a parser error message, or an austere explanation of a type error).
 %
@@
 \newcommand{\acksSection}{Acknowledgments}
-\section*{\protect\acksSection} 
+\section*{\acksSection} 
 \label{sec:acks}
@@
 %
-\clearpage
+% \clearpage
 % \todo{discussion}

### 2018-11-10 examples.tex
tweaks to spacing
format tweaking after adding copyright information

@@
 % !TEX root = hazelnut-dynamics.tex
-%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
-%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
 \newcommand{\examplesSec}{Live Programming in Hazel}
-\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
+\section{\examplesSec} 
 \label{sec:examples}
@@
 The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with a result appearing below each cell that contains an expression, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity.
-\Hazel is intended initially for use by students and instructors in introductory functional programming courses (where \Elm~ has been successful \cite{DBLP:journals/corr/abs-1805-05125}).
+\Hazel is intended initially for use by students and instructors in introductory functional programming courses (where \Elm~ has been successful \cite{DBLP:journals/corr/abs-1805-05125,zhang2018graphics}).
+
 For the sake of 

### 2018-11-10 fig-grades-example.tex
missed a file a few commits ago
format tweaking after adding copyright information

@@
 \centering
-\includegraphics[width=0.8\textwidth,interpolate=false]{images/grades-cell-mockup.png}
-\vspace{-3px}
+\includegraphics[width=\textwidth,interpolate=false]{images/grades-cell-mockup.png}
+\vspace{-10px}
 \caption{Evaluating an incomplete functional program past the first hole}
@@
 \centering
-\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-1.png}
+\includegraphics[width=0.31\textwidth,interpolate=false,valign=c]{images/grades-sidebar-1.png}
 ~${}^\blacktriangleright$
-\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-2.png}
+\includegraphics[width=0.31\textwidth,interpolate=false,valign=c]{images/grades-sidebar-2.png}
 ~${}^\blacktriangleright$
-\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-3.png}
+\includegraphics[width=0.31\textwidth,interpolate=false,valign=c]{images/grades-sidebar-3.png}
 \caption{The live context inspector communicates relevant static \emph{and} dynamic information about variables in scope.}

### 2018-11-10 hazelnut-dynamics.tex
small clarifications through sec 3, addressing R2's confusion about matching relation, and more editor states being meaningful; also changed the description of the mechanization to be consistent with our final artifact
formatting tweaks after inserting the ACM copyright stuff
tweak abstract to remove citations per the ACM formatting requirements
tweaks to spacing
format tweaking after adding copyright information

@@
 \newif\ifarxiv
-\arxivfalse
+\arxivtrue
 \ifarxiv
-\documentclass[acmsmall,nonacm]{acmart}
+\documentclass[acmsmall,screen,nonacm]{acmart}
 \else
 %%% Note: arxiv does not want line numbers (they are detected somehow, and are not allowed)
-\documentclass[acmsmall]{acmart}
+\documentclass[acmsmall,screen]{acmart}
 % \settopmatter{printfolios=false,printccs=false,printacmref=false}
@@
 \else
-\newcommand{\appendixName}{supplemental appendix}
+\newcommand{\appendixName}{extended appendix}
 \fi
@@
 
-\makeatletter\if@ACM@journal\makeatother
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2019}
-\acmMonth{1}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-% \acmConference[]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-
-\acmYear{2019}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
-
 %% Copyright information
@@
 %% see authors.acm.org) by publisher for camera-ready submission
-\setcopyright{none}             %% For review submission
+% \setcopyright{none}             %% For review submission
 %\setcopyright{acmcopyright}
@@
 %\copyrightyear{2017}           %% If different from \acmYear
+\setcopyright{rightsretained} 
+\acmJournal{PACMPL}
+\acmYear{2019} \acmVolume{3} \acmNumber{POPL} \ifarxiv \acmArticle{1} \else \acmArticle{14} \fi \acmMonth{1} \acmPrice{}\acmDOI{10.1145/3290327}
+\copyrightyear{2019}
 
@@
 \title{Live Functional Programming with Typed Holes}         %% [Short Title] is optional;
+\subtitle{Extended Version}
+\subtitlenote{The original version of this article appears as ...}
                                         %% when present, will be used in
@@
 %% comma separated list
-\keywords{live programming, gradual typing, contextual modal type theory, dynamic software updating, typed holes, structured editing}  %% \keywords is optional
+\keywords{live programming, gradual typing, contextual modal type theory, typed holes, structured editing}  %% \keywords is optional
 

### 2018-11-10 intro.tex
tweaks to spacing
format tweaking after adding copyright information

@@
 % !TEX root = hazelnut-dynamics.tex
-\vspace{-5px}
+\ifarxiv \clearpage \fi
 \newcommand{\introSec}{Introduction}
-\section{\protect\introSec} % don't like the all-caps thing that the template does, so protecting it from that
+\section{\introSec} 
 \label{sec:intro}
@@
 % The dynamic semantics of \HazelnutLive is a small-step reduction semantics equipped with clean type safety theorems (including, notably, a Progress theorem for incomplete expressions). 
-Sec.~\ref{sec:agda-mechanization} outlines our Agda-based mechanization of \HazelnutLive, which is included in the supplemental material and is also available from the following URL:
+Sec.~\ref{sec:agda-mechanization} outlines our Agda-based mechanization of \HazelnutLive, which is included in the archived artifact and is also available from the following URL:
 
@@
 It also provides some additional details on the implementation of \Hazel.
-A snapshot of the implementation is included in the supplemental material.
+A snapshot of the implementation is included in the archived artifact.
 An online version of the ongoing implementation of \Hazel is available from \url{hazel.org}, and the source code is available from the following URL:

### 2018-11-10 related-work.tex
make section titles all caps again (blah)
format tweaking after adding copyright information

@@
 % !TEX root = hazelnut-dynamics.tex
-\vspace{-5px}
 \newcommand{\relatedWorkSection}{Related and Future Work}
-\section{\protect\relatedWorkSection} % don't like the all-caps thing that the template does, so protecting it from that
+\section{\relatedWorkSection} % don't like the all-caps thing that the template does, so protecting it from that
 \label{sec:relatedWork}
-\vspace{-2px}
 
@@
 %
-There is also more general work on explicit substitutions, which records all  substitutions, not just substitutions around metavariables, explicitly \cite{Abadi:1991fr,levy1999explicit,Abadi:1990ys}.
+There is also more general work on explicit substitutions, which records all  substitutions, not just substitutions around metavariables, explicitly \cite{Abadi:1991fr,levy1999explicit}.
 \citet{DBLP:journals/corr/abs-1009-2789} have developed a theory of explicit substitutions for CMTT, which, following other work on explicit substitutions and environmental abstract machines \cite{DBLP:journals/tcs/Curien91}, could be useful in implementing \HazelnutLive more efficiently by delaying both standard and contextual substitutions until needed during evaluation.
@@
 Relatedly, \emph{prorogued programming} proposes soliciting an external source, e.g.
-the programmer, for a suitable value when a hole is encountered \cite{DBLP:conf/oopsla/AfshariBS12}.
+the programmer, for a suitable value when a hole instance is encountered \cite{DBLP:conf/oopsla/AfshariBS12}.
 

### 2018-11-12 appendix-defns.tex
started making ArXiV version and updated canonical indet forms statement

@@
 \\
+\substitute{\dexp}{x}{\halam{x}{\htau}{\dexp'}}
+&=&
+\halam{x}{\htau}{\dexp'}\\
 \substitute{\dexp}{x}{\halam{y}{\htau}{\dexp'}}
@@
 \dcastfail{(\substitute{\dexp}{x}{\dexp'})}{\htau_1}{\htau_2}
-\\
+\\[6px]
 \substitute{\dexp}{x}{\cdot}
@@
 \end{lem}
-\begin{proof} By rule induction on the first assumption in each case. The conclusion follows from the definition of substitution in each case. \end{proof}
+% \begin{proof} By rule induction on the first assumption in each case. The conclusion follows from the definition of substitution in each case. \end{proof}
 
@@
       \begin{enumerate}
-        \item[i.]
+        \item
           $\dexp=\halam{x}{\htau_1}{\dexp'}$
           where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$, or
-        \item[ii.]
+        \item
           $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1'}{\htau_2'}}{\tarr{\htau_1}{\htau_2}}$
@@
   and $\isIndet{\dexp}$
-  then either
+  then
   \begin{enumerate}[nolistsep]
     \item 
-      $\dexp=\dehole{u}{\subst}{}$
-      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
+      If $\htau = b$ then either
+        \begin{enumerate}
+          \item $\dexp = \dehole{u}{\subst}{}$ where $\Dbinding{u}{\Gamma}{b} \in \hDelta$ and $\hasType{\hDelta}{\emptyset}{\subst}{\hGamma}$, or
+          \item $\dexp = \dhole{\dexp'}{u}{\subst}{}$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$ and $\isFinal{\dexp'}$ and $\Dbinding{u}{\Gamma}{b} \in \hDelta$ and $\hasType{\hDelta}{\emptyset}{\subst}{\hGamma}$, or
+          \item $\dexp = \dap{\dexp_1}{\dexp_2}$ where $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2}{b}}$ and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2}$ and $\isIndet{\dexp_1}$ and $\isFinal{\dexp_2}$ and $\dexp_1 \neq \dcasttwo{\dexp_1'}{\tarr{\htau_3}{\htau_4}}{\tarr{\htau_3'}{\htau_4'}}$ for any $\dexp_1', \htau_3, \htau_4, \htau_3', \htau_4'$, or
+          \item $\dexp = \dcasttwo{\dexp'}{\tehole}{b}$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\tehole}$ and $\isIndet{\dexp'}$ and $\dexp' \neq \dcasttwo{\dexp''}{\htau'}{\tehole}$ for any $\dexp'', \htau'$, or
+          \item $\dexp = \dcastfail{\dexp'}{\htau'}{b}$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$ and $\isGround{\htau'}$ and $\htau' \neq b$.
+        \end{enumerate}
     \item 
-      $\dexp=\dhole{\dexp'}{u}{\subst}{}$
-      and $\isFinal{\dexp'}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$
-      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
+      If $\htau = \tarr{\htau_1}{\htau_2}$ then either 
+        \begin{enumerate}
+          \item $\dexp = \dehole{u}{\subst}{}$ where $\Dbinding{u}{\Gamma}{\tarr{\htau_1}{\htau_2}} \in \hDelta$ and $\hasType{\hDelta}{\emptyset}{\subst}{\hGamma}$, or
+          \item $\dexp = \dhole{\dexp'}{u}{\subst}{}$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$ and $\isFinal{\dexp'}$ and $\Dbinding{u}{\Gamma}{\tarr{\htau_1}{\htau_2}} \in \hDelta$ and $\hasType{\hDelta}{\emptyset}{\subst}{\hGamma}$, or
+          \item $\dexp = \dap{\dexp_1}{\dexp_2}$ where $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2'}{\tarr{\htau_1}{\htau_2}}}$ and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2'}$ and $\isIndet{\dexp_1}$ and $\isFinal{\dexp_2}$ and $\dexp_1 \neq \dcasttwo{\dexp_1'}{\tarr{\htau_3}{\htau_4}}{\tarr{\htau_3'}{\htau_4'}}$ for any $\dexp_1', \htau_3, \htau_4, \htau_3', \htau_4'$, or
+          \item $\dexp = \dcasttwo{\dexp'}{\tarr{\htau_1'}{\htau_2'}}{\tarr{\htau_1}{\htau_2}}$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\tarr{\htau_1'}{\htau_2'}}$ and $\isIndet{\dexp'}$ and $\tarr{\htau_1'}{\htau_2'} \neq \tarr{\htau_1}{\htau_2}$, or
+          \item $\dexp = \dcasttwo{\dexp'}{\tehole}{\tarr{\tehole}{\tehole}}$ and $\htau_1 = \tehole$ and $\htau_2 = \tehole$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\tehole}$ and $\isIndet{\dexp'}$ and $\dexp' \neq \dcasttwo{\dexp''}{\htau'}{\tehole}$ for any $\dexp'', \htau'$, or
+          \item $\dexp = \dcastfail{\dexp'}{\htau'}{\tarr{\tehole}{\tehole}}$ and $\htau_1 = \tehole$ and $\htau_2 = \tehole$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$ and $\isGround{\htau'}$ and $\htau' \neq \tarr{\tehole}{\tehole}$.
+        \end{enumerate}
     \item 
-      $\dexp=\dap{\dexp_1}{\dexp_2}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2}{\htau}}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2}$
-      and $\isIndet{\dexp_1}$
-      and $\isFinal{\dexp_2}$
-      and $\dexp_1\neq\dcasttwo{\dexp_1}{\tarr{\htau_3}{\htau_4}}
-                                        {\tarr{\htau_3'}{\htau_4'}}$, or
-    %% \item
-    %%   \begin{enumerate}
-    %%     \item blah
-    %%     \item blah
-    %%     \item blah
-    %%   \end{enumerate}
-    \item
-      $\htau=b$
-      and $\dexp=\dcasttwo{\dexp'}{\tehole}{b}$
-      and $\isIndet{\dexp'}$
-      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
-    \item
-      $\htau=b$
-      and $\dexp=\dcastfail{\dexp'}{\htau'}{b}$
-      and $\isGround{\htau'}$
-      and $\htau'\neq{b}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
-    \item
-      $\htau=\tarr{\htau_{11}}{\htau_{12}}$
-      and $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1}{\htau_2}}
-                                  {\tarr{\htau_{11}}{\htau_{12}}}$
-      and $\isIndet{\dexp'}$
-      and $\tarr{\htau_1}{\htau_2}\neq\tarr{\htau_{11}}{\htau_{12}}$, or
-    \item
-      $\htau=\tarr{\tehole}{\tehole}$
-      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
-      %% and $\htau_{11}=\tehole$
-      %% and $\htau_{12}=\tehole$
-      and $\dexp=\dcastthree{\dexp'}{\tehole}{\tehole}{\tehole}$
-      and $\isIndet{\dexp'}$
-      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
-    \item
-      $\htau=\tarr{\tehole}{\tehole}$
-      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
-      %% and $\htau_{11}=\tehole$
-      %% and $\htau_{12}=\tehole$
-      and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\tehole}{\tehole}}$
-      %% and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\htau_{11}}{\htau_{12}}}$
-      and $\htau'\neq\htau$
-      and $\isGround{\htau'}$
-      and $\isIndet{\dexp'}$
-      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
-    \item
-      $\htau=\tehole$
-      and $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
-      and $\isGround{\htau'}$
-      and $\isIndet{\dexp'}$.
+      If $\htau = \tehole$ then either 
+        \begin{enumerate}
+          \item $\dexp = \dehole{u}{\subst}{}$ where $\Dbinding{u}{\Gamma}{\tehole} \in \hDelta$ and $\hasType{\hDelta}{\emptyset}{\subst}{\hGamma}$, or
+          \item $\dexp = \dhole{\dexp'}{u}{\subst}{}$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$ and $\isFinal{\dexp'}$ and $\Dbinding{u}{\Gamma}{\tehole} \in \hDelta$ and $\hasType{\hDelta}{\emptyset}{\subst}{\hGamma}$, or
+          \item $\dexp = \dap{\dexp_1}{\dexp_2}$ and $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2}{\tehole}}$ and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2}$ and $\isIndet{\dexp_1}$ and $\isFinal{\dexp_2}$ and $\dexp_1 \neq \dcasttwo{\dexp_1'}{\tarr{\htau_3}{\htau_4}}{\tarr{\htau_3'}{\htau_4'}}$ for any $\dexp_1', \htau_3, \htau_4, \htau_3', \htau_4'$, or
+          \item $\dexp = \dcasttwo{\dexp'}{\htau'}{\tehole}$ where $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$ and $\isGround{\htau'}$ and $\isIndet{\dexp'}$.
         \end{enumerate}
+  \end{enumerate}
+  % \begin{enumerate}[nolistsep]
+  %   \item
+  %     $\dexp=\dehole{u}{\subst}{}$
+  %     and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
+  %   \item
+  %     $\dexp=\dhole{\dexp'}{u}{\subst}{}$
+  %     and $\isFinal{\dexp'}$
+  %     and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$
+  %     and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
+  %   \item
+  %     $\dexp=\dap{\dexp_1}{\dexp_2}$
+  %     and $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2}{\htau}}$
+  %     and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2}$
+  %     and $\isIndet{\dexp_1}$
+  %     and $\isFinal{\dexp_2}$
+  %     and $\dexp_1\neq\dcasttwo{\dexp_1}{\tarr{\htau_3}{\htau_4}}
+  %                                       {\tarr{\htau_3'}{\htau_4'}}$, or
+  %   %% \item
+  %   %%   \begin{enumerate}
+  %   %%     \item blah
+  %   %%     \item blah
+  %   %%     \item blah
+  %   %%   \end{enumerate}
+  %   \item
+  %     $\htau=b$
+  %     and $\dexp=\dcasttwo{\dexp'}{\tehole}{b}$
+  %     and $\isIndet{\dexp'}$
+  %     and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
+  %   \item
+  %     $\htau=b$
+  %     and $\dexp=\dcastfail{\dexp'}{\htau'}{b}$
+  %     and $\isGround{\htau'}$
+  %     and $\htau'\neq{b}$
+  %     and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
+  %   \item
+  %     $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+  %     and $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1}{\htau_2}}
+  %                                 {\tarr{\htau_{11}}{\htau_{12}}}$
+  %     and $\isIndet{\dexp'}$
+  %     and $\tarr{\htau_1}{\htau_2}\neq\tarr{\htau_{11}}{\htau_{12}}$, or
+  %   \item
+  %     $\htau=\tarr{\tehole}{\tehole}$
+  %     %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+  %     %% and $\htau_{11}=\tehole$
+  %     %% and $\htau_{12}=\tehole$
+  %     and $\dexp=\dcastthree{\dexp'}{\tehole}{\tehole}{\tehole}$
+  %     and $\isIndet{\dexp'}$
+  %     and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
+  %   \item
+  %     $\htau=\tarr{\tehole}{\tehole}$
+  %     %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
+  %     %% and $\htau_{11}=\tehole$
+  %     %% and $\htau_{12}=\tehole$
+  %     and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\tehole}{\tehole}}$
+  %     %% and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\htau_{11}}{\htau_{12}}}$
+  %     and $\htau'\neq\htau$
+  %     and $\isGround{\htau'}$
+  %     and $\isIndet{\dexp'}$
+  %     and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
+  %   \item
+  %     $\htau=\tehole$
+  %     and $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
+  %     and $\isGround{\htau'}$
+  %     and $\isIndet{\dexp'}$.
+  % \end{enumerate}
 \end{lem}
 
-The proofs for all three of these theorems follow by straightforward rule induction.
+% The proofs for all three of these theorems follow by straightforward rule induction.
 
@@
 
-\clearpage
 \subsection{Complete Programs}
@@
 We proceed by cases on the instruction transition assumption (no induction is needed).
-For Rule \rulename{ITBeta}, we defer to the Substitution Commutativity lemma above.
+For Rule \rulename{ITLam}, we defer to the Substitution Commutativity lemma above.
 For the remaining cases, the conclusion follows from the definition of hole filling.

### 2018-11-12 intro.tex
started making ArXiV version and updated canonical indet forms statement

@@
 \fi
-(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for the sake of space; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief discussion on defining other extensions (in part by by following the ``gradualization'' approach of \citet{DBLP:conf/popl/CiminiS16}).
+(1) provides some straightforward auxiliary definitions and proofs that were omitted from the \ifarxiv original \fi paper  for the sake of space; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief discussion on defining other extensions (in part by by following the ``gradualization'' approach of \citet{DBLP:conf/popl/CiminiS16}).
 % These are not necessary to communicate the fundamental ideas of the paper, but may be of interest to some readers.

### 2018-11-13 appendix-defns.tex
tweaks to text in proof of commutativity + fixes to num extension

@@
 
-\clearpage
 \subsection{Hole Filling}\label{sec:hole-filling}
@@
 
-We need the following auxiliary definitions, which lift hole filling to evaluation contexts taking care to consider the special situation where the mark is inside the hole that is being filled, to prove the Commutativity theorem.
+To prove the Commutativity theorem, we need the auxiliary definitions in Fig.~\ref{fig:evalctx-filling}, which lift hole filling to evaluation contexts taking care to consider the special situation where the mark is inside the hole that is being filled.
 \input{fig-evalctx-instantiation}
+
+We also need the following lemmas, which characterize how hole filling interacts with substitution and instruction transitions. 
 \begin{lem}[Substitution Commutativity]
@@
 
-We exclude these proofs and definitions from the Agda mechanization
-for two reasons.
-%
-First, fill-and-resume is merely an optimization, and unlike the meta
-theory of \Secref{sec:calculus}, these properties are generally not
-conserved by certain reasonable extensions of the core
-calculus~(e.g., reference cells and other non-commuting effects).
-%
-Second, to properly encode the hole filling operation, such a
-mechanization requires a significantly more complex representation of
-hole environments; unfortunately, Agda cannot be easily convinced that
-the definition of hole filling is well-founded (\citet{Nanevski2008}
-establish that it is in fact well-founded).
-%
-By contrast, the developments in \Secref{sec:calculus} do not require
-these more complex (and somewhat problematic) representations.
+% We excluded these proofs and definitions from the Agda mechanization
+% for two reasons.
+% %
+% First, fill-and-resume is merely an optimization, and unlike the meta
+% theory of \Secref{sec:calculus}, these properties are generally not
+% conserved by certain reasonable extensions of the core
+% calculus~(e.g., reference cells and other non-commuting effects).
+% %
+% Second, to properly encode the hole filling operation, such a
+% mechanization requires a more complex representation of
+% hole environments; unfortunately, Agda cannot be easily convinced that
+% the definition of hole filling is well-founded (\citet{Nanevski2008}
+% establish that it is in fact well-founded).
+% %
+% By contrast, the developments in \Secref{sec:calculus} do not require
+% these more complex representations.
 
-\subsection{Confluence, and Friends}\label{sec:confluence}
+\subsection{Confluence and Resumption}\label{sec:confluence}
 There are various ways to encode the intuition that ``evaluation order does not matter''. One way to do so is
@@
 The most general confluence property does not hold for the dynamic
-semantics in Sec.~\ref{sec:calculus} for the usual reason: We do not
+semantics in Sec.~\ref{sec:calculus} for the usual reason: we do not
 reduce under binders (\citet{DBLP:conf/birthday/BlancLM05} discuss the

### 2018-11-13 extensions.tex
tweaks to text in proof of commutativity + fixes to num extension

@@
 % !TEX root = hazelnut-dynamics.tex
-\clearpage
 \newcommand{\extensionsSec}{Extensions to Hazelnut Live}
@@
 The ``non-obvious'' machinery is almost entirely related to casts.
-Fortunately, there has been excellent work of late on generating ``gradualized'' specifications from standard specifications \cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/CiminiS17}.
+Fortunately, there has been progress on the problem of generating ``gradualized'' specifications from standard specifications \cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/CiminiS17}.
 The extensions below, and other extensions of interest, closely follow the output of the gradualizer: \url{http://cimini.info/gradualizerDynamicSemantics/} (which, like our work, is based on the refined account of gradual typing by \cite{DBLP:conf/snapl/SiekVCB15}).

### 2018-11-13 calculus.tex
add more specific references to sections of the appendix behind ifarxiv flag

@@
 That paper also defined sum types as an extension to the core calculus.
-We follow suit on both counts in the \appendixName.%Appendix~\ref{sec:extensions}.
+We follow suit on both counts in \ifarxiv Appendix \ref{sec:extensions}\else the \appendixName\fi.%Appendix~\ref{sec:extensions}.
 %
@@
 
-Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the \appendixName~for the full definition).
+Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see \ifarxiv Appendix \ref{sec:substitution} \else the \appendixName~\fi for the full definition).
 % Appendix~\ref{sec:additional-defns}).
@@
 The proof relies on an analogous preservation lemma for instruction
-transitions and a standard substitution lemma stated in
-the \appendixName.
+transitions and a standard substitution lemma stated in \ifarxiv Appendix \ref{sec:substitution}\else the \appendixName\fi.
 % Appendix~\ref{sec:additional-defns}.
@@
 %
-The proof relies on canonical forms lemmas stated in the \appendixName.
+The proof relies on canonical forms lemmas stated in \ifarxiv Appendix~\ref{sec:canonical-forms}\else the \appendixName\fi.
 % Appendix~\ref{sec:additional-defns}.
@@
 % Appendix~\ref{sec:additional-defns}
-The \appendixName{} defines the predicates~$\isComplete{e}$,
-$\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
+\ifarxiv Appendix~\ref{sec:complete-programs} \else The \appendixName{} \fi defines the predicates~$\isComplete{\htau}$,
+$\isComplete{\hexp}$, $\isComplete{\dexp}$ and~$\isComplete{\hGamma}$.
 %
@@
 in this section, with some simple extensions. As with many full-scale systems, there is not currently a formal
-specification for the full \Hazel language, but the \appendixName~
+specification for the full \Hazel language, but \ifarxiv Appendix~\ref{sec:extensions} \else the \appendixName~\fi 
 discusses how the standard approach for deriving a ``gradualized'' version of a
@@
 We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
-transitive closure of stepping (see the \appendixName).
+transitive closure of stepping (see \ifarxiv Appendix~\ref{sec:multi-step}\else the \appendixName\fi).
 % Appendix~\ref{sec:additional-defns}).
@@
 by stating a suitable confluence property.
-For the sake of space, we review confluence in the \appendixName.} Languages with non-commutative effects do not enjoy this property.
+For the sake of space, we review confluence in \ifarxiv Appendix \ref{sec:confluence}\else the \appendixName\fi.} Languages with non-commutative effects do not enjoy this property.
 % Appendix~\ref{sec:confluence}.)
@@
 We describe the proof, which is straightforward but involves a
-number of lemmas and definitions, in the \appendixName.
+number of lemmas and definitions, in \ifarxiv Appendix~\ref{sec:hole-filling}\else the \appendixName\fi.
 %Appendix~\ref{sec:hole-filling}.

### 2018-11-13 examples.tex
add more specific references to sections of the appendix behind ifarxiv flag

@@
 For the sake of 
-exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of  ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel (which, as of this writing, implements little more than the language features described in Sec.~\ref{sec:calculus} and the appendix).
+exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of  ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel (which, as of this writing, implements little more than the language features described in Sec.~\ref{sec:calculus} and \ifarxiv Appendix \ref{sec:extensions}\else the appendix\fi).
 These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented and all of the computations can be expressed using standard ``encoding tricks''.

### 2018-11-13 fig-complete.tex
added formal statement of every cast being an identity cast in a well-typed and complete internal expression

@@
 
-\begin{lem}[Complete Consistency] If $\tconsistent{\htau_1}{\htau_2}$ and $\isComplete{\htau_1}$ and $\isComplete{\htau_2}$ then $\htau_1 = \htau_2$.
+\begin{lem}[Complete Consistency]\label{lem:complete-consistency} If $\tconsistent{\htau_1}{\htau_2}$ and $\isComplete{\htau_1}$ and $\isComplete{\htau_2}$ then $\htau_1 = \htau_2$.
 \end{lem}
@@
 an identity cast.
+
+\begin{lem}[Complete Casts] If $\hasType{\hGamma}{\hDelta}{\dcasttwo{\dexp}{\htau_1}{\htau_2}}$ and $\isComplete{\dcasttwo{\dexp}{\htau_1}{\htau_2}}$ then $\htau_1 = \htau_2$. \end{lem}
+\begin{proof} By straightforward rule induction and Lemma~\ref{lem:complete-consistency}. \end{proof}

### 2018-11-13 fig-complete.tex
fixed typo in statement of complete casts

@@
 
-\begin{lem}[Complete Casts] If $\hasType{\hGamma}{\hDelta}{\dcasttwo{\dexp}{\htau_1}{\htau_2}}$ and $\isComplete{\dcasttwo{\dexp}{\htau_1}{\htau_2}}$ then $\htau_1 = \htau_2$. \end{lem}
+\begin{lem}[Complete Casts] If $\hasType{\hGamma}{\hDelta}{\dcasttwo{\dexp}{\htau_1}{\htau_2}}{\htau_2}$ and $\isComplete{\dcasttwo{\dexp}{\htau_1}{\htau_2}}$ then $\htau_1 = \htau_2$. \end{lem}
 \begin{proof} By straightforward rule induction and Lemma~\ref{lem:complete-consistency}. \end{proof}

### 2018-11-20 hazelnut-dynamics.tex
started making ArXiV version and updated canonical indet forms statement
add more specific references to sections of the appendix behind ifarxiv flag
added country to author affiliations per ACM request + fixed a spelling mistake in related work (both submitted in final version)

@@
 \arxivtrue
+% \arxivfalse
 \ifarxiv
@@
 
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission
-% \setcopyright{none}             %% For review submission
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2017}           %% If different from \acmYear
+\ifarxiv
 \setcopyright{rightsretained} 
@@
 \copyrightyear{2019}
+\else
+%%% The following is specific to POPL '19 and the paper
+%%% 'Live Functional Programming with Typed Holes'
+%%% by Cyrus Omar, Ian Voysey, Ravi Chugh, and Matthew A. Hammer.
+%%%
+\setcopyright{rightsretained}
+\acmPrice{}
+\acmDOI{10.1145/3290327}
+\acmYear{2019}
+\copyrightyear{2019}
+\acmJournal{PACMPL}
+\acmVolume{3}
+\acmNumber{POPL}
+\acmArticle{14}
+\acmMonth{1}
+\fi
 
@@
 \title{Live Functional Programming with Typed Holes}         %% [Short Title] is optional;
+\ifarxiv
 \subtitle{Extended Version}
-\subtitlenote{The original version of this article appears as ...}
+\subtitlenote{The original version of this article was published in the POPL 2019 edition of PACMPL \cite{HazelnutLive}. This extended version includes an additional appendix.}
+\fi
                                         %% when present, will be used in
+
                                         %% header instead of Full Title.
@@
   % \department{Department1}              %% \department is recommended
-  \institution{University of Chicago}            %% \institution is required
+  \institution{University of Chicago, USA}            %% \institution is required
   % \streetaddress{Street1 Address1}
@@
   % \department{Department1}              %% \department is recommended
-  \institution{Carnegie Mellon University}            %% \institution is required
+  \institution{Carnegie Mellon University, USA}            %% \institution is required
   % \streetaddress{Street1 Address1}
@@
   % \department{Department1}              %% \department is recommended
-  \institution{University of Colorado Boulder}            %% \institution is required
+  \institution{University of Colorado Boulder, USA}            %% \institution is required
   % \streetaddress{Street1 Address1}
@@
 %\clearpage
-\bibliography{references,all.short,hazel_NSF}
+\bibliography{all.short}
 

### 2018-11-20 related-work.tex
added country to author affiliations per ACM request + fixed a spelling mistake in related work (both submitted in final version)

@@
 %
-A program with expression holes can be interpretted as being bound under a number of these $\mathsf{letbox}$ constructs, i.e.
+A program with expression holes can be interpreted as being bound under a number of these $\mathsf{letbox}$ constructs, i.e.
 CMTT serves as a ``{development calculus}'' \cite{DBLP:phd/ethos/McBride00}.
@@
 %
-Each applied hole filling can be interpretted as inducing a new dynamic
+Each applied hole filling can be interpreted as inducing a new dynamic
 \emph{edit stage}.
